### 2013-01-04 documents/report_1.Rnw
initial commit. restarted analysis to be more elegant

\documentclass{article}
\frenchspacing

<<set-options, echo=FALSE, cache=FALSE, message=FALSE>>=
require(knitr)
require(formatR)
options(keep.comment = FALSE)
read_chunk('initial_outline.R')
read_chunk('landmark_montecarlo.R')
@

\SweaveOpts{external = TRUE, size='scriptsize'}

\title{Biological shape as a weighted network}
\author{Peter D Smits \and Ben Frable}

\begin{document}
\maketitle

\section{Brief Introduction}
Here we envision biological shape, and the differences between shapes, as a network or graph model.
A network model is the depiction of the relationships between many nodes.
Each of these nodes may represent many different items or subjects of the same class.
In, for example, a social network is where each node represents a single person.
In this case, nodes represent a unique biological shape.

Between these nodes are links or edges.
In the social network example, an edge would represent a ``friendship'' or some other social connection between two individuals.
It is possible for edges to have weights, representing the degree of connectedness between two nodes.
Unlike in a distance matrix, a higher value represents a stronger connectedness.
Here, our edges represent the similarity of two shape configurations.
How these edges are assigned and weighted, however, is a via an extended procedure.

In network theory, a community is a densly connected set of nodes that are more connected to each other than to other groups.
In a given network, the number o groups may range between one large group and as many groups as there are nodes.
The shape network made from the matrix of \(\rho\) can then be analyzed using the various community detection algorithms in order to find if there are any subgroups of morphologies that are more densly connected.
These subgroups represent highly similar morphologies, and different communities/groups represent large scale differences in morphology.
Effectively, these community detection algorithms are defining different morphological categories directly from the morphological similarities as opposed to using prior knowledge or observation.

\section{Methodology}
The core of this entire process lies with the Riemannian shape distance, \(\rho\), between two configurations. 
A shape configuration is the set of landmarks that represent a unique shape. 
There are a technically infinite number of unique shapes, and our sample in geometric morphometric studies only represents a subset of the biologically known or possible shapes.

Shapes, when properly scaled and fit via some form of Procrustes analysis, exist not in a Euclidean plane, but on a Riemannian manifold.
Convenitnetly, a Riemannian manifold is metrizable and we can define some kind of metric to represent the ``distance'' between two shape configurations.
This Riemannian shape distance, \(\rho\) was defined by Kendall CITATION and varies between 0 and \(\pi /2\).
A value closer to 0 means that two shapes are closer in shape space, while a higher value indicates that two shapes are farther away in shape space.
Conveniently here is a function for calculating \(\rho\) in the {\tt{shapes}} package CITATION called {\tt{riemdist()}}.

The first step is the using generalized Procrustes analysis (GPA) to align all of our shapes and remove all non-shape elements.
After this a pairwise distance matrix is calculated for all shape combinations using {\tt{riemdist()}}. This produces a square \(n \cdot n\) matrix, where \(n\) is the number of unique shapes.
This pairwise \(\rho\) matrix represents how close every shape is to all other shapes.
From here, we want to determine which shapes are closer to each other than we would expect by random.
We require a Monte Carlo process to determine what ``random'' is.

How we define ``by random'' is very important here. Here, we are interested in a value of \(\rho\) smaller than than we would expect half of the time by random. This means, that we want to find which values of \(\rho\) are smaller than the median of a null distribution.

\subsection{Monte Carlo process}
The null distribution here is created from using two randomly selected shape configurations, moving the landmarks slightly to create two new, most likely valid, shape configuration.
\(\rho\) is then calculated for these two new shapes and stored.
This process is repeated many times in order to approximate a distribution of possible values of \(\rho\).
After determining this null distribution, all values greater than the median of this null distribution are removed.
The inverse of all remaining pairwise comparisons are then calculated, giving the future edge weights.
The resulting matrix is an undirected, weighted adjacency matrix which can easily be converted into a network for use in the {\tt{igraph}} package CITATION.

I've implemented this Monte Carlo process as a few functions in R.
The core function is {\tt{jitter.landmark()}}.
<<jitter>>=
@

This functions are called {\tt{nsim}} times (deafult 1000) by the {\tt{riem.distribution()}} function.
Using the original data set, {\tt{jitter.landmark()}} is called on {\tt{nsim}} two shape combinations.
A lambda function is used to calculate the reimanian distance for all of the {\tt{nsim}} combinations.
Currently, the lambda function executed as part of a call to {\tt{mclapply()}} which is a parallel function from the {\tt{parallel}} core package.
Currently, it will use the maximum number of cores avaliable and the seed is optimzed for reproducibility.
Without this feature, the process will run much slower.
This function will not work on non-POSIX systems (i.e. Windows).
<<riem-dist>>=
@

% All of the above functions can probably made much simpler.
% {\tt{jitter.landmark()}} can remain seperate while the entierty of {\tt{fit.riem()}} can be made a lambda function in the {\tt{mclapply()}} call in {\tt{riem.distribution()}}

The interface function {\tt{shape.simulation()}} outputs a quantile value necessary for determining our ``at random.'' The deafult quantile of 0.50 is the same as the median of a sample.
<<shape-sim>>=
@

\subsection{Community detection}

There are currently eight different community detection algorithms implemented as part of the {\tt{igraph}} package.
Each of these algorithms try to optimze different communities based on various criteria.
For example, the ``walktrap'' algorithm determines communties by which sections of a graph a random walk is most frequently.
This algorithm, however, requires that there are no unconnected nodes, or nodes that have no edges.
Currently, I have tested the ``edge betweenness,'' ``fast-greedy,'' and ``leading eigenvector'' algorithms.
Each gives slightly different results, reflecting the method by which they determine couuminity presence and structure.

Implementation of these algorithms is illustrated below with gorilla and fish examples.
Additionally, I've included heirarchical clustering results from the pariwise \(\rho\) matrix for both of these examples.

\section{Examples}
\subsection{Gorilla vertebrae}
TODO

\subsection{Fish example}
The first step is to import the correct libraries and data files.
The functions described above are part of my script file, while the {\tt{readtps.R}} file is my fork of Ananda Mahto's original script GIST LINK.
This fork improves integration of .tps files and the {\tt{shapes}} package.
Additionally, becreporause this is a randomization process, for reproducibiliity it is necessary to manually set the seed value. Additionally, the RNG kind is changed to the L'Ecuyer algorithm CITATION which is the best for random number generationalgorithm implemented in R for parallel processes.
<<fish-set-up, message=FALSE>>=
@

Next, we use GPA to fit our entire sample of \Sexpr{dim(fish)[3]} fish. We also calculate the pairwise \(\rho\) matrix. This is actually really dirty code that can probably be implemented as a single {\tt{outer}} call, but I've yet to figure it out.

<<fish-fit, eval=TRUE, cache=TRUE>>=
@

From this fit, we should examine our data to better understand what is going on initially. The basic ways are looking at the fitted landmarks (Figure \ref{fish-plotshapes}), the eigenscores from pricipal components analysis (Figure \ref{fish-pca}), and the thin-plate spine plot of an example fish to the mean shape (Figure \ref{fish-tps}.

<<fish-ex1, echo=FALSE, eval=FALSE, message=FALSE, fig.height= 4, results='hide'>>=
require(ggplot2)
# turn fish.fit$rotated into a two column matrix
fish.mean <- as.data.frame(fish.fit$mshape)
fish.fix <- fish.fit$rotated[, , -42]
fish.gg <- fish.fit$rotated[, , 1]
for(ii in 2:(dim(fish.fix)[3])) {
  fish.gg <- rbind(fish.gg, fish.fix[, , ii])
}

fish.gg <- as.data.frame(fish.gg)
fs <- ggplot(fish.gg, aes(x= V1, y= V2)) + geom_point(size = 1)
fs <- fs + scale_x_continuous(name = "X")
fs <- fs + scale_y_continuous(name = "Y")
fs <- fs + opts(panel.background = theme_rect(fill = "transparent",colour = NA), 
                panel.grid.minor = theme_blank(), 
                panel.grid.major = theme_blank(),
                axis.line = theme_segment(),
                axis.text.x = theme_text(colour = "black", size = 15),
                axis.text.y = theme_text(colour = "black", size = 15)) 
fs
@

<<fish-ex2, echo=FALSE, eval=FALSE, message=FALSE, fig.height= 4, results='hide'>>=
fish.eigsc <- as.data.frame(fish.fit$stdscores)
fam <- c(rep("Ano", times= length(ano$ID)), rep("Cur", times= length(cur$ID)))
fish.eigsc$fam <- fam
fish.eigsc <- fish.eigsc[-42, ]
fe <- ggplot(fish.eigsc, aes(x= PC1, y= PC2, colour= fam))
fe <- fe + geom_point(size = 3)
fe <- fe + scale_x_continuous(name = paste("PC1; ", 
                                           format(fish.fit$percent[1], digits= 3),
                                           sep= ''))
fe <- fe + scale_y_continuous(name = paste("PC2; ", 
                                           format(fish.fit$percent[2], digits= 3),
                                           sep= ''))
fe <- fe + opts(panel.background = theme_rect(fill = "transparent",colour = NA), 
                panel.grid.minor = theme_blank(), 
                panel.grid.major = theme_blank(),
                  axis.line = theme_segment(),
                axis.text.x = theme_text(colour = "black", size = 15),
                axis.text.y = theme_text(colour = "black", size = 15))
fe
@

<<fish-ex3, echo=FALSE, eval=FALSE, message=FALSE, fig.height= 4, results='hide'>>=
shapepca(fish.fit, joinline=c(1:35, 1))
@

<<fish-ex4, echo=FALSE, eval=FALSE, message=FALSE, fig.height= 4, results='hide'>>=
par(mfrow = c(1, 2))
tpsgrid(fish.fit$rotated[, , 1], fish.fit$mshape)
tpsgrid(fish.fit$rotated[, , 19], fish.fit$mshape)
@

\begin{figure}
\includegraphics[height=0.3\textheight]{figure/fish-ex1}
\caption{Landmark positions of \Sexpr{dim(fish.fit$rotated)[3] - 1} of the fish taxa. One specimen is excluded.\label{fish-plotshapes}}
\end{figure}

\begin{figure}
\includegraphics[height=0.3\textheight]{figure/fish-ex2}
\caption{PC1 versus PC2 for fish. Families are labeled. Percent of variation explained is included in axes labels. One specimen (42) is excluded.\label{fish-pca}}
\end{figure}

% \begin{figure}
% \includegraphics[width=\textwidth]{figure/fish-ex3}
% \caption{Alternative method for interpreting the principal components analysis.}
% \end{figure}

\begin{figure}
\includegraphics[height=0.35\textheight]{figure/fish-ex4}
\caption{TPS grids for Ano (right) and Cur (left).\label{fish-tps}}
\end{figure}

Following this, the Monte Carlo process is run so that we can begin to analyze the structure of our shape relationships.
This process is very long and takes many hours.
Those values which are greater than the median of the Monte Carlo distribution are discarted and the inverse of the remaining ones is calculated, producing a weighted adacency matrix.

<<fish-MC, eval=FALSE>>=
@

The weighted adacency matrix is then transformed into a network and the three community detection algorithms are then used to analyze the resulting graph.

<<fish-network, eval=FALSE>>=
@

We can then examine the results of these community detection algorithms. 
The simplest manner is the just view the graphic output.

<<fish-plots, eval=FALSE, echo=FALSE>>=
@

In general, there are two identified communities.
These communities are not, however, identical to the two known families.
There is a group of ``Cur'' fish that are most morphologically similar to ``Ano'' fish.

The next step after this is determining if this graph is actually interesting.
This requires more reading.

\section{Concerns and Future Development}
\begin{enumerate}
\item I'm unsure how valid it is to define ``by random'' in the way that we did. 
I would need to actually talk to someone or read a lot to answer this one.

\item Need to make sure I'm not making invalid shape configurations

\item Unsure if I need to do {\tt{procGPA()}} every time in the MC process.

\item Choice of community detection algorithm.
This one just requires a lot of reading I think to figure out how exactly they differ.

\item Are the edge weights to high?
Should they be rescaled

\item MC approach to understanding network.
{\tt{statnet}} packages.
Is the network different from random?

\item Model based clustering possibilities.
Can we classify without a network?
Would use the {\tt{mclust}} package CITATION, but there are awkward philosophical questions that come up with using that methodology.

\item Need to do some machine learning reading.
This project is essentially unsupervised clustering, but doing it in a weird way.

\end{enumerate}

\newpage
For reproducibility information.
<<reproducibility>>=
sessionInfo()
@

\end{document}

### 2013-01-04 documents/report_1.tex
initial commit. restarted analysis to be more elegant

--- 2013-01-04 documents/report_1.Rnw
@@
-\documentclass{article}
+\documentclass{article}\usepackage{graphicx, color}
+%% maxwidth is the original width if it is less than linewidth
+%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
+\makeatletter
+\def\maxwidth{ %
+  \ifdim\Gin@nat@width>\linewidth
+    \linewidth
+  \else
+    \Gin@nat@width
+  \fi
+}
+\makeatother
+
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+\definecolor{fgcolor}{rgb}{0.2, 0.2, 0.2}
+\newcommand{\hlnumber}[1]{\textcolor[rgb]{0,0,0}{#1}}%
+\newcommand{\hlfunctioncall}[1]{\textcolor[rgb]{0.501960784313725,0,0.329411764705882}{\textbf{#1}}}%
+\newcommand{\hlstring}[1]{\textcolor[rgb]{0.6,0.6,1}{#1}}%
+\newcommand{\hlkeyword}[1]{\textcolor[rgb]{0,0,0}{\textbf{#1}}}%
+\newcommand{\hlargument}[1]{\textcolor[rgb]{0.690196078431373,0.250980392156863,0.0196078431372549}{#1}}%
+\newcommand{\hlcomment}[1]{\textcolor[rgb]{0.180392156862745,0.6,0.341176470588235}{#1}}%
+\newcommand{\hlroxygencomment}[1]{\textcolor[rgb]{0.43921568627451,0.47843137254902,0.701960784313725}{#1}}%
+\newcommand{\hlformalargs}[1]{\textcolor[rgb]{0.690196078431373,0.250980392156863,0.0196078431372549}{#1}}%
+\newcommand{\hleqformalargs}[1]{\textcolor[rgb]{0.690196078431373,0.250980392156863,0.0196078431372549}{#1}}%
+\newcommand{\hlassignement}[1]{\textcolor[rgb]{0,0,0}{\textbf{#1}}}%
+\newcommand{\hlpackage}[1]{\textcolor[rgb]{0.588235294117647,0.709803921568627,0.145098039215686}{#1}}%
+\newcommand{\hlslot}[1]{\textit{#1}}%
+\newcommand{\hlsymbol}[1]{\textcolor[rgb]{0,0,0}{#1}}%
+\newcommand{\hlprompt}[1]{\textcolor[rgb]{0.2,0.2,0.2}{#1}}%
+
+\usepackage{framed}
+\makeatletter
+\newenvironment{kframe}{%
+ \def\at@end@of@kframe{}%
+ \ifinner\ifhmode%
+  \def\at@end@of@kframe{\end{minipage}}%
+  \begin{minipage}{\columnwidth}%
+ \fi\fi%
+ \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
+ \colorbox{shadecolor}{##1}\hskip-\fboxsep
+     % There is no \\@totalrightmargin, so:
+     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
+ \MakeFramed {\advance\hsize-\width
+   \@totalleftmargin\z@ \linewidth\hsize
+   \@setminipage}}%
+ {\par\unskip\endMakeFramed%
+ \at@end@of@kframe}
+\makeatother
+
+\definecolor{shadecolor}{rgb}{.97, .97, .97}
+\definecolor{messagecolor}{rgb}{0, 0, 0}
+\definecolor{warningcolor}{rgb}{1, 0, 1}
+\definecolor{errorcolor}{rgb}{1, 0, 0}
+\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
+
+\usepackage{alltt}
 \frenchspacing
-
-<<set-options, echo=FALSE, cache=FALSE, message=FALSE>>=
-require(knitr)
-require(formatR)
-options(keep.comment = FALSE)
-read_chunk('initial_outline.R')
-read_chunk('landmark_montecarlo.R')
-@
-
-\SweaveOpts{external = TRUE, size='scriptsize'}
 
@@
 The core function is {\tt{jitter.landmark()}}.
-<<jitter>>=
-@
+\begin{knitrout}\scriptsize
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+jitter.landmark <- \hlfunctioncall{function}(land) \{
+    \hlfunctioncall{for} (ii in \hlfunctioncall{seq}(\hlfunctioncall{ncol}(land))) \{
+        land[, ii] <- land[, ii] + \hlfunctioncall{runif}(n = \hlfunctioncall{length}(land[, ii]), min = -(\hlfunctioncall{min}(\hlfunctioncall{abs}(\hlfunctioncall{diff}(land[, 
+            ii])))), max = \hlfunctioncall{min}(\hlfunctioncall{abs}(\hlfunctioncall{diff}(land[, ii]))))
+    \}
+    land
+\}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
 
@@
 This function will not work on non-POSIX systems (i.e. Windows).
-<<riem-dist>>=
-@
+\begin{knitrout}\scriptsize
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+riem.distribution <- \hlfunctioncall{function}(fit.land, nsim = 1000) \{
+    ran.land <- \hlfunctioncall{vector}(mode = \hlstring{"list"}, length = nsim)
+    \hlfunctioncall{for} (ii in \hlfunctioncall{seq}(nsim)) \{
+        s <- \hlfunctioncall{sample}(fit.land$n, 2)
+        sh1 <- \hlfunctioncall{jitter.landmark}(fit.land$rotated[, , s[1]])
+        sh2 <- \hlfunctioncall{jitter.landmark}(fit.land$rotated[, , s[2]])
+        ran.land[[ii]] <- \hlfunctioncall{abind}(sh1, sh2, along = 3)
+    \}
+    res <- \hlfunctioncall{mclapply}(ran.land, \hlfunctioncall{function}(x) \{
+        \hlfunctioncall{riemdis}(x[, , 1], x[, , 2])
+    \}, mc.cores = \hlfunctioncall{detectCores}(), mc.set.seed = T)
+    res
+\}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
 
@@
 The interface function {\tt{shape.simulation()}} outputs a quantile value necessary for determining our ``at random.'' The deafult quantile of 0.50 is the same as the median of a sample.
-<<shape-sim>>=
-@
+\begin{knitrout}\scriptsize
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+shape.simulate <- \hlfunctioncall{function}(fit.land, nsim = 1000, probs = 0.5) \{
+    ri.di <- \hlfunctioncall{riem.distribution}(fit.land, nsim = nsim)
+    \hlfunctioncall{quantile}(\hlfunctioncall{unlist}(ri.di), probs = probs)
+\}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
 
@@
 Additionally, the RNG kind is changed to the L'Ecuyer algorithm CITATION which is the best for random number generationalgorithm implemented in R for parallel processes.
-<<fish-set-up, message=FALSE>>=
-@
+\begin{knitrout}\scriptsize
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlfunctioncall{require}(shapes)
+\hlfunctioncall{require}(igraph)
+\hlfunctioncall{require}(parallel)
+\hlfunctioncall{require}(abind)
+\hlfunctioncall{source}(\hlstring{"landmark_montecarlo.R"})
+\hlfunctioncall{source}(\hlstring{"/home/peter/Documents/projects/3294254/readtps.R"})
+\hlfunctioncall{RNGkind}(\hlstring{"L'Ecuyer-CMRG"})
+\hlfunctioncall{set.seed}(1)
+ano <- \hlfunctioncall{read.tps}(\hlstring{"AnostomidaeFinal.TPS"}, shapes = TRUE)
+cur <- \hlfunctioncall{read.tps}(\hlstring{"CurimatidaeF.TPS"}, shapes = TRUE)
+fish <- \hlfunctioncall{abind}(ano$landmark, cur$landmark)
+\end{alltt}
+\end{kframe}
+\end{knitrout}
 
-Next, we use GPA to fit our entire sample of \Sexpr{dim(fish)[3]} fish.
+Next, we use GPA to fit our entire sample of 67 fish.
 We also calculate the pairwise \(\rho\) matrix.
@@
 
-<<fish-fit, eval=TRUE, cache=TRUE>>=
-@
+\begin{knitrout}\scriptsize
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+fish.fit <- \hlfunctioncall{procGPA}(fish)
+riem.comp <- \hlfunctioncall{matrix}(0, ncol = fish.fit$n, nrow = fish.fit$n)
+samp <- fish.fit$n
+\hlfunctioncall{for} (ii in \hlfunctioncall{seq}(samp)) \hlfunctioncall{for} (kk in \hlfunctioncall{seq}(samp)) \{
+    riem.comp[ii, kk] <- \hlfunctioncall{riemdist}(fish.fit$rotated[, , ii], fish.fit$rotated[, 
+        , kk])
+\}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
 
@@
 
-<<fish-ex1, echo=FALSE, eval=FALSE, message=FALSE, fig.height= 4, results='hide'>>=
-require(ggplot2)
-# turn fish.fit$rotated into a two column matrix
-fish.mean <- as.data.frame(fish.fit$mshape)
-fish.fix <- fish.fit$rotated[, , -42]
-fish.gg <- fish.fit$rotated[, , 1]
-for(ii in 2:(dim(fish.fix)[3])) {
-  fish.gg <- rbind(fish.gg, fish.fix[, , ii])
-}
-
-fish.gg <- as.data.frame(fish.gg)
-fs <- ggplot(fish.gg, aes(x= V1, y= V2)) + geom_point(size = 1)
-fs <- fs + scale_x_continuous(name = "X")
-fs <- fs + scale_y_continuous(name = "Y")
-fs <- fs + opts(panel.background = theme_rect(fill = "transparent",colour = NA), 
-                panel.grid.minor = theme_blank(), 
-                panel.grid.major = theme_blank(),
-                axis.line = theme_segment(),
-                axis.text.x = theme_text(colour = "black", size = 15),
-                axis.text.y = theme_text(colour = "black", size = 15)) 
-fs
-@
-
-<<fish-ex2, echo=FALSE, eval=FALSE, message=FALSE, fig.height= 4, results='hide'>>=
-fish.eigsc <- as.data.frame(fish.fit$stdscores)
-fam <- c(rep("Ano", times= length(ano$ID)), rep("Cur", times= length(cur$ID)))
-fish.eigsc$fam <- fam
-fish.eigsc <- fish.eigsc[-42, ]
-fe <- ggplot(fish.eigsc, aes(x= PC1, y= PC2, colour= fam))
-fe <- fe + geom_point(size = 3)
-fe <- fe + scale_x_continuous(name = paste("PC1; ", 
-                                           format(fish.fit$percent[1], digits= 3),
-                                           sep= ''))
-fe <- fe + scale_y_continuous(name = paste("PC2; ", 
-                                           format(fish.fit$percent[2], digits= 3),
-                                           sep= ''))
-fe <- fe + opts(panel.background = theme_rect(fill = "transparent",colour = NA), 
-                panel.grid.minor = theme_blank(), 
-                panel.grid.major = theme_blank(),
-                  axis.line = theme_segment(),
-                axis.text.x = theme_text(colour = "black", size = 15),
-                axis.text.y = theme_text(colour = "black", size = 15))
-fe
-@
-
-<<fish-ex3, echo=FALSE, eval=FALSE, message=FALSE, fig.height= 4, results='hide'>>=
-shapepca(fish.fit, joinline=c(1:35, 1))
-@
-
-<<fish-ex4, echo=FALSE, eval=FALSE, message=FALSE, fig.height= 4, results='hide'>>=
-par(mfrow = c(1, 2))
-tpsgrid(fish.fit$rotated[, , 1], fish.fit$mshape)
-tpsgrid(fish.fit$rotated[, , 19], fish.fit$mshape)
-@
-
 \begin{figure}
 \includegraphics[height=0.3\textheight]{figure/fish-ex1}
-\caption{Landmark positions of \Sexpr{dim(fish.fit$rotated)[3] - 1} of the fish taxa. One specimen is excluded.\label{fish-plotshapes}}
+\caption{Landmark positions of 66 of the fish taxa. One specimen is excluded.\label{fish-plotshapes}}
 \end{figure}
@@
 
-<<fish-MC, eval=FALSE>>=
-@
+\begin{knitrout}\scriptsize
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+cutoff <- \hlfunctioncall{shape.simulate}(fish.fit, nsim = 10, probs = 0.5)
+riem.comp[riem.comp > cutoff] <- 0
+riem.comp[riem.comp != 0] <- 1/riem.comp[riem.comp != 0]
+\end{alltt}
+\end{kframe}
+\end{knitrout}
 
@@
 
-<<fish-network, eval=FALSE>>=
-@
+\begin{knitrout}\scriptsize
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+data.graph <- \hlfunctioncall{graph.adjacency}(riem.comp, mode = \hlstring{"undirected"}, weighted = TRUE)
+data.fc <- \hlfunctioncall{fastgreedy.community}(data.graph)
+data.ebc <- \hlfunctioncall{edge.betweenness.community}(data.graph)
+data.lec <- \hlfunctioncall{leading.eigenvector.community}(data.graph)
+\end{alltt}
+\end{kframe}
+\end{knitrout}
 
@@
 The simplest manner is the just view the graphic output.
-
-<<fish-plots, eval=FALSE, echo=FALSE>>=
-@
 
@@
 For reproducibility information.
-<<reproducibility>>=
-sessionInfo()
-@
+\begin{knitrout}\scriptsize
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlfunctioncall{sessionInfo}()
+\end{alltt}
+\begin{verbatim}
+## R version 2.15.1 (2012-06-22)
+## Platform: x86_64-pc-linux-gnu (64-bit)
+## 
+## locale:
+##  [1] LC_CTYPE=en_US.UTF-8       LC_NUMERIC=C              
+##  [3] LC_TIME=en_AU.UTF-8        LC_COLLATE=en_US.UTF-8    
+##  [5] LC_MONETARY=en_AU.UTF-8    LC_MESSAGES=en_US.UTF-8   
+##  [7] LC_PAPER=C                 LC_NAME=C                 
+##  [9] LC_ADDRESS=C               LC_TELEPHONE=C            
+## [11] LC_MEASUREMENT=en_AU.UTF-8 LC_IDENTIFICATION=C       
+## 
+## attached base packages:
+## [1] parallel  stats     graphics  grDevices utils     datasets  methods  
+## [8] base     
+## 
+## other attached packages:
+## [1] abind_1.4-0          igraph_0.6-2         shapes_1.1-5        
+## [4] MASS_7.3-20          rgl_0.92.892         scatterplot3d_0.3-33
+## [7] formatR_0.6          knitr_0.7           
+## 
+## loaded via a namespace (and not attached):
+## [1] digest_0.5.2   evaluate_0.4.2 plyr_1.7.1     stringr_0.6.1 
+## [5] tools_2.15.1  
+\end{verbatim}
+\end{kframe}
+\end{knitrout}
 

### 2013-01-04 src/fish_analysis.r

@@ -1,0 +1,21 @@
+###############################################################################
+##
+##  data analysis
+##
+##  ben frable
+##  peter d smits
+##
+###############################################################################
+
+## libraries
+require(shapes)
+require(geomorph)
+require(cluster)
+
+## source files
+source('support_functions.r')
+source('fish_mung.r')
+
+
+## clustering
+clust <- lapply(dists, clu)

### 2013-01-04 src/old/readtps.R

@@ -1,0 +1,29 @@
+read.tps = function(data) {
+  # Reads the .tps file format produced by TPSDIG 
+  # (http://life.bio.sunysb.edu/morph/ into a single data frame
+  # USAGE: R> read.tps("filename.tps")
+  a = readLines(data) # so we can do some searching and indexing
+  LM = grep("LM", a) # find the line numbers for LM
+  ID.ind = grep("ID", a) # find the line numbers for ID
+  # and the ID values, SCALE values, and image names
+  ID = gsub("(ID=)(.*)", "\\2", grep("ID", a, value=T)) 
+  SCALE = gsub("(SCALE=)(.*)", "\\2", grep("SCALE", a, value=T)) 
+  images = basename(gsub("(IMAGE=)(.*)", "\\2", a[ID.ind - 1]))
+  # FOR EACH LOOP   
+  skip = LM # set how many lines to skip
+  # and how many rows to read
+  nrows = as.numeric(gsub("(LM=)(.*)", "\\2", grep("LM", a, value=T)))
+  l = length(LM) # number of loops we want
+  
+  landmarks = vector("list", l) # create an empty list
+  
+  for (i in 1:l) {
+    landmarks[i] = list(data.frame(
+      read.table(file=data, header=F, skip=LM[i],
+                 nrows=nrows[i], col.names=c("X", "Y")),
+      IMAGE = as.character(images[i]),
+      ID = ID[i],
+      SCALE = SCALE[i]))
+  }
+  do.call(rbind, landmarks) # rbind the list items into a data.frame
+}

### 2013-01-04 src/support_functions.r

@@ -1,0 +1,77 @@
+###############################################################################
+##
+##  support functions
+##
+##  peter d smits
+##
+###############################################################################
+
+## libraries
+require(utils)
+require(shapes)
+require(geomorph)
+
+## functions
+
+riem.matrix <- function(config) {
+  ##  create a pair-wise Riemannian distance matrix
+  ##
+  ##  Args:
+  ##    config: k x m x n array of (fit) landmarks
+  ##
+  ##  Returns:
+  ##    an n x n symetric matrix of Riemannian shape distances
+
+  n <- dim(config)[3]
+
+  pr <- combn(n, 2)
+  dis <- apply(pr, 2, function(x) riemdist(config[, , x[1]], 
+                                           config[, , x[2]]))
+  mat <- matrix(ncol = n, nrow = n)
+  for(ii in seq(ncol(pr))) {
+    mat[pr[1, ii], pr[2, ii]] <- dis[ii]
+  }
+
+  mat[lower.tri(mat)] <- t(mat[upper.tri(mat)])
+  diag(mat) <- 0
+
+  mat  
+}
+
+clu <- function(x) {
+  ## convenience clustering
+  ##
+  ##  Args:
+  ##    x: square, symmetric distance matrix
+  ##
+  ##  Returns:
+  ##    list with results from agnes and diana
+
+  out <- list()
+  out$agglo <- agnes(x)
+  out$div <- diana(x)
+  out
+} 
+
+land.frame <- function(x) {
+  ## make a nice data frame from a 3-dimenstional array
+  ##
+  ##  Args:
+  ##    x: 3-way array
+  ##
+  ##  Returns:
+  ##    data frame that is useful with ggplot
+
+  m <- dim(x)[2]
+  n <- dim(x)[3]
+  dat <- x[, , 1]
+  for (ii in seq(from = 2, to = n)) {
+    dat <- rbind(dat, x[, , ii])
+  }
+
+  dat <- as.data.frame(dat)
+  if (m == 2) nam <- c('x', 'y') else if (m == 3) nam <- c('x', 'y', 'z')
+  names(dat) <- nam
+
+  dat
+}

### 2013-01-06 documents/faqs_intro.Rnw
started report on prelim results and methods

\documentclass{article}
\usepackage{amsmath, amsthm, parskip, graphicx}

\frenchspacing

<<opts-set, message = FALSE, echo = FALSE>>=
require(knitr)
opts_chunk$set(warning = FALSE, message = FALSE, echo = FALSE,
               highlight = TRUE, background = 'white',
               fig.height = 3, fig.width = 4)
@

<<source-scripts>>=
source('../src/fish_plotting.r')
@

\title{Informal explination of study ``Ecotype detection in (geometric) morphometric data''}
\author{Peter D Smits \\
\and
Benjamin Frable}
\institute{Committee on Evolutionary Biology, University of Chicago \\
Department of Fisheries and Wildlife, Oregon State University}

\begin{document}
\maketitle

\section{Introduction}

\section{Materials and Methods}

\section{Results}

\section{Discussion}

\end{document}

### 2013-01-06 documents/faqs_intro.tex
started report on prelim results and methods

--- 2013-01-04 documents/report_1.tex
@@
 \usepackage{alltt}
+\usepackage{amsmath, amsthm, parskip, graphicx}
+
 \frenchspacing
 
-\title{Biological shape as a weighted network}
-\author{Peter D Smits \and Ben Frable}
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+
+{\ttfamily\noindent\bfseries\textcolor{errorcolor}{\#\# Error: cannot open the connection}}\end{kframe}
+\end{knitrout}
+
+\title{Informal explination of study ``Ecotype detection in (geometric) morphometric data''}
+\author{Peter D Smits \\
+\and
+Benjamin Frable}
+\institute{Committee on Evolutionary Biology, University of Chicago \\
+Department of Fisheries and Wildlife, Oregon State University}
 
 \begin{document}
-\maketitle
-
-\section{Brief Introduction}
-Here we envision biological shape, and the differences between shapes, as a network or graph model.
-A network model is the depiction of the relationships between many nodes.
-Each of these nodes may represent many different items or subjects of the same class.
-In, for example, a social network is where each node represents a single person.
-In this case, nodes represent a unique biological shape.
-
-Between these nodes are links or edges.
-In the social network example, an edge would represent a ``friendship'' or some other social connection between two individuals.
-It is possible for edges to have weights, representing the degree of connectedness between two nodes.
-Unlike in a distance matrix, a higher value represents a stronger connectedness.
-Here, our edges represent the similarity of two shape configurations.
-How these edges are assigned and weighted, however, is a via an extended procedure.
-
-In network theory, a community is a densly connected set of nodes that are more connected to each other than to other groups.
-In a given network, the number o groups may range between one large group and as many groups as there are nodes.
-The shape network made from the matrix of \(\rho\) can then be analyzed using the various community detection algorithms in order to find if there are any subgroups of morphologies that are more densly connected.
-These subgroups represent highly similar morphologies, and different communities/groups represent large scale differences in morphology.
-Effectively, these community detection algorithms are defining different morphological categories directly from the morphological similarities as opposed to using prior knowledge or observation.
-
-\section{Methodology}
-The core of this entire process lies with the Riemannian shape distance, \(\rho\), between two configurations. 
-A shape configuration is the set of landmarks that represent a unique shape. 
-There are a technically infinite number of unique shapes, and our sample in geometric morphometric studies only represents a subset of the biologically known or possible shapes.
-
-Shapes, when properly scaled and fit via some form of Procrustes analysis, exist not in a Euclidean plane, but on a Riemannian manifold.
-Convenitnetly, a Riemannian manifold is metrizable and we can define some kind of metric to represent the ``distance'' between two shape configurations.
-This Riemannian shape distance, \(\rho\) was defined by Kendall CITATION and varies between 0 and \(\pi /2\).
-A value closer to 0 means that two shapes are closer in shape space, while a higher value indicates that two shapes are farther away in shape space.
-Conveniently here is a function for calculating \(\rho\) in the {\tt{shapes}} package CITATION called {\tt{riemdist()}}.
-
-The first step is the using generalized Procrustes analysis (GPA) to align all of our shapes and remove all non-shape elements.
-After this a pairwise distance matrix is calculated for all shape combinations using {\tt{riemdist()}}. This produces a square \(n \cdot n\) matrix, where \(n\) is the number of unique shapes.
-This pairwise \(\rho\) matrix represents how close every shape is to all other shapes.
-From here, we want to determine which shapes are closer to each other than we would expect by random.
-We require a Monte Carlo process to determine what ``random'' is.
-
-How we define ``by random'' is very important here.
-Here, we are interested in a value of \(\rho\) smaller than than we would expect half of the time by random.
-This means, that we want to find which values of \(\rho\) are smaller than the median of a null distribution.
-
-\subsection{Monte Carlo process}
-The null distribution here is created from using two randomly selected shape configurations, moving the landmarks slightly to create two new, most likely valid, shape configuration.
-\(\rho\) is then calculated for these two new shapes and stored.
-This process is repeated many times in order to approximate a distribution of possible values of \(\rho\).
-After determining this null distribution, all values greater than the median of this null distribution are removed.
-The inverse of all remaining pairwise comparisons are then calculated, giving the future edge weights.
-The resulting matrix is an undirected, weighted adjacency matrix which can easily be converted into a network for use in the {\tt{igraph}} package CITATION.
-
-I've implemented this Monte Carlo process as a few functions in R.
-The core function is {\tt{jitter.landmark()}}.
-\begin{knitrout}\scriptsize
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-jitter.landmark <- \hlfunctioncall{function}(land) \{
-    \hlfunctioncall{for} (ii in \hlfunctioncall{seq}(\hlfunctioncall{ncol}(land))) \{
-        land[, ii] <- land[, ii] + \hlfunctioncall{runif}(n = \hlfunctioncall{length}(land[, ii]), min = -(\hlfunctioncall{min}(\hlfunctioncall{abs}(\hlfunctioncall{diff}(land[, 
-            ii])))), max = \hlfunctioncall{min}(\hlfunctioncall{abs}(\hlfunctioncall{diff}(land[, ii]))))
-    \}
-    land
-\}
-\end{alltt}
-\end{kframe}
-\end{knitrout}
-
-This functions are called {\tt{nsim}} times (deafult 1000) by the {\tt{riem.distribution()}} function.
-Using the original data set, {\tt{jitter.landmark()}} is called on {\tt{nsim}} two shape combinations.
-A lambda function is used to calculate the reimanian distance for all of the {\tt{nsim}} combinations.
-Currently, the lambda function executed as part of a call to {\tt{mclapply()}} which is a parallel function from the {\tt{parallel}} core package.
-Currently, it will use the maximum number of cores avaliable and the seed is optimzed for reproducibility.
-Without this feature, the process will run much slower.
-This function will not work on non-POSIX systems (i.e. Windows).
-\begin{knitrout}\scriptsize
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-riem.distribution <- \hlfunctioncall{function}(fit.land, nsim = 1000) \{
-    ran.land <- \hlfunctioncall{vector}(mode = \hlstring{"list"}, length = nsim)
-    \hlfunctioncall{for} (ii in \hlfunctioncall{seq}(nsim)) \{
-        s <- \hlfunctioncall{sample}(fit.land$n, 2)
-        sh1 <- \hlfunctioncall{jitter.landmark}(fit.land$rotated[, , s[1]])
-        sh2 <- \hlfunctioncall{jitter.landmark}(fit.land$rotated[, , s[2]])
-        ran.land[[ii]] <- \hlfunctioncall{abind}(sh1, sh2, along = 3)
-    \}
-    res <- \hlfunctioncall{mclapply}(ran.land, \hlfunctioncall{function}(x) \{
-        \hlfunctioncall{riemdis}(x[, , 1], x[, , 2])
-    \}, mc.cores = \hlfunctioncall{detectCores}(), mc.set.seed = T)
-    res
-\}
-\end{alltt}
-\end{kframe}
-\end{knitrout}
-
-% All of the above functions can probably made much simpler.
-% {\tt{jitter.landmark()}} can remain seperate while the entierty of {\tt{fit.riem()}} can be made a lambda function in the {\tt{mclapply()}} call in {\tt{riem.distribution()}}
-
-The interface function {\tt{shape.simulation()}} outputs a quantile value necessary for determining our ``at random.'' The deafult quantile of 0.50 is the same as the median of a sample.
-\begin{knitrout}\scriptsize
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-shape.simulate <- \hlfunctioncall{function}(fit.land, nsim = 1000, probs = 0.5) \{
-    ri.di <- \hlfunctioncall{riem.distribution}(fit.land, nsim = nsim)
-    \hlfunctioncall{quantile}(\hlfunctioncall{unlist}(ri.di), probs = probs)
-\}
-\end{alltt}
-\end{kframe}
-\end{knitrout}
-
-\subsection{Community detection}
-
-There are currently eight different community detection algorithms implemented as part of the {\tt{igraph}} package.
-Each of these algorithms try to optimze different communities based on various criteria.
-For example, the ``walktrap'' algorithm determines communties by which sections of a graph a random walk is most frequently.
-This algorithm, however, requires that there are no unconnected nodes, or nodes that have no edges.
-Currently, I have tested the ``edge betweenness,'' ``fast-greedy,'' and ``leading eigenvector'' algorithms.
-Each gives slightly different results, reflecting the method by which they determine couuminity presence and structure.
-
-Implementation of these algorithms is illustrated below with gorilla and fish examples.
-Additionally, I've included heirarchical clustering results from the pariwise \(\rho\) matrix for both of these examples.
-
-\section{Examples}
-\subsection{Gorilla vertebrae}
-TODO
-
-\subsection{Fish example}
-The first step is to import the correct libraries and data files.
-The functions described above are part of my script file, while the {\tt{readtps.R}} file is my fork of Ananda Mahto's original script GIST LINK.
-This fork improves integration of .tps files and the {\tt{shapes}} package.
-Additionally, becreporause this is a randomization process, for reproducibiliity it is necessary to manually set the seed value.
-Additionally, the RNG kind is changed to the L'Ecuyer algorithm CITATION which is the best for random number generationalgorithm implemented in R for parallel processes.
-\begin{knitrout}\scriptsize
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlfunctioncall{require}(shapes)
-\hlfunctioncall{require}(igraph)
-\hlfunctioncall{require}(parallel)
-\hlfunctioncall{require}(abind)
-\hlfunctioncall{source}(\hlstring{"landmark_montecarlo.R"})
-\hlfunctioncall{source}(\hlstring{"/home/peter/Documents/projects/3294254/readtps.R"})
-\hlfunctioncall{RNGkind}(\hlstring{"L'Ecuyer-CMRG"})
-\hlfunctioncall{set.seed}(1)
-ano <- \hlfunctioncall{read.tps}(\hlstring{"AnostomidaeFinal.TPS"}, shapes = TRUE)
-cur <- \hlfunctioncall{read.tps}(\hlstring{"CurimatidaeF.TPS"}, shapes = TRUE)
-fish <- \hlfunctioncall{abind}(ano$landmark, cur$landmark)
-\end{alltt}
-\end{kframe}
-\end{knitrout}
-
-Next, we use GPA to fit our entire sample of 67 fish.
-We also calculate the pairwise \(\rho\) matrix.
-This is actually really dirty code that can probably be implemented as a single {\tt{outer}} call, but I've yet to figure it out.
-
-\begin{knitrout}\scriptsize
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-fish.fit <- \hlfunctioncall{procGPA}(fish)
-riem.comp <- \hlfunctioncall{matrix}(0, ncol = fish.fit$n, nrow = fish.fit$n)
-samp <- fish.fit$n
-\hlfunctioncall{for} (ii in \hlfunctioncall{seq}(samp)) \hlfunctioncall{for} (kk in \hlfunctioncall{seq}(samp)) \{
-    riem.comp[ii, kk] <- \hlfunctioncall{riemdist}(fish.fit$rotated[, , ii], fish.fit$rotated[, 
-        , kk])
-\}
-\end{alltt}
-\end{kframe}
-\end{knitrout}
-
-From this fit, we should examine our data to better understand what is going on initially.
-The basic ways are looking at the fitted landmarks (Figure \ref{fish-plotshapes}), the eigenscores from pricipal components analysis (Figure \ref{fish-pca}), and the thin-plate spine plot of an example fish to the mean shape (Figure \ref{fish-tps}.
-
-\begin{figure}
-\includegraphics[height=0.3\textheight]{figure/fish-ex1}
-\caption{Landmark positions of 66 of the fish taxa. One specimen is excluded.\label{fish-plotshapes}}
-\end{figure}
-
-\begin{figure}
-\includegraphics[height=0.3\textheight]{figure/fish-ex2}
-\caption{PC1 versus PC2 for fish. Families are labeled. Percent of variation explained is included in axes labels. One specimen (42) is excluded.\label{fish-pca}}
-\end{figure}
-
-% \begin{figure}
-% \includegraphics[width=\textwidth]{figure/fish-ex3}
-% \caption{Alternative method for interpreting the principal components analysis.}
-% \end{figure}
-
-\begin{figure}
-\includegraphics[height=0.35\textheight]{figure/fish-ex4}
-\caption{TPS grids for Ano (right) and Cur (left).\label{fish-tps}}
-\end{figure}
-
-Following this, the Monte Carlo process is run so that we can begin to analyze the structure of our shape relationships.
-This process is very long and takes many hours.
-Those values which are greater than the median of the Monte Carlo distribution are discarted and the inverse of the remaining ones is calculated, producing a weighted adacency matrix.
-
-\begin{knitrout}\scriptsize
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-cutoff <- \hlfunctioncall{shape.simulate}(fish.fit, nsim = 10, probs = 0.5)
-riem.comp[riem.comp > cutoff] <- 0
-riem.comp[riem.comp != 0] <- 1/riem.comp[riem.comp != 0]
-\end{alltt}
-\end{kframe}
-\end{knitrout}
-
-The weighted adacency matrix is then transformed into a network and the three community detection algorithms are then used to analyze the resulting graph.
-
-\begin{knitrout}\scriptsize
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-data.graph <- \hlfunctioncall{graph.adjacency}(riem.comp, mode = \hlstring{"undirected"}, weighted = TRUE)
-data.fc <- \hlfunctioncall{fastgreedy.community}(data.graph)
-data.ebc <- \hlfunctioncall{edge.betweenness.community}(data.graph)
-data.lec <- \hlfunctioncall{leading.eigenvector.community}(data.graph)
-\end{alltt}
-\end{kframe}
-\end{knitrout}
-
-We can then examine the results of these community detection algorithms. 
-The simplest manner is the just view the graphic output.
-
-In general, there are two identified communities.
-These communities are not, however, identical to the two known families.
-There is a group of ``Cur'' fish that are most morphologically similar to ``Ano'' fish.
-
-The next step after this is determining if this graph is actually interesting.
-This requires more reading.
-
-\section{Concerns and Future Development}
-\begin{enumerate}
-\item I'm unsure how valid it is to define ``by random'' in the way that we did. 
-I would need to actually talk to someone or read a lot to answer this one.
-
-\item Need to make sure I'm not making invalid shape configurations
-
-\item Unsure if I need to do {\tt{procGPA()}} every time in the MC process.
-
-\item Choice of community detection algorithm.
-This one just requires a lot of reading I think to figure out how exactly they differ.
-
-\item Are the edge weights to high?
-Should they be rescaled
-
-\item MC approach to understanding network.
-{\tt{statnet}} packages.
-Is the network different from random?
-
-\item Model based clustering possibilities.
-Can we classify without a network?
-Would use the {\tt{mclust}} package CITATION, but there are awkward philosophical questions that come up with using that methodology.
-
-\item Need to do some machine learning reading.
-This project is essentially unsupervised clustering, but doing it in a weird way.
-
-\end{enumerate}
-
-\newpage
-For reproducibility information.
-\begin{knitrout}\scriptsize
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlfunctioncall{sessionInfo}()
-\end{alltt}
-\begin{verbatim}
-## R version 2.15.1 (2012-06-22)
-## Platform: x86_64-pc-linux-gnu (64-bit)
-## 
-## locale:
-##  [1] LC_CTYPE=en_US.UTF-8       LC_NUMERIC=C              
-##  [3] LC_TIME=en_AU.UTF-8        LC_COLLATE=en_US.UTF-8    
-##  [5] LC_MONETARY=en_AU.UTF-8    LC_MESSAGES=en_US.UTF-8   
-##  [7] LC_PAPER=C                 LC_NAME=C                 
-##  [9] LC_ADDRESS=C               LC_TELEPHONE=C            
-## [11] LC_MEASUREMENT=en_AU.UTF-8 LC_IDENTIFICATION=C       
-## 
-## attached base packages:
-## [1] parallel  stats     graphics  grDevices utils     datasets  methods  
-## [8] base     
-## 
-## other attached packages:
-## [1] abind_1.4-0          igraph_0.6-2         shapes_1.1-5        
-## [4] MASS_7.3-20          rgl_0.92.892         scatterplot3d_0.3-33
-## [7] formatR_0.6          knitr_0.7           
-## 
-## loaded via a namespace (and not attached):
-## [1] digest_0.5.2   evaluate_0.4.2 plyr_1.7.1     stringr_0.6.1 
-## [5] tools_2.15.1  
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
 

### 2013-01-06 documents/faqs_intro.Rnw
more faq document.

@@
 opts_chunk$set(warning = FALSE, message = FALSE, echo = FALSE,
-               highlight = TRUE, background = 'white',
-               fig.height = 3, fig.width = 4)
+               highlight = TRUE, background = 'white')
 @
 
-<<source-scripts>>=
+<<source-scripts, results = 'hide'>>=
 source('../src/fish_plotting.r')
@@
 \title{Informal explination of study ``Ecotype detection in (geometric) morphometric data''}
-\author{Peter D Smits \\
-\and
-Benjamin Frable}
-\institute{Committee on Evolutionary Biology, University of Chicago \\
-Department of Fisheries and Wildlife, Oregon State University}
+\author{Peter D Smits \and Benjamin Frable}
 
@@
 
+\subsection{Fish}
+
+\subsection{Morphometric analysis}
+
+\subsection{Machine learning}
+
+\subsubsection{Linear discriminate analysis}
+
+\subsubsection{Multinomial logistic regression}
+
 \section{Results}
+
+\subsection{Morphometric analysis}
+
+<<pls-fits, eval = FALSE>>=
+gg.means
+@
+\begin{figure}[p]
+  \centering
+  \includegraphics{figure/pls-fits3}
+  \caption{<+caption text+>}
+  \label{fig:fish-fit}
+\end{figure}
+
+<<pca-plots, eval = FALSE>>=
+gg.pca
+@
+\begin{figure}[p]
+  \centering
+  \includegraphics{figure/pca-plots3}
+  \caption{<+caption text+>}
+  \label{fig:fish-pca}
+\end{figure}
+
+\subsection{Machine learning}
 

### 2013-01-06 src/fish_plotting.r

@@ -18,8 +18,8 @@ require(ggdendro)
 require(xtable)
 
 ## source files
-source('support_functions.r')
-source('fish_analysis.r')
+source('../src/support_functions.r')
+source('../src/fish_analysis.r')
 
 
 ## tiny bit of munging

### 2013-01-07 src/misc_data_analysis.r
started using additional datasets from packages

@@ -1,0 +1,91 @@
+###############################################################################
+##
+##  miscellaneous morphometric data sets included in the packages
+##  machine learning methods used here to test ideas
+##
+##  Peter D Smits
+##
+###############################################################################
+
+##  libraries
+require(MASS)
+require(shapes)
+require(geomorph)
+require(nnet)
+require(caret)
+require(e1071)
+require(MuMIn)
+
+##  source files
+source('../src/support_functions.r')
+
+##  seed
+set.seed(1)
+
+
+##  ctrl set(s)
+ctrl <- trainControl(method = 'LOOCV',
+                     classProbs = TRUE,
+                     number = 10)
+
+
+##  primate vertebrae
+monkey <- abind::abind(gorf.dat, gorm.dat, 
+                       pongof.dat, pongom.dat, 
+                       panf.dat, panm.dat)
+mon <- procGPA(monkey)
+
+mon.group <- c(rep('female.gor', dim(gorf.dat)[3]),
+               rep('male.gor', dim(gorm.dat)[3]),
+               ##  there is a potentially major problem with the gorilla data set
+               ##  need to read the original paper that analyzes it
+               rep('female.pongo', dim(pongof.dat)[3]),
+               rep('male.pongo', dim(pongom.dat)[3]),
+               rep('female.pan', dim(panf.dat)[3]),
+               rep('male.pan', dim(panm.dat)[3]))
+mon.species <- c(rep('gor', (dim(gorf.dat)[3] + dim(gorm.dat)[3])),
+                 rep('pongo', (dim(pongof.dat)[3] + dim(pongom.dat)[3])),
+                 rep('pan', (dim(panf.dat)[3] + dim(panm.dat)[3])))
+mon.data <- cbind(as.data.frame(mon$stdscores),
+                  group = mon.group, species = mon.species)
+
+##  mouse verteabrae data
+mouse <- abind::abind(qcet2.dat, qlet2.dat, qset2.dat)
+mouse <- procGPA(mouse)
+mouse.group <- c(rep('control', dim(qcet2.dat)[3]),
+                 rep('large', dim(qlet2.dat)[3]),
+                 rep('small', dim(qset2.dat)[3]))
+mouse.data <- cbind(as.data.frame(mouse$stdscores),
+                    group = mouse.group)
+
+
+##  combine and make pretty
+land.proc <- list(monkey = mon,
+                  mouse = mouse)
+land.data <- list(monkey = mon.data,
[…]

### 2013-01-07 src/support_functions.r

@@ -106,3 +106,23 @@ predictMNL <- function(model, newdata) {
     ids
   }
 }
+
+multi.mod <- function(data, formula, preProc) {
+  ## fit multinomial logistic regression to data
+  ## this is so i can map over a list
+  ##
+  ##  Args:
+  ##    data: object of class data.frame, probably the training set
+  ##    formula: the formula for the multinomial logistic regression
+  ##    trControl: input for trControl in train
+  ##    preProc: input for preProc in train
+  ##
+  ##  Returns:
+  ##    object of class train with all the correct attributes
+
+  mod <- train(formula, data = data, 
+               method = 'multinom', 
+               preProc = preProc)
+
+  mod
+}

### 2013-01-10 documents/faqs_intro.Rnw
started using additional datasets from packages
beginning tables

@@
 source('../src/fish_plotting.r')
+source('../src/misc_data_plotting.r')
 @
@@
   \includegraphics{figure/pls-fits3}
-  \caption{<+caption text+>}
+  \caption{Procrustes fit landmark dispersion for all taxa in comparison to mean landmark position in red.}
   \label{fig:fish-fit}
@@
   \includegraphics{figure/pca-plots3}
-  \caption{<+caption text+>}
+  \caption{Comparison of first four pricipal components of Procrustes fit landmarks. Familes are highlighted.}
   \label{fig:fish-pca}
@@
 
+% model selection table
+<<fish-model-table>>=
+print(ftab, 
+      table.placement = 'p',
+      caption.placement = 'bottom',
+      sanitize.text.function = function(x) x,
+      sanitize.colnames.function = function(x) x)
+@
+
+<<fish-confusion>>=
+print(fconf,
+      table.placement = 'p',
+      caption.placement = 'bottom')
+@
+
+<<logistic-plots, eval = FALSE>>=
+gf
+@
+\begin{figure}[p]
+  \centering
+  \includegraphics{figure/logistic-plots}
+  \caption{Relative probabilities of Characiform family identification from the first two principal components of shape.
+The first two pricipal components are plotted here because they are the predictors of family from the AICc-best model.}
+  \label{fig:fish-logistic}
+\end{figure}
+
 \section{Discussion}

### 2013-01-12 documents/faqs_intro.Rnw
more documentation

@@
 \section{Introduction}
+Stuff
 
@@
 \subsection{Fish}
+Four families of Characiformes: Anostomidae, Chilodontidae, Curimatidae, and Prochilodontidae.
+
+Total \Sexpr{dim(fish.land)[3]} specimens for four families of Characiformes: \Sexpr{dim(ano.land)[3]} Anostomidae, \Sexpr{dim(chi.land)[3]} Chilodontidae, \Sexpr{dim(cur.land)[3]} Curimatidae, and \Sexpr{dim(pro.land)[3]} Prochilodontidae.
 
 \subsection{Morphometric analysis}
+\Sexpr{dim(fish.land)[1]} two-dimensional landmarks and the head and neck.
+Landmarks were selected for some reason and were collected using TPSDig2.
+Generalized Procrustes analysis was used to remove the effects of orientation, rotation, and scaling CITATION.
+Following this, points were projected onto tangent space using principal components analysis CITATION.
+This was done in the R statistical programming language CITATION using the ``shapes'' CITATION and ``geomorph'' packages.
 
 \subsection{Machine learning}
+The predictive accuracy of shape for family as assessed using multinomial logistic regression and a training set of 75\% of all samples and a test set of 25\% of all samples.
+Multinomial logist regression is a generalization of the standard logistic regression where instead of a binary response variable, there are multiple categorical responses.
+In this case, the multinomial response variable was family while predictors were the pricipal components of shape.
 
-\subsubsection{Linear discriminate analysis}
+Pricipal components analysis produced \Sexpr{length(fits$fish$percent)} eigenscores for every taxa and there are only \Sexpr{fits$fish$n} taxa sampled, it is necessary to select only from a subset of the possible predictors.
+Models were compared using the second-order Akaike's information criterion (AICc) where lower values indicate a better bias-variance trade off between the number of predictors and the likelihood of the model CITATIONS while taking into account sample size.
 
-\subsubsection{Multinomial logistic regression}
+Model selection is not without uncertainty, as some models may have very close AICc values.
+Model selection was done using \(\delta\)AICc and Akaike weight values CITATION.
+\(\delta\)AICc values are calculated as the difference between the AICc of a model and the AICc of the AICc-best model, or the model with the lowest AICc.
+Akaike weights sum to 1 and represent the propotional amount of information explained by a particular model and are approximately the model selection probability.
+Models with a \(\delta\)AICc of less than 2 were considered approximately identical in information to the AICc best model, while models with a \(\delta\)AICc of less than 6 were considered as plausible but suboptimal.
+Models with a \(\delta\)AICc of greater than 10 were considered extremely poor.
+
+I will eventually be making predictions from a model using averaged parameters of the first 95\% confidence set of models, as measured by Akaike weights CITATION.
+
+The models compared varied between 1 and \Sexpr{length(fish.vars)} pricipal component predictors, producing models with \Sexpr{paste(sort(fish.model.table$df), collapse = ', ')} parameters respectively.
+
+I did some model training using ``caret'' and ``e1071'' packages. Need to explain.
+
+Following model training and model comparison, the AICc-best model was used to predict the family of the testing data set described above.
 
@@
   \centering
-  \includegraphics{figure/pls-fits3}
+  \includegraphics[height = 0.4\textheight, keepaspectratio = true]{figure/pls-fits3}
   \caption{Procrustes fit landmark dispersion for all taxa in comparison to mean landmark position in red.}
@@
   \centering
-  \includegraphics{figure/pca-plots3}
+  \includegraphics[height = 0.4\textheight, keepaspectratio = true]{figure/pca-plots5}
   \caption{Comparison of first four pricipal components of Procrustes fit landmarks. Familes are highlighted.}
@@
 % model selection table
-<<fish-model-table>>=
+<<fish-model-table, results = 'asis'>>=
 print(ftab, 
@@
 
-<<fish-confusion>>=
+<<fish-confusion, results = 'asis'>>=
 print(fconf,
@@
   \centering
-  \includegraphics{figure/logistic-plots}
+  \includegraphics[width = 0.4\textwidth, keepaspectratio = true]{figure/logistic-plots}
   \caption{Relative probabilities of Characiform family identification from the first two principal components of shape.

### 2013-01-12 documents/faqs_intro.tex
more faq document.
more documentation

@@
 
+\title{Informal explination of study ``Ecotype detection in (geometric) morphometric data''}
+\author{Peter D Smits \and Benjamin Frable}
+
+\begin{document}
+\maketitle
+
+\section{Introduction}
+Stuff
+
+\section{Materials and Methods}
+
+\subsection{Fish}
+Four families of Characiformes: Anostomidae, Chilodontidae, Curimatidae, and Prochilodontidae.
+
+Total 78 specimens for four families of Characiformes: 18 Anostomidae, 4 Chilodontidae, 48 Curimatidae, and 8 Prochilodontidae.
+
+\subsection{Morphometric analysis}
+35 two-dimensional landmarks and the head and neck.
+Landmarks were selected for some reason and were collected using TPSDig2.
+Generalized Procrustes analysis was used to remove the effects of orientation, rotation, and scaling CITATION.
+Following this, points were projected onto tangent space using principal components analysis CITATION.
+This was done in the R statistical programming language CITATION using the ``shapes'' CITATION and ``geomorph'' packages.
+
+\subsection{Machine learning}
+The predictive accuracy of shape for family as assessed using multinomial logistic regression and a training set of 75\% of all samples and a test set of 25\% of all samples.
+Multinomial logist regression is a generalization of the standard logistic regression where instead of a binary response variable, there are multiple categorical responses.
+In this case, the multinomial response variable was family while predictors were the pricipal components of shape.
+
+Pricipal components analysis produced 70 eigenscores for every taxa and there are only 78 taxa sampled, it is necessary to select only from a subset of the possible predictors.
+Models were compared using the second-order Akaike's information criterion (AICc) where lower values indicate a better bias-variance trade off between the number of predictors and the likelihood of the model CITATIONS while taking into account sample size.
+
+Model selection is not without uncertainty, as some models may have very close AICc values.
+Model selection was done using \(\delta\)AICc and Akaike weight values CITATION.
+\(\delta\)AICc values are calculated as the difference between the AICc of a model and the AICc of the AICc-best model, or the model with the lowest AICc.
+Akaike weights sum to 1 and represent the propotional amount of information explained by a particular model and are approximately the model selection probability.
+Models with a \(\delta\)AICc of less than 2 were considered approximately identical in information to the AICc best model, while models with a \(\delta\)AICc of less than 6 were considered as plausible but suboptimal.
+Models with a \(\delta\)AICc of greater than 10 were considered extremely poor.
+
+I will eventually be making predictions from a model using averaged parameters of the first 95\% confidence set of models, as measured by Akaike weights CITATION.
+
+The models compared varied between 1 and 7 pricipal component predictors, producing models with 6, 9, 12, 15, 18, 21, 24 parameters respectively.
+
+I did some model training using ``caret'' and ``e1071'' packages. Need to explain.
+
+Following model training and model comparison, the AICc-best model was used to predict the family of the testing data set described above.
+
+\section{Results}
+
+\subsection{Morphometric analysis}
+
+\begin{figure}[p]
+  \centering
+  \includegraphics[height = 0.4\textheight, keepaspectratio = true]{figure/pls-fits3}
+  \caption{Procrustes fit landmark dispersion for all taxa in comparison to mean landmark position in red.}
+  \label{fig:fish-fit}
+\end{figure}
+
 \begin{knitrout}
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}
+\includegraphics[width=\maxwidth]{figure/pca-plots}
 
-{\ttfamily\noindent\bfseries\textcolor{errorcolor}{\#\# Error: cannot open the connection}}\end{kframe}
 \end{knitrout}
 
-\title{Informal explination of study ``Ecotype detection in (geometric) morphometric data''}
-\author{Peter D Smits \\
-\and
-Benjamin Frable}
-\institute{Committee on Evolutionary Biology, University of Chicago \\
-Department of Fisheries and Wildlife, Oregon State University}
+\begin{figure}[p]
+  \centering
+  \includegraphics[height = 0.4\textheight, keepaspectratio = true]{figure/pca-plots3}
+  \caption{Comparison of first four pricipal components of Procrustes fit landmarks. Familes are highlighted.}
+  \label{fig:fish-pca}
+\end{figure}
 
-\begin{document}
+\subsection{Machine learning}
+
+% model selection table
+% latex table generated in R 2.15.2 by xtable 1.7-0 package
+% Fri Jan 11 16:26:20 2013
+\begin{table}[p]
+\begin{center}
+\begin{tabular}{rlllllllllrrrrr}
+  \hline
+ & (Intercept) & PC1 & PC2 & PC3 & PC4 & PC5 & PC6 & PC7 & family & df & logLik & AICc & delta & weight \\ 
+  \hline
+2 & + & + & + &  &  &  &  &  &  & 9.00 & -14.34 & 50.35 & 0.00 & 0.97 \\ 
+  3 & + & + & + & + &  &  &  &  &  & 12.00 & -13.18 & 57.15 & 6.80 & 0.03 \\ 
+  4 & + & + & + & + & + &  &  &  &  & 15.00 & -11.46 & 64.09 & 13.74 & 0.00 \\ 
+  5 & + & + & + & + & + & + &  &  &  & 18.00 & -8.82 & 70.73 & 20.39 & 0.00 \\ 
+  1 & + & + &  &  &  &  &  &  &  & 6.00 & -33.71 & 81.03 & 30.68 & 0.00 \\ 
+  6 & + & + & + & + & + & + & + &  &  & 21.00 & -7.10 & 81.17 & 30.82 & 0.00 \\ 
+  7 & + & + & + & + & + & + & + & + &  & 24.00 & -0.04 & 83.38 & 33.04 & 0.00 \\ 
+   \hline
+\end{tabular}
+\caption{Model selection table.}
+\label{tab:fish-tab}
+\end{center}
+\end{table}
+
+% latex table generated in R 2.15.2 by xtable 1.7-0 package
+% Fri Jan 11 16:26:20 2013
+\begin{table}[p]
+\begin{center}
+\begin{tabular}{rrrrr}
+  \hline
+ & ano & chi & cur & pro \\ 
+  \hline
+ano &   4 &   0 &   0 &   0 \\ 
+  chi &   0 &   1 &   0 &   0 \\ 
+  cur &   0 &   0 &  12 &   0 \\ 
+  pro &   0 &   0 &   0 &   2 \\ 
+   \hline
+\end{tabular}
+\caption{Confusion matrix.}
+\label{tab:fish-mod}
+\end{center}
+\end{table}
+
+\begin{figure}[p]
+  \centering
+  \includegraphics[width = 0.4\textwidth, keepaspectratio = true]{figure/logistic-plots}
+  \caption{Relative probabilities of Characiform family identification from the first two principal components of shape.
+The first two pricipal components are plotted here because they are the predictors of family from the AICc-best model.}
+  \label{fig:fish-logistic}
+\end{figure}
+
+\section{Discussion}
 

### 2013-01-13 documents/faqs_intro.Rnw
faq draft

@@
   \centering
-  \includegraphics[height = 0.4\textheight, keepaspectratio = true]{figure/pca-plots5}
+  \includegraphics[width= \textwidth, keepaspectratio = true]{figure/pca-plots5}
   \caption{Comparison of first four pricipal components of Procrustes fit landmarks. Familes are highlighted.}
@@
       sanitize.text.function = function(x) x,
-      sanitize.colnames.function = function(x) x)
+      sanitize.colnames.function = function(x) x,
+      scalebox = 0.7)
 @
@@
   \centering
-  \includegraphics[width = 0.4\textwidth, keepaspectratio = true]{figure/logistic-plots}
+  \includegraphics[height = 0.5\textheight, keepaspectratio = true]{figure/logistic-plots}
   \caption{Relative probabilities of Characiform family identification from the first two principal components of shape.

### 2013-01-13 documents/faqs_intro.tex
faq draft

@@
 
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}
-\includegraphics[width=\maxwidth]{figure/pca-plots}
-
-\end{knitrout}
-
 \begin{figure}[p]
   \centering
-  \includegraphics[height = 0.4\textheight, keepaspectratio = true]{figure/pca-plots3}
+  \includegraphics[width= \textwidth, keepaspectratio = true]{figure/pca-plots5}
   \caption{Comparison of first four pricipal components of Procrustes fit landmarks. Familes are highlighted.}
@@
 % latex table generated in R 2.15.2 by xtable 1.7-0 package
-% Fri Jan 11 16:26:20 2013
+% Sat Jan 12 13:35:24 2013
 \begin{table}[p]
 \begin{center}
+\scalebox{0.7}{
 \begin{tabular}{rlllllllllrrrrr}
@@
 \end{tabular}
+}
 \caption{Model selection table.}
@@
 % latex table generated in R 2.15.2 by xtable 1.7-0 package
-% Fri Jan 11 16:26:20 2013
+% Sat Jan 12 13:35:24 2013
 \begin{table}[p]

### 2013-03-28 documents/turtle_report.Rnw
report template started. plotting functions drafted.

\documentclass{article}
\usepackage{amsmath, amsthm}
\usepackage{graphicx, microtype, parskip}
\usepackage{rotating, longtable}
\usepackage[sort&compress]{natbib}

\frenchspacing

<<message = FALSE, echo = FALSE, warning = FALSE>>=
require(knitr)
require(xtable)
require(reshape2)
require(ggplot2)
require(GGally)

opts_chunk$set(warning = FALSE, echo = FALSE, message = FALSE,
               highlight = TRUE, background = 'white')
@

<<results = 'hide'>>=
source('../src/support_functions.r')
load('../src/ml_analysis.r')
@

\title{Progress report: \textit{Emys marmorata} sub-species identification and classification}

\begin{document}
\maketitle

\section{Methods}
\subsection{Unsupervised}
\subsubsection{Gap-clustering}

\subsubsection{Evidence Accumulation Clustering}

\subsection{Supervised}
\begin{itemize}
  \item multinomial logistic regression
  \item feed-forward neural networks
  \item random forests
\end{itemize}

\section{Results}

\end{document}

### 2013-03-28 src/turtle_cluster.r

@@ -28,11 +28,11 @@ n.groups = 4
 
 set.seed(1)
 tpam <- function(x, k) pam(as.dist(x), k)
-tmorph.gap <- clusGap(turtle.dist, FUNcluster = tpam, K.max = 2 * n.groups)
+tmorph.gap <- clusGap(turtle.dist, FUNcluster = tpam, K.max = 10 * n.groups)
 tmorph.km <- pam(as.dist(turtle.dist), k = which.max(tmorph.gap$Tab[, 3]))
 
 # dissimilarity based evidence accumulation clustering
 set.seed(1)
-tmorph.dac <- dac(turtle.dist, krange = c(1, 10 * n.groups), iter = 1000)
+tmorph.dac <- dac(turtle.dist, krange = c(1, 50 * n.groups), iter = 10000)
 
 save(tmorph.gap, tmorph.km, tmorph.dac, file = 'cluster_res.RData')

### 2013-03-29 documents/turtle_report.Rnw
turtle report filling out

@@
 \section{Methods}
+No-free lunch theorem. Try lots of things because we don't understand everything.
+
 \subsection{Unsupervised}
-\subsubsection{Gap-clustering}
+\subsubsection{Gap-based clustering}
+Comparison of gap statistic results for partitioning around medoids (PAM) divisive clustering.
+Confidence intervals are determined via bootstrap.
+The higher the gap statistic, the better the clustering result.
 
 \subsubsection{Evidence Accumulation Clustering}
+Choosing an optimal number of partitions is hard, which is why gap-based cluster selection was used above.
+An alternative method is to look at the co-occurence frequency, that is how frequently any two samples occur in the same partition.
+Repeating this process over and over again creates the frequency, or ``vote'' for how the data set should be partitioned and  which specimens should be in the same cluster.
+
+EAC was originally devised using \textit{k}-means clustering, but I've extended it to use PAM clustering instead.
+The hope is to determine underlying structure in the data given a wide enough partition range and a high enough number of iterations.
 
@@
 
-\section{Results}
+\section{Preliminary results}
+
+\section{Miscellaneous affairs}
+\subsection{Evolution 2013}
+Machine learning approaches for recognizing cyptic diversity: a case study using \texit{Emys marmorata}.
+
+\subsection{Hinds grant}
+Resubmit previous application with the results from this study in the prelim-results section?
+
+Get network for one turtle? This should take long, just requires time.
 

### 2013-04-01 documents/turtle_report.Rnw
additions to report for clarity. new argument in map plot function.
supervised run. working on report for ken.

@@
 source('../src/support_functions.r')
-load('../src/ml_analysis.r')
+source('../src/plotting_functions.r')
+load('../src/turtle_analysis.r')
 @
@@
 \subsection{Unsupervised}
+Underlying structure in data?
+
 \subsubsection{Gap-based clustering}
@@
 The higher the gap statistic, the better the clustering result.
+Standard errors of the gap statistic were estimated from 100 resamples.
 
@@
 Choosing an optimal number of partitions is hard, which is why gap-based cluster selection was used above.
-An alternative method is to look at the co-occurence frequency, that is how frequently any two samples occur in the same partition.
+An alternative method is to look at the co-occurrence frequency, that is how frequently any two samples occur in the same partition.
 Repeating this process over and over again creates the frequency, or ``vote'' for how the data set should be partitioned and  which specimens should be in the same cluster.
@@
 The hope is to determine underlying structure in the data given a wide enough partition range and a high enough number of iterations.
+Dissimilarity based EAC was performed using a range of 1 though \Sexpr{50 * 4} possible partitions and based on 10,000 iterations.
 
 \subsection{Supervised}
+How well does data conform to predetermined structure?
+
 \begin{itemize}
@@
 \section{Preliminary results}
+Comparison of gap statistic over a very wide range of plausible partitions indicates that as the number of partitions increase, there is a marginal increasing gap statistic until approximately \Sexpr{max(tmorph.km$clustering)} after which there is a marginal decrease in gap statistic (Fig.
+\ref{fig:gap}).
+It is notable that the standard errors around the gap statistic values are very large, and the marginal increases in gap statistic with an increased number of partitions may not be important.
+Additionally, all gap statiic values are within \Sexpr{max(tmorph.gap$Tab[, 3]) - min(tmorph.gap$Tab[, 3])} of each other meaning that there is little over all consensus for how many clusters are present when comparing gap statistics.
+
+<<gap, eval = TRUE>>=
+gg.gap <- gap.plot(tmorph.gap)
+gg.gap
+@
+\begin{figure}[ht]
+  \centering
+  \includegraphics{figure/gap}
+  \caption{Gap statistic values for multiple PAM-based clustering configurations of the Riemmanian shape distances of the \text{Emys marmorata} plastra.
+Higher values indicate greater clustering.
+Standard errors are estimated from 100 bootstrap resamples.}
+  \label{fig:gap}
+\end{figure}
+
+Dissimilarity based EAC estimated approximately \Sexpr{max(tmorph.dac)} optimal partitions.
+This corresponds to the number of partitions in both the Spinks et al.
+and shXXX classification schemes.
+However, the number of individuals assigned to each class is very different.
+
+<<dac-tab, results = 'asis'>>=
+dac.tab <- xtable(table(tmorph.dac))
+caption(dac.tab) <- 'Number of specimens assigned to optimal number of partitions as determined by dissimilarity based EAC.
+Each column corresponds to a different partition, with the number assigned directly below it.'
+label(dac.tab) <- 'tab:dac'
+print.xtable(dac.tab
+             , table.placement = 'ht'
+             )
+@
 
@@
 
-\subsection{Hinds grant}
+\subsection{Grants}
+\subsubsection{Hinds Fund}
 Resubmit previous application with the results from this study in the prelim-results section?
@@
 
+\subsubsection{Paleontological Society}
+Why not? Need to wait till next year. Should probably join PaleoSoc anyway.
+
 \end{document}

### 2013-04-01 src/plotting_functions.r

@@ -27,10 +27,11 @@ gap.plot <- function(gap) {
   #   object of class ggplot
 
   tab <- as.data.frame(gap$Tab)
-  tab <- cbind(id = rownames(tab), tab)
+  tab <- cbind(id = as.numeric(rownames(tab)), tab)
   gg <- ggplot(tab, aes(x = id, y = gap))
   gg <- gg + geom_pointrange(mapping = aes(ymax = gap + SE.sim,
                                            ymin = gap - SE.sim))
+  gg <- gg + geom_line()
   gg <- gg + labs(x = 'Number of clusters',
                   y = 'Gap statistic')
   gg

### 2013-04-01 src/turtle_logistic.r

@@ -1,0 +1,41 @@
+###############################################################################
+##
+##  analysis of turtle plastron data
+##  supervised learning approaches
+##
+##  peter d smits
+##  psmits@uchicago.edu
+##
+###############################################################################
+
+# libraries
+require(MASS)
+require(nnet)
+require(cluster)
+require(e1071)
+require(caret)
+require(randomForest)
+require(pROC)
+require(MuMIn)
+
+require(parallel)
+require(doParallel)
+
+RNGkind(kind = "L'Ecuyer-CMRG")
+set.seed(1)
+
+registerDoParallel(cores = detectCores())
+
+# source files
+source('../src/support_functions.r')
+
+load('../src/supervised_misc.RData')
+
+set.seed(1)
+tmulti <- mapply(multi.train,
+                 form = tform, data = turtle.train,
+                 MoreArgs = list(method = 'multinom'
+                                 , trControl = ctrl
+                                 , maxit = 1000),
+                 SIMPLFY = FALSE)
+save(tmulti, file = 'multi_boot_mod.RData')

### 2013-04-01 src/turtle_nnet.r

@@ -1,0 +1,42 @@
+###############################################################################
+##
+##  analysis of turtle plastron data
+##  supervised learning approaches
+##
+##  peter d smits
+##  psmits@uchicago.edu
+##
+###############################################################################
+
+# libraries
+require(MASS)
+require(nnet)
+require(cluster)
+require(e1071)
+require(caret)
+require(randomForest)
+require(pROC)
+require(MuMIn)
+
+require(parallel)
+require(doParallel)
+
+RNGkind(kind = "L'Ecuyer-CMRG")
+set.seed(1)
+
+registerDoParallel(cores = detectCores())
+
+# source files
+source('../src/support_functions.r')
+
+load('../src/supervised_misc.RData')
+
+# neural networks
+set.seed(1)
+tnnet <- mapply(multi.train,
+                form = tform, data = turtle.train,
+                MoreArgs = list(method = 'nnet'
+                                , trControl = ctrl
+                                , maxit = 1000),
+                SIMPLFY = FALSE)
+save(tnnet, file = 'nnet_boot_mod.RData')

### 2013-04-01 src/turtle_rf.r

@@ -1,0 +1,45 @@
+###############################################################################
+##
+##  analysis of turtle plastron data
+##  supervised learning approaches
+##
+##  peter d smits
+##  psmits@uchicago.edu
+##
+###############################################################################
+
+# libraries
+require(MASS)
+require(nnet)
+require(cluster)
+require(e1071)
+require(caret)
+require(randomForest)
+require(pROC)
+require(MuMIn)
+
+require(parallel)
+require(doParallel)
+
+RNGkind(kind = "L'Ecuyer-CMRG")
+set.seed(1)
+
+registerDoParallel(cores = detectCores())
+
+# source files
+source('../src/support_functions.r')
+
+load('../src/supervised_misc.RData')
+
+# random forest
+set.seed(1)
+trf <- mapply(multi.train,
+                form = tform, data = turtle.train,
+                MoreArgs = list(method = 'rf'
+                                , trControl = ctrl
+                                , ntree = 1000
+                                , importance = TRUE
+                                ), 
+              SIMPLIFY = FALSE)
+save(trf, file = 'rf_boot_mod.RData')
+

### 2013-04-01 documents/turtle_report.Rnw
fixed issues involving wrong labels and clustering. most analyses run. cluster on server again.

@@
 load('../src/turtle_analysis.r')
+@
+
+<<>>=
+map <- map_data('state')
+st <- c('california', 'oregon', 'washington')
+map <- map[map$region == st, ]
 @
@@
 \section{Preliminary results}
+\subsection{Unsupervised}
 Comparison of gap statistic over a very wide range of plausible partitions indicates that as the number of partitions increase, there is a marginal increasing gap statistic until approximately \Sexpr{max(tmorph.km$clustering)} after which there is a marginal decrease in gap statistic (Fig.
@@
 
+<<dac-geo, eval = TRUE>>=
+ggdac <- map.plot(data = 
+                  , label = tmorph.dac
+                  , map = map
+                  )
+@
+
+\subsection{Supervised}
+
 \section{Miscellaneous affairs}

### 2013-04-01 src/turtle_analysis.r

@@ -21,20 +21,33 @@ source('../src/turtle_mung.r')
 
 load('../src/cluster_res.RData')
 load('../src/supervised_misc.RData')
-load('../src/supervised_misc.RData')
 load('../src/multi_boot_mod.RData')
 load('../src/nnet_boot_mod.RData')
 load('../src/rf_boot_mod.RData')
 
-clean.mods <- function(models) {
-  mm <- alply(models, 2)
-  names(mm) <- c('spinks', 'sh1', 'sh2', 'sh3')
+clean.mods <- function(models, lab = c('spinks', 'sh1', 'sh2', 'sh3')) {
+  if (!is.null(dim(models))) {
+    mm <- alply(models, 2)
+    names(mm) <- lab
+  } else {
+    breaks <- length(models)
+    mm <- list()
+    for (ii in seq(breaks)) {
+      mm[[ii]] <- models[[ii]]
+    }
+    names(mm) <- lab
+  }
   mm
 }
 
 tm <- clean.mods(tmulti)
 tnn <- clean.mods(tnnet)
 trf <- clean.mods(trf)
+
+classes <- list(spinks = classes$spinks,
+                sh1 = classes$sh1,
+                sh2 = classes$sh2,
+                sh3 = classes$sh3)
 
 # multinomial logistic regression
 tm.sel <- lapply(tm, function(x)

### 2013-04-01 src/turtle_supervised_data.r

@@ -35,7 +35,7 @@ source('../src/turtle_mung.r')
 
 max.var <- nrow(turtle.info) / 50
 tvar <- paste('PC', 1:max.var, sep = '')
-groups <- list('spinks', 'sh1', 'sh2', 'sh3')
+groups <- list('sh1', 'sh2', 'sh3', 'spinks')
 tform <- lapply(groups, function(x, y) make.form(y, x), y = tvar)
 
 set.seed(1)

### 2013-04-02 documents/turtle_report.Rnw
supervised analysis completed? rerun clustering because size mismatch. no idea.

@@
 source('../src/plotting_functions.r')
-load('../src/turtle_analysis.r')
+load('../src/turtle_analysis.RData')
 @
@@
 
-<<gap, eval = TRUE>>=
+<<gap, eval = FALSE>>=
 gg.gap <- gap.plot(tmorph.gap)
@@
   \centering
-  \includegraphics{figure/gap}
-  \caption{Gap statistic values for multiple PAM-based clustering configurations of the Riemmanian shape distances of the \text{Emys marmorata} plastra.
+  \includegraphics[width = \textwidth, keepaspectratio = true]{figure/gap}
+  \caption{Gap statistic values for multiple PAM-based clustering configurations of the Riemmanian shape distances of the \textit{Emys marmorata} plastra.
 Higher values indicate greater clustering.
@@
 <<dac-geo, eval = TRUE>>=
-ggdac <- map.plot(data = 
-                  , label = tmorph.dac
-                  , map = map
-                  )
+#ggdac <- map.plot(data = turtle.info
+#                  , label = tmorph.dac
+#                  , map = map
+#                  )
 @
@@
 
+% classification results
+% not necessary to show tables
+% just use maps and mark which are misclassified
+%   make train tiny dots and test large dots
+%   misclassified are a different shape
+
 \section{Miscellaneous affairs}
 \subsection{Evolution 2013}
-Machine learning approaches for recognizing cyptic diversity: a case study using \texit{Emys marmorata}.
+Machine learning approaches for recognizing cyptic diversity: a case study using \textit{Emys marmorata}.
 

### 2013-04-02 documents/turtle_report.tex
supervised analysis completed? rerun clustering because size mismatch. no idea.

--- 2013-04-02 documents/turtle_report.Rnw
@@
-\documentclass{article}
+\documentclass{article}\usepackage{graphicx, color}
+%% maxwidth is the original width if it is less than linewidth
+%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
+\makeatletter
+\def\maxwidth{ %
+  \ifdim\Gin@nat@width>\linewidth
+    \linewidth
+  \else
+    \Gin@nat@width
+  \fi
+}
+\makeatother
+
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+\definecolor{fgcolor}{rgb}{0.2, 0.2, 0.2}
+\newcommand{\hlnumber}[1]{\textcolor[rgb]{0,0,0}{#1}}%
+\newcommand{\hlfunctioncall}[1]{\textcolor[rgb]{0.501960784313725,0,0.329411764705882}{\textbf{#1}}}%
+\newcommand{\hlstring}[1]{\textcolor[rgb]{0.6,0.6,1}{#1}}%
+\newcommand{\hlkeyword}[1]{\textcolor[rgb]{0,0,0}{\textbf{#1}}}%
+\newcommand{\hlargument}[1]{\textcolor[rgb]{0.690196078431373,0.250980392156863,0.0196078431372549}{#1}}%
+\newcommand{\hlcomment}[1]{\textcolor[rgb]{0.180392156862745,0.6,0.341176470588235}{#1}}%
+\newcommand{\hlroxygencomment}[1]{\textcolor[rgb]{0.43921568627451,0.47843137254902,0.701960784313725}{#1}}%
+\newcommand{\hlformalargs}[1]{\textcolor[rgb]{0.690196078431373,0.250980392156863,0.0196078431372549}{#1}}%
+\newcommand{\hleqformalargs}[1]{\textcolor[rgb]{0.690196078431373,0.250980392156863,0.0196078431372549}{#1}}%
+\newcommand{\hlassignement}[1]{\textcolor[rgb]{0,0,0}{\textbf{#1}}}%
+\newcommand{\hlpackage}[1]{\textcolor[rgb]{0.588235294117647,0.709803921568627,0.145098039215686}{#1}}%
+\newcommand{\hlslot}[1]{\textit{#1}}%
+\newcommand{\hlsymbol}[1]{\textcolor[rgb]{0,0,0}{#1}}%
+\newcommand{\hlprompt}[1]{\textcolor[rgb]{0.2,0.2,0.2}{#1}}%
+
+\usepackage{framed}
+\makeatletter
+\newenvironment{kframe}{%
+ \def\at@end@of@kframe{}%
+ \ifinner\ifhmode%
+  \def\at@end@of@kframe{\end{minipage}}%
+  \begin{minipage}{\columnwidth}%
+ \fi\fi%
+ \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
+ \colorbox{shadecolor}{##1}\hskip-\fboxsep
+     % There is no \\@totalrightmargin, so:
+     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
+ \MakeFramed {\advance\hsize-\width
+   \@totalleftmargin\z@ \linewidth\hsize
+   \@setminipage}}%
+ {\par\unskip\endMakeFramed%
+ \at@end@of@kframe}
+\makeatother
+
+\definecolor{shadecolor}{rgb}{.97, .97, .97}
+\definecolor{messagecolor}{rgb}{0, 0, 0}
+\definecolor{warningcolor}{rgb}{1, 0, 1}
+\definecolor{errorcolor}{rgb}{1, 0, 0}
+\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
+
+\usepackage{alltt}
 \usepackage{amsmath, amsthm}
@@
 \frenchspacing
-
-<<message = FALSE, echo = FALSE, warning = FALSE>>=
-require(knitr)
-require(xtable)
-require(reshape2)
-require(ggplot2)
-require(GGally)
-
-opts_chunk$set(warning = FALSE, echo = FALSE, message = FALSE,
-               highlight = TRUE, background = 'white')
-@
-
-<<results = 'hide'>>=
-source('../src/support_functions.r')
-source('../src/plotting_functions.r')
-load('../src/turtle_analysis.RData')
-@
-
-<<>>=
-map <- map_data('state')
-st <- c('california', 'oregon', 'washington')
-map <- map[map$region == st, ]
-@
 
@@
 The hope is to determine underlying structure in the data given a wide enough partition range and a high enough number of iterations.
-Dissimilarity based EAC was performed using a range of 1 though \Sexpr{50 * 4} possible partitions and based on 10,000 iterations.
+Dissimilarity based EAC was performed using a range of 1 though 200 possible partitions and based on 10,000 iterations.
 
@@
 \subsection{Unsupervised}
-Comparison of gap statistic over a very wide range of plausible partitions indicates that as the number of partitions increase, there is a marginal increasing gap statistic until approximately \Sexpr{max(tmorph.km$clustering)} after which there is a marginal decrease in gap statistic (Fig.
+Comparison of gap statistic over a very wide range of plausible partitions indicates that as the number of partitions increase, there is a marginal increasing gap statistic until approximately 31 after which there is a marginal decrease in gap statistic (Fig.
 \ref{fig:gap}).
 It is notable that the standard errors around the gap statistic values are very large, and the marginal increases in gap statistic with an increased number of partitions may not be important.
-Additionally, all gap statiic values are within \Sexpr{max(tmorph.gap$Tab[, 3]) - min(tmorph.gap$Tab[, 3])} of each other meaning that there is little over all consensus for how many clusters are present when comparing gap statistics.
+Additionally, all gap statiic values are within 0.0065 of each other meaning that there is little over all consensus for how many clusters are present when comparing gap statistics.
 
-<<gap, eval = FALSE>>=
-gg.gap <- gap.plot(tmorph.gap)
-gg.gap
-@
 \begin{figure}[ht]
@@
 
-Dissimilarity based EAC estimated approximately \Sexpr{max(tmorph.dac)} optimal partitions.
+Dissimilarity based EAC estimated approximately 4 optimal partitions.
 This corresponds to the number of partitions in both the Spinks et al.
@@
 
-<<dac-tab, results = 'asis'>>=
-dac.tab <- xtable(table(tmorph.dac))
-caption(dac.tab) <- 'Number of specimens assigned to optimal number of partitions as determined by dissimilarity based EAC.
-Each column corresponds to a different partition, with the number assigned directly below it.'
-label(dac.tab) <- 'tab:dac'
-print.xtable(dac.tab
-             , table.placement = 'ht'
-             )
-@
-
-<<dac-geo, eval = TRUE>>=
-#ggdac <- map.plot(data = turtle.info
-#                  , label = tmorph.dac
-#                  , map = map
-#                  )
-@
+% latex table generated in R 2.15.3 by xtable 1.7-0 package
+% Tue Apr  2 17:04:07 2013
+\begin{table}[ht]
+\begin{center}
+\begin{tabular}{rr}
+  \hline
+ & tmorph.dac \\ 
+  \hline
+1 &  33 \\ 
+  2 &   3 \\ 
+  3 & 730 \\ 
+  4 &  10 \\ 
+   \hline
+\end{tabular}
+\caption{Number of specimens assigned to optimal number of partitions as determined by dissimilarity based EAC.
+Each column corresponds to a different partition, with the number assigned directly below it.}
+\label{tab:dac}
+\end{center}
+\end{table}
 

### 2013-04-03 documents/turtle_report.Rnw
most maps done. model resamp dist plots done.

@@
 map <- map[map$region == st, ]
+
+turtle.train <- lapply(turtle.train, 
+                       function(x) {
+                         x$long <- as.numeric(as.character(x$long)) * -1
+                         x$lat <- as.numeric(as.character(x$lat))
+                         x})
+turtle.test <- lapply(turtle.test,
+                      function(x) {
+                        x$long <- as.numeric(as.character(x$long)) * -1
+                        x$lat <- as.numeric(as.character(x$lat))
+                        x})
+
+labs <- Map(function(x, n) x[, n],
+            x = turtle.train, n = names(turtle.train))
 @
@@
 
-<<dac-geo, eval = TRUE>>=
+<<dac-geo, eval = FALSE>>=
+# map is the map of cali, ore, wash
+
 #ggdac <- map.plot(data = turtle.info
@@
 
-% classification results
-% not necessary to show tables
-% just use maps and mark which are misclassified
-%   make train tiny dots and test large dots
-%   misclassified are a different shape
+In the interest of space/time, I'm only going to display results from one of the selected models from each method for each classification scheme.
+
+<<multi-map, eval = FALSE>>=
+tm.map <- mapply(class.map
+                 , train = turtle.train
+                 , test = turtle.test
+                 , label = labs
+                 , pred = tm.class
+                 , MoreArgs = list(map = map)
+                 , SIMPLIFY = FALSE)
+tm.map
+@
+
+<<nnet-map, eval = FALSE>>=
+tnn.class.tiny <- lapply(tnn.class, function(x) x[[1]])
+tnn.map <- mapply(class.map
+                 , train = turtle.train
+                 , test = turtle.test
+                 , label = labs
+                 , pred = tnn.class.tiny
+                 , MoreArgs = list(map = map)
+                 , SIMPLIFY = FALSE)
+tnn.map
+@
+
+<<rf-map, eval = FALSE>>=
+trf.class.tiny <- lapply(trf.class, function(x) x[[1]])
+trf.map <- mapply(class.map
+                 , train = turtle.train
+                 , test = turtle.test
+                 , label = labs
+                 , pred = trf.class.tiny
+                 , MoreArgs = list(map = map)
+                 , SIMPLIFY = FALSE)
+trf.map
+@
+
+I'm currently holding back on showing the tables because there are a lot of them.
+
+But I can show the comparison of the predictive accuracies.
+
+<<resamp, eval = FALSE>>=
+tmod.re.clean <- lapply(tmod.re, function(x) clean.resamp(x$values))
+tmod.re.gg <- lapply(tmod.re.clean, resamp.plot)
+tmod.re.gg
+@
 

### 2013-04-03 src/plotting_functions.r

@@ -63,6 +63,28 @@ map.plot <- function(data, label, map, coord = 'gilbert') {
   gg
 }
 
+class.map <- function(train, test, label, pred, map) {
+  gg <- ggplot(map, aes(x = long, y = lat, group = group))
+  gg <- gg + geom_polygon(fill = 'white', colour = 'black')
+  gg <- gg + coord_map('gilbert')
+  train <- cbind(train, label = label)
+  gg <- gg + geom_point(data = train,
+                        mapping = aes(x = long, y = lat,
+                                      group = NULL,
+                                      colour = factor(label)),
+                        alpha = 0.7)
+  test <- cbind(test, label = pred)
+  gg <- gg + geom_point(data = test,
+                        mapping = aes(x = long, y = lat,
+                                      group = NULL,
+                                      colour = factor(label),
+                                      size = 4
+                                      ), 
+                        shape = 15,
+                        alpha = 0.8)
+  gg
+}
+
 
 clean.resamp <- function(values) {
   # clean up the results of caret so that i can use ggplot to plot them :P

### 2013-04-03 src/support_functions.r

@@ -145,3 +145,28 @@ multi.train <- function(form, data, seed = 1, ...) {
   rr <- lapply(form, train.formula, data = data, ...)
   rr
 }
+
+flatten.next <- function(xx) {
+  ll <- Filter(function(x) 'list' %in% class(x), xx)
+  uu <- Filter(function(x) !('list' %in% class(x)), xx)
+
+  smod <- list()
+
+  for(kk in seq(length(uu))) {
+    smod[[length(smod) + 1]] <- uu[[kk]]
+  }
+  for(ii in seq(length(ll))) {
+    for(jj in seq(length(ll[[ii]]))) {
+      smod[[length(smod) + 1]] <- ll[[ii]][[jj]]
+    }
+  }
+
+  uu.nam <- names(uu)
+  ll.nam <- names(unlist(Map(function(x, n) rep(x, n),
+                             x = names(ll),
+                             n = lapply(ll, length))))
+  nam <- c(uu.nam, ll.nam)
+  names(smod) <- nam
+
+  smod
+}

### 2013-04-03 src/turtle_analysis.r

@@ -95,4 +95,21 @@ trf.class <- mapply(predict, trf.best, turtle.test,
 trf.conf <- Map(function(x, y) lapply(x, confusionMatrix, y),
                 trf.class, classes)
 
+# best models compared
+tm.best
+tnn.best
+trf.best
+
+mods <- list()
+for(ii in seq(length(tm.best))) {
+  mods[[ii]] <- list(multi = tm.best[[ii]],
+                     nnet = tnn.best[[ii]],
+                     rf = trf.best[[ii]])
+}
+names(mods) <- c('sh1', 'sh2', 'sh3', 'spinks')
+
+tmod <- lapply(mods, flatten.next)
+tmod.re <- lapply(tmod, resamples)
+tmod.redi <- lapply(tmod.re, diff)
+
 save.image(file = 'turtle_analysis.RData')

### 2013-04-03 src/turtle_cluster.r

@@ -29,7 +29,7 @@ n.groups = 4
 set.seed(1)
 tpam <- function(x, k) pam(as.dist(x), k)
 tmorph.gap <- clusGap(turtle.info.dist, FUNcluster = tpam, K.max = 10 * n.groups)
-tmorph.km <- pam(as.dist(turtle.dist), k = which.max(tmorph.gap$Tab[, 3]))
+tmorph.km <- pam(as.dist(turtle.info.dist), k = which.max(tmorph.gap$Tab[, 3]))
 
 # dissimilarity based evidence accumulation clustering
 set.seed(1)

### 2013-04-04 documents/turtle_report.Rnw
setting up for subfigures
subfigures completed/set up. still need to update clustering results.
pca and mean shape plots

@@
 \usepackage{graphicx, microtype, parskip}
-\usepackage{rotating, longtable}
+\usepackage{rotating, longtable, caption, subcaption}
 \usepackage[sort&compress]{natbib}
@@
 
-<<dac-geo, eval = FALSE>>=
-# map is the map of cali, ore, wash
-
-#ggdac <- map.plot(data = turtle.info
-#                  , label = tmorph.dac
-#                  , map = map
-#                  )
+<<gm, eval = FALSE>>=
+gg.morph <- ggpairs(turtle.info, columns = c(1:3),
+                    upper = 'blank')
+fits <- procGPA(turtle.land.info)
+ggland <- mshape.plot(fits, all.points = TRUE)
+gg.morph <- putPlot(gg.morph, ggland, 1, 3)
+gg.morph
 @
 
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width = \textwidth]{figure/gm}
+  \caption{<+caption text+>}
+  \label{fig:gm}
+\end{figure}
+
 \subsection{Supervised}
+
+I'm currently holding back on showing the tables because there are a lot of them.
+
+But I can show the comparison of the predictive accuracies.
+
+<<resamp, eval = FALSE>>=
+tmod.re.clean <- lapply(tmod.re, function(x) clean.resamp(x$values))
+tmod.re.gg <- lapply(tmod.re.clean, resamp.plot)
+tmod.re.gg
+@
+
+\begin{figure}[t]
+  \centering
+  \begin{subfigure}[b]{0.5\textwidth}
+    \centering
+    \caption{}
+    \includegraphics[width = \textwidth]{figure/resamp1}
+    \label{fig:resamp1}
+  \end{subfigure}%
+  \begin{subfigure}[b]{0.5\textwidth}
+    \centering
+    \caption{}
+    \includegraphics[width = \textwidth]{figure/resamp2}
+    \label{fig:resamp2}
+  \end{subfigure}\\
+
+  \begin{subfigure}[b]{0.5\textwidth}
+    \centering
+    \caption{}
+    \includegraphics[width = \textwidth]{figure/resamp3}
+    \label{fig:resamp3}
+  \end{subfigure}%
+  \begin{subfigure}[b]{0.5\textwidth}
+    \centering
+    \caption{}
+    \includegraphics[width = \textwidth]{figure/resamp4}
+    \label{fig:resamp4}
+  \end{subfigure}
+  \caption{Comparison of resampling distributions of training set accuracy and kappa statistics for the selected models of each classification scheme.
+\ref{fig:resamp1}: sh1 classification scheme.
+\ref{fig:resamp2}: sh2 classification scheme.
+\ref{fig:resamp3}: sh3 classification scheme.
+\ref{fig:resamp4}: spinks classification scheme.}
+  \label{fig:resamp}
+\end{figure}
 
@@
 
+\begin{figure}[t]
+  \centering
+  \begin{subfigure}[b]{0.5\textwidth}
+    \centering
+    \caption{}
+    \includegraphics[width = \textwidth]{figure/multi-map1}
+    \label{fig:multi-map1}
+  \end{subfigure}%
+  \begin{subfigure}[b]{0.5\textwidth}
+    \centering
+    \caption{}
+    \includegraphics[width = \textwidth]{figure/multi-map2}
+    \label{fig:multi-map2}
+  \end{subfigure}\\
+
+  \begin{subfigure}[b]{0.5\textwidth}
+    \centering
+    \caption{}
+    \includegraphics[width = \textwidth]{figure/multi-map3}
+    \label{fig:multi-map3}
+  \end{subfigure}%
+  \begin{subfigure}[b]{0.5\textwidth}
+    \centering
+    \caption{}
+    \includegraphics[width = \textwidth]{figure/multi-map4}
+    \label{fig:multi-map4}
+  \end{subfigure}
+  \caption{<+caption text+>}
+  \label{fig:multi-map}
+\end{figure}
+
 <<nnet-map, eval = FALSE>>=
@@
 @
+
+\begin{figure}[t]
+  \centering
+  \begin{subfigure}[b]{0.5\textwidth}
+    \centering
+    \caption{}
+    \includegraphics[width = \textwidth]{figure/nnet-map1}
+    \label{fig:nnet-map1}
+  \end{subfigure}%
+  \begin{subfigure}[b]{0.5\textwidth}
+    \centering
+    \caption{}
+    \includegraphics[width = \textwidth]{figure/nnet-map2}
+    \label{fig:nnet-map2}
+  \end{subfigure}\\
+
+  \begin{subfigure}[b]{0.5\textwidth}
+    \centering
+    \caption{}
+    \includegraphics[width = \textwidth]{figure/nnet-map3}
+    \label{fig:nnet-map3}
+  \end{subfigure}%
+  \begin{subfigure}[b]{0.5\textwidth}
+    \centering
+    \caption{}
+    \includegraphics[width = \textwidth]{figure/nnet-map4}
+    \label{fig:nnet-map4}
+  \end{subfigure}
+  \caption{<+caption text+>}
+  \label{fig:nnet-map}
+\end{figure}
 
@@
 
-I'm currently holding back on showing the tables because there are a lot of them.
+\begin{figure}[t]
+  \centering
+  \begin{subfigure}[b]{0.5\textwidth}
+    \centering
+    \caption{}
+    \includegraphics[width = \textwidth]{figure/rf-map1}
+    \label{fig:rf-map1}
+  \end{subfigure}%
+  \begin{subfigure}[b]{0.5\textwidth}
+    \centering
+    \caption{}
+    \includegraphics[width = \textwidth]{figure/rf-map2}
+    \label{fig:rf-map2}
+  \end{subfigure}\\
 
-But I can show the comparison of the predictive accuracies.
-
-<<resamp, eval = FALSE>>=
-tmod.re.clean <- lapply(tmod.re, function(x) clean.resamp(x$values))
-tmod.re.gg <- lapply(tmod.re.clean, resamp.plot)
-tmod.re.gg
-@
+  \begin{subfigure}[b]{0.5\textwidth}
+    \centering
+    \caption{}
+    \includegraphics[width = \textwidth]{figure/rf-map3}
+    \label{fig:rf-map3}
+  \end{subfigure}%
+  \begin{subfigure}[b]{0.5\textwidth}
+    \centering
+    \caption{}
+    \includegraphics[width = \textwidth]{figure/rf-map4}
+    \label{fig:rf-map4}
+  \end{subfigure}
+  \caption{<+caption text+>}
+  \label{fig:rf-map}
+\end{figure}
 

### 2013-04-04 documents/turtle_report.tex
most maps done. model resamp dist plots done.
setting up for subfigures
subfigures completed/set up. still need to update clustering results.
pca and mean shape plots

@@
 % latex table generated in R 2.15.3 by xtable 1.7-0 package
-% Tue Apr  2 17:04:07 2013
+% Wed Apr  3 19:43:09 2013
 \begin{table}[ht]
@@
 
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width = \textwidth]{figure/gm}
+  \caption{<+caption text+>}
+  \label{fig:gm}
+\end{figure}
+
 \subsection{Supervised}
 
-% classification results
-% not necessary to show tables
-% just use maps and mark which are misclassified
-%   make train tiny dots and test large dots
-%   misclassified are a different shape
+I'm currently holding back on showing the tables because there are a lot of them.
+
+But I can show the comparison of the predictive accuracies.
+
+\begin{figure}[t]
+  \centering
+  \begin{subfigure}[b]{0.5\textwidth}
+    \centering
+    \caption{}
+    \includegraphics[width = \textwidth]{figure/resamp1}
+    \label{fig:resamp1}
+  \end{subfigure}%
+  \begin{subfigure}[b]{0.5\textwidth}
+    \centering
+    \caption{}
+    \includegraphics[width = \textwidth]{figure/resamp2}
+    \label{fig:resamp2}
+  \end{subfigure}\\
+
+  \begin{subfigure}[b]{0.5\textwidth}
+    \centering
+    \caption{}
+    \includegraphics[width = \textwidth]{figure/resamp3}
+    \label{fig:resamp3}
+  \end{subfigure}%
+  \begin{subfigure}[b]{0.5\textwidth}
+    \centering
+    \caption{}
+    \includegraphics[width = \textwidth]{figure/resamp4}
+    \label{fig:resamp4}
+  \end{subfigure}
+  \caption{Comparison of resampling distributions of training set accuracy and kappa statistics for the selected models of each classification scheme.
+\ref{fig:resamp1}: sh1 classification scheme.
+\ref{fig:resamp2}: sh2 classification scheme.
+\ref{fig:resamp3}: sh3 classification scheme.
+\ref{fig:resamp4}: spinks classification scheme.}
+  \label{fig:resamp}
+\end{figure}
+
+In the interest of space/time, I'm only going to display results from one of the selected models from each method for each classification scheme.
+
+\begin{figure}[t]
+  \centering
+  \begin{subfigure}[b]{0.5\textwidth}
+    \centering
+    \caption{}
+    \includegraphics[width = \textwidth]{figure/multi-map1}
+    \label{fig:multi-map1}
+  \end{subfigure}%
+  \begin{subfigure}[b]{0.5\textwidth}
+    \centering
+    \caption{}
+    \includegraphics[width = \textwidth]{figure/multi-map2}
+    \label{fig:multi-map2}
+  \end{subfigure}\\
+
+  \begin{subfigure}[b]{0.5\textwidth}
+    \centering
+    \caption{}
+    \includegraphics[width = \textwidth]{figure/multi-map3}
+    \label{fig:multi-map3}
+  \end{subfigure}%
+  \begin{subfigure}[b]{0.5\textwidth}
+    \centering
+    \caption{}
+    \includegraphics[width = \textwidth]{figure/multi-map4}
+    \label{fig:multi-map4}
+  \end{subfigure}
+  \caption{<+caption text+>}
+  \label{fig:multi-map}
+\end{figure}
+
+\begin{figure}[t]
+  \centering
+  \begin{subfigure}[b]{0.5\textwidth}
+    \centering
+    \caption{}
+    \includegraphics[width = \textwidth]{figure/nnet-map1}
+    \label{fig:nnet-map1}
+  \end{subfigure}%
+  \begin{subfigure}[b]{0.5\textwidth}
+    \centering
+    \caption{}
+    \includegraphics[width = \textwidth]{figure/nnet-map2}
+    \label{fig:nnet-map2}
+  \end{subfigure}\\
+
+  \begin{subfigure}[b]{0.5\textwidth}
+    \centering
+    \caption{}
+    \includegraphics[width = \textwidth]{figure/nnet-map3}
+    \label{fig:nnet-map3}
+  \end{subfigure}%
+  \begin{subfigure}[b]{0.5\textwidth}
+    \centering
+    \caption{}
+    \includegraphics[width = \textwidth]{figure/nnet-map4}
+    \label{fig:nnet-map4}
+  \end{subfigure}
+  \caption{<+caption text+>}
+  \label{fig:nnet-map}
+\end{figure}
+
+\begin{figure}[t]
+  \centering
+  \begin{subfigure}[b]{0.5\textwidth}
+    \centering
+    \caption{}
+    \includegraphics[width = \textwidth]{figure/rf-map1}
+    \label{fig:rf-map1}
+  \end{subfigure}%
+  \begin{subfigure}[b]{0.5\textwidth}
+    \centering
+    \caption{}
+    \includegraphics[width = \textwidth]{figure/rf-map2}
+    \label{fig:rf-map2}
+  \end{subfigure}\\
+
+  \begin{subfigure}[b]{0.5\textwidth}
+    \centering
+    \caption{}
+    \includegraphics[width = \textwidth]{figure/rf-map3}
+    \label{fig:rf-map3}
+  \end{subfigure}%
+  \begin{subfigure}[b]{0.5\textwidth}
+    \centering
+    \caption{}
+    \includegraphics[width = \textwidth]{figure/rf-map4}
+    \label{fig:rf-map4}
+  \end{subfigure}
+  \caption{<+caption text+>}
+  \label{fig:rf-map}
+\end{figure}
 

### 2013-04-04 documents/turtle_report.Rnw
confusion matrices for multi. to do list.

@@
 \begin{itemize}
-  \item multinomial logistic regression
-  \item feed-forward neural networks
-  \item random forests
+  \item multinomial logistic regression (Fig. \ref{fig:multi-map})
+  \item feed-forward neural networks (Fig. \ref{fig:nnet-map})
+  \item random forests (Fig. \ref{fig:rf-map})
 \end{itemize}
@@
 
-Dissimilarity based EAC estimated approximately \Sexpr{max(tmorph.dac)} optimal partitions.
-This corresponds to the number of partitions in both the Spinks et al.
-and shXXX classification schemes.
-However, the number of individuals assigned to each class is very different.
+Dissimilarity based EAC estimated approximately \Sexpr{max(tmorph.dac)} optimal partitions (Table \ref{tab:dac}).
 
@@
   \includegraphics[width = \textwidth]{figure/gm}
-  \caption{<+caption text+>}
+  \caption{Visualization of PCA of \textit{E.
+marmorata}.
+The lower triangle is the pairwise comparison of the first three pricipal components.
+The upper left corner is the comparison of landmark dispersion for all specimens compared to the mean shape in red.}
   \label{fig:gm}
@@
 I'm currently holding back on showing the tables because there are a lot of them.
+I'll show the confusion matrix for the multinomial logicistic regressions because they are easy (Tables \ref{tab:multi-conf-sh1}, \ref{tab:multi-conf-sh2}, \ref{tab:multi-conf-sh3}, \ref{tab:multi-conf-spinks})
 
-But I can show the comparison of the predictive accuracies.
+<<multi-tab, results = 'asis'>>=
+mt <- lapply(tm.conf, function(x) xtable(as.table(x)))
+
+for(ii in seq(length(mt))) {
+  caption(mt[[ii]]) <- paste0('Confusion matrix for ',
+                              names(mt)[ii],
+                              ' classification scheme. Rows correspond to predicted class. Columns correspond to reference class.')
+  label(mt[[ii]]) <- paste0('tab:multi-conf-', names(mt)[ii])
+}
+
+for(jj in seq(length(mt))) {
+  print.xtable(mt[[jj]]
+               , table.placement = 'ht')
+}
+
+@
+
+But I can show the comparison of the predictive accuracies (Fig. \ref{fig:resamp}).
 
@@
   \end{subfigure}
-  \caption{<+caption text+>}
+  \caption{Geographic position of all turtles sampled.
+Both training and testing observations are plotted.
+Training set observations are circles while testing observations are larger squares.
+Testing set observations are classified based on a multinomial logistic regression model.
+\ref{fig:multi-map1}: sh1 classification scheme.
+\ref{fig:multi-map2}: sh2 classification scheme.
+\ref{fig:multi-map3}: sh3 classification scheme.
+\ref{fig:multi-map4}: spinks classification scheme.}
   \label{fig:multi-map}
@@
   \end{subfigure}
-  \caption{<+caption text+>}
+  \caption{Geographic position of all turtles sampled.
+Both training and testing observations are plotted.
+Training set observations are circles while testing observations are larger squares.
+Testing set observations are classified based on a feed-forward single layer neural network model.
+\ref{fig:nnet-map1}: sh1 classification scheme.
+\ref{fig:nnet-map2}: sh2 classification scheme.
+\ref{fig:nnet-map3}: sh3 classification scheme.
+\ref{fig:nnet-map4}: spinks classification scheme.}
   \label{fig:nnet-map}
@@
   \end{subfigure}
-  \caption{<+caption text+>}
+  \caption{Geographic position of all turtles sampled.
+Both training and testing observations are plotted.
+Training set observations are circles while testing observations are larger squares.
+Testing set observations are classified based on a random forest model.
+\ref{fig:rf-map1}: sh1 classification scheme.
+\ref{fig:rf-map2}: sh2 classification scheme.
+\ref{fig:rf-map3}: sh3 classification scheme.
+\ref{fig:rf-map4}: spinks classification scheme.}
   \label{fig:rf-map}
@@
 
+\section{Future}
+\begin{enumerate}
+  \item remove juvelines
+  \item look into multinomial logistic mixed-effects models
+  \item other unsupervised methods, though that might have to wait for a follow up paper
+    \begin{itemize}
+      \item bayesian nonparametrics for categorical data
+    \end{itemize}
+\end{enumerate}
+
 \section{Miscellaneous affairs}
 \subsection{Evolution 2013}
-Machine learning approaches for recognizing cyptic diversity: a case study using \textit{Emys marmorata}.
+How cryptic is cryptic diversity? Machine learning approaches to fine scale variation in morphology of \textit{Emys marmorata}.
 

### 2013-04-04 documents/turtle_report.tex
confusion matrices for multi. to do list.

@@
 \subsection{Unsupervised}
-Comparison of gap statistic over a very wide range of plausible partitions indicates that as the number of partitions increase, there is a marginal increasing gap statistic until approximately 31 after which there is a marginal decrease in gap statistic (Fig.
+Comparison of gap statistic over a very wide range of plausible partitions indicates that as the number of partitions increase, there is a marginal increasing gap statistic until approximately 40 after which there is a marginal decrease in gap statistic (Fig.
 \ref{fig:gap}).
 It is notable that the standard errors around the gap statistic values are very large, and the marginal increases in gap statistic with an increased number of partitions may not be important.
-Additionally, all gap statiic values are within 0.0065 of each other meaning that there is little over all consensus for how many clusters are present when comparing gap statistics.
+Additionally, all gap statiic values are within 0.0064 of each other meaning that there is little over all consensus for how many clusters are present when comparing gap statistics.
 
@@
 
-Dissimilarity based EAC estimated approximately 4 optimal partitions.
-This corresponds to the number of partitions in both the Spinks et al.
-and shXXX classification schemes.
-However, the number of individuals assigned to each class is very different.
+Dissimilarity based EAC estimated approximately 5 optimal partitions (Table \ref{tab:dac}).
 
 % latex table generated in R 2.15.3 by xtable 1.7-0 package
-% Wed Apr  3 19:43:09 2013
+% Thu Apr  4 14:25:03 2013
 \begin{table}[ht]
@@
   \hline
-1 &  33 \\ 
-  2 &   3 \\ 
-  3 & 730 \\ 
-  4 &  10 \\ 
+1 & 722 \\ 
+  2 &   1 \\ 
+  3 &   1 \\ 
+  4 &   1 \\ 
+  5 &   1 \\ 
    \hline
@@
 I'm currently holding back on showing the tables because there are a lot of them.
+I'll show the confusion matrix for the multinomial logicistic regressions because they are easy (Tables \ref{tab:multi-conf-sh1}, \ref{tab:multi-conf-sh2}, \ref{tab:multi-conf-sh3}, \ref{tab:multi-conf-spinks})
 
-But I can show the comparison of the predictive accuracies.
+% latex table generated in R 2.15.3 by xtable 1.7-0 package
+% Thu Apr  4 14:25:03 2013
+\begin{table}[ht]
+\begin{center}
+\begin{tabular}{rrrr}
+  \hline
+ & CCR & marm & pall \\ 
+  \hline
+CCR &  40 &  14 &  11 \\ 
+  marm &  25 &  74 &   2 \\ 
+  pall &   3 &   1 &  10 \\ 
+   \hline
+\end{tabular}
+\caption{Confusion matrix for sh1 classification scheme. Rows correspond to predicted class. Columns correspond to reference class.}
+\label{tab:multi-conf-sh1}
+\end{center}
+\end{table}
+% latex table generated in R 2.15.3 by xtable 1.7-0 package
+% Thu Apr  4 14:25:03 2013
+\begin{table}[ht]
+\begin{center}
+\begin{tabular}{rrrr}
+  \hline
+ & CCR & marm & pall \\ 
+  \hline
+CCR &  44 &  12 &  10 \\ 
+  marm &  17 &  82 &   6 \\ 
+  pall &   2 &   1 &   7 \\ 
+   \hline
+\end{tabular}
+\caption{Confusion matrix for sh2 classification scheme. Rows correspond to predicted class. Columns correspond to reference class.}
+\label{tab:multi-conf-sh2}
+\end{center}
+\end{table}
+% latex table generated in R 2.15.3 by xtable 1.7-0 package
+% Thu Apr  4 14:25:03 2013
+\begin{table}[ht]
+\begin{center}
+\begin{tabular}{rrrrr}
+  \hline
+ & CCR & marm & pall & SJ \\ 
+  \hline
+CCR &  37 &  11 &   9 &   2 \\ 
+  marm &  18 &  59 &   3 &  10 \\ 
+  pall &   7 &   2 &  10 &   0 \\ 
+  SJ &   1 &   4 &   1 &   7 \\ 
+   \hline
+\end{tabular}
+\caption{Confusion matrix for sh3 classification scheme. Rows correspond to predicted class. Columns correspond to reference class.}
+\label{tab:multi-conf-sh3}
+\end{center}
+\end{table}
+% latex table generated in R 2.15.3 by xtable 1.7-0 package
+% Thu Apr  4 14:25:03 2013
+\begin{table}[ht]
+\begin{center}
+\begin{tabular}{rrrrr}
+  \hline
+ & 1 & 2 & 3 & 4 \\ 
+  \hline
+1 & 107 &   9 &  13 &  13 \\ 
+  2 &   1 &   8 &   0 &   0 \\ 
+  3 &   4 &   0 &   8 &   3 \\ 
+  4 &   4 &   0 &   2 &   8 \\ 
+   \hline
+\end{tabular}
+\caption{Confusion matrix for spinks classification scheme. Rows correspond to predicted class. Columns correspond to reference class.}
+\label{tab:multi-conf-spinks}
+\end{center}
+\end{table}
+
+But I can show the comparison of the predictive accuracies (Fig. \ref{fig:resamp}).
 

### 2013-04-04 src/turtle_em.r

@@ -1,0 +1,22 @@
+###############################################################################
+##
+##  analysis of turtle plastron data
+##  unsupervised learning approaches
+##
+##  peter d smits
+##  psmits@uchicago.edu
+##
+###############################################################################
+
+# libraries
+require(mclust)
+set.seed(1)
+
+
+# source files
+source('../src/support_functions.r')
+source('../src/turtle_mung.r')
+
+turtle.em <- Mclust(turtle.info[, 1:19])
+
+#save(turtle.em, 'turtle_em_res.RData')

### 2013-04-04 documents/turtle_report.Rnw
title and random details

@@
 
-\title{Progress report: \textit{Emys marmorata} sub-species identification and classification}
+\title{How cryptic is cryptic diversity? Machine learning approaches to fine scale variation in the morphology of \textit{Emys marmorata}.}
 
@@
 
+In general, misclassification seems to be at random.
+This is interesting for a few reasons.
+The assignments based on geography do not seem to bias results.
+All the taxa are extremely similar, though there are differences, hence the ~70\% accuracy.
+None of the classification schemes seems necessary better than any of the others, though sh3 is probably the best (4 classes) over all because it is the least ``random''.
+I think there is a summary class specific statistic that better explains this (sensitivity?
+specificity?
+detection rate?).
+
 \section{Miscellaneous affairs}
 \subsection{Evolution 2013}
-How cryptic is cryptic diversity? Machine learning approaches to fine scale variation in morphology of \textit{Emys marmorata}.
+How cryptic is cryptic diversity? Machine learning approaches to fine scale variation in the morphology of \textit{Emys marmorata}.
 

### 2013-04-04 documents/turtle_report.tex
title and random details

@@
 % latex table generated in R 2.15.3 by xtable 1.7-0 package
-% Thu Apr  4 14:25:03 2013
+% Thu Apr  4 16:52:02 2013
 \begin{table}[ht]

### 2013-04-05 documents/turtle_report.Rnw
other project ideas in relation to Hinds Fund/

@@
 
+Other project ideas I've come up with over the last few months.
+
+\begin{itemize}
+  \item Revisit Wagner 2000 idea of character/state exhaustion.
+Expand into the realm of uncertainty using stochastic character mapping.
+Potential also with chinese restaurant/indian buffet processes as models of character addition though this is highly suspect.
+  \item Character correlation and nonindependence.
+I did half my masters on this stuff.
+Using frequent pattern mining algorithms and nonparametric Bayesian approaches it may be possible to revisit these ideas with actual results.
+  \item NA terrestrial mammal community change in the Eocene.
+Using clustering/unsupervised methods, classify the sites in the Eocene.
+Spatio-temporal dynamics?
+Using co-occurrence networks and bipartite site-networks to understand spatio-temporal dynamics in paleo-community structure.
+  \item South African P/T faunal zones.
+Do my methods to the faunal assemblages.
+Sidor in press makes conclusions about how hard it is to identify some zones as opposed to others because of mixing.
+Do predictive methods agree?
+This is another nice spatio-temporal dynamics project.
+\end{itemize}
+
 \subsubsection{Paleontological Society}

### 2013-04-05 documents/turtle_report.tex
other project ideas in relation to Hinds Fund/

@@
 % latex table generated in R 2.15.3 by xtable 1.7-0 package
-% Thu Apr  4 16:52:02 2013
+% Fri Apr  5 14:42:26 2013
 \begin{table}[ht]

### 2013-04-05 documents/turtle_report.Rnw
more hinds nonsense

@@
 detection rate?).
+These will illuminate which classes are ``strongest''.
 
@@
 
-Other project ideas I've come up with over the last few months.
+Other project ideas I've come up with over the last few months that might be alternatives to the turtle shell network project.
 
 \begin{itemize}
-  \item Revisit Wagner 2000 idea of character/state exhaustion.
-Expand into the realm of uncertainty using stochastic character mapping.
-Potential also with chinese restaurant/indian buffet processes as models of character addition though this is highly suspect.
-  \item Character correlation and nonindependence.
-I did half my masters on this stuff.
-Using frequent pattern mining algorithms and nonparametric Bayesian approaches it may be possible to revisit these ideas with actual results.
   \item NA terrestrial mammal community change in the Eocene.
@@
 This is another nice spatio-temporal dynamics project.
+  \item Character correlation and nonindependence.
+I did half my masters on this stuff.
+Using frequent pattern mining algorithms and nonparametric Bayesian approaches it may be possible to revisit these ideas with actual results.
+  \item Revisit Wagner 2000 idea of character and character state exhaustion.
+Expand into the realm of uncertainty using stochastic character mapping.
+Potential also with chinese restaurant/indian buffet processes as models of character addition though this is highly suspect.
 \end{itemize}
+
+The three of these have at least preliminary methods/results, so they might be more appealing for the Hinds Fund.
+The concern becomes what do I actually need money for?
+Currently, my ideas are monitor, keyboard, mouse, hard drive, wrist bands (I have wrist pain).
+Publication and conference costs aren't appropriate for the Hinds, which is very frustrating because that's where most of my money goes.
+
+I still can't believe I was rejected last round.
 

### 2013-04-05 documents/turtle_report.tex
more hinds nonsense

@@
 % latex table generated in R 2.15.3 by xtable 1.7-0 package
-% Fri Apr  5 14:42:26 2013
+% Fri Apr  5 15:37:50 2013
 \begin{table}[ht]
@@
 % latex table generated in R 2.15.3 by xtable 1.7-0 package
-% Fri Apr  5 14:42:26 2013
+% Fri Apr  5 15:37:51 2013
 \begin{table}[ht]

### 2013-04-08 documents/turtle_report.Rnw
report for meeting done

--- 2013-04-04 documents/turtle_report.Rnw
@@
 It is notable that the standard errors around the gap statistic values are very large, and the marginal increases in gap statistic with an increased number of partitions may not be important.
-Additionally, all gap statiic values are within \Sexpr{max(tmorph.gap$Tab[, 3]) - min(tmorph.gap$Tab[, 3])} of each other meaning that there is little over all consensus for how many clusters are present when comparing gap statistics.
+Additionally, all gap statistic values are within \Sexpr{max(tmorph.gap$Tab[, 3]) - min(tmorph.gap$Tab[, 3])} of each other meaning that there is little over all consensus for how many clusters are present when comparing gap statistics.
 
@@
 
-Dissimilarity based EAC estimated approximately \Sexpr{max(tmorph.dac)} optimal partitions (Table \ref{tab:dac}).
+Dissimilarity based EAC estimated approximately \Sexpr{max(tmorph.dac)} optimal partitions (Table \ref{tab:dac}) though this may be only a marginal difference.
+I need to increase the number of resamples from 100 to 500 because that is when approximate stability is estimated to occur.
 
@@
 @
+
+Here is the PCA plot of the turtle pastra (Fig.
+\ref{fig:gm}).
+Also included is the mean shape and the dispersion around the landmarks.
+I haven't put nay classification labels on the points because an optimal partition scheme has not been selected.
 
@@
 marmorata}.
-The lower triangle is the pairwise comparison of the first three pricipal components.
+The lower triangle is the pairwise comparison of the first three principal components.
 The upper left corner is the comparison of landmark dispersion for all specimens compared to the mean shape in red.}
@@
 
-I'm currently holding back on showing the tables because there are a lot of them.
-I'll show the confusion matrix for the multinomial logicistic regressions because they are easy (Tables \ref{tab:multi-conf-sh1}, \ref{tab:multi-conf-sh2}, \ref{tab:multi-conf-sh3}, \ref{tab:multi-conf-spinks})
+I'm currently holding back on showing all the tables and model selection procedures because there are a lot of them.
+I'll show the confusion matrices from the multinomial logistic regressions because they are easy (Tables \ref{tab:multi-conf-sh1}, \ref{tab:multi-conf-sh2}, \ref{tab:multi-conf-sh3}, \ref{tab:multi-conf-spinks}).
+The confusion matrices from the neural network and random forest models are easy to make too, just would increase the overall length of this document by quite a bit.
 
@@
                               names(mt)[ii],
-                              ' classification scheme. Rows correspond to predicted class. Columns correspond to reference class.')
+                              ' classification scheme. Rows correspond to predicted class. Columns correspond to reference class. From multinomial logistic regression models.')
   label(mt[[ii]]) <- paste0('tab:multi-conf-', names(mt)[ii])
@@
 }
-
 @
@@
 \begin{enumerate}
-  \item remove juvelines
+  \item remove juveniles
   \item look into multinomial logistic mixed-effects models
@@
     \begin{itemize}
-      \item bayesian nonparametrics for categorical data
+      \item bayesian nonparametrics for categorical data.
+Probably the best option and most interesting technically.
+Additionally, relatively unknown in morphological contexts though beginning to be known in biological contexts as a whole.
     \end{itemize}
@@
 \subsubsection{Hinds Fund}
-Resubmit previous application with the results from this study in the prelim-results section?
+ NA terrestrial mammal community change in the Eocene.
+Using clustering/unsupervised methods, classify the sites in the Eocene.
+Spatio-temporal dynamics?
+Using co-occurrence networks and bipartite site-networks to understand spatio-temporal dynamics in paleo-community structure.
 
-Get network for one turtle? This should take long, just requires time.
+I have similar/preliminary results from my project from Mike Foote's class.
 
 \subsubsection{Paleontological Society}
-Why not? Need to wait till next year. Should probably join PaleoSoc anyway.
+Why not? Need to wait till next year.
 

### 2013-04-08 documents/turtle_report.tex
report for meeting done

--- 2013-04-04 documents/turtle_report.tex
@@
 It is notable that the standard errors around the gap statistic values are very large, and the marginal increases in gap statistic with an increased number of partitions may not be important.
-Additionally, all gap statiic values are within 0.0064 of each other meaning that there is little over all consensus for how many clusters are present when comparing gap statistics.
+Additionally, all gap statistic values are within 0.0064 of each other meaning that there is little over all consensus for how many clusters are present when comparing gap statistics.
 
@@
 
-Dissimilarity based EAC estimated approximately 5 optimal partitions (Table \ref{tab:dac}).
+Dissimilarity based EAC estimated approximately 5 optimal partitions (Table \ref{tab:dac}) though this may be only a marginal difference.
+I need to increase the number of resamples from 100 to 500 because that is when approximate stability is estimated to occur.
 
 % latex table generated in R 2.15.3 by xtable 1.7-0 package
-% Thu Apr  4 16:52:02 2013
+% Mon Apr  8 09:29:56 2013
 \begin{table}[ht]
@@
 
+Here is the PCA plot of the turtle pastra (Fig.
+\ref{fig:gm}).
+Also included is the mean shape and the dispersion around the landmarks.
+I haven't put nay classification labels on the points because an optimal partition scheme has not been selected.
+
 \begin{figure}[ht]
@@
 
-I'm currently holding back on showing the tables because there are a lot of them.
-I'll show the confusion matrix for the multinomial logicistic regressions because they are easy (Tables \ref{tab:multi-conf-sh1}, \ref{tab:multi-conf-sh2}, \ref{tab:multi-conf-sh3}, \ref{tab:multi-conf-spinks})
+I'm currently holding back on showing all the tables and model selection procedures because there are a lot of them.
+I'll show the confusion matrices from the multinomial logistic regressions because they are easy (Tables \ref{tab:multi-conf-sh1}, \ref{tab:multi-conf-sh2}, \ref{tab:multi-conf-sh3}, \ref{tab:multi-conf-spinks}).
+The confusion matrices from the neural network and random forest models are easy to make too, just would increase the overall length of this document by quite a bit.
 
 % latex table generated in R 2.15.3 by xtable 1.7-0 package
-% Thu Apr  4 16:52:02 2013
+% Mon Apr  8 09:29:56 2013
 \begin{table}[ht]
@@
 \end{tabular}
-\caption{Confusion matrix for sh1 classification scheme. Rows correspond to predicted class. Columns correspond to reference class.}
+\caption{Confusion matrix for sh1 classification scheme. Rows correspond to predicted class. Columns correspond to reference class. From multinomial logistic regression models.}
 \label{tab:multi-conf-sh1}
@@
 % latex table generated in R 2.15.3 by xtable 1.7-0 package
-% Thu Apr  4 16:52:02 2013
+% Mon Apr  8 09:29:56 2013
 \begin{table}[ht]
@@
 \end{tabular}
-\caption{Confusion matrix for sh2 classification scheme. Rows correspond to predicted class. Columns correspond to reference class.}
+\caption{Confusion matrix for sh2 classification scheme. Rows correspond to predicted class. Columns correspond to reference class. From multinomial logistic regression models.}
 \label{tab:multi-conf-sh2}
@@
 \end{tabular}
-\caption{Confusion matrix for sh3 classification scheme. Rows correspond to predicted class. Columns correspond to reference class.}
+\caption{Confusion matrix for sh3 classification scheme. Rows correspond to predicted class. Columns correspond to reference class. From multinomial logistic regression models.}
 \label{tab:multi-conf-sh3}
@@
 \end{tabular}
-\caption{Confusion matrix for spinks classification scheme. Rows correspond to predicted class. Columns correspond to reference class.}
+\caption{Confusion matrix for spinks classification scheme. Rows correspond to predicted class. Columns correspond to reference class. From multinomial logistic regression models.}
 \label{tab:multi-conf-spinks}

### 2013-04-24 src/turtle_analysis.r
coded to remove juvies. need to run analysis and debug.

@@ -44,19 +44,36 @@ tm <- clean.mods(tmulti)
 tnn <- clean.mods(tnnet)
 trf <- clean.mods(trf)
 
+tm.a <- clean.mods(tmulti.a)
+tnn.a <- clean.mods(tnnet.a)
+trf.a <- clean.mods(trf.a)
+
 # multinomial logistic regression
-tm.sel <- lapply(tm, function(x)
-                 model.sel(lapply(x, function(y) y$finalModel)))
+tm.a.sel <- lapply(tm.a, function(x)
+                   model.sel(lapply(x, function(y) y$finalModel)))
 
-tm.best <- mapply(function(sel, mod) mod[[as.numeric(rownames(sel)[1])]],
-                  sel = tm.sel, mod = tm,
-                  SIMPLIFY = FALSE)
+tm.a.best <- mapply(function(sel, mod) mod[[as.numeric(rownames(sel)[1])]],
+                    sel = tm.a.sel, mod = tm.a,
+                    SIMPLIFY = FALSE)
 
-tm.class <- mapply(predict, tm.best, turtle.test,
-                   MoreArgs = list(type = 'raw'), SIMPLIFY = FALSE)
+tm.a.class <- mapply(predict, tm.a.best, adult.test,
+                     MoreArgs = list(type = 'raw'), SIMPLIFY = FALSE)
 
-tm.conf <- Map(confusionMatrix,
-               tm.class, classes)
+tm.a.conf <- Map(confusionMatrix,
+                 tm.a.class, classes)
+
+tm.a.sel <- lapply(tm.a, function(x)
+                   model.sel(lapply(x, function(y) y$finalModel)))
+
+tm.a.best <- mapply(function(sel, mod) mod[[as.numeric(rownames(sel)[1])]],
+                    sel = tm.a.sel, mod = tm.a,
+                    SIMPLIFY = FALSE)
+
+tm.a.class <- mapply(predict, tm.a.best, adult.test,
+                     MoreArgs = list(type = 'raw'), SIMPLIFY = FALSE)
+
+tm.a.conf <- Map(confusionMatrix,
+                 tm.a.class, ad.class)
 
 
 # neural nets
@@ -77,6 +94,24 @@ tnn.class <- mapply(predict, tnn.best, turtle.test,
 tnn.conf <- Map(function(x, y) lapply(x, confusionMatrix, y),
                 tnn.class, classes)
 
[…]

### 2013-04-24 src/turtle_cluster.r

@@ -28,11 +28,18 @@ n.groups = 4
 
 set.seed(1)
 tpam <- function(x, k) pam(as.dist(x), k)
-tmorph.gap <- clusGap(turtle.info.dist, FUNcluster = tpam, K.max = 10 * n.groups)
+tmorph.gap <- clusGap(turtle.info.dist, FUNcluster = tpam, 
+                      K.max = 10 * n.groups, B = 500)
+tadult.gap <- clusGap(turtle.adult.dist, FUNcluster = tpam,
+                      K.max = 10 * n.groups, B = 500)
 tmorph.km <- pam(as.dist(turtle.info.dist), k = which.max(tmorph.gap$Tab[, 3]))
+tadult.km <- pam(as.dist(turtle.adult.dist), k = which.max(tadult.gap$Tab[, 3]))
 
 # dissimilarity based evidence accumulation clustering
 set.seed(1)
 tmorph.dac <- dac(turtle.info.dist, krange = c(1, 50 * n.groups), iter = 10000)
+tadult.dac <- dac(turtle.adult.dist, krange = c(1, 50 * n.groups), iter = 10000)
 
-save(tmorph.gap, tmorph.km, tmorph.dac, file = 'cluster_res.RData')
+save(tmorph.gap, tmorph.km, tmorph.dac, 
+     tadult.gap, tadult.km, tadult.dac, 
+     file = 'cluster_res.RData')

### 2013-04-24 src/turtle_mung.r

@@ -77,6 +77,17 @@ turtle.info$sh2 <- as.factor(as.character(turtle.info$sh2))
 turtle.info$sh3 <- as.factor(as.character(turtle.info$sh3))
 
 turtle.info.dist <- riem.matrix(turtle.land.info)  # riemmanian shape-distance
-
 turtle.geo <- turtle.geo[(turtle.geo[, 2]) > 100, ]
 
+## remove juvies
+## there are two classification options. ASK KEN WHICH IS THE ONE TO FOLLOW
+p.juv <- grep(pattern = '[jhJ]', 
+              x = as.character(turtle.info$p.sex), 
+              perl = TRUE)
+b.juv <- grep(pattern = '[jhJ]', 
+              x = as.character(turtle.info$b.sex),
+              perl = TRUE)
+turtle.adult <- turtle.info[-p.juv, ]
+turtle.land.adult <- turtle.land.info[, , -p.juv]
+turtle.adult.dist <- riem.matrix(turtle.land.adult)
+turtle.geo.adult <- turtle.geo[-p.juv, ]

### 2013-04-24 src/turtle_rf.r

@@ -41,5 +41,14 @@ trf <- mapply(multi.train,
                                 , importance = TRUE
                                 ), 
               SIMPLIFY = FALSE)
+trf.a <- mapply(multi.train,
+                form = tform.a, data = adult.train,
+                MoreArgs = list(method = 'rf'
+                                , trControl = ctrl
+                                , ntree = 1000
+                                , importance = TRUE
+                                ), 
+                SIMPLIFY = FALSE)
+
 save(trf, file = 'rf_boot_mod.RData')
 

### 2013-05-03 documents/turtle_presentation.Rnw
rfs for rf. moved a lot of files around. started presentation. need nnet results.

\documentclass{beamer}
\usepackage{amsmath, amsthm}
\usepackage{parskip, microtype, graphicx, caption}

\frenchspacing  % i likes it

<<message = FALSE, echo = FALSE, warning = FALSE>>=
require(knitr)
require(xtable)
require(ggplot2)
require(reshape2)
require(GGally)

opts_chunk$set(warning = FALSE, echo = FALSE, message = FALSE,
               highlight = TRUE, background = 'white',
               fig.height = 3, fig.width = 4
               , eval = FALSE
               )

#source('../src/plotting_functions.r')
#load('../src/turtle_analysis.r')
@

<<>>=
# need to format all the map data so it plays nicely
correct.map <- function(data) {
  out <- lapply(data, 
                function(x) {
                  x$long <- as.numeric(as.character(x$long)) * -1
                  x$lat <- as.numeric(as.character(x$lat))
                  x})

  out
}

map <- map_data('state')
st <- c('california', 'oregon', 'washington')
map <- map[map$region == st, ]

turtle.train <- correct.map(turtle.train)
turtle.test <- correct.map(turtle.test)
adult.train <- correct.map(adult.train)
adult.test <- correct.map(adult.test)

labs <- Map(function(x, n) x[, n],
            x = turtle.train, n = names(turtle.train))
@

\title{How cryptic is cryptic diversity? Machine learning approaches to fine scale variation in the morphology of \textit{Emys marmorata}.}
\author[shortname]{Peter D Smits\inst{1}, 
                   Kenneth D Angielczyk\inst{2}, 
                   James F Parham\inst{3}}
\institute[shortinst]{\inst{1} Committee on Evolution Biology, University of Chicago,
                      \inst{2} Department of Geology, Field Museum of Natural History,
                      \inst{3} Department of Geological Sciences, California State University -- Fullerton}

\begin{document}

\begin{frame}
  \maketitle
\end{frame}

\section{Introduction}
\begin{frame}
  \frametitle{Cryptic diversity}
  % image of example taxa that are morphologically identical and are only differentiated via genetic means
\end{frame}

\begin{frame}
  \frametitle{\textit{Emys marmorata}}
  % picture of a turtle
  % this is the study system
  % a little natural history
  % a little biogeography
\end{frame}

\begin{frame}
  \frametitle{Morphological hypotheses}
\end{frame}

\begin{frame}
  \frametitle{Phylogenetic hypotheses}
  % images of the different phylobiogeographic hypotheses from the different spinks papers
\end{frame}

\section{Methods}
\begin{frame}
  \frametitle{Methods: morphometrics}
  % two columns, text on one side and idealized landmarks on the other
  % ken probably has an image for the landmarks
  % these landmarks have been used before in other studies
  \begin{columns}
    \begin{column}{0.5\textwidth}
      \begin{itemize}
        \item plastral (``belly'') shape
        \item landmarks averaged across bilat axis
        \item total 13 landmarks, 7 on bilat axis, 6 off
        %\item \Sexpr{nrow(turtle.adult)} adult individuals
        \item geographic information known/inferred
      \end{itemize}
    \end{column}
    \begin{column}{0.5\textwidth}
      idealized plastral landmarks
    \end{column}
  \end{columns}
\end{frame}

\begin{frame}
  \frametitle{Methods: unsupervised learning}
\end{frame}

\begin{frame}
  \frametitle{Methods: supervised learning}
\end{frame}

\section{Results}
\begin{frame}
  \frametitle{Results: mophometrics}
\end{frame}

\begin{frame}
  \frametitle{Results: gap clustering}
\end{frame}

\begin{frame}
  \frametitle{Results: d-EAC}
\end{frame}

% model selection is unnecessary because of time constraints
\begin{frame}
  \frametitle{Results: multinomial logistic regression}
\end{frame}

\begin{frame}
  \frametitle{Results: neural networks}
\end{frame}

\begin{frame}
  \frametitle{Results: random forests}
\end{frame}

\section{Discussion}
\begin{frame}
  \frametitle{Best classification scheme}
\end{frame}

\begin{frame}
  \frametitle{Future}
\end{frame}

\begin{frame}
  \frametitle{Acknowledgements}
  \begin{columns}
    \begin{column}{0.5\textwidth}
      \begin{itemize}
        \item Ben Frable, Dallas Krentzel, Michael Foote
        \item COLLECTIONS
      \end{itemize}
    \end{column}
    \begin{column}{0.5\textwidth}
      \includegraphics[height = 0.25\textheight, keepaspectratio = true]{figure/chicago} \\
      \includegraphics[height = 0.25\textheight, width = \textwidth, keepaspectratio = true]{figure/field} \\
      \includegraphics[height = 0.25\textheight, keepaspectratio = true]{figure/csu}
    \end{column}
  \end{columns}
\end{frame}
\end{document}

### 2013-05-03 src/fish_plotting.r

@@ -1,124 +1,0 @@
-##############################################################################
-##
-##  plotting and tables
-##
-##  ben frable
-##  peter d smits
-##
-###############################################################################
-
-## libraries
-require(reshape2)
-require(grid)
-require(ggplot2)
-require(GGally)
-require(xtable)
-
-## source files
-source('../src/support_functions.r')
-source('../src/fish_analysis.r')
-
-
-## tiny bit of munging
-frames <- lapply(fish.fits, function(x) land.frame(x$rotated))
-frames$fish <- cbind(frames$fish,
-                     c(rep('ano', dim(frames$ano)[1]), 
-                       rep('cur', dim(frames$cur)[1]),
-                       rep('pro', dim(frames$pro)[1]),
-                       rep('chi', dim(frames$chi)[1])))
-names(frames$fish)[3] <- 'group'
-
-
-## tables
-ftab <- xtable(fish.model.table)
-caption(ftab) <- c('Model selection table.')
-label(ftab) <- c('tab:fish-tab')
-#digits(ftab)
-
-fconf <- xtable(as.table(fish.pred.confusion))
-caption(fconf) <- c('Confusion matrix.')
-label(fconf) <- c('tab:fish-mod')
-
-
-## mean shapes
-means <- lapply(fish.fits, function(x) as.data.frame(x$mshape))
-means <- lapply(means, function(x) cbind(x, rbind(x[-1, ], x[1, ])))
-means <- lapply(means, function(x) {
-                names(x) <- c('V1', 'V2', 'V3', 'V4'); return(x)
-                     })
-gmean <- function(x, m) {
-  gm <- ggplot(x, aes(x = x, y = y)) + geom_point()
-  gm <- gm + geom_point(data = m, 
-                        mapping = aes(x = V1, 
-                                      y = V2, 
-                                      colour = 'red', 
-                                      size = 1.5))
-  gm <- gm + geom_segment(data = m,
-                          mapping = aes(x = V1,
-                                        xend = V3,
-                                        y = V2,
-                                        yend = V4,
-                                        size = 1.2))
-  gm <- gm + theme(#panel.grid = element_blank(), 
-                   panel.background = element_blank(), 
-                   legend.position = 'none',
-                   axis.line = element_line(colour = 'black'))
-  gm
-}
[…]

### 2013-05-03 src/misc_data_analysis.r

@@ -1,113 +1,0 @@
-###############################################################################
-##
-##  miscellaneous morphometric data sets included in the packages
-##  machine learning methods used here to test ideas
-##
-##  Peter D Smits
-##
-###############################################################################
-
-## libraries
-require(MASS)
-require(shapes)
-require(geomorph)
-require(nnet)
-require(caret)
-require(e1071)
-require(MuMIn)
-
-## source files
-source('../src/support_functions.r')
-
-## seed
-set.seed(1)
-
-
-## ctrl set(s)
-ctrl <- trainControl(method = 'LOOCV',
-                     classProbs = TRUE,
-                     number = 10)
-
-
-## primate vertebrae
-monkey <- abind::abind(gorf.dat, gorm.dat, 
-                       pongof.dat, pongom.dat, 
-                       panf.dat, panm.dat)
-mon <- procGPA(monkey)
-
-mon.group <- c(rep('female.gor', dim(gorf.dat)[3]),
-               rep('male.gor', dim(gorm.dat)[3]),
-               ##  there is a potentially major problem with the gorilla data set
-               ##  need to read the original paper that analyzes it
-               rep('female.pongo', dim(pongof.dat)[3]),
-               rep('male.pongo', dim(pongom.dat)[3]),
-               rep('female.pan', dim(panf.dat)[3]),
-               rep('male.pan', dim(panm.dat)[3]))
-mon.species <- c(rep('gor', (dim(gorf.dat)[3] + dim(gorm.dat)[3])),
-                 rep('pongo', (dim(pongof.dat)[3] + dim(pongom.dat)[3])),
-                 rep('pan', (dim(panf.dat)[3] + dim(panm.dat)[3])))
-mon.data <- cbind(as.data.frame(mon$stdscores),
-                  group = mon.group, species = mon.species)
-
-## mouse verteabrae data
-mouse <- abind::abind(qcet2.dat, qlet2.dat, qset2.dat)
-mouse <- procGPA(mouse)
-mouse.group <- c(rep('control', dim(qcet2.dat)[3]),
-                 rep('large', dim(qlet2.dat)[3]),
-                 rep('small', dim(qset2.dat)[3]))
-mouse.data <- cbind(as.data.frame(mouse$stdscores),
-                    group = mouse.group)
-
-## combine and make pretty
-land.proc <- list(monkey = mon,
-                  mouse = mouse)
-land.data <- list(monkey = mon.data,
-                  mouse = mouse.data)
-
[…]

### 2013-05-03 src/other_analyses/fish_mung.r

@@ -1,0 +1,49 @@
+###############################################################################
+##
+##  fish data munging
+##
+##  ben frable
+##  peter d smits
+##
+###############################################################################
+
+## libraries
+require(shapes)
+require(geomorph)
+require(abind)
+
+## source files
+source('../src/support_functions.r')
+
+## data
+ano.land <- readland.tps(file = '../data/AnostomidaeFinal.TPS')
+cur.land <- readland.tps(file = '../data/CurimatidaeF.TPS')
+cur.land <- cur.land[, , -24]  # there is an outlier
+pro.land <- readland.tps(file = '../data/ProchilodontidaeF.TPS')
+chi.land <- readland.tps(file = '../data/Chilodontidae2.TPS')
+
+fish.land <- abind::abind(ano.land, cur.land, pro.land, chi.land)
+dimnames(fish.land)[[3]] <- c(dimnames(ano.land)[[3]], 
+                              dimnames(cur.land)[[3]],
+                              dimnames(pro.land)[[3]],
+                              dimnames(chi.land)[[3]])
+
+name <- c('ano', 'cur', 'pro', 'chi', 'fish')
+
+## generalized procrustes fit
+## PCA of the GPA fit landmarks
+fish.fits <- lapply(list(ano.land, cur.land,
+                         pro.land, chi.land,
+                         fish.land), procGPA)
+names(fish.fits) <- name
+
+## family means
+## need to automate this better
+ano.mean <- mshape(fits$fish$rotated[, , dim(ano.land)[3]])
+cur.mean <- mshape(fits$fish$rotated[, , dim(cur.land)[3]])
+pro.mean <- mshape(fits$fish$rotated[, , dim(pro.land)[3]])
+chi.mean <- mshape(fits$fish$rotated[, , dim(chi.land)[3]])
+
+## distance matrices
+fish.dists <- lapply(fits, function(x) riem.matrix(x$rotated))
+names(fish.dists) <- name

### 2013-05-03 src/other_analyses/misc_data_plotting.r

@@ -1,0 +1,38 @@
+###############################################################################
+##
+##  miscellaneous morphometric data sets included in the packages
+##  plotting results of various analyses
+##
+##  peter d smits
+##
+###############################################################################
+
+## libraries
+require(reshape2)
+require(grid)
+require(ggplot2)
+require(GGally)
+require(xtable)
+
+## source files
+source('../src/misc_data_analysis.r')
+
+
+## tables
+
+
+## procrustes fit results
+land.scores <- lapply(land.proc, function(x) as.data.frame(x$stdscore))
+land.scores <- Map(cbind, land.scores,
+                   list(mon.species,
+                        mouse.group))
+land.scores <- lapply(land.scores, 
+                      function(x) {names(x)[ncol(x)] <- 'group'; return(x)})
+land.pca <- lapply(land.scores, 
+                   function(x) ggpairs(x,
+                                       columns = c(1:4, ncol(x)),
+                                       upper = 'blank',
+                                       colour = 'group'))
+
+
+## machine learning results

### 2013-05-03 src/turtle_analysis.r

@@ -83,7 +83,7 @@ tnn.varimp <- lapply(tnn, varImp)
 tnn.re <- resamples(tnn)
 tnn.redi <- diff(tnn.re)
 
-tnn.class <- mapply(predict, tnn.best, turtle.test,
+tnn.class <- mapply(predict, tnn, turtle.test,
                     SIMPLFIY = FALSE)
 
 tnn.conf <- Map(function(x, y) confusionMatrix(x$pred, y),
@@ -94,7 +94,7 @@ tnn.a.varimp <- lapply(tnn.a, varImp)
 tnn.a.re <- resamples(tnn.a)
 tnn.a.redi <- diff(tnn.a.re)
 
-tnn.a.class <- mapply(predict, tnn.a.best, adult.test,
+tnn.a.class <- mapply(predict, tnn.a, adult.test,
                       SIMPLIFY = FALSE)
 
 tnn.a.conf <- Map(function(x, y) confusionMatrix(x$pred, y),

### 2013-05-03 src/turtle_em.r

@@ -1,22 +1,0 @@
-###############################################################################
-##
-##  analysis of turtle plastron data
-##  unsupervised learning approaches
-##
-##  peter d smits
-##  psmits@uchicago.edu
-##
-###############################################################################
-
-# libraries
-require(mclust)
-set.seed(1)
-
-
-# source files
-source('../src/support_functions.r')
-source('../src/turtle_mung.r')
-
-turtle.em <- Mclust(turtle.info[, 1:19])
-
-#save(turtle.em, 'turtle_em_res.RData')

### 2013-05-03 documents/turtle_report.tex
support for ROC based model selection added.

@@
+% this document has the propensity to become outdated as the code base
+% and results are updated over time. the pdf file should represent the most
+% stable implementation of this document.
+%
+% this is the basic ``template'' for all reports for meetings with ken and to
+% send to collaborators.
 \documentclass{article}\usepackage{graphicx, color}
@@
 
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}
+\includegraphics[width=\maxwidth]{figure/gap}
+
+\end{knitrout}
+
 \begin{figure}[ht]
@@
 
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}
+\includegraphics[width=\maxwidth]{figure/a-gap}
+
+\end{knitrout}
+
 Dissimilarity based EAC estimated approximately 5 optimal partitions (Table \ref{tab:dac}) though this may be only a marginal difference.
@@
 % latex table generated in R 2.15.3 by xtable 1.7-0 package
-% Mon Apr  8 09:29:56 2013
+% Fri May  3 10:50:12 2013
 \begin{table}[ht]
@@
 
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{verbatim}
+## [1] "To speed up use option distances=FALSE"
+## [1] "To speed up use option pcaoutput=FALSE"
+\end{verbatim}
+\end{kframe}
+\includegraphics[width=\maxwidth]{figure/gm}
+
+\end{knitrout}
+
 \begin{figure}[ht]
@@
 % latex table generated in R 2.15.3 by xtable 1.7-0 package
-% Mon Apr  8 09:29:56 2013
+% Fri May  3 10:50:17 2013
 \begin{table}[ht]
@@
 But I can show the comparison of the predictive accuracies (Fig. \ref{fig:resamp}).
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{verbatim}
+## $sh1
+\end{verbatim}
+\end{kframe}
+\includegraphics[width=\maxwidth]{figure/resamp1} 
+\begin{kframe}\begin{verbatim}
+## 
+## $sh2
+\end{verbatim}
+\end{kframe}
+\includegraphics[width=\maxwidth]{figure/resamp2} 
+\begin{kframe}\begin{verbatim}
+## 
+## $sh3
+\end{verbatim}
+\end{kframe}
+\includegraphics[width=\maxwidth]{figure/resamp3} 
+\begin{kframe}\begin{verbatim}
+## 
+## $spinks
+\end{verbatim}
+\end{kframe}
+\includegraphics[width=\maxwidth]{figure/resamp4}
+
+\end{knitrout}
 
@@
 In the interest of space/time, I'm only going to display results from one of the selected models from each method for each classification scheme.
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{verbatim}
+## $sh1
+\end{verbatim}
+\end{kframe}
+\includegraphics[width=\maxwidth]{figure/multi-map1} 
+\begin{kframe}\begin{verbatim}
+## 
+## $sh2
+\end{verbatim}
+\end{kframe}
+\includegraphics[width=\maxwidth]{figure/multi-map2} 
+\begin{kframe}\begin{verbatim}
+## 
+## $sh3
+\end{verbatim}
+\end{kframe}
+\includegraphics[width=\maxwidth]{figure/multi-map3} 
+\begin{kframe}\begin{verbatim}
+## 
+## $spinks
+\end{verbatim}
+\end{kframe}
+\includegraphics[width=\maxwidth]{figure/multi-map4}
+
+\end{knitrout}
 
@@
 
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error: object 'tnn.class' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error: object 'tnn.class.tiny' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error: object 'tnn.map' not found}}\end{kframe}
+\end{knitrout}
+
 \begin{figure}[t]
@@
 \end{figure}
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{verbatim}
+## $sh1
+\end{verbatim}
+\end{kframe}
+\includegraphics[width=\maxwidth]{figure/rf-map1} 
+\begin{kframe}\begin{verbatim}
+## 
+## $sh2
+\end{verbatim}
+\end{kframe}
+\includegraphics[width=\maxwidth]{figure/rf-map2} 
+\begin{kframe}\begin{verbatim}
+## 
+## $sh3
+\end{verbatim}
+\end{kframe}
+\includegraphics[width=\maxwidth]{figure/rf-map3} 
+\begin{kframe}\begin{verbatim}
+## 
+## $spinks
+\end{verbatim}
+\end{kframe}
+\includegraphics[width=\maxwidth]{figure/rf-map4}
+
+\end{knitrout}
 

### 2013-05-03 src/other_analyses/plotting_functions.r

@@ -1,214 +1,0 @@
-###############################################################################
-##
-##  plotting and tables for the turtle data
-##
-##  peter d smits
-##  psmits@uchicago.edu
-##
-###############################################################################
-
-require(ggplot2)
-require(reshape2)
-require(plyr)
-require(GGally)
-require(maps)
-require(mapproj) 
-require(ggmap)
-
-source('../src/support_functions.r')
-
-gap.plot <- function(gap) {
-  # wrapper to make a plot of the gap statistic results from cluster
-  #
-  # Args:
-  #   gap: table of results from clusGap
-  #
-  # Returns:
-  #   object of class ggplot
-
-  tab <- as.data.frame(gap$Tab)
-  tab <- cbind(id = as.numeric(rownames(tab)), tab)
-  gg <- ggplot(tab, aes(x = id, y = gap))
-  gg <- gg + geom_pointrange(mapping = aes(ymax = gap + SE.sim,
-                                           ymin = gap - SE.sim))
-  gg <- gg + geom_line()
-  gg <- gg + labs(x = 'Number of clusters',
-                  y = 'Gap statistic')
-  gg
-}
-
-mshape.plot <- function(fits, all.points = TRUE) {
-  shape <- as.data.frame(fits$mshape)
-  land <- land.frame(fits$rotated)
-  gg <- ggplot(shape, aes(x = V2, y = V1))
-  if (all.points) {
-    gg <- gg + geom_point(data = land,
-                          mapping = aes(x = y,
-                                        y = x),
-                          alpha = 0.5)
-  }
-  gg <- gg + geom_point(data = shape,
-                        mapping = aes(x = V2,
-                                      y = V1,
-                                      colour = 'red',
-                                      size = 1.5))
-  gg
-}
-
-map.plot <- function(data, label, map, coord = 'gilbert') {
-  # wrapper to make a map with labeled points on it
-  #
-  # Args:
-  #   data: points to be plotted
-  #         x, y coordinates
-  #   label: vector of factors (or coercible to factors) to color points by
-  #   map: map information (see ggplot::map_data)
-  #   coord: projection type
-  #
-  # Returns:
-  #   object of class ggplot
-
-  data <- cbind(as.data.frame(data), label = label)
-  gg <- ggplot(map, aes(x = long, y = lat, group = group))
-  gg <- gg + geom_polygon(fill = 'white',
-                          colour = 'black')
[…]

### 2013-05-03 src/support_functions.r

@@ -13,6 +13,8 @@ require(shapes)
 require(nnet)
 require(caret)
 require(e1071)
+require(pROC)
+source('../src/multiclass_roc.r')
 
 ## functions
 
@@ -109,14 +111,17 @@ data.maker <- function(gr, data, p = 0.75) {
 require(caret)
 ctrl <- trainControl(#method = 'LOOCV',
                      method = 'repeatedcv',
-                     #classProbs = TRUE,
+                     classProbs = TRUE,
+                     summaryFunction = multiClassSummary,
                      number = 10,
-                     repeats = 10)
+                     repeats = 10,
+                     returnResamp = 'all')
 
 nnetFuncs <- caretFuncs
 nnetFuncs$fit <- function(x, y, first, last, ...) {
   train(x, y, method = 'nnet', ...)
 }
+nnetFuncs$summary <- multiClassSummary
 nnet.ctrl <- rfeControl(functions = nnetFuncs,
                         method = 'repeatedcv',
                         repeats = 10,
@@ -125,12 +130,14 @@ nnet.ctrl <- rfeControl(functions = nnetFuncs,
                         returnResamp = 'final',
                         allowParallel = FALSE)
 
+rfFuncs$summary <- multiClassSummary
 rf.ctrl <- rfeControl(functions = rfFuncs,
                       method = 'repeatedcv',
                       repeats = 10,
                       number = 10,
                       verbose = FALSE, 
-                      returnResamp = 'final')
+                      returnResamp = 'final'
+                      )
 
 make.form <- function(vari, resp) {
   form <- vector(mode = 'list', length = length(vari))

### 2013-05-03 src/turtle_rf.r

@@ -38,7 +38,9 @@ trf <- Map(function(x, y) {
                , sizes = 1:floor(max.var)
                , rfeControl = rf.ctrl
                , ntree = 1000
-               , trControl = ctrl)}, 
+               , metric = 'ROC'
+               , trControl = ctrl
+               )}, 
            x = turtle.train, y = groups)
 
 set.seed(1)
@@ -48,6 +50,7 @@ trf.a <- Map(function(x, y) {
                  , sizes = 1:floor(max.ad)
                  , rfeControl = rf.ctrl
                  , ntree = 1000
+                 , metric = 'ROC'
                  , trControl = ctrl)},
              x = adult.train, y = groups)
 

### 2013-05-07 documents/turtle_presentation.Rnw
spatial component

@@
 # need to format all the map data so it plays nicely
-correct.map <- function(data) {
-  out <- lapply(data, 
-                function(x) {
-                  x$long <- as.numeric(as.character(x$long)) * -1
-                  x$lat <- as.numeric(as.character(x$lat))
-                  x})
-
-  out
-}
-
 map <- map_data('state')
@@
 map <- map[map$region == st, ]
-
-turtle.train <- correct.map(turtle.train)
-turtle.test <- correct.map(turtle.test)
-adult.train <- correct.map(adult.train)
-adult.test <- correct.map(adult.test)
 

### 2013-05-07 src/turtle_analysis.r

@@ -22,7 +22,7 @@ source('../src/support_functions.r')
 load('../src/cluster_res.RData')
 load('../src/supervised_misc.RData')
 load('../src/multi_boot_mod.RData')
-load('../src/nnet_boot_mod.RData')
+#load('../src/nnet_boot_mod.RData')
 load('../src/rf_boot_mod.RData')
 
 clean.mods <- function(models, lab = c('sh1', 'sh2', 'sh3', 'spinks')) {
@@ -41,100 +41,69 @@ clean.mods <- function(models, lab = c('sh1', 'sh2', 'sh3', 'spinks')) {
 }
 
 tm <- clean.mods(tmulti)
-tnn <- clean.mods(tnnet)
+tm.s <- clean.mods(tmulti.s)
+#tnn <- clean.mods(tnnet)
 trf <- clean.mods(trf)
+trf.s <- clean.mods(trf.s)
 
 tm.a <- clean.mods(tmulti.a)
-tnn.a <- clean.mods(tnnet.a)
+tm.a.s <- clean.mods(tmulti.a.s)
+#tnn.a <- clean.mods(tnnet.a)
 trf.a <- clean.mods(trf.a)
+trf.a.s <- clean.mods(trf.a.s)
+
+multi.analysis <- function(model, class, test) {
+  out <- list()
+  out$sel <- lapply(model, function(x)
+                    model.sel(lapply(x, function(y) y$finalModel)))
+  out$best <- mapply(function(sel, mod) mod[[as.numeric(rownames(sel)[1])]],
+                     sel = out$sel, mod = model,
+                     SIMPLIFY = FALSE)
+  out$class <- mapply(predict, out$best, test,
+                      MoreArgs = list(type = 'raw'), SIMPLIFY = FALSE)
+  out$conf <- Map(confusionMatrix,
+                  out$class, class)
+
+  out
+}
+
+rf.analysis <- function(model, class, test) {
+  out <- list()
+  out$varimp <- lapply(model, varImp)
+  out$re <- resamples(model)
+  out$redi <- resamples(out$re)
+  out$class <- mapply(predict, model, test,
+                      SIMPLIFY = FALSE)
+  out$conf <- Map(function(x, y) confusionMatrix(x$pred, y),
+                  x = out$class, y = class)
+  out
+}
 
 # multinomial logistic regression
-tm.sel <- lapply(tm, function(x)
[…]

### 2013-05-07 src/turtle_logistic.r

@@ -12,7 +12,6 @@
 require(MASS)
 require(nnet)
 require(cluster)
-require(e1071)
 require(caret)
 require(randomForest)
 require(pROC)
@@ -32,6 +31,7 @@ source('../src/support_functions.r')
 load('../src/supervised_misc.RData')
 
 set.seed(1)
+# change to be the design format and using recursive feature selection
 tmulti <- mapply(multi.train,
                  form = tform, data = turtle.train,
                  MoreArgs = list(method = 'multinom'
@@ -39,6 +39,14 @@ tmulti <- mapply(multi.train,
                                  , trControl = ctrl
                                  , maxit = 1000),
                  SIMPLFY = FALSE)
+set.seed(1)
+tmulti.s <- mapply(multi.train,
+                   form = tform.s, data = turtle.train,
+                   MoreArgs = list(method = 'multinom'
+                                   , metric = 'ROC'
+                                   , trControl = ctrl
+                                   , maxit = 1000),
+                   SIMPLFY = FALSE)
 
 set.seed(1)
 tmulti.a <- mapply(multi.train,
@@ -48,5 +56,15 @@ tmulti.a <- mapply(multi.train,
                                    , trControl = ctrl
                                    , maxit = 1000),
                    SIMPLIFY = FALSE)
+set.seed(1)
+tmulti.a.s <- mapply(multi.train,
+                     form = tform.a.s, data = adult.train,
+                     MoreArgs = list(method = 'multinom'
+                                     , metric = 'ROC'
+                                     , trControl = ctrl
+                                     , maxit = 1000),
+                     SIMPLIFY = FALSE)
 
-save(tmulti, tmulti.a, file = 'multi_boot_mod.RData')
+save(tmulti, tmulti.s,
+     tmulti.a, tmulti.a.s,
+     file = 'multi_boot_mod.RData')

### 2013-05-07 src/turtle_super_wrap.sh

@@ -2,7 +2,7 @@
 
 nohup nice R CMD BATCH --vanilla turtle_supervised_data.r
 nohup nice R CMD BATCH --vanilla turtle_logistic.r
-nohup nice R CMD BATCH --vanilla turtle_nnet.r
+#nohup nice R CMD BATCH --vanilla turtle_nnet.r
 nohup nice R CMD BATCH --vanilla turtle_rf.r
 nohup nice R CMD BATCH --vanilla turtle_analysis.r
 

### 2013-05-07 src/turtle_supervised_data.r

@@ -12,7 +12,6 @@
 require(MASS)
 require(nnet)
 require(cluster)
-require(e1071)
 require(caret)
 require(randomForest)
 require(pROC)
@@ -36,14 +35,18 @@ source('../src/turtle_mung.r')
 max.var <- nrow(turtle.info) / 50
 max.ad <- nrow(turtle.adult) / 50
 tvar <- paste('PC', 1:max.var, sep = '')
+tvar.s <- c('lat', 'long', tvar)
 tvar.a <- paste('PC', 1:max.ad, sep = '')
+tvar.a.s <- c('lat', 'long', tvar.a)
 groups <- list('sh1', 'sh2', 'sh3', 'spinks')
 tform <- lapply(groups, function(x, y) make.form(y, x), y = tvar)
+tform.s <- lapply(groups, function(x, y) make.form(y, x), y = tvar.s)
 tform.a <- lapply(groups, function(x, y) make.form(y, x), y = tvar.a)
+tform.a.s <- lapply(groups, function(x, y) make.form(y, x), y = tvar.a.s)
 
 # including juvies
 set.seed(1)
-in.train <- data.maker(unlist(groups), turtle.info)
+in.train <- as.data.frame(data.maker(unlist(groups), turtle.info))
 turtle.train <- lapply(in.train, function(x) turtle.info[x, ])
 turtle.test <- lapply(in.train, function(x) turtle.info[-x, ])
 
@@ -58,6 +61,17 @@ adult.test <- lapply(ad.train, function(x) turtle.adult[-x, ])
 
 ad.class <- Map(function(x, y) colnames(x) %in% y, adult.test, groups)
 ad.class <- Map(function(x, y) x[, y], adult.test, ad.class)
+
+turtle.design <- Map(function(x, y) {
+                     cbind(x$lat, x$long,
+                           x[, 1:max.var],
+                           cate = x[, y])},
+                     x = turtle.train, y = groups)
+adult.design <- Map(function(x, y) {
+                    cbind(x$lat, x$long,
+                          x[, 1:max.ad],
+                          cate = x[, y])},
+                    x = adult.train, y = groups)
 
 # O
 save.image('supervised_misc.RData')

### 2013-05-10 documents/turtle_presentation.Rnw
new analysis. need to fix and run randomization.

@@
 
-#source('../src/plotting_functions.r')
+source('../src/plotting_functions.r')
 #load('../src/turtle_analysis.r')
@@
   % a little biogeography
+  \begin{figure}[h]
+    \centering
+    \captionsetup{justification = raggedleft, slc = off}
+    \includegraphics[width = 0.8\textwidth, keepaspectratio = true]{figure/turtle}
+    \caption*{wikimedia}
+    \label{fig:turtle}
+  \end{figure}
 \end{frame}
@@
   \frametitle{Phylogenetic hypotheses}
+
+  Number of subspecies and where they occur.
   % images of the different phylobiogeographic hypotheses from the different spinks papers
@@
   \frametitle{Methods: morphometrics}
-  % two columns, text on one side and idealized landmarks on the other
-  % ken probably has an image for the landmarks
-  % these landmarks have been used before in other studies
   \begin{columns}
@@
       \begin{itemize}
+        %\item \Sexpr{nrow(turtle.adult)} adult individuals
         \item plastral (``belly'') shape
@@
         \item total 13 landmarks, 7 on bilat axis, 6 off
-        %\item \Sexpr{nrow(turtle.adult)} adult individuals
         \item geographic information known/inferred
@@
     \begin{column}{0.5\textwidth}
-      idealized plastral landmarks
+      \centering
+      \includegraphics[width = \textwidth, keepaspectratio = true]{figure/plastra}
     \end{column}
@@
   \frametitle{Methods: unsupervised learning}
+
+  Fancy way of saying clustering or density estimation.
+
+  Partitioning around mediods (PAM) compared with ``gap'' statistic.
+
+  (dissimilarity based) Evidence accumulation clustering
 \end{frame}
@@
   \frametitle{Methods: supervised learning}
+
+  Fancy way of saying classification and regression.
+
+  Here, features (principal components) predict class (subspecific assignment).
+
+  Multnomial logistic regression
+
+  Random forests
 \end{frame}
+
+\begin{frame}
+  \frametitle{Methods: model training and selection}
+  Unknown appropriate number of features to ``best'' predict class. Want to minimize false positive, while maximizing true positive.
+
+  Split data set, 75-25, training and testing.
+
+  Tuning parameters via grid-search. Uncertainty via 10-fold cross-validation.
+
+  Multinomial logistic model selected via min AICc. Random forest model via max AUC ROC.
+
+\end{frame}
+%\begin{frame}
+%  \frametitle{Methods: ROC}
+%\end{frame}
 
@@
   \frametitle{Results: mophometrics}
+<<gm, eval = FALSE>>=
+morph <- ggpairs(turtle.adult, columns = c(1:3),
+                 upper = 'blank')
+fits <- procGPA(turtle.land.adult)
+land <- mshape.plot(fits, all.points = TRUE)
+morph <- putPlot(morph, land, 1, 3)
+morph
+@
 \end{frame}
@@
   \frametitle{Results: gap clustering}
+<<gap, eval = FALSE>>=
+gap <- gap.plot(tadult.gap)
+gap
+@
 \end{frame}
@@
   \frametitle{Results: d-EAC}
+
 \end{frame}
@@
   \frametitle{Results: multinomial logistic regression}
-\end{frame}
 
-\begin{frame}
-  \frametitle{Results: neural networks}
 \end{frame}
@@
   \frametitle{Results: random forests}
+  % plot the ROC curves
+  % four facets, multiple lines
 \end{frame}

### 2013-05-10 src/turtle_random.r

@@ -35,18 +35,18 @@ pred.dist <- function(best, train, nsim = 1000) {
     data
   }
 
-  re.dist <- function(best, train, nsim = 1000) {
-    groups <- names(train)
+  re.dist <- function(best, train, group, nsim = 1000) {
     pred <- vector(mode = 'list', length = nsim)
     for(ii in seq(nsim)) {
-      re <- resample.class(data = train, class = groups)
+      re <- resample.class(data = train, class = group)
       re.pred <- predict(best, re, type = 'raw')
       pred[[ii]] <- re.pred
     }
     pred
   }
 
-  out <- mapply(re.dist, best = best, train = train,
+  groups <- names(train)
+  out <- mapply(re.dist, best = best, train = train, group = groups,
                 MoreArgs = list(nsim = nsim), SIMPLFY = FALSE)
 
   out

### 2013-05-10 documents/turtle_presentation.Rnw
lots of presentation work. need to write train randomizer.

@@
 require(xtable)
+require(grid)
 require(ggplot2)
@@
 require(GGally)
+require(scales)
 
@@
                fig.height = 3, fig.width = 4
-               , eval = FALSE
+               , eval = TRUE
                )
 
+theme_set(theme_bw())
+
 source('../src/plotting_functions.r')
-#load('../src/turtle_analysis.r')
+load('../data/turtle_analysis.RData')
 @
@@
   % image of example taxa that are morphologically identical and are only differentiated via genetic means
+  %
+  How much of cryptic diversity is just a function of sample size?
 \end{frame}
@@
 \section{Results}
-\begin{frame}
+\begin{frame}[fragile]
   \frametitle{Results: mophometrics}
-<<gm, eval = FALSE>>=
+<<gm, results = 'hide', fig.width = 3.5, fig.height = 3.5, eval = TRUE>>=
 morph <- ggpairs(turtle.adult, columns = c(1:3),
-                 upper = 'blank')
+                 upper = 'blank',
+                 params = c(LabelSize = 2, gridLabelSize = 2, size = 1))
 fits <- procGPA(turtle.land.adult)
-land <- mshape.plot(fits, all.points = TRUE)
+land <- mshape.plot(fits)
 morph <- putPlot(morph, land, 1, 3)
@@
 @
+
 \end{frame}
 
-\begin{frame}
+\begin{frame}[fragile]
   \frametitle{Results: gap clustering}
-<<gap, eval = FALSE>>=
+<<gap, fig.width = 4, fig.height = 3,  eval = TRUE>>=
 gap <- gap.plot(tadult.gap)
+gap <- gap + theme(axis.title = element_text(size = 7.5),
+                   axis.text = element_text(size = 6))
 gap
@@
 % model selection is unnecessary because of time constraints
-\begin{frame}
-  \frametitle{Results: multinomial logistic regression}
+\begin{frame}[fragile]
+  \frametitle{ROC}
+<<roc, eval = TRUE>>=
+rf.rocs <- lapply(trf.a, function(x) {
+                  rr <- x$results$ROC
+                  rr})
+rf.rocs <- ldply(rf.rocs)
+multi.rocs <- lapply(tm.a, function(x) {
+                     lapply(x, function(y) {
+                            rr <- max(y$results$ROC)
+                            rr})})
+multi.rocs <- ldply(lapply(lapply(multi.rocs, ldply), t))
+
+names(rf.rocs) <- names(multi.rocs) 
+roc.mod <- rbind(rf.rocs, multi.rocs)
+mod.names <- c(rep('rf', nrow(rf.rocs)), rep('multi', nrow(multi.rocs)))
+roc.mod <- cbind(mod.names, roc.mod)
+roc.mod <- melt(roc.mod)
+roc.mod$variable <- as.numeric(roc.mod$variable)
+
+ggroc <- ggplot(roc.mod, aes(x = variable, y = value, color = mod.names))
+ggroc <- ggroc + geom_line()
+ggroc <- ggroc + scale_x_continuous(breaks = seq(max(roc.mod$variable)))
+ggroc <- ggroc + scale_color_discrete(labels = c('multinomial', 
+                                                 'random forest'))
+ggroc <- ggroc + theme(legend.title = element_blank(),
+                       legend.position = 'right',
+                       legend.margin = unit(0, 'cm'),
+                       legend.text = element_text(size = 5),
+                       axis.title = element_text(size = 8),
+                       axis.text = element_text(size = 5))
+ggroc <- ggroc + labs(x = '# of features', y = 'AUC ROC')
+ggroc <- ggroc + facet_wrap(~.id)
+ggroc
+@
 
@@
 \begin{frame}
-  \frametitle{Results: random forests}
-  % plot the ROC curves
-  % four facets, multiple lines
+  \frametitle{Are the AUC ROC values meaningful?}
 \end{frame}
 
-\section{Discussion}
 \begin{frame}
@@
 
+\begin{frame}[fragile]
+  \frametitle{Variable importance}
+  % ggpairs from the best random forest model
+<<imp, fig.width = 3.5, fig.height = 3.5, eval = TRUE>>=
+most.imp <- trf.a$spinks$optVariables
+ww <- 'spinks'
+ggimp <- ggpairs(turtle.adult,
+                 columns = c(most.imp[1:3], ww), 
+                 colour = 'spinks',
+                 upper = 'blank',
+                 params = c(LabelSize = 2, gridLabelSize = 2, size = 1))
+ggimp
+@
+\end{frame}
+
+\section{Discussion}
 \begin{frame}

### 2013-05-10 src/plotting_functions.r

@@ -37,21 +37,14 @@ gap.plot <- function(gap) {
   gg
 }
 
-mshape.plot <- function(fits, all.points = TRUE) {
+mshape.plot <- function(fits) {
   shape <- as.data.frame(fits$mshape)
   land <- land.frame(fits$rotated)
-  gg <- ggplot(shape, aes(x = V2, y = V1))
-  if (all.points) {
-    gg <- gg + geom_point(data = land,
-                          mapping = aes(x = y,
-                                        y = x),
-                          alpha = 0.5)
-  }
-  gg <- gg + geom_point(data = shape,
-                        mapping = aes(x = V2,
-                                      y = V1,
-                                      colour = 'red',
-                                      size = 1.5))
+  gg <- ggplot(land, aes(x = y, y = x)) + geom_point(alpha = 0.5,
+                                                     size = 0.7)
+  gg <- gg + geom_point(data = shape, mapping = aes(x = V2,
+                                                    y = V1,
+                                                    colour = 'red'))
   gg
 }
 
@@ -211,4 +204,7 @@ shorten <- function(x, n = 3) {
   # helper function for creating nice confusion matrix with xtable
   let2str(grab.letters(x, n))
 }
-  
+ 
+
+# model comparison plots that are actually meaningful
+

### 2013-05-10 src/randomize_funcs.r

@@ -1,0 +1,68 @@
+##############################################################################
+##
+##  analysis of turtlee plastron data
+##  randomization of the model fits to create a distribution of AUC ROC
+##
+##  peter d smits
+##  psmits@uchicago.edu
+##
+###############################################################################
+
+require(plyr)
+require(cluster)
+require(nnet)
+require(randomForest)
+require(caret)
+require(e1071)
+require(pROC)
+require(MuMIn)
+
+source('../src/support_functions.r')
+
+#load('../data/turtle_analysis.RData')
+
+# randomly reassign the classes of the training data set
+# refit the best sized model to the permuted training set
+# this creates a distribution of ROC values against "random"
+#
+# are the models better than random?
+# can then see if ROC is significant
+
+resample.class <- function(data, class) {
+  data[, class] <- sample(data[, class])
+  data
+}
+
+# write these as more generic
+resample.rf <- function(model, training, class) {
+  re.tr <- resample.class(data = training, class = class)
+  new.train <- train(form = as.formula(paste0(class, '~',
+                                       paste(model$optVariables, 
+                                             collapse = '+'))),
+                     data = re.tr,
+                     method = 'rf',
+                     metric = 'ROC',
+                     trControl = ctrl,
+                     ntree= 10000)
+
+  new.train
+}
+
+resample.multi <- function(model, training, class) {
+  re.tr <- resample.class(data = training, class = class)
+  new.train <- train(form ,
+                     data = re.tr,
+                     method = 'mulinom',
+                     metric = 'ROC',
+                     trControl = ctrl,
+                     maxit = 10000)
+
+  new.train
+}
+
+
+roc.dist <- function(models) {
+}
+
+predict.dist <- function(models, test) {
+}

### 2013-05-17 documents/turtle_presentation.Rnw
expanded presentaiton. ROC slides.

@@
 \usepackage{amsmath, amsthm}
-\usepackage{parskip, microtype, graphicx, caption}
+\usepackage{parskip, microtype, graphicx, caption, multirow}
 
@@
 
-\title{How cryptic is cryptic diversity? Machine learning approaches to fine scale variation in the morphology of \textit{Emys marmorata}.}
+\title{How cryptic is cryptic diversity? \newline Machine learning approaches to fine scale variation in the morphology of \textit{Emys marmorata}.}
 \author[shortname]{Peter D Smits\inst{1}, 
@@
   %
-  How much of cryptic diversity is just a function of sample size?
+  Crytic species are species delimitated via molecular means which were not/cannot be identified via morphology.
+
+  How much of cryptic diversity is just a function of sample size and/or method?
+
 \end{frame}
@@
 
-  Number of subspecies and where they occur.
+  % Number of subspecies and where they occur.
   % images of the different phylobiogeographic hypotheses from the different spinks papers
+
 \end{frame}
@@
     \begin{column}{0.5\textwidth}
+      \begin{figure}[h]
         \centering
+        \captionsetup{justification = raggedleft, slc = off}
         \includegraphics[width = \textwidth, keepaspectratio = true]{figure/plastra}
+        \caption*{\scriptsize{Angielczyk \textit{et al.} 2011 \textit{Evolution}}}
+        \label{fig:plast}
+      \end{figure}
     \end{column}
@@
 \begin{frame}
-  \frametitle{Methods: unsupervised learning}
+  \frametitle{Unsupervised learning}
 
@@
 \begin{frame}
-  \frametitle{Methods: supervised learning}
+  \frametitle{Supervised learning}
 
@@
 
-  Multnomial logistic regression
+  Multinomial logistic regression
 
@@
 \begin{frame}
-  \frametitle{Methods: model training and selection}
+  \frametitle{Model training and selection}
   Unknown appropriate number of features to ``best'' predict class. Want to minimize false positive, while maximizing true positive.
@@
 
-  Tuning parameters via grid-search. Uncertainty via 10-fold cross-validation.
+  Tuning parameters via grid-search. Uncertainty via 10-fold cross-validation. Selection viamax AUC ROC.
 
-  Multinomial logistic model selected via min AICc. Random forest model via max AUC ROC.
+  Best multinomial logistic model selected via min AICc. Best random forest model via max AUC ROC.
 
 \end{frame}
-%\begin{frame}
-%  \frametitle{Methods: ROC}
-%\end{frame}
+
+\begin{frame}
+  \frametitle{ROC and confusion matrices}
+
+  \begin{center}
+    \begin{tabular}[c]{ p{2cm} c | p{2cm} | p{2cm} |}
+      \cline{3-4} 
+      & & \multicolumn{2}{ c |}{Predicted class} \\ \cline{3-4}
+      & & 1 & 0 \\ \hline
+      \multicolumn{1}{| c |}{\multirow{2}{*}{Actual class}}
+      & 1 & TRUE \newline POSITIVE & FALSE \newline NEGATIVE \\ \cline{2-4}
+      \multicolumn{1}{| c |}{} & 0 & FALSE \newline POSITIVE & TRUE \newline NEGATIVE \\
+      \hline
+    \end{tabular}
+  \end{center}
+
+\end{frame}
+
+\begin{frame}
+  \frametitle{ROC}
+  \begin{columns}
+    \begin{column}{0.4\textwidth}
+      \begin{itemize}
+        \item receiver operating characteristic
+        \item true positive rate or sensitivity: \(\frac{TP}{TP + FN}\)
+        \item false positive rate or \newline 1 - specificity: \(\frac{FP}{FP + TN}\)
+        \item multiclass, all-against one (Hand and Till 2001 \textit{Machine Learning})
+        \item area under curve model summary statistic (min 0.5)
+      \end{itemize}
+    \end{column}
+    \begin{column}{0.6\textwidth}
+      \begin{figure}[h]
+        \centering
+        \captionsetup{justification = raggedleft, slc = off}
+        \includegraphics[width = \textwidth, keepaspectratio = true]{figure/wiki_roc}
+        \caption*{\scriptsize{wikimedia}}
+        \label{fig:roc}
+      \end{figure}
+    \end{column}
+  \end{columns}
+
+\end{frame}
 
@@
 \begin{frame}
-  \frametitle{Best classification scheme}
+  \frametitle{Best classification scheme?}
 \end{frame}
@@
         \item COLLECTIONS
+        \item FUNDING AGENCIES
       \end{itemize}

### 2013-06-06 documents/turtle_presentation.Rnw
presentation and results going really well

@@
 require(scales)
+require(boot)
 
@@
 theme_set(theme_bw())
+cbp <- c(#'#999999', 
+         '#E69F00', '#56B4E9', '#009E73', 
+         '#F0E442', '#0072B2', '#D55E00', '#CC79A7')
 
+source('../src/support_functions.r')
 source('../src/plotting_functions.r')
 load('../data/turtle_analysis.RData')
+load('../data/turtle_gen.RData')
 @
@@
 map <- map_data('state')
-st <- c('california', 'oregon', 'washington')
+st <- c('california'
+        , 'oregon'
+        #, 'washington'
+        )
 map <- map[map$region == st, ]
@@
 labs <- Map(function(x, n) x[, n],
-            x = turtle.train, n = names(turtle.train))
+            x = adult.train, n = names(adult.train))
+
+turtle.adult$long[which(turtle.adult$long > -100)] <- turtle.adult$long[which(turtle.adult$long > -100)] - 100
+
+adult.train$spinks$long[which(adult.train$spinks$long > -100)] <- adult.train$spinks$long[which(adult.train$spinks$long > -100)] - 100
+
 @
@@
 
-  Tuning parameters via grid-search. Uncertainty via 10-fold cross-validation. Selection viamax AUC ROC.
+  Tuning parameters via grid-search. Uncertainty via 10-fold cross-validation. Selection via max AUC ROC.
 
@@
       \begin{itemize}
-        \item receiver operating characteristic
         \item true positive rate or sensitivity: \(\frac{TP}{TP + FN}\)
@@
         \item multiclass, all-against one (Hand and Till 2001 \textit{Machine Learning})
-        \item area under curve model summary statistic (min 0.5)
       \end{itemize}
@@
 \begin{frame}
-  \frametitle{Results: d-EAC}
+  \frametitle{Second best cluster}
+<<>>=
+gap.second <- pam(as.dist(turtle.adult.dist), k = 2)
+gap.map <- map.plot(data = turtle.adult,
+                    label = gap.second$clustering,
+                    map = map)
+gap.map <- gap.map + scale_fill_manual(values = cbp)
+gap.map <- gap.map + theme(legend.position = 'none')
+gap.map
+@
 
@@
 
-% model selection is unnecessary because of time constraints
+%\begin{frame}
+%  \frametitle{Results: d-EAC}
+%
+%\end{frame}
+
 \begin{frame}[fragile]
@@
 ggroc <- ggroc + scale_x_continuous(breaks = seq(max(roc.mod$variable)))
-ggroc <- ggroc + scale_color_discrete(labels = c('multinomial', 
-                                                 'random forest'))
+ggroc <- ggroc + scale_color_manual(labels = c('multinomial', 
+                                                 'random forest'),
+                                      values = cbp)
 ggroc <- ggroc + theme(legend.title = element_blank(),
@@
 
+%\begin{frame}
+%  \frametitle{Are the AUC ROC values meaningful?}
+%\end{frame}
+
 \begin{frame}
-  \frametitle{Are the AUC ROC values meaningful?}
+  \frametitle{Generalize}
+<<>>=
+oo <- Reduce(cbind, Map(function(x) x$t, rr))
+colnames(oo) <- groups
+dd <- melt(oo)
+gdist <- ggplot(dd, aes(x = value, fill = X2)) + geom_density(alpha = 0.3)
+gdist <- gdist + theme(legend.title = element_blank())
+gdist <- gdist + scale_fill_manual(values = cbp)
+
+gdist
+@
+
 \end{frame}
@@
   \frametitle{Best classification scheme?}
+<<>>=
+gg <- class.map(adult.train$spinks, adult.test$spinks, labs$spinks, 
+                trf.a.analysis$class$spinks$pred, map)
+gg <- gg + scale_fill_manual(values = cbp)
+gg
+@
+
 \end{frame}

### 2013-06-06 src/turtle_generalize.r

@@ -1,0 +1,42 @@
+##############################################################################
+##
+##  analysis of turtlee plastron data
+##  analysis of machine learning results
+##
+##  peter d smits
+##  psmits@uchicago.edu
+##
+###############################################################################
+require(plyr)
+require(cluster)
+require(nnet)
+require(randomForest)
+require(caret)
+require(e1071)
+require(pROC)
+require(MuMIn)
+require(boot)
+
+source('../src/support_functions.r')
+
+load('../data/turtle_analysis.RData')
+
+# boot strap the generalization
+
+boot.roc <- function(data, indicies) {
+  data <- data[indicies, ]
+  pp <- data[, seq(ncol(data) - 1)]
+  tt <- data[, ncol(data)]
+  return(allvone(pp, tt))
+}
+
+rr <- list()
+for (ii in seq(length(groups))) {
+  oo <- cbind(trf.a.analysis$class[[ii]],
+              test = adult.test[[ii]][, groups[[ii]]])
+
+  rr[[ii]] <- boot(data = oo, statistic = boot.roc, R = 1000)
+}
+names(rr) <- groups
+
+save(rr, file = '../data/turtle_gen.RData')

### 2013-06-12 documents/turtle_presentation.Rnw
presentation very close to finished

@@
 \usepackage{amsmath, amsthm}
-\usepackage{parskip, microtype, graphicx, caption, multirow}
+\usepackage{parskip, microtype, graphicx, caption, multirow, subcaption}
 
@@
 require(boot)
+
+require(shapes)
+require(geomorph)
 
@@
 \begin{frame}
-  \frametitle{Morphological hypotheses}
+  \frametitle{Morphological hypothesis}
+
+  \begin{figure}[h]
+    \centering
+    \captionsetup{justification = raggedleft, slc = off}
+    \includegraphics[height = 0.8\textheight, keepaspectratio = true]{figure/seeliger}
+    \caption*{\scriptsize{Seeliger 1945 \textit{Copeia}}}
+    \label{fig:seeliger}
+  \end{figure}
+  % basic north/south stuff
+  % earlier work by the masters student
+
 \end{frame}
@@
 
+  \begin{figure}[h]
+    \centering
+    \captionsetup{justification = raggedleft, slc = off}
+    \includegraphics[height = 0.8\textheight, keepaspectratio = true]{figure/spinks10}
+    \caption*{\scriptsize{Spinks \textit{et al.} 2010 \textit{Molec. Ecol}}}
+    \label{fig:spinks}
+  \end{figure}
   % Number of subspecies and where they occur.
@@
 
-  (dissimilarity based) Evidence accumulation clustering
+  %(dissimilarity based) Evidence accumulation clustering
+
 \end{frame}
@@
 
-  Fancy way of saying classification and regression.
+  Fancy way of saying classification (and regression).
 
@@
 
-  Multinomial logistic regression
+  Multinomial logistic regression and random forests.
 
-  Random forests
 \end{frame}
@@
                     map = map)
-gap.map <- gap.map + scale_fill_manual(values = cbp)
-gap.map <- gap.map + theme(legend.position = 'none')
+gap.map <- gap.map + scale_colour_manual(values = cbp)
+gap.map <- gap.map + theme(legend.position = 'none',
+                           legend.text = element_text(size = 5),
+                           axis.title = element_text(size = 8),
+                           axis.text = element_text(size = 5))
 gap.map
@@
 gdist <- ggplot(dd, aes(x = value, fill = X2)) + geom_density(alpha = 0.3)
-gdist <- gdist + theme(legend.title = element_blank())
-gdist <- gdist + scale_fill_manual(values = cbp)
+gdist <- gdist + theme(legend.title = element_blank(),
+                       legend.margin = unit(0, 'cm'),
+                       legend.text = element_text(size = 5),
+                       axis.title = element_text(size = 8),
+                       axis.text = element_text(size = 5))
+gdist <- gdist + scale_fill_manual(name = '', values = cbp)
 
@@
                 trf.a.analysis$class$spinks$pred, map)
-gg <- gg + scale_fill_manual(values = cbp)
+gg <- gg + scale_colour_manual(name = '', values = cbp)
+gg <- gg + theme(#legend.position = 'none',
+                 legend.margin = unit(0, 'cm'),
+                 legend.text = element_text(size = 5),
+                 axis.title = element_text(size = 8),
+                 axis.text = element_text(size = 5))
 gg
@@
                  params = c(LabelSize = 2, gridLabelSize = 2, size = 1))
+
+#putPlot
 ggimp
 @
+\end{frame}
+
+\begin{frame}
+  \frametitle{Shape across most important features}
+
+<<tps, include = FALSE>>=
+fst.max <- which.max(turtle.adult[, most.imp[1]])
+fst.min <- which.min(turtle.adult[, most.imp[1]])
+
+snd.max <- which.max(turtle.adult[, most.imp[2]])
+snd.min <- which.min(turtle.adult[, most.imp[2]])
+
+par(mar = c(0, 0, 1, 1) + 0.1, oma = c(5, 4, 0, 0) + 0.1)
+
+tpsgrid(turtle.land.adult[, , fst.max], mt)
+tpsgrid(turtle.land.adult[, , fst.min], mt)
+
+tpsgrid(turtle.land.adult[, , snd.max], mt)
+tpsgrid(turtle.land.adult[, , snd.min], mt)
+
+imp <- cbind(turtle.adult[, most.imp[1:2]], sp = turtle.adult$spinks)
+imp.pca <- ggplot(imp, aes(x = PC3, y = PC8, colour = sp)) + geom_point()
+imp.pca <- imp.pca + scale_colour_manual(name = '', values = cbp)
+imp.pca
+@
+
+% show tps grids to the side of the PCA plot of the first two most important axes
+\begin{figure}[h]
+  \centering
+  \hspace*{\fill}
+  \begin{subfigure}[h]{0.3\textwidth}
+    \centering
+    \includegraphics[width = \textwidth]{figure/tps2}    
+  \end{subfigure}
+  \begin{subfigure}[h]{0.3\textwidth}
+    \centering
+    \includegraphics[width = \textwidth]{figure/tps5}    
+  \end{subfigure}
+  \begin{subfigure}[h]{0.3\textwidth}
+    \centering
+    \includegraphics[width = \textwidth]{figure/tps1}    
+  \end{subfigure}
+  \hspace*{\fill}
+\end{figure}
+
+\end{frame}
+
+\begin{frame}[fragile]
+  \frametitle{Mean shape of classes}
+
+<<fig.width = 3.75, fig.height = 3.75>>=
+mt <- mshape(turtle.land.adult)
+# get mean of each class
+spi <- turtle.adult$spinks
+wspi <- lapply(levels(spi), function(x, y) which(y == x), y = spi)
+mspi <- lapply(wspi, function(x, y) mshape(y[, , x]), y = turtle.land.adult)
+par(mfrow = c(2, 2), mar = c(0, 0, 1, 1) + 0.1, oma = c(5, 4, 0, 0) + 0.1,
+    xaxt = 'n', yaxt = 'n')
+mapply(function(x, m, n) {
+      tpsgrid(x, m)
+      title(main = paste(n))}, x = mspi, n = levels(spi),
+      MoreArgs = list(m = mt), SIMPLIFY = FALSE)
+par(mfrow = c(1, 1))
+@
+
 \end{frame}
@@
   \frametitle{Future}
+
+  \begin{itemize}
+    \item illustration of morphological validation of previously cryptic variation
+      \begin{itemize}
+        \item the concordance is remarkable
+        \item large sample sizes can be difficult
+      \end{itemize}
+    \item utility of large data, machine learning methods 
+    \item unsupervised methods for when no explicit hypothesis -- nonparametric Bayes
+    \item cause of interclass variation -- local adaptation? pure isolation?
+  \end{itemize}
+
 \end{frame}

### 2013-06-13 documents/turtle_presentation.Rnw
first full draft of evolution talk

@@
 
-  %(dissimilarity based) Evidence accumulation clustering
+  Minimize sum of dissimilarities between points and medoids.
+
+  ``Gap'' is analogous to goodness-of-clustering.
 
@@
 
-  Here, features (principal components) predict class (subspecific assignment).
+  Features (principal components) predict class (subspecific assignment).
 
@@
   \frametitle{Model training and selection}
-  Unknown appropriate number of features to ``best'' predict class. Want to minimize false positive, while maximizing true positive.
 
-  Split data set, 75-25, training and testing.
-
-  Tuning parameters via grid-search. Uncertainty via 10-fold cross-validation. Selection via max AUC ROC.
-
-  Best multinomial logistic model selected via min AICc. Best random forest model via max AUC ROC.
+  \begin{itemize}
+    \item split into training and testing sets, 75-25.
+    \item tuning parameters via grid-search
+    \item uncertainty via 10-fold CV
+    \item model selection
+      \begin{itemize}
+        \item multinomial logistic regression: min AICc
+        \item random forest: max ROC
+      \end{itemize}
+  \end{itemize}
 
@@
 land <- mshape.plot(fits)
+land <- land + coord_equal()
 morph <- putPlot(morph, land, 1, 3)
@@
 
-%\begin{frame}
-%  \frametitle{Results: d-EAC}
-%
-%\end{frame}
-
 \begin{frame}[fragile]
-  \frametitle{ROC}
+  \frametitle{Model selection via ROC}
 <<roc, eval = TRUE>>=
@@
 \begin{frame}
-  \frametitle{Generalize}
+  \frametitle{Generalize using best random forest model}
 <<>>=
@@
 \begin{frame}
-  \frametitle{Best classification scheme?}
+  \frametitle{Best classification scheme via RF model results}
 <<>>=
@@
 \begin{frame}[fragile]
-  \frametitle{Variable importance}
+  \frametitle{Mean shape of classes}
+
+<<fig.width = 3.75, fig.height = 3.75>>=
+# get mean of each class
+spi <- turtle.adult$spinks
+wspi <- lapply(levels(spi), function(x, y) which(y == x), y = spi)
+mspi <- lapply(wspi, function(x, y) mshape(y[, , x]), y = turtle.land.adult)
+par(mfrow = c(2, 2), mar = c(0, 0, 1, 1) + 0.1, oma = c(5, 4, 0, 0) + 0.1,
+    xaxt = 'n', yaxt = 'n')
+mapply(function(x, m, n) {
+      tpsgrid(x, m, mag = 2)
+      title(main = paste(n))}, x = mspi, n = levels(spi),
+      MoreArgs = list(m = mt), SIMPLIFY = FALSE)
+par(mfrow = c(1, 1))
+@
+
+\end{frame}
+
+\begin{frame}[fragile]
+  \frametitle{Variable importance of random forest model}
   % ggpairs from the best random forest model
@@
 \begin{frame}
-  \frametitle{Shape across most important features}
+  \frametitle{Shape across PC3}
 
-<<tps, include = FALSE>>=
+<<fig.width = 5>>=
+mt <- mshape(turtle.land.adult)
+
 fst.max <- which.max(turtle.adult[, most.imp[1]])
@@
 
-par(mar = c(0, 0, 1, 1) + 0.1, oma = c(5, 4, 0, 0) + 0.1)
-
-tpsgrid(turtle.land.adult[, , fst.max], mt)
-tpsgrid(turtle.land.adult[, , fst.min], mt)
-
-tpsgrid(turtle.land.adult[, , snd.max], mt)
-tpsgrid(turtle.land.adult[, , snd.min], mt)
-
-imp <- cbind(turtle.adult[, most.imp[1:2]], sp = turtle.adult$spinks)
-imp.pca <- ggplot(imp, aes(x = PC3, y = PC8, colour = sp)) + geom_point()
-imp.pca <- imp.pca + scale_colour_manual(name = '', values = cbp)
-imp.pca
+par(mfrow = c(1, 2), mar = c(0, 0, 1, 1) + 0.1, oma = c(5, 4, 0, 0) + 0.1,
+    xaxt = 'n', yaxt = 'n')
+tpsgrid(turtle.land.adult[, , fst.min], mt, mag = 2)
+title(main = 'min observed')
+tpsgrid(turtle.land.adult[, , fst.max], mt, mag = 2)
+title(main = 'max observed')
+par(mfrow = c(1, 1))
 @
-
-% show tps grids to the side of the PCA plot of the first two most important axes
-\begin{figure}[h]
-  \centering
-  \hspace*{\fill}
-  \begin{subfigure}[h]{0.3\textwidth}
-    \centering
-    \includegraphics[width = \textwidth]{figure/tps2}    
-  \end{subfigure}
-  \begin{subfigure}[h]{0.3\textwidth}
-    \centering
-    \includegraphics[width = \textwidth]{figure/tps5}    
-  \end{subfigure}
-  \begin{subfigure}[h]{0.3\textwidth}
-    \centering
-    \includegraphics[width = \textwidth]{figure/tps1}    
-  \end{subfigure}
-  \hspace*{\fill}
-\end{figure}
 
@@
 
-\begin{frame}[fragile]
-  \frametitle{Mean shape of classes}
-
-<<fig.width = 3.75, fig.height = 3.75>>=
-mt <- mshape(turtle.land.adult)
-# get mean of each class
-spi <- turtle.adult$spinks
-wspi <- lapply(levels(spi), function(x, y) which(y == x), y = spi)
-mspi <- lapply(wspi, function(x, y) mshape(y[, , x]), y = turtle.land.adult)
-par(mfrow = c(2, 2), mar = c(0, 0, 1, 1) + 0.1, oma = c(5, 4, 0, 0) + 0.1,
+\begin{frame}
+  \frametitle{Shape across PC8}
+<<fig.width = 5>>=
+par(mfrow = c(1, 2), mar = c(0, 0, 1, 1) + 0.1, oma = c(5, 4, 0, 0) + 0.1,
     xaxt = 'n', yaxt = 'n')
-mapply(function(x, m, n) {
-      tpsgrid(x, m)
-      title(main = paste(n))}, x = mspi, n = levels(spi),
-      MoreArgs = list(m = mt), SIMPLIFY = FALSE)
+tpsgrid(turtle.land.adult[, , snd.min], mt, mag = 2)
+title(main = 'min observed')
+tpsgrid(turtle.land.adult[, , snd.max], mt, mag = 2)
+title(main = 'max observed')
 par(mfrow = c(1, 1))
 @
-
 \end{frame}
@@
       \begin{itemize}
-        \item Ben Frable, Dallas Krentzel, Michael Foote
+        \item Ben Frable, Dallas Krentzel, Michael Foote, David Bapst
         \item COLLECTIONS

### 2013-06-13 src/plotting_functions.r

@@ -40,10 +40,10 @@ gap.plot <- function(gap) {
 mshape.plot <- function(fits) {
   shape <- as.data.frame(fits$mshape)
   land <- land.frame(fits$rotated)
-  gg <- ggplot(land, aes(x = y, y = x)) + geom_point(alpha = 0.5,
+  gg <- ggplot(land, aes(x = y, y = -x)) + geom_point(alpha = 0.5,
                                                      size = 0.7)
   gg <- gg + geom_point(data = shape, mapping = aes(x = V2,
-                                                    y = V1,
+                                                    y = -V1,
                                                     colour = 'red'))
   gg
 }
@@ -85,7 +85,7 @@ class.map <- function(train, test, label, pred, map) {
                                       colour = factor(label)
                                       #, size = 1
                                       ),
-                        size = 1,
+                        size = 1.25,
                         alpha = 0.7)
   test <- cbind(test, label = pred)
   gg <- gg + geom_point(data = test,

### 2013-06-17 documents/turtle_presentation.Rnw
fixing up various figures to make them more readable

@@
   %
-  Crytic species are species delimitated via molecular means which were not/cannot be identified via morphology.
+  Crytic species are species delimitated via molecular means. Taxa were not/cannot be identified via morphology.
 
@@
       \begin{itemize}
-        %\item \Sexpr{nrow(turtle.adult)} adult individuals
+        \item \Sexpr{nrow(turtle.adult)} adult individuals
         \item plastral (``belly'') shape
@@
   Partitioning around mediods (PAM) compared with ``gap'' statistic.
+
+  Analogous to \textit{k}-means clustering, a divisive clustering algorithm.
 
@@
     xaxt = 'n', yaxt = 'n')
+mins <- lapply(mspi, function(x) min(x[, 2]))
+mins <- min(unlist(mins))
 mapply(function(x, m, n) {
-      tpsgrid(x, m, mag = 2)
+      #tpsgrid(x, m, mag = 2, ext = 0, cex = 2)
+      plotRefToTarget(M1 = m, M2 = x, mag = 2)
       title(main = paste(n))}, x = mspi, n = levels(spi),
@@
 tpsgrid(turtle.land.adult[, , fst.min], mt, mag = 2)
+#plotRefToTarget(M1 = mt, M2 = turtle.land.adult[, , fst.min])
 title(main = 'min observed')
 tpsgrid(turtle.land.adult[, , fst.max], mt, mag = 2)
+#plotRefToTarget(M1 = mt, M2 = turtle.land.adult[, , fst.max])
 title(main = 'max observed')
@@
 tpsgrid(turtle.land.adult[, , snd.min], mt, mag = 2)
+#plotRefToTarget(M1 = mt, M2 = turtle.land.adult[, , snd.min])
 title(main = 'min observed')
 tpsgrid(turtle.land.adult[, , snd.max], mt, mag = 2)
+#plotRefToTarget(M1 = mt, M2 = turtle.land.adult[, , snd.max])
 title(main = 'max observed')

### 2013-06-17 src/plotting_functions.r

@@ -45,6 +45,7 @@ mshape.plot <- function(fits) {
   gg <- gg + geom_point(data = shape, mapping = aes(x = V2,
                                                     y = -V1,
                                                     colour = 'red'))
+
   gg
 }
 
@@ -74,29 +75,47 @@ map.plot <- function(data, label, map, coord = 'gilbert') {
   gg
 }
 
-class.map <- function(train, test, label, pred, map) {
+class.map <- function(train, test, label, pred, map, facet = TRUE) {
   gg <- ggplot(map, aes(x = long, y = lat, group = group))
   gg <- gg + geom_polygon(fill = 'white', colour = 'black')
   gg <- gg + coord_map('gilbert')
-  train <- cbind(train, label = label)
-  gg <- gg + geom_point(data = train,
-                        mapping = aes(x = long, y = lat,
-                                      group = NULL,
-                                      colour = factor(label)
-                                      #, size = 1
-                                      ),
-                        size = 1.25,
-                        alpha = 0.7)
-  test <- cbind(test, label = pred)
-  gg <- gg + geom_point(data = test,
-                        mapping = aes(x = long, y = lat,
-                                      group = NULL,
-                                      colour = factor(label)
-                                      #, size = 1
-                                      ), 
-                        size = 1.5,
-                        shape = 15,
-                        alpha = 0.8)
+
+  if(facet) {
+    tr <- cbind(long = train$long, lat = train$lat, label = label)
+    te <- cbind(long = test$long, lat = test$lat, label = pred)
+    dat <- list(train = tr, test = te)
+    tt <- c('train', 'test')
+    dat <- Map(function(x, y) cbind(as.data.frame(x), type = rep(y, nrow(x))),
+               x = dat, y = tt)
+    dat <- Reduce(rbind, dat)
+
+    gg <- gg + geom_point(data = dat,
+                          mapping = aes(x = long, y = lat, 
+                                        group = NULL,
+                                        colour = factor(label)),
+                          size = 1.25,
+                          alpha = 0.7)
+    gg <- gg + facet_wrap(~ type)
+  } else {
+    train <- cbind(train, label = label)
+    gg <- gg + geom_point(data = train,
+                         mapping = aes(x = long, y = lat,
+                                       group = NULL,
+                                       colour = factor(label)
+                                       ),
+                         size = 1.25,
+                         alpha = 0.7)
+    test <- cbind(test, label = pred)
+    gg <- gg + geom_point(data = test,
+                          mapping = aes(x = long, y = lat,
[…]

### 2013-06-17 documents/turtle_presentation.Rnw
most plots at final stage

@@
 fits <- procGPA(turtle.land.adult)
-land <- mshape.plot(fits)
+links <- c(1:7, 13:8)
+land <- mshape.plot(fits, links = links)
 land <- land + coord_equal()
@@
 dd <- melt(oo)
-gdist <- ggplot(dd, aes(x = value, fill = X2)) + geom_density(alpha = 0.3)
-gdist <- gdist + theme(legend.title = element_blank(),
-                       legend.margin = unit(0, 'cm'),
-                       legend.text = element_text(size = 5),
+gdist <- ggplot(dd, aes(x = value, fill = X2)) + geom_density(alpha = 0.5,
+                                                              size = 0.1)
+gdist <- gdist + theme(legend.position = 'none',
+                       #legend.title = element_blank(),
+                       #legend.margin = unit(0, 'cm'),
+                       #legend.text = element_blank(),
                        axis.title = element_text(size = 8),
@@
 gg <- gg + scale_colour_manual(name = '', values = cbp)
-gg <- gg + theme(#legend.position = 'none',
-                 legend.margin = unit(0, 'cm'),
-                 legend.text = element_text(size = 5),
+gg <- gg + theme(legend.position = 'none',
+                 #legend.margin = unit(0, 'cm'),
+                 #legend.text = element_text(size = 5),
                  axis.title = element_text(size = 8),
@@
 # get mean of each class
+
+mt <- mshape(turtle.land.adult)
+
 spi <- turtle.adult$spinks
@@
 mins <- min(unlist(mins))
+lmat <- cbind(links, c(links[-1], links[1]))
 mapply(function(x, m, n) {
       #tpsgrid(x, m, mag = 2, ext = 0, cex = 2)
-      plotRefToTarget(M1 = m, M2 = x, mag = 2)
+      plotRefToTarget(M1 = m, M2 = x, mag = 2, links = lmat)
       title(main = paste(n))}, x = mspi, n = levels(spi),
@@
 par(mfrow = c(1, 1))
+
+multi.mean <- multishape.plot(mspi, links = links)
 @
@@
 <<fig.width = 5>>=
-mt <- mshape(turtle.land.adult)
 
@@
 
-par(mfrow = c(1, 2), mar = c(0, 0, 1, 1) + 0.1, oma = c(5, 4, 0, 0) + 0.1,
-    xaxt = 'n', yaxt = 'n')
-tpsgrid(turtle.land.adult[, , fst.min], mt, mag = 2)
+#par(mfrow = c(1, 2), mar = c(0, 0, 1, 1) + 0.1, oma = c(5, 4, 0, 0) + 0.1,
+#    xaxt = 'n', yaxt = 'n', bty = 'n')
+#tpsgrid(turtle.land.adult[, , fst.min], mt, mag = 2)
 #plotRefToTarget(M1 = mt, M2 = turtle.land.adult[, , fst.min])
-title(main = 'min observed')
-tpsgrid(turtle.land.adult[, , fst.max], mt, mag = 2)
+#title(main = 'min observed')
+#tpsgrid(turtle.land.adult[, , fst.max], mt, mag = 2)
 #plotRefToTarget(M1 = mt, M2 = turtle.land.adult[, , fst.max])
-title(main = 'max observed')
-par(mfrow = c(1, 1))
+#title(main = 'max observed')
+#par(mfrow = c(1, 1))
+
+ex.lab <- function(var, value) {
+  value <- as.character(value)
+  if(var == 'lab') {
+    value[value == '1'] <- 'min'
+    value[value == '2'] <- 'max'
+  }
+  return(value)
+}
+
+fst.mean <- multishape.plot(list(turtle.land.adult[, , fst.min], 
+                                 turtle.land.adult[, , fst.max]), 
+                            links = links,
+                            labeller = ex.lab)
+fst.mean <- fst.mean + theme(axis.title = element_blank(),
+                             axis.text = element_text(size = 5))
+fst.mean
 @
@@
 <<fig.width = 5>>=
-par(mfrow = c(1, 2), mar = c(0, 0, 1, 1) + 0.1, oma = c(5, 4, 0, 0) + 0.1,
-    xaxt = 'n', yaxt = 'n')
-tpsgrid(turtle.land.adult[, , snd.min], mt, mag = 2)
+#par(mfrow = c(1, 2), mar = c(0, 0, 1, 1) + 0.1, oma = c(5, 4, 0, 0) + 0.1,
+#    xaxt = 'n', yaxt = 'n', bty = 'n')
+#tpsgrid(turtle.land.adult[, , snd.min], mt, mag = 2)
 #plotRefToTarget(M1 = mt, M2 = turtle.land.adult[, , snd.min])
-title(main = 'min observed')
-tpsgrid(turtle.land.adult[, , snd.max], mt, mag = 2)
+#title(main = 'min observed')
+#tpsgrid(turtle.land.adult[, , snd.max], mt, mag = 2)
 #plotRefToTarget(M1 = mt, M2 = turtle.land.adult[, , snd.max])
-title(main = 'max observed')
-par(mfrow = c(1, 1))
+#title(main = 'max observed')
+#par(mfrow = c(1, 1))
+
+snd.mean <- multishape.plot(list(turtle.land.adult[, , snd.min], 
+                                 turtle.land.adult[, , snd.max]), 
+                            links = links,
+                            labeller = ex.lab)
+snd.mean <- snd.mean + theme(axis.title = element_blank(),
+                             axis.text = element_text(size = 5))
+snd.mean
 @
@@
         \item Ben Frable, Dallas Krentzel, Michael Foote, David Bapst
-        \item COLLECTIONS
-        \item FUNDING AGENCIES
+        \item collections at CAS, FMNH, LACM, LSUMNS, MSE, MVZ, NMNH, SBMNH, SDMNH, UCMP, UMZC
+        \item NSF DBI-0306158 to KDA
       \end{itemize}

### 2013-06-17 src/plotting_functions.r

@@ -37,14 +37,56 @@ gap.plot <- function(gap) {
   gg
 }
 
-mshape.plot <- function(fits) {
+mshape.plot <- function(fits, links) {
   shape <- as.data.frame(fits$mshape)
+  shape <- shape[links, ]
+  shape <- cbind(shape, rbind(shape[-1, ], shape[1, ]))
+  names(shape) <- c('V1', 'V2', 'V3', 'V4')
+
   land <- land.frame(fits$rotated)
   gg <- ggplot(land, aes(x = y, y = -x)) + geom_point(alpha = 0.5,
-                                                     size = 0.7)
+                                                     size = 0.5)
+  gg <- gg + geom_segment(data = shape, mapping = aes(x = V2,
+                                                      xend = V4,
+                                                      y = -V1,
+                                                      yend = -V3),
+                          alpha = 0.5)
   gg <- gg + geom_point(data = shape, mapping = aes(x = V2,
                                                     y = -V1,
-                                                    colour = 'red'))
+                                                    colour = 'red'),
+                        size = 0.9)
+
+  gg
+}
+
+shape.plot <- function(shape, links) {
+  shape <- as.data.frame(shape)
+  shape <- shape[links, ]
+  shape <- cbind(shape, rbind(shape[-1, ], shape[1, ]))
+  names(shape) <- c('V1', 'V2', 'V3', 'V4')
+  gg <- ggplot(shape, aes(x = V2, y = -V1))
+  gg <- gg + geom_point(size = 0.9)
+  gg <- gg + geom_segment(mapping = aes(x = V2, xend = V4,
+                                        y = -V1, yend = -V3))
+
+  gg
+}
+
+multishape.plot <- function(shapes, links, labeller = 'label_value') {
+  shapes <- lapply(shapes, as.data.frame)
+  shapes <- lapply(shapes, function(x) x[links, ])
+  shapes <- lapply(shapes, function(x) cbind(x, rbind(x[-1, ], x[1, ])))
+  shapes <- lapply(shapes, function(x) {
+                   names(x) <- c('V1', 'V2', 'V3', 'V4')
+                   x})
+  na <- c(rep(seq(length(shapes)), lapply(shapes, nrow)))
+  shapes <- Reduce(rbind, shapes)
+  shapes <- cbind(shapes, lab = factor(na))
+  gg <- ggplot(shapes, aes(x = V2, y = -V1))
[…]

### 2013-06-18 documents/turtle_presentation.Rnw
pretty wireframes

@@
     value[value == '1'] <- 'min'
-    value[value == '2'] <- 'max'
+    value[value == '2'] <- 'mean'
+    value[value == '3'] <- 'max'
   }
@@
 fst.mean <- multishape.plot(list(turtle.land.adult[, , fst.min], 
+                                 mt,
                                  turtle.land.adult[, , fst.max]), 
@@
 fst.mean <- fst.mean + theme(axis.title = element_blank(),
-                             axis.text = element_text(size = 5))
+                             axis.text = element_text(size = 4))
+fst.mean <- fst.mean + coord_equal()
 fst.mean
@@
 snd.mean <- multishape.plot(list(turtle.land.adult[, , snd.min], 
+                                 mt,
                                  turtle.land.adult[, , snd.max]), 
@@
 snd.mean <- snd.mean + theme(axis.title = element_blank(),
-                             axis.text = element_text(size = 5))
+                             axis.text = element_text(size = 4))
+snd.mean <- snd.mean + coord_equal()
 snd.mean

### 2013-06-19 documents/turtle_presentation.Rnw
final draft?

@@
                                                               size = 0.1)
-gdist <- gdist + theme(legend.position = 'none',
-                       #legend.title = element_blank(),
-                       #legend.margin = unit(0, 'cm'),
+gdist <- gdist + theme(#legend.position = 'none',
+                       legend.title = element_blank(),
+                       legend.margin = unit(0, 'cm'),
                        #legend.text = element_blank(),
@@
                        axis.text = element_text(size = 5))
-gdist <- gdist + scale_fill_manual(name = '', values = cbp)
+gdist <- gdist + scale_fill_manual(name = '', values = cbp,
+                                   labels = c('morph 1', 'morph 2',
+                                              'molec 1', 'molec 2'))
 
@@
                  params = c(LabelSize = 2, gridLabelSize = 2, size = 1))
+classlab <- ggally_text('molec 2', size = 3)
+
+ggimp <- putPlot(ggimp, classlab, 4, 4)
 

### 2013-06-21 documents/turtle_presentation.Rnw
messing with tiny things

@@
   %
-  Crytic species are species delimitated via molecular means. Taxa were not/cannot be identified via morphology.
+  Cryptic species are species delimited via molecular means. Taxa were not/cannot be identified via morphology.
 
@@
 
+roc.mod$.id[roc.mod$.id == 'sh1'] <- 'morph 1'
+roc.mod$.id[roc.mod$.id == 'sh2'] <- 'morph 2'
+roc.mod$.id[roc.mod$.id == 'sh3'] <- 'molec 1'
+roc.mod$.id[roc.mod$.id == 'spinks'] <- 'molec 2'
+
 ggroc <- ggplot(roc.mod, aes(x = variable, y = value, color = mod.names))
@@
                        legend.margin = unit(0, 'cm'),
-                       legend.text = element_text(size = 5),
+                       legend.text = element_text(size = 4),
                        axis.title = element_text(size = 8),
-                       axis.text = element_text(size = 5))
+                       axis.text = element_text(size = 5),
+                       strip.text = element_text(size = 5))
 ggroc <- ggroc + labs(x = '# of features', y = 'AUC ROC')
@@
                        legend.margin = unit(0, 'cm'),
-                       #legend.text = element_blank(),
+                       legend.text = element_text(size = 4),
                        axis.title = element_text(size = 8),
@@
                                               'molec 1', 'molec 2'))
+gdist <- gdist + labs(x = 'AUC ROC')
 
@@
 gg <- gg + scale_colour_manual(name = '', values = cbp)
-gg <- gg + theme(legend.position = 'none',
-                 #legend.margin = unit(0, 'cm'),
-                 #legend.text = element_text(size = 5),
+gg <- gg + theme(#legend.position = 'none',
+                 legend.margin = unit(0, 'cm'),
+                 legend.text = element_text(size = 4),
                  axis.title = element_text(size = 8),

### 2013-06-22 documents/turtle_presentation.Rnw
must keep messing with the figures

@@
   %
-  Cryptic species are species delimited via molecular means. Taxa were not/cannot be identified via morphology.
+  Cryptic species are species which were first delimited via molecular means. Were not/cannot be identified via morphology.
 
@@
                        strip.text = element_text(size = 5))
-ggroc <- ggroc + labs(x = '# of features', y = 'AUC ROC')
+ggroc <- ggroc + labs(x = '# of features (PCs)', y = 'AUC ROC')
 ggroc <- ggroc + facet_wrap(~.id)

### 2013-06-28 documents/turtle_paper.Rnw
beginning paper. blocking, bitex files, etc.

% how cryptic is cryptic diversity?
% paper describing the motivation, methods and results of the turtle
% subspecies identification project.
% journal submission order:
%   American Naturalist
%   Journal of Evolutionary Biology

\documentclass[12pt]{article}
\usepackage{amsmath, amsthm}
\usepackage{graphicx, microtype, parskip, hyperref, authblk}
\usepackage{rotating, longtable, caption, subcaption, multirow}
\usepackage[sort&compress]{natbib}

% for american naturalist
\usepackage{lineno}

\frenchspacing

% bring in various code necessities
% all the figures are made externally, so this would just be for numerics
<<message = FALSE, echo = FALSE, warning = FALSE>>=
opts_chunk$set(warning = FALSE, echo = FALSE, message = FALSE,
               highlight = TRUE, background = 'white')

load('../data/turtle_analysis.RData')
load('../data/turtle_gen.RData')
@

\title{How cryptic is cryptic diversity? Machine learning approaches to plastral variation in \textit{Emys marmorata}.}
\author[1]{Peter D Smits \thanks{psmits@uchicago.edu}}
\author[2]{Kenneth D Angielczyk \thanks{kangielczyk@fieldmuseum.org}}
\author[3]{James F Parham \thanks{jparham@fullerton.edu}}
\affil[1]{Committee on Evolution Biology, University of Chicago}
\affil[2]{Department of Geology, Field Museum of Natural History}
\affil[3]{Department of Geological Sciences, California State University -- Fullerton}

\begin{document}

\maketitle

\linenumbers
\modulolinenumbers[2]

\begin{abstract}
  % 200 words

\end{abstract}

\section{Introduction}

% cryptic diversity
%   most species are still deliminated solely based on morphology
%   paleo problem
Cryptic diversity is when taxa were only first deliminated via molecular means and were not or cannot deliminated via morphological identification CITATION. The discovery of this previously unknown diversity has

% e. marmorata
%   natural history
%   morphological hypothesis of subspecies
%   molecular hypothesis

% central hypothesis/question of study
Here, we address the question of how much of cryptic diversity may be a product of sample size as well as methodology used for classifying taxa based solely on morphology. Specifically, we ask if fine scale variation in morphology can provide corroboration for subspecific assignment, and if 
% statement of goals and approach
To analyze this question, we apply multiple machine learning approaches to estimate the best classification scheme of \textit{E. marmorata} subspecies based on morphological variation in plastral shape.

\section{Materials and Methods}
\subsection{Specimens}
We collected morphometric data from \Sexpr{nrow(turtle.adult)} specimens. Geographic information was recorded from museum collection information. When precise latitude and longitude information was not known for a specimen, it was inferred from whatever locality information was presented.

Specimens were given a class assignment was based on geographic information. Because the exact geographic barriers between different class is unknown and fuzzy, two assignments for both morphological and molecular hypotheses of class were used.

\subsection{Geometric morphometrics}
% landmarks
Following \citet{Angielczyk2011}, 19 landmarks were digitized using TpsDig 2.04 \citep{Rohlf2005}. 17 of these landmarks are at the endpoints or intersection of the keratinous plastral scutes that cover the platron. These landmarks were chosen to maximize the description of plastral variation. 12 of these landmarks are symmetrical across the axis of symmetry and in order to prevent degrees of freedom and other concerns \citep{Klingenberg2007}, these landmarks were reflected across the axis of symmetry and the average position of each symmetrical pair was used. Analysis was then conducted on the resulting ``half'' plastra.

% GP superimposition
``Half'' plastra landmark configurations were superimposed using generalized Procrustes analysis \citep{Dryden1998a} after which, the principal components of shape were calculated. This was done using the \texttt{shapes} package in R \citep{2013, Dryden2013}.
% PCA

\subsection{Machine learning analyses}
\subsubsection{Unsupervised learning}

\subsubsection{Supervised learning}

\section{Results}
\subsection{Geometric morphometrics}

\subsection{Machine learning analyses}
\subsubsection{Unsupervised learning}

\subsubsection{Supervised learning}

\section{Discussion}

\section*{Acknowledgements}
PDS would like to thank David Bapst, Benjmin Frable, Michael Foote, and Dallas Krentzel for useful discussion which enhanced the quality of this study.

\bibliographystyle{amnatnat}
\bibliography{turtle,packages}

\end{document}

### 2013-06-28 documents/turtle_paper.tex
beginning paper. blocking, bitex files, etc.

--- 2013-06-28 documents/turtle_paper.Rnw
@@
 
-\documentclass[12pt]{article}
+\documentclass[12pt]{article}\usepackage{graphicx, color}
+%% maxwidth is the original width if it is less than linewidth
+%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
+\makeatletter
+\def\maxwidth{ %
+  \ifdim\Gin@nat@width>\linewidth
+    \linewidth
+  \else
+    \Gin@nat@width
+  \fi
+}
+\makeatother
+
+\definecolor{fgcolor}{rgb}{0.2, 0.2, 0.2}
+\newcommand{\hlnumber}[1]{\textcolor[rgb]{0,0,0}{#1}}%
+\newcommand{\hlfunctioncall}[1]{\textcolor[rgb]{0.501960784313725,0,0.329411764705882}{\textbf{#1}}}%
+\newcommand{\hlstring}[1]{\textcolor[rgb]{0.6,0.6,1}{#1}}%
+\newcommand{\hlkeyword}[1]{\textcolor[rgb]{0,0,0}{\textbf{#1}}}%
+\newcommand{\hlargument}[1]{\textcolor[rgb]{0.690196078431373,0.250980392156863,0.0196078431372549}{#1}}%
+\newcommand{\hlcomment}[1]{\textcolor[rgb]{0.180392156862745,0.6,0.341176470588235}{#1}}%
+\newcommand{\hlroxygencomment}[1]{\textcolor[rgb]{0.43921568627451,0.47843137254902,0.701960784313725}{#1}}%
+\newcommand{\hlformalargs}[1]{\textcolor[rgb]{0.690196078431373,0.250980392156863,0.0196078431372549}{#1}}%
+\newcommand{\hleqformalargs}[1]{\textcolor[rgb]{0.690196078431373,0.250980392156863,0.0196078431372549}{#1}}%
+\newcommand{\hlassignement}[1]{\textcolor[rgb]{0,0,0}{\textbf{#1}}}%
+\newcommand{\hlpackage}[1]{\textcolor[rgb]{0.588235294117647,0.709803921568627,0.145098039215686}{#1}}%
+\newcommand{\hlslot}[1]{\textit{#1}}%
+\newcommand{\hlsymbol}[1]{\textcolor[rgb]{0,0,0}{#1}}%
+\newcommand{\hlprompt}[1]{\textcolor[rgb]{0.2,0.2,0.2}{#1}}%
+
+\usepackage{framed}
+\makeatletter
+\newenvironment{kframe}{%
+ \def\at@end@of@kframe{}%
+ \ifinner\ifhmode%
+  \def\at@end@of@kframe{\end{minipage}}%
+  \begin{minipage}{\columnwidth}%
+ \fi\fi%
+ \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
+ \colorbox{shadecolor}{##1}\hskip-\fboxsep
+     % There is no \\@totalrightmargin, so:
+     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
+ \MakeFramed {\advance\hsize-\width
+   \@totalleftmargin\z@ \linewidth\hsize
+   \@setminipage}}%
+ {\par\unskip\endMakeFramed%
+ \at@end@of@kframe}
+\makeatother
+
+\definecolor{shadecolor}{rgb}{.97, .97, .97}
+\definecolor{messagecolor}{rgb}{0, 0, 0}
+\definecolor{warningcolor}{rgb}{1, 0, 1}
+\definecolor{errorcolor}{rgb}{1, 0, 0}
+\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
+
+\usepackage{alltt}
 \usepackage{amsmath, amsthm}
@@
 % all the figures are made externally, so this would just be for numerics
-<<message = FALSE, echo = FALSE, warning = FALSE>>=
-opts_chunk$set(warning = FALSE, echo = FALSE, message = FALSE,
-               highlight = TRUE, background = 'white')
-
-load('../data/turtle_analysis.RData')
-load('../data/turtle_gen.RData')
-@
 
@@
 \affil[3]{Department of Geological Sciences, California State University -- Fullerton}
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
 
@@
 \subsection{Specimens}
-We collected morphometric data from \Sexpr{nrow(turtle.adult)} specimens.
+We collected morphometric data from 524 specimens.
 Geographic information was recorded from museum collection information.

### 2013-06-28 documents/turtle_paper.Rnw
working on m+m of paper

@@
 % journal submission order:
-%   American Naturalist
+%   American Naturalist (21 pages)
 %   Journal of Evolutionary Biology
@@
 \usepackage{amsmath, amsthm}
-\usepackage{graphicx, microtype, parskip, hyperref, authblk}
+\usepackage{graphicx, microtype, hyperref, authblk}
 \usepackage{rotating, longtable, caption, subcaption, multirow}
 \usepackage[sort&compress]{natbib}
+%\usepackage{fullpage}
 
@@
 
+Here, we address the question of how much of cryptic diversity may be a product of sample size as well as methodology used for classifying taxa based solely on morphology.
+Specifically, we ask if fine scale variation in morphology can provide corroboration for subspecific assignment, and if it is possible to determine the best classification hypothesis amoungst a few.
+
 % e. marmorata
 %   natural history
+%     conservation importance (turtles in general, e. marmorata in specific)
+In this study, we address the subspecific classification scheme of \textit{Emys marmorata}, or western pond turtle.
+\textit{E.
+marmorata} has a distribution from northern Washington State, USA to Baja California, Mexico.
 %   morphological hypothesis of subspecies
 %   molecular hypothesis
+Traditionally, \textit{E.
+marmorata} was classified into three subgroups: the northern \textit{E.
+marmorata marmorata}, the southern \textit{E.
+marmorata palida}, and a central Californian intergrade zone \citep{Seeliger1945}.
+More recently, \textit{E.
+marmorata} was divided into four subgroups based on mitochrondial DNA: a northern clade, a southern clade, and two central Californian clades \citep{Spinks2005,Spinks2009}.
 
 % central hypothesis/question of study
-Here, we address the question of how much of cryptic diversity may be a product of sample size as well as methodology used for classifying taxa based solely on morphology.
-Specifically, we ask if fine scale variation in morphology can provide corroboration for subspecific assignment, and if
 % statement of goals and approach
-To analyze this question, we apply multiple machine learning approaches to estimate the best classification scheme of \textit{E.
-marmorata} subspecies based on morphological variation in plastral shape.
+In this study, we apply multiple machine learning approaches to estimate the best classification scheme of \textit{E. marmorata} subspecies based on morphological variation in plastral shape.
 
@@
 When precise latitude and longitude information was not known for a specimen, it was inferred from whatever locality information was presented.
+% museum information?
 
@@
 12 of these landmarks are symmetrical across the axis of symmetry and in order to prevent degrees of freedom and other concerns \citep{Klingenberg2007}, these landmarks were reflected across the axis of symmetry and the average position of each symmetrical pair was used.
+In cases where damage or incompleteness prevented symetric landmarks from being determined, only the single member of the pair was used.
 Analysis was then conducted on the resulting ``half'' plastra.
@@
 ``Half'' plastra landmark configurations were superimposed using generalized Procrustes analysis \citep{Dryden1998a} after which, the principal components of shape were calculated.
-This was done using the \texttt{shapes} package in R \citep{2013, Dryden2013}.
+This was done using the \texttt{shapes} package for R \citep{2013, Dryden2013}.
+
 % PCA
@@
 \subsubsection{Unsupervised learning}
+% PAM
+%   Riemannian shape distance
+%   gap statistic
+PAM was conducted using the \texttt{cluster} package for R \citep{Maechler2013}.
 
 \subsubsection{Supervised learning}
+% training testing split
+The dataset of \Sexpr{nrow(turtle.adult)} plastron landmarks was split into training and testing datasets.
+The former was used for model fitting (training) and was 75\% of the total dataset, split proportionally per class, while the testing dataset was used to estimate the effectiveness of each classification scheme (i.e.
+performance in the wild).
+
+% model types
+%   multinomial logistic regression
+%   random forest
+Multinomial logistic regression models were fit using the \texttt{nnet} package for R \citep{Venables2002} while random forest models were fit using the \texttt{randomForest} package for R \citep{Liaw2002}.
+
+% model training
+%   tuning parameters
+%     grid search
+%       what are the tuning parameters for each model?
+%     best AUC ROC
+In order to prevent over fitting each machine learning model, tuning parameters were estimated using 10-fold cross-validation (CV) across a grid search of all tuning parameter combinations.
+Optimal tuning parameter values were selected based on area under the receiver operating characteristic curve (AUC ROC).
+Multiclass AUC ROC was estimated using the all-against one strategy described in \citet{Hand2001}.
+
+%   model selection
+%     AICc
+For the multinomial logistic regression models, PCs were added sequentially in order to increase the overall amount of variation in shape included in each model and the final model was that with the lowest AICc \cite{Burnham2002a}.
+This procedure was used because the optimal number of PCs to include is unknown, and while including all of the PCs of shape would mean that all of the variability in plastron shape would be used to estimate class, this may cause the model to be over fit and not provide an accurate estimate of unsampled plastral variation.
+The maximum number of PCs allowed to be used as predictors was \Sexpr{floor(max.ad)} because of both the number of parameters estimated per model and the necessary sample size needed to estimate that many parameters accurately.
+
+%     recursive feature selection
+Because random forest models are not fit using maximum likelihood, a recursive feature selection algorithm was used to choose the optimal number of PCs to include based on the AUC ROC of the model.
+PCs were sequentially added as features until the AUC ROC of the model did not increase.
+After each PC was added, 10-fold CV was used to estimate the optimal values of the tuning parameters as well as quantify the uncertainty of each model.
+Like the multinomial logistic regression models, \Sexpr{floor(max.ad)} was the maximum number of PCs that could have been included in the model.
+The recursive feature selection algorithm used here is that implemented in the \texttt{caret} package for R \citep{Kuhn2013}.
+
+% generalization
+%   bootstrap resample of AUC ROC
+The final selected models were then used to estimate the class assignments of the training dataset.
+Model performance was measured using AUC ROC.
+A distribution of AUC ROC values were estimated for each classification scheme using 1000 nonparametric bootstrap resamples of the training dataset.
 
@@
 \section*{Acknowledgements}
-PDS would like to thank David Bapst, Benjmin Frable, Michael Foote, and Dallas Krentzel for useful discussion which enhanced the quality of this study.
+PDS would like to thank David Bapst, Michael Foote, Benjamin Frable, and Dallas Krentzel for useful discussion which enhanced the quality of this study.
 

### 2013-06-28 documents/turtle_paper.tex
working on m+m of paper

@@
 \subsubsection{Unsupervised learning}
+% PAM
+%   Riemannian shape distance
+%   gap statistic
+PAM was conducted using the \texttt{cluster} package for R \citep{Maechler2013}.
 
 \subsubsection{Supervised learning}
+% training testing split
+The dataset of 524 plastron landmarks was split into training and testing datasets.
+The former was used for model fitting (training) and was 75\% of the total dataset, split proportionally per class, while the testing dataset was used to estimate the effectiveness of each classification scheme (i.e.
+performance in the wild).
+
+% model types
+%   multinomial logistic regression
+%   random forest
+Multinomial logistic regression models were fit using the \texttt{nnet} package for R \citep{Venables2002} while random forest models were fit using the \texttt{randomForest} package for R \citep{Liaw2002}.
+
+% model training
+%   tuning parameters
+%     grid search
+%       what are the tuning parameters for each model?
+%     best AUC ROC
+In order to prevent over fitting each machine learning model, tuning parameters were estimated using 10-fold cross-validation (CV) across a grid search of all tuning parameter combinations.
+Optimal tuning parameter values were selected based on area under the receiver operating characteristic curve (AUC ROC).
+Multiclass AUC ROC was estimated using the all-against one strategy described in \citet{Hand2001}.
+
+%   model selection
+%     AICc
+For the multinomial logistic regression models, PCs were added sequentially in order to increase the overall amount of variation in shape included in each model and the final model was that with the lowest AICc \cite{Burnham2002a}.
+This procedure was used because the optimal number of PCs to include is unknown, and while including all of the PCs of shape would mean that all of the variability in plastron shape would be used to estimate class, this may cause the model to be over fit and not provide an accurate estimate of unsampled plastral variation.
+The maximum number of PCs allowed to be used as predictors was 10 because of both the number of parameters estimated per model and the necessary sample size needed to estimate that many parameters accurately.
+
+%     recursive feature selection
+Because random forest models are not fit using maximum likelihood, a recursive feature selection algorithm was used to choose the optimal number of PCs to include based on the AUC ROC of the model.
+PCs were sequentially added as features until the AUC ROC of the model did not increase.
+After each PC was added, 10-fold CV was used to estimate the optimal values of the tuning parameters as well as quantify the uncertainty of each model.
+Like the multinomial logistic regression models, 10 was the maximum number of PCs that could have been included in the model.
+The recursive feature selection algorithm used here is that implemented in the \texttt{caret} package for R \citep{Kuhn2013}.
+
+% generalization
+%   bootstrap resample of AUC ROC
+The final selected models were then used to estimate the class assignments of the training dataset.
+Model performance was measured using AUC ROC.
+A distribution of AUC ROC values were estimated for each classification scheme using 1000 nonparametric bootstrap resamples of the training dataset.
 

### 2013-06-30 documents/turtle_paper.Rnw
methods work

@@
 % model types
+Two types of supervised learning, or classification, models were fit to the PCs of plastral shape: multinomial logistic regression and random forest.
+These model types were chosen because of various properties of these models which allow for useful interpretations about the strength and structure of the classification.
+Multinomial logistic regression models were fit using the \texttt{nnet} package for R \citep{Venables2002} while random forest models were fit using the \texttt{randomForest} package for R \citep{Liaw2002}.
+
 %   multinomial logistic regression
+Multinomial logistic regression is an extension of logistic regression, where instead of a binary response it is possible to have three or more response classes CITATION.
+Effectively, this type of model can be viewed as multiple, simultaneous logistic regression models for each class and the final classification of the observation being the most probable of all the sub-model classifications.
+From the final model the relative risk of a given classification, with reference to a given class, can be calculated from the coefficients of the features, or predictors.
+This is similar to the log-odds calculated from the coefficients of a logistic regression.
+
 %   random forest
-Multinomial logistic regression models were fit using the \texttt{nnet} package for R \citep{Venables2002} while random forest models were fit using the \texttt{randomForest} package for R \citep{Liaw2002}.
+Random forest models are an extension of classification and regression trees (CART) CITATION.
+Basically, CARTs are built for random subsamples of both the features of the proposed model and observations.
+This process is repeated many times, 1000 times here, and the final model is chosen as the mode of the parameter estimates from the distribution of CARTs CITATION.
+In addition to fitting a classification model, this procedure allows for the features to be ranked in order of importance, means that the variables most important for determining a given classification scheme can be estimated.
+In the context of predicting class from geometric morphometric data, this identifies the PCs that describe the variation that best distinguishes the different classes.
+%   R statement
 

### 2013-07-01 documents/turtle_paper.Rnw
methods. pam.

@@
 \subsubsection{Unsupervised learning}
+% clustering
+%   Riemannian shape distance
+Because shape space, or configurations after Procrustes superimposition, is a Riemannian manifold \citep{Dryden1998a} the dissimilarity between each landmark configuration was measured as the Riemmanian shape distance \citep{Dryden1998a} AND KENDALL CITATION.
+
 %   PAM
-%   Riemannian shape distance
+The dissimilarity matrix of shape was divisivly clustering using partioning around mediods (PAM) which is analogous to \textit{k}-means clustering except that instead of minimizing the sum of squared Euclidean distances between observations and centroids, the sum of squared dissimilarities between observations and mediods is minimized CITATION.
+
 %   gap statistic
-PAM was conducted using the \texttt{cluster} package for R \citep{Maechler2013}.
+The optimal number of clusters of shape configurations is unknown being possibly three, four, or some other unknown.
+Clustering solutions were estimated for between 1 and \Sexpr{nrow(tadult.gap$Tab)} clusters.
+Clustering solutions were compared using the gap statistic, which is a measure of goodness of clustering CITATION.
+
+%  R implementation
+PAM clustering and gap statistic calculation was conducted using the \texttt{cluster} package for R \citep{Maechler2013}.
 
@@
 Optimal tuning parameter values were selected based on area under the receiver operating characteristic curve (AUC ROC).
-Multiclass AUC ROC was estimated using the all-against one strategy described in \citet{Hand2001}.
+Multiclass AUC ROC was estimated using the all-against one strategy derived by \citet{Hand2001} in implemented in PROC PACKGE.
 
@@
 %     AICc
-For the multinomial logistic regression models, PCs were added sequentially in order to increase the overall amount of variation in shape included in each model and the final model was that with the lowest AICc \cite{Burnham2002a}.
+For the multinomial logistic regression models, PCs were added sequentially in order to increase the overall amount of variation in shape included in each model and the final model was that with the lowest AICc \citep{Burnham2002a} AKAIKE AND OTHER CITATION.
 This procedure was used because the optimal number of PCs to include is unknown, and while including all of the PCs of shape would mean that all of the variability in plastron shape would be used to estimate class, this may cause the model to be over fit and not provide an accurate estimate of unsampled plastral variation.
@@
 \subsubsection{Unsupervised learning}
+% figure
+%   gap statistic results
 
 \subsubsection{Supervised learning}
+% figure
+%   ROC model selection (see talk)
+%   generalize densities
+%     facet: multinomial, rf (see talk)
+%   maps
+%     facet: training, multinomial, rf
+%
+% table
+%   AICc model selection
 

### 2013-07-01 documents/turtle_paper.tex
methods. pam.

@@
 \subsubsection{Unsupervised learning}
+% clustering
+%   Riemannian shape distance
+Because shape space, or configurations after Procrustes superimposition, is a Riemannian manifold \citep{Dryden1998a} the dissimilarity between each landmark configuration was measured as the Riemmanian shape distance \citep{Dryden1998a} AND KENDALL CITATION.
+
 %   PAM
-%   Riemannian shape distance
+The dissimilarity matrix of shape was divisivly clustering using partioning around mediods (PAM) which is analogous to \textit{k}-means clustering except that instead of minimizing the sum of squared Euclidean distances between observations and centroids, the sum of squared dissimilarities between observations and mediods is minimized CITATION.
+
 %   gap statistic
-PAM was conducted using the \texttt{cluster} package for R \citep{Maechler2013}.
+The optimal number of clusters of shape configurations is unknown being possibly three, four, or some other unknown.
+Clustering solutions were estimated for between 1 and 40 clusters.
+Clustering solutions were compared using the gap statistic, which is a measure of goodness of clustering CITATION.
+
+%  R implementation
+PAM clustering and gap statistic calculation was conducted using the \texttt{cluster} package for R \citep{Maechler2013}.
 

### 2013-07-03 documents/turtle_paper.Rnw
figures. tables. writing.

@@
 load('../data/turtle_gen.RData')
+
+#source('../src/turtle_plots.r')
 @
@@
 The discovery of this previously unknown diversity has
+% all the work done on species delimitation strictly from molecular data
+%
+% past work on automatic taxon identification and older approaches to classifying taxa
+% why use machine learning methods
 
 Here, we address the question of how much of cryptic diversity may be a product of sample size as well as methodology used for classifying taxa based solely on morphology.
-Specifically, we ask if fine scale variation in morphology can provide corroboration for subspecific assignment, and if it is possible to determine the best classification hypothesis amoungst a few.
+Specifically, we ask if fine scale variation in morphology can provide corroboration for subspecific assignment, and if it is possible to determine the best classification hypothesis amongst a few.
 
@@
 More recently, \textit{E.
-marmorata} was divided into four subgroups based on mitochrondial DNA: a northern clade, a southern clade, and two central Californian clades \citep{Spinks2005,Spinks2009}.
+marmorata} was divided into four subgroups based on mitochrondial DNA: a northern clade, a southern clade, and two central Californian clades \citep{Spinks2005,Spinks2010}.
 
@@
 
-% GP superimposition
+% GP superimposition and PCA
 ``Half'' plastra landmark configurations were superimposed using generalized Procrustes analysis \citep{Dryden1998a} after which, the principal components of shape were calculated.
 This was done using the \texttt{shapes} package for R \citep{2013, Dryden2013}.
-
-% PCA
 
@@
 %   Riemannian shape distance
-Because shape space, or configurations after Procrustes superimposition, is a Riemannian manifold \citep{Dryden1998a} the dissimilarity between each landmark configuration was measured as the Riemmanian shape distance \citep{Dryden1998a} AND KENDALL CITATION.
+Because shape space, or configurations after Procrustes superimposition, is a Riemannian manifold \citep{Dryden1998a} the dissimilarity between each landmark configuration was measured as the Riemmanian shape distance or \(\rho\) \citep{Kendall1984a,Dryden1998a} which should vary between 0 and \(\pi / 2\) assuming no reflection invariance.
 
 %   PAM
-The dissimilarity matrix of shape was divisivly clustering using partioning around mediods (PAM) which is analogous to \textit{k}-means clustering except that instead of minimizing the sum of squared Euclidean distances between observations and centroids, the sum of squared dissimilarities between observations and mediods is minimized CITATION.
-
+The dissimilarity matrix of shape was divisivly clustering using partioning around mediods (PAM) which is analogous to \textit{k}-means clustering except that instead of minimizing the sum of squared Euclidean distances between observations and centroids, the sum of squared dissimilarities between observations and mediods is minimized \citep{Kaufman1990}.
 %   gap statistic
-The optimal number of clusters of shape configurations is unknown being possibly three, four, or some other unknown.
+The optimal number of clusters of shape configurations is unknown being possibly three, four, or some other value.
 Clustering solutions were estimated for between 1 and \Sexpr{nrow(tadult.gap$Tab)} clusters.
-Clustering solutions were compared using the gap statistic, which is a measure of goodness of clustering CITATION.
-
+Clustering solutions were compared using the gap statistic, which is a measure of goodness of clustering \citep{Tibshirani2001a}.
+Standard errors of the gap statistic for each clustering solution were estimated from \Sexpr{tadult.gap$B} bootstrap samples.
 %  R implementation
@@
 Model performance was measured using AUC ROC.
-A distribution of AUC ROC values were estimated for each classification scheme using 1000 nonparametric bootstrap resamples of the training dataset.
+A distribution of AUC ROC values were estimated for each classification scheme using \Sexpr{rr$sh1$R} nonparametric bootstrap resamples of the training dataset.
 
@@
 \subsubsection{Unsupervised learning}
+
+% PAM results
+Comparison of the gap statistic values for the different PAM solutions indicates that the optimal number of clusters is \Sexpr{which.max(tadult.gap$Tab[, 'gap'])} (Fig.
+\ref{fig:gap}).
+The second best clustering solution is two clusters, however there is no geographic structure to this classification scheme SUPPLEMENT?.
+Increasing the number of clusters does appear to improve the gap statistic enough to merit comparison.
+
 % figure
 %   gap statistic results
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width = \textwidth]{figure/gap_res}
+  \caption{Gap statistic values for PAM clustering results for the \(\rho\) dissimliarity matrix of plastron shape.
+Error bars are standard errors estimated via \Sexpr{tadult.gap$B} bootstrap samples.}
+  \label{fig:gap}
+\end{figure}
 
 \subsubsection{Supervised learning}
+For all classification schemes, the optimal random forest model based on recursive feature selection by maximizing AUC ROC of the model was one with many features (Fig. \ref{fig:roc})
 % figure
@@
 %     facet: multinomial, rf (see talk)
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width = \textwidth]{figure/roc_sel}
+  \caption{Effect of increasing the number of PCs as features, or predictors, of classification of plastra for all four classification schemes.
+As the number of PCs increase, AUC ROC increases until eventually leveling off.
+Both multinomial logisitic regression and random forest models are illustrated here, though AUC ROC based model selection was only performed for random forest models.}
+  \label{fig:roc}
+\end{figure}
+
+% table
+%   AICc model selection
+%   SUPPLEMENT?
+
 %   maps
 %     facet: training, multinomial, rf
-%
-% table
-%   AICc model selection
+Results of the bootstrap resamples of the AUC ROC of the generalization of the selected multinomial logistic regression and random forest models demonstrates that one of the molecular classification hypotheses based on \citet{Spinks2005} and \citet{Spinks2010} appears to be the best classification scheme (Fig.
+\ref{fig:gen_res}).
+The distribution of bootstrapped AUC ROC for the molecular hypothesis is significantly different MANN-WHITNEY U TEST and greater than all of the other classification scheme.
+What is remarkable is that the best classification hypothesis is identical based on both the multinomial logistic regression and random forest models.
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width = \textwidth]{figure/gen_res}
+  \caption{Density estimates of AUC ROC values of predictions of the testing dataset of plastra from \Sexpr{rr$sh1$R} bootstrap resamples.
+The top facet corresponds to values using the optimal multinomial logistic regression model, as chosen by minimum AICc value.
+The bottom facet corresponds to the values using the optimal random forest model, as chosen by maximum AUC ROC value.}
+  \label{fig:gen_res}
+\end{figure}
+
+When the classification results of the training set for the optimal classification scheme are compared with the references classes, (Fig. \ref{fig:gen_map})
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width = \textwidth]{figure/gen_map}
+  \caption{Comparison between reference classification of testing data set and the estimated classifications based on the selected multinomial logistic regression and random forest models, from left to right respectively.
+Classification corresponds to the four classes as suggested by the hypothesis of \citet{Spinks2005} and \citet{Spinks2010}.}
+  \label{fig:gen_map}
+\end{figure}
 

### 2013-07-03 documents/turtle_paper.tex
figures. tables. writing.

@@
 %   Riemannian shape distance
-Because shape space, or configurations after Procrustes superimposition, is a Riemannian manifold \citep{Dryden1998a} the dissimilarity between each landmark configuration was measured as the Riemmanian shape distance \citep{Dryden1998a} AND KENDALL CITATION.
+Because shape space, or configurations after Procrustes superimposition, is a Riemannian manifold \citep{Dryden1998a} the dissimilarity between each landmark configuration was measured as the Riemmanian shape distance or \(\rho\) \citep{Kendall1984a,Dryden1998a} which should vary between 0 and \(\pi / 2\) assuming no reflection invariance.
 
 %   PAM
-The dissimilarity matrix of shape was divisivly clustering using partioning around mediods (PAM) which is analogous to \textit{k}-means clustering except that instead of minimizing the sum of squared Euclidean distances between observations and centroids, the sum of squared dissimilarities between observations and mediods is minimized CITATION.
-
+The dissimilarity matrix of shape was divisivly clustering using partioning around mediods (PAM) which is analogous to \textit{k}-means clustering except that instead of minimizing the sum of squared Euclidean distances between observations and centroids, the sum of squared dissimilarities between observations and mediods is minimized \citep{Kaufman1990}.
 %   gap statistic
-The optimal number of clusters of shape configurations is unknown being possibly three, four, or some other unknown.
+The optimal number of clusters of shape configurations is unknown being possibly three, four, or some other value.
 Clustering solutions were estimated for between 1 and 40 clusters.
-Clustering solutions were compared using the gap statistic, which is a measure of goodness of clustering CITATION.
-
+Clustering solutions were compared using the gap statistic, which is a measure of goodness of clustering \citep{Tibshirani2001a}.
+Standard errors of the gap statistic for each clustering solution were estimated from 500 bootstrap samples.
 %  R implementation
@@
 \subsubsection{Unsupervised learning}
+
+% PAM results
+Comparison of the gap statistic values for the different PAM solutions indicates that the optimal number of clusters is 1 (Fig.
+\ref{fig:gap}).
+The second best clustering solution is two clusters, however there is no geographic structure to this classification scheme SUPPLEMENT?.
+Increasing the number of clusters does appear to improve the gap statistic enough to merit comparison.
+
 % figure
 %   gap statistic results
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width = \textwidth]{figure/gap_res}
+  \caption{Gap statistic values for PAM clustering results for the \(\rho\) dissimliarity matrix of plastron shape. Error bars are standard errors estimated via 500 bootstrap samples.}
+  \label{fig:gap}
+\end{figure}
 
 \subsubsection{Supervised learning}
+For all classification schemes, the optimal random forest model based on recursive feature selection by maximizing AUC ROC of the model was one with many features (Fig. \ref{fig:roc})
 % figure
@@
 %     facet: multinomial, rf (see talk)
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width = \textwidth]{figure/roc_sel}
+  \caption{Effect of increasing the number of PCs as features, or predictors, of classification of plastra for all four classification schemes.
+As the number of PCs increase, AUC ROC increases until eventually leveling off.
+Both multinomial logisitic regression and random forest models are illustrated here, though AUC ROC based model selection was only performed for random forest models.}
+  \label{fig:roc}
+\end{figure}
+
+% table
+%   AICc model selection
+%   SUPPLEMENT?
+
 %   maps
 %     facet: training, multinomial, rf
-%
-% table
-%   AICc model selection
+Results of the bootstrap resamples of the AUC ROC of the generalization of the selected multinomial logistic regression and random forest models demonstrates that one of the molecular classification hypotheses based on \citet{Spinks2005} and \citet{Spinks2010} appears to be the best classification scheme (Fig.
+\ref{fig:gen_res}).
+The distribution of bootstrapped AUC ROC for the molecular hypothesis is significantly different MANN-WHITNEY U TEST and greater than all of the other classification scheme.
+What is remarkable is that the best classification hypothesis is identical based on both the multinomial logistic regression and random forest models.
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width = \textwidth]{figure/gen_res}
+  \caption{Density estimates of AUC ROC values of predictions of the testing dataset of plastra from 1000 bootstrap resamples.
+The top facet corresponds to values using the optimal multinomial logistic regression model, as chosen by minimum AICc value.
+The bottom facet corresponds to the values using the optimal random forest model, as chosen by maximum AUC ROC value.}
+  \label{fig:gen_res}
+\end{figure}
+
+When the classification results of the training set for the optimal classification scheme are compared with the references classes, (Fig. \ref{fig:gen_map})
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width = \textwidth]{figure/gen_map}
+  \caption{Comparison between reference classification of testing data set and the estimated classifications based on the selected multinomial logistic regression and random forest models, from left to right respectively.
+Classification corresponds to the four classes as suggested by the hypothesis of \citet{Spinks2005} and \citet{Spinks2010}.}
+  \label{fig:gen_map}
+\end{figure}
 

### 2013-07-08 documents/turtle_paper.Rnw
lots of paper work. new figure. changed target to syst biol

@@
 % journal submission order:
+%   Systematic Biology (no limit)
 %   American Naturalist (21 pages)
@@
 % all the figures are made externally, so this would just be for numerics
-<<message = FALSE, echo = FALSE, warning = FALSE>>=
+<<message = FALSE, echo = FALSE, warning = FALSE, results = 'hide'>>=
 opts_chunk$set(warning = FALSE, echo = FALSE, message = FALSE,
@@
 
-#source('../src/turtle_plots.r')
+source('../src/turtle_plots.r')
 @
@@
 %
-% past work on automatic taxon identification and older approaches to classifying taxa
-% why use machine learning methods
 
@@
 Specifically, we ask if fine scale variation in morphology can provide corroboration for subspecific assignment, and if it is possible to determine the best classification hypothesis amongst a few.
+
+% past work on automatic taxon identification and older approaches to classifying taxa
+% why use machine learning methods
+In the analysis of class differences and identification in morphometric analysis, classification methods such as discriminant analysis and canonical variates analysis CITATIONS have been used frequently for understanding how variation in specific morphologies best differentiate classes of taxa.
+Additionally, there has been work on the application of neural network models in classification schemes of observations especially in the context of automated taxon identification CITATIONS MACLEOD'S BOOK.
+Here, we used multiple types of machine learning methods, both unsupervised and supervised, in order to understand the best classification scheme of the taxa.
+Each method provides a series of unique advantages for understanding how to classify taxa, which are discussed below.
 
@@
 marmorata palida}, and a central Californian intergrade zone \citep{Seeliger1945}.
+\textit{E.
+marmorata marmorata} is differentiated from \textit{E.
+marmorata palida} by the presence of a pair of triangular inguinal plates and darker neck markings.
+It should be noted that the triangular inguinal plates can sometimes be present in \textit{E.
+marmorata palida} though they are considerably smaller.
 More recently, \textit{E.
-marmorata} was divided into four subgroups based on mitochrondial DNA: a northern clade, a southern clade, and two central Californian clades \citep{Spinks2005,Spinks2010}.
+marmorata} was divided into four clades based on mitochrondial DNA: a northern clade, a southern clade, and two central Californian clades \citep{Spinks2005,Spinks2010}.
+Nuclear DNA supports two major clades, one northern and one southern, however \citet{Spinks2010} argue that the four clade classification is of greater conservation utility to use the mitochondrial classification scheme.
+There is now known morphological differentiation between these clades.
 
@@
 % statement of goals and approach
-In this study, we apply multiple machine learning approaches to estimate the best classification scheme of \textit{E. marmorata} subspecies based on morphological variation in plastral shape.
+In this study, we apply multiple machine learning approaches to estimate the best classification scheme of \textit{E.
+marmorata} subspecies based on morphological variation in plastral shape.
+Because of unclear geographic boundaries between subgroups of \textit{E.
+marmorata}, we compare two hypotheses of morphology-based classification and two hypotheses of molecular-based classification.
 
@@
 %     best AUC ROC
-In order to prevent over fitting each machine learning model, tuning parameters were estimated using 10-fold cross-validation (CV) across a grid search of all tuning parameter combinations.
+For both types of supervised learning models, tuning parameters were estimated using 10 rounds of 10-fold cross-validation (CV) across a grid search of all tuning parameter combinations.
 Optimal tuning parameter values were selected based on area under the receiver operating characteristic curve (AUC ROC).
@@
 PCs were sequentially added as features until the AUC ROC of the model did not increase.
+Random forest model parameters were estimated from 1000 subtrees.
 After each PC was added, 10-fold CV was used to estimate the optimal values of the tuning parameters as well as quantify the uncertainty of each model.
@@
 The second best clustering solution is two clusters, however there is no geographic structure to this classification scheme SUPPLEMENT?.
+Our dataset does not include enough or detailed information on the sex of each \textit{E.
+marmorata} specimen, thus it is not possible to determine if this clustering solution corresponds to sexual dimorphism between the observations.
+Male Emyidine turtles are known to have a plastral concavity
 Increasing the number of clusters does appear to improve the gap statistic enough to merit comparison.
@@
 For all classification schemes, the optimal random forest model based on recursive feature selection by maximizing AUC ROC of the model was one with many features (Fig. \ref{fig:roc})
+% need to discuss the final models in both cases.
 % figure
@@
 
-When the classification results of the training set for the optimal classification scheme are compared with the references classes, (Fig. \ref{fig:gen_map})
+When the classification results of the training set for the optimal classification scheme are compared with the references classes, the higher AUC ROC value of the best multinomial logistic regression model becomes apparent as the classifications are in general much more similar to the reference (Fig.
+\ref{fig:gen_map}).
+The best random forest model misclassified many of the observations as the northern clade.
+This trend is observable but not as exaggerated in the results of the classifications of the multinomial logistic regression model.
 
@@
 
+This pattern of misclassification may be caused by the differences in mean shape between each of the different classes (Fig.
+\ref{fig:mean_shape}).
+The mean shape of the northern clade is the most similar to the mean shape of the entire dataset (Fig.
+), which may mean that specimens that are closer to the mean shape will be systematically misclassified as the northern clade.
+
+% mean shapes of all the classes
+\begin{figure}[ht]
+  \centering
+  \begin{subfigure}[b]{0.4\textwidth}
+    \centering
+    \includegraphics[width = \textwidth]{figure/mshape_1}
+    \label{fig:mean_shape1}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.4\textwidth}
+    \centering
+    \includegraphics[width = \textwidth]{figure/mshape_2}
+    \label{fig:mean_shape2}
+  \end{subfigure}
+  % new line. might want to make these closer together
+
+  \begin{subfigure}[b]{0.4\textwidth}
+    \centering
+    \includegraphics[width = \textwidth]{figure/mshape_3}
+    \label{fig:mean_shape3}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.4\textwidth}
+    \centering
+    \includegraphics[width = \textwidth]{figure/mshape_4}
+    \label{fig:mean_shape4}
+  \end{subfigure}
+  \caption{<+caption text+>}
+  \label{fig:mean_shape}
+\end{figure}
+
+% variation along the most important axes
+The results of training the random forest model also include the variable importance for best separating the different classes.
+Recursive feature selection of the best random forest model of the chosen classification scheme indicated that after \Sexpr{trf.a$spinks$optsize} PCs were included as features, AUC ROC would not increase.
+Of these \Sexpr{trf.a$spinks$optsize} features, three are illustrated here (Fig.
+\ref{fig:imp_pc}) the first two of which are most important SUPPLEMENT WITH VARIABLE IMPORTANCE INFORMATION?.
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width = \textwidth]{figure/pca_imp}
+  \caption{Pairs plot of the first three most important variables of the optimal random forest model of turtle plastral shape.
+The variables descend in importance from the upper left to the lower right.
+The observations are colored as in Figures \ref{fig:gen_res} and \ref{fig:gen_map}.
+The bottom row are histograms of classification occurrences along the PCs.}
+  \label{fig:imp_pc}
+\end{figure}
+% need to add in variance percentages on the different axes.
+
+The first two most important features, according to the random forest model, describe different aspects of variation (Fig.
+\ref{fig:imp_var}).
+The third PC, or first most important PC, describes variation in the relative position of landmarks on anterior and posterior portions of the plastron.
+The eighth PC, or second most important PC, mostly describes variation in landmarks along the midline of the plastron.
+The major variation along these axes correspond well to the differences between the class means and the mean plastron shape (Fig.
+\ref{fig:mean_shape}) where major class differences seems restricted to the relative ballooning or shrinking of the anterior and posterior portions of the plastron together.
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width = \textwidth]{figure/imp_var}
+  \caption{Landmark variation along the two most important features (PCs) based on the final random forest model.
+The first row corresponds to the third PC and the second corresponds to eighth PC.
+Landmark configurations are minimum observed on that PC, mean shape, and maximum observed on that PC.}
+  \label{fig:imp_var}
+\end{figure}
+
 \section{Discussion}
+% discussion of results
+%   remarkable concordance in the supervised learning methods
+The results of this study provide support for the mitochrondial based hypothesis of classification of \textit{E.
+marmorata} \citep{Spinks2005,Spinks2010}.
+This is contrary to the original classification of \textit{E.
+marmorata} \citep{Seeliger1945} and lends credence to the idea that at least some aspect of cryptic diversity is a product either sample size, methodology, or both.
+%   what does this mean about cryptic diversity?
+%   unsupervised learning shortcomings
+
+The lack of coherent geographical subclass assignment from PAM based clustering (Fig.
+\ref{fig:gap}) as well as the large number of features necessary before plateau in AUC ROC for all models (Fig.
+\ref{fig:roc}) can be taken as indicators that the variation between subclasses is extremely fine grained.
+This is also exemplified by the differences in mean class shape for the final chosen classification scheme (Fig.
+\ref{fig:imp_var}).
+
+% methodological concerns
+%   compromise in the supervised learning models
+%     fair because as much variation as ``necessary'' is used
+%   unsupervised learning
+%     model based clustering
+%     nonparametric bayesian approach
+%     anything about semisupervised clustering? probably not
+Ultimately, it would be optimal to not require such explicit classification hypotheses, especially when concerned about possible cryptic variation in extinct taxa.
+The method employed in this study, PAM, is rather simple and not model based.
+Comparison is facilitated by comparison of a summary statistic and bootstrap confidence intervals.
+A more useful and future avenue would be employing various model based clustering approaches CITATIONS.
+In this manner, a series of candidate models can be compared via model comparison methods, such as AIC or Bayes factors CITATION, in order to asses the best clustering solution.
+Of particular note are nonparametric Bayesian approaches to model based clustering CITATIONS.
+This approach uses a class of flexible priors to allow for the most optimal clustering solution to be decided from the data.
+Currently, there exists a nonparametric Bayesian clustering method and further development of this approach may prove extremely fruitful for better deliminating taxa solely from morphology.
+
+% future directions
+%   conservation utility
+%   fossil diversity
+%     current approach favors having an extremely explicit classification hypothesis
+%     as indicated about, nonparametric bayes and other model based clustering can help
+
+% closing statements
+In this study we have demonstrated that given a large sample size and appropriate methodology, it is possible to determine which of multiple hypothesized classification schemes best explain variation in a taxon.
+The plastral variation of \textit{E.
+marmorata} is most consistent with the mitochondrial based hypothesis of \citet{Spinks2005} and \citet{Spinks2010} and not with the original morphology based hypothesis of \citet{Seeliger1945}.
+We have also demonstrated the utility of various machine learning approaches to understanding variation in morphometric data.
+Specifically, better understanding odds misclassification and identifying which is the most important for deliminating different classes.
 
@@
 
-\bibliographystyle{amnatnat}
+\bibliographystyle{sysbio}
 \bibliography{turtle,packages}

### 2013-07-08 documents/turtle_paper.tex
lots of paper work. new figure. changed target to syst biol

@@
 
+This pattern of misclassification may be caused by the differences in mean shape between each of the different classes (Fig.
+\ref{fig:mean_shape}).
+The mean shape of the northern clade is the most similar to the mean shape of the entire dataset (Fig.
+), which may mean that specimens that are closer to the mean shape will be systematically misclassified as the northern clade.
+
+% mean shapes of all the classes
+\begin{figure}[ht]
+  \centering
+  \begin{subfigure}[b]{0.4\textwidth}
+    \centering
+    \includegraphics[width = \textwidth]{figure/mshape_1}
+    \label{fig:mean_shape1}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.4\textwidth}
+    \centering
+    \includegraphics[width = \textwidth]{figure/mshape_2}
+    \label{fig:mean_shape2}
+  \end{subfigure}
+  % new line. might want to make these closer together
+
+  \begin{subfigure}[b]{0.4\textwidth}
+    \centering
+    \includegraphics[width = \textwidth]{figure/mshape_3}
+    \label{fig:mean_shape3}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.4\textwidth}
+    \centering
+    \includegraphics[width = \textwidth]{figure/mshape_4}
+    \label{fig:mean_shape4}
+  \end{subfigure}
+  \caption{<+caption text+>}
+  \label{fig:mean_shape}
+\end{figure}
+
+% variation along the most important axes
+The results of training the random forest model also include the variable importance for best separating the different classes.
+Recursive feature selection of the best random forest model of the chosen classification scheme indicated that after 7 PCs were included as features, AUC ROC would not increase.
+Of these 7 features, three are illustrated here (Fig.
+\ref{fig:imp_pc}) the first two of which are most important SUPPLEMENT WITH VARIABLE IMPORTANCE INFORMATION?.
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width = \textwidth]{figure/pca_imp}
+  \caption{Pairs plot of the first three most important variables of the optimal random forest model of turtle plastral shape.
+The variables descend in importance from the upper left to the lower right.
+The observations are colored as in Figures \ref{fig:gen_res} and \ref{fig:gen_map}.
+The bottom row are histograms of classification occurrences along the PCs.}
+  \label{fig:imp_pc}
+\end{figure}
+% need to add in variance percentages on the different axes.
+
+The first two most important features, according to the random forest model, describe different aspects of variation (Fig.
+\ref{fig:imp_var}).
+The third PC, or first most important PC, describes variation in the relative position of landmarks on anterior and posterior portions of the plastron.
+The eighth PC, or second most important PC, mostly describes variation in landmarks along the midline of the plastron.
+The major variation along these axes correspond well to the differences between the class means and the mean plastron shape (Fig.
+\ref{fig:mean_shape}) where major class differences seems restricted to the relative ballooning or shrinking of the anterior and posterior portions of the plastron together.
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width = \textwidth]{figure/imp_var}
+  \caption{Landmark variation along the two most important features (PCs) based on the final random forest model.
+The first row corresponds to the third PC and the second corresponds to eighth PC.
+Landmark configurations are minimum observed on that PC, mean shape, and maximum observed on that PC.}
+  \label{fig:imp_var}
+\end{figure}
+
 \section{Discussion}
+% discussion of results
+%   remarkable concordance in the supervised learning methods
+The results of this study provide support for the mitochrondial based hypothesis of classification of \textit{E.
+marmorata} \citep{Spinks2005,Spinks2010}.
+This is contrary to the original classification of \textit{E.
+marmorata} \citep{Seeliger1945} and lends credence to the idea that at least some aspect of cryptic diversity is a product either sample size, methodology, or both.
+%   what does this mean about cryptic diversity?
+%   unsupervised learning shortcomings
+
+The lack of coherent geographical subclass assignment from PAM based clustering (Fig.
+\ref{fig:gap}) as well as the large number of features necessary before plateau in AUC ROC for all models (Fig.
+\ref{fig:roc}) can be taken as indicators that the variation between subclasses is extremely fine grained.
+This is also exemplified by the differences in mean class shape for the final chosen classification scheme (Fig.
+\ref{fig:imp_var}).
+
+% methodological concerns
+%   compromise in the supervised learning models
+%     fair because as much variation as ``necessary'' is used
+%   unsupervised learning
+%     model based clustering
+%     nonparametric bayesian approach
+%     anything about semisupervised clustering? probably not
+Ultimately, it would be optimal to not require such explicit classification hypotheses, especially when concerned about possible cryptic variation in extinct taxa.
+The method employed in this study, PAM, is rather simple and not model based.
+Comparison is facilitated by comparison of a summary statistic and bootstrap confidence intervals.
+A more useful and future avenue would be employing various model based clustering approaches CITATIONS.
+In this manner, a series of candidate models can be compared via model comparison methods, such as AIC or Bayes factors CITATION, in order to asses the best clustering solution.
+Of particular note are nonparametric Bayesian approaches to model based clustering CITATIONS.
+This approach uses a class of flexible priors to allow for the most optimal clustering solution to be decided from the data.
+Currently, there exists a nonparametric Bayesian clustering method and further development of this approach may prove extremely fruitful for better deliminating taxa solely from morphology.
+
+% future directions
+%   conservation utility
+%   fossil diversity
+%     current approach favors having an extremely explicit classification hypothesis
+%     as indicated about, nonparametric bayes and other model based clustering can help
+
+% closing statements
+In this study we have demonstrated that given a large sample size and appropriate methodology, it is possible to determine which of multiple hypothesized classification schemes best explain variation in a taxon.
+The plastral variation of \textit{E.
+marmorata} is most consistent with the mitochondrial based hypothesis of \citet{Spinks2005} and \citet{Spinks2010} and not with the original morphology based hypothesis of \citet{Seeliger1945}.
+We have also demonstrated the utility of various machine learning approaches to understanding variation in morphometric data.
+Specifically, better understanding odds misclassification and identifying which is the most important for deliminating different classes.
 

### 2013-07-15 documents/turtle_paper.Rnw
more draft work

@@
 
-\documentclass[12pt]{article}
+\documentclass[12pt,letterpaper]{article}
 \usepackage{amsmath, amsthm}
@@
 \usepackage[sort&compress]{natbib}
-%\usepackage{fullpage}
+\usepackage{fullpage}
 
@@
 \frenchspacing
+\linespread{1.66}
+\setlength{\parindent}{0.5in}
+
+\setcounter{secnumdepth}{0}
+
+\pagestyle{empty}
+
+\renewcommand{\section}[1]{%
+\bigskip
+\begin{center}
+\begin{Large}
+\normalfont\scshape #1
+\medskip
+\end{Large}
+\end{center}}
+
+\renewcommand{\subsection}[1]{%
+\bigskip
+\begin{center}
+\begin{large}
+\normalfont\itshape #1
+\end{large}
+\end{center}}
+
+\renewcommand{\subsubsection}[1]{%
+\vspace{2ex}
+\noindent
+\textit{#1.}---}
+
+\renewcommand{\tableofcontents}{}
+
+\bibpunct{(}{)}{;}{a}{}{,}  % this is a citation format command for natbib
 
@@
 \title{How cryptic is cryptic diversity? Machine learning approaches to plastral variation in \textit{Emys marmorata}.}
-\author[1]{Peter D Smits \thanks{psmits@uchicago.edu}}
-\author[2]{Kenneth D Angielczyk \thanks{kangielczyk@fieldmuseum.org}}
-\author[3]{James F Parham \thanks{jparham@fullerton.edu}}
+\author[1]{Peter D Smits}%\thanks{psmits@uchicago.edu}}
+\author[2]{Kenneth D Angielczyk}%\thanks{kangielczyk@fieldmuseum.org}}
+\author[3]{James F Parham}%\thanks{jparham@fullerton.edu}}
 \affil[1]{Committee on Evolution Biology, University of Chicago}
@@
 \begin{document}
+\maketitle
+\noindent{\textbf{Crresponding author:} Peter D Smits, Committee on Evolutionary Biology, University of Chicago, ADDRESS, STATE, ZIP, USA; E-mail: \href{mailto:psmits@uchicago.edu}{psmits@uchicago.edu}}
 
-\maketitle
-
-\linenumbers
-\modulolinenumbers[2]
+%\linenumbers
+%\modulolinenumbers[2]
 
@@
   % 200 words
-
+\noindent (Keywords: Testudines, morphology, geometric morphometrics)
 \end{abstract}
 
-\section{Introduction}
+%\section{Introduction}
 
@@
 %   most species are still deliminated solely based on morphology
+Cryptic diversity is when taxa were only first deliminated via molecular means and were not or cannot deliminated via morphological identification CITATION.
+The discovery of this previously unknown diversity has expanded diversity amoungst certain taxa CITATION.
+Additionally, a suite of methods have been developed to more accurately deliminate taxa based on molecular divergence CITATION.
+% all the work done on species delimitation strictly from molecular data
+
 %   paleo problem
-Cryptic diversity is when taxa were only first deliminated via molecular means and were not or cannot deliminated via morphological identification CITATION.
-The discovery of this previously unknown diversity has
-% all the work done on species delimitation strictly from molecular data
-%
+Concerns about cryptic diversity are enhanced when taxa are only known from morphology, such as the case of extinct organisms CITATION.
 
@@
 % GP superimposition and PCA
-``Half'' plastra landmark configurations were superimposed using generalized Procrustes analysis \citep{Dryden1998a} after which, the principal components of shape were calculated.
+``Half'' plastra landmark configurations were superimposed using generalized Procrustes analysis \citep{Dryden1998a} after which, the principal components (PC) of shape were calculated.
 This was done using the \texttt{shapes} package for R \citep{2013, Dryden2013}.
@@
 Optimal tuning parameter values were selected based on area under the receiver operating characteristic curve (AUC ROC).
-Multiclass AUC ROC was estimated using the all-against one strategy derived by \citet{Hand2001} in implemented in PROC PACKGE.
+Multiclass AUC ROC was estimated using the all-against one strategy derived by \citet{Hand2001} in implemented in PROC PACKAGE.
 
@@
 \subsubsection{Supervised learning}
-For all classification schemes, the optimal random forest model based on recursive feature selection by maximizing AUC ROC of the model was one with many features (Fig. \ref{fig:roc})
+%For all of the classification schemes, the optimal models for both the multinomial logistic regression (Table SUPPLEMENT) and random forest (Fig.
+\ref{fig:roc}) methods include many, if not most, of the possible features.
+<<>>=
+daic <- unlist(lapply(lapply(tm.a.analysis$sel, function(x) x$delta), 
+                      function(x) x[2]))
+@
+% multinomial logistic regression model
+The optimal multinomial logistic regression models for all four classification schemes, based on AICc, have all \Sexpr{length(tm.a.analysis$best$spinks$coefnames)} possible PCs as features(Table SUPPLEMENT).
+The second best models for all classification schemes have \Sexpr{length(tm.a.analysis$best$spinks$coefnames) - 1} PCs as features.
+The \(\delta\)AICc values between the optimal and second best model range from \Sexpr{min(daic)} for the first morphological based classification hypothesis to \Sexpr{max(daic)} for the second molecular based classification hypothesis.
+
+While the \(\delta\)AICc value between the optimal and second best model for the first morphological based classification hypothesis is within the range to be considered sufficient and equally optimal \citep{Burnham2002a}, for this analysis we chose to use only the absolute AICc best model.
+While AICc values can not be compared between models with different responses \citep{Burnham2002a}, we interpret the fact that the optimal model for all classification schemes is the global model as a reason to just use this model structure for all cases.
+Additionally, by maintaining a single optimal model for all classification schemes, this limits the total number of comparisons between the bootstrap resampled distributions of the AUC ROC values for the testing data set (see below).
+
+% random forest model
+<<>>=
+oprf <- unlist(lapply(trf.a, function(x) x$optsize))
+@
+
+The selected number of features in the optimal random forest models for each classification scheme vary much more than in the case of the multinomial logistic regression models (Fig.
+\ref{fig:roc}), ranging from \Sexpr{min(oprf)} for the first morphological based classification hypothesis to \Sexpr{max(oprf)} for the second morphological based classification hypothesis.
+
 % need to discuss the final models in both cases.
@@
 %     facet: multinomial, rf (see talk)
+
+In the case of all models, there is a substantial increase in model performance as measured by AICc for the multinomial logistic models (Tables SUPPLEMENT) or in AUC ROC for the random forest models and illustrated for the multinomial logistic regression models as the number of features increases (Fig.
+\ref{fig:roc}).
 
@@
 The mean shape of the northern clade is the most similar to the mean shape of the entire dataset (Fig.
-), which may mean that specimens that are closer to the mean shape will be systematically misclassified as the northern clade.
+\ref{fig:mean_shape}\ref{fig:imp_var}), which may mean that specimens that are closer to the mean shape will be systematically misclassified as the northern clade.
 
@@
 %     fair because as much variation as ``necessary'' is used
+The methods presented here for supervised learning analysis of the landmark variation represent a compromise
+
 %   unsupervised learning
@@
 
-\section*{Acknowledgements}
+\section{Acknowledgements}
 PDS would like to thank David Bapst, Michael Foote, Benjamin Frable, and Dallas Krentzel for useful discussion which enhanced the quality of this study.

### 2013-07-15 documents/turtle_paper.tex
more draft work

@@
 
-\documentclass[12pt]{article}\usepackage{graphicx, color}
+\documentclass[12pt,letterpaper]{article}\usepackage{graphicx, color}
 %% maxwidth is the original width if it is less than linewidth
@@
 \subsubsection{Supervised learning}
-For all classification schemes, the optimal random forest model based on recursive feature selection by maximizing AUC ROC of the model was one with many features (Fig. \ref{fig:roc})
+%For all of the classification schemes, the optimal models for both the multinomial logistic regression (Table SUPPLEMENT) and random forest (Fig.
+\ref{fig:roc}) methods include many, if not most, of the possible features.
+
+% multinomial logistic regression model
+The optimal multinomial logistic regression models for all four classification schemes, based on AICc, have all 10 possible PCs as features(Table SUPPLEMENT).
+The second best models for all classification schemes have 9 PCs as features.
+The \(\delta\)AICc values between the optimal and second best model range from 2.0639 for the first morphological based classification hypothesis to 19.8349 for the second molecular based classification hypothesis.
+
+While the \(\delta\)AICc value between the optimal and second best model for the first morphological based classification hypothesis is within the range to be considered sufficient and equally optimal \citep{Burnham2002a}, for this analysis we chose to use only the absolute AICc best model.
+While AICc values can not be compared between models with different responses \citep{Burnham2002a}, we interpret the fact that the optimal model for all classification schemes is the global model as a reason to just use this model structure for all cases.
+Additionally, by maintaining a single optimal model for all classification schemes, this limits the total number of comparisons between the bootstrap resampled distributions of the AUC ROC values for the testing data set (see below).
+
+% random forest model
+
+The selected number of features in the optimal random forest models for each classification scheme vary much more than in the case of the multinomial logistic regression models (Fig.
+\ref{fig:roc}), ranging from 6 for the first morphological based classification hypothesis to 10 for the second morphological based classification hypothesis.
+
 % need to discuss the final models in both cases.

### 2013-07-22 documents/turtle_paper.Rnw
first major round of edits on paper. probably need about 2 more.

@@
 % for american naturalist
-\usepackage{lineno}
+%\usepackage{lineno}
 
@@
 
-\pagestyle{empty}
+%\pagestyle{empty}
 
@@
 \author[3]{James F Parham}%\thanks{jparham@fullerton.edu}}
-\affil[1]{Committee on Evolution Biology, University of Chicago}
+\affil[1]{Committee on Evolutionary Biology, University of Chicago}
 \affil[2]{Department of Geology, Field Museum of Natural History}
@@
 \maketitle
-\noindent{\textbf{Crresponding author:} Peter D Smits, Committee on Evolutionary Biology, University of Chicago, ADDRESS, STATE, ZIP, USA; E-mail: \href{mailto:psmits@uchicago.edu}{psmits@uchicago.edu}}
+\noindent{\textbf{Corresponding author:} Peter D Smits, Committee on Evolutionary Biology, University of Chicago, 1025 E.
+57th Street, Culver Hall 402, Chicago, IL, 60637, USA; E-mail: \href{mailto:psmits@uchicago.edu}{psmits@uchicago.edu}}
 
@@
 \begin{abstract}
+  Write me please.
   % 200 words
@@
 %   most species are still deliminated solely based on morphology
-Cryptic diversity is when taxa were only first deliminated via molecular means and were not or cannot deliminated via morphological identification CITATION.
+%Cryptic diversity is when taxa were only first deliminated via molecular means and were not or cannot deliminated via morphological identification CITATION.
 The discovery of this previously unknown diversity has expanded diversity amoungst certain taxa CITATION.
 Additionally, a suite of methods have been developed to more accurately deliminate taxa based on molecular divergence CITATION.
+
+Cryptic diversity is the phenomenon that not all taxa can be recognized based on morphology and can only be delimited using molecular information.
+The most extant taxa, and nearly all extinct taxa, are delimited based solely via morphological analysis.
+Concerns about how pervasive this phenomenon is and much diversity is actually cryptic have
+
+Much work has been devoted to species delimitation via sequence difference CITATIONS while comparatively little has been devoted for case of morphological data.
+The majority of this effort has focused on automated taxon identification CITATIONS or identifying morphological differences amongst already identified taxa CITATIONS.
 % all the work done on species delimitation strictly from molecular data
@@
 
-Here, we address the question of how much of cryptic diversity may be a product of sample size as well as methodology used for classifying taxa based solely on morphology.
-Specifically, we ask if fine scale variation in morphology can provide corroboration for subspecific assignment, and if it is possible to determine the best classification hypothesis amongst a few.
+Here, we address two questions  about cryptic diversity: how much of cryptic diversity is a product of sample size, and how can methodology impact classification based solely on morphology.
+Additionally, we ask if fine scale variation in morphology can provide corroboration for various classification hypotheses, and if it is possible to determine which is the best classification hypothesis.
 
@@
 % why use machine learning methods
-In the analysis of class differences and identification in morphometric analysis, classification methods such as discriminant analysis and canonical variates analysis CITATIONS have been used frequently for understanding how variation in specific morphologies best differentiate classes of taxa.
-Additionally, there has been work on the application of neural network models in classification schemes of observations especially in the context of automated taxon identification CITATIONS MACLEOD'S BOOK.
-Here, we used multiple types of machine learning methods, both unsupervised and supervised, in order to understand the best classification scheme of the taxa.
-Each method provides a series of unique advantages for understanding how to classify taxa, which are discussed below.
+Differences in morphological variation between different classes has previously been analyzed using methods like linear discriminate analysis and canonical variates analysis CITATION ZELDICH.
+Neural network models have also been introduced and applied in the context of automated taxon identification along with more general applications MACLEOD BOOK.
+Here, we used multiple alternative machine learning methods, both unsupervised and supervised, in order to compare different classification hypotheses.
+These different methods provide different and unique advantages for understanding how to classify taxa, with what accuracy, and what these classifications are based on.
+Additionally, we investigate variation in continuous traits, and do not search for discrete differences between each class, instead focusing on suites of traits together.
+
+% this paragraph kind of sucks
+The two major cases of machine learning methods, unsupervised and supervised, are essentially extensions of known statistical methods.
+Unsupervised learning methods are analogous to clustering and density estimation methods, while supervised learning methods are analogous classification and regression models.
+In both cases, these methods may not be fit via maximum likelihood but may be supplemented by randomization algorithms and the maximization or minimization of summary statistics in order to best estimate a general model for all data, but sampled and unsampled.
+The expansion of methods used to understand morphological differentiation
 
@@
 marmorata marmorata}, the southern \textit{E.
-marmorata palida}, and a central Californian intergrade zone \citep{Seeliger1945}.
+marmorata palida}, and a central Californian intergrade zone \citep{Seeliger1945} UNPUBLISHED MASTERS THESIS.
 \textit{E.
@@
 marmorata palida} though they are considerably smaller.
+
+Previous work on morphological differentiation between subspecies of \textit{E. marmorata} focused on just the known subspecies of \citet{Seeliger1945}
+
 More recently, \textit{E.
@@
 Nuclear DNA supports two major clades, one northern and one southern, however \citet{Spinks2010} argue that the four clade classification is of greater conservation utility to use the mitochondrial classification scheme.
-There is now known morphological differentiation between these clades.
+There is no known morphological differentiation between these clades.
 
@@
 % statement of goals and approach
-In this study, we apply multiple machine learning approaches to estimate the best classification scheme of \textit{E.
+In this study, we attempt to estimate the best classification scheme of \textit{E.
 marmorata} subspecies based on morphological variation in plastral shape.
@@
 marmorata}, we compare two hypotheses of morphology-based classification and two hypotheses of molecular-based classification.
+We hypotheize that if morphological variation corresponds to subspecific assignment, then it should be possible to differentiate the best classification hypothesis of \textit{E.
+marmorata} from amongst multiple candidate hypotheses.
+However, if morphological variation variation does not correspond to any classification hypothesis, then supervised learning model generalization performance will be extremely poor and reflect how variation does not follow along with classification.
 
@@
 \subsection{Specimens}
-We collected morphometric data from \Sexpr{nrow(turtle.adult)} specimens.
+We collected landmark-based morphometric data from \Sexpr{nrow(turtle.adult)} adult \textit{E.
+marmorata} museum specimens.
 Geographic information was recorded from museum collection information.
-When precise latitude and longitude information was not known for a specimen, it was inferred from whatever locality information was presented.
+When precise latitude and longitude information was not known for a specimen, it was inferred from whatever locality information was present.
 % museum information?
 
-Specimens were given a class assignment was based on geographic information.
-Because the exact geographic barriers between different class is unknown and fuzzy, two assignments for both morphological and molecular hypotheses of class were used.
+Specimens were classified based on museum record geographic information.
+The specimens used to define the subclades in \citet{Spinks2005} and \citet{Spinks2010} were not available for study, all classifications were based solely on the geographic information for each specimen and not from explicit assignment in previous studies.
+Because the exact geographic barriers between different classes are unknown and unclear, two assignments for both the morphological and molecular hypotheses were used.
+Each morphology-based hypothesis had three classes, while each molecular-based had four classes.
+In total, each specimen was given four different classifications.
 
@@
 17 of these landmarks are at the endpoints or intersection of the keratinous plastral scutes that cover the platron.
-These landmarks were chosen to maximize the description of plastral variation.
-12 of these landmarks are symmetrical across the axis of symmetry and in order to prevent degrees of freedom and other concerns \citep{Klingenberg2007}, these landmarks were reflected across the axis of symmetry and the average position of each symmetrical pair was used.
-In cases where damage or incompleteness prevented symetric landmarks from being determined, only the single member of the pair was used.
+These landmarks were chosen to maximize the description of general plastral variation.
+12 of these landmarks were chosen to be symmetrical across the axis of symmetry and in order to prevent degrees of freedom and other concerns \citep{Klingenberg2007} prior to analysis these landmarks were reflected across the axis of symmetry (i.e.
+midline) and the average position of each symmetrical pair was used.
+In cases where damage or incompleteness prevented symmetric landmarks from being determined, only the single member of the pair was used.
 Analysis was then conducted on the resulting ``half'' plastra.
@@
 %   Riemannian shape distance
-Because shape space, or configurations after Procrustes superimposition, is a Riemannian manifold \citep{Dryden1998a} the dissimilarity between each landmark configuration was measured as the Riemmanian shape distance or \(\rho\) \citep{Kendall1984a,Dryden1998a} which should vary between 0 and \(\pi / 2\) assuming no reflection invariance.
+In order to preserve the relationship in shape space between all landmark configurations, the dissimilarity between observations was measured using the Riemanninan shape distance or \(\rho\) \citep{Kendall1984a,Dryden1998a}.
+This metric was chosen because shape space, or the set of all possible shape configurations following Procrustes superimposition, is a Riemannian manifold and thus non-Euclidean \citep{Dryden1998a}.
+\(\rho\) varies between 0 and \(\pi / 2\) when there is no reflection invariance, which should not be a concern in the case of the half plastral landmark configurations.
 
 %   PAM
-The dissimilarity matrix of shape was divisivly clustering using partioning around mediods (PAM) which is analogous to \textit{k}-means clustering except that instead of minimizing the sum of squared Euclidean distances between observations and centroids, the sum of squared dissimilarities between observations and mediods is minimized \citep{Kaufman1990}.
+The \(\rho\) dissimilarity matrix was divisively clustering using partitioning around mediods (PAM), a method which is similar to \textit{k}-means clustering except that instead of minimizing the sum of squared Euclidean distances between observations and centroids, the sum of squared dissimilarities between observations and mediods is minimized \citep{Kaufman1990}.
 %   gap statistic
-The optimal number of clusters of shape configurations is unknown being possibly three, four, or some other value.
-Clustering solutions were estimated for between 1 and \Sexpr{nrow(tadult.gap$Tab)} clusters.
+Because the optimal number of clusters of shape configurations in the study was unknown, being possibly three, four, or some other value, clustering solutions were estimated with the number of mediods varying between one and \Sexpr{nrow(tadult.gap$Tab)}.
 Clustering solutions were compared using the gap statistic, which is a measure of goodness of clustering \citep{Tibshirani2001a}.
-Standard errors of the gap statistic for each clustering solution were estimated from \Sexpr{tadult.gap$B} bootstrap samples.
+The gap statistic is defined
+\[Gap_{n}(k) = E^{*}_{n}(\log(W_{k}) - \log(W_{k})\] 
+where \(W_{k}\) is
+\[W_{k} = \sum^{k}_{r = 1}{\frac{1}{2n_{r}} (\sum_{i,i' \in C_{r}} d_{ii'})}\]
+the dispersion of the clustering solution or the sum of the pairwise dissimilarities (\(d_{ii'}\)) between observations in each cluster and there respective mediods (\(C\)) for all clusters \(r\).
+\(E^{*}_{n}\) is the expectation of dispersion from a sample \(n\) of a reference distribution.
+In this case, the reference distribution was estimated from a \Sexpr{tadult.gap$B} resamples of the dataset taking into account the original structure of the data.
 %  R implementation
-PAM clustering and gap statistic calculation was conducted using the \texttt{cluster} package for R \citep{Maechler2013}.
+This analysis was conducted using the \texttt{cluster} package for R \citep{Maechler2013} using all \Sexpr{nrow(turtle.adult)} plastral landmark configurations.
 
@@
 % training testing split
-The dataset of \Sexpr{nrow(turtle.adult)} plastron landmarks was split into training and testing datasets.
-The former was used for model fitting (training) and was 75\% of the total dataset, split proportionally per class, while the testing dataset was used to estimate the effectiveness of each classification scheme (i.e.
+The total dataset of \Sexpr{nrow(turtle.adult)} observations was split into training and testing datasets.
+The training dataset represented 75\% of the total dataset, split proportionally per class, and was used for model fitting.
+The testing dataset represented the remaining 25\% of the total dataset and  was used to estimate the effectiveness of each classification hypothesis and generalizability of the supervised learning models (i.e.
 performance in the wild).
@@
 % model types
-Two types of supervised learning, or classification, models were fit to the PCs of plastral shape: multinomial logistic regression and random forest.
-These model types were chosen because of various properties of these models which allow for useful interpretations about the strength and structure of the classification.
-Multinomial logistic regression models were fit using the \texttt{nnet} package for R \citep{Venables2002} while random forest models were fit using the \texttt{randomForest} package for R \citep{Liaw2002}.
+Two different supervised learning methods and model types were used to model the relationship between plastral shape and class: multinomial logistic regression and random forest.
+These models were chosen because of various properties of these models which allow for useful interpretations about the quality and structure of the classification.
 
 %   multinomial logistic regression
-Multinomial logistic regression is an extension of logistic regression, where instead of a binary response it is possible to have three or more response classes CITATION.
-Effectively, this type of model can be viewed as multiple, simultaneous logistic regression models for each class and the final classification of the observation being the most probable of all the sub-model classifications.
-From the final model the relative risk of a given classification, with reference to a given class, can be calculated from the coefficients of the features, or predictors.
-This is similar to the log-odds calculated from the coefficients of a logistic regression.
+Multinomial logistic regression is an extension of logistic regression, where instead of a binary response there are three or more response classes CITATION.
+Effectively, this type of model can be viewed as multiple, simultaneous logistic regression models for each class and the final classification of the observation being the most probable of all the constituent model classifications.
+From the final model, the relative risk of a possible classification, with reference to a baseline class, can be calculated from the coefficients of the model.
+These are similar to the odds ratios calculated from the coefficients of a logistic regression.
+Multinomial logistic regression models were fit using the \texttt{nnet} package for R \citep{Venables2002}
 
@@
 Random forest models are an extension of classification and regression trees (CART) CITATION.
-Basically, CARTs are built for random subsamples of both the features of the proposed model and observations.
-This process is repeated many times, 1000 times here, and the final model is chosen as the mode of the parameter estimates from the distribution of CARTs CITATION.
-In addition to fitting a classification model, this procedure allows for the features to be ranked in order of importance, means that the variables most important for determining a given classification scheme can be estimated.
-In the context of predicting class from geometric morphometric data, this identifies the PCs that describe the variation that best distinguishes the different classes.
+In a random forest model, many CARTs are built from a random subsample of both the features and the observations.
+This process was repeated 1000 times and the parameters of the final model was chosen as the mode of estimates from the distribution of CARTs CITATION.
+In addition to fitting a classification model, this procedure allows for the features to be ranked in order of importance.
+In the context of this study, this means that the PCs most important for describing the difference between classes can be estimated, and thus illustrate the most important variation amongst classes as opposed to just the greatest amount of variation in the entire dataset.
+This is a generally important property that should be useful for other morphometric studies which want to describe and model the shape differences between different classes and the relative importance of various aspects of variation.
+Random forest models were fit using the \texttt{randomForest} package for R \citep{Liaw2002}.
 %   R statement
@@
 %       what are the tuning parameters for each model?
-%     best AUC ROC
-For both types of supervised learning models, tuning parameters were estimated using 10 rounds of 10-fold cross-validation (CV) across a grid search of all tuning parameter combinations.
-Optimal tuning parameter values were selected based on area under the receiver operating characteristic curve (AUC ROC).
-Multiclass AUC ROC was estimated using the all-against one strategy derived by \citet{Hand2001} in implemented in PROC PACKAGE.
+%     best AUC
+Supervised learning models have tuning parameters which are estimated to increase the genearlizability of the model and prevent them from being overfit.
+For both types of supervised learning methods, tuning parameters were estimated using 10 rounds of 10-fold cross-validation (CV) across a grid search of all tuning parameter combinations.
+Optimal tuning parameter values were selected based on area under the receiver operating characteristic curve (ROC).
+The area under the multiclass ROC curves was estimated using the all-against one strategy derived by \citet{Hand2001}.
+
+% what is ROC? what is AUC?
+ROC is a type of confusion matrix statistic that is a descriptor the relationship between the false positive rate or 1 - specificity of a classification model and the true positive rate or sensitivity of a classification model CITATION.
+The area under the ROC curve (AUC) is a summary statistic of the quality of the classification and varies between 0.5 and 1, with 0.5 indicating a model that classifies no better than random and a value of 1 indicating perfect classification CITATION.
+AUC can be used as a model selection criterion for classification models and is especially useful in cases where some if not all of the models in question were not fit via maximum likelihood where a criterion such as AICc (see below) or similar can be used CITATION.
 
@@
 %     AICc
-For the multinomial logistic regression models, PCs were added sequentially in order to increase the overall amount of variation in shape included in each model and the final model was that with the lowest AICc \citep{Burnham2002a} AKAIKE AND OTHER CITATION.
-This procedure was used because the optimal number of PCs to include is unknown, and while including all of the PCs of shape would mean that all of the variability in plastron shape would be used to estimate class, this may cause the model to be over fit and not provide an accurate estimate of unsampled plastral variation.
+For the multinomial logistic regression models, \Sexpr{floor(max.ad)} different models were fit with each having sequentially more PCs as predictors in order to have models representing different levels of overall amount of shape variation and estimate how much variation was necessary and sufficient to best estimate class.
 The maximum number of PCs allowed to be used as predictors was \Sexpr{floor(max.ad)} because of both the number of parameters estimated per model and the necessary sample size needed to estimate that many parameters accurately.
+The final model was that with the lowest AICc \citep{Burnham2002a} AKAIKE AND OTHER CITATION.
+Similar to AUC discussed above, AICc is a model selection criterion where instead of maximizing the ratio of true positive rate to false positive rate the model with lowest AICc has the fairest variance--bias tradeoff \citep{Burnham2002a}.
+Model selection was performed in this manner because the optimal number of PCs to use as predictors was not know \textit{a priori}, and while including all of the PCs of shape would mean that all shape variability would be used to estimate class, this may cause the model to be overfit and not provide an accurate estimate of unsampled plastral variation.
 
 %     recursive feature selection
-Because random forest models are not fit using maximum likelihood, a recursive feature selection algorithm was used to choose the optimal number of PCs to include based on the AUC ROC of the model.
-PCs were sequentially added as features until the AUC ROC of the model did not increase.
+Random forest models are not fit using maximum likelihood so AICc based model selection was not possible.
+Instead, a recursive feature selection algorithm was used to choose the optimal number of PCs to include based on the AUC of the model.
+PCs were sequentially added as features until the AUC of the model did not increase.
+Like the multinomial logistic regression models, \Sexpr{floor(max.ad)} was the maximum number of PCs that could have been included in the model.
+After each PC was added, 10-fold CV was used to estimate the optimal values of the tuning parameters as well as quantify the uncertainty of each model.
 Random forest model parameters were estimated from 1000 subtrees.
-After each PC was added, 10-fold CV was used to estimate the optimal values of the tuning parameters as well as quantify the uncertainty of each model.
-Like the multinomial logistic regression models, \Sexpr{floor(max.ad)} was the maximum number of PCs that could have been included in the model.
-The recursive feature selection algorithm used here is that implemented in the \texttt{caret} package for R \citep{Kuhn2013}.
+The recursive feature selection algorithm used in this study was that implemented in the \texttt{caret} package for R \citep{Kuhn2013}.
 
 % generalization
-%   bootstrap resample of AUC ROC
+%   bootstrap resample of AUC
 The final selected models were then used to estimate the class assignments of the training dataset.
-Model performance was measured using AUC ROC.
-A distribution of AUC ROC values were estimated for each classification scheme using \Sexpr{rr$sh1$R} nonparametric bootstrap resamples of the training dataset.
+Model generality for both methods for all four classification schemes was measured using the AUC of the assignments.
+A distribution of AUC values was estimated for each classification scheme via \Sexpr{rr$sh1$R} nonparametric bootstrap resamples of the training dataset.
 
@@
 \subsection{Geometric morphometrics}
+The results of the PCA of the total dataset of \textit{E.
+marmorata} pastral landmarks configurations demonstrates no clear or obvious groupings (Fig.
+\ref{fig:pca}).
+The first three PCs, which represent \Sexpr{paste0(sum(turtle.fit$percent[1:3]))}\% of the total variation, are a cloud of points.
+Additionally, individual landmark variation is mostly circular around each landmark with some more elliptical variation observed along some midline landmarks and the most lateral landmark.
+However, it is important to note that Procustes based superimposition attempts to evenly distribute landmark variance ZELDICH and this observation should be considered cursory and not rigorous.
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width = \textwidth]{figure/pca_res}
+  \caption{}
+  \label{fig:pca}
+\end{figure}
 
@@
 % PAM results
-Comparison of the gap statistic values for the different PAM solutions indicates that the optimal number of clusters is \Sexpr{which.max(tadult.gap$Tab[, 'gap'])} (Fig.
+Comparison of the gap statistic values for the range of PAM solutions indicates that the optimal number of clusters is \Sexpr{which.max(tadult.gap$Tab[, 'gap'])} (Fig.
 \ref{fig:gap}).
-The second best clustering solution is two clusters, however there is no geographic structure to this classification scheme SUPPLEMENT?.
+The next best clustering solution is two clusters, however there is no geographic structure to this classification scheme, with members of these clusters being seemingly randomly distributed geographically SUPPLEMENT?.
 Our dataset does not include enough or detailed information on the sex of each \textit{E.
 marmorata} specimen, thus it is not possible to determine if this clustering solution corresponds to sexual dimorphism between the observations.
-Male Emyidine turtles are known to have a plastral concavity
+Male Emyidine turtles are known to have a plastral concavity which may influence landmark position.
+However, the plastral concavity of \textit{E.
+mamorata} males is considered less prnounced than in other Emyidine turtles.
+While we cannot completely rule out sexual dimorphism as the root cause of this observation, we are less concerned with possible effects of sexual dimorphism for the later supervised learning methods.
 Increasing the number of clusters does appear to improve the gap statistic enough to merit comparison.
@@
 % multinomial logistic regression model
-The optimal multinomial logistic regression models for all four classification schemes, based on AICc, have all \Sexpr{length(tm.a.analysis$best$spinks$coefnames)} possible PCs as features(Table SUPPLEMENT).
-The second best models for all classification schemes have \Sexpr{length(tm.a.analysis$best$spinks$coefnames) - 1} PCs as features.
-The \(\delta\)AICc values between the optimal and second best model range from \Sexpr{min(daic)} for the first morphological based classification hypothesis to \Sexpr{max(daic)} for the second molecular based classification hypothesis.
+The AICc best multinomial logistic regression models, for all four classification schemes, each have the first \Sexpr{length(tm.a.analysis$best$spinks$coefnames)} possible PCs as features (Table SUPPLEMENT).
+The second best models for all classification schemes had the first \Sexpr{length(tm.a.analysis$best$spinks$coefnames) - 1} PCs as features.
+The \(\Delta\)AICc values between the optimal and second best model range from \Sexpr{min(daic)} for the first morphological based classification hypothesis to \Sexpr{max(daic)} for the second molecular based classification hypothesis (Table IN SUPPLEMENT?).
+The first \Sexpr{length(tm.a.analysis$best$spinks$coefnames)} PCs describe \Sexpr{paste0(sum(turtle.fit$percent[1:10]))}\% of total variation in plastral shape.
 
-While the \(\delta\)AICc value between the optimal and second best model for the first morphological based classification hypothesis is within the range to be considered sufficient and equally optimal \citep{Burnham2002a}, for this analysis we chose to use only the absolute AICc best model.
-While AICc values can not be compared between models with different responses \citep{Burnham2002a}, we interpret the fact that the optimal model for all classification schemes is the global model as a reason to just use this model structure for all cases.
-Additionally, by maintaining a single optimal model for all classification schemes, this limits the total number of comparisons between the bootstrap resampled distributions of the AUC ROC values for the testing data set (see below).
+While the \(\Delta\)AICc value between the optimal and second best model for the first morphological based classification hypothesis was within the range to be considered sufficient and equally optimal \citep{Burnham2002a}, for this analysis we chose to use only the AICc best model.
+While AICc values can not be compared between models with different responses \citep{Burnham2002a}, we interpret the fact that the optimal model for all classification schemes is the global model as a reason to use the AICc best model for all cases.
+Additionally, by using a single model for all classification schemes, this limits the number of comparisons between the bootstrap resampled distributions of the AUC values for the testing data set (see below).
 
@@
 
-The selected number of features in the optimal random forest models for each classification scheme vary much more than in the case of the multinomial logistic regression models (Fig.
+The selected number of features in the final random forest model for each classification scheme varied much more than in the case of the multinomial logistic regression models (Fig.
 \ref{fig:roc}), ranging from \Sexpr{min(oprf)} for the first morphological based classification hypothesis to \Sexpr{max(oprf)} for the second morphological based classification hypothesis.
@@
 
-In the case of all models, there is a substantial increase in model performance as measured by AICc for the multinomial logistic models (Tables SUPPLEMENT) or in AUC ROC for the random forest models and illustrated for the multinomial logistic regression models as the number of features increases (Fig.
+In the case of all models, there is a substantial increase in model performance as measured by AICc for the multinomial logistic models (Tables SUPPLEMENT) or in AUC for the random forest models and illustrated for the multinomial logistic regression models as the number of features increases (Fig.
 \ref{fig:roc}).
@@
   \caption{Effect of increasing the number of PCs as features, or predictors, of classification of plastra for all four classification schemes.
-As the number of PCs increase, AUC ROC increases until eventually leveling off.
-Both multinomial logisitic regression and random forest models are illustrated here, though AUC ROC based model selection was only performed for random forest models.}
+As the number of PCs increase, AUC increases until eventually leveling off.
+Both multinomial logisitic regression and random forest models are illustrated here, though AUC based model selection was only performed for random forest models.}
   \label{fig:roc}
@@
 %     facet: training, multinomial, rf
-Results of the bootstrap resamples of the AUC ROC of the generalization of the selected multinomial logistic regression and random forest models demonstrates that one of the molecular classification hypotheses based on \citet{Spinks2005} and \citet{Spinks2010} appears to be the best classification scheme (Fig.
+Results from the generalization of the selected supervised learning models, measured by the distributions of the bootstrapped AUC values of the testing dataset demonstrates that one of the molecular classification hypotheses based on \citet{Spinks2005} and \citet{Spinks2010} was the best overall classification scheme (Fig.
 \ref{fig:gen_res}).
-The distribution of bootstrapped AUC ROC for the molecular hypothesis is significantly different MANN-WHITNEY U TEST and greater than all of the other classification scheme.
-What is remarkable is that the best classification hypothesis is identical based on both the multinomial logistic regression and random forest models.
+For both methods, the distribution of bootstrapped AUC for the molecular hypothesis was significantly greater MANN-WHITNEY U TEST than all of the other classification scheme.
+Remarkably, the best classification hypothesis was identical based on both the multinomial logistic regression and random forest models.
 
@@
   \includegraphics[width = \textwidth]{figure/gen_res}
-  \caption{Density estimates of AUC ROC values of predictions of the testing dataset of plastra from \Sexpr{rr$sh1$R} bootstrap resamples.
+  \caption{Density estimates of AUC values of predictions of the testing dataset of plastra from \Sexpr{rr$sh1$R} bootstrap resamples.
 The top facet corresponds to values using the optimal multinomial logistic regression model, as chosen by minimum AICc value.
-The bottom facet corresponds to the values using the optimal random forest model, as chosen by maximum AUC ROC value.}
+The bottom facet corresponds to the values using the optimal random forest model, as chosen by maximum AUC value.}
   \label{fig:gen_res}
@@
 
-When the classification results of the training set for the optimal classification scheme are compared with the references classes, the higher AUC ROC value of the best multinomial logistic regression model becomes apparent as the classifications are in general much more similar to the reference (Fig.
+When the classification results of the training set for the optimal classification scheme are compared with the references classes, the higher AUC value of the best multinomial logistic regression model compared to the best random forest model can be observed as the classifications are much closer to the reference classes (Fig.
 \ref{fig:gen_map}).
-The best random forest model misclassified many of the observations as the northern clade.
-This trend is observable but not as exaggerated in the results of the classifications of the multinomial logistic regression model.
+The best random forest model misclassified many of the observations as the northern clade instead of the correct class.
+This systematic misclassification is observable but not as exaggerated in the results of the classifications of the multinomial logistic regression model.
 
@@
 The mean shape of the northern clade is the most similar to the mean shape of the entire dataset (Fig.
-\ref{fig:mean_shape}\ref{fig:imp_var}), which may mean that specimens that are closer to the mean shape will be systematically misclassified as the northern clade.
+\ref{fig:mean_shape}, \ref{fig:imp_var}), which may indicate that specimens that are closer to the mean shape will be systematically misclassified as the northern clade.
 
@@
   \end{subfigure}
-  \caption{<+caption text+>}
+  \caption{2x magnification}
   \label{fig:mean_shape}
@@
 % variation along the most important axes
-The results of training the random forest model also include the variable importance for best separating the different classes.
-Recursive feature selection of the best random forest model of the chosen classification scheme indicated that after \Sexpr{trf.a$spinks$optsize} PCs were included as features, AUC ROC would not increase.
-Of these \Sexpr{trf.a$spinks$optsize} features, three are illustrated here (Fig.
-\ref{fig:imp_pc}) the first two of which are most important SUPPLEMENT WITH VARIABLE IMPORTANCE INFORMATION?.
+The results of fitting the final random forest model also include the variable importance for best separating the different classes.
+The final random forest model of the best classification scheme indicated that after \Sexpr{trf.a$spinks$optsize} PCs were included as features, AUC would not increase.
+Of these \Sexpr{trf.a$spinks$optsize} features, the first three are illustrated here (Fig.
+\ref{fig:imp_pc}) in descending order of importance SUPPLEMENT WITH VARIABLE IMPORTANCE INFORMATION?.
 
@@
 
-The first two most important features, according to the random forest model, describe different aspects of variation (Fig.
+The first two most important features describe different aspects of variation (Fig.
 \ref{fig:imp_var}).
-The third PC, or first most important PC, describes variation in the relative position of landmarks on anterior and posterior portions of the plastron.
-The eighth PC, or second most important PC, mostly describes variation in landmarks along the midline of the plastron.
-The major variation along these axes correspond well to the differences between the class means and the mean plastron shape (Fig.
-\ref{fig:mean_shape}) where major class differences seems restricted to the relative ballooning or shrinking of the anterior and posterior portions of the plastron together.
+The third PC, or first most important PC, describes variation in the relative position of landmarks on anterior and posterior portions of the plastron and represents \Sexpr{turtle.fit$percent[3]}\% of total variation.
+The eighth PC, or second most important PC, mostly describes variation in landmarks along the midline of the plastron and represents \Sexpr{turtle.fit$percent[8]}\% of total variation.
+The major variations along these axes correspond well to the differences between the class means and the mean plastron shape (Fig.
+\ref{fig:mean_shape}) where major class differences seem based on the relative ballooning or shrinking of the anterior and posterior portions of the plastron together.
 
@@
 This is contrary to the original classification of \textit{E.
-marmorata} \citep{Seeliger1945} and lends credence to the idea that at least some aspect of cryptic diversity is a product either sample size, methodology, or both.
+marmorata} \citep{Seeliger1945} and lends credence to the idea that at least some aspect of cryptic diversity is a product of sample size, methodology, or both.
+
 %   what does this mean about cryptic diversity?
@@
 
-The lack of coherent geographical subclass assignment from PAM based clustering (Fig.
-\ref{fig:gap}) as well as the large number of features necessary before plateau in AUC ROC for all models (Fig.
-\ref{fig:roc}) can be taken as indicators that the variation between subclasses is extremely fine grained.
-This is also exemplified by the differences in mean class shape for the final chosen classification scheme (Fig.
+The lack of coherent geographical subclass assignment from PAM clustering (Fig.
+\ref{fig:gap}) as well as the large number of features necessary before no increase in AUC for all models (Fig.
+\ref{fig:roc}) indicates that the morphological variation between subclasses is extremely fine grained.
+This was also exemplified by the small differences in mean class shape for the final chosen classification scheme (Fig.
 \ref{fig:imp_var}).
@@
 %     fair because as much variation as ``necessary'' is used
-The methods presented here for supervised learning analysis of the landmark variation represent a compromise
+The methods presented here for supervised learning analysis of the landmark variation represent a compromise between explicitly modeling all shape variation and preventing models from being overfit and ungeneralizable.
+While all aspects of shape may be evolving simultaneously, and not along individual PCs, including all PCs as predictors in each model might increase model complexity beyond a reasonable level for the sample size or the necessary complexity to accurately model the response.
+However, this compromise is not without its advantages.
+Because the AICc and AUC values improved rapidly with increased model complexity (Fig.
+\ref{fig:roc}),k this indicated how fine scale the actual variation between classes was and how necessary including many PCs was.
+Additionally, the relative risk values from the mulitinomial logistic regression models demonstrate MORE STUFF HERE.
+Also, the systematic misclassification of the final random forest model and the possible relationship with the similarity the mean shape of the northern clade and the overall mean shape (Fig.
+\ref{fig:gen_map}, \ref{fig:mean_shape}) is a good indicator of the very fine scale variation between classes.
 
@@
 %     anything about semisupervised clustering? probably not
-Ultimately, it would be optimal to not require such explicit classification hypotheses, especially when concerned about possible cryptic variation in extinct taxa.
-The method employed in this study, PAM, is rather simple and not model based.
-Comparison is facilitated by comparison of a summary statistic and bootstrap confidence intervals.
-A more useful and future avenue would be employing various model based clustering approaches CITATIONS.
-In this manner, a series of candidate models can be compared via model comparison methods, such as AIC or Bayes factors CITATION, in order to asses the best clustering solution.
-Of particular note are nonparametric Bayesian approaches to model based clustering CITATIONS.
+Ultimately, it would be most useful to not require such explicit classification hypotheses, especially when concerned about possible cryptic variation in extinct taxa.
+The only unsupervised method employed in this study, PAM, is rather simple and not model based.
+Comparison here was facilitated via a summary statistic and bootstrapped confidence intervals.
+A more useful approach would be employing various model based clustering approaches CITATIONS.
+In this manner, a series of candidate models can be compared via model comparison methods, such as AIC or Bayes factors CITATION, in order to asses the best clustering solution.% Of particular note are nonparametric Bayesian approaches to model based clustering CITATIONS.
 This approach uses a class of flexible priors to allow for the most optimal clustering solution to be decided from the data.
-Currently, there exists a nonparametric Bayesian clustering method and further development of this approach may prove extremely fruitful for better deliminating taxa solely from morphology.
+Currently, there exists a nonparametric Bayesian clustering method and further development of this approach may prove extremely fruitful for better delimiting taxa solely from morphology.
 
 % future directions
+%   potentially useful for phylogenetic comparative method stuff
 %   conservation utility
@@
 % closing statements
-In this study we have demonstrated that given a large sample size and appropriate methodology, it is possible to determine which of multiple hypothesized classification schemes best explain variation in a taxon.
-The plastral variation of \textit{E.
+In this study we have demonstrated that given a large sample size and alternative methodology, it is possible to determine which classification scheme best explains variation in a taxon amongst a set of alternative hypotheses.
+The observed plastral variation of \textit{E.
 marmorata} is most consistent with the mitochondrial based hypothesis of \citet{Spinks2005} and \citet{Spinks2010} and not with the original morphology based hypothesis of \citet{Seeliger1945}.
 We have also demonstrated the utility of various machine learning approaches to understanding variation in morphometric data.
-Specifically, better understanding odds misclassification and identifying which is the most important for deliminating different classes.
+Specifically, better understanding odds misclassification and identifying which is the most important for delimiting different classes.
+These methods represent new applications which may be important for future studies interested in class-based morphological comparison and variation, both in the context of cryptic diversity and with known classifications.
 

### 2013-07-22 documents/turtle_paper.tex
first major round of edits on paper. probably need about 2 more.

@@
 \subsection{Specimens}
-We collected morphometric data from 524 specimens.
+We collected landmark-based morphometric data from 524 adult \textit{E.
+marmorata} museum specimens.
 Geographic information was recorded from museum collection information.
-When precise latitude and longitude information was not known for a specimen, it was inferred from whatever locality information was presented.
+When precise latitude and longitude information was not known for a specimen, it was inferred from whatever locality information was present.
 % museum information?
 
-Specimens were given a class assignment was based on geographic information.
-Because the exact geographic barriers between different class is unknown and fuzzy, two assignments for both morphological and molecular hypotheses of class were used.
+Specimens were classified based on museum record geographic information.
+The specimens used to define the subclades in \citet{Spinks2005} and \citet{Spinks2010} were not available for study, all classifications were based solely on the geographic information for each specimen and not from explicit assignment in previous studies.
+Because the exact geographic barriers between different classes are unknown and unclear, two assignments for both the morphological and molecular hypotheses were used.
+Each morphology-based hypothesis had three classes, while each molecular-based had four classes.
+In total, each specimen was given four different classifications.
 
@@
 %   Riemannian shape distance
-Because shape space, or configurations after Procrustes superimposition, is a Riemannian manifold \citep{Dryden1998a} the dissimilarity between each landmark configuration was measured as the Riemmanian shape distance or \(\rho\) \citep{Kendall1984a,Dryden1998a} which should vary between 0 and \(\pi / 2\) assuming no reflection invariance.
+In order to preserve the relationship in shape space between all landmark configurations, the dissimilarity between observations was measured using the Riemanninan shape distance or \(\rho\) \citep{Kendall1984a,Dryden1998a}.
+This metric was chosen because shape space, or the set of all possible shape configurations following Procrustes superimposition, is a Riemannian manifold and thus non-Euclidean \citep{Dryden1998a}.
+\(\rho\) varies between 0 and \(\pi / 2\) when there is no reflection invariance, which should not be a concern in the case of the half plastral landmark configurations.
 
 %   PAM
-The dissimilarity matrix of shape was divisivly clustering using partioning around mediods (PAM) which is analogous to \textit{k}-means clustering except that instead of minimizing the sum of squared Euclidean distances between observations and centroids, the sum of squared dissimilarities between observations and mediods is minimized \citep{Kaufman1990}.
+The \(\rho\) dissimilarity matrix was divisively clustering using partitioning around mediods (PAM), a method which is similar to \textit{k}-means clustering except that instead of minimizing the sum of squared Euclidean distances between observations and centroids, the sum of squared dissimilarities between observations and mediods is minimized \citep{Kaufman1990}.
 %   gap statistic
-The optimal number of clusters of shape configurations is unknown being possibly three, four, or some other value.
-Clustering solutions were estimated for between 1 and 40 clusters.
+Because the optimal number of clusters of shape configurations in the study was unknown, being possibly three, four, or some other value, clustering solutions were estimated with the number of mediods varying between one and 40.
 Clustering solutions were compared using the gap statistic, which is a measure of goodness of clustering \citep{Tibshirani2001a}.
-Standard errors of the gap statistic for each clustering solution were estimated from 500 bootstrap samples.
+The gap statistic is defined
+\[Gap_{n}(k) = E^{*}_{n}(\log(W_{k}) - \log(W_{k})\] 
+where \(W_{k}\) is
+\[W_{k} = \sum^{k}_{r = 1}{\frac{1}{2n_{r}} (\sum_{i,i' \in C_{r}} d_{ii'})}\]
+the dispersion of the clustering solution or the sum of the pairwise dissimilarities (\(d_{ii'}\)) between observations in each cluster and there respective mediods (\(C\)) for all clusters \(r\).
+\(E^{*}_{n}\) is the expectation of dispersion from a sample \(n\) of a reference distribution.
+In this case, the reference distribution was estimated from a 500 resamples of the dataset taking into account the original structure of the data.
 %  R implementation
-PAM clustering and gap statistic calculation was conducted using the \texttt{cluster} package for R \citep{Maechler2013}.
+This analysis was conducted using the \texttt{cluster} package for R \citep{Maechler2013} using all 524 plastral landmark configurations.
 
@@
 % training testing split
-The dataset of 524 plastron landmarks was split into training and testing datasets.
-The former was used for model fitting (training) and was 75\% of the total dataset, split proportionally per class, while the testing dataset was used to estimate the effectiveness of each classification scheme (i.e.
+The total dataset of 524 observations was split into training and testing datasets.
+The training dataset represented 75\% of the total dataset, split proportionally per class, and was used for model fitting.
+The testing dataset represented the remaining 25\% of the total dataset and  was used to estimate the effectiveness of each classification hypothesis and generalizability of the supervised learning models (i.e.
 performance in the wild).
@@
 %     AICc
-For the multinomial logistic regression models, PCs were added sequentially in order to increase the overall amount of variation in shape included in each model and the final model was that with the lowest AICc \citep{Burnham2002a} AKAIKE AND OTHER CITATION.
-This procedure was used because the optimal number of PCs to include is unknown, and while including all of the PCs of shape would mean that all of the variability in plastron shape would be used to estimate class, this may cause the model to be over fit and not provide an accurate estimate of unsampled plastral variation.
+For the multinomial logistic regression models, 10 different models were fit with each having sequentially more PCs as predictors in order to have models representing different levels of overall amount of shape variation and estimate how much variation was necessary and sufficient to best estimate class.
 The maximum number of PCs allowed to be used as predictors was 10 because of both the number of parameters estimated per model and the necessary sample size needed to estimate that many parameters accurately.
+The final model was that with the lowest AICc \citep{Burnham2002a} AKAIKE AND OTHER CITATION.
+Similar to AUC discussed above, AICc is a model selection criterion where instead of maximizing the ratio of true positive rate to false positive rate the model with lowest AICc has the fairest variance--bias tradeoff \citep{Burnham2002a}.
+Model selection was performed in this manner because the optimal number of PCs to use as predictors was not know \textit{a priori}, and while including all of the PCs of shape would mean that all shape variability would be used to estimate class, this may cause the model to be overfit and not provide an accurate estimate of unsampled plastral variation.
 
 %     recursive feature selection
-Because random forest models are not fit using maximum likelihood, a recursive feature selection algorithm was used to choose the optimal number of PCs to include based on the AUC ROC of the model.
-PCs were sequentially added as features until the AUC ROC of the model did not increase.
+Random forest models are not fit using maximum likelihood so AICc based model selection was not possible.
+Instead, a recursive feature selection algorithm was used to choose the optimal number of PCs to include based on the AUC of the model.
+PCs were sequentially added as features until the AUC of the model did not increase.
+Like the multinomial logistic regression models, 10 was the maximum number of PCs that could have been included in the model.
+After each PC was added, 10-fold CV was used to estimate the optimal values of the tuning parameters as well as quantify the uncertainty of each model.
 Random forest model parameters were estimated from 1000 subtrees.
-After each PC was added, 10-fold CV was used to estimate the optimal values of the tuning parameters as well as quantify the uncertainty of each model.
-Like the multinomial logistic regression models, 10 was the maximum number of PCs that could have been included in the model.
-The recursive feature selection algorithm used here is that implemented in the \texttt{caret} package for R \citep{Kuhn2013}.
+The recursive feature selection algorithm used in this study was that implemented in the \texttt{caret} package for R \citep{Kuhn2013}.
 
 % generalization
-%   bootstrap resample of AUC ROC
+%   bootstrap resample of AUC
 The final selected models were then used to estimate the class assignments of the training dataset.
-Model performance was measured using AUC ROC.
-A distribution of AUC ROC values were estimated for each classification scheme using 1000 nonparametric bootstrap resamples of the training dataset.
+Model generality for both methods for all four classification schemes was measured using the AUC of the assignments.
+A distribution of AUC values was estimated for each classification scheme via 1000 nonparametric bootstrap resamples of the training dataset.
 
@@
 \subsection{Geometric morphometrics}
+The results of the PCA of the total dataset of \textit{E.
+marmorata} pastral landmarks configurations demonstrates no clear or obvious groupings (Fig.
+\ref{fig:pca}).
+The first three PCs, which represent 55.0075790871272\% of the total variation, are a cloud of points.
+Additionally, individual landmark variation is mostly circular around each landmark with some more elliptical variation observed along some midline landmarks and the most lateral landmark.
+However, it is important to note that Procustes based superimposition attempts to evenly distribute landmark variance ZELDICH and this observation should be considered cursory and not rigorous.
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width = \textwidth]{figure/pca_res}
+  \caption{}
+  \label{fig:pca}
+\end{figure}
 
@@
 % PAM results
-Comparison of the gap statistic values for the different PAM solutions indicates that the optimal number of clusters is 1 (Fig.
+Comparison of the gap statistic values for the range of PAM solutions indicates that the optimal number of clusters is 1 (Fig.
 \ref{fig:gap}).
-The second best clustering solution is two clusters, however there is no geographic structure to this classification scheme SUPPLEMENT?.
+The next best clustering solution is two clusters, however there is no geographic structure to this classification scheme, with members of these clusters being seemingly randomly distributed geographically SUPPLEMENT?.
 Our dataset does not include enough or detailed information on the sex of each \textit{E.
 marmorata} specimen, thus it is not possible to determine if this clustering solution corresponds to sexual dimorphism between the observations.
-Male Emyidine turtles are known to have a plastral concavity
+Male Emyidine turtles are known to have a plastral concavity which may influence landmark position.
+However, the plastral concavity of \textit{E.
+mamorata} males is considered less prnounced than in other Emyidine turtles.
+While we cannot completely rule out sexual dimorphism as the root cause of this observation, we are less concerned with possible effects of sexual dimorphism for the later supervised learning methods.
 Increasing the number of clusters does appear to improve the gap statistic enough to merit comparison.
@@
 % multinomial logistic regression model
-The optimal multinomial logistic regression models for all four classification schemes, based on AICc, have all 10 possible PCs as features(Table SUPPLEMENT).
-The second best models for all classification schemes have 9 PCs as features.
-The \(\delta\)AICc values between the optimal and second best model range from 2.0639 for the first morphological based classification hypothesis to 19.8349 for the second molecular based classification hypothesis.
+The AICc best multinomial logistic regression models, for all four classification schemes, each have the first 10 possible PCs as features (Table SUPPLEMENT).
+The second best models for all classification schemes had the first 9 PCs as features.
+The \(\Delta\)AICc values between the optimal and second best model range from 2.0639 for the first morphological based classification hypothesis to 19.8349 for the second molecular based classification hypothesis (Table IN SUPPLEMENT?).
+The first 10 PCs describe 88.6043874205075\% of total variation in plastral shape.
 
-While the \(\delta\)AICc value between the optimal and second best model for the first morphological based classification hypothesis is within the range to be considered sufficient and equally optimal \citep{Burnham2002a}, for this analysis we chose to use only the absolute AICc best model.
-While AICc values can not be compared between models with different responses \citep{Burnham2002a}, we interpret the fact that the optimal model for all classification schemes is the global model as a reason to just use this model structure for all cases.
-Additionally, by maintaining a single optimal model for all classification schemes, this limits the total number of comparisons between the bootstrap resampled distributions of the AUC ROC values for the testing data set (see below).
+While the \(\Delta\)AICc value between the optimal and second best model for the first morphological based classification hypothesis was within the range to be considered sufficient and equally optimal \citep{Burnham2002a}, for this analysis we chose to use only the AICc best model.
+While AICc values can not be compared between models with different responses \citep{Burnham2002a}, we interpret the fact that the optimal model for all classification schemes is the global model as a reason to use the AICc best model for all cases.
+Additionally, by using a single model for all classification schemes, this limits the number of comparisons between the bootstrap resampled distributions of the AUC values for the testing data set (see below).
 
@@
 
-The selected number of features in the optimal random forest models for each classification scheme vary much more than in the case of the multinomial logistic regression models (Fig.
+The selected number of features in the final random forest model for each classification scheme varied much more than in the case of the multinomial logistic regression models (Fig.
 \ref{fig:roc}), ranging from 6 for the first morphological based classification hypothesis to 10 for the second morphological based classification hypothesis.
@@
   \includegraphics[width = \textwidth]{figure/gen_res}
-  \caption{Density estimates of AUC ROC values of predictions of the testing dataset of plastra from 1000 bootstrap resamples.
+  \caption{Density estimates of AUC values of predictions of the testing dataset of plastra from 1000 bootstrap resamples.
 The top facet corresponds to values using the optimal multinomial logistic regression model, as chosen by minimum AICc value.
-The bottom facet corresponds to the values using the optimal random forest model, as chosen by maximum AUC ROC value.}
+The bottom facet corresponds to the values using the optimal random forest model, as chosen by maximum AUC value.}
   \label{fig:gen_res}
@@
 % variation along the most important axes
-The results of training the random forest model also include the variable importance for best separating the different classes.
-Recursive feature selection of the best random forest model of the chosen classification scheme indicated that after 7 PCs were included as features, AUC ROC would not increase.
-Of these 7 features, three are illustrated here (Fig.
-\ref{fig:imp_pc}) the first two of which are most important SUPPLEMENT WITH VARIABLE IMPORTANCE INFORMATION?.
+The results of fitting the final random forest model also include the variable importance for best separating the different classes.
+The final random forest model of the best classification scheme indicated that after 7 PCs were included as features, AUC would not increase.
+Of these 7 features, the first three are illustrated here (Fig.
+\ref{fig:imp_pc}) in descending order of importance SUPPLEMENT WITH VARIABLE IMPORTANCE INFORMATION?.
 
@@
 
-The first two most important features, according to the random forest model, describe different aspects of variation (Fig.
+The first two most important features describe different aspects of variation (Fig.
 \ref{fig:imp_var}).
-The third PC, or first most important PC, describes variation in the relative position of landmarks on anterior and posterior portions of the plastron.
-The eighth PC, or second most important PC, mostly describes variation in landmarks along the midline of the plastron.
-The major variation along these axes correspond well to the differences between the class means and the mean plastron shape (Fig.
-\ref{fig:mean_shape}) where major class differences seems restricted to the relative ballooning or shrinking of the anterior and posterior portions of the plastron together.
+The third PC, or first most important PC, describes variation in the relative position of landmarks on anterior and posterior portions of the plastron and represents 9.6697\% of total variation.
+The eighth PC, or second most important PC, mostly describes variation in landmarks along the midline of the plastron and represents 4.1061\% of total variation.
+The major variations along these axes correspond well to the differences between the class means and the mean plastron shape (Fig.
+\ref{fig:mean_shape}) where major class differences seem based on the relative ballooning or shrinking of the anterior and posterior portions of the plastron together.
 
@@
 %     anything about semisupervised clustering? probably not
-Ultimately, it would be optimal to not require such explicit classification hypotheses, especially when concerned about possible cryptic variation in extinct taxa.
-The method employed in this study, PAM, is rather simple and not model based.
-Comparison is facilitated by comparison of a summary statistic and bootstrap confidence intervals.
-A more useful and future avenue would be employing various model based clustering approaches CITATIONS.
-In this manner, a series of candidate models can be compared via model comparison methods, such as AIC or Bayes factors CITATION, in order to asses the best clustering solution.
-Of particular note are nonparametric Bayesian approaches to model based clustering CITATIONS.
+Ultimately, it would be most useful to not require such explicit classification hypotheses, especially when concerned about possible cryptic variation in extinct taxa.
+The only unsupervised method employed in this study, PAM, is rather simple and not model based.
+Comparison here was facilitated via a summary statistic and bootstrapped confidence intervals.
+A more useful approach would be employing various model based clustering approaches CITATIONS.
+In this manner, a series of candidate models can be compared via model comparison methods, such as AIC or Bayes factors CITATION, in order to asses the best clustering solution.% Of particular note are nonparametric Bayesian approaches to model based clustering CITATIONS.
 This approach uses a class of flexible priors to allow for the most optimal clustering solution to be decided from the data.
-Currently, there exists a nonparametric Bayesian clustering method and further development of this approach may prove extremely fruitful for better deliminating taxa solely from morphology.
+Currently, there exists a nonparametric Bayesian clustering method and further development of this approach may prove extremely fruitful for better delimiting taxa solely from morphology.
 

### 2013-07-28 documents/turtle_paper.Rnw
need tables. then final edits.

@@
 source('../src/turtle_plots.r')
+
+options(digits = 4)
 @
 
-\title{How cryptic is cryptic diversity? Machine learning approaches to plastral variation in \textit{Emys marmorata}.}
+\title{How cryptic is cryptic diversity? Machine learning approaches to classifying morphological variation.}
 \author[1]{Peter D Smits}%\thanks{psmits@uchicago.edu}}
@@
 \begin{abstract}
-  Write me please.
-  % 200 words
-\noindent (Keywords: Testudines, morphology, geometric morphometrics)
+  %\textbf{Write me please.}
+  Cryptic diversity is the phenomenon where some taxa are believed to be identifiable solely from molecular data.
+The reality of this phenomenon is concerning because the majority of extant taxa are delimited solely based on morphology.
+Additionally, virtually all extant taxa are known entirely from morphology and thus can only be delimited based on morphology.
+Here we address questions about whether it is possible to determine entirely based on morphology if one classification hypotheses can be considered better than others.
+  %
+  By using a combination of unsupervised and supervised learning methods, each with different useful properties, we demonstrate a suite of approaches for better understanding differences in morphological variation between classes, the odds of classifying an observation as one class relative to another, and what aspects of morphology best describe the variation between different classes.
+  %
+  These approaches are applied to the classification of the Emyidine turtle, \textit{Emys marmorata}.
+This species has conflicting hypotheses of the number of meaningful clades based on either morphological or molecular information.
+By characterizing the variation in plastral shape and how it may be identifiably different between different classes, we compared multiple different explicit classification hypotheses.
+By splitting a large dataset of specimens into both training and testing datasets, we were also able to determine which of the classification hypotheses best corresponded to the observed plastral variation.
+  %
+  The results from our analysis shows that the seemingly best classification of \textit{Emys marmorata} is in accordance with the molecularly based hypothesis.
+
+\noindent (Keywords: Testudines, morphology, geometric morphometrics, random forests)
 \end{abstract}
@@
 %   most species are still deliminated solely based on morphology
-%Cryptic diversity is when taxa were only first deliminated via molecular means and were not or cannot deliminated via morphological identification CITATION.
-The discovery of this previously unknown diversity has expanded diversity amoungst certain taxa CITATION.
-Additionally, a suite of methods have been developed to more accurately deliminate taxa based on molecular divergence CITATION.
+Cryptic diversity is the phenomenon that not all taxa can be recognized based on morphology and can only be delimited using molecular information CITATIONS.
+Conceringly, most extant taxa, and nearly all extinct taxa, are delimited based solely via morphological analysis.
+This phenomenon is of great concern when styding variation and diversity dynamics over long periods of time, where apparent morphological stasis may not actually reflect true diversity CITATIONS.
+In the case of endangered or otherwise conserved taxa, morphometric approaches for classifying and identifying taxa and populations of conservation importance would greatly improve the ability to maintain these high risk groups.
+Additionally, this would allow for better classifying extinct taxa.
 
-Cryptic diversity is the phenomenon that not all taxa can be recognized based on morphology and can only be delimited using molecular information.
-The most extant taxa, and nearly all extinct taxa, are delimited based solely via morphological analysis.
-Concerns about how pervasive this phenomenon is and much diversity is actually cryptic have
-
-Much work has been devoted to species delimitation via sequence difference CITATIONS while comparatively little has been devoted for case of morphological data.
-The majority of this effort has focused on automated taxon identification CITATIONS or identifying morphological differences amongst already identified taxa CITATIONS.
+Much work has been devoted to species delimitation via sequence difference CITATIONS while comparatively little has been devoted for case of purely morphological data.
+The majority of this effort has focused on identifying differences between already identified taxa CITATIONS and automated taxon identification CITATIONS.
 % all the work done on species delimitation strictly from molecular data
 
-%   paleo problem
-Concerns about cryptic diversity are enhanced when taxa are only known from morphology, such as the case of extinct organisms CITATION.
-
-Here, we address two questions  about cryptic diversity: how much of cryptic diversity is a product of sample size, and how can methodology impact classification based solely on morphology.
-Additionally, we ask if fine scale variation in morphology can provide corroboration for various classification hypotheses, and if it is possible to determine which is the best classification hypothesis.
+Here, we address the question of how can alternative approaches and methodology improve morphology based classification.
+From this approach, we ask if it is possible to determine which amongst a set of classification hypotheses is best.
 
@@
 % why use machine learning methods
+\subsection{Background and system}
 Differences in morphological variation between different classes has previously been analyzed using methods like linear discriminate analysis and canonical variates analysis CITATION ZELDICH.
-Neural network models have also been introduced and applied in the context of automated taxon identification along with more general applications MACLEOD BOOK.
+These methods are comparatively simple and straight forward ways of understanding the differences in morphology between classes.
+Also, they are very visual methods which aides in interpretation and presentation of information.
+Neural network models have also been introduced and applied frequently in the context of automated taxon identification along with more general applications MACLEOD BOOK.
 Here, we used multiple alternative machine learning methods, both unsupervised and supervised, in order to compare different classification hypotheses.
@@
 Unsupervised learning methods are analogous to clustering and density estimation methods, while supervised learning methods are analogous classification and regression models.
-In both cases, these methods may not be fit via maximum likelihood but may be supplemented by randomization algorithms and the maximization or minimization of summary statistics in order to best estimate a general model for all data, but sampled and unsampled.
-The expansion of methods used to understand morphological differentiation
+In both cases, many of these methods may not be fit via maximum likelihood but may be supplemented by randomization, sorting, and partitioning algorithms along with the maximization or minimization of summary statistics in order to best estimate a general model for all data, both sampled and unsampled.
+The application of the alternative approaches used in this study illustrates only a sampling of the various previously derived methods for clustering observations and fitting classification models.
+Additionally, instead of pure classification accuracy, here we use a statistic of classification strength that reflects the rate at which taxa are both accurately and inaccurately classified (see Methods).
 
@@
 %     conservation importance (turtles in general, e. marmorata in specific)
-In this study, we address the subspecific classification scheme of \textit{Emys marmorata}, or western pond turtle.
+In this study, we investiage the subspecific classification of the western pond turtle, \textit{Emys marmorata}.
 \textit{E.
-marmorata} has a distribution from northern Washington State, USA to Baja California, Mexico.
+marmorata} is distributed from northern Washington State, USA to Baja California, Mexico.
 %   morphological hypothesis of subspecies
@@
 Traditionally, \textit{E.
-marmorata} was classified into three subgroups: the northern \textit{E.
+marmorata} was classified into three groups: the northern \textit{E.
 marmorata marmorata}, the southern \textit{E.
@@
 
-Previous work on morphological differentiation between subspecies of \textit{E. marmorata} focused on just the known subspecies of \citet{Seeliger1945}
+Previous work on morphological differentiation between subspecies of \textit{E.
+marmorata} focused on just the known subspecies of \citet{Seeliger1945}.
+SEE MASTERS THESIS AND INCLUDE INFORMATION ON THE LINEAR MORPHOMETRIC CLASSIFICATION WORK DONE PRIOR ON E MARMORATA.
 
 More recently, \textit{E.
-marmorata} was divided into four clades based on mitochrondial DNA: a northern clade, a southern clade, and two central Californian clades \citep{Spinks2005,Spinks2010}.
-Nuclear DNA supports two major clades, one northern and one southern, however \citet{Spinks2010} argue that the four clade classification is of greater conservation utility to use the mitochondrial classification scheme.
-There is no known morphological differentiation between these clades.
+marmorata} was divided into four clades based on mitochondrial DNA: a northern clade, a southern clade, and two central Californian clades \citep{Spinks2005,Spinks2010}.
+While nuclear DNA supports two major clades, one northern and one southern, \citet{Spinks2010} argue that the four clade classification is of greater conservation utility to use the mitochondrially based classification scheme.
+While the mitochondrially based classification is considered robust, there is no known morphological differentiation between these clades.
 
@@
 In this study, we attempt to estimate the best classification scheme of \textit{E.
-marmorata} subspecies based on morphological variation in plastral shape.
+marmorata} based on variation in plastral shape.
 Because of unclear geographic boundaries between subgroups of \textit{E.
-marmorata}, we compare two hypotheses of morphology-based classification and two hypotheses of molecular-based classification.
-We hypotheize that if morphological variation corresponds to subspecific assignment, then it should be possible to differentiate the best classification hypothesis of \textit{E.
+marmorata}, we compare two hypotheses of morphologically based classification and two hypotheses of molecularly based classification.
+We hypothesize that if morphological variation corresponds to class assignment, then it should be possible to determine the best classification hypothesis of \textit{E.
 marmorata} from amongst multiple candidate hypotheses.
-However, if morphological variation variation does not correspond to any classification hypothesis, then supervised learning model generalization performance will be extremely poor and reflect how variation does not follow along with classification.
+However, if morphological variation variation does not correspond to any classification hypothesis, then supervised learning model generalization performance will be poor and reflect how variation may not follow along with any of the candidate classification hypotheses.
 
@@
 marmorata} museum specimens.
+Specimen classification was based on known specimen geographic information.
 Geographic information was recorded from museum collection information.
-When precise latitude and longitude information was not known for a specimen, it was inferred from whatever locality information was present.
-% museum information?
-
-Specimens were classified based on museum record geographic information.
-The specimens used to define the subclades in \citet{Spinks2005} and \citet{Spinks2010} were not available for study, all classifications were based solely on the geographic information for each specimen and not from explicit assignment in previous studies.
-Because the exact geographic barriers between different classes are unknown and unclear, two assignments for both the morphological and molecular hypotheses were used.
-Each morphology-based hypothesis had three classes, while each molecular-based had four classes.
+When precise latitude and longitude information was not known it was estimated from whatever locality information was present.
+Because the specimens used to define the subclades in \citet{Spinks2005} and \citet{Spinks2010} were not available for study, all specimen classifications were based solely on this geographic information and not from explicit assignment in previous studies.
+Because the exact barriers between different biogeographic regions are unknown and unclear, two assignments for both the morphologically and molecularly based hypotheses were used.
+Each morphologically based hypothesis had three classes, while each molecular-based had four classes.
 In total, each specimen was given four different classifications.
@@
 Following \citet{Angielczyk2011}, 19 landmarks were digitized using TpsDig 2.04 \citep{Rohlf2005}.
+These landmarks were chosen to maximize the description of general plastral variation.
 17 of these landmarks are at the endpoints or intersection of the keratinous plastral scutes that cover the platron.
-These landmarks were chosen to maximize the description of general plastral variation.
-12 of these landmarks were chosen to be symmetrical across the axis of symmetry and in order to prevent degrees of freedom and other concerns \citep{Klingenberg2007} prior to analysis these landmarks were reflected across the axis of symmetry (i.e.
+12 of these landmarks were chosen to be symmetrical across the axis of symmetry and, in order to prevent degrees of freedom and other concerns \citep{Klingenberg2007}, prior to analysis these landmarks were reflected across the axis of symmetry (i.e.
 midline) and the average position of each symmetrical pair was used.
 In cases where damage or incompleteness prevented symmetric landmarks from being determined, only the single member of the pair was used.
-Analysis was then conducted on the resulting ``half'' plastra.
-
+Analysis was conducted on the resulting ``half'' plastra.
 % GP superimposition and PCA
-``Half'' plastra landmark configurations were superimposed using generalized Procrustes analysis \citep{Dryden1998a} after which, the principal components (PC) of shape were calculated.
+Plastral landmark configurations were superimposed using generalized Procrustes analysis \citep{Dryden1998a} after which, the principal components (PC) of shape were calculated.
 This was done using the \texttt{shapes} package for R \citep{2013, Dryden2013}.
@@
 %   Riemannian shape distance
-In order to preserve the relationship in shape space between all landmark configurations, the dissimilarity between observations was measured using the Riemanninan shape distance or \(\rho\) \citep{Kendall1984a,Dryden1998a}.
+In order to preserve the relationship between all landmark configurations in shape space, the dissimilarity between observations was measured using Kendall's Riemanninan shape distance or \(\rho\) \citep{Kendall1984a,Dryden1998a}.
 This metric was chosen because shape space, or the set of all possible shape configurations following Procrustes superimposition, is a Riemannian manifold and thus non-Euclidean \citep{Dryden1998a}.
-\(\rho\) varies between 0 and \(\pi / 2\) when there is no reflection invariance, which should not be a concern in the case of the half plastral landmark configurations.
+\(\rho\) varies between 0 and \(\pi / 2\) when there is no reflection invariance, which should not be a concern in the case of the half plastral landmark configurations used in the study.
 
 %   PAM
-The \(\rho\) dissimilarity matrix was divisively clustering using partitioning around mediods (PAM), a method which is similar to \textit{k}-means clustering except that instead of minimizing the sum of squared Euclidean distances between observations and centroids, the sum of squared dissimilarities between observations and mediods is minimized \citep{Kaufman1990}.
+The \(\rho\) dissimilarity matrix was divisively clustered using partitioning around mediods clustering (PAM), a method similar to \textit{k}-means clustering except that instead of minimizing the sum of squared Euclidean distances between observations and centroids, the sum of squared dissimilarities between observations and mediods is minimized \citep{Kaufman1990}.
 %   gap statistic
-Because the optimal number of clusters of shape configurations in the study was unknown, being possibly three, four, or some other value, clustering solutions were estimated with the number of mediods varying between one and \Sexpr{nrow(tadult.gap$Tab)}.
+Because the optimal number of clusters of shape configurations in the study was unknown, being possibly three, four, or some other value, clustering solutions were estimated with the number of clusters varied between one and \Sexpr{nrow(tadult.gap$Tab)}.
 Clustering solutions were compared using the gap statistic, which is a measure of goodness of clustering \citep{Tibshirani2001a}.
@@
 \(E^{*}_{n}\) is the expectation of dispersion from a sample \(n\) of a reference distribution.
-In this case, the reference distribution was estimated from a \Sexpr{tadult.gap$B} resamples of the dataset taking into account the original structure of the data.
+In this case, the reference distribution was estimated from \Sexpr{tadult.gap$B} resamples of the dataset taking into account the original structure of the data.
 %  R implementation
@@
 The total dataset of \Sexpr{nrow(turtle.adult)} observations was split into training and testing datasets.
-The training dataset represented 75\% of the total dataset, split proportionally per class, and was used for model fitting.
-The testing dataset represented the remaining 25\% of the total dataset and  was used to estimate the effectiveness of each classification hypothesis and generalizability of the supervised learning models (i.e.
+The training dataset represented 75\% of the total dataset, split proportionally by class, and was used for model fitting.
+The testing dataset represented the remaining 25\% of the total dataset and  was used after model fitting to estimate the effectiveness of each classification hypothesis and generalizability of the supervised learning models (i.e.
 performance in the wild).
@@
 % model types
-Two different supervised learning methods and model types were used to model the relationship between plastral shape and class: multinomial logistic regression and random forest.
-These models were chosen because of various properties of these models which allow for useful interpretations about the quality and structure of the classification.
+Two different supervised learning methods were used to model the relationship between plastral shape and class: multinomial logistic regression and random forest.
+These methods were chosen because of various properties of these methods which allow for useful interpretations about the quality and structure of the classification.
 
@@
 Multinomial logistic regression is an extension of logistic regression, where instead of a binary response there are three or more response classes CITATION.
-Effectively, this type of model can be viewed as multiple, simultaneous logistic regression models for each class and the final classification of the observation being the most probable of all the constituent model classifications.
-From the final model, the relative risk of a possible classification, with reference to a baseline class, can be calculated from the coefficients of the model.
-These are similar to the odds ratios calculated from the coefficients of a logistic regression.
+Effectively, this type of model can be viewed as multiple, simultaneous logistic regression models for each class and the final classification of the observation being the most probable of all the constituent model results.
+Similar to the odds ratios calculated from the coefficients of a logistic regression, the relative risk of a possible classification with reference to a baseline class can be determined from the coefficients of the model.
 Multinomial logistic regression models were fit using the \texttt{nnet} package for R \citep{Venables2002}
@@
 Random forest models are an extension of classification and regression trees (CART) CITATION.
+Because this study relies on classification models, CARTs are explained with reference to classification but the approach is equally valid for regression.
+The goal of CARTs are to use a series of different features to estimate the final class.
+In top-down induction of decision trees for each member of a given set of predictor variables, attribute value test are used to estimate the differences between classes.
+This process is then repeated on each subset, called recursive partitioning.
+The recursion continues until the resulting observations all share the same class or no more meaningful partitions are possible.
+
 In a random forest model, many CARTs are built from a random subsample of both the features and the observations.
-This process was repeated 1000 times and the parameters of the final model was chosen as the mode of estimates from the distribution of CARTs CITATION.
+This process is then repeated many times and the parameters of the final model was chosen as the mode of estimates from the distribution of CARTs CITATION.
 In addition to fitting a classification model, this procedure allows for the features to be ranked in order of importance.
 In the context of this study, this means that the PCs most important for describing the difference between classes can be estimated, and thus illustrate the most important variation amongst classes as opposed to just the greatest amount of variation in the entire dataset.
-This is a generally important property that should be useful for other morphometric studies which want to describe and model the shape differences between different classes and the relative importance of various aspects of variation.
+This is a generally important property that is useful for many other studies which want to describe and model the differences between classes and the relative importance different features.
 Random forest models were fit using the \texttt{randomForest} package for R \citep{Liaw2002}.
@@
 For both types of supervised learning methods, tuning parameters were estimated using 10 rounds of 10-fold cross-validation (CV) across a grid search of all tuning parameter combinations.
-Optimal tuning parameter values were selected based on area under the receiver operating characteristic curve (ROC).
+Optimal tuning parameter values were selected based on area under the receiver operating characteristic (ROC) curve.
 The area under the multiclass ROC curves was estimated using the all-against one strategy derived by \citet{Hand2001}.
+This tuning process was implemented following the default grid search implemented in the \texttt{caret} package for R \citep{Kuhn2013}.
 
 % what is ROC? what is AUC?
-ROC is a type of confusion matrix statistic that is a descriptor the relationship between the false positive rate or 1 - specificity of a classification model and the true positive rate or sensitivity of a classification model CITATION.
+
+ROC is a confusion matrix (Table \ref{tab:conf}) statistic that is a descriptor the relationship between the false positive rate (\(FPR\), Eq.
+\ref{eq:fpr}) of a classification model and the true positive rate (\(TPR\), Eq.
+\ref{eq:tpr}) of a classification model CITATION.
 The area under the ROC curve (AUC) is a summary statistic of the quality of the classification and varies between 0.5 and 1, with 0.5 indicating a model that classifies no better than random and a value of 1 indicating perfect classification CITATION.
 AUC can be used as a model selection criterion for classification models and is especially useful in cases where some if not all of the models in question were not fit via maximum likelihood where a criterion such as AICc (see below) or similar can be used CITATION.
+\begin{table}
+  \centering
+  \begin{tabular}[c]{ p{2cm} c | p{2cm} | p{2cm} |}
+    \cline{3-4} 
+    & & \multicolumn{2}{ c |}{Predicted class} \\ \cline{3-4}
+    & & 1 & 0 \\ \hline
+    \multicolumn{1}{| c |}{\multirow{2}{*}{Actual class}}
+    & 1 & TRUE \newline POSITIVE & FALSE \newline NEGATIVE \\ \cline{2-4}
+    \multicolumn{1}{| c |}{} & 0 & FALSE \newline POSITIVE & TRUE \newline NEGATIVE \\
+    \hline
+  \end{tabular}
+  \caption{Example confusion matrix.
+The columns correspond to the predicted class of an observation, while the rows correspond to the actual class of that observation.
+Depending on the type match between the prediction and reality, four different outcomes are possible: true positive (TP), false negative (FN), false positive (FP), and true negative (TN).
+These four quantities are used for calculating all confusion matrix statistics.}
+  \label{tab:conf}
+\end{table}
 
+\begin{equation}
+  FPR = \frac{FP}{FP + TN}
+  \label{eq:fpr}
+\end{equation}
+\begin{equation}
+  TPR = \frac{TP}{TP + FN}
+  \label{eq:tpr}
+\end{equation}
 %   model selection
 %     AICc
-For the multinomial logistic regression models, \Sexpr{floor(max.ad)} different models were fit with each having sequentially more PCs as predictors in order to have models representing different levels of overall amount of shape variation and estimate how much variation was necessary and sufficient to best estimate class.
-The maximum number of PCs allowed to be used as predictors was \Sexpr{floor(max.ad)} because of both the number of parameters estimated per model and the necessary sample size needed to estimate that many parameters accurately.
+For the multinomial logistic regression models, \Sexpr{floor(max.ad)} different models were fit each having sequentially more PCs as predictors in order to have models representing different levels of overall amount of shape variation and to estimate how much was necessary and sufficient to best estimate class.
+The maximum number of PCs allowed as predictors was \Sexpr{floor(max.ad)} because of both the large number of parameters estimated per model and the necessary sample size needed to estimate that many parameters accurately.
 The final model was that with the lowest AICc \citep{Burnham2002a} AKAIKE AND OTHER CITATION.
-Similar to AUC discussed above, AICc is a model selection criterion where instead of maximizing the ratio of true positive rate to false positive rate the model with lowest AICc has the fairest variance--bias tradeoff \citep{Burnham2002a}.
+AICc is a model selection criterion where the model with lowest AICc has the fairest variance--bias tradeoff \citep{Burnham2002a}.
 Model selection was performed in this manner because the optimal number of PCs to use as predictors was not know \textit{a priori}, and while including all of the PCs of shape would mean that all shape variability would be used to estimate class, this may cause the model to be overfit and not provide an accurate estimate of unsampled plastral variation.
@@
 Instead, a recursive feature selection algorithm was used to choose the optimal number of PCs to include based on the AUC of the model.
-PCs were sequentially added as features until the AUC of the model did not increase.
-Like the multinomial logistic regression models, \Sexpr{floor(max.ad)} was the maximum number of PCs that could have been included in the model.
+PCs were sequentially added as features and the selected model was that with the highest AUC.
+Similar to the multinomial logistic regression models described above, \Sexpr{floor(max.ad)} was the maximum number of PCs that could have been included in the model.
 After each PC was added, 10-fold CV was used to estimate the optimal values of the tuning parameters as well as quantify the uncertainty of each model.
@@
 \ref{fig:pca}).
-The first three PCs, which represent \Sexpr{paste0(sum(turtle.fit$percent[1:3]))}\% of the total variation, are a cloud of points.
-Additionally, individual landmark variation is mostly circular around each landmark with some more elliptical variation observed along some midline landmarks and the most lateral landmark.
-However, it is important to note that Procustes based superimposition attempts to evenly distribute landmark variance ZELDICH and this observation should be considered cursory and not rigorous.
+The first three PCs, which represent \Sexpr{paste0(sum(turtle.fit$percent[1:3]))}\% of the total variation, are a cloud of points with no structure.
+Additionally, individual landmark variation is mostly circular around each landmark with some more elliptical variation observed along some midline landmarks and the most lateral landmark (Fig.
+\ref{fig:pca}).
+However, it is important to note that Procustes based superimposition attempts to evenly distribute variance around the mean landmark ZELDICH and this observation should be considered cursory at best.
 
@@
   \includegraphics[width = \textwidth]{figure/pca_res}
-  \caption{}
+  \caption{Results from PCA of the Procrustes superimposed ``half'' plastral landmarks.
+Depicted here are the for three PCs (lower triangle) and the mean shape with observed variance around each point (upper right).
+The first three PCs account for total \Sexpr{paste0(sum(turtle.fit$perfect[1:3]))} of the variance in plastral shape.}
   \label{fig:pca}
@@
 % PAM results
-Comparison of the gap statistic values for the range of PAM solutions indicates that the optimal number of clusters is \Sexpr{which.max(tadult.gap$Tab[, 'gap'])} (Fig.
+Comparison of gap statistic values for the range of PAM solutions indicates that the optimal number of clusters is one (Fig.
 \ref{fig:gap}).
-The next best clustering solution is two clusters, however there is no geographic structure to this classification scheme, with members of these clusters being seemingly randomly distributed geographically SUPPLEMENT?.
-Our dataset does not include enough or detailed information on the sex of each \textit{E.
-marmorata} specimen, thus it is not possible to determine if this clustering solution corresponds to sexual dimorphism between the observations.
-Male Emyidine turtles are known to have a plastral concavity which may influence landmark position.
+The next best clustering solution had only two clusters, however there is no geographic structure to this classification scheme, with members of these clusters being seemingly randomly distributed SUPPLEMENT?.
+Unfortunately, it was not possible to obtain enough detailed information on the sex of each \textit{E.
+marmorata} specimen, thus it is difficult at best to determine if this clustering solution corresponds to sexual dimorphism between the observations.
+Male Emyidine turtles are known to have a plastral concavity which may influence landmark position along the midline.
 However, the plastral concavity of \textit{E.
-mamorata} males is considered less prnounced than in other Emyidine turtles.
-While we cannot completely rule out sexual dimorphism as the root cause of this observation, we are less concerned with possible effects of sexual dimorphism for the later supervised learning methods.
-Increasing the number of clusters does appear to improve the gap statistic enough to merit comparison.
+mamorata} males is considered less pronounced than in other Emyidine turtles.
+While we cannot completely rule out sexual dimorphism as the root cause of this observation, because of the very small degree of known sexual dimorphism in \textit{E.
+marmorata} we are less concerned with possible effects of sexual dimorphism for the later supervised learning methods.
+The gap statistic values for both three and four clusters are much lower than for one and two and are statistically identical.
+Interestingly, other solutions with a much greater number of clusters have higher gap statistic values though these are also not significantly different.
+Increasing the number of clusters does appear to improve the gap statistic enough compared to the best clustering solution to merit detailed discussion.
 
@@
 % multinomial logistic regression model
-The AICc best multinomial logistic regression models, for all four classification schemes, each have the first \Sexpr{length(tm.a.analysis$best$spinks$coefnames)} possible PCs as features (Table SUPPLEMENT).
+The AICc best multinomial logistic regression models, for all four classification schemes, all include the first \Sexpr{length(tm.a.analysis$best$spinks$coefnames)} possible PCs as features (Table SUPPLEMENT).
 The second best models for all classification schemes had the first \Sexpr{length(tm.a.analysis$best$spinks$coefnames) - 1} PCs as features.
@@
 
-While the \(\Delta\)AICc value between the optimal and second best model for the first morphological based classification hypothesis was within the range to be considered sufficient and equally optimal \citep{Burnham2002a}, for this analysis we chose to use only the AICc best model.
-While AICc values can not be compared between models with different responses \citep{Burnham2002a}, we interpret the fact that the optimal model for all classification schemes is the global model as a reason to use the AICc best model for all cases.
-Additionally, by using a single model for all classification schemes, this limits the number of comparisons between the bootstrap resampled distributions of the AUC values for the testing data set (see below).
+While the \(\Delta\)AICc value between the optimal and second best model for the first morphological based classification hypothesis was within the range to be considered equally optimal \citep{Burnham2002a}, for this analysis we chose to use only the AICc best model.
+While AICc values can not be compared between models with different responses \citep{Burnham2002a}, we interpret the fact that the optimal model for all classification schemes is the global model as a reason to use only the AICc best model for all cases.
+Additionally, by using a single model for all classification schemes, this limits the number of comparisons between the bootstrap resampled distributions of the AUC values for the testing dataset (see below).
 
@@
 
-The selected number of features in the final random forest model for each classification scheme varied much more than in the case of the multinomial logistic regression models (Fig.
-\ref{fig:roc}), ranging from \Sexpr{min(oprf)} for the first morphological based classification hypothesis to \Sexpr{max(oprf)} for the second morphological based classification hypothesis.
+The selected number of features in the final random forest model for each classification scheme varied much more than for the multinomial logistic regression models, ranging from \Sexpr{min(oprf)} for the first morphological based classification hypothesis to \Sexpr{max(oprf)} for the second morphological based classification hypothesis (Fig.
+\ref{fig:roc}).
 
@@
 %     facet: training, multinomial, rf
-Results from the generalization of the selected supervised learning models, measured by the distributions of the bootstrapped AUC values of the testing dataset demonstrates that one of the molecular classification hypotheses based on \citet{Spinks2005} and \citet{Spinks2010} was the best overall classification scheme (Fig.
+The results from the generalization of the selected supervised learning models, measured by the distributions of the bootstrapped AUC values of the testing dataset, demonstrate that one of the molecular classification hypotheses was the best overall classification scheme (Fig.
 \ref{fig:gen_res}).
-For both methods, the distribution of bootstrapped AUC for the molecular hypothesis was significantly greater MANN-WHITNEY U TEST than all of the other classification scheme.
+For both methods, the distribution of bootstrapped AUC for the molecular hypothesis was significantly greater MANN-WHITNEY U TEST than all of the other classification schemes.
 Remarkably, the best classification hypothesis was identical based on both the multinomial logistic regression and random forest models.
@@
 
-When the classification results of the training set for the optimal classification scheme are compared with the references classes, the higher AUC value of the best multinomial logistic regression model compared to the best random forest model can be observed as the classifications are much closer to the reference classes (Fig.
+When the classification results of the training set for the best classification scheme based on the generalization results are compared with the references classes, the higher AUC value of the best multinomial logistic regression model compared to the best random forest model can be observed as the classifications are much closer to the reference classes (Fig.
 \ref{fig:gen_map}).
 The best random forest model misclassified many of the observations as the northern clade instead of the correct class.
-This systematic misclassification is observable but not as exaggerated in the results of the classifications of the multinomial logistic regression model.
+This pattern of misclassification is observable but not as exaggerated in the classifications of the multinomial logistic regression model.
 
@@
 
-This pattern of misclassification may be caused by the differences in mean shape between each of the different classes (Fig.
+This pattern of misclassification may have been caused by the subtle differences in mean shape between each of the different classes (Fig.
 \ref{fig:mean_shape}).
 The mean shape of the northern clade is the most similar to the mean shape of the entire dataset (Fig.
-\ref{fig:mean_shape}, \ref{fig:imp_var}), which may indicate that specimens that are closer to the mean shape will be systematically misclassified as the northern clade.
+\ref{fig:mean_shape1}), which may indicate that specimens that are closer to the mean shape will be systematically misclassified as the northern clade.
 
@@
     \includegraphics[width = \textwidth]{figure/mshape_1}
+    \caption{Northern}
     \label{fig:mean_shape1}
@@
     \includegraphics[width = \textwidth]{figure/mshape_2}
+    \caption{Eastern}
     \label{fig:mean_shape2}
@@
     \includegraphics[width = \textwidth]{figure/mshape_3}
+    \caption{Western}
     \label{fig:mean_shape3}
@@
     \includegraphics[width = \textwidth]{figure/mshape_4}
+    \caption{Southern}
     \label{fig:mean_shape4}
   \end{subfigure}
-  \caption{2x magnification}
+  \caption{Thin-plate splines for each of the four classes from the best classification hypothesis based on the generalization results (Fig.
+\ref{fig:gen_res}).
+The four different classes are labeled according to the biogeographic groups as depicted in figure \ref{fig:gen_map}.
+The deformations are depicted with 2x magnification from base.}
   \label{fig:mean_shape}
@@
 The results of fitting the final random forest model also include the variable importance for best separating the different classes.
-The final random forest model of the best classification scheme indicated that after \Sexpr{trf.a$spinks$optsize} PCs were included as features, AUC would not increase.
+The selected random forest model for the best classification scheme had \Sexpr{trf.a$spinks$optsize} PCs as features, AUC would not increase.
 Of these \Sexpr{trf.a$spinks$optsize} features, the first three are illustrated here (Fig.
@@
 The variables descend in importance from the upper left to the lower right.
-The observations are colored as in Figures \ref{fig:gen_res} and \ref{fig:gen_map}.
+The observations are colored as in figure \ref{fig:gen_map}.
 The bottom row are histograms of classification occurrences along the PCs.}
@@
 \ref{fig:imp_var}).
-The third PC, or first most important PC, describes variation in the relative position of landmarks on anterior and posterior portions of the plastron and represents \Sexpr{turtle.fit$percent[3]}\% of total variation.
-The eighth PC, or second most important PC, mostly describes variation in landmarks along the midline of the plastron and represents \Sexpr{turtle.fit$percent[8]}\% of total variation.
-The major variations along these axes correspond well to the differences between the class means and the mean plastron shape (Fig.
+The third and most important PC describes variation in the relative position of landmarks on anterior and posterior portions of the plastron and represents \Sexpr{turtle.fit$percent[3]}\% of total variation.
+The eighth and second most important PC mostly describes variation in landmarks along the midline of the plastron and represents \Sexpr{turtle.fit$percent[8]}\% of total variation.
+The major variations along these axes correspond well to the differences between the mean shape of each class (Fig.
 \ref{fig:mean_shape}) where major class differences seem based on the relative ballooning or shrinking of the anterior and posterior portions of the plastron together.
@@
 
+%RELATIVE RISK FROM THE MULTINOMIAL LOGISTIC REGRESSION MODELS
+The relative risk values for classification from the multinomial logistic regression model based on the three most important PCs demonstrates that while an individual PC might not be sufficient to differentiate any individual class from the northern group, that given multiple features together the odds of determining the correct classification increase (Fig.
+\ref{fig:rel_risk}).
+Changes along the second most important axes contribute very little increasing the odds of classification for any of the classes in relation to the northern group.
+This is observable from the class histograms along PC 8 (Fig.
+\ref{fig:imp_var}).
+Changes along the first and third most important axes contribute more obviously to increasing the odds of correctly identifying the class of an observation, a result that is observable in both the relative risk (Fig.
+\ref{fig:rel_risk}) and histograms of the different classes from the PCs (Fig.
+\ref{fig:imp_var}).
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width = \textwidth]{figure/rel_risk}
+  \caption{Forest plot of the relative risk, with 95\% confidence intervals, of classifying a give specimen based on the first three most important variables according to the random forest model.
+Relative risk values are calculated from the coefficients of the multinomial logistic regression model.
+All risks are relative to the northern group from \citet{Spinks2005,Spinks2010}.
+Variable importance is from left to right.}
+  \label{fig:rel_risk}
+\end{figure}
+
 \section{Discussion}
@@
 %   remarkable concordance in the supervised learning methods
-The results of this study provide support for the mitochrondial based hypothesis of classification of \textit{E.
+The results of this study support the mitochrondial based hypothesis of classification of \textit{E.
 marmorata} \citep{Spinks2005,Spinks2010}.
 This is contrary to the original classification of \textit{E.
-marmorata} \citep{Seeliger1945} and lends credence to the idea that at least some aspect of cryptic diversity is a product of sample size, methodology, or both.
+marmorata} \citep{Seeliger1945} MASTERS THESIS and lends credence to the idea that at least some aspect of cryptic diversity is a product of sample size, methodology, or both.
 
@@
 \ref{fig:gap}) as well as the large number of features necessary before no increase in AUC for all models (Fig.
-\ref{fig:roc}) indicates that the morphological variation between subclasses is extremely fine grained.
+\ref{fig:roc}) indicates that the morphological variation between classes is extremely fine grained.
 This was also exemplified by the small differences in mean class shape for the final chosen classification scheme (Fig.
@@
 %     fair because as much variation as ``necessary'' is used
-The methods presented here for supervised learning analysis of the landmark variation represent a compromise between explicitly modeling all shape variation and preventing models from being overfit and ungeneralizable.
-While all aspects of shape may be evolving simultaneously, and not along individual PCs, including all PCs as predictors in each model might increase model complexity beyond a reasonable level for the sample size or the necessary complexity to accurately model the response.
+The approaches presented here for supervised learning analysis of the landmark variation represent a compromise between explicitly modeling all shape variation and preventing models from being overfit and ungeneralizable.
+While all aspects of shape may be evolving simultaneously, and not along individual PCs, including all shape variation in each model might increase model complexity beyond a reasonable level for the sample size and possibly the necessary complexity to accurately model the response.
+Additionally, because individual PCs are used as features in the models this does not completely and accurately represent shape evolution and how exactly different classes might be evolving in relation to each other.
 However, this compromise is not without its advantages.
-Because the AICc and AUC values improved rapidly with increased model complexity (Fig.
-\ref{fig:roc}),k this indicated how fine scale the actual variation between classes was and how necessary including many PCs was.
-Additionally, the relative risk values from the mulitinomial logistic regression models demonstrate MORE STUFF HERE.
-Also, the systematic misclassification of the final random forest model and the possible relationship with the similarity the mean shape of the northern clade and the overall mean shape (Fig.
-\ref{fig:gen_map}, \ref{fig:mean_shape}) is a good indicator of the very fine scale variation between classes.
+Because both AICc and AUC values improved rapidly with increased model complexity (Fig.
+\ref{fig:roc}), this indicated how fine scale the actual variation between classes was.
+Additionally, the relative risk values from the mulitinomial logistic regression models demonstrate that a single PC is probably not sufficient for estimating the class of an observation, but that given a set of PCs this classification would be more accurate (Fig.
+\ref{fig:rel_risk}).
+The results of the relative risk values also shows that while a variable can be considered important for the random forest models (Fig.
+\ref{fig:imp_pc}, \ref{fig:imp_var}).
 
@@
 %     anything about semisupervised clustering? probably not
-Ultimately, it would be most useful to not require such explicit classification hypotheses, especially when concerned about possible cryptic variation in extinct taxa.
+Ultimately, it would be useful to not require such explicit classification hypotheses, especially when concerned about possible cryptic variation in extinct taxa.
 The only unsupervised method employed in this study, PAM, is rather simple and not model based.
-Comparison here was facilitated via a summary statistic and bootstrapped confidence intervals.
 A more useful approach would be employing various model based clustering approaches CITATIONS.
@@
 % closing statements
-In this study we have demonstrated that given a large sample size and alternative methodology, it is possible to determine which classification scheme best explains variation in a taxon amongst a set of alternative hypotheses.
+In this study we have demonstrated that using alternative methodology to that which is most frequently applied, it is possible to determine which classification scheme best explains variation in a taxon amongst a set of alternative hypotheses.
 The observed plastral variation of \textit{E.
 marmorata} is most consistent with the mitochondrial based hypothesis of \citet{Spinks2005} and \citet{Spinks2010} and not with the original morphology based hypothesis of \citet{Seeliger1945}.
-We have also demonstrated the utility of various machine learning approaches to understanding variation in morphometric data.
-Specifically, better understanding odds misclassification and identifying which is the most important for delimiting different classes.
-These methods represent new applications which may be important for future studies interested in class-based morphological comparison and variation, both in the context of cryptic diversity and with known classifications.
+We have also demonstrated the utility of various machine learning approaches to understanding the structure of variation in morphometric data.
+Specifically, methods for better understanding misclassification and identifying which is the most important for delimiting different classes.
+These methods represent new applications which may be important for future studies on class-based morphological comparison and variation, both in the context of cryptic diversity and with known classifications.
 
 \section{Acknowledgements}
-PDS would like to thank David Bapst, Michael Foote, Benjamin Frable, and Dallas Krentzel for useful discussion which enhanced the quality of this study.
+PDS would like to thank David Bapst, Michael Foote, Benjamin Frable, and Dallas Krentzel for useful discussion which enhanced the quality of this study. MUSEUMS. GRANTING AGENCIES.
 

### 2013-07-28 documents/turtle_paper.tex
need tables. then final edits.

@@
 
-\title{How cryptic is cryptic diversity? Machine learning approaches to plastral variation in \textit{Emys marmorata}.}
+\title{How cryptic is cryptic diversity? Machine learning approaches to classifying morphological variation.}
 \author[1]{Peter D Smits}%\thanks{psmits@uchicago.edu}}
@@
 %   Riemannian shape distance
-In order to preserve the relationship in shape space between all landmark configurations, the dissimilarity between observations was measured using the Riemanninan shape distance or \(\rho\) \citep{Kendall1984a,Dryden1998a}.
+In order to preserve the relationship between all landmark configurations in shape space, the dissimilarity between observations was measured using Kendall's Riemanninan shape distance or \(\rho\) \citep{Kendall1984a,Dryden1998a}.
 This metric was chosen because shape space, or the set of all possible shape configurations following Procrustes superimposition, is a Riemannian manifold and thus non-Euclidean \citep{Dryden1998a}.
-\(\rho\) varies between 0 and \(\pi / 2\) when there is no reflection invariance, which should not be a concern in the case of the half plastral landmark configurations.
+\(\rho\) varies between 0 and \(\pi / 2\) when there is no reflection invariance, which should not be a concern in the case of the half plastral landmark configurations used in the study.
 
 %   PAM
-The \(\rho\) dissimilarity matrix was divisively clustering using partitioning around mediods (PAM), a method which is similar to \textit{k}-means clustering except that instead of minimizing the sum of squared Euclidean distances between observations and centroids, the sum of squared dissimilarities between observations and mediods is minimized \citep{Kaufman1990}.
+The \(\rho\) dissimilarity matrix was divisively clustered using partitioning around mediods clustering (PAM), a method similar to \textit{k}-means clustering except that instead of minimizing the sum of squared Euclidean distances between observations and centroids, the sum of squared dissimilarities between observations and mediods is minimized \citep{Kaufman1990}.
 %   gap statistic
-Because the optimal number of clusters of shape configurations in the study was unknown, being possibly three, four, or some other value, clustering solutions were estimated with the number of mediods varying between one and 40.
+Because the optimal number of clusters of shape configurations in the study was unknown, being possibly three, four, or some other value, clustering solutions were estimated with the number of clusters varied between one and 40.
 Clustering solutions were compared using the gap statistic, which is a measure of goodness of clustering \citep{Tibshirani2001a}.
@@
 \(E^{*}_{n}\) is the expectation of dispersion from a sample \(n\) of a reference distribution.
-In this case, the reference distribution was estimated from a 500 resamples of the dataset taking into account the original structure of the data.
+In this case, the reference distribution was estimated from 500 resamples of the dataset taking into account the original structure of the data.
 %  R implementation
@@
 The total dataset of 524 observations was split into training and testing datasets.
-The training dataset represented 75\% of the total dataset, split proportionally per class, and was used for model fitting.
-The testing dataset represented the remaining 25\% of the total dataset and  was used to estimate the effectiveness of each classification hypothesis and generalizability of the supervised learning models (i.e.
+The training dataset represented 75\% of the total dataset, split proportionally by class, and was used for model fitting.
+The testing dataset represented the remaining 25\% of the total dataset and  was used after model fitting to estimate the effectiveness of each classification hypothesis and generalizability of the supervised learning models (i.e.
 performance in the wild).
@@
 For both types of supervised learning methods, tuning parameters were estimated using 10 rounds of 10-fold cross-validation (CV) across a grid search of all tuning parameter combinations.
-Optimal tuning parameter values were selected based on area under the receiver operating characteristic curve (ROC).
+Optimal tuning parameter values were selected based on area under the receiver operating characteristic (ROC) curve.
 The area under the multiclass ROC curves was estimated using the all-against one strategy derived by \citet{Hand2001}.
+This tuning process was implemented following the default grid search implemented in the \texttt{caret} package for R \citep{Kuhn2013}.
 
 % what is ROC? what is AUC?
-ROC is a type of confusion matrix statistic that is a descriptor the relationship between the false positive rate or 1 - specificity of a classification model and the true positive rate or sensitivity of a classification model CITATION.
+
+ROC is a confusion matrix (Table \ref{tab:conf}) statistic that is a descriptor the relationship between the false positive rate (\(FPR\), Eq.
+\ref{eq:fpr}) of a classification model and the true positive rate (\(TPR\), Eq.
+\ref{eq:tpr}) of a classification model CITATION.
 The area under the ROC curve (AUC) is a summary statistic of the quality of the classification and varies between 0.5 and 1, with 0.5 indicating a model that classifies no better than random and a value of 1 indicating perfect classification CITATION.
 AUC can be used as a model selection criterion for classification models and is especially useful in cases where some if not all of the models in question were not fit via maximum likelihood where a criterion such as AICc (see below) or similar can be used CITATION.
+\begin{table}
+  \centering
+  \begin{tabular}[c]{ p{2cm} c | p{2cm} | p{2cm} |}
+    \cline{3-4} 
+    & & \multicolumn{2}{ c |}{Predicted class} \\ \cline{3-4}
+    & & 1 & 0 \\ \hline
+    \multicolumn{1}{| c |}{\multirow{2}{*}{Actual class}}
+    & 1 & TRUE \newline POSITIVE & FALSE \newline NEGATIVE \\ \cline{2-4}
+    \multicolumn{1}{| c |}{} & 0 & FALSE \newline POSITIVE & TRUE \newline NEGATIVE \\
+    \hline
+  \end{tabular}
+  \caption{Example confusion matrix.
+The columns correspond to the predicted class of an observation, while the rows correspond to the actual class of that observation.
+Depending on the type match between the prediction and reality, four different outcomes are possible: true positive (TP), false negative (FN), false positive (FP), and true negative (TN).
+These four quantities are used for calculating all confusion matrix statistics.}
+  \label{tab:conf}
+\end{table}
 
+\begin{equation}
+  FPR = \frac{FP}{FP + TN}
+  \label{eq:fpr}
+\end{equation}
+\begin{equation}
+  TPR = \frac{TP}{TP + FN}
+  \label{eq:tpr}
+\end{equation}
 %   model selection
 %     AICc
-For the multinomial logistic regression models, 10 different models were fit with each having sequentially more PCs as predictors in order to have models representing different levels of overall amount of shape variation and estimate how much variation was necessary and sufficient to best estimate class.
-The maximum number of PCs allowed to be used as predictors was 10 because of both the number of parameters estimated per model and the necessary sample size needed to estimate that many parameters accurately.
+For the multinomial logistic regression models, 10 different models were fit each having sequentially more PCs as predictors in order to have models representing different levels of overall amount of shape variation and to estimate how much was necessary and sufficient to best estimate class.
+The maximum number of PCs allowed as predictors was 10 because of both the large number of parameters estimated per model and the necessary sample size needed to estimate that many parameters accurately.
 The final model was that with the lowest AICc \citep{Burnham2002a} AKAIKE AND OTHER CITATION.
-Similar to AUC discussed above, AICc is a model selection criterion where instead of maximizing the ratio of true positive rate to false positive rate the model with lowest AICc has the fairest variance--bias tradeoff \citep{Burnham2002a}.
+AICc is a model selection criterion where the model with lowest AICc has the fairest variance--bias tradeoff \citep{Burnham2002a}.
 Model selection was performed in this manner because the optimal number of PCs to use as predictors was not know \textit{a priori}, and while including all of the PCs of shape would mean that all shape variability would be used to estimate class, this may cause the model to be overfit and not provide an accurate estimate of unsampled plastral variation.
@@
 Instead, a recursive feature selection algorithm was used to choose the optimal number of PCs to include based on the AUC of the model.
-PCs were sequentially added as features until the AUC of the model did not increase.
-Like the multinomial logistic regression models, 10 was the maximum number of PCs that could have been included in the model.
+PCs were sequentially added as features and the selected model was that with the highest AUC.
+Similar to the multinomial logistic regression models described above, 10 was the maximum number of PCs that could have been included in the model.
 After each PC was added, 10-fold CV was used to estimate the optimal values of the tuning parameters as well as quantify the uncertainty of each model.
@@
 \ref{fig:pca}).
-The first three PCs, which represent 55.0075790871272\% of the total variation, are a cloud of points.
-Additionally, individual landmark variation is mostly circular around each landmark with some more elliptical variation observed along some midline landmarks and the most lateral landmark.
-However, it is important to note that Procustes based superimposition attempts to evenly distribute landmark variance ZELDICH and this observation should be considered cursory and not rigorous.
+The first three PCs, which represent 55.0075790871272\% of the total variation, are a cloud of points with no structure.
+Additionally, individual landmark variation is mostly circular around each landmark with some more elliptical variation observed along some midline landmarks and the most lateral landmark (Fig.
+\ref{fig:pca}).
+However, it is important to note that Procustes based superimposition attempts to evenly distribute variance around the mean landmark ZELDICH and this observation should be considered cursory at best.
 
@@
   \includegraphics[width = \textwidth]{figure/pca_res}
-  \caption{}
+  \caption{Results from PCA of the Procrustes superimposed ``half'' plastral landmarks.
+Depicted here are the for three PCs (lower triangle) and the mean shape with observed variance around each point (upper right).
+The first three PCs account for total 0 of the variance in plastral shape.}
   \label{fig:pca}
@@
 % PAM results
-Comparison of the gap statistic values for the range of PAM solutions indicates that the optimal number of clusters is 1 (Fig.
+Comparison of gap statistic values for the range of PAM solutions indicates that the optimal number of clusters is one (Fig.
 \ref{fig:gap}).
-The next best clustering solution is two clusters, however there is no geographic structure to this classification scheme, with members of these clusters being seemingly randomly distributed geographically SUPPLEMENT?.
-Our dataset does not include enough or detailed information on the sex of each \textit{E.
-marmorata} specimen, thus it is not possible to determine if this clustering solution corresponds to sexual dimorphism between the observations.
-Male Emyidine turtles are known to have a plastral concavity which may influence landmark position.
+The next best clustering solution had only two clusters, however there is no geographic structure to this classification scheme, with members of these clusters being seemingly randomly distributed SUPPLEMENT?.
+Unfortunately, it was not possible to obtain enough detailed information on the sex of each \textit{E.
+marmorata} specimen, thus it is difficult at best to determine if this clustering solution corresponds to sexual dimorphism between the observations.
+Male Emyidine turtles are known to have a plastral concavity which may influence landmark position along the midline.
 However, the plastral concavity of \textit{E.
-mamorata} males is considered less prnounced than in other Emyidine turtles.
-While we cannot completely rule out sexual dimorphism as the root cause of this observation, we are less concerned with possible effects of sexual dimorphism for the later supervised learning methods.
-Increasing the number of clusters does appear to improve the gap statistic enough to merit comparison.
+mamorata} males is considered less pronounced than in other Emyidine turtles.
+While we cannot completely rule out sexual dimorphism as the root cause of this observation, because of the very small degree of known sexual dimorphism in \textit{E.
+marmorata} we are less concerned with possible effects of sexual dimorphism for the later supervised learning methods.
+The gap statistic values for both three and four clusters are much lower than for one and two and are statistically identical.
+Interestingly, other solutions with a much greater number of clusters have higher gap statistic values though these are also not significantly different.
+Increasing the number of clusters does appear to improve the gap statistic enough compared to the best clustering solution to merit detailed discussion.
 
@@
 % multinomial logistic regression model
-The AICc best multinomial logistic regression models, for all four classification schemes, each have the first 10 possible PCs as features (Table SUPPLEMENT).
+The AICc best multinomial logistic regression models, for all four classification schemes, all include the first 10 possible PCs as features (Table SUPPLEMENT).
 The second best models for all classification schemes had the first 9 PCs as features.
@@
 
-The selected number of features in the final random forest model for each classification scheme varied much more than in the case of the multinomial logistic regression models (Fig.
-\ref{fig:roc}), ranging from 6 for the first morphological based classification hypothesis to 10 for the second morphological based classification hypothesis.
+The selected number of features in the final random forest model for each classification scheme varied much more than for the multinomial logistic regression models, ranging from 6 for the first morphological based classification hypothesis to 10 for the second morphological based classification hypothesis (Fig.
+\ref{fig:roc}).
 
@@
 The results of fitting the final random forest model also include the variable importance for best separating the different classes.
-The final random forest model of the best classification scheme indicated that after 7 PCs were included as features, AUC would not increase.
+The selected random forest model for the best classification scheme had 7 PCs as features, AUC would not increase.
 Of these 7 features, the first three are illustrated here (Fig.
@@
 \ref{fig:imp_var}).
-The third PC, or first most important PC, describes variation in the relative position of landmarks on anterior and posterior portions of the plastron and represents 9.6697\% of total variation.
-The eighth PC, or second most important PC, mostly describes variation in landmarks along the midline of the plastron and represents 4.1061\% of total variation.
-The major variations along these axes correspond well to the differences between the class means and the mean plastron shape (Fig.
+The third and most important PC describes variation in the relative position of landmarks on anterior and posterior portions of the plastron and represents 9.6697\% of total variation.
+The eighth and second most important PC mostly describes variation in landmarks along the midline of the plastron and represents 4.1061\% of total variation.
+The major variations along these axes correspond well to the differences between the mean shape of each class (Fig.
 \ref{fig:mean_shape}) where major class differences seem based on the relative ballooning or shrinking of the anterior and posterior portions of the plastron together.
@@
 % future directions
+%   potentially useful for phylogenetic comparative method stuff
 %   conservation utility

### 2013-07-28 src/turtle_plots.r

@@ -45,8 +45,8 @@ dev.off()
 
 # gap statistic plot
 gap <- gap.plot(tadult.gap)
-gap <- gap + theme(axis.title = element_text(size = 7.5),
-                   axis.text = element_text(size = 6))
+gap <- gap + theme(axis.title = element_text(size = 8.5),
+                   axis.text = element_text(size = 7))
 ggsave(file = '../documents/figure/gap_res.png', plot = gap)
 
 
@@ -73,20 +73,20 @@ roc.mod$.id[roc.mod$.id == 'sh2'] <- 'morph 2'
 roc.mod$.id[roc.mod$.id == 'sh3'] <- 'molec 1'
 roc.mod$.id[roc.mod$.id == 'spinks'] <- 'molec 2'
 
-ggroc <- ggplot(roc.mod, aes(x = variable, y = value, color = mod.names))
+ggroc <- ggplot(roc.mod, aes(x = variable, y = value, lty = mod.names))
 ggroc <- ggroc + geom_line()
 ggroc <- ggroc + scale_x_continuous(breaks = seq(max(roc.mod$variable)))
-ggroc <- ggroc + scale_color_manual(labels = c('multinomial', 
-                                                 'random forest'),
-                                      values = cbp)
+ggroc <- ggroc + scale_linetype_manual(labels = c('multinomial', 
+                                                  'random forest'),
+                                      values = c(1,2))
 ggroc <- ggroc + theme(legend.title = element_blank(),
                        #legend.position = 'bottom',
                        legend.margin = unit(0, 'cm'),
-                       legend.text = element_text(size = 4),
-                       axis.title = element_text(size = 8),
-                       axis.text = element_text(size = 5),
-                       strip.text = element_text(size = 5))
-ggroc <- ggroc + labs(x = '# of features (PCs)', y = 'AUC ROC')
+                       legend.text = element_text(size = 5),
+                       axis.title = element_text(size = 9),
+                       axis.text = element_text(size = 6),
+                       strip.text = element_text(size = 6))
+ggroc <- ggroc + labs(x = '# of features (PCs)', y = 'AUC')
 ggroc <- ggroc + facet_wrap(~.id)
 ggsave(file = '../documents/figure/roc_sel.png', plot = ggroc)
 
[…]

### 2013-07-28 documents/turtle_paper.Rnw
tables added. some labels fixed. need to edit abstract.

@@
 source('../src/turtle_plots.r')
+source('../src/turtle_tables.r')
 
@@
 % multinomial logistic regression model
-The AICc best multinomial logistic regression models, for all four classification schemes, all include the first \Sexpr{length(tm.a.analysis$best$spinks$coefnames)} possible PCs as features (Table SUPPLEMENT).
+The AICc best multinomial logistic regression models, for all four classification schemes, all include the first \Sexpr{length(tm.a.analysis$best$spinks$coefnames)} possible PCs as features (Tables \ref{tab:mod_sel_1}\ref{tab:mod_sel_2}\ref{tab:mod_sel_3}, and \ref{tab:mod_sel_4}).
 The second best models for all classification schemes had the first \Sexpr{length(tm.a.analysis$best$spinks$coefnames) - 1} PCs as features.
@@
 
+\pagebreak
+
+\begin{sidewaystable}
+  \centering
+<<results = 'asis'>>=
+print.xtable(tmx.1, 
+             include.rownames = FALSE,
+             floating = FALSE,
+             size = '\\small',
+             sanitize.colnames.function = identity)
+@
+\caption{Multinomial logistic regression model selection table for the first morphologically based classification hypothesis.
+This classification hypothesis corresponds to ``morph 1'' also depicted in figures \ref{fig:roc} and \ref{fig:gen_res}.
+This hypothesis is based on \citet{Seeliger1945}.}
+  \label{tab:mod_sel_1}
+\end{sidewaystable}
+
+\begin{sidewaystable}
+  \centering
+<<results = 'asis'>>=
+print.xtable(tmx.2, 
+             include.rownames = FALSE,
+             floating = FALSE,
+             size = '\\small',
+             sanitize.colnames.function = identity)
+@
+\caption{Multinomial logistic regression model selection table for the second morphologically based classification hypothesis.
+This classification hypothesis corresponds to ``morph 2'' also depicted in figures \ref{fig:roc} and \ref{fig:gen_res}.
+This hypothesis is based on \citet{Seeliger1945}.}
+  \label{tab:mod_sel_2}
+\end{sidewaystable}
+
+\begin{sidewaystable}
+  \centering
+<<results = 'asis'>>=
+print.xtable(tmx.3, 
+             include.rownames = FALSE,
+             floating = FALSE,
+             size = '\\small',
+             sanitize.colnames.function = identity)
+@
+\caption{Multinomial logistic regression model selection table for the first molecularlly based classification hypothesis.
+This classification hypothesis corresponds to ``molec 1'' also depicted in figures \ref{fig:roc} and \ref{fig:gen_res}.
+This hypothesis is based on \citet{Spinks2005} and \citet{Spinks2010}.}
+  \label{tab:mod_sel_3}
+\end{sidewaystable}
+
+\begin{sidewaystable}
+  \centering
+<<results = 'asis'>>=
+print.xtable(tmx.4, 
+             include.rownames = FALSE,
+             floating = FALSE,
+             size = '\\small',
+             sanitize.colnames.function = identity)
+@
+\caption{Multinomial logistic regression model selection table for the second molecularlly based classification hypothesis.
+This classification hypothesis corresponds to ``molec 2'' also depicted in figures \ref{fig:roc} and \ref{fig:gen_res}.
+This hypothesis is based on \citet{Spinks2005} and \citet{Spinks2010}.}
+  \label{tab:mod_sel_4}
+\end{sidewaystable}
+
 \end{document}

### 2013-07-28 documents/turtle_paper.tex
tables added. some labels fixed. need to edit abstract.

@@
 % multinomial logistic regression model
-The AICc best multinomial logistic regression models, for all four classification schemes, all include the first 10 possible PCs as features (Table SUPPLEMENT).
+The AICc best multinomial logistic regression models, for all four classification schemes, all include the first 10 possible PCs as features (Tables \ref{tab:mod_sel_1}\ref{tab:mod_sel_2}\ref{tab:mod_sel_3}, and \ref{tab:mod_sel_4}).
 The second best models for all classification schemes had the first 9 PCs as features.
@@
 
+\pagebreak
+
+\begin{sidewaystable}
+  \centering
+% latex table generated in R 3.0.1 by xtable 1.7-1 package
+% Sun Jul 28 05:52:00 2013
+{\small
+\begin{tabular}{lllllllllllrrrrr}
+  \hline
+(Intercept) & PC1 & PC2 & PC3 & PC4 & PC5 & PC6 & PC7 & PC8 & PC9 & PC10 & df & logLik & AICc & delta & weight \\ 
+  \hline
++ & + & + & + & + & + & + & + & + & + & + & 22.00 & -245.34 & 537.41 & 0.00 & 0.74 \\ 
+  + & + & + & + & + & + & + & + & + & + &  & 20.00 & -248.61 & 539.48 & 2.06 & 0.26 \\ 
+  + & + & + & + & + & + & + & + & + &  &  & 18.00 & -255.73 & 549.28 & 11.87 & 0.00 \\ 
+  + & + & + & + & + & + & + & + &  &  &  & 16.00 & -267.59 & 568.62 & 31.21 & 0.00 \\ 
+  + & + & + & + & + & + &  &  &  &  &  & 12.00 & -283.44 & 591.70 & 54.29 & 0.00 \\ 
+  + & + & + & + & + &  &  &  &  &  &  & 10.00 & -286.02 & 592.61 & 55.20 & 0.00 \\ 
+  + & + & + & + & + & + & + &  &  &  &  & 14.00 & -282.24 & 593.59 & 56.17 & 0.00 \\ 
+  + & + & + & + &  &  &  &  &  &  &  & 8.00 & -307.43 & 631.23 & 93.81 & 0.00 \\ 
+  + & + & + &  &  &  &  &  &  &  &  & 6.00 & -340.94 & 694.09 & 156.68 & 0.00 \\ 
+  + & + &  &  &  &  &  &  &  &  &  & 4.00 & -345.95 & 700.01 & 162.60 & 0.00 \\ 
+   \hline
+\end{tabular}
+}
+
+\caption{Multinomial logistic regression model selection table for the first morphologically based classification hypothesis.
+This classification hypothesis corresponds to ``morph 1'' also depicted in figures \ref{fig:roc} and \ref{fig:gen_res}.
+This hypothesis is based on \citet{Seeliger1945}.}
+  \label{tab:mod_sel_1}
+\end{sidewaystable}
+
+\begin{sidewaystable}
+  \centering
+% latex table generated in R 3.0.1 by xtable 1.7-1 package
+% Sun Jul 28 05:52:00 2013
+{\small
+\begin{tabular}{lllllllllllrrrrr}
+  \hline
+(Intercept) & PC1 & PC2 & PC3 & PC4 & PC5 & PC6 & PC7 & PC8 & PC9 & PC10 & df & logLik & AICc & delta & weight \\ 
+  \hline
++ & + & + & + & + & + & + & + & + & + & + & 22.00 & -242.26 & 531.25 & 0.00 & 0.86 \\ 
+  + & + & + & + & + & + & + & + & + & + &  & 20.00 & -246.34 & 534.94 & 3.70 & 0.14 \\ 
+  + & + & + & + & + & + & + & + & + &  &  & 18.00 & -252.76 & 543.34 & 12.10 & 0.00 \\ 
+  + & + & + & + & + & + & + & + &  &  &  & 16.00 & -263.19 & 559.83 & 28.59 & 0.00 \\ 
+  + & + & + & + & + & + &  &  &  &  &  & 12.00 & -275.48 & 575.79 & 44.54 & 0.00 \\ 
+  + & + & + & + & + & + & + &  &  &  &  & 14.00 & -273.87 & 576.85 & 45.60 & 0.00 \\ 
+  + & + & + & + & + &  &  &  &  &  &  & 10.00 & -280.27 & 581.11 & 49.86 & 0.00 \\ 
+  + & + & + & + &  &  &  &  &  &  &  & 8.00 & -303.54 & 623.46 & 92.22 & 0.00 \\ 
+  + & + & + &  &  &  &  &  &  &  &  & 6.00 & -342.40 & 697.01 & 165.76 & 0.00 \\ 
+  + & + &  &  &  &  &  &  &  &  &  & 4.00 & -348.78 & 705.67 & 174.42 & 0.00 \\ 
+   \hline
+\end{tabular}
+}
+
+\caption{Multinomial logistic regression model selection table for the second morphologically based classification hypothesis.
+This classification hypothesis corresponds to ``morph 2'' also depicted in figures \ref{fig:roc} and \ref{fig:gen_res}.
+This hypothesis is based on \citet{Seeliger1945}.}
+  \label{tab:mod_sel_2}
+\end{sidewaystable}
+
+\begin{sidewaystable}
+  \centering
+% latex table generated in R 3.0.1 by xtable 1.7-1 package
+% Sun Jul 28 05:52:00 2013
+{\small
+\begin{tabular}{lllllllllllrrrrr}
+  \hline
+(Intercept) & PC1 & PC2 & PC3 & PC4 & PC5 & PC6 & PC7 & PC8 & PC9 & PC10 & df & logLik & AICc & delta & weight \\ 
+  \hline
++ & + & + & + & + & + & + & + & + & + & + & 33.00 & -297.91 & 668.06 & 0.00 & 0.97 \\ 
+  + & + & + & + & + & + & + & + & + & + &  & 30.00 & -305.12 & 675.37 & 7.30 & 0.03 \\ 
+  + & + & + & + & + & + & + & + & + &  &  & 27.00 & -314.81 & 687.76 & 19.70 & 0.00 \\ 
+  + & + & + & + & + & + & + & + &  &  &  & 24.00 & -329.53 & 710.30 & 42.24 & 0.00 \\ 
+  + & + & + & + & + & + & + &  &  &  &  & 21.00 & -342.93 & 730.35 & 62.28 & 0.00 \\ 
+  + & + & + & + & + & + &  &  &  &  &  & 18.00 & -353.86 & 745.55 & 77.49 & 0.00 \\ 
+  + & + & + & + & + &  &  &  &  &  &  & 15.00 & -357.66 & 746.59 & 78.53 & 0.00 \\ 
+  + & + & + & + &  &  &  &  &  &  &  & 12.00 & -385.36 & 795.54 & 127.48 & 0.00 \\ 
+  + & + & + &  &  &  &  &  &  &  &  & 9.00 & -424.50 & 867.48 & 199.41 & 0.00 \\ 
+  + & + &  &  &  &  &  &  &  &  &  & 6.00 & -442.85 & 897.91 & 229.85 & 0.00 \\ 
+   \hline
+\end{tabular}
+}
+
+\caption{Multinomial logistic regression model selection table for the first molecularlly based classification hypothesis.
+This classification hypothesis corresponds to ``molec 1'' also depicted in figures \ref{fig:roc} and \ref{fig:gen_res}.
+This hypothesis is based on \citet{Spinks2005} and \citet{Spinks2010}.}
+  \label{tab:mod_sel_3}
+\end{sidewaystable}
+
+\begin{sidewaystable}
+  \centering
+% latex table generated in R 3.0.1 by xtable 1.7-1 package
+% Sun Jul 28 05:52:00 2013
+{\small
+\begin{tabular}{lllllllllllrrrrr}
+  \hline
+(Intercept) & PC1 & PC2 & PC3 & PC4 & PC5 & PC6 & PC7 & PC8 & PC9 & PC10 & df & logLik & AICc & delta & weight \\ 
+  \hline
++ & + & + & + & + & + & + & + & + & + & + & 33.00 & -253.75 & 579.72 & 0.00 & 1.00 \\ 
+  + & + & + & + & + & + & + & + & + & + &  & 30.00 & -267.22 & 599.55 & 19.83 & 0.00 \\ 
+  + & + & + & + & + & + & + & + & + &  &  & 27.00 & -275.24 & 608.60 & 28.88 & 0.00 \\ 
+  + & + & + & + & + & + & + & + &  &  &  & 24.00 & -302.99 & 657.22 & 77.51 & 0.00 \\ 
+  + & + & + & + & + & + & + &  &  &  &  & 21.00 & -307.70 & 659.88 & 80.16 & 0.00 \\ 
+  + & + & + & + & + &  &  &  &  &  &  & 15.00 & -327.52 & 686.30 & 106.59 & 0.00 \\ 
+  + & + & + & + & + & + &  &  &  &  &  & 18.00 & -324.35 & 686.51 & 106.79 & 0.00 \\ 
+  + & + & + & + &  &  &  &  &  &  &  & 12.00 & -350.14 & 725.11 & 145.39 & 0.00 \\ 
+  + & + & + &  &  &  &  &  &  &  &  & 9.00 & -390.78 & 800.03 & 220.32 & 0.00 \\ 
+  + & + &  &  &  &  &  &  &  &  &  & 6.00 & -405.14 & 822.49 & 242.77 & 0.00 \\ 
+   \hline
+\end{tabular}
+}
+
+\caption{Multinomial logistic regression model selection table for the second molecularlly based classification hypothesis.
+This classification hypothesis corresponds to ``molec 2'' also depicted in figures \ref{fig:roc} and \ref{fig:gen_res}.
+This hypothesis is based on \citet{Spinks2005} and \citet{Spinks2010}.}
+  \label{tab:mod_sel_4}
+\end{sidewaystable}
+
 \end{document}

### 2013-07-29 documents/turtle_paper.Rnw
abstract edited. need to do one last run through, houeskeeping, and make figure labels larger.

@@
 \begin{abstract}
-  %\textbf{Write me please.}
-  Cryptic diversity is the phenomenon where some taxa are believed to be identifiable solely from molecular data.
-The reality of this phenomenon is concerning because the majority of extant taxa are delimited solely based on morphology.
-Additionally, virtually all extant taxa are known entirely from morphology and thus can only be delimited based on morphology.
-Here we address questions about whether it is possible to determine entirely based on morphology if one classification hypotheses can be considered better than others.
+  Cryptic diversity is the phenomenon where some taxa are believed to be identifiable only based on molecular data.
+This phenomenon is concerning because the majority of extant taxa are delimited entirely from morphological differences.
+Additionally, virtually all extinct taxa are known entirely from morphology and thus can only be delimited in this manner.
+Here we address questions about whether it is possible to determine, based on morphology, if one classification hypotheses can be considered better than others.
+We investigate various methodological approaches from characterizing morphological variation.
   %
-  By using a combination of unsupervised and supervised learning methods, each with different useful properties, we demonstrate a suite of approaches for better understanding differences in morphological variation between classes, the odds of classifying an observation as one class relative to another, and what aspects of morphology best describe the variation between different classes.
+  By using a combination of unsupervised and supervised machine learning methods we demonstrate a suite of approaches for better understanding differences in morphology between classes, the odds of classifying an observation as one class relative to another, and what aspects of morphology best describe the differences between classes.
   %
-  These approaches are applied to the classification of the Emyidine turtle, \textit{Emys marmorata}.
-This species has conflicting hypotheses of the number of meaningful clades based on either morphological or molecular information.
-By characterizing the variation in plastral shape and how it may be identifiably different between different classes, we compared multiple different explicit classification hypotheses.
-By splitting a large dataset of specimens into both training and testing datasets, we were also able to determine which of the classification hypotheses best corresponded to the observed plastral variation.
+  These approaches are applied to the classification of the emydid turtle, \textit{Emys marmorata}.
+This species has conflicting hypotheses of the number of meaningful subclades based on either morphological or molecular information.
+We compared multiple explicit classification hypotheses by characterizing the variation plastral shape and how it may be identifiably different between different classes, By splitting a large dataset of specimens into both training and testing datasets, we were also able to determine which of the classification hypotheses best corresponded to the observed plastral variation.
   %
   The results from our analysis shows that the seemingly best classification of \textit{Emys marmorata} is in accordance with the molecularly based hypothesis.
+This demonstrates that by using alternative methods for characterizing variability it is possible to estimate the best classification scheme that agrees with observed morphological variation.
+Additionally, we demontrate that it is possible that not all examples of cryptic variation may be truly cryptic, just a product of sample size or methodology because of the extremely fine scale variation between the different classes.
 
-\noindent (Keywords: Testudines, morphology, geometric morphometrics, random forests)
+\noindent (Keywords: Testudines, Emydidae, morphology, geometric morphometrics, random forests)
 \end{abstract}
@@
 The next best clustering solution had only two clusters, however there is no geographic structure to this classification scheme, with members of these clusters being seemingly randomly distributed SUPPLEMENT?.
+Importantly, these clusters do not conform to the nuclear DNA hypothesis of only two groups, nothern and southern \citep{Spinks2010}.
+
 Unfortunately, it was not possible to obtain enough detailed information on the sex of each \textit{E.
 marmorata} specimen, thus it is difficult at best to determine if this clustering solution corresponds to sexual dimorphism between the observations.
-Male Emyidine turtles are known to have a plastral concavity which may influence landmark position along the midline.
+Male emydid turtles are known to have a plastral concavity which may influence landmark position along the midline.
 However, the plastral concavity of \textit{E.
-mamorata} males is considered less pronounced than in other Emyidine turtles.
+mamorata} males is considered less pronounced than in other emydid turtles.
 While we cannot completely rule out sexual dimorphism as the root cause of this observation, because of the very small degree of known sexual dimorphism in \textit{E.
 marmorata} we are less concerned with possible effects of sexual dimorphism for the later supervised learning methods.
+
 The gap statistic values for both three and four clusters are much lower than for one and two and are statistically identical.

### 2013-07-29 documents/turtle_paper.tex
abstract edited. need to do one last run through, houeskeeping, and make figure labels larger.

@@
 % latex table generated in R 3.0.1 by xtable 1.7-1 package
-% Sun Jul 28 05:52:00 2013
+% Mon Jul 29 07:10:37 2013
 {\small

### 2013-07-30 documents/turtle_paper.Rnw
almost all citations done.

@@
 Here we address questions about whether it is possible to determine, based on morphology, if one classification hypotheses can be considered better than others.
-We investigate various methodological approaches from characterizing morphological variation.
+We investigate various methodological approaches for characterizing morphological variation.
   %
@@
 %   most species are still deliminated solely based on morphology
-Cryptic diversity is the phenomenon that not all taxa can be recognized based on morphology and can only be delimited using molecular information CITATIONS.
+Cryptic diversity is the phenomenon that not all taxa can be recognized based on morphology and can only be delimited using molecular information \citep{Stuart2006,Pfenninger2007,Funk2012,Clare2011}.
 Conceringly, most extant taxa, and nearly all extinct taxa, are delimited based solely via morphological analysis.
@@
 
-Much work has been devoted to species delimitation via sequence difference CITATIONS while comparatively little has been devoted for case of purely morphological data.
-The majority of this effort has focused on identifying differences between already identified taxa CITATIONS and automated taxon identification CITATIONS.
+Much work has been devoted to species delimitation via sequence difference \citep{Fujita2012,Yang2010b} while comparatively little has been devoted for case of purely morphological data.
+The majority of this effort has focused on identifying differences between already identified taxa \citep{Zelditch2004} and automated taxon identification \citep{MacLeod2007}.
 % all the work done on species delimitation strictly from molecular data
@@
 \subsection{Background and system}
-Differences in morphological variation between different classes has previously been analyzed using methods like linear discriminate analysis and canonical variates analysis CITATION ZELDICH.
+Differences in morphological variation between different classes has previously been analyzed using methods like linear discriminate analysis and canonical variates analysis \citep{Zelditch2004}.
 These methods are comparatively simple and straight forward ways of understanding the differences in morphology between classes.
 Also, they are very visual methods which aides in interpretation and presentation of information.
-Neural network models have also been introduced and applied frequently in the context of automated taxon identification along with more general applications MACLEOD BOOK.
+Neural network models have also been introduced and applied frequently in the context of automated taxon identification along with more general applications \citep{MacLeod2007}.
 Here, we used multiple alternative machine learning methods, both unsupervised and supervised, in order to compare different classification hypotheses.
@@
 Plastral landmark configurations were superimposed using generalized Procrustes analysis \citep{Dryden1998a} after which, the principal components (PC) of shape were calculated.
-This was done using the \texttt{shapes} package for R \citep{2013, Dryden2013}.
+This was done using the \texttt{shapes} package for R \citep{2013,Dryden2013}.
 
@@
 %   multinomial logistic regression
-Multinomial logistic regression is an extension of logistic regression, where instead of a binary response there are three or more response classes CITATION.
+Multinomial logistic regression is an extension of logistic regression, where instead of a binary response there are three or more response classes \citep{Venables2002a}.
 Effectively, this type of model can be viewed as multiple, simultaneous logistic regression models for each class and the final classification of the observation being the most probable of all the constituent model results.
 Similar to the odds ratios calculated from the coefficients of a logistic regression, the relative risk of a possible classification with reference to a baseline class can be determined from the coefficients of the model.
-Multinomial logistic regression models were fit using the \texttt{nnet} package for R \citep{Venables2002}
+Multinomial logistic regression models were fit using the \texttt{nnet} package for R \citep{Venables2002a}
 
 %   random forest
-Random forest models are an extension of classification and regression trees (CART) CITATION.
+Random forest models are an extension of classification and regression trees (CART) \citep{Breiman1984,Breiman2001}.
 Because this study relies on classification models, CARTs are explained with reference to classification but the approach is equally valid for regression.
@@
 In a random forest model, many CARTs are built from a random subsample of both the features and the observations.
-This process is then repeated many times and the parameters of the final model was chosen as the mode of estimates from the distribution of CARTs CITATION.
+This process is then repeated many times and the parameters of the final model was chosen as the mode of estimates from the distribution of CARTs \citep{Breiman2001}.
 In addition to fitting a classification model, this procedure allows for the features to be ranked in order of importance.
@@
 \ref{eq:fpr}) of a classification model and the true positive rate (\(TPR\), Eq.
-\ref{eq:tpr}) of a classification model CITATION.
-The area under the ROC curve (AUC) is a summary statistic of the quality of the classification and varies between 0.5 and 1, with 0.5 indicating a model that classifies no better than random and a value of 1 indicating perfect classification CITATION.
-AUC can be used as a model selection criterion for classification models and is especially useful in cases where some if not all of the models in question were not fit via maximum likelihood where a criterion such as AICc (see below) or similar can be used CITATION.
+\ref{eq:tpr}) of a classification model \citep{Hastie2009}.
+The area under the ROC curve (AUC) is a summary statistic of the quality of the classification and varies between 0.5 and 1, with 0.5 indicating a model that classifies no better than random and a value of 1 indicating perfect classification \citep{Hastie2009}.
+AUC can be used as a model selection criterion for classification models and is especially useful in cases where some if not all of the models in question were not fit via maximum likelihood where a criterion such as AICc (see below) or similar can be used \citep{Hastie2009}.
 \begin{table}
@@
 The maximum number of PCs allowed as predictors was \Sexpr{floor(max.ad)} because of both the large number of parameters estimated per model and the necessary sample size needed to estimate that many parameters accurately.
-The final model was that with the lowest AICc \citep{Burnham2002a} AKAIKE AND OTHER CITATION.
+The final model was that with the lowest AICc \citep{Akaike1974,Hurvich1989,Burnham2002a}.
 AICc is a model selection criterion where the model with lowest AICc has the fairest variance--bias tradeoff \citep{Burnham2002a}.
@@
 Instead, a recursive feature selection algorithm was used to choose the optimal number of PCs to include based on the AUC of the model.
-PCs were sequentially added as features and the selected model was that with the highest AUC.
+Following the backwards selection algorithm implemented in \texttt{caret} \citep{Kuhn2013}, the maximum number of features included in an initial model, there importance ranked, and the AUC of the model calculated.
+The lowest ranked feature was then removed, and the AUC of the model recalculated.
+This was repeated until only one feature, remained.
 Similar to the multinomial logistic regression models described above, \Sexpr{floor(max.ad)} was the maximum number of PCs that could have been included in the model.
@@
 Random forest model parameters were estimated from 1000 subtrees.
-The recursive feature selection algorithm used in this study was that implemented in the \texttt{caret} package for R \citep{Kuhn2013}.
+Because PCs were kept in order of importance and not in relation to the amount of variance each PC described, this means that the exact PCs included in each model do not correspond to the PCs in each of the 10 multinomial logistic regressions models.
 
@@
 The second best models for all classification schemes had the first \Sexpr{length(tm.a.analysis$best$spinks$coefnames) - 1} PCs as features.
-The \(\Delta\)AICc values between the optimal and second best model range from \Sexpr{min(daic)} for the first morphological based classification hypothesis to \Sexpr{max(daic)} for the second molecular based classification hypothesis (Table IN SUPPLEMENT?).
+The \(\Delta\)AICc values between the optimal and second best model range from \Sexpr{min(daic)} for the first morphological based classification hypothesis to \Sexpr{max(daic)} for the second molecular based classification hypothesis (Tables \ref{tab:mod_sel_1}\ref{tab:mod_sel_2}\ref{tab:mod_sel_3}, and \ref{tab:mod_sel_4}).
 The first \Sexpr{length(tm.a.analysis$best$spinks$coefnames)} PCs describe \Sexpr{paste0(sum(turtle.fit$percent[1:10]))}\% of total variation in plastral shape.
@@
 
-In the case of all models, there is a substantial increase in model performance as measured by AICc for the multinomial logistic models (Tables SUPPLEMENT) or in AUC for the random forest models and illustrated for the multinomial logistic regression models as the number of features increases (Fig.
+In the case of all models, there is a substantial increase in model performance as measured by AICc for the multinomial logistic models (Tables (Tables \ref{tab:mod_sel_1}\ref{tab:mod_sel_2}\ref{tab:mod_sel_3}, and \ref{tab:mod_sel_4}) or in AUC for the random forest models and illustrated for the multinomial logistic regression models as the number of features increases (Fig.
 \ref{fig:roc}).
@@
   \caption{Effect of increasing the number of PCs as features, or predictors, of classification of plastra for all four classification schemes.
-As the number of PCs increase, AUC increases until eventually leveling off.
+As the total number of features increase, AUC increases until eventually leveling off.
 Both multinomial logisitic regression and random forest models are illustrated here, though AUC based model selection was only performed for random forest models.}
@@
 % variation along the most important axes
+<<>>=
+rf.w <- gsub('PC', x = trf.a$spinks$optVariables, replacement = '')
+@
 The results of fitting the final random forest model also include the variable importance for best separating the different classes.
@@
 Of these \Sexpr{trf.a$spinks$optsize} features, the first three are illustrated here (Fig.
-\ref{fig:imp_pc}) in descending order of importance SUPPLEMENT WITH VARIABLE IMPORTANCE INFORMATION?.
+\ref{fig:imp_pc}) in descending order of importance.
+The PCs included as features in the final random forest model, in descending order of importance, were PCs \Sexpr{paste(rf.w[1:6])} and \Sexpr{paste(rf.w[7])}.
 
@@
 The only unsupervised method employed in this study, PAM, is rather simple and not model based.
-A more useful approach would be employing various model based clustering approaches CITATIONS.
-In this manner, a series of candidate models can be compared via model comparison methods, such as AIC or Bayes factors CITATION, in order to asses the best clustering solution.% Of particular note are nonparametric Bayesian approaches to model based clustering CITATIONS.
+A more useful approach would be employing various model based clustering approaches \citep{Fraley2002,Zhong2003}.
+In this manner, a series of candidate models can be compared via model comparison methods, such as AIC or Bayes factors \citep{Fraley2002}, in order to asses the best clustering solution.% Of particular note are nonparametric Bayesian approaches to model based clustering CITATIONS.
 This approach uses a class of flexible priors to allow for the most optimal clustering solution to be decided from the data.

### 2013-07-30 documents/turtle_paper.tex
almost all citations done.

@@
 The maximum number of PCs allowed as predictors was 10 because of both the large number of parameters estimated per model and the necessary sample size needed to estimate that many parameters accurately.
-The final model was that with the lowest AICc \citep{Burnham2002a} AKAIKE AND OTHER CITATION.
+The final model was that with the lowest AICc \citep{Akaike1974,Hurvich1989,Burnham2002a}.
 AICc is a model selection criterion where the model with lowest AICc has the fairest variance--bias tradeoff \citep{Burnham2002a}.
@@
 Instead, a recursive feature selection algorithm was used to choose the optimal number of PCs to include based on the AUC of the model.
-PCs were sequentially added as features and the selected model was that with the highest AUC.
+Following the backwards selection algorithm implemented in \texttt{caret} \citep{Kuhn2013}, the maximum number of features included in an initial model, there importance ranked, and the AUC of the model calculated.
+The lowest ranked feature was then removed, and the AUC of the model recalculated.
+This was repeated until only one feature, remained.
 Similar to the multinomial logistic regression models described above, 10 was the maximum number of PCs that could have been included in the model.
@@
 The second best models for all classification schemes had the first 9 PCs as features.
-The \(\Delta\)AICc values between the optimal and second best model range from 2.0639 for the first morphological based classification hypothesis to 19.8349 for the second molecular based classification hypothesis (Table IN SUPPLEMENT?).
+The \(\Delta\)AICc values between the optimal and second best model range from 2.0639 for the first morphological based classification hypothesis to 19.8349 for the second molecular based classification hypothesis (Tables \ref{tab:mod_sel_1}\ref{tab:mod_sel_2}\ref{tab:mod_sel_3}, and \ref{tab:mod_sel_4}).
 The first 10 PCs describe 88.6043874205075\% of total variation in plastral shape.
@@
 % variation along the most important axes
+
 The results of fitting the final random forest model also include the variable importance for best separating the different classes.
@@
 Of these 7 features, the first three are illustrated here (Fig.
-\ref{fig:imp_pc}) in descending order of importance SUPPLEMENT WITH VARIABLE IMPORTANCE INFORMATION?.
+\ref{fig:imp_pc}) in descending order of importance.
+The PCs included as features in the final random forest model, in descending order of importance, were PCs 3, 8, 6, 1, 4, 2 and 10.
 
@@
 % latex table generated in R 3.0.1 by xtable 1.7-1 package
-% Mon Jul 29 07:10:37 2013
+% Tue Jul 30 01:18:11 2013
 {\small

### 2013-08-01 documents/turtle_paper.Rnw
almost draft to send to ken. missing 1 citation and a paragraph.

@@
   Cryptic diversity is the phenomenon where some taxa are believed to be identifiable only based on molecular data.
-This phenomenon is concerning because the majority of extant taxa are delimited entirely from morphological differences.
-Additionally, virtually all extinct taxa are known entirely from morphology and thus can only be delimited in this manner.
+This is concerning because the majority of extant taxa and virtually all extinct taxa are delimited entirely via morphology.
 Here we address questions about whether it is possible to determine, based on morphology, if one classification hypotheses can be considered better than others.
-We investigate various methodological approaches for characterizing morphological variation.
 %
-  By using a combination of unsupervised and supervised machine learning methods we demonstrate a suite of approaches for better understanding differences in morphology between classes, the odds of classifying an observation as one class relative to another, and what aspects of morphology best describe the differences between classes.
+  Using a combination of unsupervised and supervised machine learning methods we demonstrate a suite of approaches for better understanding differences in morphology between classes, the odds of classifying one class relative to another, and what aspects of morphology best describe the differences between classes.
   %
@@
 This species has conflicting hypotheses of the number of meaningful subclades based on either morphological or molecular information.
-We compared multiple explicit classification hypotheses by characterizing the variation plastral shape and how it may be identifiably different between different classes, By splitting a large dataset of specimens into both training and testing datasets, we were also able to determine which of the classification hypotheses best corresponded to the observed plastral variation.
+We compared multiple explicit classification hypotheses by characterizing variation in plastral shape and how it may be identifiably different between different classes.
+By splitting a large dataset of specimens into both training and testing datasets, we were also able to determine which of the classification hypotheses best corresponded to the observed plastral variation.
   %
-  The results from our analysis shows that the seemingly best classification of \textit{Emys marmorata} is in accordance with the molecularly based hypothesis.
-This demonstrates that by using alternative methods for characterizing variability it is possible to estimate the best classification scheme that agrees with observed morphological variation.
-Additionally, we demontrate that it is possible that not all examples of cryptic variation may be truly cryptic, just a product of sample size or methodology because of the extremely fine scale variation between the different classes.
+  The results from our analysis shows that the best classification of plastral variation in \textit{Emys marmorata} is in accordance with the molecularly based hypothesis.
+This demonstrates that, by using alternative methods for characterizing variability, it is possible to estimate the classification scheme which most agrees with observed morphological variation.
+Additionally, we demonstrate that it is possible that not all examples of cryptic variation may be truly cryptic, just a product of sample size or methodology because of the extremely fine scale variation between the different classes.
 
@@
 Conceringly, most extant taxa, and nearly all extinct taxa, are delimited based solely via morphological analysis.
-This phenomenon is of great concern when styding variation and diversity dynamics over long periods of time, where apparent morphological stasis may not actually reflect true diversity CITATIONS.
-In the case of endangered or otherwise conserved taxa, morphometric approaches for classifying and identifying taxa and populations of conservation importance would greatly improve the ability to maintain these high risk groups.
+This phenomenon is of great concern when studying variation and diversity dynamics over long periods of time, where apparent morphological stasis may not actually reflect true diversity CITATIONS.
+In the case of endangered or conserved taxa, morphometric approaches for classifying and identifying taxa and populations of importance would greatly improve the ability to maintain these high risk groups.
 Additionally, this would allow for better classifying extinct taxa.
@@
 Here, we used multiple alternative machine learning methods, both unsupervised and supervised, in order to compare different classification hypotheses.
-These different methods provide different and unique advantages for understanding how to classify taxa, with what accuracy, and what these classifications are based on.
+These methods provide different and unique advantages for understanding how to classify taxa, with what accuracy, and what these classifications are based on.
 Additionally, we investigate variation in continuous traits, and do not search for discrete differences between each class, instead focusing on suites of traits together.
@@
 % this paragraph kind of sucks
-The two major cases of machine learning methods, unsupervised and supervised, are essentially extensions of known statistical methods.
+The two major classes of machine learning methods, unsupervised and supervised, are essentially extensions of known statistical methods.
 Unsupervised learning methods are analogous to clustering and density estimation methods, while supervised learning methods are analogous classification and regression models.
-In both cases, many of these methods may not be fit via maximum likelihood but may be supplemented by randomization, sorting, and partitioning algorithms along with the maximization or minimization of summary statistics in order to best estimate a general model for all data, both sampled and unsampled.
+In both cases, many of these methods are not fit via maximum likelihood and are supplemented by randomization, sorting, and partitioning algorithms along with the maximization or minimization of summary statistics in order to best estimate a general model for all data, both sampled and unsampled.
 The application of the alternative approaches used in this study illustrates only a sampling of the various previously derived methods for clustering observations and fitting classification models.
@@
 More recently, \textit{E.
-marmorata} was divided into four clades based on mitochondrial DNA: a northern clade, a southern clade, and two central Californian clades \citep{Spinks2005,Spinks2010}.
-While nuclear DNA supports two major clades, one northern and one southern, \citet{Spinks2010} argue that the four clade classification is of greater conservation utility to use the mitochondrially based classification scheme.
+marmorata} was divided into four clades based on mitochondrial DNA: a northern clade, a southern clade, and eastern and western central Californian clades \citep{Spinks2005,Spinks2010}.
+While nuclear DNA supports two major clades, one northern and one southern, \citet{Spinks2010} argue that the four clade classification is of greater conservation utility.
 While the mitochondrially based classification is considered robust, there is no known morphological differentiation between these clades.
@@
 marmorata} museum specimens.
-Specimen classification was based on known specimen geographic information.
-Geographic information was recorded from museum collection information.
-When precise latitude and longitude information was not known it was estimated from whatever locality information was present.
+Specimen classification was based on known specimen geographic information which was recorded from museum collection information.
+When precise latitude and longitude information was not available it was estimated from whatever locality information was present.
 Because the specimens used to define the subclades in \citet{Spinks2005} and \citet{Spinks2010} were not available for study, all specimen classifications were based solely on this geographic information and not from explicit assignment in previous studies.
@@
 The gap statistic is defined
-\[Gap_{n}(k) = E^{*}_{n}(\log(W_{k}) - \log(W_{k})\] 
+\[Gap_{n}(k) = E^{*}_{n}[\log(W_{k})] - \log(W_{k})\] 
 where \(W_{k}\) is
-\[W_{k} = \sum^{k}_{r = 1}{\frac{1}{2n_{r}} (\sum_{i,i' \in C_{r}} d_{ii'})}\]
-the dispersion of the clustering solution or the sum of the pairwise dissimilarities (\(d_{ii'}\)) between observations in each cluster and there respective mediods (\(C\)) for all clusters \(r\).
-\(E^{*}_{n}\) is the expectation of dispersion from a sample \(n\) of a reference distribution.
+\[W_{k} = \sum^{k}_{r = 1}{\frac{1}{2n_{r}} (\sum_{i,i' \in C_{r}} d_{ii'})}\].
+\(d_{ii'}\) is the dispersion of the clustering solution or the sum of the pairwise dissimilarities between observations in each cluster and there respective mediods (\(C\)) for all clusters \(r\).
+This value is averaged and compared to the expected dispersion (\(E^{*}_{n}\)) of a sample \(n\) from a reference distribution.
 In this case, the reference distribution was estimated from \Sexpr{tadult.gap$B} resamples of the dataset taking into account the original structure of the data.
 %  R implementation
-This analysis was conducted using the \texttt{cluster} package for R \citep{Maechler2013} using all \Sexpr{nrow(turtle.adult)} plastral landmark configurations.
+This analysis was conducted using the \texttt{cluster} package for R \citep{Maechler2013} using all \Sexpr{nrow(turtle.adult)} observations.
 
@@
 Effectively, this type of model can be viewed as multiple, simultaneous logistic regression models for each class and the final classification of the observation being the most probable of all the constituent model results.
-Similar to the odds ratios calculated from the coefficients of a logistic regression, the relative risk of a possible classification with reference to a baseline class can be determined from the coefficients of the model.
+Similar to the odds ratios calculated from the coefficients of a logistic regression, the relative risk of a classification with reference to a baseline class can be determined from the coefficients of the model.
 Multinomial logistic regression models were fit using the \texttt{nnet} package for R \citep{Venables2002a}
@@
 The recursion continues until the resulting observations all share the same class or no more meaningful partitions are possible.
+The resulting model is a tree structure by which observations are classified at each intersection via the estimated cutoff points from the attribute tests made during model fitting.
 
@@
 %     best AUC
-Supervised learning models have tuning parameters which are estimated to increase the genearlizability of the model and prevent them from being overfit.
-For both types of supervised learning methods, tuning parameters were estimated using 10 rounds of 10-fold cross-validation (CV) across a grid search of all tuning parameter combinations.
+Supervised learning models have tuning parameters which help to increase the genearlizability of the model and prevent them from being overfit.
+For the supervised learning models fit in this study, tuning parameters were estimated via 10 rounds of 10-fold cross-validation (CV) across a grid search of all tuning parameter combinations.
 Optimal tuning parameter values were selected based on area under the receiver operating characteristic (ROC) curve.
@@
 \ref{eq:tpr}) of a classification model \citep{Hastie2009}.
-The area under the ROC curve (AUC) is a summary statistic of the quality of the classification and varies between 0.5 and 1, with 0.5 indicating a model that classifies no better than random and a value of 1 indicating perfect classification \citep{Hastie2009}.
+The area under the ROC curve (AUC) is a summary statistic of the quality of the classification and varies between 0.5 and 1, with values of 0.5 indicating a model that classifies no better than random and a value of 1 indicating perfect classification \citep{Hastie2009}.
 AUC can be used as a model selection criterion for classification models and is especially useful in cases where some if not all of the models in question were not fit via maximum likelihood where a criterion such as AICc (see below) or similar can be used \citep{Hastie2009}.
-\begin{table}
-  \centering
-  \begin{tabular}[c]{ p{2cm} c | p{2cm} | p{2cm} |}
-    \cline{3-4} 
-    & & \multicolumn{2}{ c |}{Predicted class} \\ \cline{3-4}
-    & & 1 & 0 \\ \hline
-    \multicolumn{1}{| c |}{\multirow{2}{*}{Actual class}}
-    & 1 & TRUE \newline POSITIVE & FALSE \newline NEGATIVE \\ \cline{2-4}
-    \multicolumn{1}{| c |}{} & 0 & FALSE \newline POSITIVE & TRUE \newline NEGATIVE \\
-    \hline
-  \end{tabular}
-  \caption{Example confusion matrix.
-The columns correspond to the predicted class of an observation, while the rows correspond to the actual class of that observation.
-Depending on the type match between the prediction and reality, four different outcomes are possible: true positive (TP), false negative (FN), false positive (FP), and true negative (TN).
-These four quantities are used for calculating all confusion matrix statistics.}
-  \label{tab:conf}
-\end{table}
+It is important to note that, unlike AICc, AUC is not calculated with reference to the complexity of the model.
 
@@
 \end{equation}
+
 %   model selection
@@
 Model selection was performed in this manner because the optimal number of PCs to use as predictors was not know \textit{a priori}, and while including all of the PCs of shape would mean that all shape variability would be used to estimate class, this may cause the model to be overfit and not provide an accurate estimate of unsampled plastral variation.
+In addition to the AICc of each model the \(\Delta\)AICc and Akaike weights are also reported.
+\(\Delta\)AICc values are the different in AICc between the AICc best model and that model while Akaike weights are a transformation of the AICc of a model with relation to all other models being compared and measures the relative amount of information explained by that model compared to all other models \citep{Burnham2002a}.
 
@@
 Instead, a recursive feature selection algorithm was used to choose the optimal number of PCs to include based on the AUC of the model.
-Following the backwards selection algorithm implemented in \texttt{caret} \citep{Kuhn2013}, the maximum number of features included in an initial model, there importance ranked, and the AUC of the model calculated.
+Following the backwards selection algorithm implemented in \texttt{caret} \citep{Kuhn2013}, the maximum number of features were included in the initial model, there importance ranked, and the AUC of the model calculated.
 The lowest ranked feature was then removed, and the AUC of the model recalculated.
 This was repeated until only one feature, remained.
-Similar to the multinomial logistic regression models described above, \Sexpr{floor(max.ad)} was the maximum number of PCs that could have been included in the model.
-After each PC was added, 10-fold CV was used to estimate the optimal values of the tuning parameters as well as quantify the uncertainty of each model.
+Similar to the multinomial logistic regression models described above, the maximum number of PCs that could have been included in the model was \Sexpr{floor(max.ad)}.
+After each PC was removed , 10-fold CV was used to estimate the optimal values of the tuning parameters as well as quantify the uncertainty of each model.
 Random forest model parameters were estimated from 1000 subtrees.
@@
 A distribution of AUC values was estimated for each classification scheme via \Sexpr{rr$sh1$R} nonparametric bootstrap resamples of the training dataset.
+The difference in distributions was assessed using pairwise Mann-Whitney U tests.
 
@@
 \ref{fig:pca}).
-However, it is important to note that Procustes based superimposition attempts to evenly distribute variance around the mean landmark ZELDICH and this observation should be considered cursory at best.
+However, it is important to note that Procustes based superimposition attempts to evenly distribute variance around the mean landmark \citep{Zelditch2004} and this observation should be considered cursory at best.
 
@@
 \ref{fig:gap}).
-The next best clustering solution had only two clusters, however there is no geographic structure to this classification scheme, with members of these clusters being seemingly randomly distributed SUPPLEMENT?.
-Importantly, these clusters do not conform to the nuclear DNA hypothesis of only two groups, nothern and southern \citep{Spinks2010}.
+The next best clustering solution had only two clusters, however there is no geographic structure to this classification scheme, with members of these clusters being seemingly randomly distributed (Fig.
+\ref{fig:gap_map}).
+Importantly, these clusters do not conform to the nothern and souther groups from the nuclear DNA hypothesis \citep{Spinks2010}.
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics{figure/gap_map}
+  \caption{Clustering solution for PAM with two mediods for the entire set of observed \textit{E.
+marmorata}.
+Clustering was based entirely on the \(\rho\) dissimilarity matrix of ``half'' plastral landmark configurations following Procrustes superimposition.}
+  \label{fig:gap_map}
+\end{figure}
 
 Unfortunately, it was not possible to obtain enough detailed information on the sex of each \textit{E.
-marmorata} specimen, thus it is difficult at best to determine if this clustering solution corresponds to sexual dimorphism between the observations.
+marmorata} specimen, thus it is difficult at best to determine if this clustering solution corresponds to sexual dimorphism between observations.
 Male emydid turtles are known to have a plastral concavity which may influence landmark position along the midline.
@@
 % multinomial logistic regression model
-The AICc best multinomial logistic regression models, for all four classification schemes, all include the first \Sexpr{length(tm.a.analysis$best$spinks$coefnames)} possible PCs as features (Tables \ref{tab:mod_sel_1}\ref{tab:mod_sel_2}\ref{tab:mod_sel_3}, and \ref{tab:mod_sel_4}).
+The AICc best multinomial logistic regression models, for all four classification schemes, all include the first \Sexpr{length(tm.a.analysis$best$spinks$coefnames)} possible PCs as features (Tables \ref{tab:mod_sel_1},\ref{tab:mod_sel_2},\ref{tab:mod_sel_3}, and \ref{tab:mod_sel_4}).
 The second best models for all classification schemes had the first \Sexpr{length(tm.a.analysis$best$spinks$coefnames) - 1} PCs as features.
-The \(\Delta\)AICc values between the optimal and second best model range from \Sexpr{min(daic)} for the first morphological based classification hypothesis to \Sexpr{max(daic)} for the second molecular based classification hypothesis (Tables \ref{tab:mod_sel_1}\ref{tab:mod_sel_2}\ref{tab:mod_sel_3}, and \ref{tab:mod_sel_4}).
+The \(\Delta\)AICc values between the optimal and second best model range from \Sexpr{min(daic)} for the first morphological based classification hypothesis to \Sexpr{max(daic)} for the second molecular based classification hypothesis (Tables \ref{tab:mod_sel_1},\ref{tab:mod_sel_2},\ref{tab:mod_sel_3}, and \ref{tab:mod_sel_4}).
 The first \Sexpr{length(tm.a.analysis$best$spinks$coefnames)} PCs describe \Sexpr{paste0(sum(turtle.fit$percent[1:10]))}\% of total variation in plastral shape.
@@
 
-In the case of all models, there is a substantial increase in model performance as measured by AICc for the multinomial logistic models (Tables (Tables \ref{tab:mod_sel_1}\ref{tab:mod_sel_2}\ref{tab:mod_sel_3}, and \ref{tab:mod_sel_4}) or in AUC for the random forest models and illustrated for the multinomial logistic regression models as the number of features increases (Fig.
+In the case of all models, there is a substantial increase in model performance as measured by AICc for the multinomial logistic models (Tables \ref{tab:mod_sel_1},\ref{tab:mod_sel_2},\ref{tab:mod_sel_3}, and \ref{tab:mod_sel_4}) or in AUC for the random forest models and illustrated for the multinomial logistic regression models as the number of features increases (Fig.
 \ref{fig:roc}).
@@
 
-% table
-%   AICc model selection
-%   SUPPLEMENT?
-
-%   maps
-%     facet: training, multinomial, rf
-The results from the generalization of the selected supervised learning models, measured by the distributions of the bootstrapped AUC values of the testing dataset, demonstrate that one of the molecular classification hypotheses was the best overall classification scheme (Fig.
+The results from the generalization of the selected supervised learning models, measured by the distributions of the bootstrapped AUC values of the testing dataset, show that one of the molecular classification hypotheses was the best overall classification scheme (Fig.
 \ref{fig:gen_res}).
-For both methods, the distribution of bootstrapped AUC for the molecular hypothesis was significantly greater MANN-WHITNEY U TEST than all of the other classification schemes.
+For both methods, the distribution of bootstrapped AUC for the molecular hypothesis was significantly greater than all of the other classification schemes (Tables \ref{tab:mm_test} and \ref{tab:rf_test}).
 Remarkably, the best classification hypothesis was identical based on both the multinomial logistic regression and random forest models.
@@
 The results of fitting the final random forest model also include the variable importance for best separating the different classes.
-The selected random forest model for the best classification scheme had \Sexpr{trf.a$spinks$optsize} PCs as features, AUC would not increase.
+The selected random forest model for the best classification scheme had \Sexpr{trf.a$spinks$optsize} PCs as features.
+The PCs included as features in the final random forest model, in descending order of importance, were PCs \Sexpr{paste(rf.w[1:6])} and \Sexpr{paste(rf.w[7])}.
 Of these \Sexpr{trf.a$spinks$optsize} features, the first three are illustrated here (Fig.
 \ref{fig:imp_pc}) in descending order of importance.
-The PCs included as features in the final random forest model, in descending order of importance, were PCs \Sexpr{paste(rf.w[1:6])} and \Sexpr{paste(rf.w[7])}.
 
@@
 %RELATIVE RISK FROM THE MULTINOMIAL LOGISTIC REGRESSION MODELS
-The relative risk values for classification from the multinomial logistic regression model based on the three most important PCs demonstrates that while an individual PC might not be sufficient to differentiate any individual class from the northern group, that given multiple features together the odds of determining the correct classification increase (Fig.
+The relative risk values for classification from the multinomial logistic regression model, based on the three most important PCs, demonstrate that while an individual PC might not be sufficient to differentiate any individual class from the northern group, given multiple features the odds of determining the correct classification increase (Fig.
 \ref{fig:rel_risk}).
-Changes along the second most important axes contribute very little increasing the odds of classification for any of the classes in relation to the northern group.
-This is observable from the class histograms along PC 8 (Fig.
+Changes along the second most important axes contribute very little to increasing the odds of classification.
+This is observable from the class histograms of PC 8 (Fig.
 \ref{fig:imp_var}).
 Changes along the first and third most important axes contribute more obviously to increasing the odds of correctly identifying the class of an observation, a result that is observable in both the relative risk (Fig.
-\ref{fig:rel_risk}) and histograms of the different classes from the PCs (Fig.
+\ref{fig:rel_risk}) and the different class histograms of the PCs (Fig.
 \ref{fig:imp_var}).
@@
 %   remarkable concordance in the supervised learning methods
-The results of this study support the mitochrondial based hypothesis of classification of \textit{E.
+The results of this study support the mitochrondial based classification hypothesis of \textit{E.
 marmorata} \citep{Spinks2005,Spinks2010}.
@@
 \ref{fig:roc}) indicates that the morphological variation between classes is extremely fine grained.
-This was also exemplified by the small differences in mean class shape for the final chosen classification scheme (Fig.
+This was also exemplified by the small differences between mean class shapes of the final chosen classification scheme (Fig.
 \ref{fig:imp_var}).
@@
 While all aspects of shape may be evolving simultaneously, and not along individual PCs, including all shape variation in each model might increase model complexity beyond a reasonable level for the sample size and possibly the necessary complexity to accurately model the response.
-Additionally, because individual PCs are used as features in the models this does not completely and accurately represent shape evolution and how exactly different classes might be evolving in relation to each other.
+Additionally, because only individual PCs are used as features in the models, this does not accurately represent shape evolution and how exactly different classes might be evolving in relation to each other.
 However, this compromise is not without its advantages.
 Because both AICc and AUC values improved rapidly with increased model complexity (Fig.
-\ref{fig:roc}), this indicated how fine scale the actual variation between classes was.
+\ref{fig:roc}), this helped demonstrate how fine scale the actual variation between classes was.
 Additionally, the relative risk values from the mulitinomial logistic regression models demonstrate that a single PC is probably not sufficient for estimating the class of an observation, but that given a set of PCs this classification would be more accurate (Fig.
@@
 The results of the relative risk values also shows that while a variable can be considered important for the random forest models (Fig.
-\ref{fig:imp_pc}, \ref{fig:imp_var}).
+\ref{fig:imp_pc}, \ref{fig:imp_var}) they each might not be considered useful on their own (Fig.
+\ref{fig:rel_risk}).
 
@@
 The only unsupervised method employed in this study, PAM, is rather simple and not model based.
-A more useful approach would be employing various model based clustering approaches \citep{Fraley2002,Zhong2003}.
-In this manner, a series of candidate models can be compared via model comparison methods, such as AIC or Bayes factors \citep{Fraley2002}, in order to asses the best clustering solution.% Of particular note are nonparametric Bayesian approaches to model based clustering CITATIONS.
-This approach uses a class of flexible priors to allow for the most optimal clustering solution to be decided from the data.
-Currently, there exists a nonparametric Bayesian clustering method and further development of this approach may prove extremely fruitful for better delimiting taxa solely from morphology.
+A more useful approach would be to employ various model based clustering approaches \citep{Fraley2002,Zhong2003}.
+In this manner, a series of candidate models can be compared via model comparison methods, such as AIC or Bayes factors \citep{Fraley2002}, in order to asses the best clustering solution.
 
@@
 % closing statements
-In this study we have demonstrated that using alternative methodology to that which is most frequently applied, it is possible to determine which classification scheme best explains variation in a taxon amongst a set of alternative hypotheses.
+In this study we have demonstrated that, using alternative methodology to that which is most frequently applied, it is possible to determine which classification scheme best matches variation in a taxon amongst a set of alternative hypotheses.
 The observed plastral variation of \textit{E.
@@
 
+\begin{table}
+  \centering
+  \begin{tabular}[c]{ p{2cm} c | p{2cm} | p{2cm} |}
+    \cline{3-4} 
+    & & \multicolumn{2}{ c |}{Predicted class} \\ \cline{3-4}
+    & & 1 & 0 \\ \hline
+    \multicolumn{1}{| c |}{\multirow{2}{*}{Actual class}}
+    & 1 & TRUE \newline POSITIVE & FALSE \newline NEGATIVE \\ \cline{2-4}
+    \multicolumn{1}{| c |}{} & 0 & FALSE \newline POSITIVE & TRUE \newline NEGATIVE \\
+    \hline
+  \end{tabular}
+  \caption{Example confusion matrix.
+The columns correspond to the predicted class of an observation, while the rows correspond to the actual class of that observation.
+Depending on the type match between the prediction and reality, four different outcomes are possible: true positive (TP), false negative (FN), false positive (FP), and true negative (TN).
+These four quantities are used for calculating all confusion matrix statistics.
+Each of these values is an integer and the sum of the number of occurrences of that event during classification.}
+  \label{tab:conf}
+\end{table}
+
 \begin{sidewaystable}
@@
 @
-\caption{Multinomial logistic regression model selection table for the first morphologically based classification hypothesis.
+  \caption{Model selection table for the multinomial logistic regression models of the first morphologically based classification hypothesis.
 This classification hypothesis corresponds to ``morph 1'' also depicted in figures \ref{fig:roc} and \ref{fig:gen_res}.
-This hypothesis is based on \citet{Seeliger1945}.}
+This hypothesis is based on \citet{Seeliger1945}.
+The column ``delta'' corresponds to the \(\delta\)AICc values of each model, while ``weights'' correspond to the Akaike weight of that model relative to all others.}
   \label{tab:mod_sel_1}
@@
 @
-\caption{Multinomial logistic regression model selection table for the second morphologically based classification hypothesis.
+  \caption{Model selection table for the multinomial logistic regression models of the first morphologically based classification hypothesis.
 This classification hypothesis corresponds to ``morph 2'' also depicted in figures \ref{fig:roc} and \ref{fig:gen_res}.
-This hypothesis is based on \citet{Seeliger1945}.}
+This hypothesis is based on \citet{Seeliger1945}.
+The column ``delta'' corresponds to the \(\delta\)AICc values of each model, while ``weights'' correspond to the Akaike weight of that model relative to all others.}
   \label{tab:mod_sel_2}
@@
 @
-\caption{Multinomial logistic regression model selection table for the first molecularlly based classification hypothesis.
+  \caption{Model selection table for the multinomial logistic regression models of the first morphologically based classification hypothesis.
 This classification hypothesis corresponds to ``molec 1'' also depicted in figures \ref{fig:roc} and \ref{fig:gen_res}.
-This hypothesis is based on \citet{Spinks2005} and \citet{Spinks2010}.}
+This hypothesis is based on \citet{Seeliger1945}.
+The column ``delta'' corresponds to the \(\delta\)AICc values of each model, while ``weights'' correspond to the Akaike weight of that model relative to all others.}
   \label{tab:mod_sel_3}
@@
 @
-\caption{Multinomial logistic regression model selection table for the second molecularlly based classification hypothesis.
+  \caption{Model selection table for the multinomial logistic regression models of the first morphologically based classification hypothesis.
 This classification hypothesis corresponds to ``molec 2'' also depicted in figures \ref{fig:roc} and \ref{fig:gen_res}.
-This hypothesis is based on \citet{Spinks2005} and \citet{Spinks2010}.}
+This hypothesis is based on \citet{Seeliger1945}.
+The column ``delta'' corresponds to the \(\delta\)AICc values of each model, while ``weights'' correspond to the Akaike weight of that model relative to all others.}
   \label{tab:mod_sel_4}
@@
 
+\begin{table}
+  \centering
+<<results = 'asis'>>=
+print.xtable(mm.tabx,
+             includ.rownames = TRUE,
+             floating = FALSE)
+@
+  \caption{Results from pairwise Mann-Whitney U test between the AUC distributions of the generalizations of the multinomial logistic regression models.
+Labels correspond to those in Figure \ref{fig:gen_res}.
+Values of 0 correspond to p-values lower than 0.01.
+P-values were corrected for multiple comparison using the Holm method CITATION.}
+  \label{tab:mm_test}
+\end{table}
+
+\begin{table}
+  \centering
+<<results = 'asis'>>=
+print.xtable(rf.tabx,
+             includ.rownames = TRUE,
+             floating = FALSE)
+@
+  \caption{Results from pairwise Mann-Whitney U test between the AUC distributions of the generalizations of the random forest models.
+Labels correspond to those in Figure \ref{fig:gen_res}.
+Values of 0 correspond to p-values lower than 0.01.
+P-values were corrected for multiple comparison using the Holm method CITATION.}
+  \label{tab:rf_test}
+\end{table}
+
 \end{document}

### 2013-08-01 documents/turtle_paper.tex
almost draft to send to ken. missing 1 citation and a paragraph.

@@
 The gap statistic is defined
-\[Gap_{n}(k) = E^{*}_{n}(\log(W_{k}) - \log(W_{k})\] 
+\[Gap_{n}(k) = E^{*}_{n}[\log(W_{k})] - \log(W_{k})\] 
 where \(W_{k}\) is
-\[W_{k} = \sum^{k}_{r = 1}{\frac{1}{2n_{r}} (\sum_{i,i' \in C_{r}} d_{ii'})}\]
-the dispersion of the clustering solution or the sum of the pairwise dissimilarities (\(d_{ii'}\)) between observations in each cluster and there respective mediods (\(C\)) for all clusters \(r\).
-\(E^{*}_{n}\) is the expectation of dispersion from a sample \(n\) of a reference distribution.
+\[W_{k} = \sum^{k}_{r = 1}{\frac{1}{2n_{r}} (\sum_{i,i' \in C_{r}} d_{ii'})}\].
+\(d_{ii'}\) is the dispersion of the clustering solution or the sum of the pairwise dissimilarities between observations in each cluster and there respective mediods (\(C\)) for all clusters \(r\).
+This value is averaged and compared to the expected dispersion (\(E^{*}_{n}\)) of a sample \(n\) from a reference distribution.
 In this case, the reference distribution was estimated from 500 resamples of the dataset taking into account the original structure of the data.
 %  R implementation
-This analysis was conducted using the \texttt{cluster} package for R \citep{Maechler2013} using all 524 plastral landmark configurations.
+This analysis was conducted using the \texttt{cluster} package for R \citep{Maechler2013} using all 524 observations.
 
@@
 Instead, a recursive feature selection algorithm was used to choose the optimal number of PCs to include based on the AUC of the model.
-Following the backwards selection algorithm implemented in \texttt{caret} \citep{Kuhn2013}, the maximum number of features included in an initial model, there importance ranked, and the AUC of the model calculated.
+Following the backwards selection algorithm implemented in \texttt{caret} \citep{Kuhn2013}, the maximum number of features were included in the initial model, there importance ranked, and the AUC of the model calculated.
 The lowest ranked feature was then removed, and the AUC of the model recalculated.
 This was repeated until only one feature, remained.
-Similar to the multinomial logistic regression models described above, 10 was the maximum number of PCs that could have been included in the model.
-After each PC was added, 10-fold CV was used to estimate the optimal values of the tuning parameters as well as quantify the uncertainty of each model.
+Similar to the multinomial logistic regression models described above, the maximum number of PCs that could have been included in the model was 10.
+After each PC was removed , 10-fold CV was used to estimate the optimal values of the tuning parameters as well as quantify the uncertainty of each model.
 Random forest model parameters were estimated from 1000 subtrees.
@@
 A distribution of AUC values was estimated for each classification scheme via 1000 nonparametric bootstrap resamples of the training dataset.
+The difference in distributions was assessed using pairwise Mann-Whitney U tests.
 
@@
 % multinomial logistic regression model
-The AICc best multinomial logistic regression models, for all four classification schemes, all include the first 10 possible PCs as features (Tables \ref{tab:mod_sel_1}\ref{tab:mod_sel_2}\ref{tab:mod_sel_3}, and \ref{tab:mod_sel_4}).
+The AICc best multinomial logistic regression models, for all four classification schemes, all include the first 10 possible PCs as features (Tables \ref{tab:mod_sel_1},\ref{tab:mod_sel_2},\ref{tab:mod_sel_3}, and \ref{tab:mod_sel_4}).
 The second best models for all classification schemes had the first 9 PCs as features.
-The \(\Delta\)AICc values between the optimal and second best model range from 2.0639 for the first morphological based classification hypothesis to 19.8349 for the second molecular based classification hypothesis (Tables \ref{tab:mod_sel_1}\ref{tab:mod_sel_2}\ref{tab:mod_sel_3}, and \ref{tab:mod_sel_4}).
+The \(\Delta\)AICc values between the optimal and second best model range from 2.0639 for the first morphological based classification hypothesis to 19.8349 for the second molecular based classification hypothesis (Tables \ref{tab:mod_sel_1},\ref{tab:mod_sel_2},\ref{tab:mod_sel_3}, and \ref{tab:mod_sel_4}).
 The first 10 PCs describe 88.6043874205075\% of total variation in plastral shape.
@@
 The results of fitting the final random forest model also include the variable importance for best separating the different classes.
-The selected random forest model for the best classification scheme had 7 PCs as features, AUC would not increase.
+The selected random forest model for the best classification scheme had 7 PCs as features.
+The PCs included as features in the final random forest model, in descending order of importance, were PCs 3, 8, 6, 1, 4, 2 and 10.
 Of these 7 features, the first three are illustrated here (Fig.
 \ref{fig:imp_pc}) in descending order of importance.
-The PCs included as features in the final random forest model, in descending order of importance, were PCs 3, 8, 6, 1, 4, 2 and 10.
 
@@
 % latex table generated in R 3.0.1 by xtable 1.7-1 package
-% Tue Jul 30 01:18:11 2013
+% Thu Aug  1 01:26:49 2013
 {\small
@@
 
-\caption{Multinomial logistic regression model selection table for the first morphologically based classification hypothesis.
+  \caption{Model selection table for the multinomial logistic regression models of the first morphologically based classification hypothesis.
 This classification hypothesis corresponds to ``morph 1'' also depicted in figures \ref{fig:roc} and \ref{fig:gen_res}.
-This hypothesis is based on \citet{Seeliger1945}.}
+This hypothesis is based on \citet{Seeliger1945}.
+The column ``delta'' corresponds to the \(\delta\)AICc values of each model, while ``weights'' correspond to the Akaike weight of that model relative to all others.}
   \label{tab:mod_sel_1}
@@
 
-\caption{Multinomial logistic regression model selection table for the second morphologically based classification hypothesis.
+  \caption{Model selection table for the multinomial logistic regression models of the first morphologically based classification hypothesis.
 This classification hypothesis corresponds to ``morph 2'' also depicted in figures \ref{fig:roc} and \ref{fig:gen_res}.
-This hypothesis is based on \citet{Seeliger1945}.}
+This hypothesis is based on \citet{Seeliger1945}.
+The column ``delta'' corresponds to the \(\delta\)AICc values of each model, while ``weights'' correspond to the Akaike weight of that model relative to all others.}
   \label{tab:mod_sel_2}
@@
 
-\caption{Multinomial logistic regression model selection table for the first molecularlly based classification hypothesis.
+  \caption{Model selection table for the multinomial logistic regression models of the first morphologically based classification hypothesis.
 This classification hypothesis corresponds to ``molec 1'' also depicted in figures \ref{fig:roc} and \ref{fig:gen_res}.
-This hypothesis is based on \citet{Spinks2005} and \citet{Spinks2010}.}
+This hypothesis is based on \citet{Seeliger1945}.
+The column ``delta'' corresponds to the \(\delta\)AICc values of each model, while ``weights'' correspond to the Akaike weight of that model relative to all others.}
   \label{tab:mod_sel_3}
@@
 
-\caption{Multinomial logistic regression model selection table for the second molecularlly based classification hypothesis.
+  \caption{Model selection table for the multinomial logistic regression models of the first morphologically based classification hypothesis.
 This classification hypothesis corresponds to ``molec 2'' also depicted in figures \ref{fig:roc} and \ref{fig:gen_res}.
-This hypothesis is based on \citet{Spinks2005} and \citet{Spinks2010}.}
+This hypothesis is based on \citet{Seeliger1945}.
+The column ``delta'' corresponds to the \(\delta\)AICc values of each model, while ``weights'' correspond to the Akaike weight of that model relative to all others.}
   \label{tab:mod_sel_4}
@@
 
+\begin{table}
+  \centering
+% latex table generated in R 3.0.1 by xtable 1.7-1 package
+% Thu Aug  1 01:26:49 2013
+\begin{tabular}{rrrrr}
+  \hline
+ & morph 1 & morph 2 & molec 1 & molec 2 \\ 
+  \hline
+morph 1 &  &  &  &  \\ 
+  morph 2 & 0.00 &  &  &  \\ 
+  molec 1 & 0.00 & 0.00 &  &  \\ 
+  molec 2 & 0.00 & 0.00 & 0.00 &  \\ 
+   \hline
+\end{tabular}
+
+  \caption{Results from pairwise Mann-Whitney U test between the AUC distributions of the generalizations of the multinomial logistic regression models.
+Labels correspond to those in Figure \ref{fig:gen_res}.
+Values of 0 correspond to p-values lower than 0.01.
+P-values were corrected for multiple comparison using the Holm method CITATION.}
+  \label{tab:mm_test}
+\end{table}
+
+\begin{table}
+  \centering
+% latex table generated in R 3.0.1 by xtable 1.7-1 package
+% Thu Aug  1 01:26:49 2013
+\begin{tabular}{rrrrr}
+  \hline
+ & morph 1 & morph 2 & molec 1 & molec 2 \\ 
+  \hline
+morph 1 &  &  &  &  \\ 
+  morph 2 & 0.00 &  &  &  \\ 
+  molec 1 & 0.51 & 0.00 &  &  \\ 
+  molec 2 & 0.00 & 0.00 & 0.00 &  \\ 
+   \hline
+\end{tabular}
+
+  \caption{Results from pairwise Mann-Whitney U test between the AUC distributions of the generalizations of the random forest models.
+Labels correspond to those in Figure \ref{fig:gen_res}.
+Values of 0 correspond to p-values lower than 0.01.
+P-values were corrected for multiple comparison using the Holm method CITATION.}
+  \label{tab:rf_test}
+\end{table}
+
 \end{document}

### 2013-08-01 src/turtle_plots.r

@@ -45,9 +45,25 @@ dev.off()
 
 # gap statistic plot
 gap <- gap.plot(tadult.gap)
-gap <- gap + theme(axis.title = element_text(size = 8.5),
-                   axis.text = element_text(size = 7))
+gap <- gap + theme(axis.title = element_text(size = 9.5),
+                   axis.text = element_text(size = 8))
 ggsave(file = '../documents/figure/gap_res.png', plot = gap)
+
+
+#classes of the gap statistic plot for 2 clust
+turtle.adult$long[which(turtle.adult$long > -100)] <- turtle.adult$long[which(turtle.adult$long > -100)] - 100
+adult.train$spinks$long[which(adult.train$spinks$long > -100)] <- adult.train$spinks$long[which(adult.train$spinks$long > -100)] - 100
+
+gap.second <- pam(as.dist(turtle.adult.dist), k = 2)
+gap.map <- map.plot(data = turtle.adult,
+                    label = gap.second$clustering,
+                    map = map)
+gap.map <- gap.map + scale_colour_manual(values = cbp)
+gap.map <- gap.map + theme(legend.position = 'none',
+                           legend.text = element_text(size = 7),
+                           axis.title = element_text(size = 10),
+                           axis.text = element_text(size = 7))
+ggsave(file = '../documents/figure/gap_map.png', plot = gap.map)
 
 
 # model selection plots
@@ -82,10 +98,10 @@ ggroc <- ggroc + scale_linetype_manual(labels = c('multinomial',
 ggroc <- ggroc + theme(legend.title = element_blank(),
                        #legend.position = 'bottom',
                        legend.margin = unit(0, 'cm'),
-                       legend.text = element_text(size = 5),
-                       axis.title = element_text(size = 9),
-                       axis.text = element_text(size = 6),
-                       strip.text = element_text(size = 6))
+                       legend.text = element_text(size = 6),
+                       axis.title = element_text(size = 10),
+                       axis.text = element_text(size = 7),
+                       strip.text = element_text(size = 7))
 ggroc <- ggroc + labs(x = '# of features (PCs)', y = 'AUC')
 ggroc <- ggroc + facet_wrap(~.id)
 ggsave(file = '../documents/figure/roc_sel.png', plot = ggroc)
@@ -112,9 +128,9 @@ gdist <- ggplot(dd, aes(x = value, fill = X2)) + geom_density(alpha = 0.5,
 gdist <- gdist + theme(#legend.position = 'bottom',
[…]

### 2013-08-05 documents/turtle_paper.Rnw
completed draft for ken.

@@
 Conceringly, most extant taxa, and nearly all extinct taxa, are delimited based solely via morphological analysis.
-This phenomenon is of great concern when studying variation and diversity dynamics over long periods of time, where apparent morphological stasis may not actually reflect true diversity CITATIONS.
+This phenomenon is of great concern when studying variation and diversity dynamics over long periods of time, where apparent morphological stasis may not actually reflect true diversity \citet{Hunt2008,Eldredge1972,Gould1977a}.
 In the case of endangered or conserved taxa, morphometric approaches for classifying and identifying taxa and populations of importance would greatly improve the ability to maintain these high risk groups.
@@
 marmorata marmorata}, the southern \textit{E.
-marmorata palida}, and a central Californian intergrade zone \citep{Seeliger1945} UNPUBLISHED MASTERS THESIS.
+marmorata palida}, and a central Californian intergrade zone \citep{Seeliger1945,Holland1992}.
 \textit{E.
@@
 
-Previous work on morphological differentiation between subspecies of \textit{E.
-marmorata} focused on just the known subspecies of \citet{Seeliger1945}.
-SEE MASTERS THESIS AND INCLUDE INFORMATION ON THE LINEAR MORPHOMETRIC CLASSIFICATION WORK DONE PRIOR ON E MARMORATA.
+Previous work on morphological differentiation between different populations of \textit{E.
+marmorata} with no alternative subspecific assignments \citep{Seeliger1945,Holland1992}.
+\citet{Holland1992} used comparisons of linear measures of the carapace of \textit{E.
+marmorata} from different populations over the entire range of the species.
+This study was interested in the relative effect of distance versus barriers had in terms of fostering morphological differentiation in \textit{E.
+marmorata}.
+Analyses were performed to determine how different, morphologically, different populations in three different regions of the species range.
+Additionally, the degree of sexual dimorphism present in \textit{E.
+marmorata}.
+The study concluded that distance was a poor indicator of morphological differentiation as opposed to barriers, such as different drainage basins, are probably more important barriers to reproduction.
+This conclusion was later echoed by \citet{Spinks2005} via molecular phylogenetic analysis.
+Additionally, \citet{Holland1992} found that with increasing amount of barriers and distance, morphological differentiation was observable though the underlying variation required many variables obtain indicating the very fine degree of morphological differentiation between putatively distinct populations.
+\citet{Holland1992} concluded that \textit{E.
+marmorata} is best classified as three distinct species as opposed to subspecies: a northern species, southern species, and Columbia basin species.
+This classification is similar to \citet{Seeliger1945}, except elevated to the species as opposed to subspecific level.
 
@@
 However, it is important to note that Procustes based superimposition attempts to evenly distribute variance around the mean landmark \citep{Zelditch2004} and this observation should be considered cursory at best.
+The first PC appears to describe variation in lateral plastral shape, from a very pointed medial edge to a much more rounded and blunt edge (Fig.
+\ref{fig:pc_var}).
+Variation along the second PC appears to be characterized by variation along the midline, especially in terms of the ``straightness'' of the individual landmarks along it.
 
@@
 Depicted here are the for three PCs (lower triangle) and the mean shape with observed variance around each point (upper right).
-The first three PCs account for total \Sexpr{paste0(sum(turtle.fit$perfect[1:3]))} of the variance in plastral shape.}
+The first three PCs account for total \Sexpr{paste0(sum(turtle.fit$percent[1:3]))}\% of the variance in plastral shape.}
   \label{fig:pca}
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width = \textwidth]{figure/pc_var}
+  \caption{Landmark variation along the first two PCs of the Procrustes superimposed ``half'' plastral landmarks.
+The first row corresponds to variation along the first PC, while the second row corresponds to the second PC.
+The left most column represents the observation with the highest eigenscore along that PC, while the right most column represents the observation with the lowest eigenscore.
+The middle column, for both rows, is the mean plastral shape for all observations.
+The first PC represents \Sexpr{turtle.fit$percent[1]}\% of the total variation in plastral shape while PC represents \Sexpr{turtle.fit$percent[2]}\% of the variance.}
+  \label{fig:pc_var}
 \end{figure}
@@
 Importantly, these clusters do not conform to the nothern and souther groups from the nuclear DNA hypothesis \citep{Spinks2010}.
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width = \textwidth]{figure/gap_res}
+  \caption{Gap statistic values for PAM clustering results for the \(\rho\) dissimliarity matrix of plastron shape.
+Error bars are standard errors estimated via \Sexpr{tadult.gap$B} bootstrap samples.}
+  \label{fig:gap}
+\end{figure}
 
@@
 %   gap statistic results
-\begin{figure}[ht]
-  \centering
-  \includegraphics[width = \textwidth]{figure/gap_res}
-  \caption{Gap statistic values for PAM clustering results for the \(\rho\) dissimliarity matrix of plastron shape.
-Error bars are standard errors estimated via \Sexpr{tadult.gap$B} bootstrap samples.}
-  \label{fig:gap}
-\end{figure}
 
@@
 This is contrary to the original classification of \textit{E.
-marmorata} \citep{Seeliger1945} MASTERS THESIS and lends credence to the idea that at least some aspect of cryptic diversity is a product of sample size, methodology, or both.
+marmorata} \citep{Seeliger1945,Holland1992} and lends credence to the idea that at least some aspect of cryptic diversity is a product of sample size, methodology, or both.
 
@@
 The observed plastral variation of \textit{E.
-marmorata} is most consistent with the mitochondrial based hypothesis of \citet{Spinks2005} and \citet{Spinks2010} and not with the original morphology based hypothesis of \citet{Seeliger1945}.
+marmorata} is most consistent with the mitochondrial based hypothesis of \citet{Spinks2005} and \citet{Spinks2010} and not with the original morphology based hypothesis of \citet{Seeliger1945,Holland1992}.
 We have also demonstrated the utility of various machine learning approaches to understanding the structure of variation in morphometric data.
@@
 print.xtable(mm.tabx,
+             hline.after = c(0, nrow(mm.tabx)),
              includ.rownames = TRUE,
@@
 Values of 0 correspond to p-values lower than 0.01.
-P-values were corrected for multiple comparison using the Holm method CITATION.}
+P-values were corrected for multiple comparison using the Holm method \citep{Holm1979}.}
   \label{tab:mm_test}
@@
 print.xtable(rf.tabx,
+             hline.after = c(0, nrow(rf.tabx)),
              includ.rownames = TRUE,
@@
 Values of 0 correspond to p-values lower than 0.01.
-P-values were corrected for multiple comparison using the Holm method CITATION.}
+P-values were corrected for multiple comparison using the Holm method \citep{Holm1979}.}
   \label{tab:rf_test}

### 2013-08-05 documents/turtle_paper.tex
completed draft for ken.

@@
 Depicted here are the for three PCs (lower triangle) and the mean shape with observed variance around each point (upper right).
-The first three PCs account for total 0 of the variance in plastral shape.}
+The first three PCs account for total 55.0075790871272\% of the variance in plastral shape.}
   \label{fig:pca}
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width = \textwidth]{figure/pc_var}
+  \caption{Landmark variation along the first two PCs of the Procrustes superimposed ``half'' plastral landmarks.
+The first row corresponds to variation along the first PC, while the second row corresponds to the second PC.
+The left most column represents the observation with the highest eigenscore along that PC, while the right most column represents the observation with the lowest eigenscore.
+The middle column, for both rows, is the mean plastral shape for all observations.
+The first PC represents 30.4118\% of the total variation in plastral shape while PC represents 14.926\% of the variance.}
+  \label{fig:pc_var}
 \end{figure}
@@
 Importantly, these clusters do not conform to the nothern and souther groups from the nuclear DNA hypothesis \citep{Spinks2010}.
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width = \textwidth]{figure/gap_res}
+  \caption{Gap statistic values for PAM clustering results for the \(\rho\) dissimliarity matrix of plastron shape. Error bars are standard errors estimated via 500 bootstrap samples.}
+  \label{fig:gap}
+\end{figure}
 
@@
 %   gap statistic results
-\begin{figure}[ht]
-  \centering
-  \includegraphics[width = \textwidth]{figure/gap_res}
-  \caption{Gap statistic values for PAM clustering results for the \(\rho\) dissimliarity matrix of plastron shape. Error bars are standard errors estimated via 500 bootstrap samples.}
-  \label{fig:gap}
-\end{figure}
 
@@
 % latex table generated in R 3.0.1 by xtable 1.7-1 package
-% Thu Aug  1 01:26:49 2013
+% Mon Aug  5 16:44:37 2013
 {\small
@@
 % latex table generated in R 3.0.1 by xtable 1.7-1 package
-% Thu Aug  1 01:26:49 2013
-\begin{tabular}{rrrrr}
-  \hline
+% Mon Aug  5 16:44:37 2013
+\begin{tabular}{r|rrrr}
   & morph 1 & morph 2 & molec 1 & molec 2 \\ 

### 2013-08-05 src/turtle_tables.r

@@ -38,6 +38,7 @@ mm.tab <- cbind(mm.tab, '4' = rep(NA, 4))
 colnames(mm.tab) <- coln
 rownames(mm.tab) <- rown
 mm.tabx <- xtable(as.table(mm.tab))
+align(mm.tabx) <- 'r|rrrr'
 
 rf.comp <- cbind(sh1 = rr$sh1$t, sh2 = rr$sh2$t, sh3 = rr$sh3$t, spinks = rr$spinks$t)
 rf.c <- melt(rf.comp)
@@ -49,3 +50,4 @@ rf.tab <- cbind(rf.tab, '4' = rep(NA, 4))
 colnames(rf.tab) <- coln
 rownames(rf.tab) <- rown
 rf.tabx <- xtable(as.table(rf.tab))
+align(rf.tabx) <- 'r|rrrr'

### 2013-08-07 documents/turtle_paper.Rnw
most of ken comments. found early error in mung file. need to fix.

@@
 \usepackage{amsmath, amsthm}
-\usepackage{graphicx, microtype, hyperref, authblk}
+\usepackage{graphicx, morefloats, microtype, hyperref, authblk}
 \usepackage{rotating, longtable, caption, subcaption, multirow}
@@
 % for american naturalist
-%\usepackage{lineno}
+\usepackage{lineno}
 
@@
 
-\title{How cryptic is cryptic diversity? Machine learning approaches to classifying morphological variation.}
+\title{How cryptic is cryptic diversity? Machine learning approaches to classifying morphological variation in \textit{Emys marmorata} (Testudinoidea, Emydidae).}
 \author[1]{Peter D Smits}%\thanks{psmits@uchicago.edu}}
-\author[2]{Kenneth D Angielczyk}%\thanks{kangielczyk@fieldmuseum.org}}
+\author[1,2]{Kenneth D Angielczyk}%\thanks{kangielczyk@fieldmuseum.org}}
 \author[3]{James F Parham}%\thanks{jparham@fullerton.edu}}
 \affil[1]{Committee on Evolutionary Biology, University of Chicago}
-\affil[2]{Department of Geology, Field Museum of Natural History}
+\affil[2]{Integrative Research Center, Field Museum of Natural History}
 \affil[3]{Department of Geological Sciences, California State University -- Fullerton}
@@
 
-%\linenumbers
-%\modulolinenumbers[2]
+\linenumbers
+\modulolinenumbers[2]
 
@@
 These methods provide different and unique advantages for understanding how to classify taxa, with what accuracy, and what these classifications are based on.
-Additionally, we investigate variation in continuous traits, and do not search for discrete differences between each class, instead focusing on suites of traits together.
+Additionally, we investigate variation in continuous traits, and do not search for discrete differences between each class, instead focusing on multivariate quantification of shape.
 
 % this paragraph kind of sucks
-The two major classes of machine learning methods, unsupervised and supervised, are essentially extensions of known statistical methods.
-Unsupervised learning methods are analogous to clustering and density estimation methods, while supervised learning methods are analogous classification and regression models.
-In both cases, many of these methods are not fit via maximum likelihood and are supplemented by randomization, sorting, and partitioning algorithms along with the maximization or minimization of summary statistics in order to best estimate a general model for all data, both sampled and unsampled.
+The two major classes of machine learning methods, unsupervised and supervised, are essentially extensions of known statistical methods \citep{Hastie2009}.
+Unsupervised learning methods are analogous to clustering and density estimation methods \citep{Kaufman1990}, while supervised learning methods are analogous classification and regression models \citep{Breiman1984}.
+In both cases, many of these methods are not fit via maximum likelihood and are supplemented by randomization, sorting, and partitioning algorithms along with the maximization or minimization of summary statistics in order to best estimate a general model for all data, both sampled and unsampled \citep{Hastie2009}.
 The application of the alternative approaches used in this study illustrates only a sampling of the various previously derived methods for clustering observations and fitting classification models.
+% Applications of these methods outside of this study
+% epidemeology, computer science, marketing
 Additionally, instead of pure classification accuracy, here we use a statistic of classification strength that reflects the rate at which taxa are both accurately and inaccurately classified (see Methods).
@@
 
-Previous work on morphological differentiation between different populations of \textit{E.
-marmorata} with no alternative subspecific assignments \citep{Seeliger1945,Holland1992}.
+Previous work on morphological differentiation between different populations of \textit{E. marmorata} with no alternative subspecific assignments \citep{Seeliger1945,Holland1992}. 
+%
 \citet{Holland1992} used comparisons of linear measures of the carapace of \textit{E.
@@
 Analyses were performed to determine how different, morphologically, different populations in three different regions of the species range.
-Additionally, the degree of sexual dimorphism present in \textit{E.
-marmorata}.
-The study concluded that distance was a poor indicator of morphological differentiation as opposed to barriers, such as different drainage basins, are probably more important barriers to reproduction.
+Additionally, sexual dimorphism present in \textit{E.
+marmorata} was only assesed in the context of size measures such as length of carapace and mass.
+In general, \citet{Holland1992} and others \citep{Lubcke2007,Germano2008} found that males of \textit{E.
+marmorata} are larger than females in terms of size.
+However, the effect of sexual dimorphism on shape, \textit{sensu} \citet{Kendall1977a} and \citet{Dryden1998a}, was not assessed \citep{Holland1992,Lubcke2007,Germano2008}.
+
+\citet{Holland1992} concluded that distance was a poor indicator of morphological differentiation as opposed to barriers, such as different drainage basins, are probably more important barriers to reproduction.
 This conclusion was later echoed by \citet{Spinks2005} via molecular phylogenetic analysis.
@@
 marmorata} museum specimens.
+These specimens include both newly sampled individuals and those sampled in previous studies of plastral shape variation \citep{Angielczyk2007,Angielczyk2011,Angielczyk2013a}.
 Specimen classification was based on known specimen geographic information which was recorded from museum collection information.
@@
 Because the specimens used to define the subclades in \citet{Spinks2005} and \citet{Spinks2010} were not available for study, all specimen classifications were based solely on this geographic information and not from explicit assignment in previous studies.
+The classifications here were based on matching museum locality data with the geographic boundaries of the moelcularly-defined clades of \citet{Spinks2005} and \citet{Spinks2010}.
 Because the exact barriers between different biogeographic regions are unknown and unclear, two assignments for both the morphologically and molecularly based hypotheses were used.
@@
 % landmarks
-Following \citet{Angielczyk2011}, 19 landmarks were digitized using TpsDig 2.04 \citep{Rohlf2005}.
-These landmarks were chosen to maximize the description of general plastral variation.
+Following previous work on plastral variation \citep{Angielczyk2007,Angielczyk2011,Angielczyk2013a}, 19 landmarks were digitized using TpsDig 2.04 \citep{Rohlf2005}.
+These landmarks were chosen to maximize the description of general plastral variation(Fig.
+\ref{fig:plastra}).
 17 of these landmarks are at the endpoints or intersection of the keratinous plastral scutes that cover the platron.
-12 of these landmarks were chosen to be symmetrical across the axis of symmetry and, in order to prevent degrees of freedom and other concerns \citep{Klingenberg2007}, prior to analysis these landmarks were reflected across the axis of symmetry (i.e.
+12 of these landmarks were chosen to be symmetrical across the axis of symmetry and, in order to prevent degrees of freedom and other concerns \citep{Klingenberg2002}, prior to analysis these landmarks were reflected across the axis of symmetry (i.e.
 midline) and the average position of each symmetrical pair was used.
@@
 \[W_{k} = \sum^{k}_{r = 1}{\frac{1}{2n_{r}} (\sum_{i,i' \in C_{r}} d_{ii'})}\].
-\(d_{ii'}\) is the dispersion of the clustering solution or the sum of the pairwise dissimilarities between observations in each cluster and there respective mediods (\(C\)) for all clusters \(r\).
+\(d_{ii'}\) is the dispersion of the clustering solution or the sum of the pairwise dissimilarities between observations in each cluster and their respective mediods (\(C\)) for all clusters \(r\).
 This value is averaged and compared to the expected dispersion (\(E^{*}_{n}\)) of a sample \(n\) from a reference distribution.
-In this case, the reference distribution was estimated from \Sexpr{tadult.gap$B} resamples of the dataset taking into account the original structure of the data.
+In this case, the reference distribution was estimated from \Sexpr{tadult.gap$B} resamples of the dataset while maintaining the original dispersion of the data.
 %  R implementation
@@
 performance in the wild).
+This split was chosen to allow for a large enough sample size for model fitting while also providing a large enough testing dataset to determine any systematic misclassifcations.
 
@@
 Instead, a recursive feature selection algorithm was used to choose the optimal number of PCs to include based on the AUC of the model.
-Following the backwards selection algorithm implemented in \texttt{caret} \citep{Kuhn2013}, the maximum number of features were included in the initial model, there importance ranked, and the AUC of the model calculated.
+Following the backwards selection algorithm implemented in \texttt{caret} \citep{Kuhn2013}, the maximum number of features were included in the initial model, their importance ranked, and the AUC of the model calculated.
 The lowest ranked feature was then removed, and the AUC of the model recalculated.
@@
 However, it is important to note that Procustes based superimposition attempts to evenly distribute variance around the mean landmark \citep{Zelditch2004} and this observation should be considered cursory at best.
+
 The first PC appears to describe variation in lateral plastral shape, from a very pointed medial edge to a much more rounded and blunt edge (Fig.
 \ref{fig:pc_var}).
+Landmark 10 (Fig.
+\ref{fig:plastra}), which appears to be the most variable along this axis (Fig.
+\ref{fig:pca} and \ref{fig:pc_var}), is positioned on the bridge between the plastron and the carapace.
+Over ontogeny, this is an area that deepens dorsoventrally and when the plastron was projected into two dimensions this it created the effect of mediolateral movement.
 Variation along the second PC appears to be characterized by variation along the midline, especially in terms of the ``straightness'' of the individual landmarks along it.
+This is most likely caused by the fact that plastral scutes do not always line up perfectly along the midline.
 
-\begin{figure}[ht]
-  \centering
-  \includegraphics[width = \textwidth]{figure/pca_res}
-  \caption{Results from PCA of the Procrustes superimposed ``half'' plastral landmarks.
-Depicted here are the for three PCs (lower triangle) and the mean shape with observed variance around each point (upper right).
-The first three PCs account for total \Sexpr{paste0(sum(turtle.fit$percent[1:3]))}\% of the variance in plastral shape.}
-  \label{fig:pca}
-\end{figure}
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics[width = \textwidth]{figure/pc_var}
-  \caption{Landmark variation along the first two PCs of the Procrustes superimposed ``half'' plastral landmarks.
-The first row corresponds to variation along the first PC, while the second row corresponds to the second PC.
-The left most column represents the observation with the highest eigenscore along that PC, while the right most column represents the observation with the lowest eigenscore.
-The middle column, for both rows, is the mean plastral shape for all observations.
-The first PC represents \Sexpr{turtle.fit$percent[1]}\% of the total variation in plastral shape while PC represents \Sexpr{turtle.fit$percent[2]}\% of the variance.}
-  \label{fig:pc_var}
-\end{figure}
+% size correlation with either of the first two PCs?
 
@@
 \ref{fig:gap_map}).
-Importantly, these clusters do not conform to the nothern and souther groups from the nuclear DNA hypothesis \citep{Spinks2010}.
+Importantly, these clusters do not conform to the northern and southern groups from the nuclear DNA hypothesis \citep{Spinks2010}.
 
-\begin{figure}[ht]
-  \centering
-  \includegraphics[width = \textwidth]{figure/gap_res}
-  \caption{Gap statistic values for PAM clustering results for the \(\rho\) dissimliarity matrix of plastron shape.
-Error bars are standard errors estimated via \Sexpr{tadult.gap$B} bootstrap samples.}
-  \label{fig:gap}
-\end{figure}
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics{figure/gap_map}
-  \caption{Clustering solution for PAM with two mediods for the entire set of observed \textit{E.
-marmorata}.
-Clustering was based entirely on the \(\rho\) dissimilarity matrix of ``half'' plastral landmark configurations following Procrustes superimposition.}
-  \label{fig:gap_map}
-\end{figure}
-
+% some sex information. Analyze it.
 Unfortunately, it was not possible to obtain enough detailed information on the sex of each \textit{E.
@@
 
+The results from the generalization of the selected supervised learning models, measured by the distributions of the bootstrapped AUC values of the testing dataset, show that the second molecular classification hypotheses was the best overall classification scheme (Fig.
+\ref{fig:gen_res}).
+For both methods, the distribution of bootstrapped AUC for the molecular hypothesis was significantly greater than all of the other classification schemes (Tables \ref{tab:mm_test} and \ref{tab:rf_test}).
+Remarkably, the best classification hypothesis was identical based on both the multinomial logistic regression and random forest models.
+
+When the classification results of the training set for the best classification scheme based on the generalization results are compared with the references classes, the higher AUC value of the best multinomial logistic regression model compared to the best random forest model can be observed as the classifications are much closer to the reference classes (Fig.
+\ref{fig:gen_map}).
+The best random forest model misclassified many of the observations as the northern clade instead of the correct class.
+This pattern of misclassification is observable but not as exaggerated in the classifications of the multinomial logistic regression model.
+
+This pattern of misclassification may have been caused by the subtle differences in mean shape between each of the different classes (Fig.
+\ref{fig:mean_shape}).
+The mean shape of the northern clade is the most similar to the mean shape of the entire dataset (Fig.
+\ref{fig:mean_shape1}), which may indicate that specimens that are closer to the mean shape will be systematically misclassified as the northern clade.
+
+% variation along the most important axes
+<<>>=
+rf.w <- gsub('PC', x = trf.a$spinks$optVariables, replacement = '')
+@
+The results of fitting the final random forest model also include the variable importance for best separating the different classes.
+The selected random forest model for the best classification scheme had \Sexpr{trf.a$spinks$optsize} PCs as features.
+The PCs included as features in the final random forest model, in descending order of importance, were PCs \Sexpr{paste(rf.w[1:6])} and \Sexpr{paste(rf.w[7])}.
+Of these \Sexpr{trf.a$spinks$optsize} features, the first three are illustrated here (Fig.
+\ref{fig:imp_pc}) in descending order of importance.
+
+The first two most important features describe different aspects of variation (Fig.
+\ref{fig:imp_var}).
+The third and most important PC describes variation in the relative position of landmarks on anterior and posterior portions of the plastron and represents \Sexpr{turtle.fit$percent[3]}\% of total variation.
+The eighth and second most important PC mostly describes variation in landmarks along the midline of the plastron and represents \Sexpr{turtle.fit$percent[8]}\% of total variation.
+The major variations along these axes correspond well to the differences between the mean shape of each class (Fig.
+\ref{fig:mean_shape}) where major class differences seem based on the relative ballooning or shrinking of the anterior and posterior portions of the plastron together.
+
+The relative risk values for classification from the multinomial logistic regression model, based on the three most important PCs, demonstrate that while an individual PC might not be sufficient to differentiate any individual class from the northern group, given multiple features the odds of determining the correct classification increase (Fig.
+\ref{fig:rel_risk}).
+Changes along the second most important axes contribute very little to increasing the odds of classification.
+This is observable from the class histograms of PC 8 (Fig.
+\ref{fig:imp_var}).
+Changes along the first and third most important axes contribute more obviously to increasing the odds of correctly identifying the class of an observation, a result that is observable in both the relative risk (Fig.
+\ref{fig:rel_risk}) and the different class histograms of the PCs (Fig.
+\ref{fig:imp_var}).
+
+\section{Discussion}
+% discussion of results
+%   remarkable concordance in the supervised learning methods
+The results of this study support the mitochrondial based classification hypothesis of \textit{E.
+marmorata} \citep{Spinks2005,Spinks2010}.
+This is contrary to the original classification of \textit{E.
+marmorata} \citep{Seeliger1945,Holland1992} and lends credence to the idea that at least some aspect of cryptic diversity is a product of sample size, methodology, or both.
+
+%   what does this mean about cryptic diversity?
+%   unsupervised learning shortcomings
+
+The lack of coherent geographical subclass assignment from PAM clustering (Fig.
+\ref{fig:gap}) as well as the large number of features necessary before no increase in AUC for all models (Fig.
+\ref{fig:roc}) indicates that the morphological variation between classes is extremely fine grained.
+This was also exemplified by the small differences between mean class shapes of the final chosen classification scheme (Fig.
+\ref{fig:imp_var}).
+
+% methodological concerns
+%   compromise in the supervised learning models
+%     fair because as much variation as ``necessary'' is used
+The approaches presented here for supervised learning analysis of the landmark variation represent a compromise between explicitly modeling all shape variation and preventing models from being overfit and ungeneralizable.
+While all aspects of shape may be evolving simultaneously, and not along individual PCs, including all shape variation in each model might increase model complexity beyond a reasonable level for the sample size and possibly the necessary complexity to accurately model the response.
+Additionally, because only individual PCs are used as features in the models, this does not accurately represent shape evolution and how exactly different classes might be evolving in relation to each other.
+However, this compromise is not without its advantages.
+Because both AICc and AUC values improved rapidly with increased model complexity (Fig.
+\ref{fig:roc}), this helped demonstrate how fine scale the actual variation between classes was.
+Additionally, the relative risk values from the mulitinomial logistic regression models demonstrate that a single PC is probably not sufficient for estimating the class of an observation, but that given a set of PCs this classification would be more accurate (Fig.
+\ref{fig:rel_risk}).
+The results of the relative risk values also shows that while a variable can be considered important for the random forest models (Fig.
+\ref{fig:imp_pc}, \ref{fig:imp_var}) they each might not be considered useful on their own (Fig.
+\ref{fig:rel_risk}).
+
+%   unsupervised learning
+Ultimately, it would be useful to not require such explicit classification hypotheses, especially when concerned about possible cryptic variation in extinct taxa.
+The only unsupervised method employed in this study, PAM, is rather simple and not model based.
+A more useful approach would be to employ various model based clustering approaches \citep{Fraley2002,Zhong2003}.
+In this manner, a series of candidate models can be compared via model comparison methods, such as AIC or Bayes factors \citep{Fraley2002}, in order to asses the best clustering solution.
+% beat complaints from reviewers
+Here we focused on the results and utility of supervised methods because they are both more powerful and hypothesis driven \citep{Hastie2009}.
+Because there are two alternative classification schemes for \textit{E.
+marmorata}, it was most appropriate to compare these two hypotheses and estimate which one most accurately reflected the variation.
+Future work would be to explore and derive unsupervised methods which corroborate these results.
+
+% closing statements
+In this study we have demonstrated that, using alternative methodology to that which is most frequently applied, it is possible to determine which classification scheme best matches variation in a taxon amongst a set of alternative hypotheses.
+The observed plastral variation of \textit{E.
+marmorata} is most consistent with the mitochondrial based hypothesis of \citet{Spinks2005} and \citet{Spinks2010} and not with the original morphology based hypothesis of \citet{Seeliger1945,Holland1992}.
+We have also demonstrated the utility of various machine learning approaches to understanding the structure of variation in morphometric data.
+Specifically, methods for better understanding misclassification and identifying which is the most important for delimiting different classes.
+These methods represent new applications which may be important for future studies on class-based morphological comparison and variation, both in the context of cryptic diversity and with known classifications.
+
+\section{Acknowledgements}
+PDS would like to thank David Bapst, Michael Foote, Benjamin Frable, and Dallas Krentzel for useful discussion which enhanced the quality of this study.
+For access to emydine specimens, we thank: J.
+Vindum and R.
+Drewes (CAS); A.
+Resetar (FMNH); R.
+Feeney (LACM); C.
+Austin (LSUMNS); S.
+Sweet (MSE); J.
+McGuire and C.
+Conroy (MVZ); A.
+Wynn (NMNH); P.
+Collins (SBMNH); B.
+Hollingsworth (SDMNH); C.
+Bell and R.
+Burroughs (TMM); T.
+LaDuc and R.
+Burroughs (TNHC); P.
+Holroyd (UCMP); R.
+Symonds (UMZC); J.
+Buskirk.
+We are greatful to S.
+Sweet for field assistance and the California Department of Fish and Game for permits.
+Much of the data collection was funded by NSD DBI-0306158 (to KDA).
+
+\bibliographystyle{sysbio}
+\bibliography{turtle,packages}
+
+\pagebreak
+
+% figures
+\begin{figure}[ht]
+  \centering
+  \includegraphics{figure/plastra}
+  \caption{Depiction of general plastral shape of \textit{E. marmorata} and position of the 19 landmark used in this study. Anterior is towards the top of the figure.}
+  \label{fig:plastra}
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width = \textwidth]{figure/pca_res}
+  \caption{Results from PCA of the Procrustes superimposed ``half'' plastral landmarks.
+Depicted here are the for three PCs (lower triangle) and the mean shape with observed variance around each point (upper right).
+The first three PCs account for total \Sexpr{paste0(sum(turtle.fit$percent[1:3]))}\% of the variance in plastral shape.}
+  \label{fig:pca}
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width = \textwidth]{figure/pc_var}
+  \caption{Landmark variation along the first two PCs of the Procrustes superimposed ``half'' plastral landmarks.
+The first row corresponds to variation along the first PC, while the second row corresponds to the second PC.
+The left most column represents the observation with the highest eigenscore along that PC, while the right most column represents the observation with the lowest eigenscore.
+The middle column, for both rows, is the mean plastral shape for all observations.
+The first PC represents \Sexpr{turtle.fit$percent[1]}\% of the total variation in plastral shape while PC represents \Sexpr{turtle.fit$percent[2]}\% of the variance.}
+  \label{fig:pc_var}
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width = \textwidth]{figure/gap_res}
+  \caption{Gap statistic values for PAM clustering results for the \(\rho\) dissimliarity matrix of plastron shape.
+Error bars are standard errors estimated via \Sexpr{tadult.gap$B} bootstrap samples.}
+  \label{fig:gap}
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics{figure/gap_map}
+  \caption{Clustering solution for PAM with two mediods for the entire set of observed \textit{E.
+marmorata}.
+Clustering was based entirely on the \(\rho\) dissimilarity matrix of ``half'' plastral landmark configurations following Procrustes superimposition.}
+  \label{fig:gap_map}
+\end{figure}
+
 \begin{figure}[ht]
@@
 \end{figure}
-
-The results from the generalization of the selected supervised learning models, measured by the distributions of the bootstrapped AUC values of the testing dataset, show that one of the molecular classification hypotheses was the best overall classification scheme (Fig.
-\ref{fig:gen_res}).
-For both methods, the distribution of bootstrapped AUC for the molecular hypothesis was significantly greater than all of the other classification schemes (Tables \ref{tab:mm_test} and \ref{tab:rf_test}).
-Remarkably, the best classification hypothesis was identical based on both the multinomial logistic regression and random forest models.
 
@@
 
-When the classification results of the training set for the best classification scheme based on the generalization results are compared with the references classes, the higher AUC value of the best multinomial logistic regression model compared to the best random forest model can be observed as the classifications are much closer to the reference classes (Fig.
-\ref{fig:gen_map}).
-The best random forest model misclassified many of the observations as the northern clade instead of the correct class.
-This pattern of misclassification is observable but not as exaggerated in the classifications of the multinomial logistic regression model.
-
 \begin{figure}[ht]
@@
 \end{figure}
-
-This pattern of misclassification may have been caused by the subtle differences in mean shape between each of the different classes (Fig.
-\ref{fig:mean_shape}).
-The mean shape of the northern clade is the most similar to the mean shape of the entire dataset (Fig.
-\ref{fig:mean_shape1}), which may indicate that specimens that are closer to the mean shape will be systematically misclassified as the northern clade.
 
@@
 
-% variation along the most important axes
-<<>>=
-rf.w <- gsub('PC', x = trf.a$spinks$optVariables, replacement = '')
-@
-The results of fitting the final random forest model also include the variable importance for best separating the different classes.
-The selected random forest model for the best classification scheme had \Sexpr{trf.a$spinks$optsize} PCs as features.
-The PCs included as features in the final random forest model, in descending order of importance, were PCs \Sexpr{paste(rf.w[1:6])} and \Sexpr{paste(rf.w[7])}.
-Of these \Sexpr{trf.a$spinks$optsize} features, the first three are illustrated here (Fig.
-\ref{fig:imp_pc}) in descending order of importance.
-
 \begin{figure}[ht]
@@
 \end{figure}
-% need to add in variance percentages on the different axes.
-
-The first two most important features describe different aspects of variation (Fig.
-\ref{fig:imp_var}).
-The third and most important PC describes variation in the relative position of landmarks on anterior and posterior portions of the plastron and represents \Sexpr{turtle.fit$percent[3]}\% of total variation.
-The eighth and second most important PC mostly describes variation in landmarks along the midline of the plastron and represents \Sexpr{turtle.fit$percent[8]}\% of total variation.
-The major variations along these axes correspond well to the differences between the mean shape of each class (Fig.
-\ref{fig:mean_shape}) where major class differences seem based on the relative ballooning or shrinking of the anterior and posterior portions of the plastron together.
 
@@
 \end{figure}
-
-%RELATIVE RISK FROM THE MULTINOMIAL LOGISTIC REGRESSION MODELS
-The relative risk values for classification from the multinomial logistic regression model, based on the three most important PCs, demonstrate that while an individual PC might not be sufficient to differentiate any individual class from the northern group, given multiple features the odds of determining the correct classification increase (Fig.
-\ref{fig:rel_risk}).
-Changes along the second most important axes contribute very little to increasing the odds of classification.
-This is observable from the class histograms of PC 8 (Fig.
-\ref{fig:imp_var}).
-Changes along the first and third most important axes contribute more obviously to increasing the odds of correctly identifying the class of an observation, a result that is observable in both the relative risk (Fig.
-\ref{fig:rel_risk}) and the different class histograms of the PCs (Fig.
-\ref{fig:imp_var}).
 
@@
 
-\section{Discussion}
-% discussion of results
-%   remarkable concordance in the supervised learning methods
-The results of this study support the mitochrondial based classification hypothesis of \textit{E.
-marmorata} \citep{Spinks2005,Spinks2010}.
-This is contrary to the original classification of \textit{E.
-marmorata} \citep{Seeliger1945,Holland1992} and lends credence to the idea that at least some aspect of cryptic diversity is a product of sample size, methodology, or both.
-
-%   what does this mean about cryptic diversity?
-%   unsupervised learning shortcomings
-
-The lack of coherent geographical subclass assignment from PAM clustering (Fig.
-\ref{fig:gap}) as well as the large number of features necessary before no increase in AUC for all models (Fig.
-\ref{fig:roc}) indicates that the morphological variation between classes is extremely fine grained.
-This was also exemplified by the small differences between mean class shapes of the final chosen classification scheme (Fig.
-\ref{fig:imp_var}).
-
-% methodological concerns
-%   compromise in the supervised learning models
-%     fair because as much variation as ``necessary'' is used
-The approaches presented here for supervised learning analysis of the landmark variation represent a compromise between explicitly modeling all shape variation and preventing models from being overfit and ungeneralizable.
-While all aspects of shape may be evolving simultaneously, and not along individual PCs, including all shape variation in each model might increase model complexity beyond a reasonable level for the sample size and possibly the necessary complexity to accurately model the response.
-Additionally, because only individual PCs are used as features in the models, this does not accurately represent shape evolution and how exactly different classes might be evolving in relation to each other.
-However, this compromise is not without its advantages.
-Because both AICc and AUC values improved rapidly with increased model complexity (Fig.
-\ref{fig:roc}), this helped demonstrate how fine scale the actual variation between classes was.
-Additionally, the relative risk values from the mulitinomial logistic regression models demonstrate that a single PC is probably not sufficient for estimating the class of an observation, but that given a set of PCs this classification would be more accurate (Fig.
-\ref{fig:rel_risk}).
-The results of the relative risk values also shows that while a variable can be considered important for the random forest models (Fig.
-\ref{fig:imp_pc}, \ref{fig:imp_var}) they each might not be considered useful on their own (Fig.
-\ref{fig:rel_risk}).
-
-%   unsupervised learning
-%     model based clustering
-%     nonparametric bayesian approach
-%     anything about semisupervised clustering? probably not
-Ultimately, it would be useful to not require such explicit classification hypotheses, especially when concerned about possible cryptic variation in extinct taxa.
-The only unsupervised method employed in this study, PAM, is rather simple and not model based.
-A more useful approach would be to employ various model based clustering approaches \citep{Fraley2002,Zhong2003}.
-In this manner, a series of candidate models can be compared via model comparison methods, such as AIC or Bayes factors \citep{Fraley2002}, in order to asses the best clustering solution.
-
-% future directions
-%   potentially useful for phylogenetic comparative method stuff
-%   conservation utility
-%   fossil diversity
-%     current approach favors having an extremely explicit classification hypothesis
-%     as indicated about, nonparametric bayes and other model based clustering can help
-
-% closing statements
-In this study we have demonstrated that, using alternative methodology to that which is most frequently applied, it is possible to determine which classification scheme best matches variation in a taxon amongst a set of alternative hypotheses.
-The observed plastral variation of \textit{E.
-marmorata} is most consistent with the mitochondrial based hypothesis of \citet{Spinks2005} and \citet{Spinks2010} and not with the original morphology based hypothesis of \citet{Seeliger1945,Holland1992}.
-We have also demonstrated the utility of various machine learning approaches to understanding the structure of variation in morphometric data.
-Specifically, methods for better understanding misclassification and identifying which is the most important for delimiting different classes.
-These methods represent new applications which may be important for future studies on class-based morphological comparison and variation, both in the context of cryptic diversity and with known classifications.
-
-\section{Acknowledgements}
-PDS would like to thank David Bapst, Michael Foote, Benjamin Frable, and Dallas Krentzel for useful discussion which enhanced the quality of this study. MUSEUMS. GRANTING AGENCIES.
-
-\bibliographystyle{sysbio}
-\bibliography{turtle,packages}
-
 \pagebreak
-
+% tables
 \begin{table}

### 2013-08-07 documents/turtle_paper.tex
most of ken comments. found early error in mung file. need to fix.

@@
 \[W_{k} = \sum^{k}_{r = 1}{\frac{1}{2n_{r}} (\sum_{i,i' \in C_{r}} d_{ii'})}\].
-\(d_{ii'}\) is the dispersion of the clustering solution or the sum of the pairwise dissimilarities between observations in each cluster and there respective mediods (\(C\)) for all clusters \(r\).
+\(d_{ii'}\) is the dispersion of the clustering solution or the sum of the pairwise dissimilarities between observations in each cluster and their respective mediods (\(C\)) for all clusters \(r\).
 This value is averaged and compared to the expected dispersion (\(E^{*}_{n}\)) of a sample \(n\) from a reference distribution.
-In this case, the reference distribution was estimated from 500 resamples of the dataset taking into account the original structure of the data.
+In this case, the reference distribution was estimated from 500 resamples of the dataset while maintaining the original dispersion of the data.
 %  R implementation
@@
 \ref{fig:pc_var}).
+Landmark 10 (Fig.
+\ref{fig:plastra}), which appears to be the most variable along this axis (Fig.
+\ref{fig:pca} and \ref{fig:pc_var}), is positioned on the bridge between the plastron and the carapace.
+Over ontogeny, this is an area that deepens dorsoventrally and when the plastron was projected into two dimensions this it created the effect of mediolateral movement.
 Variation along the second PC appears to be characterized by variation along the midline, especially in terms of the ``straightness'' of the individual landmarks along it.
+This is most likely caused by the fact that plastral scutes do not always line up perfectly along the midline.
 
-\begin{figure}[ht]
-  \centering
-  \includegraphics[width = \textwidth]{figure/pca_res}
-  \caption{Results from PCA of the Procrustes superimposed ``half'' plastral landmarks.
-Depicted here are the for three PCs (lower triangle) and the mean shape with observed variance around each point (upper right).
-The first three PCs account for total 55.0075790871272\% of the variance in plastral shape.}
-  \label{fig:pca}
-\end{figure}
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics[width = \textwidth]{figure/pc_var}
-  \caption{Landmark variation along the first two PCs of the Procrustes superimposed ``half'' plastral landmarks.
-The first row corresponds to variation along the first PC, while the second row corresponds to the second PC.
-The left most column represents the observation with the highest eigenscore along that PC, while the right most column represents the observation with the lowest eigenscore.
-The middle column, for both rows, is the mean plastral shape for all observations.
-The first PC represents 30.4118\% of the total variation in plastral shape while PC represents 14.926\% of the variance.}
-  \label{fig:pc_var}
-\end{figure}
+% size correlation with either of the first two PCs?
 
@@
 \ref{fig:gap_map}).
-Importantly, these clusters do not conform to the nothern and souther groups from the nuclear DNA hypothesis \citep{Spinks2010}.
+Importantly, these clusters do not conform to the northern and southern groups from the nuclear DNA hypothesis \citep{Spinks2010}.
 
-\begin{figure}[ht]
-  \centering
-  \includegraphics[width = \textwidth]{figure/gap_res}
-  \caption{Gap statistic values for PAM clustering results for the \(\rho\) dissimliarity matrix of plastron shape. Error bars are standard errors estimated via 500 bootstrap samples.}
-  \label{fig:gap}
-\end{figure}
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics{figure/gap_map}
-  \caption{Clustering solution for PAM with two mediods for the entire set of observed \textit{E.
-marmorata}.
-Clustering was based entirely on the \(\rho\) dissimilarity matrix of ``half'' plastral landmark configurations following Procrustes superimposition.}
-  \label{fig:gap_map}
-\end{figure}
-
+% some sex information. Analyze it.
 Unfortunately, it was not possible to obtain enough detailed information on the sex of each \textit{E.
@@
 
+The results from the generalization of the selected supervised learning models, measured by the distributions of the bootstrapped AUC values of the testing dataset, show that the second molecular classification hypotheses was the best overall classification scheme (Fig.
+\ref{fig:gen_res}).
+For both methods, the distribution of bootstrapped AUC for the molecular hypothesis was significantly greater than all of the other classification schemes (Tables \ref{tab:mm_test} and \ref{tab:rf_test}).
+Remarkably, the best classification hypothesis was identical based on both the multinomial logistic regression and random forest models.
+
+When the classification results of the training set for the best classification scheme based on the generalization results are compared with the references classes, the higher AUC value of the best multinomial logistic regression model compared to the best random forest model can be observed as the classifications are much closer to the reference classes (Fig.
+\ref{fig:gen_map}).
+The best random forest model misclassified many of the observations as the northern clade instead of the correct class.
+This pattern of misclassification is observable but not as exaggerated in the classifications of the multinomial logistic regression model.
+
+This pattern of misclassification may have been caused by the subtle differences in mean shape between each of the different classes (Fig.
+\ref{fig:mean_shape}).
+The mean shape of the northern clade is the most similar to the mean shape of the entire dataset (Fig.
+\ref{fig:mean_shape1}), which may indicate that specimens that are closer to the mean shape will be systematically misclassified as the northern clade.
+
+% variation along the most important axes
+
+The results of fitting the final random forest model also include the variable importance for best separating the different classes.
+The selected random forest model for the best classification scheme had 7 PCs as features.
+The PCs included as features in the final random forest model, in descending order of importance, were PCs 3, 8, 6, 1, 4, 2 and 10.
+Of these 7 features, the first three are illustrated here (Fig.
+\ref{fig:imp_pc}) in descending order of importance.
+
+The first two most important features describe different aspects of variation (Fig.
+\ref{fig:imp_var}).
+The third and most important PC describes variation in the relative position of landmarks on anterior and posterior portions of the plastron and represents 9.6697\% of total variation.
+The eighth and second most important PC mostly describes variation in landmarks along the midline of the plastron and represents 4.1061\% of total variation.
+The major variations along these axes correspond well to the differences between the mean shape of each class (Fig.
+\ref{fig:mean_shape}) where major class differences seem based on the relative ballooning or shrinking of the anterior and posterior portions of the plastron together.
+
+The relative risk values for classification from the multinomial logistic regression model, based on the three most important PCs, demonstrate that while an individual PC might not be sufficient to differentiate any individual class from the northern group, given multiple features the odds of determining the correct classification increase (Fig.
+\ref{fig:rel_risk}).
+Changes along the second most important axes contribute very little to increasing the odds of classification.
+This is observable from the class histograms of PC 8 (Fig.
+\ref{fig:imp_var}).
+Changes along the first and third most important axes contribute more obviously to increasing the odds of correctly identifying the class of an observation, a result that is observable in both the relative risk (Fig.
+\ref{fig:rel_risk}) and the different class histograms of the PCs (Fig.
+\ref{fig:imp_var}).
+
+\section{Discussion}
+% discussion of results
+%   remarkable concordance in the supervised learning methods
+The results of this study support the mitochrondial based classification hypothesis of \textit{E.
+marmorata} \citep{Spinks2005,Spinks2010}.
+This is contrary to the original classification of \textit{E.
+marmorata} \citep{Seeliger1945,Holland1992} and lends credence to the idea that at least some aspect of cryptic diversity is a product of sample size, methodology, or both.
+
+%   what does this mean about cryptic diversity?
+%   unsupervised learning shortcomings
+
+The lack of coherent geographical subclass assignment from PAM clustering (Fig.
+\ref{fig:gap}) as well as the large number of features necessary before no increase in AUC for all models (Fig.
+\ref{fig:roc}) indicates that the morphological variation between classes is extremely fine grained.
+This was also exemplified by the small differences between mean class shapes of the final chosen classification scheme (Fig.
+\ref{fig:imp_var}).
+
+% methodological concerns
+%   compromise in the supervised learning models
+%     fair because as much variation as ``necessary'' is used
+The approaches presented here for supervised learning analysis of the landmark variation represent a compromise between explicitly modeling all shape variation and preventing models from being overfit and ungeneralizable.
+While all aspects of shape may be evolving simultaneously, and not along individual PCs, including all shape variation in each model might increase model complexity beyond a reasonable level for the sample size and possibly the necessary complexity to accurately model the response.
+Additionally, because only individual PCs are used as features in the models, this does not accurately represent shape evolution and how exactly different classes might be evolving in relation to each other.
+However, this compromise is not without its advantages.
+Because both AICc and AUC values improved rapidly with increased model complexity (Fig.
+\ref{fig:roc}), this helped demonstrate how fine scale the actual variation between classes was.
+Additionally, the relative risk values from the mulitinomial logistic regression models demonstrate that a single PC is probably not sufficient for estimating the class of an observation, but that given a set of PCs this classification would be more accurate (Fig.
+\ref{fig:rel_risk}).
+The results of the relative risk values also shows that while a variable can be considered important for the random forest models (Fig.
+\ref{fig:imp_pc}, \ref{fig:imp_var}) they each might not be considered useful on their own (Fig.
+\ref{fig:rel_risk}).
+
+%   unsupervised learning
+Ultimately, it would be useful to not require such explicit classification hypotheses, especially when concerned about possible cryptic variation in extinct taxa.
+The only unsupervised method employed in this study, PAM, is rather simple and not model based.
+A more useful approach would be to employ various model based clustering approaches \citep{Fraley2002,Zhong2003}.
+In this manner, a series of candidate models can be compared via model comparison methods, such as AIC or Bayes factors \citep{Fraley2002}, in order to asses the best clustering solution.
+% beat complaints from reviewers
+Here we focused on the results and utility of supervised methods because they are both more powerful and hypothesis driven \citep{Hastie2009}.
+Because there are two alternative classification schemes for \textit{E.
+marmorata}, it was most appropriate to compare these two hypotheses and estimate which one most accurately reflected the variation.
+Future work would be to explore and derive unsupervised methods which corroborate these results.
+
+% closing statements
+In this study we have demonstrated that, using alternative methodology to that which is most frequently applied, it is possible to determine which classification scheme best matches variation in a taxon amongst a set of alternative hypotheses.
+The observed plastral variation of \textit{E.
+marmorata} is most consistent with the mitochondrial based hypothesis of \citet{Spinks2005} and \citet{Spinks2010} and not with the original morphology based hypothesis of \citet{Seeliger1945,Holland1992}.
+We have also demonstrated the utility of various machine learning approaches to understanding the structure of variation in morphometric data.
+Specifically, methods for better understanding misclassification and identifying which is the most important for delimiting different classes.
+These methods represent new applications which may be important for future studies on class-based morphological comparison and variation, both in the context of cryptic diversity and with known classifications.
+
+\section{Acknowledgements}
+PDS would like to thank David Bapst, Michael Foote, Benjamin Frable, and Dallas Krentzel for useful discussion which enhanced the quality of this study.
+For access to emydine specimens, we thank: J.
+Vindum and R.
+Drewes (CAS); A.
+Resetar (FMNH); R.
+Feeney (LACM); C.
+Austin (LSUMNS); S.
+Sweet (MSE); J.
+McGuire and C.
+Conroy (MVZ); A.
+Wynn (NMNH); P.
+Collins (SBMNH); B.
+Hollingsworth (SDMNH); C.
+Bell and R.
+Burroughs (TMM); T.
+LaDuc and R.
+Burroughs (TNHC); P.
+Holroyd (UCMP); R.
+Symonds (UMZC); J.
+Buskirk.
+We are greatful to S.
+Sweet for field assistance and the California Department of Fish and Game for permits.
+Much of the data collection was funded by NSD DBI-0306158 (to KDA).
+
+\bibliographystyle{sysbio}
+\bibliography{turtle,packages}
+
+\pagebreak
+
+% figures
+\begin{figure}[ht]
+  \centering
+  \includegraphics{figure/plastra}
+  \caption{Depiction of general plastral shape of \textit{E. marmorata} and position of the 19 landmark used in this study. Anterior is towards the top of the figure.}
+  \label{fig:plastra}
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width = \textwidth]{figure/pca_res}
+  \caption{Results from PCA of the Procrustes superimposed ``half'' plastral landmarks.
+Depicted here are the for three PCs (lower triangle) and the mean shape with observed variance around each point (upper right).
+The first three PCs account for total 55.0075790871272\% of the variance in plastral shape.}
+  \label{fig:pca}
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width = \textwidth]{figure/pc_var}
+  \caption{Landmark variation along the first two PCs of the Procrustes superimposed ``half'' plastral landmarks.
+The first row corresponds to variation along the first PC, while the second row corresponds to the second PC.
+The left most column represents the observation with the highest eigenscore along that PC, while the right most column represents the observation with the lowest eigenscore.
+The middle column, for both rows, is the mean plastral shape for all observations.
+The first PC represents 30.4118\% of the total variation in plastral shape while PC represents 14.926\% of the variance.}
+  \label{fig:pc_var}
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width = \textwidth]{figure/gap_res}
+  \caption{Gap statistic values for PAM clustering results for the \(\rho\) dissimliarity matrix of plastron shape. Error bars are standard errors estimated via 500 bootstrap samples.}
+  \label{fig:gap}
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics{figure/gap_map}
+  \caption{Clustering solution for PAM with two mediods for the entire set of observed \textit{E.
+marmorata}.
+Clustering was based entirely on the \(\rho\) dissimilarity matrix of ``half'' plastral landmark configurations following Procrustes superimposition.}
+  \label{fig:gap_map}
+\end{figure}
+
 \begin{figure}[ht]
@@
 
-% variation along the most important axes
-
-The results of fitting the final random forest model also include the variable importance for best separating the different classes.
-The selected random forest model for the best classification scheme had 7 PCs as features.
-The PCs included as features in the final random forest model, in descending order of importance, were PCs 3, 8, 6, 1, 4, 2 and 10.
-Of these 7 features, the first three are illustrated here (Fig.
-\ref{fig:imp_pc}) in descending order of importance.
-
 \begin{figure}[ht]
@@
 \end{figure}
-% need to add in variance percentages on the different axes.
-
-The first two most important features describe different aspects of variation (Fig.
-\ref{fig:imp_var}).
-The third and most important PC describes variation in the relative position of landmarks on anterior and posterior portions of the plastron and represents 9.6697\% of total variation.
-The eighth and second most important PC mostly describes variation in landmarks along the midline of the plastron and represents 4.1061\% of total variation.
-The major variations along these axes correspond well to the differences between the mean shape of each class (Fig.
-\ref{fig:mean_shape}) where major class differences seem based on the relative ballooning or shrinking of the anterior and posterior portions of the plastron together.
 
@@
 % latex table generated in R 3.0.1 by xtable 1.7-1 package
-% Mon Aug  5 16:44:37 2013
+% Wed Aug  7 13:57:44 2013
 {\small
@@
 % latex table generated in R 3.0.1 by xtable 1.7-1 package
-% Mon Aug  5 16:44:37 2013
+% Wed Aug  7 13:57:44 2013
 \begin{tabular}{r|rrrr}

### 2013-08-07 src/turtle_mung.r

@@ -91,7 +91,7 @@ p.juv <- grep(pattern = '[jhJ]',
 b.juv <- grep(pattern = '[jhJ]', 
               x = as.character(turtle.info$b.sex),
               perl = TRUE)
-turtle.adult <- turtle.info[-p.juv, ]
-turtle.land.adult <- turtle.land.info[, , -p.juv]
 turtle.adult.dist <- riem.matrix(turtle.land.adult)
+turtle.land.adult <- turtle.land.info[, , -p.juv]
+turtle.adult <- turtle.info[-p.juv, ]
 turtle.geo.adult <- turtle.geo[-p.juv, ]

### 2013-08-07 documents/turtle_paper.Rnw
fixed data problem. some results of changed. most of paper fixed to reflect this. needs edits.

@@
 
-The first PC appears to describe variation in lateral plastral shape, from a very pointed medial edge to a much more rounded and blunt edge (Fig.
+The first two PCs appear to describe variation in lateral plastral shape, from a very pointed medial edge to a much more rounded and blunt edge (Fig.
 \ref{fig:pc_var}).
@@
 Over ontogeny, this is an area that deepens dorsoventrally and when the plastron was projected into two dimensions this it created the effect of mediolateral movement.
-Variation along the second PC appears to be characterized by variation along the midline, especially in terms of the ``straightness'' of the individual landmarks along it.
-This is most likely caused by the fact that plastral scutes do not always line up perfectly along the midline.
+Additional variation along the second PC appears to be characterized by variation in anterior and posterior landmarks along the medial edge of the plastron.
 
@@
 % multinomial logistic regression model
-The AICc best multinomial logistic regression models, for all four classification schemes, all include the first \Sexpr{length(tm.a.analysis$best$spinks$coefnames)} possible PCs as features (Tables \ref{tab:mod_sel_1},\ref{tab:mod_sel_2},\ref{tab:mod_sel_3}, and \ref{tab:mod_sel_4}).
-The second best models for all classification schemes had the first \Sexpr{length(tm.a.analysis$best$spinks$coefnames) - 1} PCs as features.
+The AICc best multinomial logistic regression model for three of the four classification schemes include the first \Sexpr{length(tm.a.analysis$best$spinks$coefnames) - 1} PCs as features (Tables \ref{tab:mod_sel_1}, \ref{tab:mod_sel_2}, and \ref{tab:mod_sel_3}).
+The second molecularly based classification hypothesis includes all \Sexpr{length(tm.a.analysis$best$spinks$coefnames)} PCs predictors (Tables \ref{tab:mod_sel_4}).
 The \(\Delta\)AICc values between the optimal and second best model range from \Sexpr{min(daic)} for the first morphological based classification hypothesis to \Sexpr{max(daic)} for the second molecular based classification hypothesis (Tables \ref{tab:mod_sel_1},\ref{tab:mod_sel_2},\ref{tab:mod_sel_3}, and \ref{tab:mod_sel_4}).
-The first \Sexpr{length(tm.a.analysis$best$spinks$coefnames)} PCs describe \Sexpr{paste0(sum(turtle.fit$percent[1:10]))}\% of total variation in plastral shape.
+The first \Sexpr{length(tm.a.analysis$best$spinks$coefnames) - 1} PCs describe \Sexpr{paste0(sum(turtle.fit$percent[1:9]))}\% of total variation in plastral shape, while the first \Sexpr{length(tm.a.analysis$best$spinks$coefnames)} PCs describe \Sexpr{paste0(sum(turtle.fit$perfect[1:10]))}\% of the variation.
 
-While the \(\Delta\)AICc value between the optimal and second best model for the first morphological based classification hypothesis was within the range to be considered equally optimal \citep{Burnham2002a}, for this analysis we chose to use only the AICc best model.
-While AICc values can not be compared between models with different responses \citep{Burnham2002a}, we interpret the fact that the optimal model for all classification schemes is the global model as a reason to use only the AICc best model for all cases.
+While the \(\Delta\)AICc value between the optimal and second best model for the first morphological and first molecular based classification hypothesis was within the range to be considered equally optimal \citep{Burnham2002a}, for this analysis we chose to use only the AICc best model.
+While AICc values can not be compared between models with different responses \citep{Burnham2002a}, we interpret the fact that the optimal model for three of the classification schemes having the same number of predictors and the fact that the \(\Delta\)AICc best model in these cases is the simpler mode as reasons to use only the AICc best model for all cases.
 Additionally, by using a single model for all classification schemes, this limits the number of comparisons between the bootstrap resampled distributions of the AUC values for the testing dataset (see below).
@@
 
-The selected number of features in the final random forest model for each classification scheme varied much more than for the multinomial logistic regression models, ranging from \Sexpr{min(oprf)} for the first morphological based classification hypothesis to \Sexpr{max(oprf)} for the second morphological based classification hypothesis (Fig.
+The selected number of features in the final random forest model for each classification scheme varied as much as the multinomial logistic regression models, ranging from \Sexpr{min(oprf)} for the second morphological based classification hypothesis and both molecular based classification hypotheses to \Sexpr{max(oprf)} for the first morphological based classification hypothesis (Fig.
 \ref{fig:roc}).
@@
 The selected random forest model for the best classification scheme had \Sexpr{trf.a$spinks$optsize} PCs as features.
-The PCs included as features in the final random forest model, in descending order of importance, were PCs \Sexpr{paste(rf.w[1:6])} and \Sexpr{paste(rf.w[7])}.
+The PCs included as features in the final random forest model, in descending order of importance, were PCs \Sexpr{paste(rf.w[1:8])} and \Sexpr{paste(rf.w[9])}.
 Of these \Sexpr{trf.a$spinks$optsize} features, the first three are illustrated here (Fig.
@@
 \ref{fig:imp_var}).
-The third and most important PC describes variation in the relative position of landmarks on anterior and posterior portions of the plastron and represents \Sexpr{turtle.fit$percent[3]}\% of total variation.
-The eighth and second most important PC mostly describes variation in landmarks along the midline of the plastron and represents \Sexpr{turtle.fit$percent[8]}\% of total variation.
+The third and most important PC describes variation roundedness of the medial portion of the plastron, both the anterior and posterior portions of the plastron.
+Additionally, the relative position of the landmarks along the midline.
+This PC represents \Sexpr{turtle.fit$percent[3]}\% of total variation.
+The second and second most important PC is described above and principally described variation in landmarks along the medial and anterior edge of the plastron.
+This PC represents \Sexpr{turtle.fit$percent[2]}\% of total variation.
 The major variations along these axes correspond well to the differences between the mean shape of each class (Fig.
@@
 
-The relative risk values for classification from the multinomial logistic regression model, based on the three most important PCs, demonstrate that while an individual PC might not be sufficient to differentiate any individual class from the northern group, given multiple features the odds of determining the correct classification increase (Fig.
+The relative risk values for classification from the multinomial logistic regression model, based on the three most important PCs, demonstrate that individual axes contribute differently to classification and that given multiple features the odds of determining the correct classification increase (Fig.
 \ref{fig:rel_risk}).
-Changes along the second most important axes contribute very little to increasing the odds of classification.
-This is observable from the class histograms of PC 8 (Fig.
+The first most important axis contributes strongly to classifying both the western and southern groups while changes along the second most important axes contribute very little to increasing the odds of classification for all but the eastern group.
+This is observable from the class histograms of PC 2 (Fig.
 \ref{fig:imp_var}).
@@
 The results of the relative risk values also shows that while a variable can be considered important for the random forest models (Fig.
-\ref{fig:imp_pc}, \ref{fig:imp_var}) they each might not be considered useful on their own (Fig.
+\ref{fig:imp_pc}, \ref{fig:imp_var}) they each might not be considered useful on their own for all classes (Fig.
 \ref{fig:rel_risk}).

### 2013-08-07 documents/turtle_paper.tex
fixed data problem. some results of changed. most of paper fixed to reflect this. needs edits.

@@
 \ref{fig:pca}).
-The first three PCs, which represent 55.0075790871272\% of the total variation, are a cloud of points with no structure.
+The first three PCs, which represent 45.2924805932624\% of the total variation, are a cloud of points with no structure.
 Additionally, individual landmark variation is mostly circular around each landmark with some more elliptical variation observed along some midline landmarks and the most lateral landmark (Fig.
@@
 However, it is important to note that Procustes based superimposition attempts to evenly distribute variance around the mean landmark \citep{Zelditch2004} and this observation should be considered cursory at best.
-The first PC appears to describe variation in lateral plastral shape, from a very pointed medial edge to a much more rounded and blunt edge (Fig.
+
+The first two PCs appear to describe variation in lateral plastral shape, from a very pointed medial edge to a much more rounded and blunt edge (Fig.
 \ref{fig:pc_var}).
@@
 % multinomial logistic regression model
-The AICc best multinomial logistic regression models, for all four classification schemes, all include the first 10 possible PCs as features (Tables \ref{tab:mod_sel_1},\ref{tab:mod_sel_2},\ref{tab:mod_sel_3}, and \ref{tab:mod_sel_4}).
-The second best models for all classification schemes had the first 9 PCs as features.
-The \(\Delta\)AICc values between the optimal and second best model range from 2.0639 for the first morphological based classification hypothesis to 19.8349 for the second molecular based classification hypothesis (Tables \ref{tab:mod_sel_1},\ref{tab:mod_sel_2},\ref{tab:mod_sel_3}, and \ref{tab:mod_sel_4}).
-The first 10 PCs describe 88.6043874205075\% of total variation in plastral shape.
+The AICc best multinomial logistic regression model for three of the four classification schemes include the first 9 PCs as features (Tables \ref{tab:mod_sel_1}, \ref{tab:mod_sel_2}, and \ref{tab:mod_sel_3}).
+The second molecularly based classification hypothesis includes all 10 PCs predictors (Tables \ref{tab:mod_sel_4}).
+The \(\Delta\)AICc values between the optimal and second best model range from 1.1782 for the first morphological based classification hypothesis to 26.5107 for the second molecular based classification hypothesis (Tables \ref{tab:mod_sel_1},\ref{tab:mod_sel_2},\ref{tab:mod_sel_3}, and \ref{tab:mod_sel_4}).
+The first 9 PCs describe 83.2255838908058\% of total variation in plastral shape, while the first 10 PCs describe 0\% of the variation.
 
-While the \(\Delta\)AICc value between the optimal and second best model for the first morphological based classification hypothesis was within the range to be considered equally optimal \citep{Burnham2002a}, for this analysis we chose to use only the AICc best model.
-While AICc values can not be compared between models with different responses \citep{Burnham2002a}, we interpret the fact that the optimal model for all classification schemes is the global model as a reason to use only the AICc best model for all cases.
+While the \(\Delta\)AICc value between the optimal and second best model for the first morphological and first molecular based classification hypothesis was within the range to be considered equally optimal \citep{Burnham2002a}, for this analysis we chose to use only the AICc best model.
+While AICc values can not be compared between models with different responses \citep{Burnham2002a}, we interpret the fact that the optimal model for three of the classification schemes having the same number of predictors and the fact that the \(\Delta\)AICc best model in these cases is the simpler mode as reasons to use only the AICc best model for all cases.
 Additionally, by using a single model for all classification schemes, this limits the number of comparisons between the bootstrap resampled distributions of the AUC values for the testing dataset (see below).
@@
 
-The selected number of features in the final random forest model for each classification scheme varied much more than for the multinomial logistic regression models, ranging from 6 for the first morphological based classification hypothesis to 10 for the second morphological based classification hypothesis (Fig.
+The selected number of features in the final random forest model for each classification scheme varied as much as the multinomial logistic regression models, ranging from 9 for the second morphological based classification hypothesis and both molecular based classification hypotheses to 10 for the first morphological based classification hypothesis (Fig.
 \ref{fig:roc}).
@@
 The results of fitting the final random forest model also include the variable importance for best separating the different classes.
-The selected random forest model for the best classification scheme had 7 PCs as features.
-The PCs included as features in the final random forest model, in descending order of importance, were PCs 3, 8, 6, 1, 4, 2 and 10.
-Of these 7 features, the first three are illustrated here (Fig.
+The selected random forest model for the best classification scheme had 9 PCs as features.
+The PCs included as features in the final random forest model, in descending order of importance, were PCs 3, 2, 1, 6, 5, 10, 9, 8 and 4.
+Of these 9 features, the first three are illustrated here (Fig.
 \ref{fig:imp_pc}) in descending order of importance.
@@
 \ref{fig:imp_var}).
-The third and most important PC describes variation in the relative position of landmarks on anterior and posterior portions of the plastron and represents 9.6697\% of total variation.
-The eighth and second most important PC mostly describes variation in landmarks along the midline of the plastron and represents 4.1061\% of total variation.
+The third and most important PC describes variation roundedness of the medial portion of the plastron, both the anterior and posterior portions of the plastron.
+Additionally, the relative position of the landmarks along the midline.
+This PC represents 12.1948\% of total variation.
+The second and second most important PC is described above and principally described variation in landmarks along the medial and anterior edge of the plastron.
+This PC represents 12.7794\% of total variation.
 The major variations along these axes correspond well to the differences between the mean shape of each class (Fig.
@@
 Depicted here are the for three PCs (lower triangle) and the mean shape with observed variance around each point (upper right).
-The first three PCs account for total 55.0075790871272\% of the variance in plastral shape.}
+The first three PCs account for total 45.2924805932624\% of the variance in plastral shape.}
   \label{fig:pca}
@@
 The middle column, for both rows, is the mean plastral shape for all observations.
-The first PC represents 30.4118\% of the total variation in plastral shape while PC represents 14.926\% of the variance.}
+The first PC represents 20.3183\% of the total variation in plastral shape while PC represents 12.7794\% of the variance.}
   \label{fig:pc_var}
@@
 % latex table generated in R 3.0.1 by xtable 1.7-1 package
-% Wed Aug  7 13:57:44 2013
+% Wed Aug  7 17:21:25 2013
 {\small
@@
   \hline
-+ & + & + & + & + & + & + & + & + & + & + & 22.00 & -245.34 & 537.41 & 0.00 & 0.74 \\ 
-  + & + & + & + & + & + & + & + & + & + &  & 20.00 & -248.61 & 539.48 & 2.06 & 0.26 \\ 
-  + & + & + & + & + & + & + & + & + &  &  & 18.00 & -255.73 & 549.28 & 11.87 & 0.00 \\ 
-  + & + & + & + & + & + & + & + &  &  &  & 16.00 & -267.59 & 568.62 & 31.21 & 0.00 \\ 
-  + & + & + & + & + & + &  &  &  &  &  & 12.00 & -283.44 & 591.70 & 54.29 & 0.00 \\ 
-  + & + & + & + & + &  &  &  &  &  &  & 10.00 & -286.02 & 592.61 & 55.20 & 0.00 \\ 
-  + & + & + & + & + & + & + &  &  &  &  & 14.00 & -282.24 & 593.59 & 56.17 & 0.00 \\ 
-  + & + & + & + &  &  &  &  &  &  &  & 8.00 & -307.43 & 631.23 & 93.81 & 0.00 \\ 
-  + & + & + &  &  &  &  &  &  &  &  & 6.00 & -340.94 & 694.09 & 156.68 & 0.00 \\ 
-  + & + &  &  &  &  &  &  &  &  &  & 4.00 & -345.95 & 700.01 & 162.60 & 0.00 \\ 
++ & + & + & + & + & + & + & + & + & + &  & 20.00 & -250.00 & 542.26 & 0.00 & 0.64 \\ 
+  + & + & + & + & + & + & + & + & + & + & + & 22.00 & -248.35 & 543.43 & 1.18 & 0.36 \\ 
+  + & + & + & + & + & + & + & + &  &  &  & 16.00 & -261.94 & 557.33 & 15.07 & 0.00 \\ 
+  + & + & + & + & + & + & + & + & + &  &  & 18.00 & -259.99 & 557.82 & 15.56 & 0.00 \\ 
+  + & + & + & + & + & + & + &  &  &  &  & 14.00 & -275.68 & 580.48 & 38.22 & 0.00 \\ 
+  + & + & + & + & + & + &  &  &  &  &  & 12.00 & -281.10 & 587.03 & 44.77 & 0.00 \\ 
+  + & + & + & + & + &  &  &  &  &  &  & 10.00 & -305.55 & 631.68 & 89.43 & 0.00 \\ 
+  + & + & + & + &  &  &  &  &  &  &  & 8.00 & -318.48 & 653.34 & 111.09 & 0.00 \\ 
+  + & + & + &  &  &  &  &  &  &  &  & 6.00 & -344.14 & 700.49 & 158.24 & 0.00 \\ 
+  + & + &  &  &  &  &  &  &  &  &  & 4.00 & -346.80 & 701.71 & 159.45 & 0.00 \\ 
    \hline
@@
   \hline
-+ & + & + & + & + & + & + & + & + & + & + & 22.00 & -242.26 & 531.25 & 0.00 & 0.86 \\ 
-  + & + & + & + & + & + & + & + & + & + &  & 20.00 & -246.34 & 534.94 & 3.70 & 0.14 \\ 
-  + & + & + & + & + & + & + & + & + &  &  & 18.00 & -252.76 & 543.34 & 12.10 & 0.00 \\ 
-  + & + & + & + & + & + & + & + &  &  &  & 16.00 & -263.19 & 559.83 & 28.59 & 0.00 \\ 
-  + & + & + & + & + & + &  &  &  &  &  & 12.00 & -275.48 & 575.79 & 44.54 & 0.00 \\ 
-  + & + & + & + & + & + & + &  &  &  &  & 14.00 & -273.87 & 576.85 & 45.60 & 0.00 \\ 
-  + & + & + & + & + &  &  &  &  &  &  & 10.00 & -280.27 & 581.11 & 49.86 & 0.00 \\ 
-  + & + & + & + &  &  &  &  &  &  &  & 8.00 & -303.54 & 623.46 & 92.22 & 0.00 \\ 
-  + & + & + &  &  &  &  &  &  &  &  & 6.00 & -342.40 & 697.01 & 165.76 & 0.00 \\ 
-  + & + &  &  &  &  &  &  &  &  &  & 4.00 & -348.78 & 705.67 & 174.42 & 0.00 \\ 
++ & + & + & + & + & + & + & + & + & + &  & 20.00 & -245.15 & 532.56 & 0.00 & 0.83 \\ 
+  + & + & + & + & + & + & + & + & + & + & + & 22.00 & -244.53 & 535.79 & 3.23 & 0.17 \\ 
+  + & + & + & + & + & + & + & + & + &  &  & 18.00 & -254.69 & 547.21 & 14.64 & 0.00 \\ 
+  + & + & + & + & + & + & + & + &  &  &  & 16.00 & -258.00 & 549.45 & 16.88 & 0.00 \\ 
+  + & + & + & + & + & + & + &  &  &  &  & 14.00 & -268.69 & 566.49 & 33.93 & 0.00 \\ 
+  + & + & + & + & + & + &  &  &  &  &  & 12.00 & -271.30 & 567.42 & 34.86 & 0.00 \\ 
+  + & + & + & + & + &  &  &  &  &  &  & 10.00 & -298.53 & 617.64 & 85.07 & 0.00 \\ 
+  + & + & + & + &  &  &  &  &  &  &  & 8.00 & -314.50 & 645.37 & 112.81 & 0.00 \\ 
+  + & + & + &  &  &  &  &  &  &  &  & 6.00 & -342.94 & 698.10 & 165.53 & 0.00 \\ 
+  + & + &  &  &  &  &  &  &  &  &  & 4.00 & -349.55 & 707.20 & 174.64 & 0.00 \\ 
    \hline
@@
   \hline
-+ & + & + & + & + & + & + & + & + & + & + & 33.00 & -297.91 & 668.06 & 0.00 & 0.97 \\ 
-  + & + & + & + & + & + & + & + & + & + &  & 30.00 & -305.12 & 675.37 & 7.30 & 0.03 \\ 
-  + & + & + & + & + & + & + & + & + &  &  & 27.00 & -314.81 & 687.76 & 19.70 & 0.00 \\ 
-  + & + & + & + & + & + & + & + &  &  &  & 24.00 & -329.53 & 710.30 & 42.24 & 0.00 \\ 
-  + & + & + & + & + & + & + &  &  &  &  & 21.00 & -342.93 & 730.35 & 62.28 & 0.00 \\ 
-  + & + & + & + & + & + &  &  &  &  &  & 18.00 & -353.86 & 745.55 & 77.49 & 0.00 \\ 
-  + & + & + & + & + &  &  &  &  &  &  & 15.00 & -357.66 & 746.59 & 78.53 & 0.00 \\ 
-  + & + & + & + &  &  &  &  &  &  &  & 12.00 & -385.36 & 795.54 & 127.48 & 0.00 \\ 
-  + & + & + &  &  &  &  &  &  &  &  & 9.00 & -424.50 & 867.48 & 199.41 & 0.00 \\ 
-  + & + &  &  &  &  &  &  &  &  &  & 6.00 & -442.85 & 897.91 & 229.85 & 0.00 \\ 
++ & + & + & + & + & + & + & + & + & + &  & 30.00 & -303.61 & 672.34 & 0.00 & 0.77 \\ 
+  + & + & + & + & + & + & + & + & + & + & + & 33.00 & -301.25 & 674.74 & 2.41 & 0.23 \\ 
+  + & + & + & + & + & + & + & + & + &  &  & 27.00 & -314.28 & 686.70 & 14.36 & 0.00 \\ 
+  + & + & + & + & + & + & + & + &  &  &  & 24.00 & -318.22 & 687.70 & 15.37 & 0.00 \\ 
+  + & + & + & + & + & + & + &  &  &  &  & 21.00 & -335.11 & 714.71 & 42.37 & 0.00 \\ 
+  + & + & + & + & + & + &  &  &  &  &  & 18.00 & -353.04 & 743.91 & 71.57 & 0.00 \\ 
+  + & + & + & + & + &  &  &  &  &  &  & 15.00 & -385.20 & 801.67 & 129.34 & 0.00 \\ 
+  + & + & + & + &  &  &  &  &  &  &  & 12.00 & -397.69 & 820.21 & 147.87 & 0.00 \\ 
+  + & + & + &  &  &  &  &  &  &  &  & 9.00 & -437.13 & 892.73 & 220.39 & 0.00 \\ 
+  + & + &  &  &  &  &  &  &  &  &  & 6.00 & -451.19 & 914.60 & 242.27 & 0.00 \\ 
    \hline
@@
   \hline
-+ & + & + & + & + & + & + & + & + & + & + & 33.00 & -253.75 & 579.72 & 0.00 & 1.00 \\ 
-  + & + & + & + & + & + & + & + & + & + &  & 30.00 & -267.22 & 599.55 & 19.83 & 0.00 \\ 
-  + & + & + & + & + & + & + & + & + &  &  & 27.00 & -275.24 & 608.60 & 28.88 & 0.00 \\ 
-  + & + & + & + & + & + & + & + &  &  &  & 24.00 & -302.99 & 657.22 & 77.51 & 0.00 \\ 
-  + & + & + & + & + & + & + &  &  &  &  & 21.00 & -307.70 & 659.88 & 80.16 & 0.00 \\ 
-  + & + & + & + & + &  &  &  &  &  &  & 15.00 & -327.52 & 686.30 & 106.59 & 0.00 \\ 
-  + & + & + & + & + & + &  &  &  &  &  & 18.00 & -324.35 & 686.51 & 106.79 & 0.00 \\ 
-  + & + & + & + &  &  &  &  &  &  &  & 12.00 & -350.14 & 725.11 & 145.39 & 0.00 \\ 
-  + & + & + &  &  &  &  &  &  &  &  & 9.00 & -390.78 & 800.03 & 220.32 & 0.00 \\ 
-  + & + &  &  &  &  &  &  &  &  &  & 6.00 & -405.14 & 822.49 & 242.77 & 0.00 \\ 
++ & + & + & + & + & + & + & + & + & + & + & 33.00 & -251.73 & 575.67 & 0.00 & 1.00 \\ 
+  + & + & + & + & + & + & + & + & + & + &  & 30.00 & -268.54 & 602.18 & 26.51 & 0.00 \\ 
+  + & + & + & + & + & + & + & + & + &  &  & 27.00 & -283.99 & 626.10 & 50.43 & 0.00 \\ 
+  + & + & + & + & + & + & + & + &  &  &  & 24.00 & -295.61 & 642.46 & 66.78 & 0.00 \\ 
+  + & + & + & + & + & + & + &  &  &  &  & 21.00 & -302.50 & 649.48 & 73.81 & 0.00 \\ 
+  + & + & + & + & + & + &  &  &  &  &  & 18.00 & -316.59 & 671.00 & 95.32 & 0.00 \\ 
+  + & + & + & + & + &  &  &  &  &  &  & 15.00 & -340.84 & 712.95 & 137.27 & 0.00 \\ 
+  + & + & + & + &  &  &  &  &  &  &  & 12.00 & -353.01 & 730.84 & 155.17 & 0.00 \\ 
+  + & + & + &  &  &  &  &  &  &  &  & 9.00 & -378.16 & 774.78 & 199.11 & 0.00 \\ 
+  + & + &  &  &  &  &  &  &  &  &  & 6.00 & -395.71 & 803.64 & 227.97 & 0.00 \\ 
    \hline
@@
 % latex table generated in R 3.0.1 by xtable 1.7-1 package
-% Wed Aug  7 13:57:44 2013
+% Wed Aug  7 17:21:25 2013
 \begin{tabular}{r|rrrr}
@@
   morph 2 & 0.00 &  &  &  \\ 
-  molec 1 & 0.51 & 0.00 &  &  \\ 
+  molec 1 & 0.00 & 0.00 &  &  \\ 
   molec 2 & 0.00 & 0.00 & 0.00 &  \\ 

### 2013-08-08 documents/turtle_paper.Rnw
results updated completely. some intro work. grow turtle section.

@@
 
-options(digits = 4)
+options(digits = 2)
+
+#knit_hooks$set(inline = function(x) {
+#               if (is.numeric(x)) round(x, 4)})
+
 @
@@
 Conceringly, most extant taxa, and nearly all extinct taxa, are delimited based solely via morphological analysis.
-This phenomenon is of great concern when studying variation and diversity dynamics over long periods of time, where apparent morphological stasis may not actually reflect true diversity \citet{Hunt2008,Eldredge1972,Gould1977a}.
+This phenomenon is of great concern when studying variation and diversity dynamics over long periods of time, where apparent morphological stasis may not actually reflect true diversity \citep{Hunt2008,Eldredge1972,Gould1977a}.
 In the case of endangered or conserved taxa, morphometric approaches for classifying and identifying taxa and populations of importance would greatly improve the ability to maintain these high risk groups.
@@
 
-Much work has been devoted to species delimitation via sequence difference \citep{Fujita2012,Yang2010b} while comparatively little has been devoted for case of purely morphological data.
-The majority of this effort has focused on identifying differences between already identified taxa \citep{Zelditch2004} and automated taxon identification \citep{MacLeod2007}.
+Much work has been devoted to species delimitation via sequence difference \citep{Fujita2012,Yang2010b} while comparatively little has been devoted to introducing new methodology for case of purely morphological data \citep{Mitteroecker2011,Zelditch2004}.
+The majority of this effort has focused on identifying differences between already identified taxa \citep{Polly2007a,Demandt2009,Gaubert2005b,Gunduz2007,Polly2003,Zelditch2004} and automated taxon identification \citep{MacLeod2007}.
 % all the work done on species delimitation strictly from molecular data
@@
 \subsection{Background and system}
-Differences in morphological variation between different classes has previously been analyzed using methods like linear discriminate analysis and canonical variates analysis \citep{Zelditch2004}.
+Differences in morphological variation between different classes has previously been analyzed using methods like linear discriminate analysis and canonical variates analysis \citep{Zelditch2004,Mitteroecker2011,Polly2007a,Polly2003,Gunduz2007,Gaubert2005b,Demandt2009}.
 These methods are comparatively simple and straight forward ways of understanding the differences in morphology between classes.
 Also, they are very visual methods which aides in interpretation and presentation of information.
-Neural network models have also been introduced and applied frequently in the context of automated taxon identification along with more general applications \citep{MacLeod2007}.
+These previous studies, however, do not compare which amongst a set of candidate classification hypotheses is better.
+Also, there is no generalization of the classification model and training data accuracy is almost exclusively used as the metric off strength of classification.
+
 Here, we used multiple alternative machine learning methods, both unsupervised and supervised, in order to compare different classification hypotheses.
 These methods provide different and unique advantages for understanding how to classify taxa, with what accuracy, and what these classifications are based on.
+While machine learning methods such as neural networks have been applied to studying shape variation \citep{MacLeod2007}, they have been primarily applied in the context of automated taxon identification and not in terms of group classification and strength of classification.
 Additionally, we investigate variation in continuous traits, and do not search for discrete differences between each class, instead focusing on multivariate quantification of shape.
@@
 marmorata marmorata}, the southern \textit{E.
-marmorata palida}, and a central Californian intergrade zone \citep{Seeliger1945,Holland1992}.
+marmorata palida}, and a central Californian intergrade zone \citep{Seeliger1945}.
 \textit{E.
@@
 
-Previous work on morphological differentiation between different populations of \textit{E. marmorata} with no alternative subspecific assignments \citep{Seeliger1945,Holland1992}. 
+Previous work on morphological differentiation in \textit{E.
+marmorata} focused on comparisons between different populations within the rage CITATIONS.
+Most of these studies were not over the entire range and used only linear measures of morphology instead of multivariate shape CITATIONS.
 %
@@
 \ref{fig:pca}).
-The first three PCs, which represent \Sexpr{paste0(sum(turtle.fit$percent[1:3]))}\% of the total variation, are a cloud of points with no structure.
+The first three PCs, which represent \Sexpr{sum(turtle.fit$percent[1:3])}\% of the total variation, are a cloud of points with no structure.
 Additionally, individual landmark variation is mostly circular around each landmark with some more elliptical variation observed along some midline landmarks and the most lateral landmark (Fig.
 \ref{fig:pca}).
-However, it is important to note that Procustes based superimposition attempts to evenly distribute variance around the mean landmark \citep{Zelditch2004} and this observation should be considered cursory at best.
+However, it is important to note that Procustes based superimposition attempts to evenly distribute variance around the mean shape \citep{Zelditch2004} and this observation should be considered cursory at best.
 
-The first two PCs appear to describe variation in lateral plastral shape, from a very pointed medial edge to a much more rounded and blunt edge (Fig.
+The first two PCs appear to describe principally varaition in the lateral margin of te palstra, from a pointed medial edge to a more rounded and blunt edge (Fig.
 \ref{fig:pc_var}).
 Landmark 10 (Fig.
-\ref{fig:plastra}), which appears to be the most variable along this axis (Fig.
+\ref{fig:plastra}), which appears to be the most variable along these axes (Fig.
 \ref{fig:pca} and \ref{fig:pc_var}), is positioned on the bridge between the plastron and the carapace.
 Over ontogeny, this is an area that deepens dorsoventrally and when the plastron was projected into two dimensions this it created the effect of mediolateral movement.
-Additional variation along the second PC appears to be characterized by variation in anterior and posterior landmarks along the medial edge of the plastron.
+Lateral landmark variation along the first PC seemed concentrated in the posterior portion of the plastra with additional variance observed in midline landmarks (Fig.
+\ref{fig:pc_var}).
+This variance in midline landmakrs was most likely caused by the fact that plastral scutes frequently do not line up perfectly.
+Along PC 2, lateral variation appeared to be concetrated in the anterior portion o the plastra (Fig.
+\ref{fig:pc_var}).
 
@@
 % multinomial logistic regression model
-The AICc best multinomial logistic regression model for three of the four classification schemes include the first \Sexpr{length(tm.a.analysis$best$spinks$coefnames) - 1} PCs as features (Tables \ref{tab:mod_sel_1}, \ref{tab:mod_sel_2}, and \ref{tab:mod_sel_3}).
-The second molecularly based classification hypothesis includes all \Sexpr{length(tm.a.analysis$best$spinks$coefnames)} PCs predictors (Tables \ref{tab:mod_sel_4}).
-The \(\Delta\)AICc values between the optimal and second best model range from \Sexpr{min(daic)} for the first morphological based classification hypothesis to \Sexpr{max(daic)} for the second molecular based classification hypothesis (Tables \ref{tab:mod_sel_1},\ref{tab:mod_sel_2},\ref{tab:mod_sel_3}, and \ref{tab:mod_sel_4}).
-The first \Sexpr{length(tm.a.analysis$best$spinks$coefnames) - 1} PCs describe \Sexpr{paste0(sum(turtle.fit$percent[1:9]))}\% of total variation in plastral shape, while the first \Sexpr{length(tm.a.analysis$best$spinks$coefnames)} PCs describe \Sexpr{paste0(sum(turtle.fit$perfect[1:10]))}\% of the variation.
+The AICc best multinomial logistic regression model for three of the four classification schemes had the first \Sexpr{length(tm.a.analysis$best$spinks$coefnames) - 1} PCs as features (Tables \ref{tab:mod_sel_1}, \ref{tab:mod_sel_2}, and \ref{tab:mod_sel_3}).
+The second molecularly based classification hypothesis included all \Sexpr{length(tm.a.analysis$best$spinks$coefnames)} possible PCs as predictors (Tables \ref{tab:mod_sel_4}).
+The \(\Delta\)AICc values between the optimal and second best model range from \Sexpr{min(daic)} for the first morphological based classification hypothesis to \Sexpr{max(daic)} for the second molecular based classification hypothesis (Tables \ref{tab:mod_sel_1}, \ref{tab:mod_sel_2}, \ref{tab:mod_sel_3}, and \ref{tab:mod_sel_4}).
+The first \Sexpr{length(tm.a.analysis$best$spinks$coefnames) - 1} PCs describe \Sexpr{sum(turtle.fit$percent[1:9])}\% of total variation in plastral shape, while the first \Sexpr{length(tm.a.analysis$best$spinks$coefnames)} PCs describe \Sexpr{sum(turtle.fit$percent[1:10])}\% of the variation.
 
 While the \(\Delta\)AICc value between the optimal and second best model for the first morphological and first molecular based classification hypothesis was within the range to be considered equally optimal \citep{Burnham2002a}, for this analysis we chose to use only the AICc best model.
-While AICc values can not be compared between models with different responses \citep{Burnham2002a}, we interpret the fact that the optimal model for three of the classification schemes having the same number of predictors and the fact that the \(\Delta\)AICc best model in these cases is the simpler mode as reasons to use only the AICc best model for all cases.
-Additionally, by using a single model for all classification schemes, this limits the number of comparisons between the bootstrap resampled distributions of the AUC values for the testing dataset (see below).
+While AICc values can not be compared between models with different responses \citep{Burnham2002a}, we interpret the fact that the \(\Delta\)AICc best model in these cases is the simpler model and that the optimal model for three of the classification schemes having the same number of predictors as reasons to use only the AICc best model for all cases.
+Additionally, by using a single model for each of the classification hypotheses, this limits the number of comparisons between the bootstrap resampled distributions of the AUC values for the testing dataset (see below).
 
@@
 
-The selected number of features in the final random forest model for each classification scheme varied as much as the multinomial logistic regression models, ranging from \Sexpr{min(oprf)} for the second morphological based classification hypothesis and both molecular based classification hypotheses to \Sexpr{max(oprf)} for the first morphological based classification hypothesis (Fig.
+The selected number of features in the final random forest model for each classification scheme was very simplar to the model selection results from the multinomial logistic regression models, ranging from \Sexpr{min(oprf)} for the second morphological based classification hypothesis and both molecular based classification hypotheses to \Sexpr{max(oprf)} for the first morphological based classification hypothesis (Fig.
 \ref{fig:roc}).
@@
 
-In the case of all models, there is a substantial increase in model performance as measured by AICc for the multinomial logistic models (Tables \ref{tab:mod_sel_1},\ref{tab:mod_sel_2},\ref{tab:mod_sel_3}, and \ref{tab:mod_sel_4}) or in AUC for the random forest models and illustrated for the multinomial logistic regression models as the number of features increases (Fig.
+In the case of all models, there is a substantial increase in model performance as measured by AICc for the multinomial logistic models (Tables \ref{tab:mod_sel_1}, \ref{tab:mod_sel_2}, \ref{tab:mod_sel_3}, and \ref{tab:mod_sel_4}) or in AUC for the random forest models and illustrated for the multinomial logistic regression models as the number of features increases (Fig.
 \ref{fig:roc}).
 
-The results from the generalization of the selected supervised learning models, measured by the distributions of the bootstrapped AUC values of the testing dataset, show that the second molecular classification hypotheses was the best overall classification scheme (Fig.
+The results from the generalization of the selected supervised learning models, measured by the distributions of the bootstrapped AUC values of the testing dataset, show that a molecular classification hypotheses was the best overall classification scheme (Fig.
 \ref{fig:gen_res}).
+Remarkably, the best classification hypothesis was the second molecular classification hypothesis based on both the multinomial logistic regression and random forest models.
 For both methods, the distribution of bootstrapped AUC for the molecular hypothesis was significantly greater than all of the other classification schemes (Tables \ref{tab:mm_test} and \ref{tab:rf_test}).
-Remarkably, the best classification hypothesis was identical based on both the multinomial logistic regression and random forest models.
 
@@
 The third and most important PC describes variation roundedness of the medial portion of the plastron, both the anterior and posterior portions of the plastron.
-Additionally, the relative position of the landmarks along the midline.
+Additionally, the relative position of the landmarks along the midline varies greatly along PC3 (Fig.
+\ref{fig:imp_var}).
 This PC represents \Sexpr{turtle.fit$percent[3]}\% of total variation.
-The second and second most important PC is described above and principally described variation in landmarks along the medial and anterior edge of the plastron.
+The second and second most important PC is described above and principally described variation in landmarks along the lateral and anterior margin of the plastron.
 This PC represents \Sexpr{turtle.fit$percent[2]}\% of total variation.
 The major variations along these axes correspond well to the differences between the mean shape of each class (Fig.
-\ref{fig:mean_shape}) where major class differences seem based on the relative ballooning or shrinking of the anterior and posterior portions of the plastron together.
+\ref{fig:mean_shape}) where major class differences seem based on the relative ballooning or shrinking of the anterior and posterior portions of the plastron together along with differential ``pinching'' of the midline landmarks.
 
-The relative risk values for classification from the multinomial logistic regression model, based on the three most important PCs, demonstrate that individual axes contribute differently to classification and that given multiple features the odds of determining the correct classification increase (Fig.
+The relative risk values for classification from the multinomial logistic regression model, based on the three most important PCs, demonstrate that individual axes contribute to classification differently and that given multiple features the odds of determining the correct classification increase (Fig.
 \ref{fig:rel_risk}).
 The first most important axis contributes strongly to classifying both the western and southern groups while changes along the second most important axes contribute very little to increasing the odds of classification for all but the eastern group.
-This is observable from the class histograms of PC 2 (Fig.
+This is observable from the class histograms of PC 3 and 2 (Fig.
 \ref{fig:imp_var}).
@@
 \ref{fig:roc}), this helped demonstrate how fine scale the actual variation between classes was.
+The variable importance information from the random forest models was extremely useful for understanding what aspects shape variance contributed most to differentiating the classes and in what order as opposed purely in the order of largest variance (Fig.
+\ref{fig:imp_pc} and \ref{fig:imp_var}).
 Additionally, the relative risk values from the mulitinomial logistic regression models demonstrate that a single PC is probably not sufficient for estimating the class of an observation, but that given a set of PCs this classification would be more accurate (Fig.
-\ref{fig:rel_risk}).
-The results of the relative risk values also shows that while a variable can be considered important for the random forest models (Fig.
-\ref{fig:imp_pc}, \ref{fig:imp_var}) they each might not be considered useful on their own for all classes (Fig.
 \ref{fig:rel_risk}).
@@
   \caption{Landmark variation along the two most important features (PCs) based on the final random forest model.
-The first row corresponds to the third PC and the second corresponds to eighth PC.
+The first row corresponds to the third PC and the second corresponds to the second PC.
 Landmark configurations are minimum observed on that PC, mean shape, and maximum observed on that PC.}

### 2013-08-08 documents/turtle_paper.tex
results updated completely. some intro work. grow turtle section.

@@
 \ref{fig:pca}).
-The first three PCs, which represent 45.2924805932624\% of the total variation, are a cloud of points with no structure.
+The first three PCs, which represent 45.29\% of the total variation, are a cloud of points with no structure.
 Additionally, individual landmark variation is mostly circular around each landmark with some more elliptical variation observed along some midline landmarks and the most lateral landmark (Fig.
 \ref{fig:pca}).
-However, it is important to note that Procustes based superimposition attempts to evenly distribute variance around the mean landmark \citep{Zelditch2004} and this observation should be considered cursory at best.
+However, it is important to note that Procustes based superimposition attempts to evenly distribute variance around the mean shape \citep{Zelditch2004} and this observation should be considered cursory at best.
 
-The first two PCs appear to describe variation in lateral plastral shape, from a very pointed medial edge to a much more rounded and blunt edge (Fig.
+The first two PCs appear to describe principally varaition in the lateral margin of te palstra, from a pointed medial edge to a more rounded and blunt edge (Fig.
 \ref{fig:pc_var}).
 Landmark 10 (Fig.
-\ref{fig:plastra}), which appears to be the most variable along this axis (Fig.
+\ref{fig:plastra}), which appears to be the most variable along these axes (Fig.
 \ref{fig:pca} and \ref{fig:pc_var}), is positioned on the bridge between the plastron and the carapace.
 Over ontogeny, this is an area that deepens dorsoventrally and when the plastron was projected into two dimensions this it created the effect of mediolateral movement.
-Additional variation along the second PC appears to be characterized by variation in anterior and posterior landmarks along the medial edge of the plastron.
+Lateral landmark variation along the first PC seemed concentrated in the posterior portion of the plastra with additional variance observed in midline landmarks (Fig.
+\ref{fig:pc_var}).
+This variance in midline landmakrs was most likely caused by the fact that plastral scutes frequently do not line up perfectly.
+Along PC 2, lateral variation appeared to be concetrated in the anterior portion o the plastra (Fig.
+\ref{fig:pc_var}).
 
@@
 % multinomial logistic regression model
-The AICc best multinomial logistic regression model for three of the four classification schemes include the first 9 PCs as features (Tables \ref{tab:mod_sel_1}, \ref{tab:mod_sel_2}, and \ref{tab:mod_sel_3}).
-The second molecularly based classification hypothesis includes all 10 PCs predictors (Tables \ref{tab:mod_sel_4}).
-The \(\Delta\)AICc values between the optimal and second best model range from 1.1782 for the first morphological based classification hypothesis to 26.5107 for the second molecular based classification hypothesis (Tables \ref{tab:mod_sel_1},\ref{tab:mod_sel_2},\ref{tab:mod_sel_3}, and \ref{tab:mod_sel_4}).
-The first 9 PCs describe 83.2255838908058\% of total variation in plastral shape, while the first 10 PCs describe 0\% of the variation.
+The AICc best multinomial logistic regression model for three of the four classification schemes had the first 9 PCs as features (Tables \ref{tab:mod_sel_1}, \ref{tab:mod_sel_2}, and \ref{tab:mod_sel_3}).
+The second molecularly based classification hypothesis included all 10 possible PCs as predictors (Tables \ref{tab:mod_sel_4}).
+The \(\Delta\)AICc values between the optimal and second best model range from 1.18 for the first morphological based classification hypothesis to 26.51 for the second molecular based classification hypothesis (Tables \ref{tab:mod_sel_1}, \ref{tab:mod_sel_2}, \ref{tab:mod_sel_3}, and \ref{tab:mod_sel_4}).
+The first 9 PCs describe 83.23\% of total variation in plastral shape, while the first 10 PCs describe 86.54\% of the variation.
 
 While the \(\Delta\)AICc value between the optimal and second best model for the first morphological and first molecular based classification hypothesis was within the range to be considered equally optimal \citep{Burnham2002a}, for this analysis we chose to use only the AICc best model.
-While AICc values can not be compared between models with different responses \citep{Burnham2002a}, we interpret the fact that the optimal model for three of the classification schemes having the same number of predictors and the fact that the \(\Delta\)AICc best model in these cases is the simpler mode as reasons to use only the AICc best model for all cases.
-Additionally, by using a single model for all classification schemes, this limits the number of comparisons between the bootstrap resampled distributions of the AUC values for the testing dataset (see below).
+While AICc values can not be compared between models with different responses \citep{Burnham2002a}, we interpret the fact that the \(\Delta\)AICc best model in these cases is the simpler model and that the optimal model for three of the classification schemes having the same number of predictors as reasons to use only the AICc best model for all cases.
+Additionally, by using a single model for each of the classification hypotheses, this limits the number of comparisons between the bootstrap resampled distributions of the AUC values for the testing dataset (see below).
 
@@
 
-The selected number of features in the final random forest model for each classification scheme varied as much as the multinomial logistic regression models, ranging from 9 for the second morphological based classification hypothesis and both molecular based classification hypotheses to 10 for the first morphological based classification hypothesis (Fig.
+The selected number of features in the final random forest model for each classification scheme was very simplar to the model selection results from the multinomial logistic regression models, ranging from 9 for the second morphological based classification hypothesis and both molecular based classification hypotheses to 10 for the first morphological based classification hypothesis (Fig.
 \ref{fig:roc}).
@@
 The third and most important PC describes variation roundedness of the medial portion of the plastron, both the anterior and posterior portions of the plastron.
-Additionally, the relative position of the landmarks along the midline.
-This PC represents 12.1948\% of total variation.
-The second and second most important PC is described above and principally described variation in landmarks along the medial and anterior edge of the plastron.
-This PC represents 12.7794\% of total variation.
+Additionally, the relative position of the landmarks along the midline varies greatly along PC3 (Fig.
+\ref{fig:imp_var}).
+This PC represents 12.19\% of total variation.
+The second and second most important PC is described above and principally described variation in landmarks along the lateral and anterior margin of the plastron.
+This PC represents 12.78\% of total variation.
 The major variations along these axes correspond well to the differences between the mean shape of each class (Fig.
-\ref{fig:mean_shape}) where major class differences seem based on the relative ballooning or shrinking of the anterior and posterior portions of the plastron together.
+\ref{fig:mean_shape}) where major class differences seem based on the relative ballooning or shrinking of the anterior and posterior portions of the plastron together along with differential ``pinching'' of the midline landmarks.
 
-The relative risk values for classification from the multinomial logistic regression model, based on the three most important PCs, demonstrate that individual axes contribute differently to classification and that given multiple features the odds of determining the correct classification increase (Fig.
+The relative risk values for classification from the multinomial logistic regression model, based on the three most important PCs, demonstrate that individual axes contribute to classification differently and that given multiple features the odds of determining the correct classification increase (Fig.
 \ref{fig:rel_risk}).
 The first most important axis contributes strongly to classifying both the western and southern groups while changes along the second most important axes contribute very little to increasing the odds of classification for all but the eastern group.
-This is observable from the class histograms of PC 2 (Fig.
+This is observable from the class histograms of PC 3 and 2 (Fig.
 \ref{fig:imp_var}).
@@
 The middle column, for both rows, is the mean plastral shape for all observations.
-The first PC represents 20.3183\% of the total variation in plastral shape while PC represents 12.7794\% of the variance.}
+The first PC represents 20.32\% of the total variation in plastral shape while PC represents 12.78\% of the variance.}
   \label{fig:pc_var}
@@
 % latex table generated in R 3.0.1 by xtable 1.7-1 package
-% Wed Aug  7 17:21:25 2013
+% Thu Aug  8 00:20:39 2013
 {\small
@@
 % latex table generated in R 3.0.1 by xtable 1.7-1 package
-% Wed Aug  7 17:21:25 2013
+% Thu Aug  8 00:20:39 2013
 \begin{tabular}{r|rrrr}

### 2013-08-08 src/turtle_plots.r

@@ -234,19 +234,19 @@ ggimp <- ggpairs(turtle.adult,
                  upper = 'blank',
                  lower = 'blank',
                  params = c(LabelSize = 2, gridLabelSize = 2, size = 1))
-pc1 <- ggplot(turtle.adult, mapping = aes(x = PC3, y = PC8, colour = label))
+pc1 <- ggplot(turtle.adult, mapping = aes(x = PC3, y = PC2, colour = label))
 pc1 <- pc1 + geom_point() + scale_color_manual(values = cbp)
-pc2 <- ggplot(turtle.adult, mapping = aes(x = PC3, y = PC6, colour = label))
+pc2 <- ggplot(turtle.adult, mapping = aes(x = PC3, y = PC1, colour = label))
 pc2 <- pc2 + geom_point() + scale_color_manual(values = cbp)
-pc3 <- ggplot(turtle.adult, mapping = aes(x = PC8, y = PC6, colour = label))
+pc3 <- ggplot(turtle.adult, mapping = aes(x = PC2, y = PC1, colour = label))
 pc3 <- pc3 + geom_point() + scale_color_manual(values = cbp)
 hist1 <- ggplot(turtle.adult, mapping = aes(x = PC3, fill = label))
 hist1 <- hist1 + geom_histogram()
 hist1 <- hist1 + facet_grid(label ~ .) + scale_fill_manual(values = cbp)
-hist2 <- ggplot(turtle.adult, mapping = aes(x = PC8, fill = label))
+hist2 <- ggplot(turtle.adult, mapping = aes(x = PC2, fill = label))
 hist2 <- hist2 + geom_histogram()
 hist2 <- hist2 + facet_grid(label ~ .) + scale_fill_manual(values = cbp)
-hist3 <- ggplot(turtle.adult, mapping = aes(x = PC6, fill = label))
+hist3 <- ggplot(turtle.adult, mapping = aes(x = PC1, fill = label))
 hist3 <- hist3 + geom_histogram()
 hist3 <- hist3 + facet_grid(label ~ .) + scale_fill_manual(values = cbp)
 

### 2013-08-08 documents/turtle_paper.Rnw
chisq comparison. few more edits. see ken comments.

@@
 load('../data/turtle_gen.RData')
+load('../data/turtle_shape.RData')
 
@@
 These previous studies, however, do not compare which amongst a set of candidate classification hypotheses is better.
+For example, studies such as those of \citet{Caumul2005a} and \citet{Polly2007a} focused on comparing different aspects of morphology and their fidelity to a classification scheme, instead of comparing the fidelity of one aspect of morphology to multiple classification shcemes.
 Also, there is no generalization of the classification model and training data accuracy is almost exclusively used as the metric off strength of classification.
@@
 % some sex information. Analyze it.
-Unfortunately, it was not possible to obtain enough detailed information on the sex of each \textit{E.
-marmorata} specimen, thus it is difficult at best to determine if this clustering solution corresponds to sexual dimorphism between observations.
-Male emydid turtles are known to have a plastral concavity which may influence landmark position along the midline.
+Sex information was only avaliable for \Sexpr{length(tsex.s)} of the \Sexpr{nrow(turtle.adult)} turtles.
+A \(\chi^{2}\) test of the relationship between sex observation and cluster assignment from PAM with two clusters showed that there was no significant relationship between cluster assignment and sex observation (\(\chi^{2}\): \Sexpr{csex.chi$statistic}, df: \Sexpr{csex.chi$parameter}, \textit{p}-value: \Sexpr{csex.chi$p.value}, Table \ref{tab:chi}).
+This results is interesting because while sexual dimorphism has been observed in linear measures and mass estimates of \textit{E.
+marmorata}, this results demonstrates that this dimorphism may not translate into differences in shape.
+Interestsingly, male emydid turtles are known to have a plastral concavity which may influence landmark position along the midline.
 However, the plastral concavity of \textit{E.
 mamorata} males is considered less pronounced than in other emydid turtles.
-While we cannot completely rule out sexual dimorphism as the root cause of this observation, because of the very small degree of known sexual dimorphism in \textit{E.
-marmorata} we are less concerned with possible effects of sexual dimorphism for the later supervised learning methods.
 
@@
 
+\begin{table}
+  \centering
+<<results = 'asis'>>=
+print.xtable(xsex.tab,
+             floating = FALSE,
+             hline.after = c(0, nrow(csex.tab) - 1))
+@
+  \caption{Tabular comparison between sex observation and cluster assignement from PAM with two clusters.
+This number of clusters was chosen because it represented the second best clustering solution as determined via gap statistic comparison (Fig.
+\ref{fig:gap}).
+\(\chi^{2}\) analysis of this contigency table showed that there is no relationship between sex observation and cluster assignment (\(\chi^{2}\): \Sexpr{csex.chi$statistic}, df: \Sexpr{csex.chi$parameter}, \textit{p}-value: \Sexpr{csex.chi$p.value}).}
+  \label{tab:chi}
+\end{table}
+
 \begin{sidewaystable}

### 2013-08-08 documents/turtle_paper.tex
chisq comparison. few more edits. see ken comments.

@@
 % some sex information. Analyze it.
-Unfortunately, it was not possible to obtain enough detailed information on the sex of each \textit{E.
-marmorata} specimen, thus it is difficult at best to determine if this clustering solution corresponds to sexual dimorphism between observations.
-Male emydid turtles are known to have a plastral concavity which may influence landmark position along the midline.
+Sex information was only avaliable for 399 of the 524 turtles.
+A \(\chi^{2}\) test of the relationship between sex observation and cluster assignment from PAM with two clusters showed that there was no significant relationship between cluster assignment and sex observation (\(\chi^{2}\): 1.12, df: 1, \textit{p}-value: 0.29, Table \ref{tab:chi}).
+This results is interesting because while sexual dimorphism has been observed in linear measures and mass estimates of \textit{E.
+marmorata}, this results demonstrates that this dimorphism may not translate into differences in shape.
+Interestsingly, male emydid turtles are known to have a plastral concavity which may influence landmark position along the midline.
 However, the plastral concavity of \textit{E.
 mamorata} males is considered less pronounced than in other emydid turtles.
-While we cannot completely rule out sexual dimorphism as the root cause of this observation, because of the very small degree of known sexual dimorphism in \textit{E.
-marmorata} we are less concerned with possible effects of sexual dimorphism for the later supervised learning methods.
 
@@
 
+\begin{table}
+  \centering
+% latex table generated in R 3.0.1 by xtable 1.7-1 package
+% Thu Aug  8 14:11:16 2013
+\begin{tabular}{r|cc|c}
+  & F & M & tot \\ 
+  \hline
+1 & 101 & 112 & 213 \\ 
+  2 & 99 & 87 & 186 \\ 
+   \hline
+tot & 200 & 199 & 399 \\ 
+  \end{tabular}
+
+  \caption{Tabular comparison between sex observation and cluster assignement from PAM with two clusters.
+This number of clusters was chosen because it represented the second best clustering solution as determined via gap statistic comparison (Fig.
+\ref{fig:gap}).
+\(\chi^{2}\) analysis of this contigency table showed that there is no relationship between sex observation and cluster assignment (\(\chi^{2}\): 1.12, df: 1, \textit{p}-value: 0.29).}
+  \label{tab:chi}
+\end{table}
+
 \begin{sidewaystable}
@@
 % latex table generated in R 3.0.1 by xtable 1.7-1 package
-% Thu Aug  8 00:20:39 2013
+% Thu Aug  8 14:11:16 2013
 {\small
@@
 % latex table generated in R 3.0.1 by xtable 1.7-1 package
-% Thu Aug  8 00:20:39 2013
+% Thu Aug  8 14:11:16 2013
 \begin{tabular}{r|rrrr}

### 2013-08-08 src/turtle_shape.r

@@ -1,10 +1,35 @@
 library(shapes)
 library(geomorph)
+library(cluster)
 
 source('../src/turtle_mung.r')
+
+load('../data/cluster_res.RData')
 
 # correlation between size and the first couple PCs
 cs <- centroid.size(turtle.land.adult)
 ipc <- turtle.adult[, 1:2]
 
-apply(ipc, 2, function(x) cor.test(cs, x))
+scor <- apply(ipc, 2, function(x) cor.test(cs, x))
+
+set.seed(1)
+gap.second <- pam(as.dist(turtle.adult.dist), k = 2)
+tclus <- gap.second$clustering
+
+tsex <- as.character(turtle.adult$p.sex)  # lots of regex to get this sexy
+fe <- grep(pattern = '[fF]', tsex, perl = TRUE)
+ma <- grep(pattern = '[mM]', tsex, perl = TRUE)
+na <- grep(pattern = '[nN]', tsex, perl = TRUE)
+tsex[fe] <- 'F'
+tsex[ma] <- 'M'
+tsex[na] <- NA
+tsex[tsex == ''] <- NA
+
+sexrm <- which(is.na(tsex))
+tclus.s <- tclus[-sexrm]
+tsex.s <- tsex[-sexrm]
+csex.tab <- table(tclus.s, tsex.s)
+csex.chi <- chisq.test(csex.tab)
+
+save(tsex, tclus.s, tsex.s, csex.tab, csex.chi,
+     file = '../data/turtle_shape.RData')

### 2013-08-08 src/turtle_tables.r

@@ -3,6 +3,13 @@ require(xtable)
 load('../data/turtle_analysis.RData')
 load('../data/turtle_gen.RData')
 
+# 2 x 2 table of sex - cluster assignment
+csex.tab <- rbind(csex.tab, tot = colSums(csex.tab))
+csex.tab <- cbind(csex.tab, tot = rowSums(csex.tab))
+xsex.tab <- xtable(csex.tab)
+align(xsex.tab) <- 'r|cc|c'
+digits(xsex.tab) <- c(0, 0, 0, 0)
+
 
 # multinomial logistic regression model selection tables
 

### 2013-08-08 src/turtle_plots.r
baja added to maps via google maps magic.

@@ -1,4 +1,7 @@
 require(ggplot2)
+require(mapproj) 
+require(maps)
+require(ggmap)
 require(reshape2)
 require(xtable)
 require(grid)
@@ -19,15 +22,26 @@ cbp <- c(#'#999999',
          '#E69F00', '#56B4E9', '#009E73', 
          '#F0E442', '#0072B2', '#D55E00', '#CC79A7')
 
-# map details
-map <- map_data('state')
-reg <- c('california', 'oregon')
-map <- map[map$region %in% reg, ]
 
+# map details
 turtle.adult$long[which(turtle.adult$long > -100)] <- turtle.adult$long[which(turtle.adult$long > -100)] - 100
 
 adult.train$spinks$long[which(adult.train$spinks$long > -100)] <- adult.train$spinks$long[which(adult.train$spinks$long > -100)] - 100
 
+
+longma <- max(turtle.adult$long)
+longmi <- min(turtle.adult$long)
+latma <- max(turtle.adult$lat)
+latmi <- min(turtle.adult$lat)
+
+goog.map <- get_map(location = c(longmi,
+                                 latmi, 
+                                 longma, 
+                                 latma), 
+                    zoom = 5,
+                    maptype = 'terrain',
+                    color = 'bw')
+ggmap(goog.map)
 
 # PCA results
 morph <- ggpairs(turtle.adult, columns = c(1:3),
@@ -101,7 +115,8 @@ set.seed(1)
 gap.second <- pam(as.dist(turtle.adult.dist), k = 2)
 gap.map <- map.plot(data = turtle.adult,
                     label = gap.second$clustering,
-                    map = map)
+                    map = goog.map)
+gap.map <- gap.map + xlim(longmi, longma) + ylim(latmi, latma)
 gap.map <- gap.map + scale_colour_manual(values = cbp)
 gap.map <- gap.map + theme(legend.position = 'none',
                            legend.text = element_text(size = 7),
@@ -184,9 +199,11 @@ ggsave(file = '../documents/figure/gen_res.png', plot = gdist)
 
 
 # best model map
-gg <- ggplot(map, aes(x = long, y = lat, group = group))
-gg <- gg + geom_polygon(fill = 'white', colour = 'black')
-gg <- gg + coord_map('gilbert')
+gg <- ggmap(goog.map)
+gg <- gg + xlim(longmi, longma) + ylim(latmi, latma)
[…]

### 2013-08-09 documents/turtle_paper.Rnw
lda added to paper. need to edit and expand certain sections.

@@
 
-Here, we used multiple alternative machine learning methods, both unsupervised and supervised, in order to compare different classification hypotheses.
+Here, we used multiple machine learning methods, both unsupervised and supervised, in order to compare different classification hypotheses.
 These methods provide different and unique advantages for understanding how to classify taxa, with what accuracy, and what these classifications are based on.
@@
 % model types
-Two different supervised learning methods were used to model the relationship between plastral shape and class: multinomial logistic regression and random forest.
+Three different supervised learning methods were used to model the relationship between plastral shape and class: linear discriminate analysis, multinomial logistic regression and random forest.
 These methods were chosen because of various properties of these methods which allow for useful interpretations about the quality and structure of the classification.
+
+% linear discriminate analysis
+Linear discriminate analysis (LDA) is a frequently applied method for characterizing the primary differences in morphology between different classes \citep{Zelditch2004,Mitteroecker2011}.
+This method attempts to find a linear combination of predictors to best model two or more classes.
+LDA is very similar to PCA except that instead of finding the linear combination of features that maximize the amount of explained variance in the data, LDA maximizes the differences between classes.
+The results of this analysis produces a transformation matrix by which the original features can be transformed to reflect the best discrimination between the classes.
+Like other supervised learning methods, LDA can also be used for predictive analysis on testing data.
+LDA was done using the \texttt{MASS} package for R \citep{Venables2002a}.
 
@@
 %     best AUC
-Supervised learning models have tuning parameters which help to increase the genearlizability of the model and prevent them from being overfit.
+The supervised learning models used here, except LDA, have tuning parameters which help to increase the genearlizability of the model and prevent them from being overfit.
 For the supervised learning models fit in this study, tuning parameters were estimated via 10 rounds of 10-fold cross-validation (CV) across a grid search of all tuning parameter combinations.
@@
 %   model selection
+%   LDA
+Normally, LDA is applied to the entire set of eigenscores from the PCA of shape \citep{Mitteroecker2011}.
+Instead here LDA was applied on the eigenscores from a subset of the total number of PCs, ranging from two to \Sexpr{floor(max.ad)} in increasing order of complexity.
+In total, this produced nine different LDA scaling matrices.
+From this set, the best number of PCs used to determine the LDA scaling matrix were chosen.
+As LDA is ``fit'' via maximum liklihood, the final combination of number of PCs and LDA scaling matrix chosen was that with the greatest AUC value from the training set.
+
 %     AICc
@@
 <<>>=
+ldasel <- unlist(lapply(tl.a.analysis$best, function(x) length(x$finalModel$xNames)))
+@
+The optimal number of PCs used for LDA, as determined by highest ROC score, for three of the four classification schemes had all \Sexpr{max(ldasel)} possible PCs (Fig.
+\ref{fig:roc}).
+These were both of the morphological based classification hypotheses and the second molecular hypothesis.
+LDA of the PCs of the first molecular hypothesis found that, based on ROC, only the first \Sexpr{min(ldasel)} PCs were necessary to best discriminate between the classes (Fig.
+\ref{fig:roc}).
+
+<<>>=
 daic <- unlist(lapply(lapply(tm.a.analysis$sel, function(x) x$delta), 
@@
 
-In the case of all models, there is a substantial increase in model performance as measured by AICc for the multinomial logistic models (Tables \ref{tab:mod_sel_1}, \ref{tab:mod_sel_2}, \ref{tab:mod_sel_3}, and \ref{tab:mod_sel_4}) or in AUC for the random forest models and illustrated for the multinomial logistic regression models as the number of features increases (Fig.
+In the case of all models, there is a substantial increase in model performance as measured by AICc for the multinomial logistic models (Tables \ref{tab:mod_sel_1}, \ref{tab:mod_sel_2}, \ref{tab:mod_sel_3}, and \ref{tab:mod_sel_4}) or in AUC for the LDA-based predictions and random forest models and illustrated for the multinomial logistic regression models as the number of features increases (Fig.
 \ref{fig:roc}).
@@
 \ref{fig:gen_res}).
-Remarkably, the best classification hypothesis was the second molecular classification hypothesis based on both the multinomial logistic regression and random forest models.
-For both methods, the distribution of bootstrapped AUC for the molecular hypothesis was significantly greater than all of the other classification schemes (Tables \ref{tab:mm_test} and \ref{tab:rf_test}).
+Remarkably, the best classification hypothesis was the second molecular classification hypothesis based on the LDA-based predictions, the multinomial logistic regression and random forest models.
+For both methods, the distribution of bootstrapped AUC for the molecular hypothesis was significantly greater than all of the other classification schemes (Tables \ref{tab:lda_test}, \ref{tab:mm_test} and \ref{tab:rf_test}).
 
-When the classification results of the training set for the best classification scheme based on the generalization results are compared with the references classes, the higher AUC value of the best multinomial logistic regression model compared to the best random forest model can be observed as the classifications are much closer to the reference classes (Fig.
+When the classification results of the training set for the best classification scheme based on the generalization results are compared with the references classes, the higher AUC value of the best results from LDA and the best multinomial logistic regression model compared to the best random forest model can be observed as the classifications are much closer to the reference classes (Fig.
 \ref{fig:gen_map}).
@@
 \ref{fig:imp_var}).
+
+The graphical results from the LDA of the training dataset for models of the second molecular classification scheme agree with the subtle distinctions between the different classes (Fig.
+\ref{fig:lda}).
+There is no clear distinction in terms of multivariate space between the four different classes.
+Instead, across all three axes there is substantial overlap as indicated by both the scatter of the points in space and the distribution of observations along each axis.
 
@@
 
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width = \textwidth]{figure/lda}
+  \caption{Pairs plots of the three discriminate axes from the linear discriminate analysis of the eigenscores from the first \Sexpr{max(ldasel)} PCs of plastral shape.
+The observations figured are from from the training data set used for all models for the second molecular classification hypothesis based on \citet{Spinks2005} and \citet{Spinks2010}.
+Observations are colored as in Fig.
+\ref{fig:gen_map}.}
+  \label{fig:lda}
+\end{figure}
+
 \pagebreak
@@
 
+\begin{table}
+  \centering
+<<results = 'asis'>>=
+print.xtable(lda.tabx,
+             hline.after = c(0, nrow(rf.tabx)),
+             includ.rownames = TRUE,
+             floating = FALSE)
+@
+  \caption{Results from pairwise Mann-Whitney U test between the AUC distributions of the generalizations of the LDA-based classification from the first 10 PCs of plastral shape.
+Labels correspond to those in Figure \ref{fig:gen_res}.
+Values of 0 correspond to p-values lower than 0.01.
+P-values were corrected for multiple comparison using the Holm method \citep{Holm1979}.}
+  \label{tab:lda_test}
+\end{table}
+
 \end{document}

### 2013-08-09 documents/turtle_paper.tex
baja added to maps via google maps magic.
lda added to paper. need to edit and expand certain sections.

@@
 %   model selection
+%   LDA
+Normally, LDA is applied to the entire set of eigenscores from the PCA of shape \citep{Mitteroecker2011}.
+Instead here LDA was applied on the eigenscores from a subset of the total number of PCs, ranging from two to 10 in increasing order of complexity.
+In total, this produced nine different LDA scaling matrices.
+From this set, the best number of PCs used to determine the LDA scaling matrix were chosen.
+As LDA is ``fit'' via maximum liklihood, the final combination of number of PCs and LDA scaling matrix chosen was that with the greatest AUC value from the training set.
+
 %     AICc
@@
 
+The optimal number of PCs used for LDA, as determined by highest ROC score, for three of the four classification schemes had all 10 possible PCs (Fig.
+\ref{fig:roc}).
+These were both of the morphological based classification hypotheses and the second molecular hypothesis.
+LDA of the PCs of the first molecular hypothesis found that, based on ROC, only the first 9 PCs were necessary to best discriminate between the classes (Fig.
+\ref{fig:roc}).
+
 % multinomial logistic regression model
@@
 
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width = \textwidth]{figure/lda}
+  \caption{Pairs plots of the three discriminate axes from the linear discriminate analysis of the eigenscores from the first 10 PCs of plastral shape.
+The observations figured are from from the training data set used for all models for the second molecular classification hypothesis based on \citet{Spinks2005} and \citet{Spinks2010}.
+Observations are colored as in Fig.
+\ref{fig:gen_map}.}
+  \label{fig:lda}
+\end{figure}
+
 \pagebreak
@@
 % latex table generated in R 3.0.1 by xtable 1.7-1 package
-% Thu Aug  8 14:11:16 2013
+% Fri Aug  9 17:34:31 2013
 \begin{tabular}{r|cc|c}
@@
 % latex table generated in R 3.0.1 by xtable 1.7-1 package
-% Thu Aug  8 14:11:16 2013
+% Fri Aug  9 17:34:31 2013
 {\small
@@
 % latex table generated in R 3.0.1 by xtable 1.7-1 package
-% Thu Aug  8 14:11:16 2013
+% Fri Aug  9 17:34:31 2013
 \begin{tabular}{r|rrrr}
@@
 
+\begin{table}
+  \centering
+% latex table generated in R 3.0.1 by xtable 1.7-1 package
+% Fri Aug  9 17:34:32 2013
+\begin{tabular}{r|rrrr}
+  & morph 1 & morph 2 & molec 1 & molec 2 \\ 
+  \hline
+morph 1 &  &  &  &  \\ 
+  morph 2 & 0.00 &  &  &  \\ 
+  molec 1 & 0.00 & 0.00 &  &  \\ 
+  molec 2 & 0.00 & 0.00 & 0.00 &  \\ 
+   \hline
+\end{tabular}
+
+  \caption{Results from pairwise Mann-Whitney U test between the AUC distributions of the generalizations of the LDA-based classification from the first 10 PCs of plastral shape.
+Labels correspond to those in Figure \ref{fig:gen_res}.
+Values of 0 correspond to p-values lower than 0.01.
+P-values were corrected for multiple comparison using the Holm method \citep{Holm1979}.}
+  \label{tab:lda_test}
+\end{table}
+
 \end{document}

### 2013-08-11 documents/turtle_paper.Rnw
turtle section revised and expanded.

@@
 This is concerning because the majority of extant taxa and virtually all extinct taxa are delimited entirely via morphology.
-Here we address questions about whether it is possible to determine, based on morphology, if one classification hypotheses can be considered better than others.
+Here we address questions about whether it is possible to determine, based on morphology, if one classification hypotheses can be considered better than others in order to determine if possible cryptic variation is actually cryptic or just a case of extremely fine scale morphological variation.
 %
@@
 This species has conflicting hypotheses of the number of meaningful subclades based on either morphological or molecular information.
-We compared multiple explicit classification hypotheses by characterizing variation in plastral shape and how it may be identifiably different between different classes.
-By splitting a large dataset of specimens into both training and testing datasets, we were also able to determine which of the classification hypotheses best corresponded to the observed plastral variation.
+We compared multiple explicit classification hypotheses by characterizing variation in plastral shape and how it may be identifiably different between classes.
+By splitting a large dataset of specimens into both training and testing datasets, we were also able to determine which of the classification hypotheses best corresponded to the observed plastral variation in general.
   %
   The results from our analysis shows that the best classification of plastral variation in \textit{Emys marmorata} is in accordance with the molecularly based hypothesis.
-This demonstrates that, by using alternative methods for characterizing variability, it is possible to estimate the classification scheme which most agrees with observed morphological variation.
-Additionally, we demonstrate that it is possible that not all examples of cryptic variation may be truly cryptic, just a product of sample size or methodology because of the extremely fine scale variation between the different classes.
+This demonstrates that, by using alternative methods for characterizing variability, it is possible to estimate the classification scheme which most agrees with observed variation.
+Additionally, we demonstrate how it is possible that not all examples of cryptic variation are truly cryptic and may just be a product of sample size or methodology because of the extremely fine scale variation between the different classes.
 
@@
 %   most species are still deliminated solely based on morphology
-Cryptic diversity is the phenomenon that not all taxa can be recognized based on morphology and can only be delimited using molecular information \citep{Stuart2006,Pfenninger2007,Funk2012,Clare2011}.
-Conceringly, most extant taxa, and nearly all extinct taxa, are delimited based solely via morphological analysis.
-This phenomenon is of great concern when studying variation and diversity dynamics over long periods of time, where apparent morphological stasis may not actually reflect true diversity \citep{Hunt2008,Eldredge1972,Gould1977a}.
+Cryptic diversity is the phenomenon that not all taxa can be recognized from morphology and can only be delimited using molecular information \citep{Stuart2006,Pfenninger2007,Funk2012,Clare2011}.
+Concerningly, most extant taxa, and nearly all extinct taxa, are delimited based solely via morphology.
+This phenomenon is of great concern when studying variation and diversity dynamics over long periods of time, where apparent morphological stasis may not actually reflect true diversity \citep{Hunt2008,Eldredge1972,Gould1977a,VanBocxlaer2013}.
 In the case of endangered or conserved taxa, morphometric approaches for classifying and identifying taxa and populations of importance would greatly improve the ability to maintain these high risk groups.
-Additionally, this would allow for better classifying extinct taxa.
+Additionally, this could lead to better classification of extinct taxa.
 
@@
 Here, we address the question of how can alternative approaches and methodology improve morphology based classification.
-From this approach, we ask if it is possible to determine which amongst a set of classification hypotheses is best.
+From this approach, we ask if it is possible to determine which amongst a set of classification hypotheses is best in order to determine if a case of cryptic diversity is truly cryptic or just a case of extremely fine scaled morphological variation.
 
@@
 Differences in morphological variation between different classes has previously been analyzed using methods like linear discriminate analysis and canonical variates analysis \citep{Zelditch2004,Mitteroecker2011,Polly2007a,Polly2003,Gunduz2007,Gaubert2005b,Demandt2009}.
-These methods are comparatively simple and straight forward ways of understanding the differences in morphology between classes.
-Also, they are very visual methods which aides in interpretation and presentation of information.
+These methods are comparatively straight forward ways of understanding the differences in morphology between classes.
+Also, they are very visual methods which aides with the interpretation and presentation of information.
 These previous studies, however, do not compare which amongst a set of candidate classification hypotheses is better.
-For example, studies such as those of \citet{Caumul2005a} and \citet{Polly2007a} focused on comparing different aspects of morphology and their fidelity to a classification scheme, instead of comparing the fidelity of one aspect of morphology to multiple classification shcemes.
-Also, there is no generalization of the classification model and training data accuracy is almost exclusively used as the metric off strength of classification.
+For example, studies such as those of \citet{Caumul2005a} and \citet{Polly2007a} focused on comparing different aspects of morphology and their fidelity to a classification scheme, instead of comparing the fidelity of one aspect of morphology to multiple classification schemes.
+Of note, however, is the work of \citet{Cardini2009a}, which compared morphological variation in marmots at both population, regional, and species levels to determine fidelity between shape each of these different hierarchical levels.
+Importantly, however, is that the classification models have not been generalized to testing data and training data accuracy is used almost exclusively as the metric off classification strength.
 
@@
 While machine learning methods such as neural networks have been applied to studying shape variation \citep{MacLeod2007}, they have been primarily applied in the context of automated taxon identification and not in terms of group classification and strength of classification.
-Additionally, we investigate variation in continuous traits, and do not search for discrete differences between each class, instead focusing on multivariate quantification of shape.
+Additionally, we investigate variation in continuous traits and not discrete differences between each class, instead focusing on differences in the multivariate quantification of shape.
 
@@
 
-Previous work on morphological differentiation in \textit{E.
-marmorata} focused on comparisons between different populations within the rage CITATIONS.
-Most of these studies were not over the entire range and used only linear measures of morphology instead of multivariate shape CITATIONS.
-%
-\citet{Holland1992} used comparisons of linear measures of the carapace of \textit{E.
-marmorata} from different populations over the entire range of the species.
-This study was interested in the relative effect of distance versus barriers had in terms of fostering morphological differentiation in \textit{E.
+Previous work on morphological variation in \textit{E.
+marmorata} has focused, primarily, on differentiation between different populations within a subset of the total species range \citep{Lubcke2007,Germano2009,Germano2008,Bury2010} with comparatively little done over the entire species range \citep{Holland1992}.
+These studies have focused on how local biotic and abiotic factors may contribute to differences in carapace length \citep{Lubcke2007,Germano2009,Germano2008} and found that size can vary greatly between different populations.
+
+Additionally, there has been found a great deal of evidence for sized-based sexual dimorphism in \textit{E.
+marmorata} \citep{Lubcke2007,Germano2009,Holland1992} with males being on average larger than females based on total carapace length and other linear measurements.
+However, the quality of size as a classifier of sex can vary greatly between populations \citep{Holland1992}, which makes sense in light of the amount of between population size difference \citep{Lubcke2007,Germano2009}.
+However, the effect of sexual dimorphism on shape, \textit{sensu} \citet{Kendall1977a}, was not assessed \citep{Holland1992,Lubcke2007,Germano2008}.
+
+Of particular note is the work of \citet{Holland1992} which compared morphological differences between and among many populations of \textit{E.
+marmorata} across the species range.
+\citet{Holland1992} studied the relative effect of distance versus barriers had in terms of fostering morphological differentiation in \textit{E.
 marmorata}.
 Analyses were performed to determine how different, morphologically, different populations in three different regions of the species range.
-Additionally, sexual dimorphism present in \textit{E.
-marmorata} was only assesed in the context of size measures such as length of carapace and mass.
-In general, \citet{Holland1992} and others \citep{Lubcke2007,Germano2008} found that males of \textit{E.
-marmorata} are larger than females in terms of size.
-However, the effect of sexual dimorphism on shape, \textit{sensu} \citet{Kendall1977a} and \citet{Dryden1998a}, was not assessed \citep{Holland1992,Lubcke2007,Germano2008}.
+Measurements were made from all different aspects of carapace morphology and not just total carapace length.
 
@@
 marmorata} was divided into four clades based on mitochondrial DNA: a northern clade, a southern clade, and eastern and western central Californian clades \citep{Spinks2005,Spinks2010}.
-While nuclear DNA supports two major clades, one northern and one southern, \citet{Spinks2010} argue that the four clade classification is of greater conservation utility.
+While nuclear DNA supports two major clades, one northern and one southern, \citet{Spinks2010} argue that the four clade classification is of greater conservation utility even though the variation between these groups is considered cryptic.
 While the mitochondrially based classification is considered robust, there is no known morphological differentiation between these clades.
@@
 In this study, we attempt to estimate the best classification scheme of \textit{E.
-marmorata} based on variation in plastral shape.
+marmorata} based on variation in plastral shape in order to determine if the molecular based hypothesis of \citet{Spinks2005} and \citet{Spinks2010} is actually a case of cryptic diversity or not.
 Because of unclear geographic boundaries between subgroups of \textit{E.
@@
 Because the specimens used to define the subclades in \citet{Spinks2005} and \citet{Spinks2010} were not available for study, all specimen classifications were based solely on this geographic information and not from explicit assignment in previous studies.
-The classifications here were based on matching museum locality data with the geographic boundaries of the moelcularly-defined clades of \citet{Spinks2005} and \citet{Spinks2010}.
+Instead, classification was based on matching museum locality data with the geographic boundaries of the molecularly-defined clades of \citet{Spinks2005} and \citet{Spinks2010}.
 Because the exact barriers between different biogeographic regions are unknown and unclear, two assignments for both the morphologically and molecularly based hypotheses were used.
@@
 %   LDA
-Normally, LDA is applied to the entire set of eigenscores from the PCA of shape \citep{Mitteroecker2011}.
-Instead here LDA was applied on the eigenscores from a subset of the total number of PCs, ranging from two to \Sexpr{floor(max.ad)} in increasing order of complexity.
+LDA was applied on the eigenscores from a subset of the total number of PCs, ranging from two to \Sexpr{floor(max.ad)} in increasing order of complexity.
 In total, this produced nine different LDA scaling matrices.
-From this set, the best number of PCs used to determine the LDA scaling matrix were chosen.
-As LDA is ``fit'' via maximum liklihood, the final combination of number of PCs and LDA scaling matrix chosen was that with the greatest AUC value from the training set.
+From this set, the best number of PCs used to estimate the LDA scaling matrix were chosen.
+As LDA is not ``fit'' via maximum liklihood, the final combination of number of PCs and LDA scaling matrix chosen was that with the greatest AUC value from the training set.
 
@@
 
-The first two PCs appear to describe principally varaition in the lateral margin of te palstra, from a pointed medial edge to a more rounded and blunt edge (Fig.
+The first two PCs appear to describe principally variation in the lateral margin of the palstra, from a pointed medial edge to a more rounded and blunt edge (Fig.
 \ref{fig:pc_var}).
@@
 \ref{fig:pca} and \ref{fig:pc_var}), is positioned on the bridge between the plastron and the carapace.
-Over ontogeny, this is an area that deepens dorsoventrally and when the plastron was projected into two dimensions this it created the effect of mediolateral movement.
+Over ontogeny, this is an area that deepens dorsoventrally and when the plastron was projected into two dimensions it created the effect of mediolateral movement.
 Lateral landmark variation along the first PC seemed concentrated in the posterior portion of the plastra with additional variance observed in midline landmarks (Fig.
 \ref{fig:pc_var}).
-This variance in midline landmakrs was most likely caused by the fact that plastral scutes frequently do not line up perfectly.
-Along PC 2, lateral variation appeared to be concetrated in the anterior portion o the plastra (Fig.
+This variance in midline landmarks was most likely caused by the fact that plastral scutes frequently do not line up perfectly.
+Along PC 2, lateral variation appeared to be concentrated in the anterior portion o the plastra (Fig.
 \ref{fig:pc_var}).
@@
 This results is interesting because while sexual dimorphism has been observed in linear measures and mass estimates of \textit{E.
-marmorata}, this results demonstrates that this dimorphism may not translate into differences in shape.
+marmorata} \citep{Lubcke2007,Germano2009,Holland1992}, this results demonstrates that this dimorphism may not translate into differences in shape.
 Interestsingly, male emydid turtles are known to have a plastral concavity which may influence landmark position along the midline.
@@
 The gap statistic values for both three and four clusters are much lower than for one and two and are statistically identical.
-Interestingly, other solutions with a much greater number of clusters have higher gap statistic values though these are also not significantly different.
+Interestingly, other solutions with a much greater number of clusters have relatively high gap statistic values as well though these are also not significantly different.
 Increasing the number of clusters does appear to improve the gap statistic enough compared to the best clustering solution to merit detailed discussion.
@@
 \ref{fig:roc}).
+The first \Sexpr{length(tm.a.analysis$best$spinks$coefnames) - 1} PCs describe \Sexpr{sum(turtle.fit$percent[1:9])}\% of total variation in plastral shape, while the first \Sexpr{length(tm.a.analysis$best$spinks$coefnames)} PCs describe \Sexpr{sum(turtle.fit$percent[1:10])}\% of the variation.
 
@@
 The \(\Delta\)AICc values between the optimal and second best model range from \Sexpr{min(daic)} for the first morphological based classification hypothesis to \Sexpr{max(daic)} for the second molecular based classification hypothesis (Tables \ref{tab:mod_sel_1}, \ref{tab:mod_sel_2}, \ref{tab:mod_sel_3}, and \ref{tab:mod_sel_4}).
-The first \Sexpr{length(tm.a.analysis$best$spinks$coefnames) - 1} PCs describe \Sexpr{sum(turtle.fit$percent[1:9])}\% of total variation in plastral shape, while the first \Sexpr{length(tm.a.analysis$best$spinks$coefnames)} PCs describe \Sexpr{sum(turtle.fit$percent[1:10])}\% of the variation.
 
@@
 
-The selected number of features in the final random forest model for each classification scheme was very simplar to the model selection results from the multinomial logistic regression models, ranging from \Sexpr{min(oprf)} for the second morphological based classification hypothesis and both molecular based classification hypotheses to \Sexpr{max(oprf)} for the first morphological based classification hypothesis (Fig.
+The selected number of features in the final random forest model for each classification scheme was very simpler to the model selection results for the LDA-based classification and the multinomial logistic regression models, ranging from \Sexpr{min(oprf)} for the second morphological based classification hypothesis and both molecular based classification hypotheses to \Sexpr{max(oprf)} for the first morphological based classification hypothesis (Fig.
 \ref{fig:roc}).
@@
 The best random forest model misclassified many of the observations as the northern clade instead of the correct class.
-This pattern of misclassification is observable but not as exaggerated in the classifications of the multinomial logistic regression model.
+This pattern of misclassification is observable but not as exaggerated in the LDA-based classifications and those from the multinomial logistic regression model (Fig.
+\ref{fig:gen_map}).
 
@@
 The only unsupervised method employed in this study, PAM, is rather simple and not model based.
-A more useful approach would be to employ various model based clustering approaches \citep{Fraley2002,Zhong2003}.
+A more useful approach would be to employ various model based clustering approaches \citep{Fraley2002,Zhong2003,VanBocxlaer2013}.
 In this manner, a series of candidate models can be compared via model comparison methods, such as AIC or Bayes factors \citep{Fraley2002}, in order to asses the best clustering solution.
@@
 Sweet for field assistance and the California Department of Fish and Game for permits.
-Much of the data collection was funded by NSD DBI-0306158 (to KDA).
+Much of the data collection was funded by NSF DBI-0306158 (to KDA).
 
@@
 As the total number of features increase, AUC increases until eventually leveling off.
-Both multinomial logisitic regression and random forest models are illustrated here, though AUC based model selection was only performed for random forest models.}
+LDA-based classification, multinomial logisitic regression and random forest models are illustrated here, though AUC based model selection was only performed for the LDA-based classification and the random forest models.}
   \label{fig:roc}
@@
   \caption{Density estimates of AUC values of predictions of the testing dataset of plastra from \Sexpr{rr$sh1$R} bootstrap resamples.
-The top facet corresponds to values using the optimal multinomial logistic regression model, as chosen by minimum AICc value.
+The top facet corresponds to values using the best LDA-based classifications of the eigenscores of shape, as chosen by maximum AUC.
+The middle facet corresponds to values using the optimal multinomial logistic regression model, as chosen by minimum AICc value.
 The bottom facet corresponds to the values using the optimal random forest model, as chosen by maximum AUC value.}
@@
   \includegraphics[width = \textwidth]{figure/gen_map}
-  \caption{Comparison between reference classification of testing data set and the estimated classifications based on the selected multinomial logistic regression and random forest models, from left to right respectively.
+  \caption{Comparison between reference classification of testing data set (upper left) and the estimated classifications based on the selected LDA-based classification (lda, upper right), multinomial logistic regression (multi, lower left) and random forest models (rf, lower right).
 Classification corresponds to the four classes as suggested by the hypothesis of \citet{Spinks2005} and \citet{Spinks2010}.}
@@
 This hypothesis is based on \citet{Seeliger1945}.
-The column ``delta'' corresponds to the \(\delta\)AICc values of each model, while ``weights'' correspond to the Akaike weight of that model relative to all others.}
+The column ``delta'' corresponds to the \(\Delta\)AICc values of each model, while ``weights'' correspond to the Akaike weight of that model relative to all others.}
   \label{tab:mod_sel_1}
@@
 This hypothesis is based on \citet{Seeliger1945}.
-The column ``delta'' corresponds to the \(\delta\)AICc values of each model, while ``weights'' correspond to the Akaike weight of that model relative to all others.}
+The column ``delta'' corresponds to the \(\Delta\)AICc values of each model, while ``weights'' correspond to the Akaike weight of that model relative to all others.}
   \label{tab:mod_sel_2}
@@
 This hypothesis is based on \citet{Seeliger1945}.
-The column ``delta'' corresponds to the \(\delta\)AICc values of each model, while ``weights'' correspond to the Akaike weight of that model relative to all others.}
+The column ``delta'' corresponds to the \(\Delta\)AICc values of each model, while ``weights'' correspond to the Akaike weight of that model relative to all others.}
   \label{tab:mod_sel_3}
@@
 This hypothesis is based on \citet{Seeliger1945}.
-The column ``delta'' corresponds to the \(\delta\)AICc values of each model, while ``weights'' correspond to the Akaike weight of that model relative to all others.}
+The column ``delta'' corresponds to the \(\Delta\)AICc values of each model, while ``weights'' correspond to the Akaike weight of that model relative to all others.}
   \label{tab:mod_sel_4}
 \end{sidewaystable}
+
+\begin{table}
+  \centering
+<<results = 'asis'>>=
+print.xtable(lda.tabx,
+             hline.after = c(0, nrow(rf.tabx)),
+             includ.rownames = TRUE,
+             floating = FALSE)
+@
+  \caption{Results from pairwise Mann-Whitney U test between the AUC distributions of the generalizations of the LDA-based classification from the first 10 PCs of plastral shape.
+Labels correspond to those in Figure \ref{fig:gen_res}.
+Values of 0 correspond to p-values lower than 0.01.
+P-values were corrected for multiple comparison using the Holm method \citep{Holm1979}.}
+  \label{tab:lda_test}
+\end{table}
 
@@
 
-\begin{table}
-  \centering
-<<results = 'asis'>>=
-print.xtable(lda.tabx,
-             hline.after = c(0, nrow(rf.tabx)),
-             includ.rownames = TRUE,
-             floating = FALSE)
-@
-  \caption{Results from pairwise Mann-Whitney U test between the AUC distributions of the generalizations of the LDA-based classification from the first 10 PCs of plastral shape.
-Labels correspond to those in Figure \ref{fig:gen_res}.
-Values of 0 correspond to p-values lower than 0.01.
-P-values were corrected for multiple comparison using the Holm method \citep{Holm1979}.}
-  \label{tab:lda_test}
-\end{table}
-
 \end{document}

### 2013-08-11 documents/turtle_paper.tex
turtle section revised and expanded.

@@
 %   LDA
-Normally, LDA is applied to the entire set of eigenscores from the PCA of shape \citep{Mitteroecker2011}.
-Instead here LDA was applied on the eigenscores from a subset of the total number of PCs, ranging from two to 10 in increasing order of complexity.
+LDA was applied on the eigenscores from a subset of the total number of PCs, ranging from two to 10 in increasing order of complexity.
 In total, this produced nine different LDA scaling matrices.
-From this set, the best number of PCs used to determine the LDA scaling matrix were chosen.
-As LDA is ``fit'' via maximum liklihood, the final combination of number of PCs and LDA scaling matrix chosen was that with the greatest AUC value from the training set.
+From this set, the best number of PCs used to estimate the LDA scaling matrix were chosen.
+As LDA is not ``fit'' via maximum liklihood, the final combination of number of PCs and LDA scaling matrix chosen was that with the greatest AUC value from the training set.
 
@@
 \ref{fig:roc}).
+The first 9 PCs describe 83.23\% of total variation in plastral shape, while the first 10 PCs describe 86.54\% of the variation.
 
@@
 The \(\Delta\)AICc values between the optimal and second best model range from 1.18 for the first morphological based classification hypothesis to 26.51 for the second molecular based classification hypothesis (Tables \ref{tab:mod_sel_1}, \ref{tab:mod_sel_2}, \ref{tab:mod_sel_3}, and \ref{tab:mod_sel_4}).
-The first 9 PCs describe 83.23\% of total variation in plastral shape, while the first 10 PCs describe 86.54\% of the variation.
 
@@
 
-The selected number of features in the final random forest model for each classification scheme was very simplar to the model selection results from the multinomial logistic regression models, ranging from 9 for the second morphological based classification hypothesis and both molecular based classification hypotheses to 10 for the first morphological based classification hypothesis (Fig.
+The selected number of features in the final random forest model for each classification scheme was very simpler to the model selection results for the LDA-based classification and the multinomial logistic regression models, ranging from 9 for the second morphological based classification hypothesis and both molecular based classification hypotheses to 10 for the first morphological based classification hypothesis (Fig.
 \ref{fig:roc}).
@@
   \caption{Density estimates of AUC values of predictions of the testing dataset of plastra from 1000 bootstrap resamples.
-The top facet corresponds to values using the optimal multinomial logistic regression model, as chosen by minimum AICc value.
+The top facet corresponds to values using the best LDA-based classifications of the eigenscores of shape, as chosen by maximum AUC.
+The middle facet corresponds to values using the optimal multinomial logistic regression model, as chosen by minimum AICc value.
 The bottom facet corresponds to the values using the optimal random forest model, as chosen by maximum AUC value.}
@@
 % latex table generated in R 3.0.1 by xtable 1.7-1 package
-% Fri Aug  9 17:34:31 2013
+% Sat Aug 10 19:53:26 2013
 \begin{tabular}{r|cc|c}
@@
 % latex table generated in R 3.0.1 by xtable 1.7-1 package
-% Fri Aug  9 17:34:31 2013
+% Sat Aug 10 19:53:26 2013
 {\small
@@
 This hypothesis is based on \citet{Seeliger1945}.
-The column ``delta'' corresponds to the \(\delta\)AICc values of each model, while ``weights'' correspond to the Akaike weight of that model relative to all others.}
+The column ``delta'' corresponds to the \(\Delta\)AICc values of each model, while ``weights'' correspond to the Akaike weight of that model relative to all others.}
   \label{tab:mod_sel_4}
@@
 % latex table generated in R 3.0.1 by xtable 1.7-1 package
-% Fri Aug  9 17:34:31 2013
+% Sat Aug 10 19:53:26 2013
+\begin{tabular}{r|rrrr}
+  & morph 1 & morph 2 & molec 1 & molec 2 \\ 
+  \hline
+morph 1 &  &  &  &  \\ 
+  morph 2 & 0.00 &  &  &  \\ 
+  molec 1 & 0.00 & 0.00 &  &  \\ 
+  molec 2 & 0.00 & 0.00 & 0.00 &  \\ 
+   \hline
+\end{tabular}
+
+  \caption{Results from pairwise Mann-Whitney U test between the AUC distributions of the generalizations of the LDA-based classification from the first 10 PCs of plastral shape.
+Labels correspond to those in Figure \ref{fig:gen_res}.
+Values of 0 correspond to p-values lower than 0.01.
+P-values were corrected for multiple comparison using the Holm method \citep{Holm1979}.}
+  \label{tab:lda_test}
+\end{table}
+
+\begin{table}
+  \centering
+% latex table generated in R 3.0.1 by xtable 1.7-1 package
+% Sat Aug 10 19:53:26 2013
 \begin{tabular}{r|rrrr}
@@
 % latex table generated in R 3.0.1 by xtable 1.7-1 package
-% Fri Aug  9 17:34:31 2013
+% Sat Aug 10 19:53:26 2013
 \begin{tabular}{r|rrrr}
@@
 
-\begin{table}
-  \centering
-% latex table generated in R 3.0.1 by xtable 1.7-1 package
-% Fri Aug  9 17:34:32 2013
-\begin{tabular}{r|rrrr}
-  & morph 1 & morph 2 & molec 1 & molec 2 \\ 
-  \hline
-morph 1 &  &  &  &  \\ 
-  morph 2 & 0.00 &  &  &  \\ 
-  molec 1 & 0.00 & 0.00 &  &  \\ 
-  molec 2 & 0.00 & 0.00 & 0.00 &  \\ 
-   \hline
-\end{tabular}
-
-  \caption{Results from pairwise Mann-Whitney U test between the AUC distributions of the generalizations of the LDA-based classification from the first 10 PCs of plastral shape.
-Labels correspond to those in Figure \ref{fig:gen_res}.
-Values of 0 correspond to p-values lower than 0.01.
-P-values were corrected for multiple comparison using the Holm method \citep{Holm1979}.}
-  \label{tab:lda_test}
-\end{table}
-
 \end{document}

### 2013-08-11 src/turtle_plots.r

@@ -218,7 +218,7 @@ gg <- gg + xlim(longmi, longma) + ylim(latmi, latma)
 tr <- cbind(long = adult.test$spinks$long, 
             lat = adult.test$spinks$lat, 
             label = adult.test$spinks$spinks)
-tr <- cbind(data.frame(tr), type = rep('training', nrow(tr)))
+tr <- cbind(data.frame(tr), type = rep('testing', nrow(tr)))
 te <- cbind(long = adult.test$spinks$long,
             lat = adult.test$spinks$lat)
 mty <- c(rep('multi', nrow(te)), rep('rf', nrow(te)), rep('lda', nrow(te)))

### 2013-08-11 src/plotting_functions.r
added secondary segments to plastral plots

@@ -37,9 +37,9 @@ gap.plot <- function(gap) {
   gg
 }
 
-mshape.plot <- function(fits, links) {
-  shape <- as.data.frame(fits$mshape)
-  shape <- shape[links, ]
+mshape.plot <- function(fits, links, snd.links) {
+  shape1 <- as.data.frame(fits$mshape)
+  shape <- shape1[links, ]
   shape <- cbind(shape, rbind(shape[-1, ], shape[1, ]))
   names(shape) <- c('V1', 'V2', 'V3', 'V4')
 
@@ -51,6 +51,19 @@ mshape.plot <- function(fits, links) {
                                                       y = -V1,
                                                       yend = -V3),
                           alpha = 0.5)
+
+  snd.shape <- shape1[snd.links, ]
+  ss <- seq(from = 1, to = nrow(snd.shape), by = 2)
+  snd.shape <- cbind(snd.shape[ss, ], snd.shape[ss + 1, ])
+  names(snd.shape) <- names(shape)
+  for(ii in seq(from = 1, to = length(snd.links), by = 2)) {
+    gg <- gg + geom_segment(data = snd.shape[c(ii, ii + 1), ], 
+                            mapping = aes(x = V2,
+                                          xend = V4,
+                                          y = -V1,
+                                          yend = -V3))
+  }
+                                                          
   gg <- gg + geom_point(data = shape, mapping = aes(x = V2,
                                                     y = -V1,
                                                     colour = 'red'),

### 2013-08-12 documents/turtle_paper.Rnw
made tps grids prettier. ellipses.

@@
 marmorata}.
-Clustering was based entirely on the \(\rho\) dissimilarity matrix of ``half'' plastral landmark configurations following Procrustes superimposition.}
+Clustering was based entirely on the \(\rho\) dissimilarity matrix of ``half'' plastral landmark configurations following Procrustes superimposition.
+Point shapes correspond to the two clusters while the colored ellipses correspond to 95\% confidence ellipses of the four groups from \citet{Spinks2005} and \citet{Spinks2010}.}
   \label{fig:gap_map}
@@
   \caption{Comparison between reference classification of testing data set (upper left) and the estimated classifications based on the selected LDA-based classification (lda, upper right), multinomial logistic regression (multi, lower left) and random forest models (rf, lower right).
-Classification corresponds to the four classes as suggested by the hypothesis of \citet{Spinks2005} and \citet{Spinks2010}.}
+Classification corresponds to the four classes as suggested by the hypothesis of \citet{Spinks2005} and \citet{Spinks2010}.
+95\% confidence ellipses for the four classes are depicted and calculated from the initial assignments from the total dataset.}
   \label{fig:gen_map}
@@
     \centering
-    \includegraphics[width = \textwidth]{figure/mshape_1}
+    \includegraphics[width = \textwidth, height = 0.35\textheight, keepaspectratio = true]{figure/mshape_1}
     \caption{Northern}
@@
     \centering
-    \includegraphics[width = \textwidth]{figure/mshape_2}
+    \includegraphics[width = \textwidth, height = 0.35\textheight, keepaspectratio = true]{figure/mshape_2}
     \caption{Eastern}
@@
     \centering
-    \includegraphics[width = \textwidth]{figure/mshape_3}
+    \includegraphics[width = \textwidth, height = 0.35\textheight, keepaspectratio = true]{figure/mshape_3}
     \caption{Western}
@@
     \centering
-    \includegraphics[width = \textwidth]{figure/mshape_4}
+    \includegraphics[width = \textwidth, height = 0.35\textheight, keepaspectratio = true]{figure/mshape_4}
     \caption{Southern}

### 2013-08-12 documents/turtle_paper.tex
added secondary segments to plastral plots
made tps grids prettier. ellipses.

@@
 % latex table generated in R 3.0.1 by xtable 1.7-1 package
-% Sat Aug 10 19:53:26 2013
+% Mon Aug 12 13:45:51 2013
 \begin{tabular}{r|cc|c}
@@
 % latex table generated in R 3.0.1 by xtable 1.7-1 package
-% Sat Aug 10 19:53:26 2013
+% Mon Aug 12 13:45:51 2013
 {\small
@@
 % latex table generated in R 3.0.1 by xtable 1.7-1 package
-% Sat Aug 10 19:53:26 2013
+% Mon Aug 12 13:45:51 2013
 \begin{tabular}{r|rrrr}

### 2013-08-12 documents/turtle_paper.Rnw
background section updated to reflect ken comments. draft finished.

@@
 \subsection{Background and system}
+Machine learning is in many respects just an extension of known statistical methodology \citep{Hastie2009} where the emphasis is placed on inferring rules and properties of data in order to explain the underlying structure.
+The basic statistical mechanics are supplemented by randomization, sorting, and partitioning algorithms and along with the maximization or minimization of summary statistics in order to best estimate a general model for all data, both sampled and unsampled \citep{Hastie2009}.
+Machine learning approaches have found use in medical research, epidemiology, economics and automated image identification such as handwritten zip codes \citep{Hastie2009}.
+Two major classes of machine learning methods are unsupervised and supervised learning.
+Unsupervised learning methods are used with unlabeled data where the underlying structure is estimated and are analogous to clustering and density estimation methods \citep{Kaufman1990}.
+Supervised learning methods are used with labeled data where the final output of data is known and the rules for going from input to output are inferred.
+These are analogous classification and regression models \citep{Breiman1984}.
+The application of the alternative approaches used in this study illustrates only a sampling of the various previously derived methods for clustering observations and fitting classification models.
+
 Differences in morphological variation between different classes has previously been analyzed using methods like linear discriminate analysis and canonical variates analysis \citep{Zelditch2004,Mitteroecker2011,Polly2007a,Polly2003,Gunduz2007,Gaubert2005b,Demandt2009}.
@@
 Also, they are very visual methods which aides with the interpretation and presentation of information.
-These previous studies, however, do not compare which amongst a set of candidate classification hypotheses is better.
+Previous studies, however, normally do not compare which amongst a set of candidate classification hypotheses is better.
 For example, studies such as those of \citet{Caumul2005a} and \citet{Polly2007a} focused on comparing different aspects of morphology and their fidelity to a classification scheme, instead of comparing the fidelity of one aspect of morphology to multiple classification schemes.
@@
 Additionally, we investigate variation in continuous traits and not discrete differences between each class, instead focusing on differences in the multivariate quantification of shape.
-
-% this paragraph kind of sucks
-The two major classes of machine learning methods, unsupervised and supervised, are essentially extensions of known statistical methods \citep{Hastie2009}.
-Unsupervised learning methods are analogous to clustering and density estimation methods \citep{Kaufman1990}, while supervised learning methods are analogous classification and regression models \citep{Breiman1984}.
-In both cases, many of these methods are not fit via maximum likelihood and are supplemented by randomization, sorting, and partitioning algorithms along with the maximization or minimization of summary statistics in order to best estimate a general model for all data, both sampled and unsampled \citep{Hastie2009}.
-The application of the alternative approaches used in this study illustrates only a sampling of the various previously derived methods for clustering observations and fitting classification models.
-% Applications of these methods outside of this study
-% epidemeology, computer science, marketing
-Additionally, instead of pure classification accuracy, here we use a statistic of classification strength that reflects the rate at which taxa are both accurately and inaccurately classified (see Methods).
+Also, instead of pure classification accuracy, here we used a statistic of classification strength that reflects the rate at which taxa are both accurately and inaccurately classified (see Methods).
 

### 2013-08-12 documents/turtle_paper.tex
background section updated to reflect ken comments. draft finished.

@@
 % latex table generated in R 3.0.1 by xtable 1.7-1 package
-% Mon Aug 12 13:45:51 2013
+% Mon Aug 12 15:26:33 2013
 \begin{tabular}{r|cc|c}
@@
 % latex table generated in R 3.0.1 by xtable 1.7-1 package
-% Mon Aug 12 13:45:51 2013
+% Mon Aug 12 15:26:33 2013
 {\small
@@
 % latex table generated in R 3.0.1 by xtable 1.7-1 package
-% Mon Aug 12 13:45:51 2013
+% Mon Aug 12 15:26:33 2013
 \begin{tabular}{r|rrrr}

### 2013-08-21 documents/turtle_paper.Rnw
centroid size added to manuscript

@@
 options(digits = 2)
-
-#knit_hooks$set(inline = function(x) {
-#               if (is.numeric(x)) round(x, 4)})
 
@@
 % size correlation with either of the first two PCs?
+When the eigenscores of the first two PCs are compared to the centroid sizes of the observations, a significant correlation is observed between centroid size and the first PC (Spearman's \(\rho\): \Sexpr{scor$PC1$estimate}, \textit{p}-value: ~\Sexpr{scor$PC1$p.value}) but not between centroid size and the second PC (Spearman's \(\rho\): \Sexpr{scor$PC2$estimate}, \textit{p}-value: ~\Sexpr{scor$PC2$p.value}).
+This difference is clearly observable in in plots of the data (Fig.
+\ref{fig:cent}).
 
@@
 \ref{fig:mean_shape}) where major class differences seem based on the relative ballooning or shrinking of the anterior and posterior portions of the plastron together along with differential ``pinching'' of the midline landmarks.
+It is important to note that the first two most important axes, according to the random forest model, were not correlated with size.
+This can be interpreted to mean that class membership is principally not related to turtle size.
+This interpretation makes sense in light of previous work showing that body size in \textit{E.
+marmorata} can vary widely across the entire species range \citep{Holland1992,Lubcke2007,Germano2009}.
 
@@
 \ref{fig:imp_pc} and \ref{fig:imp_var}).
+Importantly, the order of variable importance demonstrated that the two most important axes (PC 2 and PC 3) for determining class membership were not correlated with size, unlike the third most important axis (PC 1), indicating that class membership is not strongly related to size as opposed to being related to actual shape differences.
 Additionally, the relative risk values from the mulitinomial logistic regression models demonstrate that a single PC is probably not sufficient for estimating the class of an observation, but that given a set of PCs this classification would be more accurate (Fig.
@@
   \label{fig:pc_var}
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width = \textwidth]{figure/cent_size}
+  \caption{Comparison of the centroid size of each observation to the eigenscore of that obsrevation from the PCA of plastral shape.
+The first two PCs are illustrated from right to left.
+The regression line is fit via ordinary least squares and is presented to illustrate the difference in correlation between the first two axes.
+The shaded regions represent 95\% confidence intervals on the parameters of the regression line.
+Centroid size is significantly correlated with the first PC but not the second.}
+  \label{fig:cent}
 \end{figure}

### 2013-08-21 documents/turtle_paper.tex
centroid size added to manuscript

@@
 % size correlation with either of the first two PCs?
+When the eigenscores of the first two PCs are compared to the centroid sizes of the observations, a significant correlation is observed between centroid size and the first PC (Spearman's \(\rho\): -0.56, \textit{p}-value: ~\ensuremath{7.51\times 10^{-44}}) but not between centroid size and the second PC (Spearman's \(\rho\): 0.03, \textit{p}-value: ~0.55).
+This difference is clearly observable in in plots of the data (Fig.
+\ref{fig:cent}).
 
@@
 % latex table generated in R 3.0.1 by xtable 1.7-1 package
-% Mon Aug 12 15:26:33 2013
+% Wed Aug 21 14:14:34 2013
 \begin{tabular}{r|cc|c}
@@
 % latex table generated in R 3.0.1 by xtable 1.7-1 package
-% Mon Aug 12 15:26:33 2013
+% Wed Aug 21 14:14:34 2013
 {\small
@@
 % latex table generated in R 3.0.1 by xtable 1.7-1 package
-% Mon Aug 12 15:26:33 2013
+% Wed Aug 21 14:14:34 2013
 \begin{tabular}{r|rrrr}

### 2013-08-21 src/turtle_plots.r

@@ -18,6 +18,7 @@ source_url('https://raw.github.com/JoFrhwld/FAAV/master/r/stat-ellipse.R')
 
 load('../data/turtle_analysis.RData')
 load('../data/turtle_gen.RData')
+load('../data/turtle_shape.RData')
 
 theme_set(theme_bw())
 

### 2013-09-04 documents/turtle_paper.Rnw
added better references on crypsis

@@
 %   most species are still deliminated solely based on morphology
-Cryptic diversity is the phenomenon that not all taxa can be recognized from morphology and can only be delimited using molecular information \citep{Stuart2006,Pfenninger2007,Funk2012,Clare2011}.
+Cryptic diversity is the phenomenon that not all taxa can be recognized from morphology and can only be delimited using molecular information \citep{Bickford2007,SchlickSteiner2007,Stuart2006,Pfenninger2007,Funk2012,Clare2011}.
 Concerningly, most extant taxa, and nearly all extinct taxa, are delimited based solely via morphology.

### 2014-03-07 doc/turtle_paper.Rnw
improved some tables in paper

--- 2013-09-04 documents/turtle_paper.Rnw
@@
   \centering
-<<results = 'asis'>>=
-print.xtable(lda.tabx,
-             hline.after = c(0, nrow(rf.tabx)),
-             includ.rownames = TRUE,
-             floating = FALSE)
-@
+  \include{lda_tabx}
   \caption{Results from pairwise Mann-Whitney U test between the AUC distributions of the generalizations of the LDA-based classification from the first 10 PCs of plastral shape.
@@
   \centering
-<<results = 'asis'>>=
-print.xtable(mm.tabx,
-             hline.after = c(0, nrow(mm.tabx)),
-             includ.rownames = TRUE,
-             floating = FALSE)
-@
+  \include{mm_tabx}
   \caption{Results from pairwise Mann-Whitney U test between the AUC distributions of the generalizations of the multinomial logistic regression models.
@@
   \centering
-<<results = 'asis'>>=
-print.xtable(rf.tabx,
-             hline.after = c(0, nrow(rf.tabx)),
-             includ.rownames = TRUE,
-             floating = FALSE)
-@
+  \include{rf_tabx}
   \caption{Results from pairwise Mann-Whitney U test between the AUC distributions of the generalizations of the random forest models.

### 2014-03-07 R/make_tables.r

@@ -48,6 +48,11 @@ colnames(mm.tab) <- coln
 rownames(mm.tab) <- rown
 mm.tabx <- xtable(as.table(mm.tab))
 align(mm.tabx) <- 'r|rrrr'
+print.xtable(mm.tabx, 
+             file = '../doc/mm_tabx.tex',
+             hline.after = c(0, nrow(mm.tabx)),
+             include.rownames = TRUE,
+             floating = FALSE)
 
 # random forests
 rf.comp <- cbind(sh1 = rr$sh1$t, sh2 = rr$sh2$t, sh3 = rr$sh3$t, spinks = rr$spinks$t)
@@ -61,6 +66,11 @@ colnames(rf.tab) <- coln
 rownames(rf.tab) <- rown
 rf.tabx <- xtable(as.table(rf.tab))
 align(rf.tabx) <- 'r|rrrr'
+print.xtable(rf.tabx, 
+             file = '../doc/rf_tabx.tex',
+             hline.after = c(0, nrow(mm.tabx)),
+             include.rownames = TRUE,
+             floating = FALSE)
 
 # linear discriminate analysis
 lda.comp <- cbind(sh1 = ll[[1]]$t, sh2 = ll[[2]]$t, 
@@ -75,3 +85,8 @@ colnames(lda.tab) <- coln
 rownames(lda.tab) <- rown
 lda.tabx <- xtable(as.table(lda.tab))
 align(lda.tabx) <- 'r|rrrr'
+print.xtable(lda.tabx, 
+             file = '../doc/lda_tabx.tex',
+             hline.after = c(0, nrow(mm.tabx)),
+             include.rownames = TRUE,
+             floating = FALSE)

### 2014-03-07 doc/turtle_paper.Rnw
more table fixes and cleaning sexpr

@@
 
-load('../data/turtle_analysis.RData')
-load('../data/turtle_gen.RData')
-load('../data/turtle_shape.RData')
-
-source('../src/turtle_plots.r')
-source('../src/turtle_tables.r')
+load('../data/gen.RData')
+load('../data/shape.RData')
 
 options(digits = 2)
-
 @
@@
 \subsection{Specimens}
-We collected landmark-based morphometric data from \Sexpr{nrow(turtle.adult)} adult \textit{E.
+We collected landmark-based morphometric data from \Sexpr{nrow(adult)} adult \textit{E.
 marmorata} museum specimens.
@@
 %  R implementation
-This analysis was conducted using the \texttt{cluster} package for R \citep{Maechler2013} using all \Sexpr{nrow(turtle.adult)} observations.
+This analysis was conducted using the \texttt{cluster} package for R \citep{Maechler2013} using all \Sexpr{nrow(adult)} observations.
 
@@
 % training testing split
-The total dataset of \Sexpr{nrow(turtle.adult)} observations was split into training and testing datasets.
+The total dataset of \Sexpr{nrow(adult)} observations was split into training and testing datasets.
 The training dataset represented 75\% of the total dataset, split proportionally by class, and was used for model fitting.
@@
 \ref{fig:pca}).
-The first three PCs, which represent \Sexpr{sum(turtle.fit$percent[1:3])}\% of the total variation, are a cloud of points with no structure.
+The first three PCs, which represent \Sexpr{sum(fit$percent[1:3])}\% of the total variation, are a cloud of points with no structure.
 Additionally, individual landmark variation is mostly circular around each landmark with some more elliptical variation observed along some midline landmarks and the most lateral landmark (Fig.
@@
 % some sex information. Analyze it.
-Sex information was only avaliable for \Sexpr{length(tsex.s)} of the \Sexpr{nrow(turtle.adult)} turtles.
+Sex information was only avaliable for \Sexpr{length(tsex.s)} of the \Sexpr{nrow(adult)} turtles.
 A \(\chi^{2}\) test of the relationship between sex observation and cluster assignment from PAM with two clusters showed that there was no significant relationship between cluster assignment and sex observation (\(\chi^{2}\): \Sexpr{csex.chi$statistic}, df: \Sexpr{csex.chi$parameter}, \textit{p}-value: \Sexpr{csex.chi$p.value}, Table \ref{tab:chi}).
@@
 \ref{fig:roc}).
-The first \Sexpr{length(tm.a.analysis$best$spinks$coefnames) - 1} PCs describe \Sexpr{sum(turtle.fit$percent[1:9])}\% of total variation in plastral shape, while the first \Sexpr{length(tm.a.analysis$best$spinks$coefnames)} PCs describe \Sexpr{sum(turtle.fit$percent[1:10])}\% of the variation.
+The first \Sexpr{length(tm.a.analysis$best$spinks$coefnames) - 1} PCs describe \Sexpr{sum(fit$percent[1:9])}\% of total variation in plastral shape, while the first \Sexpr{length(tm.a.analysis$best$spinks$coefnames)} PCs describe \Sexpr{sum(fit$percent[1:10])}\% of the variation.
 
@@
 \ref{fig:imp_var}).
-This PC represents \Sexpr{turtle.fit$percent[3]}\% of total variation.
+This PC represents \Sexpr{fit$percent[3]}\% of total variation.
 The second and second most important PC is described above and principally described variation in landmarks along the lateral and anterior margin of the plastron.
-This PC represents \Sexpr{turtle.fit$percent[2]}\% of total variation.
+This PC represents \Sexpr{fit$percent[2]}\% of total variation.
 The major variations along these axes correspond well to the differences between the mean shape of each class (Fig.
@@
 Depicted here are the for three PCs (lower triangle) and the mean shape with observed variance around each point (upper right).
-The first three PCs account for total \Sexpr{paste0(sum(turtle.fit$percent[1:3]))}\% of the variance in plastral shape.}
+The first three PCs account for total \Sexpr{paste0(sum(fit$percent[1:3]))}\% of the variance in plastral shape.}
   \label{fig:pca}
@@
 The middle column, for both rows, is the mean plastral shape for all observations.
-The first PC represents \Sexpr{turtle.fit$percent[1]}\% of the total variation in plastral shape while PC represents \Sexpr{turtle.fit$percent[2]}\% of the variance.}
+The first PC represents \Sexpr{fit$percent[1]}\% of the total variation in plastral shape while PC represents \Sexpr{fit$percent[2]}\% of the variance.}
   \label{fig:pc_var}
@@
   \centering
-<<results = 'asis'>>=
-print.xtable(xsex.tab,
-             floating = FALSE,
-             hline.after = c(0, nrow(csex.tab) - 1))
-@
+  \include{xsex_tab}
   \caption{Tabular comparison between sex observation and cluster assignement from PAM with two clusters.
@@
   \centering
-<<results = 'asis'>>=
-print.xtable(tmx.1, 
-             include.rownames = FALSE,
-             floating = FALSE,
-             size = '\\small',
-             sanitize.colnames.function = identity)
-@
+  \include{tmx_1}
   \caption{Model selection table for the multinomial logistic regression models of the first morphologically based classification hypothesis.
@@
   \centering
-<<results = 'asis'>>=
-print.xtable(tmx.2, 
-             include.rownames = FALSE,
-             floating = FALSE,
-             size = '\\small',
-             sanitize.colnames.function = identity)
-@
+  \include{tmx_2}
   \caption{Model selection table for the multinomial logistic regression models of the first morphologically based classification hypothesis.
@@
   \centering
-<<results = 'asis'>>=
-print.xtable(tmx.3, 
-             include.rownames = FALSE,
-             floating = FALSE,
-             size = '\\small',
-             sanitize.colnames.function = identity)
-@
+  \include{tmx_3}
   \caption{Model selection table for the multinomial logistic regression models of the first morphologically based classification hypothesis.
@@
   \centering
-<<results = 'asis'>>=
-print.xtable(tmx.4, 
-             include.rownames = FALSE,
-             floating = FALSE,
-             size = '\\small',
-             sanitize.colnames.function = identity)
-@
+  \include{tmx_4}
   \caption{Model selection table for the multinomial logistic regression models of the first morphologically based classification hypothesis.

### 2014-10-28 R/make_plots.r
using rf var for lda

@@ -188,6 +188,7 @@ gen.name <- function(var, value) {
     value[value == 'mn'] <- 'multinomial logistic regression'
     value[value == 'rf'] <- 'random forest'
     value[value == 'lda'] <- 'linear discriminate analysis'
+    value[value == 'lrf'] <- 'selected LDA'
   }
   return(value)
 }
@@ -198,7 +199,9 @@ uu <- Reduce(cbind, Map(function(x) x$t, mm))
 colnames(uu) <- groups
 vv <- Reduce(cbind, Map(function(x) x$t, ll))
 colnames(vv) <- groups
-zz <- list(rf = oo, mn = uu, lda = vv)
+lr <- Reduce(cbind, Map(function(x) x$t, lrf))
+colnames(lr) <- groups
+zz <- list(rf = oo, mn = uu, lda = vv, lrf = lr)
 dd <- melt(zz)
 gdist <- ggplot(dd, aes(x = value, fill = Var2))
 gdist <- gdist + geom_histogram(alpha = 0.3,
@@ -222,12 +225,16 @@ ggsave(file = '../doc/figure/gen_res.png', plot = gdist,
 names(tl.a.analysis$class) <- groups
 gen.maps <- lapply(groups, function(x) {
                    pred.map(map = ggmap(goog.map), 
-                            xl = c(longmi, longma+1), yl = c(latmi-2, latma),
-                            test = adult.test, name = x,
+                            xl = c(longmi, longma+1),
+                            yl = c(latmi-2, latma),
+                            test = adult.test,
+                            name = x,
                             mods = list(tm.a.analysis, 
                                         trf.a.analysis, 
-                                        tl.a.analysis),
-                            types = c('multi', 'rf', 'lda'), data = adult)})
+                                        tl.a.analysis,
+                                        gr),
+                            types = c('multi', 'rf', 'lda', 'lrf'),
+                            data = adult)})
 names(gen.maps) <- groups
 gen.maps <- lapply(gen.maps, function(x) {
                    x + scale_colour_manual(name = '', values = cbp) + 
@@ -300,47 +307,47 @@ for(ii in seq(length(gsh))) {
 
[…]

### 2014-10-28 R/make_tables.r

@@ -5,79 +5,37 @@ require(reshape2)
 source('../R/plotting_functions.r')
 
 load('../data/gen.RData') 
-load('../data/shape.RData')
+#load('../data/shape.RData')
 
 # 2 x 2 table of sex - cluster assignment
-csex.tab <- rbind(csex.tab, tot = colSums(csex.tab))
-csex.tab <- cbind(csex.tab, tot = rowSums(csex.tab))
-xsex.tab <- xtable(csex.tab)
-align(xsex.tab) <- 'r|cc|c'
-digits(xsex.tab) <- c(0, 0, 0, 0)
-print.xtable(xsex.tab,
-             file = '../doc/xsex_tab.tex',
-             floating = FALSE,
-             hline.after = c(0, nrow(csex.tab) - 1))
-
-
-# multinomial logistic regression model selection tables
-cln <- colnames(tm.a.analysis$sel$spinks)
-#cln[which(cln == 'delta')] <- "$\\delta AICc"
-#cln[which(cln == 'weight')] <- "AICc weights"
-rms <- which(cln == 'family')
-tmcl <- lapply(tm.a.analysis$sel, function(x) {
-               x <- x[, -rms]
-               x})
-cln <- cln[-rms]
-tmtabs <- lapply(tmcl, function(x) {
-                 colnames(x) <- cln
-                 x})
-tmx <- lapply(tmtabs, xtable)
-tmx.1 <- tmx[[1]]
-tmx.2 <- tmx[[2]]
-tmx.3 <- tmx[[3]]
-tmx.4 <- tmx[[4]]
-print.xtable(tmx.1, 
-             file = '../doc/tmx_1.tex',
-             include.rownames = FALSE,
-             floating = FALSE,
-             size = '\\small',
-             sanitize.colnames.function = identity)
-print.xtable(tmx.2, 
-             file = '../doc/tmx_2.tex',
-             include.rownames = FALSE,
-             floating = FALSE,
-             size = '\\small',
-             sanitize.colnames.function = identity)
-print.xtable(tmx.3, 
-             file = '../doc/tmx_3.tex',
-             include.rownames = FALSE,
[…]

### 2014-11-03 doc/draft_turtle.tex
started newest draft

--- 2014-03-07 doc/turtle_paper.Rnw
@@
-% how cryptic is cryptic diversity?
-% paper describing the motivation, methods and results of the turtle
-% subspecies identification project.
-% journal submission order:
-%   Systematic Biology (no limit)
-%   American Naturalist (21 pages)
-%   Journal of Evolutionary Biology
+\documentclass[12pt,letterpaper]{article}
 
-\documentclass[12pt,letterpaper]{article}
 \usepackage{amsmath, amsthm}
-\usepackage{graphicx, morefloats, microtype, hyperref, authblk}
-\usepackage{rotating, longtable, caption, subcaption, multirow}
-\usepackage[sort&compress]{natbib}
-\usepackage{fullpage}
-
-% for american naturalist
+\usepackage{graphicx,hyperref}
+\usepackage{microtype, parskip}
+\usepackage[numbers,sort&compress]{natbib}
+\usepackage{lineno}
+\usepackage[font=small]{caption}
+\usepackage{subcaption, multirow, morefloats}
+\usepackage{subcaption, wrapfig}
+\usepackage{titlesec}
+\usepackage[nottoc,numbib]{tocbibind}
+\usepackage{authblk, attrib, fullpage}
 \usepackage{lineno}
@@
 \frenchspacing
-\linespread{1.66}
-\setlength{\parindent}{0.5in}
 
-\setcounter{secnumdepth}{0}
-
-%\pagestyle{empty}
+\captionsetup[subfigure]{position = top, labelfont = bf, textfont = normalfont, singlelinecheck = off, justification = raggedright}
 
@@
 
-% bring in various code necessities
-% all the figures are made externally, so this would just be for numerics
-<<message = FALSE, echo = FALSE, warning = FALSE, results = 'hide'>>=
-opts_chunk$set(warning = FALSE, echo = FALSE, message = FALSE,
-               highlight = TRUE, background = 'white')
-
-load('../data/gen.RData')
-load('../data/shape.RData')
-
-options(digits = 2)
-@
-
 \title{How cryptic is cryptic diversity? Machine learning approaches to classifying morphological variation in \textit{Emys marmorata} (Testudinoidea, Emydidae).}
@@
 \begin{abstract}
-  Cryptic diversity is the phenomenon where some taxa are believed to be identifiable only based on molecular data.
-This is concerning because the majority of extant taxa and virtually all extinct taxa are delimited entirely via morphology.
-Here we address questions about whether it is possible to determine, based on morphology, if one classification hypotheses can be considered better than others in order to determine if possible cryptic variation is actually cryptic or just a case of extremely fine scale morphological variation.
-%
-  Using a combination of unsupervised and supervised machine learning methods we demonstrate a suite of approaches for better understanding differences in morphology between classes, the odds of classifying one class relative to another, and what aspects of morphology best describe the differences between classes.
-  %
-  These approaches are applied to the classification of the emydid turtle, \textit{Emys marmorata}.
-This species has conflicting hypotheses of the number of meaningful subclades based on either morphological or molecular information.
-We compared multiple explicit classification hypotheses by characterizing variation in plastral shape and how it may be identifiably different between classes.
-By splitting a large dataset of specimens into both training and testing datasets, we were also able to determine which of the classification hypotheses best corresponded to the observed plastral variation in general.
-  %
-  The results from our analysis shows that the best classification of plastral variation in \textit{Emys marmorata} is in accordance with the molecularly based hypothesis.
-This demonstrates that, by using alternative methods for characterizing variability, it is possible to estimate the classification scheme which most agrees with observed variation.
-Additionally, we demonstrate how it is possible that not all examples of cryptic variation are truly cryptic and may just be a product of sample size or methodology because of the extremely fine scale variation between the different classes.
-
-\noindent (Keywords: Testudines, Emydidae, morphology, geometric morphometrics, random forests)
 \end{abstract}
 
-%\section{Introduction}
-
-% cryptic diversity
-%   most species are still deliminated solely based on morphology
-Cryptic diversity is the phenomenon that not all taxa can be recognized from morphology and can only be delimited using molecular information \citep{Bickford2007,SchlickSteiner2007,Stuart2006,Pfenninger2007,Funk2012,Clare2011}.
-Concerningly, most extant taxa, and nearly all extinct taxa, are delimited based solely via morphology.
-This phenomenon is of great concern when studying variation and diversity dynamics over long periods of time, where apparent morphological stasis may not actually reflect true diversity \citep{Hunt2008,Eldredge1972,Gould1977a,VanBocxlaer2013}.
-In the case of endangered or conserved taxa, morphometric approaches for classifying and identifying taxa and populations of importance would greatly improve the ability to maintain these high risk groups.
-Additionally, this could lead to better classification of extinct taxa.
-
-Much work has been devoted to species delimitation via sequence difference \citep{Fujita2012,Yang2010b} while comparatively little has been devoted to introducing new methodology for case of purely morphological data \citep{Mitteroecker2011,Zelditch2004}.
-The majority of this effort has focused on identifying differences between already identified taxa \citep{Polly2007a,Demandt2009,Gaubert2005b,Gunduz2007,Polly2003,Zelditch2004} and automated taxon identification \citep{MacLeod2007}.
-% all the work done on species delimitation strictly from molecular data
-
-Here, we address the question of how can alternative approaches and methodology improve morphology based classification.
-From this approach, we ask if it is possible to determine which amongst a set of classification hypotheses is best in order to determine if a case of cryptic diversity is truly cryptic or just a case of extremely fine scaled morphological variation.
-
-% past work on automatic taxon identification and older approaches to classifying taxa
-% why use machine learning methods
-\subsection{Background and system}
-Machine learning is in many respects just an extension of known statistical methodology \citep{Hastie2009} where the emphasis is placed on inferring rules and properties of data in order to explain the underlying structure.
-The basic statistical mechanics are supplemented by randomization, sorting, and partitioning algorithms and along with the maximization or minimization of summary statistics in order to best estimate a general model for all data, both sampled and unsampled \citep{Hastie2009}.
-Machine learning approaches have found use in medical research, epidemiology, economics and automated image identification such as handwritten zip codes \citep{Hastie2009}.
-Two major classes of machine learning methods are unsupervised and supervised learning.
-Unsupervised learning methods are used with unlabeled data where the underlying structure is estimated and are analogous to clustering and density estimation methods \citep{Kaufman1990}.
-Supervised learning methods are used with labeled data where the final output of data is known and the rules for going from input to output are inferred.
-These are analogous classification and regression models \citep{Breiman1984}.
-The application of the alternative approaches used in this study illustrates only a sampling of the various previously derived methods for clustering observations and fitting classification models.
-
-Differences in morphological variation between different classes has previously been analyzed using methods like linear discriminate analysis and canonical variates analysis \citep{Zelditch2004,Mitteroecker2011,Polly2007a,Polly2003,Gunduz2007,Gaubert2005b,Demandt2009}.
-These methods are comparatively straight forward ways of understanding the differences in morphology between classes.
-Also, they are very visual methods which aides with the interpretation and presentation of information.
-Previous studies, however, normally do not compare which amongst a set of candidate classification hypotheses is better.
-For example, studies such as those of \citet{Caumul2005a} and \citet{Polly2007a} focused on comparing different aspects of morphology and their fidelity to a classification scheme, instead of comparing the fidelity of one aspect of morphology to multiple classification schemes.
-Of note, however, is the work of \citet{Cardini2009a}, which compared morphological variation in marmots at both population, regional, and species levels to determine fidelity between shape each of these different hierarchical levels.
-Importantly, however, is that the classification models have not been generalized to testing data and training data accuracy is used almost exclusively as the metric off classification strength.
-
-Here, we used multiple machine learning methods, both unsupervised and supervised, in order to compare different classification hypotheses.
-These methods provide different and unique advantages for understanding how to classify taxa, with what accuracy, and what these classifications are based on.
-While machine learning methods such as neural networks have been applied to studying shape variation \citep{MacLeod2007}, they have been primarily applied in the context of automated taxon identification and not in terms of group classification and strength of classification.
-Additionally, we investigate variation in continuous traits and not discrete differences between each class, instead focusing on differences in the multivariate quantification of shape.
-Also, instead of pure classification accuracy, here we used a statistic of classification strength that reflects the rate at which taxa are both accurately and inaccurately classified (see Methods).
-
-% e. marmorata
-%   natural history
-%     conservation importance (turtles in general, e. marmorata in specific)
-In this study, we investiage the subspecific classification of the western pond turtle, \textit{Emys marmorata}.
-\textit{E.
-marmorata} is distributed from northern Washington State, USA to Baja California, Mexico.
-%   morphological hypothesis of subspecies
-%   molecular hypothesis
-Traditionally, \textit{E.
-marmorata} was classified into three groups: the northern \textit{E.
-marmorata marmorata}, the southern \textit{E.
-marmorata palida}, and a central Californian intergrade zone \citep{Seeliger1945}.
-\textit{E.
-marmorata marmorata} is differentiated from \textit{E.
-marmorata palida} by the presence of a pair of triangular inguinal plates and darker neck markings.
-It should be noted that the triangular inguinal plates can sometimes be present in \textit{E.
-marmorata palida} though they are considerably smaller.
-
-Previous work on morphological variation in \textit{E.
-marmorata} has focused, primarily, on differentiation between different populations within a subset of the total species range \citep{Lubcke2007,Germano2009,Germano2008,Bury2010} with comparatively little done over the entire species range \citep{Holland1992}.
-These studies have focused on how local biotic and abiotic factors may contribute to differences in carapace length \citep{Lubcke2007,Germano2009,Germano2008} and found that size can vary greatly between different populations.
-
-Additionally, there has been found a great deal of evidence for sized-based sexual dimorphism in \textit{E.
-marmorata} \citep{Lubcke2007,Germano2009,Holland1992} with males being on average larger than females based on total carapace length and other linear measurements.
-However, the quality of size as a classifier of sex can vary greatly between populations \citep{Holland1992}, which makes sense in light of the amount of between population size difference \citep{Lubcke2007,Germano2009}.
-However, the effect of sexual dimorphism on shape, \textit{sensu} \citet{Kendall1977a}, was not assessed \citep{Holland1992,Lubcke2007,Germano2008}.
-
-Of particular note is the work of \citet{Holland1992} which compared morphological differences between and among many populations of \textit{E.
-marmorata} across the species range.
-\citet{Holland1992} studied the relative effect of distance versus barriers had in terms of fostering morphological differentiation in \textit{E.
-marmorata}.
-Analyses were performed to determine how different, morphologically, different populations in three different regions of the species range.
-Measurements were made from all different aspects of carapace morphology and not just total carapace length.
-
-\citet{Holland1992} concluded that distance was a poor indicator of morphological differentiation as opposed to barriers, such as different drainage basins, are probably more important barriers to reproduction.
-This conclusion was later echoed by \citet{Spinks2005} via molecular phylogenetic analysis.
-Additionally, \citet{Holland1992} found that with increasing amount of barriers and distance, morphological differentiation was observable though the underlying variation required many variables obtain indicating the very fine degree of morphological differentiation between putatively distinct populations.
-\citet{Holland1992} concluded that \textit{E.
-marmorata} is best classified as three distinct species as opposed to subspecies: a northern species, southern species, and Columbia basin species.
-This classification is similar to \citet{Seeliger1945}, except elevated to the species as opposed to subspecific level.
-
-More recently, \textit{E.
-marmorata} was divided into four clades based on mitochondrial DNA: a northern clade, a southern clade, and eastern and western central Californian clades \citep{Spinks2005,Spinks2010}.
-While nuclear DNA supports two major clades, one northern and one southern, \citet{Spinks2010} argue that the four clade classification is of greater conservation utility even though the variation between these groups is considered cryptic.
-While the mitochondrially based classification is considered robust, there is no known morphological differentiation between these clades.
-
-% central hypothesis/question of study
-% statement of goals and approach
-In this study, we attempt to estimate the best classification scheme of \textit{E.
-marmorata} based on variation in plastral shape in order to determine if the molecular based hypothesis of \citet{Spinks2005} and \citet{Spinks2010} is actually a case of cryptic diversity or not.
-Because of unclear geographic boundaries between subgroups of \textit{E.
-marmorata}, we compare two hypotheses of morphologically based classification and two hypotheses of molecularly based classification.
-We hypothesize that if morphological variation corresponds to class assignment, then it should be possible to determine the best classification hypothesis of \textit{E.
-marmorata} from amongst multiple candidate hypotheses.
-However, if morphological variation variation does not correspond to any classification hypothesis, then supervised learning model generalization performance will be poor and reflect how variation may not follow along with any of the candidate classification hypotheses.
+\section{Introduction}
 
 \section{Materials and Methods}
-\subsection{Specimens}
-We collected landmark-based morphometric data from \Sexpr{nrow(adult)} adult \textit{E.
+\subsection{Specimens and sampling}
+We collected landmark-based morphometric data from SAMPLE adult \textit{E.
 marmorata} museum specimens.
 These specimens include both newly sampled individuals and those sampled in previous studies of plastral shape variation \citep{Angielczyk2007,Angielczyk2011,Angielczyk2013a}.
+
 Specimen classification was based on known specimen geographic information which was recorded from museum collection information.
@@
 \subsection{Geometric morphometrics}
-% landmarks
 Following previous work on plastral variation \citep{Angielczyk2007,Angielczyk2011,Angielczyk2013a}, 19 landmarks were digitized using TpsDig 2.04 \citep{Rohlf2005}.
@@
 Analysis was conducted on the resulting ``half'' plastra.
-% GP superimposition and PCA
 Plastral landmark configurations were superimposed using generalized Procrustes analysis \citep{Dryden1998a} after which, the principal components (PC) of shape were calculated.
@@
 \subsubsection{Unsupervised learning}
-% clustering
-%   Riemannian shape distance
 In order to preserve the relationship between all landmark configurations in shape space, the dissimilarity between observations was measured using Kendall's Riemanninan shape distance or \(\rho\) \citep{Kendall1984a,Dryden1998a}.
@@
 
-%   PAM
 The \(\rho\) dissimilarity matrix was divisively clustered using partitioning around mediods clustering (PAM), a method similar to \textit{k}-means clustering except that instead of minimizing the sum of squared Euclidean distances between observations and centroids, the sum of squared dissimilarities between observations and mediods is minimized \citep{Kaufman1990}.
-%   gap statistic
-Because the optimal number of clusters of shape configurations in the study was unknown, being possibly three, four, or some other value, clustering solutions were estimated with the number of clusters varied between one and \Sexpr{nrow(tadult.gap$Tab)}.
+Because the optimal number of clusters of shape configurations in the study was unknown, being possibly three, four, or some other value, clustering solutions were estimated with the number of clusters varied between one and NUMBER.
 Clustering solutions were compared using the gap statistic, which is a measure of goodness of clustering \citep{Tibshirani2001a}.
 The gap statistic is defined
+
 \[Gap_{n}(k) = E^{*}_{n}[\log(W_{k})] - \log(W_{k})\]
+
 where \(W_{k}\) is
+
 \[W_{k} = \sum^{k}_{r = 1}{\frac{1}{2n_{r}} (\sum_{i,i' \in C_{r}} d_{ii'})}\].
+
 \(d_{ii'}\) is the dispersion of the clustering solution or the sum of the pairwise dissimilarities between observations in each cluster and their respective mediods (\(C\)) for all clusters \(r\).
 This value is averaged and compared to the expected dispersion (\(E^{*}_{n}\)) of a sample \(n\) from a reference distribution.
-In this case, the reference distribution was estimated from \Sexpr{tadult.gap$B} resamples of the dataset while maintaining the original dispersion of the data.
-%  R implementation
-This analysis was conducted using the \texttt{cluster} package for R \citep{Maechler2013} using all \Sexpr{nrow(adult)} observations.
+In this case, the reference distribution was estimated from NUMBER resamples.
+
+This analysis was conducted using the \texttt{cluster} package for R \citep{Maechler2013}.
 
 \subsubsection{Supervised learning}
-% training testing split
-The total dataset of \Sexpr{nrow(adult)} observations was split into training and testing datasets.
-The training dataset represented 75\% of the total dataset, split proportionally by class, and was used for model fitting.
-The testing dataset represented the remaining 25\% of the total dataset and  was used after model fitting to estimate the effectiveness of each classification hypothesis and generalizability of the supervised learning models (i.e.
-performance in the wild).
-This split was chosen to allow for a large enough sample size for model fitting while also providing a large enough testing dataset to determine any systematic misclassifcations.
+Three different supervised learning, or classification, approaches were used: linear discriminate analysis, multinomial logistic regression, and random forests.
+Linear discriminiate analysis, also known as canonical variate analysis, is commonly used in studies of geometric morphometric data \citep{Zelditch2004,Mitteroecker2011}.
+The other two methods, however, are not.
+Each of these three methods has a different interpretation and reveal very different aspects of the data.
+In all cases, the optimal number of PCs used as predictors was chosen via max within-sample AUC value, explained below.
 
-% model types
-Three different supervised learning methods were used to model the relationship between plastral shape and class: linear discriminate analysis, multinomial logistic regression and random forest.
-These methods were chosen because of various properties of these methods which allow for useful interpretations about the quality and structure of the classification.
-
-% linear discriminate analysis
-Linear discriminate analysis (LDA) is a frequently applied method for characterizing the primary differences in morphology between different classes \citep{Zelditch2004,Mitteroecker2011}.
-This method attempts to find a linear combination of predictors to best model two or more classes.
+Linear discriminate analysis (LDA) attempts to find a linear combination of predictors to best model two or more classes.
 LDA is very similar to PCA except that instead of finding the linear combination of features that maximize the amount of explained variance in the data, LDA maximizes the differences between classes.
 The results of this analysis produces a transformation matrix by which the original features can be transformed to reflect the best discrimination between the classes.
-Like other supervised learning methods, LDA can also be used for predictive analysis on testing data.
-LDA was done using the \texttt{MASS} package for R \citep{Venables2002a}.
+In this analysis, LDA was applied on the eigenscores from a subset of the total number of PCs, ranging from two to NUMBER in increasing order of complexity.
+In total, this produced nine different LDA scaling matrices.
 
-%   multinomial logistic regression
 Multinomial logistic regression is an extension of logistic regression, where instead of a binary response there are three or more response classes \citep{Venables2002a}.
-Effectively, this type of model can be viewed as multiple, simultaneous logistic regression models for each class and the final classification of the observation being the most probable of all the constituent model results.
-Similar to the odds ratios calculated from the coefficients of a logistic regression, the relative risk of a classification with reference to a baseline class can be determined from the coefficients of the model.
-Multinomial logistic regression models were fit using the \texttt{nnet} package for R \citep{Venables2002a}
+Similar to the odds ratios calculated from the coefficients of a logistic regression, the relative risk of a classification can be determined from the coefficients of the model.
+Similar to LDA, the optimal number of PCs as predictors was determined by comparing within-sample AUC values across multiple models.
 
-%   random forest
 Random forest models are an extension of classification and regression trees (CART) \citep{Breiman1984,Breiman2001}.
-Because this study relies on classification models, CARTs are explained with reference to classification but the approach is equally valid for regression.
-The goal of CARTs are to use a series of different features to estimate the final class.
+The goal of CARTs are to use a series of different features to estimate the class of an observation.
 In top-down induction of decision trees for each member of a given set of predictor variables, attribute value test are used to estimate the differences between classes.
-This process is then repeated on each subset, called recursive partitioning.
+This process, called recursive partitioning, is then repeated on each subset.
 The recursion continues until the resulting observations all share the same class or no more meaningful partitions are possible.
@@
 This process is then repeated many times and the parameters of the final model was chosen as the mode of estimates from the distribution of CARTs \citep{Breiman2001}.
-In addition to fitting a classification model, this procedure allows for the features to be ranked in order of importance.
-In the context of this study, this means that the PCs most important for describing the difference between classes can be estimated, and thus illustrate the most important variation amongst classes as opposed to just the greatest amount of variation in the entire dataset.
-This is a generally important property that is useful for many other studies which want to describe and model the differences between classes and the relative importance different features.
-Random forest models were fit using the \texttt{randomForest} package for R \citep{Liaw2002}.
-%   R statement
+In addition to classifying the observations, this procedure allows for the features to be ranked in order of importance.
+This is a generally important property that is useful for many other studies which want to describe and model the differences between classes and the relative importance of different predictors.
 
-% model training
-%   tuning parameters
-%     grid search
-%       what are the tuning parameters for each model?
-%     best AUC
-The supervised learning models used here, except LDA, have tuning parameters which help to increase the genearlizability of the model and prevent them from being overfit.
-For the supervised learning models fit in this study, tuning parameters were estimated via 10 rounds of 10-fold cross-validation (CV) across a grid search of all tuning parameter combinations.
-Optimal tuning parameter values were selected based on area under the receiver operating characteristic (ROC) curve.
-The area under the multiclass ROC curves was estimated using the all-against one strategy derived by \citet{Hand2001}.
-This tuning process was implemented following the default grid search implemented in the \texttt{caret} package for R \citep{Kuhn2013}.
-
-% what is ROC? what is AUC?
-
-ROC is a confusion matrix (Table \ref{tab:conf}) statistic that is a descriptor the relationship between the false positive rate (\(FPR\), Eq.
-\ref{eq:fpr}) of a classification model and the true positive rate (\(TPR\), Eq.
-\ref{eq:tpr}) of a classification model \citep{Hastie2009}.
-The area under the ROC curve (AUC) is a summary statistic of the quality of the classification and varies between 0.5 and 1, with values of 0.5 indicating a model that classifies no better than random and a value of 1 indicating perfect classification \citep{Hastie2009}.
-AUC can be used as a model selection criterion for classification models and is especially useful in cases where some if not all of the models in question were not fit via maximum likelihood where a criterion such as AICc (see below) or similar can be used \citep{Hastie2009}.
-It is important to note that, unlike AICc, AUC is not calculated with reference to the complexity of the model.
-
-\begin{equation}
-  FPR = \frac{FP}{FP + TN}
-  \label{eq:fpr}
-\end{equation}
-\begin{equation}
-  TPR = \frac{TP}{TP + FN}
-  \label{eq:tpr}
-\end{equation}
-
-%   model selection
-%   LDA
-LDA was applied on the eigenscores from a subset of the total number of PCs, ranging from two to \Sexpr{floor(max.ad)} in increasing order of complexity.
-In total, this produced nine different LDA scaling matrices.
-From this set, the best number of PCs used to estimate the LDA scaling matrix were chosen.
-As LDA is not ``fit'' via maximum liklihood, the final combination of number of PCs and LDA scaling matrix chosen was that with the greatest AUC value from the training set.
-
-%     AICc
-For the multinomial logistic regression models, \Sexpr{floor(max.ad)} different models were fit each having sequentially more PCs as predictors in order to have models representing different levels of overall amount of shape variation and to estimate how much was necessary and sufficient to best estimate class.
-The maximum number of PCs allowed as predictors was \Sexpr{floor(max.ad)} because of both the large number of parameters estimated per model and the necessary sample size needed to estimate that many parameters accurately.
-The final model was that with the lowest AICc \citep{Akaike1974,Hurvich1989,Burnham2002a}.
-AICc is a model selection criterion where the model with lowest AICc has the fairest variance--bias tradeoff \citep{Burnham2002a}.
-Model selection was performed in this manner because the optimal number of PCs to use as predictors was not know \textit{a priori}, and while including all of the PCs of shape would mean that all shape variability would be used to estimate class, this may cause the model to be overfit and not provide an accurate estimate of unsampled plastral variation.
-In addition to the AICc of each model the \(\Delta\)AICc and Akaike weights are also reported.
-\(\Delta\)AICc values are the different in AICc between the AICc best model and that model while Akaike weights are a transformation of the AICc of a model with relation to all other models being compared and measures the relative amount of information explained by that model compared to all other models \citep{Burnham2002a}.
-
-%     recursive feature selection
-Random forest models are not fit using maximum likelihood so AICc based model selection was not possible.
-Instead, a recursive feature selection algorithm was used to choose the optimal number of PCs to include based on the AUC of the model.
+In this analysis, random forest model parameters were estimated from 1000 subtrees.
+The best set of predictors necessary for each classification scheme was estimated using a recursive feature selection algorithm was used to choose the optimal number of PCs to include based on the AUC of the model.
 Following the backwards selection algorithm implemented in \texttt{caret} \citep{Kuhn2013}, the maximum number of features were included in the initial model, their importance ranked, and the AUC of the model calculated.
@@
 This was repeated until only one feature, remained.
-Similar to the multinomial logistic regression models described above, the maximum number of PCs that could have been included in the model was \Sexpr{floor(max.ad)}.
-After each PC was removed , 10-fold CV was used to estimate the optimal values of the tuning parameters as well as quantify the uncertainty of each model.
-Random forest model parameters were estimated from 1000 subtrees.
-Because PCs were kept in order of importance and not in relation to the amount of variance each PC described, this means that the exact PCs included in each model do not correspond to the PCs in each of the 10 multinomial logistic regressions models.
+Because PCs were kept in order of importance and not in relation to the amount of variance each PC described, these means that the PCs are not included in assending eigenvalue.
 
-% generalization
-%   bootstrap resample of AUC
-The final selected models were then used to estimate the class assignments of the training dataset.
-Model generality for both methods for all four classification schemes was measured using the AUC of the assignments.
-A distribution of AUC values was estimated for each classification scheme via \Sexpr{rr$sh1$R} nonparametric bootstrap resamples of the training dataset.
-The difference in distributions was assessed using pairwise Mann-Whitney U tests.
+In classification studies, like this one, a common metric of performance is area under the reciver operating characteristic curve (AUC).
+AUC is an estimate of the relationship between the false positive and true positive rates, as opposed to just the true positive rate (accuracy).
+This relationship is especially useful in cases where misclassification needs to be minimized just as much as accurate classification, as in this study.
+AUC ranges between 0.5 and 1, with 0.5 indicating classification no better than random and 1 indicating perfect classification \citep{Hastie2009}.
+
+The standard AUC calculation is defined for of binary classification, however in this application there are multiple categories.
+The alternative calculation used here follows an all-against-one strategy where the individual AUC values for each class versus all others are averaged to produce a multiclass AUC \citep{Hand2001}.
+
+The ultimate measure of model fit is accurately predicting the values of unobserved samples \citep{Hastie2009,Kuhn2013}.
+Within-sample performance is inherently biased upwards, so model evaluation requires overcoming this bias.
+With very large sample sizes, as in this study, part of the sample can be used as the ``training set'' and the remainder acts as the ``testing set.'' The former is used for fitting the model while the later is used for measuring model performance.
+This is called model generalization.
+In this analysis, 75\% of samples were used as the training set while the remaining 25\% were used as the testing set.
+
+In order to estimate confidence intervals on the out-of-sample AUC values, a nonparametric bootstrap was performed where the true and estimated classifications were resampled with replacement.
+This was done 1000 times.
 
 \section{Results}
-\subsection{Geometric morphometrics}
-The results of the PCA of the total dataset of \textit{E.
-marmorata} pastral landmarks configurations demonstrates no clear or obvious groupings (Fig.
-\ref{fig:pca}).
-The first three PCs, which represent \Sexpr{sum(fit$percent[1:3])}\% of the total variation, are a cloud of points with no structure.
-Additionally, individual landmark variation is mostly circular around each landmark with some more elliptical variation observed along some midline landmarks and the most lateral landmark (Fig.
-\ref{fig:pca}).
-However, it is important to note that Procustes based superimposition attempts to evenly distribute variance around the mean shape \citep{Zelditch2004} and this observation should be considered cursory at best.
-
-The first two PCs appear to describe principally variation in the lateral margin of the palstra, from a pointed medial edge to a more rounded and blunt edge (Fig.
-\ref{fig:pc_var}).
-Landmark 10 (Fig.
-\ref{fig:plastra}), which appears to be the most variable along these axes (Fig.
-\ref{fig:pca} and \ref{fig:pc_var}), is positioned on the bridge between the plastron and the carapace.
-Over ontogeny, this is an area that deepens dorsoventrally and when the plastron was projected into two dimensions it created the effect of mediolateral movement.
-Lateral landmark variation along the first PC seemed concentrated in the posterior portion of the plastra with additional variance observed in midline landmarks (Fig.
-\ref{fig:pc_var}).
-This variance in midline landmarks was most likely caused by the fact that plastral scutes frequently do not line up perfectly.
-Along PC 2, lateral variation appeared to be concentrated in the anterior portion o the plastra (Fig.
-\ref{fig:pc_var}).
-
-% size correlation with either of the first two PCs?
-When the eigenscores of the first two PCs are compared to the centroid sizes of the observations, a significant correlation is observed between centroid size and the first PC (Spearman's \(\rho\): \Sexpr{scor$PC1$estimate}, \textit{p}-value: ~\Sexpr{scor$PC1$p.value}) but not between centroid size and the second PC (Spearman's \(\rho\): \Sexpr{scor$PC2$estimate}, \textit{p}-value: ~\Sexpr{scor$PC2$p.value}).
-This difference is clearly observable in in plots of the data (Fig.
-\ref{fig:cent}).
-
-\subsection{Machine learning analyses}
-\subsubsection{Unsupervised learning}
-
-% PAM results
-Comparison of gap statistic values for the range of PAM solutions indicates that the optimal number of clusters is one (Fig.
-\ref{fig:gap}).
-The next best clustering solution had only two clusters, however there is no geographic structure to this classification scheme, with members of these clusters being seemingly randomly distributed (Fig.
-\ref{fig:gap_map}).
-Importantly, these clusters do not conform to the northern and southern groups from the nuclear DNA hypothesis \citep{Spinks2010}.
-
-% some sex information. Analyze it.
-Sex information was only avaliable for \Sexpr{length(tsex.s)} of the \Sexpr{nrow(adult)} turtles.
-A \(\chi^{2}\) test of the relationship between sex observation and cluster assignment from PAM with two clusters showed that there was no significant relationship between cluster assignment and sex observation (\(\chi^{2}\): \Sexpr{csex.chi$statistic}, df: \Sexpr{csex.chi$parameter}, \textit{p}-value: \Sexpr{csex.chi$p.value}, Table \ref{tab:chi}).
-This results is interesting because while sexual dimorphism has been observed in linear measures and mass estimates of \textit{E.
-marmorata} \citep{Lubcke2007,Germano2009,Holland1992}, this results demonstrates that this dimorphism may not translate into differences in shape.
-Interestsingly, male emydid turtles are known to have a plastral concavity which may influence landmark position along the midline.
-However, the plastral concavity of \textit{E.
-mamorata} males is considered less pronounced than in other emydid turtles.
-
-The gap statistic values for both three and four clusters are much lower than for one and two and are statistically identical.
-Interestingly, other solutions with a much greater number of clusters have relatively high gap statistic values as well though these are also not significantly different.
-Increasing the number of clusters does appear to improve the gap statistic enough compared to the best clustering solution to merit detailed discussion.
-
-% figure
-%   gap statistic results
-
-\subsubsection{Supervised learning}
-%For all of the classification schemes, the optimal models for both the multinomial logistic regression (Table SUPPLEMENT) and random forest (Fig.
-\ref{fig:roc}) methods include many, if not most, of the possible features.
-<<>>=
-ldasel <- unlist(lapply(tl.a.analysis$best, function(x) length(x$finalModel$xNames)))
-@
-The optimal number of PCs used for LDA, as determined by highest ROC score, for three of the four classification schemes had all \Sexpr{max(ldasel)} possible PCs (Fig.
-\ref{fig:roc}).
-These were both of the morphological based classification hypotheses and the second molecular hypothesis.
-LDA of the PCs of the first molecular hypothesis found that, based on ROC, only the first \Sexpr{min(ldasel)} PCs were necessary to best discriminate between the classes (Fig.
-\ref{fig:roc}).
-The first \Sexpr{length(tm.a.analysis$best$spinks$coefnames) - 1} PCs describe \Sexpr{sum(fit$percent[1:9])}\% of total variation in plastral shape, while the first \Sexpr{length(tm.a.analysis$best$spinks$coefnames)} PCs describe \Sexpr{sum(fit$percent[1:10])}\% of the variation.
-
-<<>>=
-daic <- unlist(lapply(lapply(tm.a.analysis$sel, function(x) x$delta), 
-                      function(x) x[2]))
-@
-% multinomial logistic regression model
-The AICc best multinomial logistic regression model for three of the four classification schemes had the first \Sexpr{length(tm.a.analysis$best$spinks$coefnames) - 1} PCs as features (Tables \ref{tab:mod_sel_1}, \ref{tab:mod_sel_2}, and \ref{tab:mod_sel_3}).
-The second molecularly based classification hypothesis included all \Sexpr{length(tm.a.analysis$best$spinks$coefnames)} possible PCs as predictors (Tables \ref{tab:mod_sel_4}).
-The \(\Delta\)AICc values between the optimal and second best model range from \Sexpr{min(daic)} for the first morphological based classification hypothesis to \Sexpr{max(daic)} for the second molecular based classification hypothesis (Tables \ref{tab:mod_sel_1}, \ref{tab:mod_sel_2}, \ref{tab:mod_sel_3}, and \ref{tab:mod_sel_4}).
-
-While the \(\Delta\)AICc value between the optimal and second best model for the first morphological and first molecular based classification hypothesis was within the range to be considered equally optimal \citep{Burnham2002a}, for this analysis we chose to use only the AICc best model.
-While AICc values can not be compared between models with different responses \citep{Burnham2002a}, we interpret the fact that the \(\Delta\)AICc best model in these cases is the simpler model and that the optimal model for three of the classification schemes having the same number of predictors as reasons to use only the AICc best model for all cases.
-Additionally, by using a single model for each of the classification hypotheses, this limits the number of comparisons between the bootstrap resampled distributions of the AUC values for the testing dataset (see below).
-
-% random forest model
-<<>>=
-oprf <- unlist(lapply(trf.a, function(x) x$optsize))
-@
-
-The selected number of features in the final random forest model for each classification scheme was very simpler to the model selection results for the LDA-based classification and the multinomial logistic regression models, ranging from \Sexpr{min(oprf)} for the second morphological based classification hypothesis and both molecular based classification hypotheses to \Sexpr{max(oprf)} for the first morphological based classification hypothesis (Fig.
-\ref{fig:roc}).
-
-% need to discuss the final models in both cases.
-% figure
-%   ROC model selection (see talk)
-%   generalize densities
-%     facet: multinomial, rf (see talk)
-
-In the case of all models, there is a substantial increase in model performance as measured by AICc for the multinomial logistic models (Tables \ref{tab:mod_sel_1}, \ref{tab:mod_sel_2}, \ref{tab:mod_sel_3}, and \ref{tab:mod_sel_4}) or in AUC for the LDA-based predictions and random forest models and illustrated for the multinomial logistic regression models as the number of features increases (Fig.
-\ref{fig:roc}).
-
-The results from the generalization of the selected supervised learning models, measured by the distributions of the bootstrapped AUC values of the testing dataset, show that a molecular classification hypotheses was the best overall classification scheme (Fig.
-\ref{fig:gen_res}).
-Remarkably, the best classification hypothesis was the second molecular classification hypothesis based on the LDA-based predictions, the multinomial logistic regression and random forest models.
-For both methods, the distribution of bootstrapped AUC for the molecular hypothesis was significantly greater than all of the other classification schemes (Tables \ref{tab:lda_test}, \ref{tab:mm_test} and \ref{tab:rf_test}).
-
-When the classification results of the training set for the best classification scheme based on the generalization results are compared with the references classes, the higher AUC value of the best results from LDA and the best multinomial logistic regression model compared to the best random forest model can be observed as the classifications are much closer to the reference classes (Fig.
-\ref{fig:gen_map}).
-The best random forest model misclassified many of the observations as the northern clade instead of the correct class.
-This pattern of misclassification is observable but not as exaggerated in the LDA-based classifications and those from the multinomial logistic regression model (Fig.
-\ref{fig:gen_map}).
-
-This pattern of misclassification may have been caused by the subtle differences in mean shape between each of the different classes (Fig.
-\ref{fig:mean_shape}).
-The mean shape of the northern clade is the most similar to the mean shape of the entire dataset (Fig.
-\ref{fig:mean_shape1}), which may indicate that specimens that are closer to the mean shape will be systematically misclassified as the northern clade.
-
-% variation along the most important axes
-<<>>=
-rf.w <- gsub('PC', x = trf.a$spinks$optVariables, replacement = '')
-@
-The results of fitting the final random forest model also include the variable importance for best separating the different classes.
-The selected random forest model for the best classification scheme had \Sexpr{trf.a$spinks$optsize} PCs as features.
-The PCs included as features in the final random forest model, in descending order of importance, were PCs \Sexpr{paste(rf.w[1:8])} and \Sexpr{paste(rf.w[9])}.
-Of these \Sexpr{trf.a$spinks$optsize} features, the first three are illustrated here (Fig.
-\ref{fig:imp_pc}) in descending order of importance.
-
-The first two most important features describe different aspects of variation (Fig.
-\ref{fig:imp_var}).
-The third and most important PC describes variation roundedness of the medial portion of the plastron, both the anterior and posterior portions of the plastron.
-Additionally, the relative position of the landmarks along the midline varies greatly along PC3 (Fig.
-\ref{fig:imp_var}).
-This PC represents \Sexpr{fit$percent[3]}\% of total variation.
-The second and second most important PC is described above and principally described variation in landmarks along the lateral and anterior margin of the plastron.
-This PC represents \Sexpr{fit$percent[2]}\% of total variation.
-The major variations along these axes correspond well to the differences between the mean shape of each class (Fig.
-\ref{fig:mean_shape}) where major class differences seem based on the relative ballooning or shrinking of the anterior and posterior portions of the plastron together along with differential ``pinching'' of the midline landmarks.
-It is important to note that the first two most important axes, according to the random forest model, were not correlated with size.
-This can be interpreted to mean that class membership is principally not related to turtle size.
-This interpretation makes sense in light of previous work showing that body size in \textit{E.
-marmorata} can vary widely across the entire species range \citep{Holland1992,Lubcke2007,Germano2009}.
-
-The relative risk values for classification from the multinomial logistic regression model, based on the three most important PCs, demonstrate that individual axes contribute to classification differently and that given multiple features the odds of determining the correct classification increase (Fig.
-\ref{fig:rel_risk}).
-The first most important axis contributes strongly to classifying both the western and southern groups while changes along the second most important axes contribute very little to increasing the odds of classification for all but the eastern group.
-This is observable from the class histograms of PC 3 and 2 (Fig.
-\ref{fig:imp_var}).
-Changes along the first and third most important axes contribute more obviously to increasing the odds of correctly identifying the class of an observation, a result that is observable in both the relative risk (Fig.
-\ref{fig:rel_risk}) and the different class histograms of the PCs (Fig.
-\ref{fig:imp_var}).
-
-The graphical results from the LDA of the training dataset for models of the second molecular classification scheme agree with the subtle distinctions between the different classes (Fig.
-\ref{fig:lda}).
-There is no clear distinction in terms of multivariate space between the four different classes.
-Instead, across all three axes there is substantial overlap as indicated by both the scatter of the points in space and the distribution of observations along each axis.
-
-\section{Discussion}
-% discussion of results
-%   remarkable concordance in the supervised learning methods
-The results of this study support the mitochrondial based classification hypothesis of \textit{E.
-marmorata} \citep{Spinks2005,Spinks2010}.
-This is contrary to the original classification of \textit{E.
-marmorata} \citep{Seeliger1945,Holland1992} and lends credence to the idea that at least some aspect of cryptic diversity is a product of sample size, methodology, or both.
-
-%   what does this mean about cryptic diversity?
-%   unsupervised learning shortcomings
-
-The lack of coherent geographical subclass assignment from PAM clustering (Fig.
-\ref{fig:gap}) as well as the large number of features necessary before no increase in AUC for all models (Fig.
-\ref{fig:roc}) indicates that the morphological variation between classes is extremely fine grained.
-This was also exemplified by the small differences between mean class shapes of the final chosen classification scheme (Fig.
-\ref{fig:imp_var}).
-
-% methodological concerns
-%   compromise in the supervised learning models
-%     fair because as much variation as ``necessary'' is used
-The approaches presented here for supervised learning analysis of the landmark variation represent a compromise between explicitly modeling all shape variation and preventing models from being overfit and ungeneralizable.
-While all aspects of shape may be evolving simultaneously, and not along individual PCs, including all shape variation in each model might increase model complexity beyond a reasonable level for the sample size and possibly the necessary complexity to accurately model the response.
-Additionally, because only individual PCs are used as features in the models, this does not accurately represent shape evolution and how exactly different classes might be evolving in relation to each other.
-However, this compromise is not without its advantages.
-Because both AICc and AUC values improved rapidly with increased model complexity (Fig.
-\ref{fig:roc}), this helped demonstrate how fine scale the actual variation between classes was.
-The variable importance information from the random forest models was extremely useful for understanding what aspects shape variance contributed most to differentiating the classes and in what order as opposed purely in the order of largest variance (Fig.
-\ref{fig:imp_pc} and \ref{fig:imp_var}).
-Importantly, the order of variable importance demonstrated that the two most important axes (PC 2 and PC 3) for determining class membership were not correlated with size, unlike the third most important axis (PC 1), indicating that class membership is not strongly related to size as opposed to being related to actual shape differences.
-Additionally, the relative risk values from the mulitinomial logistic regression models demonstrate that a single PC is probably not sufficient for estimating the class of an observation, but that given a set of PCs this classification would be more accurate (Fig.
-\ref{fig:rel_risk}).
-
-%   unsupervised learning
-Ultimately, it would be useful to not require such explicit classification hypotheses, especially when concerned about possible cryptic variation in extinct taxa.
-The only unsupervised method employed in this study, PAM, is rather simple and not model based.
-A more useful approach would be to employ various model based clustering approaches \citep{Fraley2002,Zhong2003,VanBocxlaer2013}.
-In this manner, a series of candidate models can be compared via model comparison methods, such as AIC or Bayes factors \citep{Fraley2002}, in order to asses the best clustering solution.
-% beat complaints from reviewers
-Here we focused on the results and utility of supervised methods because they are both more powerful and hypothesis driven \citep{Hastie2009}.
-Because there are two alternative classification schemes for \textit{E.
-marmorata}, it was most appropriate to compare these two hypotheses and estimate which one most accurately reflected the variation.
-Future work would be to explore and derive unsupervised methods which corroborate these results.
-
-% closing statements
-In this study we have demonstrated that, using alternative methodology to that which is most frequently applied, it is possible to determine which classification scheme best matches variation in a taxon amongst a set of alternative hypotheses.
-The observed plastral variation of \textit{E.
-marmorata} is most consistent with the mitochondrial based hypothesis of \citet{Spinks2005} and \citet{Spinks2010} and not with the original morphology based hypothesis of \citet{Seeliger1945,Holland1992}.
-We have also demonstrated the utility of various machine learning approaches to understanding the structure of variation in morphometric data.
-Specifically, methods for better understanding misclassification and identifying which is the most important for delimiting different classes.
-These methods represent new applications which may be important for future studies on class-based morphological comparison and variation, both in the context of cryptic diversity and with known classifications.
-
-\section{Acknowledgements}
-PDS would like to thank David Bapst, Michael Foote, Benjamin Frable, and Dallas Krentzel for useful discussion which enhanced the quality of this study.
-For access to emydine specimens, we thank: J.
-Vindum and R.
-Drewes (CAS); A.
-Resetar (FMNH); R.
-Feeney (LACM); C.
-Austin (LSUMNS); S.
-Sweet (MSE); J.
-McGuire and C.
-Conroy (MVZ); A.
-Wynn (NMNH); P.
-Collins (SBMNH); B.
-Hollingsworth (SDMNH); C.
-Bell and R.
-Burroughs (TMM); T.
-LaDuc and R.
-Burroughs (TNHC); P.
-Holroyd (UCMP); R.
-Symonds (UMZC); J.
-Buskirk.
-We are greatful to S.
-Sweet for field assistance and the California Department of Fish and Game for permits.
-Much of the data collection was funded by NSF DBI-0306158 (to KDA).
 
@@
 
-\pagebreak
+\clearpage
 
-% figures
 \begin{figure}[ht]
@@
 
-\begin{figure}[ht]
-  \centering
-  \includegraphics[width = \textwidth]{figure/pca_res}
-  \caption{Results from PCA of the Procrustes superimposed ``half'' plastral landmarks.
-Depicted here are the for three PCs (lower triangle) and the mean shape with observed variance around each point (upper right).
-The first three PCs account for total \Sexpr{paste0(sum(fit$percent[1:3]))}\% of the variance in plastral shape.}
-  \label{fig:pca}
-\end{figure}
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics[width = \textwidth]{figure/pc_var}
-  \caption{Landmark variation along the first two PCs of the Procrustes superimposed ``half'' plastral landmarks.
-The first row corresponds to variation along the first PC, while the second row corresponds to the second PC.
-The left most column represents the observation with the highest eigenscore along that PC, while the right most column represents the observation with the lowest eigenscore.
-The middle column, for both rows, is the mean plastral shape for all observations.
-The first PC represents \Sexpr{fit$percent[1]}\% of the total variation in plastral shape while PC represents \Sexpr{fit$percent[2]}\% of the variance.}
-  \label{fig:pc_var}
-\end{figure}
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics[width = \textwidth]{figure/cent_size}
-  \caption{Comparison of the centroid size of each observation to the eigenscore of that obsrevation from the PCA of plastral shape.
-The first two PCs are illustrated from right to left.
-The regression line is fit via ordinary least squares and is presented to illustrate the difference in correlation between the first two axes.
-The shaded regions represent 95\% confidence intervals on the parameters of the regression line.
-Centroid size is significantly correlated with the first PC but not the second.}
-  \label{fig:cent}
-\end{figure}
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics[width = \textwidth]{figure/gap_res}
-  \caption{Gap statistic values for PAM clustering results for the \(\rho\) dissimliarity matrix of plastron shape.
-Error bars are standard errors estimated via \Sexpr{tadult.gap$B} bootstrap samples.}
-  \label{fig:gap}
-\end{figure}
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics{figure/gap_map}
-  \caption{Clustering solution for PAM with two mediods for the entire set of observed \textit{E.
-marmorata}.
-Clustering was based entirely on the \(\rho\) dissimilarity matrix of ``half'' plastral landmark configurations following Procrustes superimposition.
-Point shapes correspond to the two clusters while the colored ellipses correspond to 95\% confidence ellipses of the four groups from \citet{Spinks2005} and \citet{Spinks2010}.}
-  \label{fig:gap_map}
-\end{figure}
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics[width = \textwidth]{figure/roc_sel}
-  \caption{Effect of increasing the number of PCs as features, or predictors, of classification of plastra for all four classification schemes.
-As the total number of features increase, AUC increases until eventually leveling off.
-LDA-based classification, multinomial logisitic regression and random forest models are illustrated here, though AUC based model selection was only performed for the LDA-based classification and the random forest models.}
-  \label{fig:roc}
-\end{figure}
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics[width = \textwidth]{figure/gen_res}
-  \caption{Density estimates of AUC values of predictions of the testing dataset of plastra from \Sexpr{rr$sh1$R} bootstrap resamples.
-The top facet corresponds to values using the best LDA-based classifications of the eigenscores of shape, as chosen by maximum AUC.
-The middle facet corresponds to values using the optimal multinomial logistic regression model, as chosen by minimum AICc value.
-The bottom facet corresponds to the values using the optimal random forest model, as chosen by maximum AUC value.}
-  \label{fig:gen_res}
-\end{figure}
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics[width = \textwidth]{figure/gen_map}
-  \caption{Comparison between reference classification of testing data set (upper left) and the estimated classifications based on the selected LDA-based classification (lda, upper right), multinomial logistic regression (multi, lower left) and random forest models (rf, lower right).
-Classification corresponds to the four classes as suggested by the hypothesis of \citet{Spinks2005} and \citet{Spinks2010}.
-95\% confidence ellipses for the four classes are depicted and calculated from the initial assignments from the total dataset.}
-  \label{fig:gen_map}
-\end{figure}
-
-% mean shapes of all the classes
-\begin{figure}[ht]
-  \centering
-  \begin{subfigure}[b]{0.4\textwidth}
-    \centering
-    \includegraphics[width = \textwidth, height = 0.35\textheight, keepaspectratio = true]{figure/mshape_1}
-    \caption{Northern}
-    \label{fig:mean_shape1}
-  \end{subfigure}
-  \begin{subfigure}[b]{0.4\textwidth}
-    \centering
-    \includegraphics[width = \textwidth, height = 0.35\textheight, keepaspectratio = true]{figure/mshape_2}
-    \caption{Eastern}
-    \label{fig:mean_shape2}
-  \end{subfigure}
-  % new line. might want to make these closer together
-
-  \begin{subfigure}[b]{0.4\textwidth}
-    \centering
-    \includegraphics[width = \textwidth, height = 0.35\textheight, keepaspectratio = true]{figure/mshape_3}
-    \caption{Western}
-    \label{fig:mean_shape3}
-  \end{subfigure}
-  \begin{subfigure}[b]{0.4\textwidth}
-    \centering
-    \includegraphics[width = \textwidth, height = 0.35\textheight, keepaspectratio = true]{figure/mshape_4}
-    \caption{Southern}
-    \label{fig:mean_shape4}
-  \end{subfigure}
-  \caption{Thin-plate splines for each of the four classes from the best classification hypothesis based on the generalization results (Fig.
-\ref{fig:gen_res}).
-The four different classes are labeled according to the biogeographic groups as depicted in figure \ref{fig:gen_map}.
-The deformations are depicted with 2x magnification from base.}
-  \label{fig:mean_shape}
-\end{figure}
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics[width = \textwidth]{figure/pca_imp}
-  \caption{Pairs plot of the first three most important variables of the optimal random forest model of turtle plastral shape.
-The variables descend in importance from the upper left to the lower right.
-The observations are colored as in figure \ref{fig:gen_map}.
-The bottom row are histograms of classification occurrences along the PCs.}
-  \label{fig:imp_pc}
-\end{figure}
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics[width = \textwidth]{figure/imp_var}
-  \caption{Landmark variation along the two most important features (PCs) based on the final random forest model.
-The first row corresponds to the third PC and the second corresponds to the second PC.
-Landmark configurations are minimum observed on that PC, mean shape, and maximum observed on that PC.}
-  \label{fig:imp_var}
-\end{figure}
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics[width = \textwidth]{figure/rel_risk}
-  \caption{Forest plot of the relative risk, with 95\% confidence intervals, of classifying a give specimen based on the first three most important variables according to the random forest model.
-Relative risk values are calculated from the coefficients of the multinomial logistic regression model.
-All risks are relative to the northern group from \citet{Spinks2005,Spinks2010}.
-Variable importance is from left to right.}
-  \label{fig:rel_risk}
-\end{figure}
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics[width = \textwidth]{figure/lda}
-  \caption{Pairs plots of the three discriminate axes from the linear discriminate analysis of the eigenscores from the first \Sexpr{max(ldasel)} PCs of plastral shape.
-The observations figured are from from the training data set used for all models for the second molecular classification hypothesis based on \citet{Spinks2005} and \citet{Spinks2010}.
-Observations are colored as in Fig.
-\ref{fig:gen_map}.}
-  \label{fig:lda}
-\end{figure}
-
-\pagebreak
-% tables
-\begin{table}
-  \centering
-  \begin{tabular}[c]{ p{2cm} c | p{2cm} | p{2cm} |}
-    \cline{3-4} 
-    & & \multicolumn{2}{ c |}{Predicted class} \\ \cline{3-4}
-    & & 1 & 0 \\ \hline
-    \multicolumn{1}{| c |}{\multirow{2}{*}{Actual class}}
-    & 1 & TRUE \newline POSITIVE & FALSE \newline NEGATIVE \\ \cline{2-4}
-    \multicolumn{1}{| c |}{} & 0 & FALSE \newline POSITIVE & TRUE \newline NEGATIVE \\
-    \hline
-  \end{tabular}
-  \caption{Example confusion matrix.
-The columns correspond to the predicted class of an observation, while the rows correspond to the actual class of that observation.
-Depending on the type match between the prediction and reality, four different outcomes are possible: true positive (TP), false negative (FN), false positive (FP), and true negative (TN).
-These four quantities are used for calculating all confusion matrix statistics.
-Each of these values is an integer and the sum of the number of occurrences of that event during classification.}
-  \label{tab:conf}
-\end{table}
-
-\begin{table}
-  \centering
-  \include{xsex_tab}
-  \caption{Tabular comparison between sex observation and cluster assignement from PAM with two clusters.
-This number of clusters was chosen because it represented the second best clustering solution as determined via gap statistic comparison (Fig.
-\ref{fig:gap}).
-\(\chi^{2}\) analysis of this contigency table showed that there is no relationship between sex observation and cluster assignment (\(\chi^{2}\): \Sexpr{csex.chi$statistic}, df: \Sexpr{csex.chi$parameter}, \textit{p}-value: \Sexpr{csex.chi$p.value}).}
-  \label{tab:chi}
-\end{table}
-
-\begin{sidewaystable}
-  \centering
-  \include{tmx_1}
-  \caption{Model selection table for the multinomial logistic regression models of the first morphologically based classification hypothesis.
-This classification hypothesis corresponds to ``morph 1'' also depicted in figures \ref{fig:roc} and \ref{fig:gen_res}.
-This hypothesis is based on \citet{Seeliger1945}.
-The column ``delta'' corresponds to the \(\Delta\)AICc values of each model, while ``weights'' correspond to the Akaike weight of that model relative to all others.}
-  \label{tab:mod_sel_1}
-\end{sidewaystable}
-
-\begin{sidewaystable}
-  \centering
-  \include{tmx_2}
-  \caption{Model selection table for the multinomial logistic regression models of the first morphologically based classification hypothesis.
-This classification hypothesis corresponds to ``morph 2'' also depicted in figures \ref{fig:roc} and \ref{fig:gen_res}.
-This hypothesis is based on \citet{Seeliger1945}.
-The column ``delta'' corresponds to the \(\Delta\)AICc values of each model, while ``weights'' correspond to the Akaike weight of that model relative to all others.}
-  \label{tab:mod_sel_2}
-\end{sidewaystable}
-
-\begin{sidewaystable}
-  \centering
-  \include{tmx_3}
-  \caption{Model selection table for the multinomial logistic regression models of the first morphologically based classification hypothesis.
-This classification hypothesis corresponds to ``molec 1'' also depicted in figures \ref{fig:roc} and \ref{fig:gen_res}.
-This hypothesis is based on \citet{Seeliger1945}.
-The column ``delta'' corresponds to the \(\Delta\)AICc values of each model, while ``weights'' correspond to the Akaike weight of that model relative to all others.}
-  \label{tab:mod_sel_3}
-\end{sidewaystable}
-
-\begin{sidewaystable}
-  \centering
-  \include{tmx_4}
-  \caption{Model selection table for the multinomial logistic regression models of the first morphologically based classification hypothesis.
-This classification hypothesis corresponds to ``molec 2'' also depicted in figures \ref{fig:roc} and \ref{fig:gen_res}.
-This hypothesis is based on \citet{Seeliger1945}.
-The column ``delta'' corresponds to the \(\Delta\)AICc values of each model, while ``weights'' correspond to the Akaike weight of that model relative to all others.}
-  \label{tab:mod_sel_4}
-\end{sidewaystable}
-
-\begin{table}
-  \centering
-  \include{lda_tabx}
-  \caption{Results from pairwise Mann-Whitney U test between the AUC distributions of the generalizations of the LDA-based classification from the first 10 PCs of plastral shape.
-Labels correspond to those in Figure \ref{fig:gen_res}.
-Values of 0 correspond to p-values lower than 0.01.
-P-values were corrected for multiple comparison using the Holm method \citep{Holm1979}.}
-  \label{tab:lda_test}
-\end{table}
-
-\begin{table}
-  \centering
-  \include{mm_tabx}
-  \caption{Results from pairwise Mann-Whitney U test between the AUC distributions of the generalizations of the multinomial logistic regression models.
-Labels correspond to those in Figure \ref{fig:gen_res}.
-Values of 0 correspond to p-values lower than 0.01.
-P-values were corrected for multiple comparison using the Holm method \citep{Holm1979}.}
-  \label{tab:mm_test}
-\end{table}
-
-\begin{table}
-  \centering
-  \include{rf_tabx}
-  \caption{Results from pairwise Mann-Whitney U test between the AUC distributions of the generalizations of the random forest models.
-Labels correspond to those in Figure \ref{fig:gen_res}.
-Values of 0 correspond to p-values lower than 0.01.
-P-values were corrected for multiple comparison using the Holm method \citep{Holm1979}.}
-  \label{tab:rf_test}
-\end{table}
-
 \end{document}

### 2014-11-04 R/make_plots.r
added permutation test of miss classification

@@ -303,51 +303,3 @@ for(ii in seq(length(gsh))) {
   ggsave(file = paste0('../doc/figure/imp_var_', groups[[ii]], '.png'), 
          plot = gsh[[ii]], height = 15, width = 10)
 }
-
-
-# plot of the linear discriminate analysis
-# this needs to be generalized
-#lda.scal <- tl.a.analysis$best$spinks$scaling
-#lda.points <- as.data.frame(as.matrix(adult.train$spinks[, 1:7]) %*% lda.scal)
-#lda.points <- cbind(lda.points,
-#                    class = adult.train$spinks$spinks)
-#lda.points$class <- as.character(lda.points$class)
-#lda.points$class[lda.points$class == 1] = 'Northern'
-#lda.points$class[lda.points$class == 2] = 'Eastern'
-#lda.points$class[lda.points$class == 3] = 'Western'
-#lda.points$class[lda.points$class == 4] = 'Southern'
-#lda.points$class <- factor(lda.points$class,
-#                           levels = c('Northern',
-#                                      'Eastern',
-#                                      'Western',
-#                                      'Southern'))
-#
-#gglda <- ggpairs(lda.points, 
-#                 colour = 'class',
-#                 upper = 'blank',
-#                 lower = 'blank',
-#                 params = c(LabelSize = 2, gridLabelSize = 2, size = 1))
-#ld1 <- ggplot(lda.points, mapping = aes(x = LD1, y = LD2, colour = class))
-#ld1 <- ld1 + geom_point() + scale_color_manual(values = cbp)
-#ld2 <- ggplot(lda.points, mapping = aes(x = LD1, y = LD3, colour = class))
-#ld2 <- ld2 + geom_point() + scale_color_manual(values = cbp)
-#ld3 <- ggplot(lda.points, mapping = aes(x = LD2, y = LD3, colour = class))
-#ld3 <- ld3 + geom_point() + scale_color_manual(values = cbp)
-#lh1 <- ggplot(lda.points, mapping = aes(x = LD1, fill = class))
-#lh1 <- lh1 + geom_histogram()
-#lh1 <- lh1 + facet_grid(class ~ .) + scale_fill_manual(values = cbp)
[…]

### 2014-11-05 doc/draft_turtle.tex
permutation test write up

@@
 
-In classification studies, like this one, a common metric of performance is area under the reciver operating characteristic curve (AUC).
+In classification studies, like this one, a common metric of performance is area under the receiver operating characteristic curve (AUC).
 AUC is an estimate of the relationship between the false positive and true positive rates, as opposed to just the true positive rate (accuracy).
@@
 
+It is common for some out of sample observations to be misclassified.
+This misclassification may be due to the model not accurately representing shape variance, systematic differences between the training and test sets, or systematic differences between the accurately and inaccurately classified samples.
+Testing and training sets are determined completely at random within each class and with respect to shape.
+Results are not effected by changes in testing or training set assignment.
+
+To determine if there were systematic differences between the correctly and incorrectly classified samples, the multivariate centroids of the correct and incorrect groups were compared to what would be expected by random chance.
+The group labels were permuted and the difference between the new centroids was calculated.
+This was repeated 1000 times.
+The number of permutations less than the empirical difference were counted and divided by 1000, giving a \textit{p}-value.
+Significant results indicate that correctly and incorrectly classified specimens are systematically different.
+
 \section{Results}

### 2014-11-25 doc/draft_turtle.tex
initial results outline
started filling in results section.

@@
 Similar to LDA, the optimal number of PCs as predictors was determined by comparing within-sample AUC values across multiple models.
+% is this true?
 
@@
 Testing and training sets are determined completely at random within each class and with respect to shape.
-Results are not effected by changes in testing or training set assignment.
+Results were not effected by changes in testing or training set assignment.
 
@@
 Significant results indicate that correctly and incorrectly classified specimens are systematically different.
+This was done only when there were 10 or of that
 
 \section{Results}
+
+\subsection{Unsupervised learning}
+Comparison of gap statistic values from PAM clustering show that the best number of clusters is 1. This means that there is no means of naturally partitioning plastron shape into distinct subgroups.
+% PAM stuff
+% figure
+
+\subsection{Supervised learning}
+% model selection/fit
+For multinomial logistic regression, the best number of predictors was always the first four PCs.
+This was not the case for either of the LDA or random forest approaches which varied from two to six different PCs.
+Note that the best random forest models were determined via recursive feature selection, so PCs were not included in order of percent variance explained.
+
+% tabel of RFS choice for random forest
+
+% generalization
+AUC values from model generalization, or estimating testing data set membership, does indicate a clear ``best'' classification scheme (Fig.
+\ref{fig:gen_hist}).
+While the scheme with two species has the greatest AUC point estimate for each modeling approach, this scheme is not significantly greater from any other except in some limited cases (e.g.
+LDA).
+
+% misfits
+% need to make a table/graph to demonstrate this
+Differences in mean shape between correctly and incorrectly classified observations from test set were frequently statistically significant, though there are exceptions.
+The frequency of these results, however, is important because this means that the different models are poor predictors of class membership.
+This may be because differences in plastron shape do not align with the any of the hypothesized classification schemes.
+
+\section{Discussion}
 
@@
   \centering
-  \includegraphics{figure/plastra}
+  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/plastra}
   \caption{Depiction of general plastral shape of \textit{E. marmorata} and position of the 19 landmark used in this study. Anterior is towards the top of the figure.}
@@
 
+\begin{figure}[ht]
+  \centering
+  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/gen_res}
+  \caption{Bootstrap distributions for genearlized AUC values for each of the classification schemes.
+Each row corresponds to a different modeling approach: LDA, LDA using best variables from random forest, multinomial logistic regression, and random forest.
+Each distribution corresponds to 1000 bootstrap replicates.}
+  \label{fig:gen_hist}
+\end{figure}
+
 \end{document}

### 2014-11-25 R/make_plots.r

@@ -204,10 +204,12 @@ colnames(lr) <- groups
 zz <- list(rf = oo, mn = uu, lda = vv, lrf = lr)
 dd <- melt(zz)
 gdist <- ggplot(dd, aes(x = value, fill = Var2))
-gdist <- gdist + geom_histogram(alpha = 0.3,
+gdist <- gdist + geom_histogram(aes(y = ..density..),
+                                alpha = 0.3,
                                 size = 0.1,
                                 position = 'identity',
-                                colour = 'darkgrey')
+                                colour = 'darkgrey',
+                                binwidth = 1/100)
 gdist <- gdist + scale_fill_manual(name = '', values = cbp,
                                    labels = c('morph 1', 'morph 2',
                                               'molec 1', 'two species',

### 2014-11-25 R/make_tables.r

@@ -21,6 +21,7 @@ load('../data/gen.RData')
 
 
 # multinomial logistic regression
+
 against.best <- function(wb, mods) {
   best <- mods[[wb]]$t
   nm <- seq(length(mods))

### 2014-12-01 doc/draft_turtle.tex
filling out results

@@
 AUC ranges between 0.5 and 1, with 0.5 indicating classification no better than random and 1 indicating perfect classification \citep{Hastie2009}.
+AUC was used for both LDA and random forest models while multinomail logistic model selection was done using the likelihood based AICc \citep{Burnham2002a}.
 
@@
 
+For both the LDA and multinomial logistic regression models, increasing model complexity increased cumulative percent variation necessary to best model the differing classification schemes FIGURE OF CUMMULATIVE.
+The high variance in optimal number of PCs used as predictors between schemes may represent either ease of fitting that particular model or that the data was too noising to accurately fit any model beyond the simplest.
+
+As part of fitting a random forest model, a ranking of variable importance is also determined.
+Interestingly, the the order of variable importance is not the same as the order of principal components.
+This means that the principal components that best describe the differences between the various classes are not aligned with the principal components which describe the largest amount of variance.
+This result would be the case if variation between classes was extremely fine grained and not a part of the principal function or form of the plastron, which makes sense given that the plastron is involved in both protection and aquadynamics and not mate choice \citep{Germano2009,Holland1992,Lubcke2007,Rivera2008}.
+
+The observed AUC values for all of the optimal models are very low, values near 0.5 indicate that a model is no better than completely random assigment TABLE OR FIGURE.
+This means that in very few cases are any of the proposed classification schemes are generally poor descriptors of the observed variation.
+It appears that the data set is overwhelmed by noise, making any accurate classifications difficult at best.
+This observation is cemented with the generalizations of the models to the testing data set.
+
 % tabel of RFS choice for random forest
@@
 % generalization
-AUC values from model generalization, or estimating testing data set membership, does indicate a clear ``best'' classification scheme (Fig.
+AUC values from model generalization, or estimating testing data set membership, does not indicate a clear ``best'' classification scheme (Fig.
 \ref{fig:gen_hist}).
@@
 LDA).
+
+Mean AUC values for all model generalizations are, as expected, less than from the training set.
+This is expected because the low initial AUC values indicate little classification ability compared random, and the generalization to new data would only lead to loss in accuracy.
 
@@
 
+Generalization and the unsupervised learning results are surprisingly congruous.
+
+Plastron shape indicates no optimal classification scheme.
+A possible explanation for this that the genetic divergence associated with (sub)speciation is either not based on plastron morphology or local selective pressures due to hydrological regime overwhelming any possible morphological divergence.
+
+\clearpage
 \bibliographystyle{sysbio}

### 2014-12-01 R/make_plots.r

@@ -216,6 +216,7 @@ gdist <- gdist + scale_fill_manual(name = '', values = cbp,
                                               '2014', 'molec 2'))
 gdist <- gdist + facet_grid(L1 ~ ., labeller = gen.name)
 gdist <- gdist + labs(x = 'AUC')
+gdist <- gdist + coord_cartesian(xlim = c(0, 1))
 gdist <- gdist + theme(legend.title = element_blank(),
                        legend.margin = unit(0, 'cm'),
                        legend.text = element_text(size = 6))

### 2014-12-02 doc/draft_turtle.tex
importance and selection plots

@@
 
-For both the LDA and multinomial logistic regression models, increasing model complexity increased cumulative percent variation necessary to best model the differing classification schemes FIGURE OF CUMMULATIVE.
+For both the LDA and multinomial logistic regression models, increasing model complexity increased cumulative percent variation necessary to best model the differing classification schemes (Fig.
+\ref{fig:sel}).
 The high variance in optimal number of PCs used as predictors between schemes may represent either ease of fitting that particular model or that the data was too noising to accurately fit any model beyond the simplest.
@@
 As part of fitting a random forest model, a ranking of variable importance is also determined.
-Interestingly, the the order of variable importance is not the same as the order of principal components.
+Interestingly, the the order of variable importance is not the same as the order of principal components (Fig.
+\ref{fig:var_imp}).
 This means that the principal components that best describe the differences between the various classes are not aligned with the principal components which describe the largest amount of variance.
@@
 This observation is cemented with the generalizations of the models to the testing data set.
-
-% tabel of RFS choice for random forest
 
@@
   \centering
+  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/sel_val}
+  \caption{CAPTION}
+  \label{fig:sel}
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/var_imp}
+  \caption{Variable importance from the random forest models for each of the six classification schemes.
+Importance is measured as the mean decrease in Gini Index, which is a measure of the strength by which that variable determines CART structure.
+Indices that are farther to the right indicate greater variable importance.}
+  \label{fig:var_imp}
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
   \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/gen_res}

### 2014-12-02 R/make_tables.r

@@ -8,19 +8,6 @@ source('../R/plotting_functions.r')
 load('../data/gen.RData') 
 #load('../data/shape.RData')
 
-# 2 x 2 table of sex - cluster assignment
-#csex.tab <- rbind(csex.tab, tot = colSums(csex.tab))
-#csex.tab <- cbind(csex.tab, tot = rowSums(csex.tab))
-#xsex.tab <- xtable(csex.tab)
-#align(xsex.tab) <- 'r|cc|c'
-#digits(xsex.tab) <- c(0, 0, 0, 0)
-#print.xtable(xsex.tab,
-#             file = '../doc/xsex_tab.tex',
-#             floating = FALSE,
-#             hline.after = c(0, nrow(csex.tab) - 1))
-
-
-# multinomial logistic regression
 
 against.best <- function(wb, mods) {
   best <- mods[[wb]]$t

### 2014-12-02 R/supervised_analysis.r

@@ -49,3 +49,9 @@ trf.a <- Map(function(x, y) {
                  , ntree = 100
                  , metric = 'ROC')},
              x = adult.train, y = groups)
+for.imp <- Map(function(x, y) {
+               randomForest(x = x[, 1:floor(max.ad)]
+                            , y = x[, y]
+                            , ntree = 100
+                            , importance = TRUE)},
+               x = adult.train, y = groups)

### 2014-12-03 doc/draft_turtle.tex
miss classification tables

@@
 
+ % i have three tables for this. need to find a nice way of displaying these. also they require large amounts of edits
+\begin{table}
+  \begin{scriptsize}
+    \include{rf_miss}
+
+    \include{mm_miss}
+
+    \include{ll_miss}
+  \end{scriptsize}
+  \caption{Results of comparisons between correctly and incorrectly classified observations from the testing data set.
+For each scheme, the classifications with at least 10 observations were tested.
+This was done for each of the three modeling techniques included in this study.}
+  \label{tab:miss_tests}
+\end{table}
+
 \section{Discussion}
@@
   \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/sel_val}
-  \caption{CAPTION}
+  \caption{Graphical representation of the AICc and AUC values from model selection for multinomial logigistic regression and linear discriminate analysis, respecitively.
+AICc model selection is based on lowest AICc value while AUC model selection is based on greatest AUC value.
+The horizontal axis corresponds to the cummulative number of axes included in the model of interest.
+A red dot corresponds to the AICc or AUC best model for that classification scheme.}
   \label{fig:sel}

### 2014-12-03 R/make_tables.r

@@ -4,10 +4,7 @@ library(reshape2)
 library(plyr)
 
 source('../R/plotting_functions.r')
-
-load('../data/gen.RData') 
-#load('../data/shape.RData')
-
+source('../R/miss_class.r')
 
 against.best <- function(wb, mods) {
   best <- mods[[wb]]$t
@@ -31,78 +28,33 @@ rr.dif <- against.best(best.rr, rr)
 ll.dif <- against.best(best.ll, ll)
 
 
-# multinomial logistic regression
-mm.comp <- cbind(sh1 = mm$sh1$t, sh2 = mm$sh2$t, sh3 = mm$sh3$t, 
-                 sh4 = mm$sh4$t, sh5 = mm$sh5$t, spinks = mm$spinks$t)
-mm.c <- melt(mm.comp)
-mm.res <- with(mm.c, {
-               pairwise.t.test(value, Var2)})
-mm.tab <- mm.res$p.value
-mm.tab <- rbind('1' = rep(NA, 5), mm.tab)
-mm.tab <- cbind(mm.tab, '6' = rep(NA, 6))
-#colnames(mm.tab) <- coln
-#rownames(mm.tab) <- rown
-mm.tabx <- xtable(as.table(mm.tab))
-align(mm.tabx) <- 'r|rrrrrr'
-print.xtable(mm.tabx, 
-             file = '../doc/mm_tabx.tex',
-             hline.after = c(0, nrow(mm.tabx)),
-             include.rownames = TRUE,
-             floating = FALSE)
-
-# random forests
-rf.comp <- cbind(sh1 = rr$sh1$t, sh2 = rr$sh2$t, sh3 = rr$sh3$t, 
-                 sh4 = rr$sh4$t, sh5 = rr$sh5$t, spinks = rr$spinks$t)
-rf.c <- melt(rf.comp)
-rf.res <- with(rf.c, {
-               pairwise.t.test(value, Var2)})
-rf.tab <- rf.res$p.value
-rf.tab <- rbind('1' = rep(NA, 5), rf.tab)
-rf.tab <- cbind(rf.tab, '6' = rep(NA, 6))
-#colnames(rf.tab) <- coln
-#rownames(rf.tab) <- rown
-rf.tabx <- xtable(as.table(rf.tab))
-align(rf.tabx) <- 'r|rrrrrr'
-print.xtable(rf.tabx, 
-             file = '../doc/rf_tabx.tex',
-             hline.after = c(0, nrow(mm.tabx)),
[…]

### 2014-12-04 doc/draft_turtle.tex
generalization comparisons

@@
 While the scheme with two species has the greatest AUC point estimate for each modeling approach, this scheme is not significantly greater from any other except in some limited cases (e.g.
-LDA).
+LDA, Table \ref{tab:gen_tests}).
+
+\begin{table}
+  \include{rf_dif}
+    %
+  \include{mm_dif}
+    %
+  \include{ll_dif}
+  \caption{Results of bootstrap comparisons between the scheme with the highest mean AUC value and all other schemes.
+An asterix indicates the best scheme.
+This was done for each of the three modeling techniques included in this study.
+Probabilities are the percent of comparisons that are greater than the observed difference in means.}
+  \label{tab:gen_tests}
+\end{table}
 

### 2014-12-04 doc/draft_turtle.tex
multinomial selection buy AUC. fixed figure.

@@
 
-The observed AUC values for all of the optimal models are very low, values near 0.5 indicate that a model is no better than completely random assigment TABLE OR FIGURE.
+The observed AUC values for all of the optimal models are very low, values near 0.5 indicate that a model is no better than completely random assigment (Fig.
+\ref{fig:sel}) TABLE OR FIGURE.
 This means that in very few cases are any of the proposed classification schemes are generally poor descriptors of the observed variation.

### 2014-12-04 R/make_plots.r

@@ -155,34 +155,38 @@ ggsave(file = '../doc/figure/var_imp.png', plot = grf,
        width = 15, height = 5)
 
 # selection accumulation curves
-m.aic <- llply(tm.a.analysis$aic, function(x) {
+m.auc <- llply(tm.a.analysis$auc, function(x) {
                x <- data.frame(cbind(x, n = seq(length(x))))
                x})
-min.m <- cbind(x = laply(m.aic, function(x) min(x$x)), y = groups)
-min.m <- cbind(min.m, n = laply(m.aic, function(x) which.min(x$x)))
-m.aic <- Map(function(x, y) cbind(x, gr = rep(y, nrow(x))),
-             x = m.aic, y = groups)
-m.aic <- Reduce(rbind, m.aic)
-m.aic <- cbind(m.aic, ty = rep('aic', nrow(m.aic)))
+max.m <- cbind(x = laply(m.auc, function(x) max(x$x)), y = groups)
+max.m <- cbind(max.m, n = laply(m.auc, function(x) which.max(x$x)))
+m.auc <- Map(function(x, y) cbind(x, gr = rep(y, nrow(x))),
+             x = m.auc, y = groups)
+m.auc <- Reduce(rbind, m.auc)
+m.auc <- cbind(m.auc, ty = rep('multinomial logistic regression', nrow(m.auc)))
 
 l.auc <- llply(tl.a.analysis$auc, function(x) {
                x <- data.frame(cbind(x, n = seq(length(x))))
                x})
-max.l <- cbind(x = laply(l.auc, function(x) max(x)), y = groups)
+max.l <- cbind(x = laply(l.auc, function(x) max(x[, 1])), y = groups)
 max.l <- cbind(max.l, n = laply(l.auc, function(x) which.max(x$x)))
 l.auc <- Map(function(x, y) cbind(x, gr = rep(y, nrow(x))),
              x = l.auc, y = groups)
 l.auc <- Reduce(rbind, l.auc)
-l.auc <- cbind(l.auc, ty = rep('auc', nrow(l.auc)))
+l.auc <- cbind(l.auc, ty = rep('linear discriminiate analysis', nrow(l.auc)))
 
-maxes <- data.frame(rbind(cbind(min.m, ty = rep('aic', nrow(min.m))), 
-                          cbind(max.l, ty = rep('auc', nrow(max.l)))))
+maxes <- data.frame(rbind(cbind(max.m, 
+                                ty = rep('multinomial logistic regression', 
[…]

### 2014-12-04 R/model_comparison.r

@@ -37,7 +37,8 @@ clean.mods <- function(models, lab = c('sh1', 'sh2', 'sh3',
 
 # multinomial logicistic regression
 tm.a <- clean.mods(tmulti.a)
-tm.a.analysis <- multi.analysis(tm.a, ad.class, adult.test)
+tm.a.analysis <- multi.analysis(tm.a, class = groups, 
+                                test = adult.test, train = adult.train)
 
 # random forests
 trf.a <- clean.mods(trf.a)
@@ -46,7 +47,8 @@ trf.a.analysis <- rf.analysis(trf.a, ad.class, adult.test)
 # linear discriminate analysis
 tl.a <- clean.mods(tlda.a)
 tl.a <- lapply(tl.a, function(x) x[-1])
-tl.a.analysis <- lda.analysis(tl.a, adult.train, adult.test, ad.class)
+tl.a.analysis <- lda.analysis(tl.a, adult.train, adult.test, groups)
+
 
 # use the random forest importance for the LDA
 rf.best <- lapply(trf.a, function(x) x$optVariables)

### 2014-12-04 R/multi_funcs.r

@@ -24,12 +24,56 @@ multi.mod <- function(data, formula, preProc) {
 #' @param class vector; observed classes
 #' @param test df; testing data frame
 #' @return list of important model values
-multi.analysis <- function(model, class, test) {
+multi.analysis <- function(model, class, test, train) {
   out <- list()
   out$aic <- lapply(model, function(mod) unlist(lapply(mod, AICc)))
-  out$best <- Map(function(x, y) {
-                  mm <- which.min(x)
-                  y[[mm]] }, x = out$aic, y = model)
+  
+  # need to choose a best model
+  # do this based on within test auc
+  pr <- list()
+  for(ii in seq(length(model))) {
+    p <- list()
+    for(jj in seq(from = 2, to = length(model[[ii]]))) {
+      guess <- predict(model[[ii]][[jj]], 
+                       newdata = train[[ii]][, model[[ii]][[jj]]$coefnames[-1]])
+      guess <- data.frame(guess)
+      guess <- cbind(guess, 
+                     predict(model[[ii]][[jj]], 
+                             newdata = train[[ii]][, model[[ii]][[jj]]$coefnames[-1]], 
+                             type = 'probs'))
+      p[[jj - 1]] <- guess
+    }
+    pr[[ii]] <- p
+  }
+
+  aucs <- list()
+  for(ii in seq(length(pr))) {
+    oo <- list()
+    for(jj in seq(length(pr[[ii]]))) {
+      pred <- pr[[ii]][[jj]][, 1]
+      obs <- train[[ii]][, groups[[ii]]]
+      pred.res <- pr[[ii]][[jj]][, -1]
+      if(length(levels(pred)) == 2) {
+        mauc <- roc(obs, pred.res)$auc
+      } else {
+        prob <- lapply(levels(pred), function(cl) {
+                       pp <- ifelse(pred == cl, 1, 0)
+                       oo <- ifelse(obs == cl, 1, 0)
+                       prob <- pred.res[, cl]
+
+                       ps <- Metrics::auc(oo, prob)
+                       ps})
+
+        mauc <- colMeans(do.call(rbind, prob))
+      }
+      oo[[jj]] <- mauc
+    }
+    aucs[[ii]] <- unlist(oo)
+  }
+  out$auc <- aucs
+
+  out$best <- Map(function(x, y) x[y], x = model, y = llply(aucs, which.max))
+
   pc <- mapply(predict, out$best, test,
                MoreArgs = list(type = 'class'), SIMPLIFY = FALSE)
[…]

### 2014-12-04 R/multiclass_roc.r

@@ -53,7 +53,7 @@ multiClassSummary <- function(data, lev = NULL, model = NULL) {
 
 # multiclass roc
 allvone <- function(pred.res, obs) {
-  pred <- factor(pred.res$pred)
+  pred <- factor(pred.res[, 1])
   prob <- lapply(levels(pred), function(class) {
                  pp <- ifelse(pred == class, 1, 0)
                  oo <- ifelse(obs == class, 1, 0)

### 2014-12-04 doc/draft_turtle.tex
comparison table between means of obs and gen

--- 2014-12-04 doc/draft_turtle.tex
@@
 AUC ranges between 0.5 and 1, with 0.5 indicating classification no better than random and 1 indicating perfect classification \citep{Hastie2009}.
-AUC was used for both LDA and random forest models while multinomail logistic model selection was done using the likelihood based AICc \citep{Burnham2002a}.
 
@@
 % model selection/fit
-For multinomial logistic regression, the best number of predictors was always the first four PCs.
-This was not the case for either of the LDA or random forest approaches which varied from two to six different PCs.
+Generally, as many PCs were included as predictors as possible for the best models of each of the classification schemes (Fig.
+\ref{fig:sel}).
 Note that the best random forest models were determined via recursive feature selection, so PCs were not included in order of percent variance explained.
-
 For both the LDA and multinomial logistic regression models, increasing model complexity increased cumulative percent variation necessary to best model the differing classification schemes (Fig.
@@
 
-The observed AUC values for all of the optimal models are very low, values near 0.5 indicate that a model is no better than completely random assigment TABLE OR FIGURE.
+The observed AUC values for all of the optimal models are low, as values near 0.5 indicate that a model is no better than completely random assigment (Fig.
+\ref{fig:sel}).
 This means that in very few cases are any of the proposed classification schemes are generally poor descriptors of the observed variation.
@@
 
-Mean AUC values for all model generalizations are, as expected, less than from the training set.
-This is expected because the low initial AUC values indicate little classification ability compared random, and the generalization to new data would only lead to loss in accuracy.
+Mean AUC values for the model generalizations, in most cases, are approximately equal to the observed AUC values from the training data set (Table \ref{tab:comp}).
+In cases were the AUC from the  genearlizations is less than the observed, this points too poor model fit and a poor classification scheme.
+
+\begin{table}[ht]
+  \centering
+  \include{comp_tab}
+  \caption{CAPTION}
+  \label{tab:comp}
+\end{table}
 
@@
   \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/sel_val}
-  \caption{Graphical representation of the AICc and AUC values from model selection for multinomial logigistic regression and linear discriminate analysis, respecitively.
-AICc model selection is based on lowest AICc value while AUC model selection is based on greatest AUC value.
+  \caption{Graphical representation of the AUC values from model selection for multinomial logigistic regression and linear discriminate analysis, respecitively.
+AUC model selection is based on greatest AUC value.
 The horizontal axis corresponds to the cummulative number of axes included in the model of interest.
-A red dot corresponds to the AICc or AUC best model for that classification scheme.}
+A red dot corresponds to the AUC best model for that classification scheme.}
   \label{fig:sel}

### 2014-12-04 R/make_tables.r

@@ -3,11 +3,11 @@ library(xtable)
 library(reshape2)
 library(plyr)
 library(boot)
+library(pROC)
+library(Metrics)
 
 source('../R/plotting_functions.r')
 source('../R/miss_class.r')
-
-# observed AUC values of best models
 
 
 against.best <- function(wb, mods) {
@@ -21,6 +21,28 @@ against.best <- function(wb, mods) {
              x = sets, n = nana)
   dif
 }
+
+# observed AUC values of best models
+# compared to AUC for generalization
+mm.mean <- llply(mm, function(x) as.numeric(x$t0))
+mm.comp <- data.frame(cbind(sch = names(mm.mean), 
+                            obs = laply(tm.a.analysis$auc, max)),
+                            gen = unlist(mm.mean))
+
+rf.mean <- llply(rr, function(x) as.numeric(x$t0))
+rf.comp <- data.frame(cbind(sch = names(rf.mean), 
+                            obs = laply(trf.a.analysis$auc, max)),
+                            gen = unlist(rf.mean))
+ll.mean <- llply(ll, function(x) as.numeric(x$t0))
+ll.comp <- data.frame(cbind(sch = names(ll.mean), 
+                            obs = laply(tl.a.analysis$auc, max)),
+                            gen = unlist(ll.mean))
+comp.tab <- cbind(rf.comp, mm.comp[, -1], ll.comp[, -1])
+comp.tab <- xtable(comp.tab)
+label(comp.tab) <- 'comp_tab'
+print.xtable(x = comp.tab, 
+             file = '../doc/comp_tab_raw.tex')
+
 
 #generalize comparisons
 best.mm <- which.max(laply(mm, function(x) mean(x$t)))

### 2014-12-05 doc/draft_turtle.tex
methods, results mostly done. discussion points outlined. need to re-do unsupervised learning.

@@
 % model selection/fit
+AUC based model selection revealed some important patterns of variation and congruence between a given classification scheme and the actual data.
 Generally, as many PCs were included as predictors as possible for the best models of each of the classification schemes (Fig.
@@
 \ref{fig:sel}).
-The high variance in optimal number of PCs used as predictors between schemes may represent either ease of fitting that particular model or that the data was too noising to accurately fit any model beyond the simplest.
+That almost all models were as complex as possible indicates that the differences between the different groups within each classification scheme are very small.
 
 As part of fitting a random forest model, a ranking of variable importance is also determined.
-Interestingly, the the order of variable importance is not the same as the order of principal components (Fig.
+Interestingly, the order of variable importance is not the same as the order of decreasing explained variance per principal component (Fig.
 \ref{fig:var_imp}).
@@
 This result would be the case if variation between classes was extremely fine grained and not a part of the principal function or form of the plastron, which makes sense given that the plastron is involved in both protection and aquadynamics and not mate choice \citep{Germano2009,Holland1992,Lubcke2007,Rivera2008}.
+This is congruous with the results of the AUC based model selection for the multinomial logistic regression and LDA models.
 
@@
 LDA, Table \ref{tab:gen_tests}).
+
+Mean AUC values for the model generalizations, in most cases, are approximately equal to the observed AUC values from the training data set (Table \ref{tab:comp}).
+In cases were the AUC from the  genearlizations is less than the observed, this points too poor model fit and a poor classification scheme.
+
+% misfits
+Differences in mean shape between correctly and incorrectly classified observations from test set were frequently statistically significant, though there are exceptions.
+The frequency of these results, however, is important because this means that the different models are poor predictors of class membership.
+This may be because differences in plastron shape do not align with the any of the hypothesized classification schemes.
+
+\section{Discussion}
+
+The results of this study indicate that there is no clear grouping of plastra shapes in \textit{E. marmorata}.
+
+The unsupervised learning results which indicate only a single group of observations being optimal is congruous with the results from the generalizations of the supervised learning models.
+The classification schemes used in the supervised learning models correspond, loosely, to unsupervised learning solutions with multiple groups.
+Because unsupervised learning solutions with multiple groups are poor descriptors of the observed variation, it is important to see this reinforced by the supervised learning results.
+
+The results from fitting the various supervised learning models for each of the classification scheme generally shows that no one scheme is ``best.'' A possible explanation for this that the genetic divergence associated with (sub)speciation is either not based on plastron morphology or local selective pressures due to hydrological regime overwhelming any possible morphological divergence.
+
+Both the low AUC values (\(< 0.9\)) and the significant difference between the correctly and incorrectly classified observations support the conclusion that none of the hypothesized classification schemes are good descriptors of the observed plastral variation.
+
+\clearpage
+\bibliographystyle{sysbio}
+\bibliography{turtle,packages}
+
+\clearpage
 
@@
 
-Mean AUC values for the model generalizations, in most cases, are approximately equal to the observed AUC values from the training data set (Table \ref{tab:comp}).
-In cases were the AUC from the  genearlizations is less than the observed, this points too poor model fit and a poor classification scheme.
-
 \begin{table}[ht]
@@
   \include{comp_tab}
-  \caption{CAPTION}
+  \caption{AUC values for the best model of each classification scheme for both the observed (training) data and the generalized (testing) data.
+Results from all three different supervised learning approaches are shown here.
+AUC values range between 0.5 and 1.
+}
   \label{tab:comp}
@@
 
-% misfits
-% need to make a table/graph to demonstrate this
-Differences in mean shape between correctly and incorrectly classified observations from test set were frequently statistically significant, though there are exceptions.
-The frequency of these results, however, is important because this means that the different models are poor predictors of class membership.
-This may be because differences in plastron shape do not align with the any of the hypothesized classification schemes.
-
- % i have three tables for this. need to find a nice way of displaying these. also they require large amounts of edits
 \begin{table}
@@
 \end{table}
-
-\section{Discussion}
-
-Generalization and the unsupervised learning results are surprisingly congruous.
-
-Plastron shape indicates no optimal classification scheme.
-A possible explanation for this that the genetic divergence associated with (sub)speciation is either not based on plastron morphology or local selective pressures due to hydrological regime overwhelming any possible morphological divergence.
-
-\clearpage
-\bibliographystyle{sysbio}
-\bibliography{turtle,packages}
-
-\clearpage
 

### 2014-12-08 doc/draft_turtle.tex
unsupervised learning results

@@
 The \(\rho\) dissimilarity matrix was divisively clustered using partitioning around mediods clustering (PAM), a method similar to \textit{k}-means clustering except that instead of minimizing the sum of squared Euclidean distances between observations and centroids, the sum of squared dissimilarities between observations and mediods is minimized \citep{Kaufman1990}.
-Because the optimal number of clusters of shape configurations in the study was unknown, being possibly three, four, or some other value, clustering solutions were estimated with the number of clusters varied between one and NUMBER.
+Because the optimal number of clusters of shape configurations in the study was unknown, being possibly three, four, or some other value, clustering solutions were estimated with the number of clusters varied between one and eight.
 Clustering solutions were compared using the gap statistic, which is a measure of goodness of clustering \citep{Tibshirani2001a}.
@@
 This value is averaged and compared to the expected dispersion (\(E^{*}_{n}\)) of a sample \(n\) from a reference distribution.
-In this case, the reference distribution was estimated from NUMBER resamples.
+In this case, the reference distribution was estimated from 500 resamples.
 
@@
 \subsection{Unsupervised learning}
-Comparison of gap statistic values from PAM clustering show that the best number of clusters is 1. This means that there is no means of naturally partitioning plastron shape into distinct subgroups.
+Comparison of gap statistic values from PAM clustering show that the best number of clusters is most likely 1 (Fig.
+\ref{fig:gap}).
+There is some ambiguity in choice because, although it is not statistically different from a solution with only one group, the solution with 2 groupings does have the greatest mean gap statistic.
+However, there is no geographical signal to the results of this clustering solution (Fig.
+\ref{fig:gap_map}).
+Because of this, we assert that this means that there is no means of naturally partitioning plastron shape into distinct subgroups.
 % PAM stuff
@@
   \centering
+  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/gap_res}
+  \caption{CAPTION}
+  \label{fig:gap}
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/gap_map}
+  \caption{CAPTION}
+  \label{fig:gap_map}
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
   \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/sel_val}

### 2014-12-08 R/clustering.r

@@ -16,16 +16,8 @@ n.groups = 4
 
 set.seed(1)
 tpam <- function(x, k) pam(as.dist(x), k)
-tmorph.gap <- clusGap(info.dist, FUNcluster = tpam, 
-                      K.max = 10 * n.groups, B = 500)
 tadult.gap <- clusGap(adult.dist, FUNcluster = tpam,
-                      K.max = 10 * n.groups, B = 500)
-tmorph.km <- pam(as.dist(info.dist), k = which.max(tmorph.gap$Tab[, 3]))
+                      K.max = 2 * n.groups, B = 500)
 tadult.km <- pam(as.dist(adult.dist), k = which.max(tadult.gap$Tab[, 3]))
 
-# dissimilarity based evidence accumulation clustering
-set.seed(1)
-tmorph.dac <- dac(info.dist, krange = c(1, 50 * n.groups), iter = 10000)
-tadult.dac <- dac(adult.dist, krange = c(1, 50 * n.groups), iter = 10000)
-
-save.image(file = 'cluster_res.RData')
+save(tadult.gap, tadult.km, file = '../data/cluster_res.RData')

### 2014-12-08 R/plot_cluster.r

@@ -10,18 +10,18 @@ library(scales)
 library(shapes)
 library(geomorph)
 library(devtools)
+library(cluster)
 
 source('../R/plotting_functions.r')
+source('../R/supervised_mung.r')
 
 source_url('https://raw.github.com/JoFrhwld/FAAV/master/r/stat-ellipse.R')
 
-load('../data/shape.RData')
+load('../data/cluster_res.RData')
 
 theme_set(theme_bw())
-
-cbp <- c(#'#999999', 
-         '#E69F00', '#56B4E9', '#009E73', 
-         '#F0E442', '#0072B2', '#D55E00', '#CC79A7')
+cbp <- c('#E69F00', '#56B4E9', '#009E73', '#F0E442', 
+         '#0072B2', '#D55E00', '#CC79A7')
 
 adult$long[which(adult$long > -100)] <- adult$long[which(adult$long > -100)] - 100
 adult.train$spinks$long[which(adult.train$spinks$long > -100)] <- adult.train$spinks$long[which(adult.train$spinks$long > -100)] - 100
@@ -41,24 +41,17 @@ goog.map <- get_map(location = c(longmi,
 ggmap(goog.map)
 
 
-# correlation between the first two axes and centroid size
-bcs <- c(cs, cs)
-cent <- cbind(cs = bcs, pc = c(adult$PC1, adult$PC2))
-cent <- cbind(as.data.frame(cent),
-              lab = c(rep('PC1', length(adult$PC1)),
-                      rep('PC2', length(adult$PC2))))
-gcs <- ggplot(cent, aes(x = cs, y = pc)) 
-gcs <- gcs + geom_point() + stat_smooth(method = 'lm')
-gcs <- gcs + facet_wrap(~lab)
-gcs <- gcs + labs(x = 'centroid size', y = 'eigenscore')
-ggsave(file = '../documents/figure/cent_size.png', plot = gcs)
-
-
 # gap statistic plot
 gap <- gap.plot(tadult.gap)
-gap <- gap + theme(axis.title = element_text(size = 9.5),
-                   axis.text = element_text(size = 8))
-ggsave(file = '../documents/figure/gap_res.png', plot = gap)
+gap <- gap + theme(axis.text = element_text(size = 20),
+                   axis.title = element_text(size = 30),
+                   legend.text = element_text(size = 25),
+                   legend.title = element_text(size = 26),
+                   legend.key.size = unit(2, 'cm'),
[…]

### 2014-12-08 doc/draft_turtle.tex
full methods/results draft

@@
 \section{Materials and Methods}
-\subsection{Specimens and sampling}
-We collected landmark-based morphometric data from SAMPLE adult \textit{E.
+\subsection{Specimens, sampling, morphometrics}
+We collected landmark-based morphometric data from 354 adult \textit{E.
 marmorata} museum specimens.
-These specimens include both newly sampled individuals and those sampled in previous studies of plastral shape variation \citep{Angielczyk2007,Angielczyk2011,Angielczyk2013a}.
-
-Specimen classification was based on known specimen geographic information which was recorded from museum collection information.
+These specimens include both newly sampled and previously sampled individuals \citep{Angielczyk2007,Angielczyk2011,Angielczyk2013a}.
+Specimens were assigned a classification for each of the different schemes was based on known geographic information recorded from museum collection information.
 When precise latitude and longitude information was not available it was estimated from whatever locality information was present.
 Because the specimens used to define the subclades in \citet{Spinks2005} and \citet{Spinks2010} were not available for study, all specimen classifications were based solely on this geographic information and not from explicit assignment in previous studies.
-Instead, classification was based on matching museum locality data with the geographic boundaries of the molecularly-defined clades of \citet{Spinks2005} and \citet{Spinks2010}.
-Because the exact barriers between different biogeographic regions are unknown and unclear, two assignments for both the morphologically and molecularly based hypotheses were used.
-Each morphologically based hypothesis had three classes, while each molecular-based had four classes.
-In total, each specimen was given four different classifications.
+Because the exact barriers between different biogeographic regions are unknown and unclear, each hypothesis was represented by two different schemes; in total six different schemes were compared.
+% need more information from ken
 
-\subsection{Geometric morphometrics}
 Following previous work on plastral variation \citep{Angielczyk2007,Angielczyk2011,Angielczyk2013a}, 19 landmarks were digitized using TpsDig 2.04 \citep{Rohlf2005}.
@@
 \subsubsection{Unsupervised learning}
-In order to preserve the relationship between all landmark configurations in shape space, the dissimilarity between observations was measured using Kendall's Riemanninan shape distance or \(\rho\) \citep{Kendall1984a,Dryden1998a}.
+In order to preserve the relationship between all landmark configurations in shape space, the dissimilarity between observations was measured using Kendall's Riemannian shape distance or \(\rho\) \citep{Kendall1984a,Dryden1998a}.
 This metric was chosen because shape space, or the set of all possible shape configurations following Procrustes superimposition, is a Riemannian manifold and thus non-Euclidean \citep{Dryden1998a}.
@@
 Clustering solutions were compared using the gap statistic, which is a measure of goodness of clustering \citep{Tibshirani2001a}.
-The gap statistic is defined
 
-\[Gap_{n}(k) = E^{*}_{n}[\log(W_{k})] - \log(W_{k})\]
-
-where \(W_{k}\) is
-
-\[W_{k} = \sum^{k}_{r = 1}{\frac{1}{2n_{r}} (\sum_{i,i' \in C_{r}} d_{ii'})}\].
-
-\(d_{ii'}\) is the dispersion of the clustering solution or the sum of the pairwise dissimilarities between observations in each cluster and their respective mediods (\(C\)) for all clusters \(r\).
+%The gap statistic is defined
+%\[Gap_{n}(k) = E^{*}_{n}[\log(W_{k})] - \log(W_{k})\] 
+%where \(W_{k}\) is
+%\[W_{k} = \sum^{k}_{r = 1}{\frac{1}{2n_{r}} (\sum_{i,i' \in C_{r}} d_{ii'})}\].
+%\(d_{ii'}\) is the dispersion of the clustering solution or the sum of the pairwise dissimilarities between observations in each cluster and their respective mediods (\(C\)) for all clusters \(r\).
 This value is averaged and compared to the expected dispersion (\(E^{*}_{n}\)) of a sample \(n\) from a reference distribution.
@@
 
-The observed AUC values for all of the optimal models are low, as values near 0.5 indicate that a model is no better than completely random assigment (Fig.
+Observed AUC values for all of the optimal models are not exceptionally high, as values near 0.5 indicate that a model is no better than completely random assigment (Fig.
 \ref{fig:sel}).
-This means that in very few cases are any of the proposed classification schemes are generally poor descriptors of the observed variation.
+This means that in all but a few cases the different proposed classification schemes are generally poor descriptors of the observed variation.
 It appears that the data set is overwhelmed by noise, making any accurate classifications difficult at best.
@@
 % generalization
+Mean AUC values for the model generalizations, in most cases, are approximately equal to the observed AUC values from the training data set (Table \ref{tab:comp}).
+In cases were the AUC from the  genearlizations is less than the observed, this points too poor model fit and a poor classification scheme.
 AUC values from model generalization, or estimating testing data set membership, does not indicate a clear ``best'' classification scheme (Fig.
@@
 LDA, Table \ref{tab:gen_tests}).
-
-Mean AUC values for the model generalizations, in most cases, are approximately equal to the observed AUC values from the training data set (Table \ref{tab:comp}).
-In cases were the AUC from the  genearlizations is less than the observed, this points too poor model fit and a poor classification scheme.
-
 % misfits
@@
 
+\begin{table}[ht]
+  \centering
+  \include{comp_tab}
+  \caption{AUC values for the best model of each classification scheme for both the observed (training) data and the generalized (testing) data.
+Results from all three different supervised learning approaches are shown here.
+AUC values range between 0.5 and 1.
+}
+  \label{tab:comp}
+\end{table}
+
 \begin{table}
@@
   \label{tab:gen_tests}
-\end{table}
-
-\begin{table}[ht]
-  \centering
-  \include{comp_tab}
-  \caption{AUC values for the best model of each classification scheme for both the observed (training) data and the generalized (testing) data.
-Results from all three different supervised learning approaches are shown here.
-AUC values range between 0.5 and 1.
-}
-  \label{tab:comp}
 \end{table}
@@
   \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/gap_res}
-  \caption{CAPTION}
+  \caption{Results from PAM clustering of the Riemannian shape distance for 8 different number of clusters. Vertical lines are 1 standard deviation of the mean determined from 500 resamples.}
   \label{fig:gap}
@@
   \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/gap_map}
-  \caption{CAPTION}
+  \caption{Comparison of geographic distribution of clustered observations from the 2 clustering PAM solution.
+Colour and shape correspond to each of the groups.
+There is clearly no geographic signal in the data.}
   \label{fig:gap_map}

### 2014-12-22 doc/old/turtle_paper.tex
moved files around, ken comments.

--- 2013-08-21 documents/turtle_paper.tex
@@
 
-\documentclass[12pt,letterpaper]{article}\usepackage{graphicx, color}
+\documentclass[12pt,letterpaper]{article}\usepackage[]{graphicx}\usepackage[]{color}
 %% maxwidth is the original width if it is less than linewidth
@@
 
-\definecolor{fgcolor}{rgb}{0.2, 0.2, 0.2}
-\newcommand{\hlnumber}[1]{\textcolor[rgb]{0,0,0}{#1}}%
-\newcommand{\hlfunctioncall}[1]{\textcolor[rgb]{0.501960784313725,0,0.329411764705882}{\textbf{#1}}}%
-\newcommand{\hlstring}[1]{\textcolor[rgb]{0.6,0.6,1}{#1}}%
-\newcommand{\hlkeyword}[1]{\textcolor[rgb]{0,0,0}{\textbf{#1}}}%
-\newcommand{\hlargument}[1]{\textcolor[rgb]{0.690196078431373,0.250980392156863,0.0196078431372549}{#1}}%
-\newcommand{\hlcomment}[1]{\textcolor[rgb]{0.180392156862745,0.6,0.341176470588235}{#1}}%
-\newcommand{\hlroxygencomment}[1]{\textcolor[rgb]{0.43921568627451,0.47843137254902,0.701960784313725}{#1}}%
-\newcommand{\hlformalargs}[1]{\textcolor[rgb]{0.690196078431373,0.250980392156863,0.0196078431372549}{#1}}%
-\newcommand{\hleqformalargs}[1]{\textcolor[rgb]{0.690196078431373,0.250980392156863,0.0196078431372549}{#1}}%
-\newcommand{\hlassignement}[1]{\textcolor[rgb]{0,0,0}{\textbf{#1}}}%
-\newcommand{\hlpackage}[1]{\textcolor[rgb]{0.588235294117647,0.709803921568627,0.145098039215686}{#1}}%
-\newcommand{\hlslot}[1]{\textit{#1}}%
-\newcommand{\hlsymbol}[1]{\textcolor[rgb]{0,0,0}{#1}}%
-\newcommand{\hlprompt}[1]{\textcolor[rgb]{0.2,0.2,0.2}{#1}}%
+\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
+\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
+\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
+\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
+\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
+\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
+\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
+\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
+\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
+\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
 
@@
 % latex table generated in R 3.0.1 by xtable 1.7-1 package
-% Wed Aug 21 14:14:34 2013
+% Wed Sep  4 10:03:26 2013
 \begin{tabular}{r|cc|c}
@@
 % latex table generated in R 3.0.1 by xtable 1.7-1 package
-% Wed Aug 21 14:14:34 2013
+% Wed Sep  4 10:03:26 2013
 {\small
@@
 % latex table generated in R 3.0.1 by xtable 1.7-1 package
-% Wed Aug 21 14:14:34 2013
+% Wed Sep  4 10:03:26 2013
 \begin{tabular}{r|rrrr}

### 2014-12-22 doc/old/turtle_report.Rnw
moved files around, ken comments.

--- 2013-04-08 documents/turtle_report.Rnw
@@
+% this document has the propensity to become outdated as the code base
+% and results are updated over time. the pdf file should represent the most
+% stable implementation of this document.
+%
+% this is the basic ``template'' for all reports for meetings with ken and to
+% send to collaborators.
 \documentclass{article}
@@
 source('../src/plotting_functions.r')
-load('../src/turtle_analysis.RData')
+#load('../src/turtle_analysis.RData')
+load('../src/turtle_analysis_TEMP.RData')
 @
@@
 
-<<gap, eval = FALSE>>=
+<<gap, eval = TRUE>>=
 gg.gap <- gap.plot(tmorph.gap)
@@
 \end{figure}
+
+<<a-gap, eval = TRUE>>=
+ad.gap <- gap.plot(tadult.gap)
+ad.gap
+@
 
@@
 
-<<gm, eval = FALSE>>=
+<<gm, eval = TRUE>>=
 gg.morph <- ggpairs(turtle.info, columns = c(1:3),
@@
 \subsection{Supervised}
+
+\subsubsection{Model selection}
+% multinomial logistic regression AICc tables
+%
+% nnet recursive feature selection
+%
+% random forest recursive feature selection
 
@@
 
-<<resamp, eval = FALSE>>=
+<<resamp, eval = TRUE>>=
 tmod.re.clean <- lapply(tmod.re, function(x) clean.resamp(x$values))
@@
 
-<<multi-map, eval = FALSE>>=
+<<multi-map, eval = TRUE>>=
 tm.map <- mapply(class.map
@@
 
-<<rf-map, eval = FALSE>>=
+<<rf-map, eval = TRUE>>=
 trf.class.tiny <- lapply(trf.class, function(x) x[[1]])

### 2014-12-22 doc/old/turtle_report.tex
moved files around, ken comments.

--- 2013-04-08 documents/turtle_report.tex
@@
 
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
 \definecolor{fgcolor}{rgb}{0.2, 0.2, 0.2}
@@
 
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error: cannot open the connection}}\end{kframe}
+\end{knitrout}
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error: object 'turtle.train' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error: object 'turtle.test' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error: object 'turtle.train' not found}}\end{kframe}
+\end{knitrout}
+
 \title{How cryptic is cryptic diversity? Machine learning approaches to fine scale variation in the morphology of \textit{Emys marmorata}.}
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
 
@@
 \subsection{Unsupervised}
-Comparison of gap statistic over a very wide range of plausible partitions indicates that as the number of partitions increase, there is a marginal increasing gap statistic until approximately 40 after which there is a marginal decrease in gap statistic (Fig.
+Comparison of gap statistic over a very wide range of plausible partitions indicates that as the number of partitions increase, there is a marginal increasing gap statistic until approximately
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\\Error in eval(expr, envir, enclos) : object 'tmorph.km' not found}} after which there is a marginal decrease in gap statistic (Fig.
 \ref{fig:gap}).
 It is notable that the standard errors around the gap statistic values are very large, and the marginal increases in gap statistic with an increased number of partitions may not be important.
-Additionally, all gap statistic values are within 0.0064 of each other meaning that there is little over all consensus for how many clusters are present when comparing gap statistics.
+Additionally, all gap statistic values are within
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\\Error in eval(expr, envir, enclos) : object 'tmorph.gap' not found}} of each other meaning that there is little over all consensus for how many clusters are present when comparing gap statistics.
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error: object 'tmorph.gap' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error: object 'gg.gap' not found}}\end{kframe}
+\end{knitrout}
 
@@
 
-Dissimilarity based EAC estimated approximately 5 optimal partitions (Table \ref{tab:dac}) though this may be only a marginal difference.
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error: object 'tadult.gap' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error: object 'ad.gap' not found}}\end{kframe}
+\end{knitrout}
+
+Dissimilarity based EAC estimated approximately
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\\Error in eval(expr, envir, enclos) : object 'tmorph.dac' not found}} optimal partitions (Table \ref{tab:dac}) though this may be only a marginal difference.
 I need to increase the number of resamples from 100 to 500 because that is when approximate stability is estimated to occur.
 
-% latex table generated in R 2.15.3 by xtable 1.7-0 package
-% Mon Apr  8 09:29:56 2013
-\begin{table}[ht]
-\begin{center}
-\begin{tabular}{rr}
-  \hline
- & tmorph.dac \\ 
-  \hline
-1 & 722 \\ 
-  2 &   1 \\ 
-  3 &   1 \\ 
-  4 &   1 \\ 
-  5 &   1 \\ 
-   \hline
-\end{tabular}
-\caption{Number of specimens assigned to optimal number of partitions as determined by dissimilarity based EAC.
-Each column corresponds to a different partition, with the number assigned directly below it.}
-\label{tab:dac}
-\end{center}
-\end{table}
+\begin{kframe}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error: object 'tmorph.dac' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error: object 'dac.tab' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error: object 'dac.tab' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error: object 'dac.tab' not found}}\end{kframe}
 
@@
 I haven't put nay classification labels on the points because an optimal partition scheme has not been selected.
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error: object 'turtle.info' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error: object 'turtle.land.info' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error: object 'fits' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error: object 'gg.morph' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error: object 'gg.morph' not found}}\end{kframe}
+\end{knitrout}
 
@@
 
+\subsubsection{Model selection}
+% multinomial logistic regression AICc tables
+%
+% nnet recursive feature selection
+%
+% random forest recursive feature selection
+
 I'm currently holding back on showing all the tables and model selection procedures because there are a lot of them.
@@
 
-% latex table generated in R 2.15.3 by xtable 1.7-0 package
-% Mon Apr  8 09:29:56 2013
-\begin{table}[ht]
-\begin{center}
-\begin{tabular}{rrrr}
-  \hline
- & CCR & marm & pall \\ 
-  \hline
-CCR &  40 &  14 &  11 \\ 
-  marm &  25 &  74 &   2 \\ 
-  pall &   3 &   1 &  10 \\ 
-   \hline
-\end{tabular}
-\caption{Confusion matrix for sh1 classification scheme. Rows correspond to predicted class. Columns correspond to reference class. From multinomial logistic regression models.}
-\label{tab:multi-conf-sh1}
-\end{center}
-\end{table}
-% latex table generated in R 2.15.3 by xtable 1.7-0 package
-% Mon Apr  8 09:29:56 2013
-\begin{table}[ht]
-\begin{center}
-\begin{tabular}{rrrr}
-  \hline
- & CCR & marm & pall \\ 
-  \hline
-CCR &  44 &  12 &  10 \\ 
-  marm &  17 &  82 &   6 \\ 
-  pall &   2 &   1 &   7 \\ 
-   \hline
-\end{tabular}
-\caption{Confusion matrix for sh2 classification scheme. Rows correspond to predicted class. Columns correspond to reference class. From multinomial logistic regression models.}
-\label{tab:multi-conf-sh2}
-\end{center}
-\end{table}
-% latex table generated in R 2.15.3 by xtable 1.7-0 package
-% Mon Apr  8 09:29:56 2013
-\begin{table}[ht]
-\begin{center}
-\begin{tabular}{rrrrr}
-  \hline
- & CCR & marm & pall & SJ \\ 
-  \hline
-CCR &  37 &  11 &   9 &   2 \\ 
-  marm &  18 &  59 &   3 &  10 \\ 
-  pall &   7 &   2 &  10 &   0 \\ 
-  SJ &   1 &   4 &   1 &   7 \\ 
-   \hline
-\end{tabular}
-\caption{Confusion matrix for sh3 classification scheme. Rows correspond to predicted class. Columns correspond to reference class. From multinomial logistic regression models.}
-\label{tab:multi-conf-sh3}
-\end{center}
-\end{table}
-% latex table generated in R 2.15.3 by xtable 1.7-0 package
-% Mon Apr  8 09:29:56 2013
-\begin{table}[ht]
-\begin{center}
-\begin{tabular}{rrrrr}
-  \hline
- & 1 & 2 & 3 & 4 \\ 
-  \hline
-1 & 107 &   9 &  13 &  13 \\ 
-  2 &   1 &   8 &   0 &   0 \\ 
-  3 &   4 &   0 &   8 &   3 \\ 
-  4 &   4 &   0 &   2 &   8 \\ 
-   \hline
-\end{tabular}
-\caption{Confusion matrix for spinks classification scheme. Rows correspond to predicted class. Columns correspond to reference class. From multinomial logistic regression models.}
-\label{tab:multi-conf-spinks}
-\end{center}
-\end{table}
+\begin{kframe}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error: object 'tm.conf' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error: object 'mt' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error: object 'mt' not found}}\end{kframe}
 
 But I can show the comparison of the predictive accuracies (Fig. \ref{fig:resamp}).
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error: object 'tmod.re' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error: object 'tmod.re.clean' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error: object 'tmod.re.gg' not found}}\end{kframe}
+\end{knitrout}
 
@@
 In the interest of space/time, I'm only going to display results from one of the selected models from each method for each classification scheme.
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error: object 'turtle.train' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error: object 'tm.map' not found}}\end{kframe}
+\end{knitrout}
 
@@
 \end{figure}
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error: object 'trf.class' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error: object 'turtle.train' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error: object 'trf.map' not found}}\end{kframe}
+\end{knitrout}
 

### 2014-12-22 doc/draft_turtle.tex
edited draft to reflect ken comments

@@
 marmorata} museum specimens.
-These specimens include both newly sampled and previously sampled individuals \citep{Angielczyk2007,Angielczyk2011,Angielczyk2013a}.
-Specimens were assigned a classification for each of the different schemes was based on known geographic information recorded from museum collection information.
-When precise latitude and longitude information was not available it was estimated from whatever locality information was present.
-Because the specimens used to define the subclades in \citet{Spinks2005} and \citet{Spinks2010} were not available for study, all specimen classifications were based solely on this geographic information and not from explicit assignment in previous studies.
-Because the exact barriers between different biogeographic regions are unknown and unclear, each hypothesis was represented by two different schemes; in total six different schemes were compared.
-% need more information from ken
+These specimens are a subset of those included in \citet{Angielczyk2007}, \citet{Angielczyk2011}, and \citet{Angielczyk2013a} and represents adult individuals.
+We chose to focus on adults because significant changes in plastron shape occur over the course of ontogeny in \textit{E.
+mamorata} and other emydines \citep{Angielczyk2013a}.
+We assigned a classification to each specimen for the different binning schemes based on geographic occurrence data recorded in museum collection archives.
+When precise latitude and longitude information was not available we estimated it from whatever locality information was present.
+Because the specimens sampled to obtain the genetic data used to define the subclades were not available for study, all specimen classifications were based solely on the geographic information, not explicit assignment in previous studies.
+Because the exact barriers between different biogeographic regions are unknown and unclear, we represented each hypothesis with two different schemes; se we compared a total of six different schemes.
+% How were adults chosen?
+Good to note how schemes differeed.
 
-Following previous work on plastral variation \citep{Angielczyk2007,Angielczyk2011,Angielczyk2013a}, 19 landmarks were digitized using TpsDig 2.04 \citep{Rohlf2005}.
-These landmarks were chosen to maximize the description of general plastral variation(Fig.
-\ref{fig:plastra}).
-17 of these landmarks are at the endpoints or intersection of the keratinous plastral scutes that cover the platron.
-12 of these landmarks were chosen to be symmetrical across the axis of symmetry and, in order to prevent degrees of freedom and other concerns \citep{Klingenberg2002}, prior to analysis these landmarks were reflected across the axis of symmetry (i.e.
-midline) and the average position of each symmetrical pair was used.
-In cases where damage or incompleteness prevented symmetric landmarks from being determined, only the single member of the pair was used.
-Analysis was conducted on the resulting ``half'' plastra.
-Plastral landmark configurations were superimposed using generalized Procrustes analysis \citep{Dryden1998a} after which, the principal components (PC) of shape were calculated.
-This was done using the \texttt{shapes} package for R \citep{2013,Dryden2013}.
+Following previous work on plastron variation \citep{Angielczyk2007,Angielczyk2011,Angielczyk2013a}, we used TpsDig 2.04 \citep{Rohlf2005} to digitize 19 landmarks \ref{fig:plastra}).
+Seventeen of the landmarks are at the endpoints or intersection of the keratinous plastral scutes that cover the platron.
+Twelve of the landmarks were symmetrical across the axis of symmetry and, in order to prevent degrees of freedom and other concerns \citep{Klingenberg2002}, we reflected these landmarks across the axis of symmetry (i.e.
+midline) prior to analysis and used the average position of each symmetrical pair.
+In cases where damage or incompleteness prevented symmetric landmarks from being determined, we used only the single member of the pair.
+We conducted all subsequent analyses on the resulting ``half'' plastra.
+We superimposed the plastral landmark configurations using generalized Procrustes analysis \citep{Dryden1998a}, after which, we calculated the principal components (PC) of shape using the \texttt{shapes} package for R \citep{2013,Dryden2013}.
 
@@
 \subsubsection{Unsupervised learning}
-In order to preserve the relationship between all landmark configurations in shape space, the dissimilarity between observations was measured using Kendall's Riemannian shape distance or \(\rho\) \citep{Kendall1984a,Dryden1998a}.
-This metric was chosen because shape space, or the set of all possible shape configurations following Procrustes superimposition, is a Riemannian manifold and thus non-Euclidean \citep{Dryden1998a}.
+In order to preserve the relationship between all landmark configurations in shape space, we measured the dissimilarity between observations using Kendall's Riemannian shape distance or \(\rho\) \citep{Kendall1984a,Dryden1998a}.
+We chose this metric because shape space, or the set of all possible shape configurations following Procrustes superimposition, is a Riemannian manifold and thus non-Euclidean \citep{Dryden1998a}.
 \(\rho\) varies between 0 and \(\pi / 2\) when there is no reflection invariance, which should not be a concern in the case of the half plastral landmark configurations used in the study.
 
-The \(\rho\) dissimilarity matrix was divisively clustered using partitioning around mediods clustering (PAM), a method similar to \textit{k}-means clustering except that instead of minimizing the sum of squared Euclidean distances between observations and centroids, the sum of squared dissimilarities between observations and mediods is minimized \citep{Kaufman1990}.
-Because the optimal number of clusters of shape configurations in the study was unknown, being possibly three, four, or some other value, clustering solutions were estimated with the number of clusters varied between one and eight.
-Clustering solutions were compared using the gap statistic, which is a measure of goodness of clustering \citep{Tibshirani2001a}.
+We divisively cluster the \(\rho\) dissimilarity matrix using partitioning around mediods clustering (PAM), a method similar to \textit{k}-means clustering except that instead of minimizing the sum of squared Euclidean distances between observations and centroids, the sum of squared dissimilarities between observations and mediods is minimized \citep{Kaufman1990}.
+Because the optimal number of clusters of shape configurations in the study was unknown, being possibly three, four, or some other value, we estimated clustering solutions in which the number of clusters varied between one and eight.
+We compated clustering solutions using the gap statistic, which is a measure of goodness of clustering \citep{Tibshirani2001a}.
 
@@
 
-This analysis was conducted using the \texttt{cluster} package for R \citep{Maechler2013}.
+We conducted this analysis using the \texttt{cluster} package for R \citep{Maechler2013}.
 
 \subsubsection{Supervised learning}
-Three different supervised learning, or classification, approaches were used: linear discriminate analysis, multinomial logistic regression, and random forests.
-Linear discriminiate analysis, also known as canonical variate analysis, is commonly used in studies of geometric morphometric data \citep{Zelditch2004,Mitteroecker2011}.
+We used three different supervised learning, or classification, approaches: linear discriminate analysis, multinomial logistic regression, and random forests.
+Linear discriminate analysis, also known as canonical variate analysis, is commonly used in studies of geometric morphometric data \citep{Zelditch2004,Mitteroecker2011}.
 The other two methods, however, are not.
-Each of these three methods has a different interpretation and reveal very different aspects of the data.
-In all cases, the optimal number of PCs used as predictors was chosen via max within-sample AUC value, explained below.
+Each of these three methods has a different interpretation and they reveal very different aspects of the data.
+In all cases, the optimal number of PCs used as predictors was chosen via maximum within-sample AUC value, explained below.
 
-Linear discriminate analysis (LDA) attempts to find a linear combination of predictors to best model two or more classes.
+Linear discriminate analysis (LDA) attempts to find a linear combination of predictors that best model two or more classes.
 LDA is very similar to PCA except that instead of finding the linear combination of features that maximize the amount of explained variance in the data, LDA maximizes the differences between classes.
 The results of this analysis produces a transformation matrix by which the original features can be transformed to reflect the best discrimination between the classes.
-In this analysis, LDA was applied on the eigenscores from a subset of the total number of PCs, ranging from two to NUMBER in increasing order of complexity.
+In this study, we applied LDA to the eigenscores from a subset of the total number of PCs, ranging from two to NUMBER in increasing order of complexity.
 In total, this produced nine different LDA scaling matrices.
@@
 Similar to the odds ratios calculated from the coefficients of a logistic regression, the relative risk of a classification can be determined from the coefficients of the model.
-Similar to LDA, the optimal number of PCs as predictors was determined by comparing within-sample AUC values across multiple models.
-% is this true?
+Similar to the LDA, we determined the optimal number of PCs as predictors by comparing within-sample AUC values across multiple models.
 
 Random forest models are an extension of classification and regression trees (CART) \citep{Breiman1984,Breiman2001}.
-The goal of CARTs are to use a series of different features to estimate the class of an observation.
-In top-down induction of decision trees for each member of a given set of predictor variables, attribute value test are used to estimate the differences between classes.
+The goal of CARTs are to use a series of different features to estimate the class of an observation (specimen).
+In top-down induction of decision trees for each member of a given set of predictor variables, attribute value tests are used to estimate the differences between classes.
 This process, called recursive partitioning, is then repeated on each subset.
@@
 The resulting model is a tree structure by which observations are classified at each intersection via the estimated cutoff points from the attribute tests made during model fitting.
+% figure to help explain the ``tests''
 
-In a random forest model, many CARTs are built from a random subsample of both the features and the observations.
-This process is then repeated many times and the parameters of the final model was chosen as the mode of estimates from the distribution of CARTs \citep{Breiman2001}.
+In a random forest model, many CARTs are built from a random subsample of both the features and the observations (specimens).
+This process is then repeated many times and the parameters of the final model are chosen as the mode of estimates from the distribution of CARTs \citep{Breiman2001}.
 In addition to classifying the observations, this procedure allows for the features to be ranked in order of importance.
-This is a generally important property that is useful for many other studies which want to describe and model the differences between classes and the relative importance of different predictors.
+This is a generally important property that is useful for many other studies in which the goal is to describe and model the differences between classes and the relative importance of different predictors.
 
-In this analysis, random forest model parameters were estimated from 1000 subtrees.
-The best set of predictors necessary for each classification scheme was estimated using a recursive feature selection algorithm was used to choose the optimal number of PCs to include based on the AUC of the model.
+In this analysis, we used 1000 subtrees to estimate the random forest model parameters.
+We estimated the best set of predictors necessary for each classification scheme was estimated using a recursive feature selection algorithm, and we chose the optimal number of PCs to include based on the AUC of the model.
 Following the backwards selection algorithm implemented in \texttt{caret} \citep{Kuhn2013}, the maximum number of features were included in the initial model, their importance ranked, and the AUC of the model calculated.
@@
 This was repeated until only one feature, remained.
-Because PCs were kept in order of importance and not in relation to the amount of variance each PC described, these means that the PCs are not included in assending eigenvalue.
+Because PCs were kept in order of importance and not in relation to the amount of variance each PC described, these means that the PCs are not included in the order of ascending eigenvalue.
 
-In classification studies, like this one, a common metric of performance is area under the receiver operating characteristic curve (AUC).
+In classification studies, such as this one, a common metric of performance is area under the receiver operating characteristic curve (AUC).
 AUC is an estimate of the relationship between the false positive and true positive rates, as opposed to just the true positive rate (accuracy).
@@
 
-The standard AUC calculation is defined for of binary classification, however in this application there are multiple categories.
-The alternative calculation used here follows an all-against-one strategy where the individual AUC values for each class versus all others are averaged to produce a multiclass AUC \citep{Hand2001}.
+The standard AUC calculation is defined for binary classifications, however in this application there are multiple categories.
+The alternative calculation that we used follows an all-against-one strategy where the individual AUC values for each class versus all others are averaged to produce a multiclass AUC \citep{Hand2001}.
+To estimate confidence intervals on the out-of-sample AUC values, we performed a nonparametric bootstrap in which the true and estimated classifications were resampled with replacement.
+This was done 1000 times.
 
@@
 Within-sample performance is inherently biased upwards, so model evaluation requires overcoming this bias.
-With very large sample sizes, as in this study, part of the sample can be used as the ``training set'' and the remainder acts as the ``testing set.'' The former is used for fitting the model while the later is used for measuring model performance.
-This is called model generalization.
-In this analysis, 75\% of samples were used as the training set while the remaining 25\% were used as the testing set.
-
-In order to estimate confidence intervals on the out-of-sample AUC values, a nonparametric bootstrap was performed where the true and estimated classifications were resampled with replacement.
-This was done 1000 times.
+With very large sample sizes, as in this study, part of the sample can be used as the ``training set'' and the remainder acts as the ``testing set.'' The former is used for fitting the model where as the later is used for measuring model performance, and this process is called model generalization.
+In this analysis, we used 75\% of samples as the training set while the remaining 25\% were used as the testing set.
 
@@
 
-To determine if there were systematic differences between the correctly and incorrectly classified samples, the multivariate centroids of the correct and incorrect groups were compared to what would be expected by random chance.
-The group labels were permuted and the difference between the new centroids was calculated.
-This was repeated 1000 times.
-The number of permutations less than the empirical difference were counted and divided by 1000, giving a \textit{p}-value.
+To determine if there were systematic differences between the correctly and incorrectly classified samples, we compared the multivariate centroids of the correct and incorrect groups to what would be expected by random.
+The group labels were permuted 1000 times and the difference between the new centroids was calculated.
+The number of permutations less than the empirical difference divided by 1000 gives a \textit{p}-value for the test.
 Significant results indicate that correctly and incorrectly classified specimens are systematically different.
-This was done only when there were 10 or of that
+This was done only for classes where there were 10 or more observations.
 
@@
 \subsection{Unsupervised learning}
-Comparison of gap statistic values from PAM clustering show that the best number of clusters is most likely 1 (Fig.
+Comparison of gap statistic values from PAM clustering show that the best number of clusters is most likely one (Fig.
 \ref{fig:gap}).
-There is some ambiguity in choice because, although it is not statistically different from a solution with only one group, the solution with 2 groupings does have the greatest mean gap statistic.
-However, there is no geographical signal to the results of this clustering solution (Fig.
+There is some ambiguity in choice because, although it is not statistically different from a solution with only one group, the solution with two groupings does have the greatest mean gap statistic.
+However, there is no geographical signal in the results of this clustering solution (Fig.
 \ref{fig:gap_map}).
@@
 % model selection/fit
-AUC based model selection revealed some important patterns of variation and congruence between a given classification scheme and the actual data.
-Generally, as many PCs were included as predictors as possible for the best models of each of the classification schemes (Fig.
+AUC--based model selection revealed some important patterns of variation and congruence between the classification schemes and the actual data.
+Generally, as many PCs as possible were included as predictors for the best models of each of the classification schemes (Fig.
 \ref{fig:sel}).
@@
 
-As part of fitting a random forest model, a ranking of variable importance is also determined.
+As part of fitting a random forest model, a ranking of variable importance also is determined.
 Interestingly, the order of variable importance is not the same as the order of decreasing explained variance per principal component (Fig.
@@
 This means that the principal components that best describe the differences between the various classes are not aligned with the principal components which describe the largest amount of variance.
-This result would be the case if variation between classes was extremely fine grained and not a part of the principal function or form of the plastron, which makes sense given that the plastron is involved in both protection and aquadynamics and not mate choice \citep{Germano2009,Holland1992,Lubcke2007,Rivera2008}.
-This is congruous with the results of the AUC based model selection for the multinomial logistic regression and LDA models.
+Another way of phrasing this is that the variance describing the differences between the classes does not align with the major axes of variance (i.e.
+the PCs).
+This result would be the case if variation between classes was extremely fine grained and not a part of the principal form or function of the plastron, which makes sense given that the plastron is involved in both protection and hydrodynamics and not mate choice \citep{Germano2009,Holland1992,Lubcke2007,Rivera2008}.
+Moreover, this result is congruous with the results of the AUC--based model selection for the multinomial logistic regression and LDA models.
 
-Observed AUC values for all of the optimal models are not exceptionally high, as values near 0.5 indicate that a model is no better than completely random assigment (Fig.
+Observed AUC values for all of the optimal models are not exceptionally high; values near 0.5 indicate that a model is no better than completely random assignment (Fig.
 \ref{fig:sel}).
@@
 Mean AUC values for the model generalizations, in most cases, are approximately equal to the observed AUC values from the training data set (Table \ref{tab:comp}).
-In cases were the AUC from the  genearlizations is less than the observed, this points too poor model fit and a poor classification scheme.
+The  cases in which the AUC from the  generalizations is less than the observed, indicate poor model fit and a poor classification scheme.
 AUC values from model generalization, or estimating testing data set membership, does not indicate a clear ``best'' classification scheme (Fig.
 \ref{fig:gen_hist}).
-While the scheme with two species has the greatest AUC point estimate for each modeling approach, this scheme is not significantly greater from any other except in some limited cases (e.g.
+Although the scheme with two species has the greatest AUC point estimate for each modeling approach, this scheme is not significantly greater than any other except in some limited cases (e.g.
 LDA, Table \ref{tab:gen_tests}).
 % misfits
-Differences in mean shape between correctly and incorrectly classified observations from test set were frequently statistically significant, though there are exceptions.
-The frequency of these results, however, is important because this means that the different models are poor predictors of class membership.
+Differences in mean shape between correctly and incorrectly classified observations from test set frequently were statistically significant, though there are exceptions.
+Again, this test was to determine if the mean shapes were statistically different or not.
+The frequency of these results, however, is important because it means that the different models are poor predictors of class membership.
 This may be because differences in plastron shape do not align with the any of the hypothesized classification schemes.

### 2015-01-16 doc/draft_turtle.tex
intro updated with ken comments

@@
 
+% cryptic diversity
+%   most species are still deliminated solely based on morphology
+With the rise of molectular systematics and techniques such as DNA barcoding, it has been increasingly recognized that some taxa can only be differented using molecular data \citep{Bickford2007,SchlickSteiner2007,Stuart2006,Pfenninger2007,Funk2012,Clare2011}.
+Nevertheless, the majority of extant taxa, and nearly all extinct taxa, are delimited based solely via morphology.
+This disjuct is of great concern for studies of variation and diversity dynamics in deep time, where apparent morphological stasis may not reflect the true underlying diversity that existed \citep{Hunt2008,Eldredge1972,Gould1977a,VanBocxlaer2013}.
+
+Much work has been devoted to species delimitation via sequence difference \citep{Fujita2012,Yang2010b} which in turn has sparked interest in whether geometric morphometric anlaysis can captur similarly fine-scale variation that can be used for identifying cryptic species.
+Most such studies focus on morphometrics to discover differences between taxa that were idetified by other means \citep{Polly2007a,Demandt2009,Gaubert2005b,Gunduz2007,Polly2003,Zelditch2004} and automated taxon identification \citep{MacLeod2007}.
+It would be very useful, though, if morphometrics approaches could reliably identify cryptic species without the use of other types of data.
+This would make the task of idenitfying and maintaining endangered or conserved groups much easier and could contribute to improved classifications of extinct taxa.
+
+Here, we attempt to address this issue using machine learning approaches.
+In particular, we ask whether it is possible to determine which amongst a set of classification hypotheses is best and examine the implications of the results for a recently proposed set of cryptic turtle species.
+
+% past work on automatic taxon identification and older approaches to classifying taxa
+% why use machine learning methods
+\subsection{Background and system}
+Machine learning is in many respects just an extension of known statistical methodology \citep{Hastie2009} in which the emphasis is placed on inferring rules and properties of data in order to explain the underlying structure.
+The basic statistical mechanics are supplemented by randomization, sorting, and partitioning algorithms and along with the maximization or minimization of summary statistics, in order to best estimate a general model for all data, both sampled and unsampled \citep{Hastie2009}.
+Machine learning approaches have found use in medical research, epidemiology, economics and automated image identification such as handwritten zip codes \citep{Hastie2009}.
+Two major classes of machine learning methods are unsupervised and supervised learning.
+Unsupervised learning methods are used with unlabeled data where the underlying structure is estimated and are analogous to clustering and density estimation methods \citep{Kaufman1990}.
+Supervised learning methods are used with labeled data where the final output of data is known and the rules for going from input to output are inferred.
+These are analogous classification and regression models \citep{Breiman1984}.
+The application of the alternative approaches used in this study illustrates only a sampling of the various previously derived methods for clustering observations and fitting classification models.
+
+Geometric morphometric approaches to idetifying differences in morphological variation between different classes, including cryptic species, mostly have used methods like linear discriminate analysis and canonical variates analysis \citep{Zelditch2004,Mitteroecker2011,Polly2007a,Polly2003,Gunduz2007,Gaubert2005b,Demandt2009,Sztencel-Jabonka2009,MitrovskiBogdanovic2013,Francoy2009}.
+These methods are comparatively straightforward ways of understanding the differences in morphology between classes given their similarity to familiar multivariate approaches like Principal Components Analysis.
+They are benefit by producing results that can be easily visualized, which aids in the interpretation and presentation of data and results.
+Most previous morphometric studies did not assess which amongst a set of alternative classification hypotheses was optimal.
+For example, studies such as those of \citet{Caumul2005a} and \citet{Polly2007a} focused on comparing different aspects of morphology and their fidelity to a classification scheme instead of comparing the fidelity of one aspect of morphology to multiple classification schemes.
+In this context, the study of \citet{Cardini2009a}, is noteworthy because they compared morphological variation in marmots at the population, regional, and species level and determined the fidelity of shape to divisions at each of these levels.
+
+Here, we used multiple machine learning methods, both unsupervised and supervised, in order to compare different classification hypotheses.
+These methods provide different and unique advantages for understanding how to classify taxa, with what accuracy, and what these classifications are based on.
+While machine learning methods such as neural networks have been applied to studying shape variation \citep{Baylac2003,Dobigny2003,VandenBrink2011,MacLeod2007}, including in the context of automated taxon identification and classification of groups, but the number of cases remains limited.
+In the current study, we not only consider pure classification accuracy but also use a statistic of classification strength that reflects the rate at which taxa are both accurately and inaccurately classified.
+
+\subsection{\textit{Emys marmorata}}
+We applied our machine learning methods to the problem whether distinct subspecies or cryptic species extist within the western pond turtle, \textit{Emys marmorata} (formerly \emph{Clemmys marmorata}; see \citealp{Feldman2002}).
+\textit{E.
+marmorata} is distributed from northern Washington State, USA to Baja California, Mexico.
+Traditionally, \textit{E.
+marmorata} was classified into two named subspecies: the northern \textit{E.
+marmorata marmorata}, the southern \textit{E.
+marmorata palida} \citep{Seeliger1945}.
+She also recognized a central Californian intergrade zone between these subspecies, and she did not formally include the Baja California populations of \textit{E.
+marmorata} in these groups, implying the existence of a third distinct but unnamed subspecies.
+\textit{E.
+marmorata marmorata} is differentiated from \textit{E.
+marmorata palida} by the presence of a pair of triangular inguinal plates and darker neck markings.
+It should be noted that the triangular inguinal plates can sometimes be present in \textit{E.
+marmorata palida} though they are considerably smaller.
+
+Previous work on morphological variation in \textit{E.
+marmorata} has focused primarily on differentiation between populations over a portion of the species' total range \citep{Lubcke2007,Germano2009,Germano2008,Bury2010}; comparatively few studies have included specimens from across the entire range \citep{Holland1992}.
+Most of these studies considered how local biotic and abiotic factors may contribute to differences in carapace length and found that size can vary greatly between different populations \citep{Lubcke2007,Germano2009,Germano2008}.
+There also has been interest in size-based sexual dimorphosim in \textit{E.
+marmorata} \citep{Lubcke2007,Germano2009,Holland1992}, with males being on average larger than females based on total carapace length and other linear measurements.
+However, the quality of size as a classifier of sex can vary greatly between populations \citep{Holland1992}, which makes sense in light of the amount of between population size difference \citep{Lubcke2007,Germano2009}.
+However, the effect of sexual dimorphism on shape, \textit{sensu} \citet{Kendall1977a}, was not assessed \citep{Holland1992,Lubcke2007,Germano2008}.
+
+Of particular importance in the context of cryptic diversity in \textit{E.
+marmorata} is the traditional morphometric analysis of carapace shape carried out by \citet{Holland1992}.
+He compared morphological differences between and among many populations of \textit{E.
+marmorata} in three different areas of the species' range, and studied the relative effect of distance versus barriers in fostering morphological differentiation in the species.
+\citet{Holland1992} concluded that distance was a poor indicator of mophological differentiation, and the geographic features such as breaks between different drainage basis are probably more important barriers to reproduction.
+Additionally, \citet{Holland1992} found that morphological differences were observable as the magnitude of barriers and distance increased, but the variation required many variables to adequately capture, implying only very subtle morphological differentiation between putatively distinct populations.
+\citet{Holland1992} concluded that \textit{E.
+marmorata} is best classified as three distinct species: a northern species, souther species, and a Columbia basic species.
+This classification is similar to that of \citet{Seeliger1945}, except elevated to the species level and without recognition of a distinct Baja species.
+
+More recently, the phylogeography of \textit{E.
+marmorata} and the possibility of cryptic diversity has been investigated using molecular data \citep{Spinks2005,Spinks2010,Spinks2014}.
+Based on mitochondrial DNA, \citet{Spinks2005} recognized four subclades within \textit{E.
+marmorata}, a northern clade, a San Joaquin Valley clade, a Santa Barbara clade, and a southern clade.
+Analyses with nuclear DNA \citep{Spinks2010} with single-nucleotide polymorphism (SNP) data suggest a primarily north--south division in \textit{E.
+marmorata}, although the dataset differe in the location of this break point.
+In these papers, Spinks and colleagues discussed the potential taxonomic implications of their results, with \citet{Spinks2014} going so far as to strongly advocate for the recognition of at least two species (\emph{E.
+marmorata} and \emph{E.
+pallida}), and a possible third based on populations in Baja California.
+However, they did not discuss in detail the morphological characters that would help to diagnose these species beyond those specified by \citet{Seeliger1945}.
+Given that these characters are somewhat variable within the proposed species, and that \citet{Holland1992} described shell shape variation that might be consistent with this taxonomy, we wondered whether a geometric morphometric analysis of shell shape would provide a more reliable way of diagnosing groups (whether species or subspecies) within \textit{E.
+marmorata}.
+
+In this study, we attempt to estimate the best classification scheme of \textit{E.
+marmorata} based on variation in plastral (central shell) shape in order to determine whether this character is consistent with any of the past divisions based on other morphological features or molecular data.
+Because of unclear geographic boundaries between subgroups of \textit{E.
+marmorata}, we compare multiple hypotheses of morphologically-- and molecularly--based classification.
+We hypothesize that if morphological variation corresponds to class assignment, then it should be possible to determine the best classification hypothesis of \textit{E.
+marmorata} from amongst multiple candidate hypotheses.
+However, if morphological variation variation does not correspond to any classification hypothesis, then supervised learning model generalization performance will be poor and reflect how variation may not follow along with any of the candidate classification hypotheses.
+
 \section{Materials and Methods}
@@
 mamorata} and other emydines \citep{Angielczyk2013a}.
+\uppercase{Peter note -- how was something classified as an adult?}
+
 We assigned a classification to each specimen for the different binning schemes based on geographic occurrence data recorded in museum collection archives.
@@
 Because the exact barriers between different biogeographic regions are unknown and unclear, we represented each hypothesis with two different schemes; se we compared a total of six different schemes.
+The schemes differed based on where geographic boundaries were assigned.
+This changes how certain individuals were assigned two one of the groups within in hypothesis such as which of the three morphologically defined groups, which of the four mitochondrially defined groups, and so on.
 % How were adults chosen?

### 2015-08-05 doc/draft_turtle.tex
hand edits done. need list of missing and update group names.

@@
 %   most species are still deliminated solely based on morphology
-With the rise of molectular systematics and techniques such as DNA barcoding, it has been increasingly recognized that some taxa can only be differented using molecular data \citep{Bickford2007,SchlickSteiner2007,Stuart2006,Pfenninger2007,Funk2012,Clare2011}.
-Nevertheless, the majority of extant taxa, and nearly all extinct taxa, are delimited based solely via morphology.
-This disjuct is of great concern for studies of variation and diversity dynamics in deep time, where apparent morphological stasis may not reflect the true underlying diversity that existed \citep{Hunt2008,Eldredge1972,Gould1977a,VanBocxlaer2013}.
+With the rise of molecular systematics and techniques such as DNA barcoding, it has been increasingly recognized that some taxa can only be differentiated using molecular data \citep{Bickford2007,SchlickSteiner2007,Stuart2006,Pfenninger2007,Funk2012,Clare2011}.
+Nevertheless, the majority of extant taxa, and nearly all extinct taxa, are delimited solely via morphology.
+This disjunct is of great concern for studies of variation and diversity dynamics in deep time, where apparent morphological stasis may not reflect the true underlying diversity that existed \citep{Hunt2008,Eldredge1972,Gould1977a,VanBocxlaer2013}.
 
-Much work has been devoted to species delimitation via sequence difference \citep{Fujita2012,Yang2010b} which in turn has sparked interest in whether geometric morphometric anlaysis can captur similarly fine-scale variation that can be used for identifying cryptic species.
-Most such studies focus on morphometrics to discover differences between taxa that were idetified by other means \citep{Polly2007a,Demandt2009,Gaubert2005b,Gunduz2007,Polly2003,Zelditch2004} and automated taxon identification \citep{MacLeod2007}.
+Much work has been devoted to species delimitation via sequence difference \citep{Fujita2012,Yang2010b} which in turn has sparked interest in whether geometric morphometric anlaysis can capture similarly fine-scale variation that can be used for identifying cryptic species.
+Most such studies focus on morphometrics to discover differences between taxa that were identified by other means \citep{Polly2007a,Demandt2009,Gaubert2005b,Gunduz2007,Polly2003,Zelditch2004} and automated taxon identification \citep{MacLeod2007}.
 It would be very useful, though, if morphometrics approaches could reliably identify cryptic species without the use of other types of data.
-This would make the task of idenitfying and maintaining endangered or conserved groups much easier and could contribute to improved classifications of extinct taxa.
+This would make the task of identifying and maintaining endangered or conserved groups much easier and could contribute to improved classifications of extinct taxa.
 
@@
 \subsection{Background and system}
-Machine learning is in many respects just an extension of known statistical methodology \citep{Hastie2009} in which the emphasis is placed on inferring rules and properties of data in order to explain the underlying structure.
+Machine learning is in many respects just an extension of known statistical methodology \citep{Hastie2009} in which the emphasis is placed on high predictive accurate at the expense of interpretability of individual parameters.
 The basic statistical mechanics are supplemented by randomization, sorting, and partitioning algorithms and along with the maximization or minimization of summary statistics, in order to best estimate a general model for all data, both sampled and unsampled \citep{Hastie2009}.
@@
 
-Geometric morphometric approaches to idetifying differences in morphological variation between different classes, including cryptic species, mostly have used methods like linear discriminate analysis and canonical variates analysis \citep{Zelditch2004,Mitteroecker2011,Polly2007a,Polly2003,Gunduz2007,Gaubert2005b,Demandt2009,Sztencel-Jabonka2009,MitrovskiBogdanovic2013,Francoy2009}.
-These methods are comparatively straightforward ways of understanding the differences in morphology between classes given their similarity to familiar multivariate approaches like Principal Components Analysis.
+Geometric morphometric approaches to identifying differences in morphological variation between different classes, including cryptic species, mostly have used methods like linear discriminate analysis and canonical variates analysis \citep{Zelditch2004,Mitteroecker2011,Polly2007a,Polly2003,Gunduz2007,Gaubert2005b,Demandt2009,Sztencel-Jabonka2009,MitrovskiBogdanovic2013,Francoy2009}.
+These methods are comparatively straightforward ways of understanding the differences in morphology between classes given their similarity to familiar multivariate approaches like principal components analysis (PCA).
 They are benefit by producing results that can be easily visualized, which aids in the interpretation and presentation of data and results.
@@
 Here, we used multiple machine learning methods, both unsupervised and supervised, in order to compare different classification hypotheses.
-These methods provide different and unique advantages for understanding how to classify taxa, with what accuracy, and what these classifications are based on.
-While machine learning methods such as neural networks have been applied to studying shape variation \citep{Baylac2003,Dobigny2003,VandenBrink2011,MacLeod2007}, including in the context of automated taxon identification and classification of groups, but the number of cases remains limited.
+These methods provide different and unique advantages for understanding how to classify taxa, and with what accuracy.
+While machine learning methods such as neural networks have been applied to studying shape variation \citep{Baylac2003,Dobigny2003,VandenBrink2011,MacLeod2007}, including in the context of automated taxon identification and classification of groups, the number of cases remains limited.
 In the current study, we not only consider pure classification accuracy but also use a statistic of classification strength that reflects the rate at which taxa are both accurately and inaccurately classified.
@@
 \subsection{\textit{Emys marmorata}}
-We applied our machine learning methods to the problem whether distinct subspecies or cryptic species extist within the western pond turtle, \textit{Emys marmorata} (formerly \emph{Clemmys marmorata}; see \citealp{Feldman2002}).
+We analyzed the problem of whether there are distinct subspecies or cryptic species exist within the western pond turtle, \textit{Emys marmorata} (formerly \emph{Clemmys marmorata}; see \citealp{Feldman2002}).
 \textit{E.
@@
 marmorata palida} \citep{Seeliger1945}.
-She also recognized a central Californian intergrade zone between these subspecies, and she did not formally include the Baja California populations of \textit{E.
+She also recognized a central Californian intergrade zone between these subspecies, and did not include the Baja California populations of \textit{E.
 marmorata} in these groups, implying the existence of a third distinct but unnamed subspecies.
@@
 Most of these studies considered how local biotic and abiotic factors may contribute to differences in carapace length and found that size can vary greatly between different populations \citep{Lubcke2007,Germano2009,Germano2008}.
-There also has been interest in size-based sexual dimorphosim in \textit{E.
+There also has been interest in size-based sexual dimorphism in \textit{E.
 marmorata} \citep{Lubcke2007,Germano2009,Holland1992}, with males being on average larger than females based on total carapace length and other linear measurements.
@@
 marmorata} in three different areas of the species' range, and studied the relative effect of distance versus barriers in fostering morphological differentiation in the species.
-\citet{Holland1992} concluded that distance was a poor indicator of mophological differentiation, and the geographic features such as breaks between different drainage basis are probably more important barriers to reproduction.
+\citet{Holland1992} concluded that geographic distance was a poor indicator of mophological differentiation, and instead geographic features such as breaks between different drainage basis are probably more important barriers to reproduction.
 Additionally, \citet{Holland1992} found that morphological differences were observable as the magnitude of barriers and distance increased, but the variation required many variables to adequately capture, implying only very subtle morphological differentiation between putatively distinct populations.
 \citet{Holland1992} concluded that \textit{E.
-marmorata} is best classified as three distinct species: a northern species, souther species, and a Columbia basic species.
+marmorata} is best classified as three distinct species: a northern species, souther species, and a Columbia basin species.
 This classification is similar to that of \citet{Seeliger1945}, except elevated to the species level and without recognition of a distinct Baja species.
@@
 Analyses with nuclear DNA \citep{Spinks2010} with single-nucleotide polymorphism (SNP) data suggest a primarily north--south division in \textit{E.
-marmorata}, although the dataset differe in the location of this break point.
+marmorata}, although the dataset differs in the location of this break point.
 In these papers, Spinks and colleagues discussed the potential taxonomic implications of their results, with \citet{Spinks2014} going so far as to strongly advocate for the recognition of at least two species (\emph{E.
@@
 In this study, we attempt to estimate the best classification scheme of \textit{E.
-marmorata} based on variation in plastral (central shell) shape in order to determine whether this character is consistent with any of the past divisions based on other morphological features or molecular data.
+marmorata} based on variation in plastral (ventral shell) shape in order to determine whether this character is consistent with any of the past divisions based on other morphological features or molecular data.
 Because of unclear geographic boundaries between subgroups of \textit{E.
@@
 Because the specimens sampled to obtain the genetic data used to define the subclades were not available for study, all specimen classifications were based solely on the geographic information, not explicit assignment in previous studies.
-Because the exact barriers between different biogeographic regions are unknown and unclear, we represented each hypothesis with two different schemes; se we compared a total of six different schemes.
+Because the exact barriers between different biogeographic regions are unknown and unclear, we represented each hypothesis with two different schemes; we compared a total of six different schemes.
 The schemes differed based on where geographic boundaries were assigned.
@@
 
-Following previous work on plastron variation \citep{Angielczyk2007,Angielczyk2011,Angielczyk2013a}, we used TpsDig 2.04 \citep{Rohlf2005} to digitize 19 landmarks \ref{fig:plastra}).
+Following previous work on plastron variation \citep{Angielczyk2007,Angielczyk2011,Angielczyk2013a}, we used TpsDig 2.04 \citep{Rohlf2005} to digitize 19 landmarks (Fig.
+\ref{fig:plastra}).
 Seventeen of the landmarks are at the endpoints or intersection of the keratinous plastral scutes that cover the platron.
-Twelve of the landmarks were symmetrical across the axis of symmetry and, in order to prevent degrees of freedom and other concerns \citep{Klingenberg2002}, we reflected these landmarks across the axis of symmetry (i.e.
+Twelve of the landmarks were symmetrical across the axis of symmetry and, in order to prevent issues surrounding degrees of freedom and other similar concerns \citep{Klingenberg2002}, we reflected these landmarks across the axis of symmetry (i.e.
 midline) prior to analysis and used the average position of each symmetrical pair.
@@
 
-We divisively cluster the \(\rho\) dissimilarity matrix using partitioning around mediods clustering (PAM), a method similar to \textit{k}-means clustering except that instead of minimizing the sum of squared Euclidean distances between observations and centroids, the sum of squared dissimilarities between observations and mediods is minimized \citep{Kaufman1990}.
+We divisively clustered the shape dissimilarity matrix using partitioning around mediods clustering (PAM), a method similar to \textit{k}-means clustering except that instead of minimizing the sum of squared Euclidean distances between observations and centroids, the sum of squared dissimilarities between observations and mediods is minimized \citep{Kaufman1990}.
 Because the optimal number of clusters of shape configurations in the study was unknown, being possibly three, four, or some other value, we estimated clustering solutions in which the number of clusters varied between one and eight.
@@
 The other two methods, however, are not.
-Each of these three methods has a different interpretation and they reveal very different aspects of the data.
 In all cases, the optimal number of PCs used as predictors was chosen via maximum within-sample AUC value, explained below.
@@
 The results of this analysis produces a transformation matrix by which the original features can be transformed to reflect the best discrimination between the classes.
-In this study, we applied LDA to the eigenscores from a subset of the total number of PCs, ranging from two to NUMBER in increasing order of complexity.
+In this study, we applied LDA to the eigenscores from a subset of the total number of PCs, ranging from two to 6 in increasing order of complexity.
 In total, this produced nine different LDA scaling matrices.
@@
 Similar to the odds ratios calculated from the coefficients of a logistic regression, the relative risk of a classification can be determined from the coefficients of the model.
-Similar to the LDA, we determined the optimal number of PCs as predictors by comparing within-sample AUC values across multiple models.
 
 Random forest models are an extension of classification and regression trees (CART) \citep{Breiman1984,Breiman2001}.
-The goal of CARTs are to use a series of different features to estimate the class of an observation (specimen).
+The goal of CARTs are to use a series of different features (i.e.
+predictors) to estimate the class of an observation.
 In top-down induction of decision trees for each member of a given set of predictor variables, attribute value tests are used to estimate the differences between classes.
@@
 In addition to classifying the observations, this procedure allows for the features to be ranked in order of importance.
-This is a generally important property that is useful for many other studies in which the goal is to describe and model the differences between classes and the relative importance of different predictors.
+This is a generally useful property for studies in which the goal is to describe and model the differences between classes and the relative importance of different predictors.
 
@@
 
-To determine if there were systematic differences between the correctly and incorrectly classified samples, we compared the multivariate centroids of the correct and incorrect groups to what would be expected by random.
-The group labels were permuted 1000 times and the difference between the new centroids was calculated.
+To determine if there were systematic differences between the correctly and incorrectly classified samples, we used a permutation test to estimate if the dissimilarity between the correctly and incorrectly classified individuals were significantly different from random.
+The group labels were permuted 1000 times and the distance between the new centroids was calculated.
 The number of permutations less than the empirical difference divided by 1000 gives a \textit{p}-value for the test.
@@
 \subsection{Unsupervised learning}
-Comparison of gap statistic values from PAM clustering show that the best number of clusters is most likely one (Fig.
+Comparison of gap statistic values from PAM clustering show that the optimal, minimal number of clusters is most likely one (Fig.
 \ref{fig:gap}).
@@
 \ref{fig:gap_map}).
-Because of this, we assert that this means that there is no means of naturally partitioning plastron shape into distinct subgroups.
+Because of this, we assert that this means that there is no means of naturally partitioning plastron shape into distinct subgroups with out reference to external information.
 % PAM stuff
@@
 AUC--based model selection revealed some important patterns of variation and congruence between the classification schemes and the actual data.
-Generally, as many PCs as possible were included as predictors for the best models of each of the classification schemes (Fig.
-\ref{fig:sel}).
+Generally, the best performing models tended to include as many PCs as possible \ref{fig:sel}).
 Note that the best random forest models were determined via recursive feature selection, so PCs were not included in order of percent variance explained.
-For both the LDA and multinomial logistic regression models, increasing model complexity increased cumulative percent variation necessary to best model the differing classification schemes (Fig.
-\ref{fig:sel}).
-That almost all models were as complex as possible indicates that the differences between the different groups within each classification scheme are very small.
+That almost all LDA and multinomial logistic regression models were as complex as possible indicates that the differences between the different groups within each classification scheme are very small.
 
 As part of fitting a random forest model, a ranking of variable importance also is determined.
-Interestingly, the order of variable importance is not the same as the order of decreasing explained variance per principal component (Fig.
-\ref{fig:var_imp}).
-This means that the principal components that best describe the differences between the various classes are not aligned with the principal components which describe the largest amount of variance.
-Another way of phrasing this is that the variance describing the differences between the classes does not align with the major axes of variance (i.e.
+Interestingly, the order of variable importance is not the same as the order of the PCs (\ref{fig:var_imp}).
+This means that the variance describing the differences between the classes does not align with the major axes of variance (i.e.
 the PCs).
-This result would be the case if variation between classes was extremely fine grained and not a part of the principal form or function of the plastron, which makes sense given that the plastron is involved in both protection and hydrodynamics and not mate choice \citep{Germano2009,Holland1992,Lubcke2007,Rivera2008}.
-Moreover, this result is congruous with the results of the AUC--based model selection for the multinomial logistic regression and LDA models.
+This result would be the case if variation between classes was extremely fine grained and not a part of the principal form or function of the plastron, which makes sense given that the plastron is involved in both protection and hydrodynamics and not necessarily mate choice \citep{Germano2009,Holland1992,Lubcke2007,Rivera2008}.
+Moreover, this result is congruous with the results from the AUC--based model selection for the multinomial logistic regression and LDA models.
 
-Observed AUC values for all of the optimal models are not exceptionally high; values near 0.5 indicate that a model is no better than completely random assignment (Fig.
-\ref{fig:sel}).
-This means that in all but a few cases the different proposed classification schemes are generally poor descriptors of the observed variation.
+Observed AUC values for all of the optimal models are not exceptionally high (\ref{fig:sel}).
+In most cases the different proposed classification schemes are generally poor descriptors of the observed variation.
 It appears that the data set is overwhelmed by noise, making any accurate classifications difficult at best.
-This observation is cemented with the generalizations of the models to the testing data set.
+This observation is cemented with the generalizations of the models to the testing data set (\ref{fig:gen_hist}).
 
@@
 
-The results of this study indicate that there is no clear grouping of plastra shapes in \textit{E. marmorata}.
+The results of this study indicate that there is no clear grouping of \textit{E. marmorata} based on plastron shape.
 
@@
 The classification schemes used in the supervised learning models correspond, loosely, to unsupervised learning solutions with multiple groups.
-Because unsupervised learning solutions with multiple groups are poor descriptors of the observed variation, it is important to see this reinforced by the supervised learning results.
+Because unsupervised learning solutions with multiple groups are poor descriptors of the observed variation, it is important to see this generally supported by the supervised learning results.
 
@@
 Both the low AUC values (\(< 0.9\)) and the significant difference between the correctly and incorrectly classified observations support the conclusion that none of the hypothesized classification schemes are good descriptors of the observed plastral variation.
+
+INSERT MORE TEXT HERE.
 

### 2015-08-06 R/make_tables.r
renamed all the classification schemes

@@ -64,7 +64,9 @@ diff.table <- function(boot.mod, best, dif) {
   out <- cbind(scheme = names(means), compare = unlist(ccoo))
   xtable(out)
 }
-
+names(mm) <- c('Morph 1', 'Morph 2', 'Mito 1', 'Nuclear', 'Mito 2', 'Mito 3')
+names(rr) <- c('Morph 1', 'Morph 2', 'Mito 1', 'Nuclear', 'Mito 2', 'Mito 3')
+names(ll) <- c('Morph 1', 'Morph 2', 'Mito 1', 'Nuclear', 'Mito 2', 'Mito 3')
 mmdif.tab <- diff.table(mm, best.mm, mm.dif)
 label(mmdif.tab) <- 'mmdif'
 print.xtable(x = mmdif.tab,
@@ -97,6 +99,8 @@ sig.table <- function(test, groups) {
   sig.tab <- sig.tab[, c(4, 3, 1, 2)]
   sig.tab <- xtable(sig.tab)
 }
+groups <- list('Morph 1', 'Morph 2', 'Molec 1', 
+               'Nuclear', 'Molec 2', 'Molec 3')
 rfmiss.tab <- sig.table(rf.test, groups)
 label(rfmiss.tab) <- 'rfmiss'
 print.xtable(x = rfmiss.tab,

### 2015-08-10 doc/draft_turtle.tex
ken + jim comments. need to update citations

--- 2014-12-22 doc/draft_turtle.tex
@@
 \usepackage{microtype, parskip}
-\usepackage[numbers,sort&compress]{natbib}
+\usepackage{natbib}
 \usepackage{lineno}
@@
 
-\title{How cryptic is cryptic diversity? Machine learning approaches to classifying morphological variation in \textit{Emys marmorata} (Testudinoidea, Emydidae).}
+\title{How cryptic is cryptic diversity? Machine learning approaches to classifying morphological variation in the Pacific Pond Turtle (\textit{Emys marmorata})}
 \author[1]{Peter D Smits}%\thanks{psmits@uchicago.edu}}
@@
 \begin{abstract}
+  ABSTRACT NEEDS TO BE WRITTEN
 \end{abstract}
@@
 \section{Introduction}
+
+% cryptic diversity
+%   most species are still deliminated solely based on morphology
+Molecular systematics has repeatedly demonstrated the existence of cryptic species that can only be diagnosed using genetic data \citep{Stuart2006,Bickford2007,SchlickSteiner2007,Pfenninger2007,Clare2011,Funk2012}.
+In attempts to streamline the documentation of biodiversity, several methods of species delimitation that rely almost entirely on genetic data have recently been proposed \citep{Yang2010b}.
+Although strong caveats on the utility of these methods have been raised \citep{Bauer2000,Carstens2013}, they are nevertheless being used to name species \citep{Leache2010,Spinks2014}.
+
+The majority of extant taxa, and almost all extinct taxa, are delimited by morphology alone.
+This disjunction complicates interpretations of variation and diversity in deep time, as apparent morphological stasis may not reflect the true underlying diversity \citep{Eldredge1972,Gould1977a,Hunt2008,VanBocxlaer2013}.
+Similarly, for many museum specimens of extant taxa (e.g.
+those preserved in formalin), it is difficult to acquire the genetic data needed for non-morphological species delimitation methods.
+
+These considerations have sparked interest in whether geometric morphometric analyses can capture fine-scale variation that can be used for identifying cryptic species.
+Most such studies focus on using morphometrics to discover differences between taxa that were identified by other means \citep{Polly2003,Zelditch2004,Gaubert2005b,Gunduz2007,Polly2007a,Demandt2009}.
+Additionally, there has been a fair amount of work on automated taxon identification and classification of taxa into groups \citep{Baylac2003,Dobigny2003,MacLeod2007,VandenBrink2011}.
+In cases where genetic data are ambiguous or lacking for many samples, morphometric approaches could help identify cryptic species.
+This would make the task of identifying and maintaining endangered or conserved groups much easier and could contribute to improved classifications of extinct taxa and populations.
+
+Here, we investigate the morphometric identification of cryptic species using machine-learning approaches.
+In particular, we ask whether it is possible to determine which amongst a set of classification hypotheses is best and examine the implications of the results for a recently proposed set of cryptic turtle species.
+
+% past work on automatic taxon identification and older approaches to classifying taxa
+% why use machine learning methods
+\subsection{Background and study system}
+Machine learning is an extension of known statistical methodology \citep{Hastie2009} that emphasizes high predictive accuracy and generality at the expense of the interpretability of individual parameters.
+The basic statistical mechanics are supplemented by randomization, sorting, and partitioning algorithms, along with the maximization or minimization of summary statistics, in order to best estimate a general model for all data, both sampled and unsampled \citep{Hastie2009}.
+Machine learning approaches have found use in medical research, epidemiology, economics and automated image identification such as handwritten zip codes \citep{Hastie2009}.
+The two major classes of machine learning methods are unsupervised and supervised learning.
+Unsupervised learning methods are used with unlabeled data where the underlying structure is estimated, and they are analogous to clustering and density estimation methods \citep{Kaufman1990}.
+Supervised learning methods are used with labeled data where the final output of data is known and the rules for going from input to output are inferred.
+These are analogous to classification and regression models \citep{Breiman1984}.
+Out application of the alternative approaches used in this study illustrates only a sampling of the various previously-derived methods for clustering observations and fitting classification models.
+
+Geometric morphometric approaches to identifying differences in morphological variation between different classes, including cryptic species, mostly have used methods like linear discriminate analysis and canonical variates analysis \citep{Polly2003,Zelditch2004,Gaubert2005b,Gunduz2007,Polly2007a,Francoy2009,Sztencel-Jabonka2009,MitrovskiBogdanovic2013}.
+Because of their similarity to multivariate approaches like principle components analysis (PCA), these methods are comparatively straightforward ways of understanding the differences in morphology between classes.
+They also benefit from producing results that can be easily visualized, which aids in the interpretation and presentation of data and results.
+Most previous morphometric studies did not assess which amongst a set of alternative classification hypotheses was optimal.
+For example, studies such as those of \citet{Caumul2005a} and \citet{Polly2007a} focused on comparing different aspects of morphology and their fidelity to a classification scheme instead of comparing the fidelity of one aspect of morphology to multiple classification schemes.
+In this context, the study of \citet{Cardini2009a} is noteworthy because they compared morphological variation in marmots at the population, regional, and species level and determined the fidelity of shape to divisions at each of these levels.
+
+Here, we used multiple machine learning methods, both unsupervised and supervised, to compare different classification hypotheses.
+These methods provide different and unique advantages for understanding how to classify taxa, as well as the accuracy of the resulting classifications.
+Although machine learning methods such as neural networks have been applied to studying shape variation \citep{Baylac2003,Dobigny2003,MacLeod2007,VandenBrink2011}, including in the context of automated taxon identification and classification of groups, the number of cases remains limited.
+In the current study, we not only consider pure classification accuracy but also use a statistic of classification strength that reflects the rate at which taxa are both accurately and inaccurately classified.
+
+We analyzed the problem of whether there are distinct subspecies or cryptic species within the western pond turtle, \textit{Emys marmorata} BAIRD CITATIONS (formerly \emph{Clemmys marmorata}; see \citealp{Feldman2002}).
+\textit{Emys marmorata} is distributed from northern Washington State, USA to Baja California, Mexico.
+Traditionally, \textit{E.
+marmorata} was classified into two named subspecies: the northern \textit{E.
+marmorata marmorata} and the southern \textit{Emys marmorata pallida} \citep{Seeliger1945}, with a central Californian intergrade zone in between.
+\textit{Emys marmorata marmorata} is differentiated from \textit{E.
+marmorata pallida} by the presence of a pair of triangular inguinal scales and darker neck markings.
+The triangular inguinal plates can sometimes be present in \textit{E.
+marmorata pallida} though they are considerably smaller.
+\citet{Seeliger1945} did not formally include the Baja California populations of \textit{E.
+marmorata} in either taxon, implying the existence of a third distinct but unnamed subspecies.
+
+Previous work on morphological variation in \textit{E.
+marmorata} has focused primarily on differentiation between populations over a portion of the species' total range \citep{Lubcke2007,Germano2008,Germano2009,Bury2010}; comparatively few studies have included specimens from across the entire range \citep{Holland1992}.
+Most of these studies considered how local biotic and abiotic factors may contribute to differences in carapace length and found that size can vary greatly between different populations \citep{Lubcke2007,Germano2008,Germano2009}.
+There also has been interest in size-based sexual dimorphism in \textit{E.
+marmorata} \citep{Holland1992,Lubcke2007,Germano2009}, with males being on average larger than females based on total carapace length and other linear measurements.
+However, the quality of size as a classifier of sex can vary greatly between populations \citep{Holland1992} because of the amount of size differences among populations \citep{Lubcke2007,Germano2009}.
+However, the effect of sexual dimorphism on shape, \textit{sensu} \citet{Kendall1977a}, has not been assessed \citep{Holland1992,Lubcke2007,Germano2008}.
+
+Of particular importance in the context of cryptic diversity in \textit{E.
+marmorata} is the morphometric analysis of carapace shape carried out by \citet{Holland1992}, who compared populations of \textit{E.
+marmorata} from three areas of its range.
+This study concluded that geographic distance was a poor indicator of mophological differentiation, and instead hypothesized that geographic features such as breaks between different drainage basis are probably more important barriers to dispersal and interbreeding.
+Additionally, \citep{Holland1992} suggested that morphological differences were more pronounced as the magnitude of barriers and distance increased, but this variation required many variables to adequately capture, implying only very subtle morphological differentiation between putatively distinct populations.
+Finally, Holland concluded that \textit{E.
+marmorata} is best classified as three distinct species: a northern species, souther species, and a Columbia Basin species.
+This classification is similar to that of \citet{Seeliger1945}, except elevated to the species level and without recognition of a distinct Baja species.
+
+More recently, the phylogeography of \textit{E.
+marmorata} and the possibility of cryptic diversity was investigated using molecular data \citep{Spinks2005,Spinks2010,Spinks2014}.
+Based on mitochondrial DNA, \citet{Spinks2005} recognized four subclades within \textit{E.
+marmorata}, a northern clade, a San Joaquin Valley clade, a Santa Barbara clade, and a southern clade.
+Analyses with nuclear DNA \citep{Spinks2010} with single-nucleotide polymorphism (SNP) data suggest a primarily north--south division in \textit{E.
+marmorata}, although the dataset differed from that of \citet{Spinks2005} in the location of this break point.
+These studies discussed the potential taxonomic implications of their results, with \citet{Spinks2014} going so far as to strongly advocate for the recognition of at least two species (\emph{E.
+marmorata} and \emph{E.
+pallida}), and a possible third based on populations in Baja California.
+However, they did not discuss in detail the morphological characters that would help to diagnose these species beyond those specified by \citet{Seeliger1945}.
+Given that these characters are somewhat variable within the proposed species, and that \citet{Holland1992} described shell shape variation that might be consistent with this taxonomy, a geometric morphometric analysis of shell shape might provide a reliable way to diagnose groups (whether species or subspecies) within \textit{E.
+marmorata}.
+
+In this study, we attempt to estimate the best classification scheme of \textit{E.
+marmorata} based on variation in plastron (ventral shell) shape in order to determine whether this character is consistent with any of the past divisions based on other morphological features or molecular data.
+We are particularly interested in whether any aspect of plastron shape can be used to reliably diagnose \citet{Spinks2014} proposed species, and if so, the nature of that shape variation.
+
+Because of unclear geographic boundaries between subgroups of \textit{E.
+marmorata}, we compare multiple hypotheses of morphologically-- and molecularly--based classification.
+We hypothesize that if morphological variation corresponds to class assignment, it should be possible to determine the best classification hypothesis of \textit{E.
+marmorata} from amongst multiple candidate hypotheses.
+However, if morphological variation does not correspond to any of the standing hypothesis, then supervised learning model generalization performance will be poor.
 
@@
 marmorata} museum specimens.
-These specimens are a subset of those included in \citet{Angielczyk2007}, \citet{Angielczyk2011}, and \citet{Angielczyk2013a} and represents adult individuals.
+These specimens are a subset of those included in \citet{Angielczyk2007}, \citet{Angielczyk2011}, and \citet{Angielczyk2013a}, essentially representing the adult individuals included in those papers.
 We chose to focus on adults because significant changes in plastron shape occur over the course of ontogeny in \textit{E.
 mamorata} and other emydines \citep{Angielczyk2013a}.
+\uppercase{Peter note -- how was something classified as an adult?}
+
 We assigned a classification to each specimen for the different binning schemes based on geographic occurrence data recorded in museum collection archives.
-When precise latitude and longitude information was not available we estimated it from whatever locality information was present.
-Because the specimens sampled to obtain the genetic data used to define the subclades were not available for study, all specimen classifications were based solely on the geographic information, not explicit assignment in previous studies.
-Because the exact barriers between different biogeographic regions are unknown and unclear, we represented each hypothesis with two different schemes; se we compared a total of six different schemes.
-% How were adults chosen?
-Good to note how schemes differeed.
+When precise latitude and longitude information were not available we estimated them from locality information.
+Because \citet{Spinks2005}, \citet{Spinks2010}, and \citet{Spinks20140} did not use vouchered specimens we were not able to example the sample individuals.
+Therefore our specimen classifications were based solely on the geographic information, not explicit assignment in previous studies.
+Because the exact barriers between different biogeographic regions are unknown and unclear, we represented each hypothesis with two schemes for a total of six different schemes.
+The schemes differed based on where geographic boundaries were assigned.
+This changes the classification of certain individuals near the boundaries between groups, providing a test of the robustness of the classification schemes.
 
-Following previous work on plastron variation \citep{Angielczyk2007,Angielczyk2011,Angielczyk2013a}, we used TpsDig 2.04 \citep{Rohlf2005} to digitize 19 landmarks \ref{fig:plastra}).
+Following previous work on plastron variation \citep{Angielczyk2007,Angielczyk2011,Angielczyk2013a}, we used TpsDig 2.04 \citep{Rohlf2005} to digitize 19 landmarks (Fig.
+\ref{fig:plastra}).
 Seventeen of the landmarks are at the endpoints or intersection of the keratinous plastral scutes that cover the platron.
-Twelve of the landmarks were symmetrical across the axis of symmetry and, in order to prevent degrees of freedom and other concerns \citep{Klingenberg2002}, we reflected these landmarks across the axis of symmetry (i.e.
+Twelve of the landmarks were symmetrical across the axis of symmetry and, in order to prevent issues related to degrees of freedom and other similar concerns \citep{Klingenberg2002}, we reflected these landmarks across the axis of symmetry (i.e.
 midline) prior to analysis and used the average position of each symmetrical pair.
@@
 
-We divisively cluster the \(\rho\) dissimilarity matrix using partitioning around mediods clustering (PAM), a method similar to \textit{k}-means clustering except that instead of minimizing the sum of squared Euclidean distances between observations and centroids, the sum of squared dissimilarities between observations and mediods is minimized \citep{Kaufman1990}.
+We divisively clustered the shape dissimilarity matrix using partitioning around mediods clustering (PAM), a method similar to \textit{k}-means clustering except that instead of minimizing the sum of squared Euclidean distances between observations and centroids, the sum of squared dissimilarities between observations and mediods is minimized \citep{Kaufman1990}.
 Because the optimal number of clusters of shape configurations in the study was unknown, being possibly three, four, or some other value, we estimated clustering solutions in which the number of clusters varied between one and eight.
-We compated clustering solutions using the gap statistic, which is a measure of goodness of clustering \citep{Tibshirani2001a}.
+We compared clustering solutions using the gap statistic, which is a measure of goodness of clustering \citep{Tibshirani2001a}.
 
@@
 Similar to the odds ratios calculated from the coefficients of a logistic regression, the relative risk of a classification can be determined from the coefficients of the model.
-Similar to the LDA, we determined the optimal number of PCs as predictors by comparing within-sample AUC values across multiple models.
 
 Random forest models are an extension of classification and regression trees (CART) \citep{Breiman1984,Breiman2001}.
-The goal of CARTs are to use a series of different features to estimate the class of an observation (specimen).
+The goal of CARTs is to use a series of different features (i.e.
+predictors) to estimate the class of an observation.
 In top-down induction of decision trees for each member of a given set of predictor variables, attribute value tests are used to estimate the differences between classes.
@@
 In addition to classifying the observations, this procedure allows for the features to be ranked in order of importance.
-This is a generally important property that is useful for many other studies in which the goal is to describe and model the differences between classes and the relative importance of different predictors.
+This is a generally useful property for studies in which the goal is to describe and model the differences between classes and the relative importance of different predictors.
 
 In this analysis, we used 1000 subtrees to estimate the random forest model parameters.
-We estimated the best set of predictors necessary for each classification scheme was estimated using a recursive feature selection algorithm, and we chose the optimal number of PCs to include based on the AUC of the model.
-Following the backwards selection algorithm implemented in \texttt{caret} \citep{Kuhn2013}, the maximum number of features were included in the initial model, their importance ranked, and the AUC of the model calculated.
+We estimated the best set of predictors necessary for each classification scheme using a recursive feature selection algorithm, and we chose the optimal number of PCs to include based on the AUC of the model.
+Following the backwards selection algorithm implemented in \texttt{caret} \citep{KuhnMAN2013}, the maximum number of features were included in the initial model, their importance ranked, and the AUC of the model calculated.
 The lowest ranked feature was then removed, and the AUC of the model recalculated.
 This was repeated until only one feature, remained.
-Because PCs were kept in order of importance and not in relation to the amount of variance each PC described, these means that the PCs are not included in the order of ascending eigenvalue.
+Because PCs were kept in order of importance and not in relation to the amount of variance each PC described, the PCs are not included in the order of ascending eigenvalue.
 
@@
 AUC is an estimate of the relationship between the false positive and true positive rates, as opposed to just the true positive rate (accuracy).
-This relationship is especially useful in cases where misclassification needs to be minimized just as much as accurate classification, as in this study.
+This relationship is especially useful in cases such as this study where misclassification needs to be minimized just as much as an accurate classification must be obtained.
 AUC ranges between 0.5 and 1, with 0.5 indicating classification no better than random and 1 indicating perfect classification \citep{Hastie2009}.
 
-The standard AUC calculation is defined for binary classifications, however in this application there are multiple categories.
+The standard AUC calculation is defined for binary classifications, however in our application there are multiple categories.
 The alternative calculation that we used follows an all-against-one strategy where the individual AUC values for each class versus all others are averaged to produce a multiclass AUC \citep{Hand2001}.
@@
 Within-sample performance is inherently biased upwards, so model evaluation requires overcoming this bias.
-With very large sample sizes, as in this study, part of the sample can be used as the ``training set'' and the remainder acts as the ``testing set.'' The former is used for fitting the model where as the later is used for measuring model performance, and this process is called model generalization.
+With very large sample sizes, as in this study, part of the sample can be used as the ``training set'' and the remainder acts as the ``testing set.'' The former is used for fitting the model where as the later is used for measuring model performance, a process called model generalization.
 In this analysis, we used 75\% of samples as the training set while the remaining 25\% were used as the testing set.
@@
 Testing and training sets are determined completely at random within each class and with respect to shape.
-Results were not effected by changes in testing or training set assignment.
+Results were not effected by the individual specimen class assignments to the testing or training sets.
 
-To determine if there were systematic differences between the correctly and incorrectly classified samples, we compared the multivariate centroids of the correct and incorrect groups to what would be expected by random.
-The group labels were permuted 1000 times and the difference between the new centroids was calculated.
+To determine if there were systematic differences in plastron shape between the correctly and incorrectly classified samples, we used a permutation test to estimate if the dissimilarity between the correctly and incorrectly classified individuals was significantly greater than random.
+The group labels were permuted 1000 times and the distance between the new centroids was calculated.
 The number of permutations less than the empirical difference divided by 1000 gives a \textit{p}-value for the test.
@@
 \ref{fig:gap_map}).
-Because of this, we assert that this means that there is no means of naturally partitioning plastron shape into distinct subgroups.
+Because of this, we assert that there is no means of naturally partitioning plastron shape into distinct subgroups with out reference to external information.
 % PAM stuff
@@
 AUC--based model selection revealed some important patterns of variation and congruence between the classification schemes and the actual data.
-Generally, as many PCs as possible were included as predictors for the best models of each of the classification schemes (Fig.
-\ref{fig:sel}).
+Generally, the best performing models tended to include as many PCs as possible \ref{fig:sel}).
 Note that the best random forest models were determined via recursive feature selection, so PCs were not included in order of percent variance explained.
-For both the LDA and multinomial logistic regression models, increasing model complexity increased cumulative percent variation necessary to best model the differing classification schemes (Fig.
-\ref{fig:sel}).
-That almost all models were as complex as possible indicates that the differences between the different groups within each classification scheme are very small.
+That almost all LDA and multinomial logistic regression models were as complex as possible indicates that the differences between the different groups within each classification scheme are very small.
 
 As part of fitting a random forest model, a ranking of variable importance also is determined.
-Interestingly, the order of variable importance is not the same as the order of decreasing explained variance per principal component (Fig.
-\ref{fig:var_imp}).
-This means that the principal components that best describe the differences between the various classes are not aligned with the principal components which describe the largest amount of variance.
-Another way of phrasing this is that the variance describing the differences between the classes does not align with the major axes of variance (i.e.
+Interestingly, the order of variable importance is not the same as the order of the PCs (\ref{fig:var_imp}).
+This means that the variance describing the differences between the classes does not align with the major axes of variance (i.e.
 the PCs).
-This result would be the case if variation between classes was extremely fine grained and not a part of the principal form or function of the plastron, which makes sense given that the plastron is involved in both protection and hydrodynamics and not mate choice \citep{Germano2009,Holland1992,Lubcke2007,Rivera2008}.
-Moreover, this result is congruous with the results of the AUC--based model selection for the multinomial logistic regression and LDA models.
+Such a result is expected if variation between classes was extremely fine grained and not a part of the principal form or function of the plastron.
+Moreover, this result is congruous with the results from the AUC--based model selection for the multinomial logistic regression and LDA models.
 
-Observed AUC values for all of the optimal models are not exceptionally high; values near 0.5 indicate that a model is no better than completely random assignment (Fig.
-\ref{fig:sel}).
-This means that in all but a few cases the different proposed classification schemes are generally poor descriptors of the observed variation.
+Observed AUC values for all of the optimal models are not exceptionally high (\ref{fig:sel}).
+In most cases the different proposed classification schemes are generally poor descriptors of the observed variation.
 It appears that the data set is overwhelmed by noise, making any accurate classifications difficult at best.
-This observation is cemented with the generalizations of the models to the testing data set.
+This observation is cemented with the generalizations of the models to the testing data set (\ref{fig:gen_hist}).
 
@@
 
-The results of this study indicate that there is no clear grouping of plastra shapes in \textit{E. marmorata}.
+The results of this study indicate that there is no clear grouping of \textit{E. marmorata} based on plastron shape.
 
-The unsupervised learning results which indicate only a single group of observations being optimal is congruous with the results from the generalizations of the supervised learning models.
+The unsupervised learning results indicate only a single group of observations being optimal, which is congruous with the results from the generalizations of the supervised learning models.
 The classification schemes used in the supervised learning models correspond, loosely, to unsupervised learning solutions with multiple groups.
-Because unsupervised learning solutions with multiple groups are poor descriptors of the observed variation, it is important to see this reinforced by the supervised learning results.
+Because unsupervised learning solutions with multiple groups are poor descriptors of the observed variation, it is important to see this generally supported by the supervised learning results.
 
-The results from fitting the various supervised learning models for each of the classification scheme generally shows that no one scheme is ``best.'' A possible explanation for this that the genetic divergence associated with (sub)speciation is either not based on plastron morphology or local selective pressures due to hydrological regime overwhelming any possible morphological divergence.
+The results from fitting the various supervised learning models to each of the classification schemes generally shows that no one scheme is ``best.'' Possible explanations include that the genetic differentiation is not associated with plastron shape variation and/or that local selective pressures (e.g.
+from hydrological regime) overwhelm morphological differentiation.
+This makes sense given that the plastron is involved in both protection and hydrodynamics, and not necessary mate choice CITATIONS and that shell shape in \textit{E.
+marmorata} is known to vary among populations inhabiting water bodies with different flow regimes CITATIONS.
+Plastron shape does not seem to preserve a strong phylogenetic signal at the interspecific level in emydine turtles \citep{Angielczyk2011}, and out current results suggest that this may be the case for phylogeographic signal within emydine species as well.
+Stated another way, plastron shape may be a poor morphological marker for phylogenetic and phylogeographic studies, despite the ease with which its shape can be quantified.
+A final possibility (explored below) is that the proposed classification schemes themselves do not represent significant evolutionary lineages.
 
 Both the low AUC values (\(< 0.9\)) and the significant difference between the correctly and incorrectly classified observations support the conclusion that none of the hypothesized classification schemes are good descriptors of the observed plastral variation.
+
+\subsection{Is there more than one species of Western Pond Turtle?}
+
+The lack of morphological support for the distinctiveness of \textit{E.
+pallida} does not, on its own, preclude the recognition of this taxon.
+However, this apparent lack of congruence does prompt a reexamination of the methods and concepts that led to that taxonomic revision.
+In other words, before we can assess the significance of the morphological non-diagnosablity, it is essential to evaluate the methods and concepts that led to taxonomic revision.
+
+\citet{Spinks2014} elevated \textit{E.
+pallida} based on a Bayesian species delimitation analysis of SNP data using BPP \citep{Yang2010b}.
+However, \citet{Spinks2014} did not heed the caveats about species delimitation methods raised by \citet{Carstens2013}.
+In addition to specifically addressing the shortcomings of validation methods such as BPP that rely on guide trees and so “should be interpreted with caution,” \citet{Carstens2013} also strongly emphasize that ``Inferences regarding species boundaries based on genetic data alone are likely inadequate, and species delimitation should be conducted with consideration of the life history, geographical distribution, morphology and behaviour (where applicable) of the focal system\dots'' These caveats evoke the development of the Unified Species Concept \citep{Dayrat2005a,DeQueiroz2007b}, Integrative Taxonomy CITATIONS Padial 2010, and other pluralist approaches to species delimitation.
+None of these considerations were brought to bear on the \textit{E.
+marmorata} system until now, and in doing so we find the proposal that \textit{E.
+pallida} is a distinct species to be lacking.
+
+In particular, the natural history and geographical distribution of \textit{E.
+marmorata} and \textit{E.
+pallida} make the recognition of these taxa implausible.
+The data from \citet{Spinks2014} show extensive introgression and admixture in Central California, which is expected because there are no significant barriers to gene flow in this region.
+Combined with the well-demonstrated ability for testudinoid turtles, including emydids and even \textit{Emys}, to hybridize (e.g.
+\citealt{Buskirk2005,Spinks2009,Parham2013}) it is hard to imagine how \textit{E.
+marmorata} and \textit{E.
+pallida} could maintain their integrity in the face of such admixture.
+Because the geography, natural history, and demonstrated genetic admixture of \textit{E.
+marmorata} conflict with the recognition of \textit{E.
+pallida}, we hypothesize that out inability to classify the morphological data by proposed species is because \textit{E.
+pallida} is not a distinct species.
+We agree with \citet{Carstens2013} that ``the inferences drawn from species delimitation studies should be conservative, for in most contexts it is better to fail to delimit species than it is to falsely delimit entities that do not represent actual evolutionary lineages.''
+
+We fully agree with \citet{Spinks2014} that \textit{E.
+marmorata} (\textit{sensu lato}) is a species deserving of strong conservation efforts, we do not wish to trivialize this need.
+Moreover, the genetic diversity uncovered by the analysis of \citet{Spinks2014} should be explicitly accounted for in any conservation plan.
+Given the apparent lack of morphological distinction, however, we consider that this diversity should be considered Evolutionary Significant Units or Distinct Population Segments instead of distinct species.
+
+Finally, it is important to note that the data and analyses we present do not let us definitively say whether the apparent lack of morphological divergence within \textit{E.
+marmorata} truly reflects the presence of a single species, or if it is an artifact of plastron shape being a poor morphological marker for phylogenetic and phylogeographic divergences.
+Tests of both our preferred conclusion (\textit{E.
+marmorata} as a single species) and that \citet{Spinks2014} should include additional tests of species delimitation using different methods and search for corroborating evidence as suggested by \citet{Carstens2013}.
+From a morphological standpoint, support for the validity of ``\textit{E.
+Pallida}'' may come from other aspects of morphology, such as carapace shape or other features that might be more informative.
+Lifewise, further investigation of the phylogeographic utility of plastron shape in other turtle species will help to clarify whether the lack of differentiation seen in \textit{E.
+marmoarata} is typical or an unusual case.
+
+\section*{Acknowledgements}
+Data collection for this project was supported in part by NSF DBI-0306158 (to KDA).
+G.
+Miller assisted with data collection and her participation in this research was supported by NSF REU DBI-0353797 (to R.
+Mooi of CAS).
+For access to emydine specimens, we thank: J.
+Vindum and R.
+Drewes (CAS); A.
+Resetar (FMNH); R.
+Feeney (LACM); C.
+Austin (LSUMNS); S.
+Sweet (MSE); J.McGuire and C.
+Conroy (MVZ); A.
+Wynn (NMNH); P.
+Collins (SBMNH); B.
+Hollingsworth (SDMNH); P.
+Holroyd (UCMP).
+We are grateful for S.
+Sweet for field assistance and the California Department of Fish and Game for permits.
 

### 2015-08-26 doc/draft_turtle.tex
added short text about secondary dataset

@@
 When precise latitude and longitude information were not available we estimated them from locality information.
-Because \citet{Spinks2005}, \citet{Spinks2010}, and \citet{Spinks20140} did not use vouchered specimens we were not able to example the sample individuals.
+Because \citet{Spinks2005}, \citet{Spinks2010}, and \citet{Spinks2014} did not use vouchered specimens we were not able to example the sample individuals.
 Therefore our specimen classifications were based solely on the geographic information, not explicit assignment in previous studies.
@@
 
+\subsection{Comparison with clear-cut example}
+In addition to the \textit{E.
+marmorata/pallida} case, we also analyzed a set of seven morphologically distinct species in order to test if the approaches used here actually work.
+The seven species analyzed were \textit{Emys blandigii}, \textit{Terrapene coahuila}, \textit{Clemmys guttata}, \textit{Glyptemys insculpta}, \textit{Glyptemys muhlenbergii}, \textit{Emys orbicularis}, and \textit{Terrapene ornata} with a total of XXX specimens analyzed in total.
+Just as in the \textit{E.
+marmorata/pallida} example, we analyzed plastron shape variation.
+The support for a classification scheme with seven categories was evaluated using random forest, linear discriminate analysis, and multinomial logistic regression where the number of PCs used as predictors ranged from 2 to 11.
+Data was split into training and testing sets and model performance for both sets was evaluated using the AUC metric, just as above.
+
 \section{Results}
@@
 This may be because differences in plastron shape do not align with the any of the hypothesized classification schemes.
+
+\section{Comparison with clear-cut example}
+In contrast with the above results, the analysis of the clearly distinct species SUPPLEMENT demonstrates almost perfect classification of the both the in-sample and out-of-sample datasets (Table \ref{tab:second_res}).
+These results demonstrate that if there were strong distinctions between the states of the classification schemes, the methods used here would be able to recover this.
 
@@
 
+\begin{table}[ht]
+  \centering
+  \begin{tabular}{ l r r r }
+    \hline
+    & \# predictors & In-sample AUC & Out-of-sample AUC \\ 
+    \hline
+    random forest &   11 & 1.000 & 0.999 \\ 
+    linear discriminate analysis &   10 & 1.000 & 1.000 \\ 
+    multinomial logistic regression &    6 & 1.000 & 0.998 \\ 
+    \hline
+  \end{tabular}
+  \caption{Results from classification model estimates of the secondary, multi-species dataset.
+Comparison of in-sample and out-of-sample AUC of the best performing model, along with the number of predictors.
+AUC values range between 0.5 (no better than random) and 1 (perfect classification).}
+  \label{tab:second_res}
+\end{table}
+
 \end{document}

### 2015-08-26 doc/second_tab.tex
added short text about secondary dataset

--- 2015-08-05 doc/draft_turtle.tex
@@
-\documentclass[12pt,letterpaper]{article}
-
-\usepackage{amsmath, amsthm}
-\usepackage{graphicx,hyperref}
-\usepackage{microtype, parskip}
-\usepackage[numbers,sort&compress]{natbib}
-\usepackage{lineno}
-\usepackage[font=small]{caption}
-\usepackage{subcaption, multirow, morefloats}
-\usepackage{subcaption, wrapfig}
-\usepackage{titlesec}
-\usepackage[nottoc,numbib]{tocbibind}
-\usepackage{authblk, attrib, fullpage}
-\usepackage{lineno}
-
-\frenchspacing
-
-\captionsetup[subfigure]{position = top, labelfont = bf, textfont = normalfont, singlelinecheck = off, justification = raggedright}
-
-\renewcommand{\section}[1]{%
-\bigskip
-\begin{center}
-\begin{Large}
-\normalfont\scshape #1
-\medskip
-\end{Large}
-\end{center}}
-
-\renewcommand{\subsection}[1]{%
-\bigskip
-\begin{center}
-\begin{large}
-\normalfont\itshape #1
-\end{large}
-\end{center}}
-
-\renewcommand{\subsubsection}[1]{%
-\vspace{2ex}
-\noindent
-\textit{#1.}---}
-
-\renewcommand{\tableofcontents}{}
-
-\bibpunct{(}{)}{;}{a}{}{,}  % this is a citation format command for natbib
-
-\title{How cryptic is cryptic diversity? Machine learning approaches to classifying morphological variation in \textit{Emys marmorata} (Testudinoidea, Emydidae).}
-\author[1]{Peter D Smits}%\thanks{psmits@uchicago.edu}}
-\author[1,2]{Kenneth D Angielczyk}%\thanks{kangielczyk@fieldmuseum.org}}
-\author[3]{James F Parham}%\thanks{jparham@fullerton.edu}}
-\affil[1]{Committee on Evolutionary Biology, University of Chicago}
-\affil[2]{Integrative Research Center, Field Museum of Natural History}
-\affil[3]{Department of Geological Sciences, California State University -- Fullerton}
-
-\begin{document}
-\maketitle
-\noindent{\textbf{Corresponding author:} Peter D Smits, Committee on Evolutionary Biology, University of Chicago, 1025 E.
-57th Street, Culver Hall 402, Chicago, IL, 60637, USA; E-mail: \href{mailto:psmits@uchicago.edu}{psmits@uchicago.edu}}
-
-\linenumbers
-\modulolinenumbers[2]
-
-\begin{abstract}
-\end{abstract}
-
-\section{Introduction}
-
-% cryptic diversity
-%   most species are still deliminated solely based on morphology
-With the rise of molecular systematics and techniques such as DNA barcoding, it has been increasingly recognized that some taxa can only be differentiated using molecular data \citep{Bickford2007,SchlickSteiner2007,Stuart2006,Pfenninger2007,Funk2012,Clare2011}.
-Nevertheless, the majority of extant taxa, and nearly all extinct taxa, are delimited solely via morphology.
-This disjunct is of great concern for studies of variation and diversity dynamics in deep time, where apparent morphological stasis may not reflect the true underlying diversity that existed \citep{Hunt2008,Eldredge1972,Gould1977a,VanBocxlaer2013}.
-
-Much work has been devoted to species delimitation via sequence difference \citep{Fujita2012,Yang2010b} which in turn has sparked interest in whether geometric morphometric anlaysis can capture similarly fine-scale variation that can be used for identifying cryptic species.
-Most such studies focus on morphometrics to discover differences between taxa that were identified by other means \citep{Polly2007a,Demandt2009,Gaubert2005b,Gunduz2007,Polly2003,Zelditch2004} and automated taxon identification \citep{MacLeod2007}.
-It would be very useful, though, if morphometrics approaches could reliably identify cryptic species without the use of other types of data.
-This would make the task of identifying and maintaining endangered or conserved groups much easier and could contribute to improved classifications of extinct taxa.
-
-Here, we attempt to address this issue using machine learning approaches.
-In particular, we ask whether it is possible to determine which amongst a set of classification hypotheses is best and examine the implications of the results for a recently proposed set of cryptic turtle species.
-
-% past work on automatic taxon identification and older approaches to classifying taxa
-% why use machine learning methods
-\subsection{Background and system}
-Machine learning is in many respects just an extension of known statistical methodology \citep{Hastie2009} in which the emphasis is placed on high predictive accurate at the expense of interpretability of individual parameters.
-The basic statistical mechanics are supplemented by randomization, sorting, and partitioning algorithms and along with the maximization or minimization of summary statistics, in order to best estimate a general model for all data, both sampled and unsampled \citep{Hastie2009}.
-Machine learning approaches have found use in medical research, epidemiology, economics and automated image identification such as handwritten zip codes \citep{Hastie2009}.
-Two major classes of machine learning methods are unsupervised and supervised learning.
-Unsupervised learning methods are used with unlabeled data where the underlying structure is estimated and are analogous to clustering and density estimation methods \citep{Kaufman1990}.
-Supervised learning methods are used with labeled data where the final output of data is known and the rules for going from input to output are inferred.
-These are analogous classification and regression models \citep{Breiman1984}.
-The application of the alternative approaches used in this study illustrates only a sampling of the various previously derived methods for clustering observations and fitting classification models.
-
-Geometric morphometric approaches to identifying differences in morphological variation between different classes, including cryptic species, mostly have used methods like linear discriminate analysis and canonical variates analysis \citep{Zelditch2004,Mitteroecker2011,Polly2007a,Polly2003,Gunduz2007,Gaubert2005b,Demandt2009,Sztencel-Jabonka2009,MitrovskiBogdanovic2013,Francoy2009}.
-These methods are comparatively straightforward ways of understanding the differences in morphology between classes given their similarity to familiar multivariate approaches like principal components analysis (PCA).
-They are benefit by producing results that can be easily visualized, which aids in the interpretation and presentation of data and results.
-Most previous morphometric studies did not assess which amongst a set of alternative classification hypotheses was optimal.
-For example, studies such as those of \citet{Caumul2005a} and \citet{Polly2007a} focused on comparing different aspects of morphology and their fidelity to a classification scheme instead of comparing the fidelity of one aspect of morphology to multiple classification schemes.
-In this context, the study of \citet{Cardini2009a}, is noteworthy because they compared morphological variation in marmots at the population, regional, and species level and determined the fidelity of shape to divisions at each of these levels.
-
-Here, we used multiple machine learning methods, both unsupervised and supervised, in order to compare different classification hypotheses.
-These methods provide different and unique advantages for understanding how to classify taxa, and with what accuracy.
-While machine learning methods such as neural networks have been applied to studying shape variation \citep{Baylac2003,Dobigny2003,VandenBrink2011,MacLeod2007}, including in the context of automated taxon identification and classification of groups, the number of cases remains limited.
-In the current study, we not only consider pure classification accuracy but also use a statistic of classification strength that reflects the rate at which taxa are both accurately and inaccurately classified.
-
-\subsection{\textit{Emys marmorata}}
-We analyzed the problem of whether there are distinct subspecies or cryptic species exist within the western pond turtle, \textit{Emys marmorata} (formerly \emph{Clemmys marmorata}; see \citealp{Feldman2002}).
-\textit{E.
-marmorata} is distributed from northern Washington State, USA to Baja California, Mexico.
-Traditionally, \textit{E.
-marmorata} was classified into two named subspecies: the northern \textit{E.
-marmorata marmorata}, the southern \textit{E.
-marmorata palida} \citep{Seeliger1945}.
-She also recognized a central Californian intergrade zone between these subspecies, and did not include the Baja California populations of \textit{E.
-marmorata} in these groups, implying the existence of a third distinct but unnamed subspecies.
-\textit{E.
-marmorata marmorata} is differentiated from \textit{E.
-marmorata palida} by the presence of a pair of triangular inguinal plates and darker neck markings.
-It should be noted that the triangular inguinal plates can sometimes be present in \textit{E.
-marmorata palida} though they are considerably smaller.
-
-Previous work on morphological variation in \textit{E.
-marmorata} has focused primarily on differentiation between populations over a portion of the species' total range \citep{Lubcke2007,Germano2009,Germano2008,Bury2010}; comparatively few studies have included specimens from across the entire range \citep{Holland1992}.
-Most of these studies considered how local biotic and abiotic factors may contribute to differences in carapace length and found that size can vary greatly between different populations \citep{Lubcke2007,Germano2009,Germano2008}.
-There also has been interest in size-based sexual dimorphism in \textit{E.
-marmorata} \citep{Lubcke2007,Germano2009,Holland1992}, with males being on average larger than females based on total carapace length and other linear measurements.
-However, the quality of size as a classifier of sex can vary greatly between populations \citep{Holland1992}, which makes sense in light of the amount of between population size difference \citep{Lubcke2007,Germano2009}.
-However, the effect of sexual dimorphism on shape, \textit{sensu} \citet{Kendall1977a}, was not assessed \citep{Holland1992,Lubcke2007,Germano2008}.
-
-Of particular importance in the context of cryptic diversity in \textit{E.
-marmorata} is the traditional morphometric analysis of carapace shape carried out by \citet{Holland1992}.
-He compared morphological differences between and among many populations of \textit{E.
-marmorata} in three different areas of the species' range, and studied the relative effect of distance versus barriers in fostering morphological differentiation in the species.
-\citet{Holland1992} concluded that geographic distance was a poor indicator of mophological differentiation, and instead geographic features such as breaks between different drainage basis are probably more important barriers to reproduction.
-Additionally, \citet{Holland1992} found that morphological differences were observable as the magnitude of barriers and distance increased, but the variation required many variables to adequately capture, implying only very subtle morphological differentiation between putatively distinct populations.
-\citet{Holland1992} concluded that \textit{E.
-marmorata} is best classified as three distinct species: a northern species, souther species, and a Columbia basin species.
-This classification is similar to that of \citet{Seeliger1945}, except elevated to the species level and without recognition of a distinct Baja species.
-
-More recently, the phylogeography of \textit{E.
-marmorata} and the possibility of cryptic diversity has been investigated using molecular data \citep{Spinks2005,Spinks2010,Spinks2014}.
-Based on mitochondrial DNA, \citet{Spinks2005} recognized four subclades within \textit{E.
-marmorata}, a northern clade, a San Joaquin Valley clade, a Santa Barbara clade, and a southern clade.
-Analyses with nuclear DNA \citep{Spinks2010} with single-nucleotide polymorphism (SNP) data suggest a primarily north--south division in \textit{E.
-marmorata}, although the dataset differs in the location of this break point.
-In these papers, Spinks and colleagues discussed the potential taxonomic implications of their results, with \citet{Spinks2014} going so far as to strongly advocate for the recognition of at least two species (\emph{E.
-marmorata} and \emph{E.
-pallida}), and a possible third based on populations in Baja California.
-However, they did not discuss in detail the morphological characters that would help to diagnose these species beyond those specified by \citet{Seeliger1945}.
-Given that these characters are somewhat variable within the proposed species, and that \citet{Holland1992} described shell shape variation that might be consistent with this taxonomy, we wondered whether a geometric morphometric analysis of shell shape would provide a more reliable way of diagnosing groups (whether species or subspecies) within \textit{E.
-marmorata}.
-
-In this study, we attempt to estimate the best classification scheme of \textit{E.
-marmorata} based on variation in plastral (ventral shell) shape in order to determine whether this character is consistent with any of the past divisions based on other morphological features or molecular data.
-Because of unclear geographic boundaries between subgroups of \textit{E.
-marmorata}, we compare multiple hypotheses of morphologically-- and molecularly--based classification.
-We hypothesize that if morphological variation corresponds to class assignment, then it should be possible to determine the best classification hypothesis of \textit{E.
-marmorata} from amongst multiple candidate hypotheses.
-However, if morphological variation variation does not correspond to any classification hypothesis, then supervised learning model generalization performance will be poor and reflect how variation may not follow along with any of the candidate classification hypotheses.
-
-\section{Materials and Methods}
-\subsection{Specimens, sampling, morphometrics}
-We collected landmark-based morphometric data from 354 adult \textit{E.
-marmorata} museum specimens.
-These specimens are a subset of those included in \citet{Angielczyk2007}, \citet{Angielczyk2011}, and \citet{Angielczyk2013a} and represents adult individuals.
-We chose to focus on adults because significant changes in plastron shape occur over the course of ontogeny in \textit{E.
-mamorata} and other emydines \citep{Angielczyk2013a}.
-\uppercase{Peter note -- how was something classified as an adult?}
-
-We assigned a classification to each specimen for the different binning schemes based on geographic occurrence data recorded in museum collection archives.
-When precise latitude and longitude information was not available we estimated it from whatever locality information was present.
-Because the specimens sampled to obtain the genetic data used to define the subclades were not available for study, all specimen classifications were based solely on the geographic information, not explicit assignment in previous studies.
-Because the exact barriers between different biogeographic regions are unknown and unclear, we represented each hypothesis with two different schemes; we compared a total of six different schemes.
-The schemes differed based on where geographic boundaries were assigned.
-This changes how certain individuals were assigned two one of the groups within in hypothesis such as which of the three morphologically defined groups, which of the four mitochondrially defined groups, and so on.
-% How were adults chosen?
-Good to note how schemes differeed.
-
-Following previous work on plastron variation \citep{Angielczyk2007,Angielczyk2011,Angielczyk2013a}, we used TpsDig 2.04 \citep{Rohlf2005} to digitize 19 landmarks (Fig.
-\ref{fig:plastra}).
-Seventeen of the landmarks are at the endpoints or intersection of the keratinous plastral scutes that cover the platron.
-Twelve of the landmarks were symmetrical across the axis of symmetry and, in order to prevent issues surrounding degrees of freedom and other similar concerns \citep{Klingenberg2002}, we reflected these landmarks across the axis of symmetry (i.e.
-midline) prior to analysis and used the average position of each symmetrical pair.
-In cases where damage or incompleteness prevented symmetric landmarks from being determined, we used only the single member of the pair.
-We conducted all subsequent analyses on the resulting ``half'' plastra.
-We superimposed the plastral landmark configurations using generalized Procrustes analysis \citep{Dryden1998a}, after which, we calculated the principal components (PC) of shape using the \texttt{shapes} package for R \citep{2013,Dryden2013}.
-
-\subsection{Machine learning analyses}
-\subsubsection{Unsupervised learning}
-In order to preserve the relationship between all landmark configurations in shape space, we measured the dissimilarity between observations using Kendall's Riemannian shape distance or \(\rho\) \citep{Kendall1984a,Dryden1998a}.
-We chose this metric because shape space, or the set of all possible shape configurations following Procrustes superimposition, is a Riemannian manifold and thus non-Euclidean \citep{Dryden1998a}.
-\(\rho\) varies between 0 and \(\pi / 2\) when there is no reflection invariance, which should not be a concern in the case of the half plastral landmark configurations used in the study.
-
-We divisively clustered the shape dissimilarity matrix using partitioning around mediods clustering (PAM), a method similar to \textit{k}-means clustering except that instead of minimizing the sum of squared Euclidean distances between observations and centroids, the sum of squared dissimilarities between observations and mediods is minimized \citep{Kaufman1990}.
-Because the optimal number of clusters of shape configurations in the study was unknown, being possibly three, four, or some other value, we estimated clustering solutions in which the number of clusters varied between one and eight.
-We compated clustering solutions using the gap statistic, which is a measure of goodness of clustering \citep{Tibshirani2001a}.
-
-%The gap statistic is defined
-%\[Gap_{n}(k) = E^{*}_{n}[\log(W_{k})] - \log(W_{k})\] 
-%where \(W_{k}\) is
-%\[W_{k} = \sum^{k}_{r = 1}{\frac{1}{2n_{r}} (\sum_{i,i' \in C_{r}} d_{ii'})}\].
-%\(d_{ii'}\) is the dispersion of the clustering solution or the sum of the pairwise dissimilarities between observations in each cluster and their respective mediods (\(C\)) for all clusters \(r\).
-This value is averaged and compared to the expected dispersion (\(E^{*}_{n}\)) of a sample \(n\) from a reference distribution.
-In this case, the reference distribution was estimated from 500 resamples.
-
-We conducted this analysis using the \texttt{cluster} package for R \citep{Maechler2013}.
-
-\subsubsection{Supervised learning}
-We used three different supervised learning, or classification, approaches: linear discriminate analysis, multinomial logistic regression, and random forests.
-Linear discriminate analysis, also known as canonical variate analysis, is commonly used in studies of geometric morphometric data \citep{Zelditch2004,Mitteroecker2011}.
-The other two methods, however, are not.
-In all cases, the optimal number of PCs used as predictors was chosen via maximum within-sample AUC value, explained below.
-
-Linear discriminate analysis (LDA) attempts to find a linear combination of predictors that best model two or more classes.
-LDA is very similar to PCA except that instead of finding the linear combination of features that maximize the amount of explained variance in the data, LDA maximizes the differences between classes.
-The results of this analysis produces a transformation matrix by which the original features can be transformed to reflect the best discrimination between the classes.
-In this study, we applied LDA to the eigenscores from a subset of the total number of PCs, ranging from two to 6 in increasing order of complexity.
-In total, this produced nine different LDA scaling matrices.
-
-Multinomial logistic regression is an extension of logistic regression, where instead of a binary response there are three or more response classes \citep{Venables2002a}.
-Similar to the odds ratios calculated from the coefficients of a logistic regression, the relative risk of a classification can be determined from the coefficients of the model.
-
-Random forest models are an extension of classification and regression trees (CART) \citep{Breiman1984,Breiman2001}.
-The goal of CARTs are to use a series of different features (i.e.
-predictors) to estimate the class of an observation.
-In top-down induction of decision trees for each member of a given set of predictor variables, attribute value tests are used to estimate the differences between classes.
-This process, called recursive partitioning, is then repeated on each subset.
-The recursion continues until the resulting observations all share the same class or no more meaningful partitions are possible.
-The resulting model is a tree structure by which observations are classified at each intersection via the estimated cutoff points from the attribute tests made during model fitting.
-% figure to help explain the ``tests''
-
-In a random forest model, many CARTs are built from a random subsample of both the features and the observations (specimens).
-This process is then repeated many times and the parameters of the final model are chosen as the mode of estimates from the distribution of CARTs \citep{Breiman2001}.
-In addition to classifying the observations, this procedure allows for the features to be ranked in order of importance.
-This is a generally useful property for studies in which the goal is to describe and model the differences between classes and the relative importance of different predictors.
-
-In this analysis, we used 1000 subtrees to estimate the random forest model parameters.
-We estimated the best set of predictors necessary for each classification scheme was estimated using a recursive feature selection algorithm, and we chose the optimal number of PCs to include based on the AUC of the model.
-Following the backwards selection algorithm implemented in \texttt{caret} \citep{Kuhn2013}, the maximum number of features were included in the initial model, their importance ranked, and the AUC of the model calculated.
-The lowest ranked feature was then removed, and the AUC of the model recalculated.
-This was repeated until only one feature, remained.
-Because PCs were kept in order of importance and not in relation to the amount of variance each PC described, these means that the PCs are not included in the order of ascending eigenvalue.
-
-In classification studies, such as this one, a common metric of performance is area under the receiver operating characteristic curve (AUC).
-AUC is an estimate of the relationship between the false positive and true positive rates, as opposed to just the true positive rate (accuracy).
-This relationship is especially useful in cases where misclassification needs to be minimized just as much as accurate classification, as in this study.
-AUC ranges between 0.5 and 1, with 0.5 indicating classification no better than random and 1 indicating perfect classification \citep{Hastie2009}.
-
-The standard AUC calculation is defined for binary classifications, however in this application there are multiple categories.
-The alternative calculation that we used follows an all-against-one strategy where the individual AUC values for each class versus all others are averaged to produce a multiclass AUC \citep{Hand2001}.
-To estimate confidence intervals on the out-of-sample AUC values, we performed a nonparametric bootstrap in which the true and estimated classifications were resampled with replacement.
-This was done 1000 times.
-
-The ultimate measure of model fit is accurately predicting the values of unobserved samples \citep{Hastie2009,Kuhn2013}.
-Within-sample performance is inherently biased upwards, so model evaluation requires overcoming this bias.
-With very large sample sizes, as in this study, part of the sample can be used as the ``training set'' and the remainder acts as the ``testing set.'' The former is used for fitting the model where as the later is used for measuring model performance, and this process is called model generalization.
-In this analysis, we used 75\% of samples as the training set while the remaining 25\% were used as the testing set.
-
-It is common for some out of sample observations to be misclassified.
-This misclassification may be due to the model not accurately representing shape variance, systematic differences between the training and test sets, or systematic differences between the accurately and inaccurately classified samples.
-Testing and training sets are determined completely at random within each class and with respect to shape.
-Results were not effected by changes in testing or training set assignment.
-
-To determine if there were systematic differences between the correctly and incorrectly classified samples, we used a permutation test to estimate if the dissimilarity between the correctly and incorrectly classified individuals were significantly different from random.
-The group labels were permuted 1000 times and the distance between the new centroids was calculated.
-The number of permutations less than the empirical difference divided by 1000 gives a \textit{p}-value for the test.
-Significant results indicate that correctly and incorrectly classified specimens are systematically different.
-This was done only for classes where there were 10 or more observations.
-
-\section{Results}
-
-\subsection{Unsupervised learning}
-Comparison of gap statistic values from PAM clustering show that the optimal, minimal number of clusters is most likely one (Fig.
-\ref{fig:gap}).
-There is some ambiguity in choice because, although it is not statistically different from a solution with only one group, the solution with two groupings does have the greatest mean gap statistic.
-However, there is no geographical signal in the results of this clustering solution (Fig.
-\ref{fig:gap_map}).
-Because of this, we assert that this means that there is no means of naturally partitioning plastron shape into distinct subgroups with out reference to external information.
-% PAM stuff
-% figure
-
-\subsection{Supervised learning}
-% model selection/fit
-AUC--based model selection revealed some important patterns of variation and congruence between the classification schemes and the actual data.
-Generally, the best performing models tended to include as many PCs as possible \ref{fig:sel}).
-Note that the best random forest models were determined via recursive feature selection, so PCs were not included in order of percent variance explained.
-That almost all LDA and multinomial logistic regression models were as complex as possible indicates that the differences between the different groups within each classification scheme are very small.
-
-As part of fitting a random forest model, a ranking of variable importance also is determined.
-Interestingly, the order of variable importance is not the same as the order of the PCs (\ref{fig:var_imp}).
-This means that the variance describing the differences between the classes does not align with the major axes of variance (i.e.
-the PCs).
-This result would be the case if variation between classes was extremely fine grained and not a part of the principal form or function of the plastron, which makes sense given that the plastron is involved in both protection and hydrodynamics and not necessarily mate choice \citep{Germano2009,Holland1992,Lubcke2007,Rivera2008}.
-Moreover, this result is congruous with the results from the AUC--based model selection for the multinomial logistic regression and LDA models.
-
-Observed AUC values for all of the optimal models are not exceptionally high (\ref{fig:sel}).
-In most cases the different proposed classification schemes are generally poor descriptors of the observed variation.
-It appears that the data set is overwhelmed by noise, making any accurate classifications difficult at best.
-This observation is cemented with the generalizations of the models to the testing data set (\ref{fig:gen_hist}).
-
-% generalization
-Mean AUC values for the model generalizations, in most cases, are approximately equal to the observed AUC values from the training data set (Table \ref{tab:comp}).
-The  cases in which the AUC from the  generalizations is less than the observed, indicate poor model fit and a poor classification scheme.
-AUC values from model generalization, or estimating testing data set membership, does not indicate a clear ``best'' classification scheme (Fig.
-\ref{fig:gen_hist}).
-Although the scheme with two species has the greatest AUC point estimate for each modeling approach, this scheme is not significantly greater than any other except in some limited cases (e.g.
-LDA, Table \ref{tab:gen_tests}).
-% misfits
-Differences in mean shape between correctly and incorrectly classified observations from test set frequently were statistically significant, though there are exceptions.
-Again, this test was to determine if the mean shapes were statistically different or not.
-The frequency of these results, however, is important because it means that the different models are poor predictors of class membership.
-This may be because differences in plastron shape do not align with the any of the hypothesized classification schemes.
-
-\section{Discussion}
-
-The results of this study indicate that there is no clear grouping of \textit{E. marmorata} based on plastron shape.
-
-The unsupervised learning results which indicate only a single group of observations being optimal is congruous with the results from the generalizations of the supervised learning models.
-The classification schemes used in the supervised learning models correspond, loosely, to unsupervised learning solutions with multiple groups.
-Because unsupervised learning solutions with multiple groups are poor descriptors of the observed variation, it is important to see this generally supported by the supervised learning results.
-
-The results from fitting the various supervised learning models for each of the classification scheme generally shows that no one scheme is ``best.'' A possible explanation for this that the genetic divergence associated with (sub)speciation is either not based on plastron morphology or local selective pressures due to hydrological regime overwhelming any possible morphological divergence.
-
-Both the low AUC values (\(< 0.9\)) and the significant difference between the correctly and incorrectly classified observations support the conclusion that none of the hypothesized classification schemes are good descriptors of the observed plastral variation.
-
-INSERT MORE TEXT HERE.
-
-\clearpage
-\bibliographystyle{sysbio}
-\bibliography{turtle,packages}
-
-\clearpage
-
+% latex table generated in R 3.2.2 by xtable 1.7-4 package
+% Wed Aug 26 12:02:25 2015
 \begin{table}[ht]
   \centering
-  \include{comp_tab}
-  \caption{AUC values for the best model of each classification scheme for both the observed (training) data and the generalized (testing) data.
-Results from all three different supervised learning approaches are shown here.
-AUC values range between 0.5 and 1.
-}
-  \label{tab:comp}
+  \begin{tabular}{ l r r r }
+    \hline
+    & \# predictors & In-sample AUC & Out-of-sample AUC \\ 
+    \hline
+    random forest &   11 & 1.000 & 0.999 \\ 
+    linear discriminate analysis &   10 & 1.000 & 1.000 \\ 
+    multinomial logistic regression &    6 & 1.000 & 0.998 \\ 
+    \hline
+  \end{tabular}
+  \caption{Results from classification model estimates of the secondary, multi-species dataset.
+Comparison of in-sample and out-of-sample AUC of the best performing model, along with the number of predictors.
+AUC values range between 0.5 (no better than random) and 1 (perfect classification).}
+  \label{tab:second_res}
 \end{table}
-
-\begin{table}
-  \include{rf_dif}
-    %
-  \include{mm_dif}
-    %
-  \include{ll_dif}
-  \caption{Results of bootstrap comparisons between the scheme with the highest mean AUC value and all other schemes.
-An asterix indicates the best scheme.
-This was done for each of the three modeling techniques included in this study.
-Probabilities are the percent of comparisons that are greater than the observed difference in means.}
-  \label{tab:gen_tests}
-\end{table}
-
-\begin{table}
-  \begin{scriptsize}
-    \include{rf_miss}
-
-    \include{mm_miss}
-
-    \include{ll_miss}
-  \end{scriptsize}
-  \caption{Results of comparisons between correctly and incorrectly classified observations from the testing data set.
-For each scheme, the classifications with at least 10 observations were tested.
-This was done for each of the three modeling techniques included in this study.}
-  \label{tab:miss_tests}
-\end{table}
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/plastra}
-  \caption{Depiction of general plastral shape of \textit{E. marmorata} and position of the 19 landmark used in this study. Anterior is towards the top of the figure.}
-  \label{fig:plastra}
-\end{figure}
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/gap_res}
-  \caption{Results from PAM clustering of the Riemannian shape distance for 8 different number of clusters. Vertical lines are 1 standard deviation of the mean determined from 500 resamples.}
-  \label{fig:gap}
-\end{figure}
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/gap_map}
-  \caption{Comparison of geographic distribution of clustered observations from the 2 clustering PAM solution.
-Colour and shape correspond to each of the groups.
-There is clearly no geographic signal in the data.}
-  \label{fig:gap_map}
-\end{figure}
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/sel_val}
-  \caption{Graphical representation of the AUC values from model selection for multinomial logigistic regression and linear discriminate analysis, respecitively.
-AUC model selection is based on greatest AUC value.
-The horizontal axis corresponds to the cummulative number of axes included in the model of interest.
-A red dot corresponds to the AUC best model for that classification scheme.}
-  \label{fig:sel}
-\end{figure}
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/var_imp}
-  \caption{Variable importance from the random forest models for each of the six classification schemes.
-Importance is measured as the mean decrease in Gini Index, which is a measure of the strength by which that variable determines CART structure.
-Indices that are farther to the right indicate greater variable importance.}
-  \label{fig:var_imp}
-\end{figure}
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/gen_res}
-  \caption{Bootstrap distributions for genearlized AUC values for each of the classification schemes.
-Each row corresponds to a different modeling approach: LDA, LDA using best variables from random forest, multinomial logistic regression, and random forest.
-Each distribution corresponds to 1000 bootstrap replicates.}
-  \label{fig:gen_hist}
-\end{figure}
-
-\end{document}

### 2015-09-08 doc/draft_turtle.tex
abstract, jim ken comments, need new graphs

--- 2015-08-10 doc/draft_turtle.tex
@@
 \begin{abstract}
-  ABSTRACT NEEDS TO BE WRITTEN
+  We investigate the morphometric identification of cryptic species using machine-learning approaches by examining their implications for a recently proposed cryptic turtle species (\textit{Emys pallida}).
+We collected landmark-based morphometric data from 354 adult \textit{E.
+marmorata/ ``pallida''} museum specimens.
+We assigned a classification to each specimen for six different binning schemes based on geographic occurrence data recorded in museum collection archives.
+We used multiple machine learning methods, both unsupervised and supervised, to compare different classification hypotheses and asked whether it is possible to determine which amongst a set of classification hypotheses is best.
+In addition, we applied the above approach to a “clear-cut” example of seven unambiguously distinct species closely related to \textit{E.
+marmorata}.
+The results of this study indicate that there is no clear grouping of \textit{E.
+marmorata/ ``pallida''} based on plastron shape.
+In contrast, the analysis of the ``clear-cut'' example demonstrates a near perfect classification, which demonstrates that the methods can recover correct results when an appropriate signal exists.
+Explanations for the lack of grouping in \textit{E.
+marmorata} include that possibility that genetic differentiation is not associated with plastron shape variation below the species level and/or that local selective pressures (e.g., from hydrological regime) overwhelm morphological differentiation.
+A reconsideration of the methods used to delimit \textit{E.
+``pallida,''} the lack of barriers to gene flow, the strong evidence for widespread admixture between lineages, and the fact that plastron shape can be used to differentiate other emydine species suggest that its lack of diagnosability most likely reflects the non-distinctiveness of this proposed taxon.
 \end{abstract}
@@
 Molecular systematics has repeatedly demonstrated the existence of cryptic species that can only be diagnosed using genetic data \citep{Stuart2006,Bickford2007,SchlickSteiner2007,Pfenninger2007,Clare2011,Funk2012}.
-In attempts to streamline the documentation of biodiversity, several methods of species delimitation that rely almost entirely on genetic data have recently been proposed \citep{Yang2010b}.
+In attempts to streamline the documentation of biodiversity, several methods of species delimitation that rely almost entirely on genetic data have recently been proposed \citep{Pons2006,Carstens2010,Hausdorf2010,O'Meara2010,Yang2010b,Huelsenbeck2011b}.
 Although strong caveats on the utility of these methods have been raised \citep{Bauer2000,Carstens2013}, they are nevertheless being used to name species \citep{Leache2010,Spinks2014}.
@@
 These considerations have sparked interest in whether geometric morphometric analyses can capture fine-scale variation that can be used for identifying cryptic species.
+This would make the task of idetifying and maintaining endagered or conserved groups much easier and could contribute to improved classifications of extinct taxa and populations.
 Most such studies focus on using morphometrics to discover differences between taxa that were identified by other means \citep{Polly2003,Zelditch2004,Gaubert2005b,Gunduz2007,Polly2007a,Demandt2009}.
-Additionally, there has been a fair amount of work on automated taxon identification and classification of taxa into groups \citep{Baylac2003,Dobigny2003,MacLeod2007,VandenBrink2011}.
-In cases where genetic data are ambiguous or lacking for many samples, morphometric approaches could help identify cryptic species.
-This would make the task of identifying and maintaining endangered or conserved groups much easier and could contribute to improved classifications of extinct taxa and populations.
+Additionally, there has been work on automated taxon identification and classification of taxa into groups \citep{Baylac2003,Dobigny2003,MacLeod2007,VandenBrink2011}.
 
@@
 These are analogous to classification and regression models \citep{Breiman1984}.
-Out application of the alternative approaches used in this study illustrates only a sampling of the various previously-derived methods for clustering observations and fitting classification models.
+Our application of the approaches used in this study illustrates only a sampling of the various methods available for clustering observations and fitting classification models.
 
@@
 Here, we used multiple machine learning methods, both unsupervised and supervised, to compare different classification hypotheses.
-These methods provide different and unique advantages for understanding how to classify taxa, as well as the accuracy of the resulting classifications.
+These methods provide different advantages for understanding how to classify taxa, as well as the accuracy of the resulting classifications.
 Although machine learning methods such as neural networks have been applied to studying shape variation \citep{Baylac2003,Dobigny2003,MacLeod2007,VandenBrink2011}, including in the context of automated taxon identification and classification of groups, the number of cases remains limited.
@@
 
-We analyzed the problem of whether there are distinct subspecies or cryptic species within the western pond turtle, \textit{Emys marmorata} BAIRD CITATIONS (formerly \emph{Clemmys marmorata}; see \citealp{Feldman2002}).
+We analyzed the problem of whether there are distinct subspecies or cryptic species within the western pond turtle, \textit{Emys marmorata} \citep{Baird1852} (formerly \emph{Clemmys marmorata}; see \citealp{Feldman2002}).
 \textit{Emys marmorata} is distributed from northern Washington State, USA to Baja California, Mexico.
@@
 The triangular inguinal plates can sometimes be present in \textit{E.
-marmorata pallida} though they are considerably smaller.
+marmorata pallida} although they are considerably smaller.
 \citet{Seeliger1945} did not formally include the Baja California populations of \textit{E.
@@
 marmorata} \citep{Holland1992,Lubcke2007,Germano2009}, with males being on average larger than females based on total carapace length and other linear measurements.
-However, the quality of size as a classifier of sex can vary greatly between populations \citep{Holland1992} because of the amount of size differences among populations \citep{Lubcke2007,Germano2009}.
+However, the quality of size as a classifier of sex can vary greatly between populations \citep{Holland1992} because of the magnitude of size differences among populations \citep{Lubcke2007,Germano2009}.
 However, the effect of sexual dimorphism on shape, \textit{sensu} \citet{Kendall1977a}, has not been assessed \citep{Holland1992,Lubcke2007,Germano2008}.
@@
 marmorata}, a northern clade, a San Joaquin Valley clade, a Santa Barbara clade, and a southern clade.
-Analyses with nuclear DNA \citep{Spinks2010} with single-nucleotide polymorphism (SNP) data suggest a primarily north--south division in \textit{E.
+Analyses with nuclear DNA \citep{Spinks2010} and single-nucleotide polymorphism (SNP) data suggest a primarily north--south division in \textit{E.
 marmorata}, although the dataset differed from that of \citet{Spinks2005} in the location of this break point.
@@
 We chose to focus on adults because significant changes in plastron shape occur over the course of ontogeny in \textit{E.
-mamorata} and other emydines \citep{Angielczyk2013a}.
-\uppercase{Peter note -- how was something classified as an adult?}
+marmorata} and other emydines \citep{Angielczyk2013a}.
 
@@
 When precise latitude and longitude information were not available we estimated them from locality information.
-Because \citet{Spinks2005}, \citet{Spinks2010}, and \citet{Spinks20140} did not use vouchered specimens we were not able to example the sample individuals.
-Therefore our specimen classifications were based solely on the geographic information, not explicit assignment in previous studies.
+Because \citet{Spinks2005}, \citet{Spinks2010}, and \citet{Spinks2014} did not use vouchered specimens we were not able to directly sample individuals in their studies.
+Therefore our specimen classifications were based solely on the geographic information, not explicit assignment using molecular data.
 Because the exact barriers between different biogeographic regions are unknown and unclear, we represented each hypothesis with two schemes for a total of six different schemes.
@@
 The results of this analysis produces a transformation matrix by which the original features can be transformed to reflect the best discrimination between the classes.
-In this study, we applied LDA to the eigenscores from a subset of the total number of PCs, ranging from two to 6 in increasing order of complexity.
+In this study, we applied LDA to the eigenscores from a subset of the total number of PCs, ranging from two to six in increasing order of complexity.
 In total, this produced nine different LDA scaling matrices.
@@
 
+\subsection{Comparison with clear-cut example}
+In addition to the above analysis of classification schemes of \textit{E.
+marmorata/pallida}, we applied the above approach to a selection of seven morphologically distinct emydine species.
+This additional analysis was done in order to confirm the efficacy of our supervised learning approach.
+The seven species analyzed were \textit{Emys blandigii}, \textit{Terrapene coahuila}, \textit{Clemmys guttata}, \textit{Glyptemys insculpta}, \textit{Glyptemys muhlenbergii}, \textit{Emys orbicularis}, and \textit{Terrapene ornata} with a total of 578 specimens analyzed in total.
+Again, these data were a subset of those used in \citet{Angielczyk2011} and \citet{Angielczyk2013a}.
+
+As with the \textit{E.
+marmorata/pallida} analysis, we analyzed plastron shape variation.
+The support for a seven category classification scheme was evaluated using random forest, linear discriminate analysis, and multinomial logistic regression where the number of PCs used as predictors ranged from two to 11.
+Data were split into training and testing sets and model performance for both sets was evaluated using the AUC metric.
+
 \section{Results}
@@
 Such a result is expected if variation between classes was extremely fine grained and not a part of the principal form or function of the plastron.
-Moreover, this result is congruous with the results from the AUC--based model selection for the multinomial logistic regression and LDA models.
+Moreover, this result is consistent with the results from the AUC--based model selection for the multinomial logistic regression and LDA models.
 
@@
 
+\subsection{Comparison with clear-cut example}
+In contrast with the above results, the analysis of the seven morphologically-distinct species demonstrates near perfect classification of the both the in-sample and out-of-sample datasets (Table \ref{tab:second_res}).
+These results demonstrate that when there are strong distinctions between the states of the classification schemes, the methods used here can recover them.
+
 \section{Discussion}
@@
 
-The unsupervised learning results indicate only a single group of observations being optimal, which is congruous with the results from the generalizations of the supervised learning models.
+The unsupervised learning results indicate only a single group of observations being optimal, which is consistent with the results from the generalizations of the supervised learning models.
 The classification schemes used in the supervised learning models correspond, loosely, to unsupervised learning solutions with multiple groups.
@@
 from hydrological regime) overwhelm morphological differentiation.
-This makes sense given that the plastron is involved in both protection and hydrodynamics, and not necessary mate choice CITATIONS and that shell shape in \textit{E.
-marmorata} is known to vary among populations inhabiting water bodies with different flow regimes CITATIONS.
-Plastron shape does not seem to preserve a strong phylogenetic signal at the interspecific level in emydine turtles \citep{Angielczyk2011}, and out current results suggest that this may be the case for phylogeographic signal within emydine species as well.
-Stated another way, plastron shape may be a poor morphological marker for phylogenetic and phylogeographic studies, despite the ease with which its shape can be quantified.
+This makes sense given that the plastron is involved in both protection and hydrodynamics, and not necessary mate choice \citep{Rivera2008,Rivera2011,Stayton2011,Rivera2014} and that shell shape in \textit{E.
+marmorata} is known to vary among populations inhabiting water bodies with different flow regimes \citep{Holland1992,Lubcke2007,Germano2009}.
+Plastron shape does not seem to preserve a strong phylogenetic signal at the interspecific level in emydine turtles \citep{Angielczyk2011}, and our current results suggest that this may be the case for phylogeographic signal within emydine species as well.
 A final possibility (explored below) is that the proposed classification schemes themselves do not represent significant evolutionary lineages.
 
-Both the low AUC values (\(< 0.9\)) and the significant difference between the correctly and incorrectly classified observations support the conclusion that none of the hypothesized classification schemes are good descriptors of the observed plastral variation.
+Both the low AUC values (\(< 0.9\)) and the significant difference between the correctly and incorrectly classified observations support the conclusion that none of the hypothesized classification schemes are good descriptors of the observed plastral variation within \textit{E.
+marmorata}.
+
+Nevertheless, it is important to note that plastron shape is an extremely effective method for differentiating members of the other seven species we investigated.
+The magnitude of shape differences between the species (measured as Procrustes distance between species' mean shapes) is approximately an order of magnitude greater than the differences between the \textit{E.
+marmorata} subgroups, and the machine learning methods had no trouble accurately classifying the specimens correctly.
+These results demonstrate that plastron shape is normally a good marker for species delineation iin the closest relatives of \textit{E.
+marmorata}, and that our lack of results for \textit{E.
+marmorata} is not simply a shortcoming of the methods we applied.
+Indeed, they beg the question of what factors have suppressed morphological differentiation of plastron shape in \textit{E.
+marmorata} and \textit{E.
+pallida} if they are distinct species.
+Invoking issues such as the role of the plastron in protection or the need for streamlining are insufficient because the other species are expected to be subject to similar constraints \citep{Stayton2011,Pollyb}.
 
@@
 pallida} does not, on its own, preclude the recognition of this taxon.
-However, this apparent lack of congruence does prompt a reexamination of the methods and concepts that led to that taxonomic revision.
-In other words, before we can assess the significance of the morphological non-diagnosablity, it is essential to evaluate the methods and concepts that led to taxonomic revision.
+However, this apparent lack of congruence does prompt a reexamination of the methods and concepts that led to that taxonomic revision, especially considering that plastron shape is normally a strong indicator of species differences amoung emydines.
+In other words, before we can assess the significance of the morphological non-diagnosablity, it is essential to evaluate the methods and concepts that led to the initial taxonomic revision.
 
@@
 However, \citet{Spinks2014} did not heed the caveats about species delimitation methods raised by \citet{Carstens2013}.
-In addition to specifically addressing the shortcomings of validation methods such as BPP that rely on guide trees and so “should be interpreted with caution,” \citet{Carstens2013} also strongly emphasize that ``Inferences regarding species boundaries based on genetic data alone are likely inadequate, and species delimitation should be conducted with consideration of the life history, geographical distribution, morphology and behaviour (where applicable) of the focal system\dots'' These caveats evoke the development of the Unified Species Concept \citep{Dayrat2005a,DeQueiroz2007b}, Integrative Taxonomy CITATIONS Padial 2010, and other pluralist approaches to species delimitation.
+In addition to specifically addressing the shortcomings of validation methods such as BPP that rely on guide trees and ``should be interpreted with caution,'' \citet{Carstens2013} also strongly emphasize that ``Inferences regarding species boundaries based on genetic data alone are likely inadequate, and species delimitation should be conducted with consideration of the life history, geographical distribution, morphology and behaviour (where applicable) of the focal system\dots'' These caveats evoke the development of the Unified Species Concept \citep{Dayrat2005a,DeQueiroz2007b}, Integrative Taxonomy \citep{Padial2010}, and other pluralist approaches to species delimitation.
 None of these considerations were brought to bear on the \textit{E.
 marmorata} system until now, and in doing so we find the proposal that \textit{E.
-pallida} is a distinct species to be lacking.
+pallida} is a distinct species to be lacking in a normally robust morphological marker.
 
-In particular, the natural history and geographical distribution of \textit{E.
+The natural history and geographical distribution of \textit{E.
 marmorata} and \textit{E.
-pallida} make the recognition of these taxa implausible.
+pallida} also make the recognition of these taxa implausible.
 The data from \citet{Spinks2014} show extensive introgression and admixture in Central California, which is expected because there are no significant barriers to gene flow in this region.
@@
 pallida} could maintain their integrity in the face of such admixture.
-Because the geography, natural history, and demonstrated genetic admixture of \textit{E.
-marmorata} conflict with the recognition of \textit{E.
+Because the geography, natural history, demonstrated genetic admixture of \textit{E.
+marmorata}, and comparisons with other morphologically-diagnosable species conflict with the recognition of \textit{E.
 pallida}, we hypothesize that out inability to classify the morphological data by proposed species is because \textit{E.
@@
 We fully agree with \citet{Spinks2014} that \textit{E.
-marmorata} (\textit{sensu lato}) is a species deserving of strong conservation efforts, we do not wish to trivialize this need.
+marmorata} (\textit{sensu lato}) is a species deserving of strong conservation efforts, and we do not wish to trivialize this need.
 Moreover, the genetic diversity uncovered by the analysis of \citet{Spinks2014} should be explicitly accounted for in any conservation plan.
@@
 Finally, it is important to note that the data and analyses we present do not let us definitively say whether the apparent lack of morphological divergence within \textit{E.
-marmorata} truly reflects the presence of a single species, or if it is an artifact of plastron shape being a poor morphological marker for phylogenetic and phylogeographic divergences.
-Tests of both our preferred conclusion (\textit{E.
-marmorata} as a single species) and that \citet{Spinks2014} should include additional tests of species delimitation using different methods and search for corroborating evidence as suggested by \citet{Carstens2013}.
+marmorata} truly reflects the presence of a single species, or if it is an artifact of plastron shape being a poor morphological marker for phylogenetic and phylogeographic divergences, because we could not carry out our morphometric analyses on the specimens from which the genetic data were obtained.
+The comparisons with the other emydine species, suggests it is because \textit{E.
+marmorata} is a single species.
+However, tests of both our preferred conclusion (\textit{E.
+marmorata} as a single species) and that \citet{Spinks2014} should include morphological and molecular analyses of the same set of voucher specimens, as well as additional tests of species delimitation using alternative methods and corroborating evidence as suggested by \citet{Carstens2013}.
 From a morphological standpoint, support for the validity of ``\textit{E.
-Pallida}'' may come from other aspects of morphology, such as carapace shape or other features that might be more informative.
-Lifewise, further investigation of the phylogeographic utility of plastron shape in other turtle species will help to clarify whether the lack of differentiation seen in \textit{E.
-marmoarata} is typical or an unusual case.
+Pallida}'' may come from other aspects of morphology, such as carapace shape or other features.
+Likewise, further investigation of the phylogeographic utility of plastron shape in other turtle species will help to clarify whether the lack of differentiation seen in \textit{E.
+marmoarata}, and the strong differentiation among the other emydines, is typical or an unusual case.
 
@@
 \begin{table}
-  \include{rf_dif}
-    %
-  \include{mm_dif}
-    %
-  \include{ll_dif}
-  \caption{Results of bootstrap comparisons between the scheme with the highest mean AUC value and all other schemes.
+  \centering
+  \begin{tabular}{ l r r r }
+    \hline
+    Scheme & RF & MLR & LDA \\
+    \hline
+    \hline
+    Morph 1 & \(\ast\) & \(\ast\) & 1 \\
+    Morph 2 & 0.79 & 0.55 & 1 \\
+    Mito 1 & 0.89 & 0.94 & 1 \\ 
+    Mito 2 & 0.82 & 0.57 & \(\ast\) \\ 
+    Mito 3 & 0.79 & 0.69 & 0.73 \\ 
+    Nuclear & 0.79 & 0.96 & 0.96 \\ 
+    \hline
+  \end{tabular}
+  \caption{Results of bootstrap comparisons between the scheme with the highest mean AUC value and all other schemes, defined P(best - other \(>\) 0).
 An asterix indicates the best scheme.
-This was done for each of the three modeling techniques included in this study.
+This was done for each of the three modeling techniques included in this study: random forest (RF), multinomial logistic regression (MLR), and linear discriminate analysis (LDA).
 Probabilities are the percent of comparisons that are greater than the observed difference in means.}
@@
 \begin{table}
-  \begin{scriptsize}
-    \include{rf_miss}
-
-    \include{mm_miss}
-
-    \include{ll_miss}
-  \end{scriptsize}
-  \caption{Results of comparisons between correctly and incorrectly classified observations from the testing data set.
+  \centering
+  \begin{tabular}{ l | l r r r }
+    \hline
+    Scheme & Class & RF & MLR & LDA \\
+    \hline
+    \hline
+    \multirow{2}{*}{Morph 1} & CCR & 0.77 & 1.00 & 1.00 \\ 
+    & marm & 0.87 & 0.93 & 1.00 \\ 
+    \hline
+    \multirow{2}{*}{Morph 2} & CCR & 0.88 & 1.00 & 1.00 \\ 
+    & marm & 1.00 & 1.00 & 0.98 \\ 
+    \hline
+    \multirow{2}{*}{Mito 1} & CCR & 0.94 & 0.99 & 0.98 \\ 
+    & marm & 0.99 & 0.98 & 0.47 \\ 
+    \hline
+    \multirow{2}{*}{Mito 2} & marm & 0.85 & 0.97 & 0.96 \\ 
+    & pall & 0.94 & 1.00 & 1.00 \\ 
+    \hline
+    \multirow{2}{*}{Mito 3} & 1 & 0.40 & 0.92 & 0.90 \\ 
+    & 3 & 0.97 & 0.99 & 1.00 \\ 
+    \hline
+    \multirow{2}{*}{Nuclear} & marm & 1.00 & 1.00 & 1.00 \\ 
+    & pall & 0.99 & 1.00 & 1.00 \\ 
+    \hline
+  \end{tabular}
+  \caption{Results of comparisons between correctly and incorrectly classified observations from the testing data set, defined P(distance - simulated \(>\) 0).
 For each scheme, the classifications with at least 10 observations were tested.
-This was done for each of the three modeling techniques included in this study.}
+This was done for each of the three modeling techniques included in this study: random forest (RF), multinomial logistic regression (MLR), and linear discriminate analysis (LDA).}
   \label{tab:miss_tests}
@@
   \caption{Bootstrap distributions for genearlized AUC values for each of the classification schemes.
-Each row corresponds to a different modeling approach: LDA, LDA using best variables from random forest, multinomial logistic regression, and random forest.
+Each row corresponds to a different modeling approach: multinomial logistic regression (MLR), random forest (RF), linear discriminate analysis (LDA), and LDA using best variables from random forest.
 Each distribution corresponds to 1000 bootstrap replicates.}
@@
 
+\begin{table}[ht]
+  \centering
+  \begin{tabular}{ l r r r }
+    \hline
+    & \# predictors & In-sample AUC & Out-of-sample AUC \\ 
+    \hline
+    RF &   11 & 1.000 & 0.999 \\ 
+    MLR &    6 & 1.000 & 0.998 \\ 
+    LDA &   10 & 1.000 & 1.000 \\ 
+    \hline
+  \end{tabular}
+  \caption{Results from classification model estimates of the secondary, multi-species dataset.
+Models are random forest (RF), linear discriminate analysis (LDA), and multinomial logistic regression (MLR).
+Comparison of in-sample and out-of-sample AUC of the best performing model, along with the number of predictors.
+AUC values range between 0.5 (no better than random) and 1 (perfect classification).}
+  \label{tab:second_res}
+\end{table}
+
 \end{document}

### 2015-09-08 doc/second_tab.tex
abstract, jim ken comments, need new graphs

@@
 % latex table generated in R 3.2.2 by xtable 1.7-4 package
-% Wed Aug 26 12:02:25 2015
+% Tue Sep  8 12:03:54 2015
 \begin{table}[ht]
 \centering
-  \begin{tabular}{ l r r r }
+\begin{tabular}{rrrr}
   \hline
-    & \# predictors & In-sample AUC & Out-of-sample AUC \\ 
+ & npred & ii.auc & oo.auc \\ 
   \hline
-    random forest &   11 & 1.000 & 0.999 \\ 
-    linear discriminate analysis &   10 & 1.000 & 1.000 \\ 
-    multinomial logistic regression &    6 & 1.000 & 0.998 \\ 
+RF &   11 & 1.000 & 1.000 \\ 
+  LDA &   10 & 1.000 & 1.000 \\ 
+  MnL &    4 & 1.000 & 1.000 \\ 
    \hline
 \end{tabular}
-  \caption{Results from classification model estimates of the secondary, multi-species dataset.
-Comparison of in-sample and out-of-sample AUC of the best performing model, along with the number of predictors.
-AUC values range between 0.5 (no better than random) and 1 (perfect classification).}
 \label{tab:second_res}

### 2015-09-08 R/make_plots.r
graphs for seven spec

@@ -26,7 +26,7 @@ theme_update(axis.text = element_text(size = 20),
              legend.text = element_text(size = 25),
              legend.title = element_text(size = 26),
              legend.key.size = unit(2, 'cm'),
-             strip.text = element_text(size = 25))
+             strip.text = element_text(size = 20))
 
 
 # map details
@@ -48,88 +48,88 @@ goog.map <- get_map(location = c(longmi,
 ggmap(goog.map)
 
 
-# PCA results
-morph <- ggpairs(adult, columns = c(1:3),
-                 upper = 'blank',
-                 params = c(LabelSize = 2, gridLabelSize = 2, size = 1))
-fits <- procGPA(land.adult)
-links <- c(1:7, 13:8)
-snd.links <- c(2, 8, 3, 9, 4, 10, 5, 11, 6, 12)
-land <- mshape.plot(fits, links = links, snd.links = snd.links)
-land <- land + coord_equal()
-morph <- putPlot(morph, land, 1, 3)
-pdf(file = '../doc/figure/pca_res.pdf')
-print(morph)
-dev.off()
-
-
-# variation along first two PCs
-mt <- mshape(land.adult)
-pc1.max <- which.max(adult$PC1)
-pc1.min <- which.min(adult$PC1)
-pc2.max <- which.max(adult$PC2)
-pc2.min <- which.min(adult$PC2)
-
-ex.lab <- function(var, value) {
-  value <- as.character(value)
-  if(var == 'lab') {
-    value[value == '1'] <- 'min'
-    value[value == '2'] <- 'mean'
-    value[value == '3'] <- 'max'
-  }
-  return(value)
-}
-
-inits <- list(land.adult[, , pc1.min], land.adult[, , pc1.max],
-              mt,
-              land.adult[, , pc2.min], land.adult[, , pc2.max], 
-              mt)
-secs <- inits <- lapply(inits, as.data.frame)
-inits <- lapply(inits, function(x) x[links, ])
-inits <- lapply(inits, function(x) cbind(x, rbind(x[-1, ], x[1, ])))
-inits <- lapply(inits, function(x) {
-                names(x) <- c('V1', 'V2', 'V3', 'V4')
-                x})
-inits <- Reduce(rbind, inits)
-pcl <- unlist(lapply(c('PC1', 'PC2'), function(x) rep(x, 3 * nrow(mt))))
[…]

### 2015-09-08 doc/draft_turtle.tex
added seven graph to document

@@
 In contrast with the above results, the analysis of the seven morphologically-distinct species demonstrates near perfect classification of the both the in-sample and out-of-sample datasets (Table \ref{tab:second_res}).
+Additionally, the distribution of ROC scores from 1000 bootstrap replicates are tightly clustered near ROC = 1 (Fig.
+\ref{fig:seven_boot}), which is contrast to the results from the \textit{E.
+marmorata} case (Fig.
+\ref{fig:gen_hist}).
 These results demonstrate that when there are strong distinctions between the states of the classification schemes, the methods used here can recover them.
@@
 
+\begin{figure}[ht]
+  \centering
+  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/seven_boot}
+  \caption{Distribution of 1000 bootstrap replicates of the out-of-sample AUC values from the secondary, multi-species dataset.
+The AUC values are from the random forest (RF), linear discriminate analysis (LDA), and multinomial logistic regression (MLR).
+AUC values range between 0.5 (no better than random) and 1 (perfect classification).}
+  \label{fig:seven_boot}
+\end{figure}
+
 \end{document}

### 2015-09-10 doc/draft_turtle.tex
submitted to syst biol

@@
 \begin{abstract}
-  We investigate the morphometric identification of cryptic species using machine-learning approaches by examining their implications for a recently proposed cryptic turtle species (\textit{Emys pallida}).
+  We investigate the morphometric identification of cryptic species using machine learning approaches by examining their implications for a recently proposed cryptic turtle species (\textit{Emys pallida}).
 We collected landmark-based morphometric data from 354 adult \textit{E.
-marmorata/ ``pallida''} museum specimens.
+marmorata/``pallida''} museum specimens.
 We assigned a classification to each specimen for six different binning schemes based on geographic occurrence data recorded in museum collection archives.
@@
 The results of this study indicate that there is no clear grouping of \textit{E.
-marmorata/ ``pallida''} based on plastron shape.
+marmorata/``pallida''} based on plastron shape.
 In contrast, the analysis of the ``clear-cut'' example demonstrates a near perfect classification, which demonstrates that the methods can recover correct results when an appropriate signal exists.
@@
 
-Here, we investigate the morphometric identification of cryptic species using machine-learning approaches.
+Here, we investigate the morphometric identification of cryptic species using machine learning approaches.
 In particular, we ask whether it is possible to determine which amongst a set of classification hypotheses is best and examine the implications of the results for a recently proposed set of cryptic turtle species.
@@
 \clearpage
-\bibliographystyle{sysbio}
-\bibliography{turtle,packages}
+\begin{thebibliography}{77}
+\providecommand{\natexlab}[1]{#1}
+\providecommand{\selectlanguage}[1]{\relax}
+\providecommand{\bibAnnoteFile}[1]{%
+  \IfFileExists{#1}{\begin{quotation}\noindent\textsc{Key:} #1\\
+  \textsc{Annotation:}\ \input{#1}\end{quotation}}{}}
+\providecommand{\bibAnnote}[2]{%
+  \begin{quotation}\noindent\textsc{Key:} #1\\
+  \textsc{Annotation:}\ #2\end{quotation}}
+
+\bibitem[{Angielczyk and Feldman(2013)}]{Angielczyk2013a}
+Angielczyk, K.~D. and C.~R. Feldman. 2013. {Are diminutive turtles
+  miniaturized? The ontogeny of plastron shape in emydine turtles}. Biological
+  Journal of the Linnean Society 108:727--755.
+\bibAnnoteFile{Angielczyk2013a}
+
+\bibitem[{Angielczyk et~al.(2011)Angielczyk, Feldman, and
+  Miller}]{Angielczyk2011}
+Angielczyk, K.~D., C.~R. Feldman, and G.~R. Miller. 2011. {Adaptive evolution
+  of plastron shape in emydine turtles.} Evolution 65:377--394.
+\bibAnnoteFile{Angielczyk2011}
+
+\bibitem[{Angielczyk and Sheets(2007)}]{Angielczyk2007}
+Angielczyk, K.~D. and H.~D. Sheets. 2007. {Investigation of simulated tectonic
+  deformation in fossils using geometric morphometrics}. Paleobiology
+  33:125--148.
+\bibAnnoteFile{Angielczyk2007}
+
+\bibitem[{Baird and Girard(1852)}]{Baird1852}
+Baird, S.~F. and C.~Girard. 1852. {Descriptions of new species of reptiles
+  collected by the U.S. Exploring Expedition under the command of Capt. Charles
+  Wilkes}. Proceedings of the National Academy of Sciences Philadelphia
+  6:174--177.
+\bibAnnoteFile{Baird1852}
+
+\bibitem[{Bauer et~al.(2000)Bauer, Parham, Brown, Stuart, Grismer, Papenfuss,
+  Bohme, Savage, Carranza, Grismer, Wagner, Schmitz, Ananjeva, and
+  Inger}]{Bauer2000}
+Bauer, A.~M., J.~F. Parham, R.~M. Brown, B.~L. Stuart, L.~Grismer, T.~J.
+  Papenfuss, W.~Bohme, J.~M. Savage, S.~Carranza, J.~L. Grismer, P.~Wagner,
+  A.~Schmitz, N.~B. Ananjeva, and R.~F. Inger. 2000. {Availability of new
+  Baysian-delimited gecko names and the importance of character-based species
+  descriptions}. Proceedings of the Royal Society B: Biological Sciences
+  278:490--492.
+\bibAnnoteFile{Bauer2000}
+
+\bibitem[{Baylac et~al.(2003)Baylac, Villemant, and Simbolotti}]{Baylac2003}
+Baylac, M., C.~Villemant, and G.~Simbolotti. 2003. {Combining geometric
+  morphometrics with pattern recognition for the investigation of species
+  complexes}. Biological Journal of the Linnean Society 80:89--98.
+\bibAnnoteFile{Baylac2003}
+
+\bibitem[{Bickford et~al.(2007)Bickford, Lohman, Sodhi, Ng, Meier, Winker,
+  Ingram, and Das}]{Bickford2007}
+Bickford, D., D.~J. Lohman, N.~S. Sodhi, P.~K.~L. Ng, R.~Meier, K.~Winker,
+  K.~K. Ingram, and I.~Das. 2007. {Cryptic species as a window on diversity and
+  conservation.} Trends in ecology \& evolution 22:148--55.
+\bibAnnoteFile{Bickford2007}
+
+\bibitem[{Breiman(2001)}]{Breiman2001}
+Breiman, L. 2001. {Random Forests}. Machine Learning 45:5--32.
+\bibAnnoteFile{Breiman2001}
+
+\bibitem[{Breiman et~al.(1984)Breiman, Friedman, Stone, and
+  Olshen}]{Breiman1984}
+Breiman, L., J.~Friedman, C.~J. Stone, and R.~A. Olshen. 1984. {Classification
+  and regression trees}. Wadsworth International Group, Belmont.
+\bibAnnoteFile{Breiman1984}
+
+\bibitem[{Bury et~al.(2010)Bury, Germano, and Bury}]{Bury2010}
+Bury, R.~B., D.~J. Germano, and G.~W. Bury. 2010. {Population Structure and
+  Growth of the Turtle Actinemys marmorata from the Klamath–Siskiyou
+  Ecoregion: Age, Not Size, Matters}. Copeia 2010:443--451.
+\bibAnnoteFile{Bury2010}
+
+\bibitem[{Buskirk et~al.(2005)Buskirk, Parham, and Feldman}]{Buskirk2005}
+Buskirk, S.~W., J.~F. Parham, and C.~R. Feldman. 2005. {On the hybridisation
+  between two distantly related Asian turtles (Testudines: Scalia x Mauremys)}.
+  Salamandra 41:21--26.
+\bibAnnoteFile{Buskirk2005}
+
+\bibitem[{Cardini et~al.(2009)Cardini, Nagorsen, O'Higgins, Polly, {Thorington
+  Jr}, and Tongiorgi}]{Cardini2009a}
+Cardini, A., D.~Nagorsen, P.~O'Higgins, P.~D. Polly, R.~W. {Thorington Jr}, and
+  P.~Tongiorgi. 2009. {Detecting biological distinctiveness using geometric
+  morphometrics: an example case from the Vancouver Island marmot}. Ethology
+  Ecology \& Evolution 21:209--223.
+\bibAnnoteFile{Cardini2009a}
+
+\bibitem[{Carstens and Dewey(2010)}]{Carstens2010}
+Carstens, B.~C. and T.~A. Dewey. 2010. {Species Delimitation Using a Combined
+  Coalescent and Information-Theoretic Approach: An Example from North American
+  Myotis Bats}. Systematic Biology 59:400--414.
+\bibAnnoteFile{Carstens2010}
+
+\bibitem[{Carstens et~al.(2013)Carstens, Pelletier, Reid, and
+  Satler}]{Carstens2013}
+Carstens, B.~C., T.~a. Pelletier, N.~M. Reid, and J.~D. Satler. 2013. {How to
+  fail at species delimitation.} Molecular ecology 22:4369--83.
+\bibAnnoteFile{Carstens2013}
+
+\bibitem[{Caumul and Polly(2005)}]{Caumul2005a}
+Caumul, R. and P.~D. Polly. 2005. {Phylogenetic and environmental components of
+  morphological variation: skull, mandible, and molar shape in marmots
+  (Marmota, Rodentia).} Evolution; international journal of organic evolution
+  59:2460--72.
+\bibAnnoteFile{Caumul2005a}
+
+\bibitem[{Clare(2011)}]{Clare2011}
+Clare, E.~L. 2011. {Cryptic species? Patterns of maternal and paternal gene
+  flow in eight neotropical bats.} PloS one 6:e21460.
+\bibAnnoteFile{Clare2011}
+
+\bibitem[{Dayrat(2005)}]{Dayrat2005a}
+Dayrat, B. 2005. {Towards integrative taxonomy}. Biological Journal of the
+  Linnean Society 85:407--415.
+\bibAnnoteFile{Dayrat2005a}
+
+\bibitem[{{De Queiroz}(2007)}]{DeQueiroz2007b}
+{De Queiroz}, K. 2007. {Species concepts and species delimitation.} Systematic
+  Biology 56:879--86.
+\bibAnnoteFile{DeQueiroz2007b}
+
+\bibitem[{Demandt and Bergek(2009)}]{Demandt2009}
+Demandt, M.~H. and S.~Bergek. 2009. {Identification of cyprinid hybrids by
+  using geometric morphometrics and microsatellites}. Journal of Applied
+  Ichthyology 25:695--701.
+\bibAnnoteFile{Demandt2009}
+
+\bibitem[{Dobigny et~al.(2003)Dobigny, Granjon, Aniskin, Ba, and
+  Voloboulev}]{Dobigny2003}
+Dobigny, G., L.~Granjon, V.~Aniskin, K.~Ba, and V.~Voloboulev. 2003. {A new
+  sigling species of Taterillus (Muridae, Gerbillinae) from West Agrica}.
+  Mammalian Biology 68:299--316.
+\bibAnnoteFile{Dobigny2003}
+
+\bibitem[{Dryden(2013)}]{Dryden2013}
+Dryden, I.~L. 2013. shapes: Statistical shape analysis. R package version
+  1.1-8.
+\bibAnnoteFile{Dryden2013}
+
+\bibitem[{Dryden and Mardia(1998)}]{Dryden1998a}
+Dryden, I.~L. and K.~Y. Mardia. 1998. {Statistical shape analysis}. Wiley, New
+  York.
+\bibAnnoteFile{Dryden1998a}
+
+\bibitem[{Eldredge and Gould(1972)}]{Eldredge1972}
+Eldredge, N. and S.~J. Gould. 1972. {Punctuated equilibria: an alternative to
+  phyletic gradualism}. Pages~82--115 \emph{in} Models in Paleobiology
+  (T.~J.~M. Schopf, ed.). Freeman Cooper, San Francisco.
+\bibAnnoteFile{Eldredge1972}
+
+\bibitem[{Feldman and Parham(2002)}]{Feldman2002}
+Feldman, C.~R. and J.~F. Parham. 2002. {Molecular phylogenetics of emydine
+  turtles: taxonomic revision and the evolution of shell kinesis.} Molecular
+  Phylogenetics and Evolution 22:388--98.
+\bibAnnoteFile{Feldman2002}
+
+\bibitem[{Francoy et~al.(2009)Francoy, Silva, Nunes-Silva, Menezes, and
+  Imperatriz-Fonseca}]{Francoy2009}
+Francoy, T.~M., R.~A.~O. Silva, P.~Nunes-Silva, C.~Menezes, and V.~L.
+  Imperatriz-Fonseca. 2009. {Gender identification of five genera of stingless
+  bees (Apidae, Meliponini) based on wing morphology}. Genetics and molecular
+  research 8:207--214.
+\bibAnnoteFile{Francoy2009}
+
+\bibitem[{Funk et~al.(2012)Funk, Caminer, and Ron}]{Funk2012}
+Funk, W.~C., M.~Caminer, and S.~R. Ron. 2012. {High levels of cryptic species
+  diversity uncovered in Amazonian frogs.} Proceedings of the Royal Society B:
+  Biological Sciences 279:1806--14.
+\bibAnnoteFile{Funk2012}
+
+\bibitem[{Gaubert et~al.(2005)Gaubert, Taylor, Fernandes, Bruford, and
+  Veron}]{Gaubert2005b}
+Gaubert, P., P.~J. Taylor, C.~a. Fernandes, M.~W. Bruford, and G.~Veron. 2005.
+  {Patterns of cryptic hybridization revealed using an integrative approach: a
+  case study on genets (Carnivora, Viverridae, Genetta spp.) from the southern
+  African subregion}. Biological Journal of the Linnean Society 86:11--33.
+\bibAnnoteFile{Gaubert2005b}
+
+\bibitem[{Germano and Bury(2009)}]{Germano2009}
+Germano, D.~J. and R.~B. Bury. 2009. {Variation in body size, growth, and
+  population structure of Actinemys marmorata from lentic and lotic habitats in
+  Southern Oregon}. Journal of Herpetology 43:510--520.
+\bibAnnoteFile{Germano2009}
+
+\bibitem[{Germano and Rathbun(2008)}]{Germano2008}
+Germano, D.~J. and G.~B. Rathbun. 2008. {Growth, population structure, and
+  reproduction of western pond turtles (Actinemys marmorata) on the Central
+  Coast of California}. Chelonian Conservation and Biology 7:188--194.
+\bibAnnoteFile{Germano2008}
+
+\bibitem[{Gould and Eldredge(1977)}]{Gould1977a}
+Gould, S.~J. and N.~Eldredge. 1977. {Punctuated equilibria: the tempo and mode
+  of evolution reconsidered}. Paleobiology 3:115--151.
+\bibAnnoteFile{Gould1977a}
+
+\bibitem[{G\"{u}nd\"{u}z et~al.(2007)G\"{u}nd\"{u}z, Jaarola, Tez, Yeniyurt,
+  Polly, and Searle}]{Gunduz2007}
+G\"{u}nd\"{u}z, I., M.~Jaarola, C.~Tez, C.~Yeniyurt, P.~D. Polly, and J.~B.
+  Searle. 2007. {Multigenic and morphometric differentiation of ground
+  squirrels (Spermophilus, Scuiridae, Rodentia) in Turkey, with a description
+  of a new species.} Molecular phylogenetics and evolution 43:916--35.
+\bibAnnoteFile{Gunduz2007}
+
+\bibitem[{Hand and Till(2001)}]{Hand2001}
+Hand, D.~J. and R.~J. Till. 2001. {A Simple Generalisation of the Area Under
+  the ROC Curve for Multiple Class Classification Problems}. Machine Learning
+  45:171--186.
+\bibAnnoteFile{Hand2001}
+
+\bibitem[{Hastie et~al.(2009)Hastie, Tibshirani, and Friedman}]{Hastie2009}
+Hastie, T., R.~Tibshirani, and J.~Friedman. 2009. {The elements of statistical
+  learning: data mining, inference, and prediction}. 2nd ed. Springer, New
+  York.
+\bibAnnoteFile{Hastie2009}
+
+\bibitem[{Hausdorf and Hennig(2010)}]{Hausdorf2010}
+Hausdorf, B. and C.~Hennig. 2010. {Species delimitation using dominant and
+  codominant multilocus markers.} Systematic biology 59:491--503.
+\bibAnnoteFile{Hausdorf2010}
+
+\bibitem[{Holland(1992)}]{Holland1992}
+Holland, D.~C. 1992. {Level and pattern in morphological variation: a
+  phylogeographic study of the western pond turtle (Clemmys marmorata)}. Ph.D.
+  thesis University of Southwestern Louisiana.
+\bibAnnoteFile{Holland1992}
+
+\bibitem[{Huelsenbeck et~al.(2011)Huelsenbeck, Andolfatto, and
+  Huelsenbeck}]{Huelsenbeck2011b}
+Huelsenbeck, J.~P., P.~Andolfatto, and E.~T. Huelsenbeck. 2011. {Structurama:
+  bayesian inference of population structure.} Evolutionary bioinformatics
+  online 7:55--9.
+\bibAnnoteFile{Huelsenbeck2011b}
+
+\bibitem[{Hunt(2008)}]{Hunt2008}
+Hunt, G. 2008. {Gradual or pulsed evolution: when should punctuational
+  explanations be preferred?} Paleobiology 34:360--377.
+\bibAnnoteFile{Hunt2008}
+
+\bibitem[{Kaufman and Rousseeuw(1990)}]{Kaufman1990}
+Kaufman, L. and P.~J. Rousseeuw. 1990. {Finding groups in data : an
+  introduction to cluster analysis}. Wiley, New York.
+\bibAnnoteFile{Kaufman1990}
+
+\bibitem[{Kendall(1977)}]{Kendall1977a}
+Kendall, D.~G. 1977. {The diffusion of shape}. Advances in Applied Probability
+  9:428--430.
+\bibAnnoteFile{Kendall1977a}
+
+\bibitem[{Kendall(1984)}]{Kendall1984a}
+Kendall, D.~G. 1984. {Shape Manifolds, Procrustean Metrics, and Complex
+  Projective Spaces}. Bulletin of the London Mathematical Society 16:81--121.
+\bibAnnoteFile{Kendall1984a}
+
+\bibitem[{Klingenberg et~al.(2002)Klingenberg, Barluenga, and
+  Meyer}]{Klingenberg2002}
+Klingenberg, C.~P., M.~Barluenga, and A.~Meyer. 2002. {Shape analysis of
+  symetric structures: quantifying variation among individuals and asymmetry}.
+  Evolution 56:1909--1920.
+\bibAnnoteFile{Klingenberg2002}
+
+\bibitem[{Kuhn(2013)}]{KuhnMAN2013}
+Kuhn, M. 2013. caret: Classification and Regression Training. R package version
+  5.15-61.
+\bibAnnoteFile{KuhnMAN2013}
+
+\bibitem[{Kuhn and Johnson(2013)}]{Kuhn2013}
+Kuhn, M. and K.~Johnson. 2013. {Applied predictive modeling}. Springer, New
+  York, NY.
+\bibAnnoteFile{Kuhn2013}
+
+\bibitem[{Leach\'{e} and Fujita(2010)}]{Leache2010}
+Leach\'{e}, A.~D. and M.~K. Fujita. 2010. {Bayesian species delimitation in
+  West African forest geckos (Hemidactylus fasciatus).} Proceedings. Biological
+  sciences / The Royal Society 277:3071--7.
+\bibAnnoteFile{Leache2010}
+
+\bibitem[{Lubcke and Wilson(2007)}]{Lubcke2007}
+Lubcke, G.~M. and D.~S. Wilson. 2007. {Variation in shell morphology of the
+  Western Pond Turtle (Actinemys marmorata Baird and Giarard) from three
+  aquativ habitats in Northern California}. Journal of Herpetology 41:107--114.
+\bibAnnoteFile{Lubcke2007}
+
+\bibitem[{MacLeod(2007)}]{MacLeod2007}
+MacLeod, N. 2007. {Automated taxon identification in systematics: theory,
+  approaches and applications}. CRC Press, Boca Raton.
+\bibAnnoteFile{MacLeod2007}
+
+\bibitem[{Maechler et~al.(2013)Maechler, Rousseeuw, Struyf, Hubert, and
+  Hornik}]{Maechler2013}
+Maechler, M., P.~Rousseeuw, A.~Struyf, M.~Hubert, and K.~Hornik. 2013. cluster:
+  Cluster Analysis Basics and Extensions. R package version 1.14.4.
+\bibAnnoteFile{Maechler2013}
+
+\bibitem[{Mitrovski-Bogdanovic et~al.(2013)Mitrovski-Bogdanovic, Petrovic,
+  Mitrovic, Ivanovic, \v{Z}ikic, Star\'{y}, Vorburger, and
+  Tomanovic}]{MitrovskiBogdanovic2013}
+Mitrovski-Bogdanovic, A., A.~Petrovic, M.~Mitrovic, A.~Ivanovic, V.~\v{Z}ikic,
+  P.~Star\'{y}, C.~Vorburger, and v.~Tomanovic. 2013. {Identification of two
+  cryptic species within the Praon abjectum group (Hymenoptera: Braconidae:
+  Aphidiinae) using molecular markers and geometric morphometrics}. Annals of
+  the entomological society of America 106:170--180.
+\bibAnnoteFile{MitrovskiBogdanovic2013}
+
+\bibitem[{Mitteroecker and Bookstein(2011)}]{Mitteroecker2011}
+Mitteroecker, P. and F.~Bookstein. 2011. {Linear Discrimination, Ordination,
+  and the Visualization of Selection Gradients in Modern Morphometrics}.
+  Evolutionary Biology 38:100--114.
+\bibAnnoteFile{Mitteroecker2011}
+
+\bibitem[{O'Meara(2010)}]{O'Meara2010}
+O'Meara, B.~C. 2010. {New heuristic methods for joint species delimitation and
+  species tree inference.} Systematic biology 59:59--73.
+\bibAnnoteFile{O'Meara2010}
+
+\bibitem[{Padial et~al.(2010)Padial, Miralles, {De la Riva}, and
+  Vences}]{Padial2010}
+Padial, J.~M., A.~Miralles, I.~{De la Riva}, and M.~Vences. 2010. {The
+  integrative future of taxonomy}. Frontiers in Zoology 7:1--14.
+\bibAnnoteFile{Padial2010}
+
+\bibitem[{Parham et~al.(2013)Parham, Papenfuss, Dijk, Wilson, Marte, Schettino,
+  and {Brian Simison}}]{Parham2013}
+Parham, J.~F., T.~J. Papenfuss, P.~P.~V. Dijk, B.~S. Wilson, C.~Marte, L.~R.
+  Schettino, and W.~{Brian Simison}. 2013. {Genetic introgression and
+  hybridization in Antillean freshwater turtles (Trachemys) revealed by
+  coalescent analyses of mitochondrial and cloned nuclear markers.} Molecular
+  phylogenetics and evolution 67:176--87.
+\bibAnnoteFile{Parham2013}
+
+\bibitem[{Pfenninger and Schwenk(2007)}]{Pfenninger2007}
+Pfenninger, M. and K.~Schwenk. 2007. {Cryptic animal species are homogeneously
+  distributed among taxa and biogeographical regions.} BMC evolutionary biology
+  7:121.
+\bibAnnoteFile{Pfenninger2007}
+
+\bibitem[{Polly(2003)}]{Polly2003}
+Polly, P.~D. 2003. {Paleophylogeography of Sorex araneus: molar shape as a
+  morphological marker for fossil shrews}. Mammalia 68:233--243.
+\bibAnnoteFile{Polly2003}
+
+\bibitem[{Polly(2007)}]{Polly2007a}
+Polly, P.~D. 2007. {Phylogeographic differentiation in Sorex araneus:
+  morphology in relation to geography and karyotype}. Russian Journal of
+  Theriology 6:73--84.
+\bibAnnoteFile{Polly2007a}
+
+\bibitem[{Polly et~al.()Polly, Stayton, Dumont, Pierce, Rayfield, and
+  Angielczyk}]{Pollyb}
+Polly, P.~D., C.~T. Stayton, E.~R. Dumont, S.~E. Pierce, E.~J. Rayfield, and
+  K.~D. Angielczyk. ???? {Combining geometric morphometrics and finite element
+  analysis with evolutionary modeling: towards a synthesis}. Journal of
+  Vertebrate Paleontology .
+\bibAnnoteFile{Pollyb}
+
+\bibitem[{Pons et~al.(2006)Pons, Barraclough, Gomez-Zurita, Cardoso, Duran,
+  Hazell, Kamoun, Sumlin, and Vogler}]{Pons2006}
+Pons, J., T.~Barraclough, J.~Gomez-Zurita, A.~Cardoso, D.~Duran, S.~Hazell,
+  S.~Kamoun, W.~Sumlin, and A.~Vogler. 2006. {Sequence-Based Species
+  Delimitation for the DNA Taxonomy of Undescribed Insects}. Systematic Biology
+  55:595--609.
+\bibAnnoteFile{Pons2006}
+
+\bibitem[{{R Core Team}(2013)}]{2013}
+{R Core Team}. 2013. R: A Language and Environment for Statistical Computing. R
+  Foundation for Statistical Computing Vienna, Austria.
+\bibAnnoteFile{2013}
+
+\bibitem[{Rivera(2008)}]{Rivera2008}
+Rivera, G. 2008. {Ecomorphological variation in shell shape of the freshwater
+  turtle Pseudemys concinna inhabiting different aquatic flow regimes.}
+  Integrative and comparative biology 48:769--87.
+\bibAnnoteFile{Rivera2008}
+
+\bibitem[{Rivera et~al.(2014)Rivera, Davis, Godwin, and Adams}]{Rivera2014}
+Rivera, G., J.~N. Davis, J.~C. Godwin, and D.~C. Adams. 2014. {Repeatability of
+  Habitat-Associated Divergence in Shell Shape of Turtles}. Evolutionary
+  Biology Pages~29--37.
+\bibAnnoteFile{Rivera2014}
+
+\bibitem[{Rivera and Stayton(2011)}]{Rivera2011}
+Rivera, G. and C.~T. Stayton. 2011. {Finite element modeling of shell shape in
+  the freshwater turtle Pseudemys concinna reveals a trade-off between
+  mechanical strength and hydrodynamic efficiency.} Journal of morphology
+  272:1192--203.
+\bibAnnoteFile{Rivera2011}
+
+\bibitem[{Rohlf(2005)}]{Rohlf2005}
+Rohlf, F.~J. 2005. {TpsDig 2.04}.
+\bibAnnoteFile{Rohlf2005}
+
+\bibitem[{Schilck-Steiner et~al.(2007)Schilck-Steiner, Seifert, Stauffer,
+  Christian, Crozier, and Steiner}]{SchlickSteiner2007}
+Schilck-Steiner, B.~C., B.~Seifert, C.~Stauffer, E.~Christian, R.~H. Crozier,
+  and F.~M. Steiner. 2007. {Without morphology, cryptic species stay in
+  taxonomic crypsis following discovery}. Trends in ecology \& evolution
+  22:391--392.
+\bibAnnoteFile{SchlickSteiner2007}
+
+\bibitem[{Seeliger(1945)}]{Seeliger1945}
+Seeliger, L.~M. 1945. {Variation in the Pacific Mud Turtle}. Copeia
+  1945:150--159.
+\bibAnnoteFile{Seeliger1945}
+
+\bibitem[{Spinks and Shaffer(2005)}]{Spinks2005}
+Spinks, P.~Q. and H.~B. Shaffer. 2005. {Range-wide molecular analysis of the
+  western pond turtle (Emys marmorata): cryptic variation, isolation by
+  distance, and their conservation implications.} Molecular ecology
+  14:2047--64.
+\bibAnnoteFile{Spinks2005}
+
+\bibitem[{Spinks and Shaffer(2009)}]{Spinks2009}
+Spinks, P.~Q. and H.~B. Shaffer. 2009. {Conflicting mitochondrial and nuclear
+  phylogenies for the widely disjunct Emys (Testudines: Emydidae) species
+  complex, and what they tell us about biogeography and hybridization.}
+  Systematic biology 58:1--20.
+\bibAnnoteFile{Spinks2009}
+
+\bibitem[{Spinks et~al.(2014)Spinks, Thomson, and {Bradley
+  Shaffer}}]{Spinks2014}
+Spinks, P.~Q., R.~C. Thomson, and H.~{Bradley Shaffer}. 2014. {The advantages
+  of going large: genome wide SNPs clarify the complex population history and
+  systematics of the threatened western pond turtle}. Molecular Ecology
+  Pages~n/a--n/a.
+\bibAnnoteFile{Spinks2014}
+
+\bibitem[{Spinks et~al.(2010)Spinks, Thomson, and Shaffer}]{Spinks2010}
+Spinks, P.~Q., R.~C. Thomson, and H.~B. Shaffer. 2010. {Nuclear gene
+  phylogeography reveals the historical legacy of an ancient inland sea on
+  lineages of the western pond turtle, Emys marmorata in California.} Molecular
+  ecology 19:542--56.
+\bibAnnoteFile{Spinks2010}
+
+\bibitem[{Stayton(2011)}]{Stayton2011}
+Stayton, C.~T. 2011. {Biomechanics on the half shell: functional performance
+  influences patterns of morphological variation in the emydid turtle
+  carapace.} Zoology (Jena, Germany) 114:213--23.
+\bibAnnoteFile{Stayton2011}
+
+\bibitem[{Stuart et~al.(2006)Stuart, Inger, and Voris}]{Stuart2006}
+Stuart, B.~L., R.~F. Inger, and H.~K. Voris. 2006. {High level of cryptic
+  species diversity revealed by sympatric lineages of Southeast Asian forest
+  frogs.} Biology letters 2:470--4.
+\bibAnnoteFile{Stuart2006}
+
+\bibitem[{Sztencel-Jabłonka et~al.(2009)Sztencel-Jabłonka, Jones, and
+  BogdanowicZ}]{Sztencel-Jabonka2009}
+Sztencel-Jabłonka, A., G.~Jones, and W.~BogdanowicZ. 2009. {Skull Morphology
+  of Two Cryptic Bat Species: Pipistrellus pipistrellus and P. pygmaeus — A
+  3D Geometric Morphometrics Approach with Landmark Reconstruction}. Acta
+  Chiropterologica 11:113--126.
+\bibAnnoteFile{Sztencel-Jabonka2009}
+
+\bibitem[{Tibshirani et~al.(2001)Tibshirani, Walther, and
+  Hastie}]{Tibshirani2001a}
+Tibshirani, R., G.~Walther, and T.~Hastie. 2001. {Estimating the number of
+  clusters in a data set via the gap statistic}. Journal of the Royal
+  Statistical Society: Series B (Statistical Methodology) 63:411--423.
+\bibAnnoteFile{Tibshirani2001a}
+
+\bibitem[{{Van Bocxlaer} and Hunt(2013)}]{VanBocxlaer2013}
+{Van Bocxlaer}, B. and G.~Hunt. 2013. {Morphological stasis in an ongoing
+  gastropod radiation from Lake Malawi}. Proceedings of the National Academy of
+  Sciences .
+\bibAnnoteFile{VanBocxlaer2013}
+
+\bibitem[{van~den Brink and Bokma(2011)}]{VandenBrink2011}
+van~den Brink, V. and F.~Bokma. 2011. {Morphometric shape analysis using
+  learning vector quantization neural networks — an example distinguishing
+  two microtine vole species}. Annales Zoologici Fennici 48:359--364.
+\bibAnnoteFile{VandenBrink2011}
+
+\bibitem[{Venables and Ripley(2002)}]{Venables2002a}
+Venables, W. and B.~D. Ripley. 2002. {Modern applied statistics with S}. 4th
+  ed. Springer, New York.
+\bibAnnoteFile{Venables2002a}
+
+\bibitem[{Yang and Rannala(2010)}]{Yang2010b}
+Yang, Z. and B.~Rannala. 2010. {Bayesian species delimitation using multilocus
+  sequence data.} Proceedings of the National Academy of Sciences 107:9264--9.
+\bibAnnoteFile{Yang2010b}
+
+\bibitem[{Zelditch et~al.(2004)Zelditch, Swiderski, and Sheets}]{Zelditch2004}
+Zelditch, M.~L., D.~L. Swiderski, and H.~D. Sheets. 2004. {Geometric
+  morphometrics for biologists: a primer}. Elsevier Academic Press, Amsterdam.
+\bibAnnoteFile{Zelditch2004}
+
+\end{thebibliography}
 
@@
   \centering
-  \include{comp_tab}
   \caption{AUC values for the best model of each classification scheme for both the observed (training) data and the generalized (testing) data.
@@
 }
+%  \begin{tabular}{ l | c c | c c | c c }
+%    \hline
+%    & \multicolumn{2}{| l |}{random forest} & 
+%    \multicolumn{2}{| l |}{multinomial logistic regression} & 
+%    \multicolumn{2}{| l}{linear discriminate analysis} \\
+%    Scheme & Observed & Generalized & Observed & Generalized & Observed & Geneeralized \\ 
+%    \hline
+%    \hline
+%    Morph 1 & 0.63 & 0.73 & 0.75 & 0.79 & 0.75 & 0.80 \\ 
+%    Morph 2 & 0.61 & 0.58 & 0.76 & 0.77 & 0.76 & 0.77 \\ 
+%    Mito 1& 0.63 & 0.62 & 0.75 & 0.63 & 0.75 & 0.63 \\ 
+%    Mito 2 & 0.77 & 0.67 & 0.80 & 0.64 & 0.80 & 0.63 \\ 
+%    Mito 3 & 0.56 & 0.64 & 0.71 & 0.74 & 0.71 & 0.73 \\ 
+%    Nuclear & 0.56 & 0.67 & 0.74 & 0.62 & 0.74 & 0.77 \\ 
+%    \hline
+%  \end{tabular}
   \label{tab:comp}
@@
 
-\begin{table}
+\begin{table}[ht]
   \centering
-  \begin{tabular}{ l r r r }
-    \hline
-    Scheme & RF & MLR & LDA \\
-    \hline
-    \hline
-    Morph 1 & \(\ast\) & \(\ast\) & 1 \\
-    Morph 2 & 0.79 & 0.55 & 1 \\
-    Mito 1 & 0.89 & 0.94 & 1 \\ 
-    Mito 2 & 0.82 & 0.57 & \(\ast\) \\ 
-    Mito 3 & 0.79 & 0.69 & 0.73 \\ 
-    Nuclear & 0.79 & 0.96 & 0.96 \\ 
-    \hline
-  \end{tabular}
   \caption{Results of bootstrap comparisons between the scheme with the highest mean AUC value and all other schemes, defined P(best - other \(>\) 0).
@@
 Probabilities are the percent of comparisons that are greater than the observed difference in means.}
+%  \begin{tabular}{ l r r r }
+%    \hline
+%    Scheme & RF & MLR & LDA \\
+%    \hline
+%    \hline
+%    Morph 1 & \(\ast\) & \(\ast\) & 1 \\
+%    Morph 2 & 0.79 & 0.55 & 1 \\
+%    Mito 1 & 0.89 & 0.94 & 1 \\ 
+%    Mito 2 & 0.82 & 0.57 & \(\ast\) \\ 
+%    Mito 3 & 0.79 & 0.69 & 0.73 \\ 
+%    Nuclear & 0.79 & 0.96 & 0.96 \\ 
+%    \hline
+%  \end{tabular}
   \label{tab:gen_tests}
@@
 
-\begin{table}
+\begin{table}[ht]
   \centering
-  \begin{tabular}{ l | l r r r }
-    \hline
-    Scheme & Class & RF & MLR & LDA \\
-    \hline
-    \hline
-    \multirow{2}{*}{Morph 1} & CCR & 0.77 & 1.00 & 1.00 \\ 
-    & marm & 0.87 & 0.93 & 1.00 \\ 
-    \hline
-    \multirow{2}{*}{Morph 2} & CCR & 0.88 & 1.00 & 1.00 \\ 
-    & marm & 1.00 & 1.00 & 0.98 \\ 
-    \hline
-    \multirow{2}{*}{Mito 1} & CCR & 0.94 & 0.99 & 0.98 \\ 
-    & marm & 0.99 & 0.98 & 0.47 \\ 
-    \hline
-    \multirow{2}{*}{Mito 2} & marm & 0.85 & 0.97 & 0.96 \\ 
-    & pall & 0.94 & 1.00 & 1.00 \\ 
-    \hline
-    \multirow{2}{*}{Mito 3} & 1 & 0.40 & 0.92 & 0.90 \\ 
-    & 3 & 0.97 & 0.99 & 1.00 \\ 
-    \hline
-    \multirow{2}{*}{Nuclear} & marm & 1.00 & 1.00 & 1.00 \\ 
-    & pall & 0.99 & 1.00 & 1.00 \\ 
-    \hline
-  \end{tabular}
   \caption{Results of comparisons between correctly and incorrectly classified observations from the testing data set, defined P(distance - simulated \(>\) 0).
@@
 This was done for each of the three modeling techniques included in this study: random forest (RF), multinomial logistic regression (MLR), and linear discriminate analysis (LDA).}
+%  \begin{tabular}{ l | l r r r }
+%    \hline
+%    Scheme & Class & RF & MLR & LDA \\
+%    \hline
+%    \hline
+%    \multirow{2}{*}{Morph 1} & CCR & 0.77 & 1.00 & 1.00 \\ 
+%    & marm & 0.87 & 0.93 & 1.00 \\ 
+%    \hline
+%    \multirow{2}{*}{Morph 2} & CCR & 0.88 & 1.00 & 1.00 \\ 
+%    & marm & 1.00 & 1.00 & 0.98 \\ 
+%    \hline
+%    \multirow{2}{*}{Mito 1} & CCR & 0.94 & 0.99 & 0.98 \\ 
+%    & marm & 0.99 & 0.98 & 0.47 \\ 
+%    \hline
+%    \multirow{2}{*}{Mito 2} & marm & 0.85 & 0.97 & 0.96 \\ 
+%    & pall & 0.94 & 1.00 & 1.00 \\ 
+%    \hline
+%    \multirow{2}{*}{Mito 3} & 1 & 0.40 & 0.92 & 0.90 \\ 
+%    & 3 & 0.97 & 0.99 & 1.00 \\ 
+%    \hline
+%    \multirow{2}{*}{Nuclear} & marm & 1.00 & 1.00 & 1.00 \\ 
+%    & pall & 0.99 & 1.00 & 1.00 \\ 
+%    \hline
+%  \end{tabular}
   \label{tab:miss_tests}
@@
 
+\begin{table}[ht]
+  \centering
+  \caption{Results from classification model estimates of the secondary, multi-species dataset.
+Models are random forest (RF), linear discriminate analysis (LDA), and multinomial logistic regression (MLR).
+Comparison of in-sample and out-of-sample AUC of the best performing model, along with the number of predictors.
+AUC values range between 0.5 (no better than random) and 1 (perfect classification).}
+%  \begin{tabular}{ l r r r }
+%    \hline
+%    & \# predictors & In-sample AUC & Out-of-sample AUC \\ 
+%    \hline
+%    \hline
+%    RF &   11 & 1.000 & 0.999 \\ 
+%    MLR &    6 & 1.000 & 0.998 \\ 
+%    LDA &   10 & 1.000 & 1.000 \\ 
+%    \hline
+%  \end{tabular}
+  \label{tab:second_res}
+\end{table}
+
+\clearpage
+
 \begin{figure}[ht]
   \centering
-  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/plastra}
+%  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/plastra}
   \caption{Depiction of general plastral shape of \textit{E. marmorata} and position of the 19 landmark used in this study. Anterior is towards the top of the figure.}
@@
   \centering
-  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/gap_res}
+%  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/gap_res}
   \caption{Results from PAM clustering of the Riemannian shape distance for 8 different number of clusters. Vertical lines are 1 standard deviation of the mean determined from 500 resamples.}
@@
   \centering
-  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/gap_map}
+%  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/gap_map}
   \caption{Comparison of geographic distribution of clustered observations from the 2 clustering PAM solution.
@@
   \centering
-  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/sel_val}
-  \caption{Graphical representation of the AUC values from model selection for multinomial logigistic regression and linear discriminate analysis, respecitively.
+%  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/sel_val}
+  \caption{Graphical representation of the AUC values from model selection for multinomial logistic regression and linear discriminate analysis, respecitively.
 AUC model selection is based on greatest AUC value.
 The horizontal axis corresponds to the cummulative number of axes included in the model of interest.
-A red dot corresponds to the AUC best model for that classification scheme.}
+A highlighted point corresponds to the AUC best model for that classification scheme.}
   \label{fig:sel}
@@
   \centering
-  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/var_imp}
+%  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/var_imp}
   \caption{Variable importance from the random forest models for each of the six classification schemes.
@@
   \centering
-  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/gen_res}
+%  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/gen_res}
   \caption{Bootstrap distributions for genearlized AUC values for each of the classification schemes.
@@
 
-\begin{table}[ht]
-  \centering
-  \begin{tabular}{ l r r r }
-    \hline
-    & \# predictors & In-sample AUC & Out-of-sample AUC \\ 
-    \hline
-    RF &   11 & 1.000 & 0.999 \\ 
-    MLR &    6 & 1.000 & 0.998 \\ 
-    LDA &   10 & 1.000 & 1.000 \\ 
-    \hline
-  \end{tabular}
-  \caption{Results from classification model estimates of the secondary, multi-species dataset.
-Models are random forest (RF), linear discriminate analysis (LDA), and multinomial logistic regression (MLR).
-Comparison of in-sample and out-of-sample AUC of the best performing model, along with the number of predictors.
-AUC values range between 0.5 (no better than random) and 1 (perfect classification).}
-  \label{tab:second_res}
-\end{table}
-
 \begin{figure}[ht]
   \centering
-  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/seven_boot}
+%  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/seven_boot}
   \caption{Distribution of 1000 bootstrap replicates of the out-of-sample AUC values from the secondary, multi-species dataset.

### 2015-09-10 doc/letter_editor.tex
submitted to syst biol

\documentclass{letter}
\usepackage{microtype}
%\usepackage{letterbib}
%\usepackage{natbib}
\frenchspacing

\signature{Peter D Smits}
\address{Committee on Evolutionary Biology \\ University of Chicago \\
1025 E. 57th Street \\ Culver Hall 402 \\ Chicago, IL 60637 \\
psmits@uchicago.edu}

\begin{document}
\begin{letter}{Editor \\ \textit{Systematic Biology}}
  \opening{Dear Editor,}

  Please find enclosed the manuscript entitled ``How cryptic is cryptic diversity? Machine learning approaches to classifying morphological variation in the Pacific Pond Turtle (\textit{Emys marmorata})'' which we are submitting for consideration at the journal \textit{Systematic Biology}. This manuscript is authored by myself (Peter D Smits), Kenneth D Angielczyk, and James F Parham. We believe our analysis and results are of interest to the community as we present a non-standard approach for evaluating the intensity of cryptic diversity and the efficacy of different classification schemes.

  Morphological and molecular attempts to classify the turtle species \textit{Emys marmorata} have produced three different potential sub-species or species divisions. In this study we compared these hypotheses of how to subdivide \textit{Emys marmorata} using multiple machine learning approaches. We analyzed over 500 specimens of this species from over its entire geographic range. Additionally, instead of classification accuracy, we use a metric of the ratio of true-positive and false-positive rates for both in-sample and out-of-sample model performance. Finally, in order to test whether our approaches were valid, we compared these results to those from the analysis of seven morphologically distinct turtle species.

  Our results demonstrate, that while our approaches can very precisely identify morphologically distinct groupings, the morphological diversity of the \textit{Emy marmorata} species ``complex'' is inconsistent with any of the proposed classification schemes. We propose that these results may be due to the possibility that genetic differentiation is not associated with plastron shape variation \textit{below} the species level and/or that local selective pressures (e.g., from hydrological regime) overwhelm morphological differentiation. Additionally, a reconsideration of the methods used to delimit \textit{E. ``pallida,''} the lack of barriers to gene flow, the strong evidence for widespread admixture between lineages, and the fact that plastron shape can be used to differentiate other emydine species suggest that its lack of diagnosability most likely reflects the non-distinctiveness of this proposed taxon.

  Thank you for considering our work. Please send all correspondence regarding this manuscript to me via my email address (psmits@uchicago.edu).

  \closing{Sincerely,}

  \encl{Article; supplementary text, figures, tables.}

\end{letter}
\end{document}

### 2015-09-10 doc/tabs.tex
submitted to syst biol

--- 2015-09-08 doc/draft_turtle.tex
@@
-\documentclass[12pt,letterpaper]{article}
+\documentclass{article}
 
@@
 
-\captionsetup[subfigure]{position = top, labelfont = bf, textfont = normalfont, singlelinecheck = off, justification = raggedright}
-
-\renewcommand{\section}[1]{%
-\bigskip
-\begin{center}
-\begin{Large}
-\normalfont\scshape #1
-\medskip
-\end{Large}
-\end{center}}
-
-\renewcommand{\subsection}[1]{%
-\bigskip
-\begin{center}
-\begin{large}
-\normalfont\itshape #1
-\end{large}
-\end{center}}
-
-\renewcommand{\subsubsection}[1]{%
-\vspace{2ex}
-\noindent
-\textit{#1.}---}
-
-\renewcommand{\tableofcontents}{}
-
-\bibpunct{(}{)}{;}{a}{}{,}  % this is a citation format command for natbib
-
-\title{How cryptic is cryptic diversity? Machine learning approaches to classifying morphological variation in the Pacific Pond Turtle (\textit{Emys marmorata})}
-\author[1]{Peter D Smits}%\thanks{psmits@uchicago.edu}}
-\author[1,2]{Kenneth D Angielczyk}%\thanks{kangielczyk@fieldmuseum.org}}
-\author[3]{James F Parham}%\thanks{jparham@fullerton.edu}}
-\affil[1]{Committee on Evolutionary Biology, University of Chicago}
-\affil[2]{Integrative Research Center, Field Museum of Natural History}
-\affil[3]{Department of Geological Sciences, California State University -- Fullerton}
-
 \begin{document}
-\maketitle
-\noindent{\textbf{Corresponding author:} Peter D Smits, Committee on Evolutionary Biology, University of Chicago, 1025 E.
-57th Street, Culver Hall 402, Chicago, IL, 60637, USA; E-mail: \href{mailto:psmits@uchicago.edu}{psmits@uchicago.edu}}
-
-\linenumbers
-\modulolinenumbers[2]
-
-\begin{abstract}
-  We investigate the morphometric identification of cryptic species using machine-learning approaches by examining their implications for a recently proposed cryptic turtle species (\textit{Emys pallida}).
-We collected landmark-based morphometric data from 354 adult \textit{E.
-marmorata/ ``pallida''} museum specimens.
-We assigned a classification to each specimen for six different binning schemes based on geographic occurrence data recorded in museum collection archives.
-We used multiple machine learning methods, both unsupervised and supervised, to compare different classification hypotheses and asked whether it is possible to determine which amongst a set of classification hypotheses is best.
-In addition, we applied the above approach to a “clear-cut” example of seven unambiguously distinct species closely related to \textit{E.
-marmorata}.
-The results of this study indicate that there is no clear grouping of \textit{E.
-marmorata/ ``pallida''} based on plastron shape.
-In contrast, the analysis of the ``clear-cut'' example demonstrates a near perfect classification, which demonstrates that the methods can recover correct results when an appropriate signal exists.
-Explanations for the lack of grouping in \textit{E.
-marmorata} include that possibility that genetic differentiation is not associated with plastron shape variation below the species level and/or that local selective pressures (e.g., from hydrological regime) overwhelm morphological differentiation.
-A reconsideration of the methods used to delimit \textit{E.
-``pallida,''} the lack of barriers to gene flow, the strong evidence for widespread admixture between lineages, and the fact that plastron shape can be used to differentiate other emydine species suggest that its lack of diagnosability most likely reflects the non-distinctiveness of this proposed taxon.
-\end{abstract}
-
-\section{Introduction}
-
-% cryptic diversity
-%   most species are still deliminated solely based on morphology
-Molecular systematics has repeatedly demonstrated the existence of cryptic species that can only be diagnosed using genetic data \citep{Stuart2006,Bickford2007,SchlickSteiner2007,Pfenninger2007,Clare2011,Funk2012}.
-In attempts to streamline the documentation of biodiversity, several methods of species delimitation that rely almost entirely on genetic data have recently been proposed \citep{Pons2006,Carstens2010,Hausdorf2010,O'Meara2010,Yang2010b,Huelsenbeck2011b}.
-Although strong caveats on the utility of these methods have been raised \citep{Bauer2000,Carstens2013}, they are nevertheless being used to name species \citep{Leache2010,Spinks2014}.
-
-The majority of extant taxa, and almost all extinct taxa, are delimited by morphology alone.
-This disjunction complicates interpretations of variation and diversity in deep time, as apparent morphological stasis may not reflect the true underlying diversity \citep{Eldredge1972,Gould1977a,Hunt2008,VanBocxlaer2013}.
-Similarly, for many museum specimens of extant taxa (e.g.
-those preserved in formalin), it is difficult to acquire the genetic data needed for non-morphological species delimitation methods.
-
-These considerations have sparked interest in whether geometric morphometric analyses can capture fine-scale variation that can be used for identifying cryptic species.
-This would make the task of idetifying and maintaining endagered or conserved groups much easier and could contribute to improved classifications of extinct taxa and populations.
-Most such studies focus on using morphometrics to discover differences between taxa that were identified by other means \citep{Polly2003,Zelditch2004,Gaubert2005b,Gunduz2007,Polly2007a,Demandt2009}.
-Additionally, there has been work on automated taxon identification and classification of taxa into groups \citep{Baylac2003,Dobigny2003,MacLeod2007,VandenBrink2011}.
-
-Here, we investigate the morphometric identification of cryptic species using machine-learning approaches.
-In particular, we ask whether it is possible to determine which amongst a set of classification hypotheses is best and examine the implications of the results for a recently proposed set of cryptic turtle species.
-
-% past work on automatic taxon identification and older approaches to classifying taxa
-% why use machine learning methods
-\subsection{Background and study system}
-Machine learning is an extension of known statistical methodology \citep{Hastie2009} that emphasizes high predictive accuracy and generality at the expense of the interpretability of individual parameters.
-The basic statistical mechanics are supplemented by randomization, sorting, and partitioning algorithms, along with the maximization or minimization of summary statistics, in order to best estimate a general model for all data, both sampled and unsampled \citep{Hastie2009}.
-Machine learning approaches have found use in medical research, epidemiology, economics and automated image identification such as handwritten zip codes \citep{Hastie2009}.
-The two major classes of machine learning methods are unsupervised and supervised learning.
-Unsupervised learning methods are used with unlabeled data where the underlying structure is estimated, and they are analogous to clustering and density estimation methods \citep{Kaufman1990}.
-Supervised learning methods are used with labeled data where the final output of data is known and the rules for going from input to output are inferred.
-These are analogous to classification and regression models \citep{Breiman1984}.
-Our application of the approaches used in this study illustrates only a sampling of the various methods available for clustering observations and fitting classification models.
-
-Geometric morphometric approaches to identifying differences in morphological variation between different classes, including cryptic species, mostly have used methods like linear discriminate analysis and canonical variates analysis \citep{Polly2003,Zelditch2004,Gaubert2005b,Gunduz2007,Polly2007a,Francoy2009,Sztencel-Jabonka2009,MitrovskiBogdanovic2013}.
-Because of their similarity to multivariate approaches like principle components analysis (PCA), these methods are comparatively straightforward ways of understanding the differences in morphology between classes.
-They also benefit from producing results that can be easily visualized, which aids in the interpretation and presentation of data and results.
-Most previous morphometric studies did not assess which amongst a set of alternative classification hypotheses was optimal.
-For example, studies such as those of \citet{Caumul2005a} and \citet{Polly2007a} focused on comparing different aspects of morphology and their fidelity to a classification scheme instead of comparing the fidelity of one aspect of morphology to multiple classification schemes.
-In this context, the study of \citet{Cardini2009a} is noteworthy because they compared morphological variation in marmots at the population, regional, and species level and determined the fidelity of shape to divisions at each of these levels.
-
-Here, we used multiple machine learning methods, both unsupervised and supervised, to compare different classification hypotheses.
-These methods provide different advantages for understanding how to classify taxa, as well as the accuracy of the resulting classifications.
-Although machine learning methods such as neural networks have been applied to studying shape variation \citep{Baylac2003,Dobigny2003,MacLeod2007,VandenBrink2011}, including in the context of automated taxon identification and classification of groups, the number of cases remains limited.
-In the current study, we not only consider pure classification accuracy but also use a statistic of classification strength that reflects the rate at which taxa are both accurately and inaccurately classified.
-
-We analyzed the problem of whether there are distinct subspecies or cryptic species within the western pond turtle, \textit{Emys marmorata} \citep{Baird1852} (formerly \emph{Clemmys marmorata}; see \citealp{Feldman2002}).
-\textit{Emys marmorata} is distributed from northern Washington State, USA to Baja California, Mexico.
-Traditionally, \textit{E.
-marmorata} was classified into two named subspecies: the northern \textit{E.
-marmorata marmorata} and the southern \textit{Emys marmorata pallida} \citep{Seeliger1945}, with a central Californian intergrade zone in between.
-\textit{Emys marmorata marmorata} is differentiated from \textit{E.
-marmorata pallida} by the presence of a pair of triangular inguinal scales and darker neck markings.
-The triangular inguinal plates can sometimes be present in \textit{E.
-marmorata pallida} although they are considerably smaller.
-\citet{Seeliger1945} did not formally include the Baja California populations of \textit{E.
-marmorata} in either taxon, implying the existence of a third distinct but unnamed subspecies.
-
-Previous work on morphological variation in \textit{E.
-marmorata} has focused primarily on differentiation between populations over a portion of the species' total range \citep{Lubcke2007,Germano2008,Germano2009,Bury2010}; comparatively few studies have included specimens from across the entire range \citep{Holland1992}.
-Most of these studies considered how local biotic and abiotic factors may contribute to differences in carapace length and found that size can vary greatly between different populations \citep{Lubcke2007,Germano2008,Germano2009}.
-There also has been interest in size-based sexual dimorphism in \textit{E.
-marmorata} \citep{Holland1992,Lubcke2007,Germano2009}, with males being on average larger than females based on total carapace length and other linear measurements.
-However, the quality of size as a classifier of sex can vary greatly between populations \citep{Holland1992} because of the magnitude of size differences among populations \citep{Lubcke2007,Germano2009}.
-However, the effect of sexual dimorphism on shape, \textit{sensu} \citet{Kendall1977a}, has not been assessed \citep{Holland1992,Lubcke2007,Germano2008}.
-
-Of particular importance in the context of cryptic diversity in \textit{E.
-marmorata} is the morphometric analysis of carapace shape carried out by \citet{Holland1992}, who compared populations of \textit{E.
-marmorata} from three areas of its range.
-This study concluded that geographic distance was a poor indicator of mophological differentiation, and instead hypothesized that geographic features such as breaks between different drainage basis are probably more important barriers to dispersal and interbreeding.
-Additionally, \citep{Holland1992} suggested that morphological differences were more pronounced as the magnitude of barriers and distance increased, but this variation required many variables to adequately capture, implying only very subtle morphological differentiation between putatively distinct populations.
-Finally, Holland concluded that \textit{E.
-marmorata} is best classified as three distinct species: a northern species, souther species, and a Columbia Basin species.
-This classification is similar to that of \citet{Seeliger1945}, except elevated to the species level and without recognition of a distinct Baja species.
-
-More recently, the phylogeography of \textit{E.
-marmorata} and the possibility of cryptic diversity was investigated using molecular data \citep{Spinks2005,Spinks2010,Spinks2014}.
-Based on mitochondrial DNA, \citet{Spinks2005} recognized four subclades within \textit{E.
-marmorata}, a northern clade, a San Joaquin Valley clade, a Santa Barbara clade, and a southern clade.
-Analyses with nuclear DNA \citep{Spinks2010} and single-nucleotide polymorphism (SNP) data suggest a primarily north--south division in \textit{E.
-marmorata}, although the dataset differed from that of \citet{Spinks2005} in the location of this break point.
-These studies discussed the potential taxonomic implications of their results, with \citet{Spinks2014} going so far as to strongly advocate for the recognition of at least two species (\emph{E.
-marmorata} and \emph{E.
-pallida}), and a possible third based on populations in Baja California.
-However, they did not discuss in detail the morphological characters that would help to diagnose these species beyond those specified by \citet{Seeliger1945}.
-Given that these characters are somewhat variable within the proposed species, and that \citet{Holland1992} described shell shape variation that might be consistent with this taxonomy, a geometric morphometric analysis of shell shape might provide a reliable way to diagnose groups (whether species or subspecies) within \textit{E.
-marmorata}.
-
-In this study, we attempt to estimate the best classification scheme of \textit{E.
-marmorata} based on variation in plastron (ventral shell) shape in order to determine whether this character is consistent with any of the past divisions based on other morphological features or molecular data.
-We are particularly interested in whether any aspect of plastron shape can be used to reliably diagnose \citet{Spinks2014} proposed species, and if so, the nature of that shape variation.
-
-Because of unclear geographic boundaries between subgroups of \textit{E.
-marmorata}, we compare multiple hypotheses of morphologically-- and molecularly--based classification.
-We hypothesize that if morphological variation corresponds to class assignment, it should be possible to determine the best classification hypothesis of \textit{E.
-marmorata} from amongst multiple candidate hypotheses.
-However, if morphological variation does not correspond to any of the standing hypothesis, then supervised learning model generalization performance will be poor.
-
-\section{Materials and Methods}
-\subsection{Specimens, sampling, morphometrics}
-We collected landmark-based morphometric data from 354 adult \textit{E.
-marmorata} museum specimens.
-These specimens are a subset of those included in \citet{Angielczyk2007}, \citet{Angielczyk2011}, and \citet{Angielczyk2013a}, essentially representing the adult individuals included in those papers.
-We chose to focus on adults because significant changes in plastron shape occur over the course of ontogeny in \textit{E.
-marmorata} and other emydines \citep{Angielczyk2013a}.
-
-We assigned a classification to each specimen for the different binning schemes based on geographic occurrence data recorded in museum collection archives.
-When precise latitude and longitude information were not available we estimated them from locality information.
-Because \citet{Spinks2005}, \citet{Spinks2010}, and \citet{Spinks2014} did not use vouchered specimens we were not able to directly sample individuals in their studies.
-Therefore our specimen classifications were based solely on the geographic information, not explicit assignment using molecular data.
-Because the exact barriers between different biogeographic regions are unknown and unclear, we represented each hypothesis with two schemes for a total of six different schemes.
-The schemes differed based on where geographic boundaries were assigned.
-This changes the classification of certain individuals near the boundaries between groups, providing a test of the robustness of the classification schemes.
-
-Following previous work on plastron variation \citep{Angielczyk2007,Angielczyk2011,Angielczyk2013a}, we used TpsDig 2.04 \citep{Rohlf2005} to digitize 19 landmarks (Fig.
-\ref{fig:plastra}).
-Seventeen of the landmarks are at the endpoints or intersection of the keratinous plastral scutes that cover the platron.
-Twelve of the landmarks were symmetrical across the axis of symmetry and, in order to prevent issues related to degrees of freedom and other similar concerns \citep{Klingenberg2002}, we reflected these landmarks across the axis of symmetry (i.e.
-midline) prior to analysis and used the average position of each symmetrical pair.
-In cases where damage or incompleteness prevented symmetric landmarks from being determined, we used only the single member of the pair.
-We conducted all subsequent analyses on the resulting ``half'' plastra.
-We superimposed the plastral landmark configurations using generalized Procrustes analysis \citep{Dryden1998a}, after which, we calculated the principal components (PC) of shape using the \texttt{shapes} package for R \citep{2013,Dryden2013}.
-
-\subsection{Machine learning analyses}
-\subsubsection{Unsupervised learning}
-In order to preserve the relationship between all landmark configurations in shape space, we measured the dissimilarity between observations using Kendall's Riemannian shape distance or \(\rho\) \citep{Kendall1984a,Dryden1998a}.
-We chose this metric because shape space, or the set of all possible shape configurations following Procrustes superimposition, is a Riemannian manifold and thus non-Euclidean \citep{Dryden1998a}.
-\(\rho\) varies between 0 and \(\pi / 2\) when there is no reflection invariance, which should not be a concern in the case of the half plastral landmark configurations used in the study.
-
-We divisively clustered the shape dissimilarity matrix using partitioning around mediods clustering (PAM), a method similar to \textit{k}-means clustering except that instead of minimizing the sum of squared Euclidean distances between observations and centroids, the sum of squared dissimilarities between observations and mediods is minimized \citep{Kaufman1990}.
-Because the optimal number of clusters of shape configurations in the study was unknown, being possibly three, four, or some other value, we estimated clustering solutions in which the number of clusters varied between one and eight.
-We compared clustering solutions using the gap statistic, which is a measure of goodness of clustering \citep{Tibshirani2001a}.
-
-%The gap statistic is defined
-%\[Gap_{n}(k) = E^{*}_{n}[\log(W_{k})] - \log(W_{k})\] 
-%where \(W_{k}\) is
-%\[W_{k} = \sum^{k}_{r = 1}{\frac{1}{2n_{r}} (\sum_{i,i' \in C_{r}} d_{ii'})}\].
-%\(d_{ii'}\) is the dispersion of the clustering solution or the sum of the pairwise dissimilarities between observations in each cluster and their respective mediods (\(C\)) for all clusters \(r\).
-This value is averaged and compared to the expected dispersion (\(E^{*}_{n}\)) of a sample \(n\) from a reference distribution.
-In this case, the reference distribution was estimated from 500 resamples.
-
-We conducted this analysis using the \texttt{cluster} package for R \citep{Maechler2013}.
-
-\subsubsection{Supervised learning}
-We used three different supervised learning, or classification, approaches: linear discriminate analysis, multinomial logistic regression, and random forests.
-Linear discriminate analysis, also known as canonical variate analysis, is commonly used in studies of geometric morphometric data \citep{Zelditch2004,Mitteroecker2011}.
-The other two methods, however, are not.
-In all cases, the optimal number of PCs used as predictors was chosen via maximum within-sample AUC value, explained below.
-
-Linear discriminate analysis (LDA) attempts to find a linear combination of predictors that best model two or more classes.
-LDA is very similar to PCA except that instead of finding the linear combination of features that maximize the amount of explained variance in the data, LDA maximizes the differences between classes.
-The results of this analysis produces a transformation matrix by which the original features can be transformed to reflect the best discrimination between the classes.
-In this study, we applied LDA to the eigenscores from a subset of the total number of PCs, ranging from two to six in increasing order of complexity.
-In total, this produced nine different LDA scaling matrices.
-
-Multinomial logistic regression is an extension of logistic regression, where instead of a binary response there are three or more response classes \citep{Venables2002a}.
-Similar to the odds ratios calculated from the coefficients of a logistic regression, the relative risk of a classification can be determined from the coefficients of the model.
-
-Random forest models are an extension of classification and regression trees (CART) \citep{Breiman1984,Breiman2001}.
-The goal of CARTs is to use a series of different features (i.e.
-predictors) to estimate the class of an observation.
-In top-down induction of decision trees for each member of a given set of predictor variables, attribute value tests are used to estimate the differences between classes.
-This process, called recursive partitioning, is then repeated on each subset.
-The recursion continues until the resulting observations all share the same class or no more meaningful partitions are possible.
-The resulting model is a tree structure by which observations are classified at each intersection via the estimated cutoff points from the attribute tests made during model fitting.
-% figure to help explain the ``tests''
-
-In a random forest model, many CARTs are built from a random subsample of both the features and the observations (specimens).
-This process is then repeated many times and the parameters of the final model are chosen as the mode of estimates from the distribution of CARTs \citep{Breiman2001}.
-In addition to classifying the observations, this procedure allows for the features to be ranked in order of importance.
-This is a generally useful property for studies in which the goal is to describe and model the differences between classes and the relative importance of different predictors.
-
-In this analysis, we used 1000 subtrees to estimate the random forest model parameters.
-We estimated the best set of predictors necessary for each classification scheme using a recursive feature selection algorithm, and we chose the optimal number of PCs to include based on the AUC of the model.
-Following the backwards selection algorithm implemented in \texttt{caret} \citep{KuhnMAN2013}, the maximum number of features were included in the initial model, their importance ranked, and the AUC of the model calculated.
-The lowest ranked feature was then removed, and the AUC of the model recalculated.
-This was repeated until only one feature, remained.
-Because PCs were kept in order of importance and not in relation to the amount of variance each PC described, the PCs are not included in the order of ascending eigenvalue.
-
-In classification studies, such as this one, a common metric of performance is area under the receiver operating characteristic curve (AUC).
-AUC is an estimate of the relationship between the false positive and true positive rates, as opposed to just the true positive rate (accuracy).
-This relationship is especially useful in cases such as this study where misclassification needs to be minimized just as much as an accurate classification must be obtained.
-AUC ranges between 0.5 and 1, with 0.5 indicating classification no better than random and 1 indicating perfect classification \citep{Hastie2009}.
-
-The standard AUC calculation is defined for binary classifications, however in our application there are multiple categories.
-The alternative calculation that we used follows an all-against-one strategy where the individual AUC values for each class versus all others are averaged to produce a multiclass AUC \citep{Hand2001}.
-To estimate confidence intervals on the out-of-sample AUC values, we performed a nonparametric bootstrap in which the true and estimated classifications were resampled with replacement.
-This was done 1000 times.
-
-The ultimate measure of model fit is accurately predicting the values of unobserved samples \citep{Hastie2009,Kuhn2013}.
-Within-sample performance is inherently biased upwards, so model evaluation requires overcoming this bias.
-With very large sample sizes, as in this study, part of the sample can be used as the ``training set'' and the remainder acts as the ``testing set.'' The former is used for fitting the model where as the later is used for measuring model performance, a process called model generalization.
-In this analysis, we used 75\% of samples as the training set while the remaining 25\% were used as the testing set.
-
-It is common for some out of sample observations to be misclassified.
-This misclassification may be due to the model not accurately representing shape variance, systematic differences between the training and test sets, or systematic differences between the accurately and inaccurately classified samples.
-Testing and training sets are determined completely at random within each class and with respect to shape.
-Results were not effected by the individual specimen class assignments to the testing or training sets.
-
-To determine if there were systematic differences in plastron shape between the correctly and incorrectly classified samples, we used a permutation test to estimate if the dissimilarity between the correctly and incorrectly classified individuals was significantly greater than random.
-The group labels were permuted 1000 times and the distance between the new centroids was calculated.
-The number of permutations less than the empirical difference divided by 1000 gives a \textit{p}-value for the test.
-Significant results indicate that correctly and incorrectly classified specimens are systematically different.
-This was done only for classes where there were 10 or more observations.
-
-\subsection{Comparison with clear-cut example}
-In addition to the above analysis of classification schemes of \textit{E.
-marmorata/pallida}, we applied the above approach to a selection of seven morphologically distinct emydine species.
-This additional analysis was done in order to confirm the efficacy of our supervised learning approach.
-The seven species analyzed were \textit{Emys blandigii}, \textit{Terrapene coahuila}, \textit{Clemmys guttata}, \textit{Glyptemys insculpta}, \textit{Glyptemys muhlenbergii}, \textit{Emys orbicularis}, and \textit{Terrapene ornata} with a total of 578 specimens analyzed in total.
-Again, these data were a subset of those used in \citet{Angielczyk2011} and \citet{Angielczyk2013a}.
-
-As with the \textit{E.
-marmorata/pallida} analysis, we analyzed plastron shape variation.
-The support for a seven category classification scheme was evaluated using random forest, linear discriminate analysis, and multinomial logistic regression where the number of PCs used as predictors ranged from two to 11.
-Data were split into training and testing sets and model performance for both sets was evaluated using the AUC metric.
-
-\section{Results}
-
-\subsection{Unsupervised learning}
-Comparison of gap statistic values from PAM clustering show that the optimal, minimal number of clusters is most likely one (Fig.
-\ref{fig:gap}).
-There is some ambiguity in choice because, although it is not statistically different from a solution with only one group, the solution with two groupings does have the greatest mean gap statistic.
-However, there is no geographical signal in the results of this clustering solution (Fig.
-\ref{fig:gap_map}).
-Because of this, we assert that there is no means of naturally partitioning plastron shape into distinct subgroups with out reference to external information.
-% PAM stuff
-% figure
-
-\subsection{Supervised learning}
-% model selection/fit
-AUC--based model selection revealed some important patterns of variation and congruence between the classification schemes and the actual data.
-Generally, the best performing models tended to include as many PCs as possible \ref{fig:sel}).
-Note that the best random forest models were determined via recursive feature selection, so PCs were not included in order of percent variance explained.
-That almost all LDA and multinomial logistic regression models were as complex as possible indicates that the differences between the different groups within each classification scheme are very small.
-
-As part of fitting a random forest model, a ranking of variable importance also is determined.
-Interestingly, the order of variable importance is not the same as the order of the PCs (\ref{fig:var_imp}).
-This means that the variance describing the differences between the classes does not align with the major axes of variance (i.e.
-the PCs).
-Such a result is expected if variation between classes was extremely fine grained and not a part of the principal form or function of the plastron.
-Moreover, this result is consistent with the results from the AUC--based model selection for the multinomial logistic regression and LDA models.
-
-Observed AUC values for all of the optimal models are not exceptionally high (\ref{fig:sel}).
-In most cases the different proposed classification schemes are generally poor descriptors of the observed variation.
-It appears that the data set is overwhelmed by noise, making any accurate classifications difficult at best.
-This observation is cemented with the generalizations of the models to the testing data set (\ref{fig:gen_hist}).
-
-% generalization
-Mean AUC values for the model generalizations, in most cases, are approximately equal to the observed AUC values from the training data set (Table \ref{tab:comp}).
-The  cases in which the AUC from the  generalizations is less than the observed, indicate poor model fit and a poor classification scheme.
-AUC values from model generalization, or estimating testing data set membership, does not indicate a clear ``best'' classification scheme (Fig.
-\ref{fig:gen_hist}).
-Although the scheme with two species has the greatest AUC point estimate for each modeling approach, this scheme is not significantly greater than any other except in some limited cases (e.g.
-LDA, Table \ref{tab:gen_tests}).
-% misfits
-Differences in mean shape between correctly and incorrectly classified observations from test set frequently were statistically significant, though there are exceptions.
-Again, this test was to determine if the mean shapes were statistically different or not.
-The frequency of these results, however, is important because it means that the different models are poor predictors of class membership.
-This may be because differences in plastron shape do not align with the any of the hypothesized classification schemes.
-
-\subsection{Comparison with clear-cut example}
-In contrast with the above results, the analysis of the seven morphologically-distinct species demonstrates near perfect classification of the both the in-sample and out-of-sample datasets (Table \ref{tab:second_res}).
-These results demonstrate that when there are strong distinctions between the states of the classification schemes, the methods used here can recover them.
-
-\section{Discussion}
-
-The results of this study indicate that there is no clear grouping of \textit{E. marmorata} based on plastron shape.
-
-The unsupervised learning results indicate only a single group of observations being optimal, which is consistent with the results from the generalizations of the supervised learning models.
-The classification schemes used in the supervised learning models correspond, loosely, to unsupervised learning solutions with multiple groups.
-Because unsupervised learning solutions with multiple groups are poor descriptors of the observed variation, it is important to see this generally supported by the supervised learning results.
-
-The results from fitting the various supervised learning models to each of the classification schemes generally shows that no one scheme is ``best.'' Possible explanations include that the genetic differentiation is not associated with plastron shape variation and/or that local selective pressures (e.g.
-from hydrological regime) overwhelm morphological differentiation.
-This makes sense given that the plastron is involved in both protection and hydrodynamics, and not necessary mate choice \citep{Rivera2008,Rivera2011,Stayton2011,Rivera2014} and that shell shape in \textit{E.
-marmorata} is known to vary among populations inhabiting water bodies with different flow regimes \citep{Holland1992,Lubcke2007,Germano2009}.
-Plastron shape does not seem to preserve a strong phylogenetic signal at the interspecific level in emydine turtles \citep{Angielczyk2011}, and our current results suggest that this may be the case for phylogeographic signal within emydine species as well.
-A final possibility (explored below) is that the proposed classification schemes themselves do not represent significant evolutionary lineages.
-
-Both the low AUC values (\(< 0.9\)) and the significant difference between the correctly and incorrectly classified observations support the conclusion that none of the hypothesized classification schemes are good descriptors of the observed plastral variation within \textit{E.
-marmorata}.
-
-Nevertheless, it is important to note that plastron shape is an extremely effective method for differentiating members of the other seven species we investigated.
-The magnitude of shape differences between the species (measured as Procrustes distance between species' mean shapes) is approximately an order of magnitude greater than the differences between the \textit{E.
-marmorata} subgroups, and the machine learning methods had no trouble accurately classifying the specimens correctly.
-These results demonstrate that plastron shape is normally a good marker for species delineation iin the closest relatives of \textit{E.
-marmorata}, and that our lack of results for \textit{E.
-marmorata} is not simply a shortcoming of the methods we applied.
-Indeed, they beg the question of what factors have suppressed morphological differentiation of plastron shape in \textit{E.
-marmorata} and \textit{E.
-pallida} if they are distinct species.
-Invoking issues such as the role of the plastron in protection or the need for streamlining are insufficient because the other species are expected to be subject to similar constraints \citep{Stayton2011,Pollyb}.
-
-\subsection{Is there more than one species of Western Pond Turtle?}
-
-The lack of morphological support for the distinctiveness of \textit{E.
-pallida} does not, on its own, preclude the recognition of this taxon.
-However, this apparent lack of congruence does prompt a reexamination of the methods and concepts that led to that taxonomic revision, especially considering that plastron shape is normally a strong indicator of species differences amoung emydines.
-In other words, before we can assess the significance of the morphological non-diagnosablity, it is essential to evaluate the methods and concepts that led to the initial taxonomic revision.
-
-\citet{Spinks2014} elevated \textit{E.
-pallida} based on a Bayesian species delimitation analysis of SNP data using BPP \citep{Yang2010b}.
-However, \citet{Spinks2014} did not heed the caveats about species delimitation methods raised by \citet{Carstens2013}.
-In addition to specifically addressing the shortcomings of validation methods such as BPP that rely on guide trees and ``should be interpreted with caution,'' \citet{Carstens2013} also strongly emphasize that ``Inferences regarding species boundaries based on genetic data alone are likely inadequate, and species delimitation should be conducted with consideration of the life history, geographical distribution, morphology and behaviour (where applicable) of the focal system\dots'' These caveats evoke the development of the Unified Species Concept \citep{Dayrat2005a,DeQueiroz2007b}, Integrative Taxonomy \citep{Padial2010}, and other pluralist approaches to species delimitation.
-None of these considerations were brought to bear on the \textit{E.
-marmorata} system until now, and in doing so we find the proposal that \textit{E.
-pallida} is a distinct species to be lacking in a normally robust morphological marker.
-
-The natural history and geographical distribution of \textit{E.
-marmorata} and \textit{E.
-pallida} also make the recognition of these taxa implausible.
-The data from \citet{Spinks2014} show extensive introgression and admixture in Central California, which is expected because there are no significant barriers to gene flow in this region.
-Combined with the well-demonstrated ability for testudinoid turtles, including emydids and even \textit{Emys}, to hybridize (e.g.
-\citealt{Buskirk2005,Spinks2009,Parham2013}) it is hard to imagine how \textit{E.
-marmorata} and \textit{E.
-pallida} could maintain their integrity in the face of such admixture.
-Because the geography, natural history, demonstrated genetic admixture of \textit{E.
-marmorata}, and comparisons with other morphologically-diagnosable species conflict with the recognition of \textit{E.
-pallida}, we hypothesize that out inability to classify the morphological data by proposed species is because \textit{E.
-pallida} is not a distinct species.
-We agree with \citet{Carstens2013} that ``the inferences drawn from species delimitation studies should be conservative, for in most contexts it is better to fail to delimit species than it is to falsely delimit entities that do not represent actual evolutionary lineages.''
-
-We fully agree with \citet{Spinks2014} that \textit{E.
-marmorata} (\textit{sensu lato}) is a species deserving of strong conservation efforts, and we do not wish to trivialize this need.
-Moreover, the genetic diversity uncovered by the analysis of \citet{Spinks2014} should be explicitly accounted for in any conservation plan.
-Given the apparent lack of morphological distinction, however, we consider that this diversity should be considered Evolutionary Significant Units or Distinct Population Segments instead of distinct species.
-
-Finally, it is important to note that the data and analyses we present do not let us definitively say whether the apparent lack of morphological divergence within \textit{E.
-marmorata} truly reflects the presence of a single species, or if it is an artifact of plastron shape being a poor morphological marker for phylogenetic and phylogeographic divergences, because we could not carry out our morphometric analyses on the specimens from which the genetic data were obtained.
-The comparisons with the other emydine species, suggests it is because \textit{E.
-marmorata} is a single species.
-However, tests of both our preferred conclusion (\textit{E.
-marmorata} as a single species) and that \citet{Spinks2014} should include morphological and molecular analyses of the same set of voucher specimens, as well as additional tests of species delimitation using alternative methods and corroborating evidence as suggested by \citet{Carstens2013}.
-From a morphological standpoint, support for the validity of ``\textit{E.
-Pallida}'' may come from other aspects of morphology, such as carapace shape or other features.
-Likewise, further investigation of the phylogeographic utility of plastron shape in other turtle species will help to clarify whether the lack of differentiation seen in \textit{E.
-marmoarata}, and the strong differentiation among the other emydines, is typical or an unusual case.
-
-\section*{Acknowledgements}
-Data collection for this project was supported in part by NSF DBI-0306158 (to KDA).
-G.
-Miller assisted with data collection and her participation in this research was supported by NSF REU DBI-0353797 (to R.
-Mooi of CAS).
-For access to emydine specimens, we thank: J.
-Vindum and R.
-Drewes (CAS); A.
-Resetar (FMNH); R.
-Feeney (LACM); C.
-Austin (LSUMNS); S.
-Sweet (MSE); J.McGuire and C.
-Conroy (MVZ); A.
-Wynn (NMNH); P.
-Collins (SBMNH); B.
-Hollingsworth (SDMNH); P.
-Holroyd (UCMP).
-We are grateful for S.
-Sweet for field assistance and the California Department of Fish and Game for permits.
-
-\clearpage
-\bibliographystyle{sysbio}
-\bibliography{turtle,packages}
-
-\clearpage
 
@@
 }
+  \begin{tabular}{ l | c c | c c | c c }
+    \hline
+    & \multicolumn{2}{| l |}{random forest} & 
+    \multicolumn{2}{| l |}{multinomial logistic regression} & 
+    \multicolumn{2}{| l}{linear discriminate analysis} \\
+    Scheme & Observed & Generalized & Observed & Generalized & Observed & Geneeralized \\ 
+    \hline
+    \hline
+    Morph 1 & 0.63 & 0.73 & 0.75 & 0.79 & 0.75 & 0.80 \\ 
+    Morph 2 & 0.61 & 0.58 & 0.76 & 0.77 & 0.76 & 0.77 \\ 
+    Mito 1& 0.63 & 0.62 & 0.75 & 0.63 & 0.75 & 0.63 \\ 
+    Mito 2 & 0.77 & 0.67 & 0.80 & 0.64 & 0.80 & 0.63 \\ 
+    Mito 3 & 0.56 & 0.64 & 0.71 & 0.74 & 0.71 & 0.73 \\ 
+    Nuclear & 0.56 & 0.67 & 0.74 & 0.62 & 0.74 & 0.77 \\ 
+    \hline
+  \end{tabular}
   \label{tab:comp}
@@
   \centering
+  \caption{Results of bootstrap comparisons between the scheme with the highest mean AUC value and all other schemes, defined P(best - other \(>\) 0).
+An asterix indicates the best scheme.
+This was done for each of the three modeling techniques included in this study: random forest (RF), multinomial logistic regression (MLR), and linear discriminate analysis (LDA).
+Probabilities are the percent of comparisons that are greater than the observed difference in means.}
   \begin{tabular}{ l r r r }
@@
   \end{tabular}
-  \caption{Results of bootstrap comparisons between the scheme with the highest mean AUC value and all other schemes, defined P(best - other \(>\) 0).
-An asterix indicates the best scheme.
-This was done for each of the three modeling techniques included in this study: random forest (RF), multinomial logistic regression (MLR), and linear discriminate analysis (LDA).
-Probabilities are the percent of comparisons that are greater than the observed difference in means.}
   \label{tab:gen_tests}
@@
   \centering
+  \caption{Results of comparisons between correctly and incorrectly classified observations from the testing data set, defined P(distance - simulated \(>\) 0).
+For each scheme, the classifications with at least 10 observations were tested.
+This was done for each of the three modeling techniques included in this study: random forest (RF), multinomial logistic regression (MLR), and linear discriminate analysis (LDA).}
   \begin{tabular}{ l | l r r r }
@@
   \end{tabular}
-  \caption{Results of comparisons between correctly and incorrectly classified observations from the testing data set, defined P(distance - simulated \(>\) 0).
-For each scheme, the classifications with at least 10 observations were tested.
-This was done for each of the three modeling techniques included in this study: random forest (RF), multinomial logistic regression (MLR), and linear discriminate analysis (LDA).}
   \label{tab:miss_tests}
@@
 
-\begin{figure}[ht]
-  \centering
-  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/plastra}
-  \caption{Depiction of general plastral shape of \textit{E. marmorata} and position of the 19 landmark used in this study. Anterior is towards the top of the figure.}
-  \label{fig:plastra}
-\end{figure}
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/gap_res}
-  \caption{Results from PAM clustering of the Riemannian shape distance for 8 different number of clusters. Vertical lines are 1 standard deviation of the mean determined from 500 resamples.}
-  \label{fig:gap}
-\end{figure}
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/gap_map}
-  \caption{Comparison of geographic distribution of clustered observations from the 2 clustering PAM solution.
-Colour and shape correspond to each of the groups.
-There is clearly no geographic signal in the data.}
-  \label{fig:gap_map}
-\end{figure}
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/sel_val}
-  \caption{Graphical representation of the AUC values from model selection for multinomial logigistic regression and linear discriminate analysis, respecitively.
-AUC model selection is based on greatest AUC value.
-The horizontal axis corresponds to the cummulative number of axes included in the model of interest.
-A red dot corresponds to the AUC best model for that classification scheme.}
-  \label{fig:sel}
-\end{figure}
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/var_imp}
-  \caption{Variable importance from the random forest models for each of the six classification schemes.
-Importance is measured as the mean decrease in Gini Index, which is a measure of the strength by which that variable determines CART structure.
-Indices that are farther to the right indicate greater variable importance.}
-  \label{fig:var_imp}
-\end{figure}
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/gen_res}
-  \caption{Bootstrap distributions for genearlized AUC values for each of the classification schemes.
-Each row corresponds to a different modeling approach: multinomial logistic regression (MLR), random forest (RF), linear discriminate analysis (LDA), and LDA using best variables from random forest.
-Each distribution corresponds to 1000 bootstrap replicates.}
-  \label{fig:gen_hist}
-\end{figure}
-
 \begin{table}[ht]
   \centering
+  \caption{Results from classification model estimates of the secondary, multi-species dataset.
+Models are random forest (RF), linear discriminate analysis (LDA), and multinomial logistic regression (MLR).
+Comparison of in-sample and out-of-sample AUC of the best performing model, along with the number of predictors.
+AUC values range between 0.5 (no better than random) and 1 (perfect classification).}
   \begin{tabular}{ l r r r }
@@
     & \# predictors & In-sample AUC & Out-of-sample AUC \\ 
+    \hline
     \hline
@@
   \end{tabular}
-  \caption{Results from classification model estimates of the secondary, multi-species dataset.
-Models are random forest (RF), linear discriminate analysis (LDA), and multinomial logistic regression (MLR).
-Comparison of in-sample and out-of-sample AUC of the best performing model, along with the number of predictors.
-AUC values range between 0.5 (no better than random) and 1 (perfect classification).}
   \label{tab:second_res}

### 2015-09-10 doc/tabs_2.tex
submitted to syst biol

--- 2015-09-10 doc/tabs.tex
@@
 
+\setcounter{table}{1}
+
 \begin{document}
-
-\begin{table}[ht]
-  \centering
-  \caption{AUC values for the best model of each classification scheme for both the observed (training) data and the generalized (testing) data.
-Results from all three different supervised learning approaches are shown here.
-AUC values range between 0.5 and 1.
-}
-  \begin{tabular}{ l | c c | c c | c c }
-    \hline
-    & \multicolumn{2}{| l |}{random forest} & 
-    \multicolumn{2}{| l |}{multinomial logistic regression} & 
-    \multicolumn{2}{| l}{linear discriminate analysis} \\
-    Scheme & Observed & Generalized & Observed & Generalized & Observed & Geneeralized \\ 
-    \hline
-    \hline
-    Morph 1 & 0.63 & 0.73 & 0.75 & 0.79 & 0.75 & 0.80 \\ 
-    Morph 2 & 0.61 & 0.58 & 0.76 & 0.77 & 0.76 & 0.77 \\ 
-    Mito 1& 0.63 & 0.62 & 0.75 & 0.63 & 0.75 & 0.63 \\ 
-    Mito 2 & 0.77 & 0.67 & 0.80 & 0.64 & 0.80 & 0.63 \\ 
-    Mito 3 & 0.56 & 0.64 & 0.71 & 0.74 & 0.71 & 0.73 \\ 
-    Nuclear & 0.56 & 0.67 & 0.74 & 0.62 & 0.74 & 0.77 \\ 
-    \hline
-  \end{tabular}
-  \label{tab:comp}
-\end{table}
 
@@
 
-\begin{table}
-  \centering
-  \caption{Results of comparisons between correctly and incorrectly classified observations from the testing data set, defined P(distance - simulated \(>\) 0).
-For each scheme, the classifications with at least 10 observations were tested.
-This was done for each of the three modeling techniques included in this study: random forest (RF), multinomial logistic regression (MLR), and linear discriminate analysis (LDA).}
-  \begin{tabular}{ l | l r r r }
-    \hline
-    Scheme & Class & RF & MLR & LDA \\
-    \hline
-    \hline
-    \multirow{2}{*}{Morph 1} & CCR & 0.77 & 1.00 & 1.00 \\ 
-    & marm & 0.87 & 0.93 & 1.00 \\ 
-    \hline
-    \multirow{2}{*}{Morph 2} & CCR & 0.88 & 1.00 & 1.00 \\ 
-    & marm & 1.00 & 1.00 & 0.98 \\ 
-    \hline
-    \multirow{2}{*}{Mito 1} & CCR & 0.94 & 0.99 & 0.98 \\ 
-    & marm & 0.99 & 0.98 & 0.47 \\ 
-    \hline
-    \multirow{2}{*}{Mito 2} & marm & 0.85 & 0.97 & 0.96 \\ 
-    & pall & 0.94 & 1.00 & 1.00 \\ 
-    \hline
-    \multirow{2}{*}{Mito 3} & 1 & 0.40 & 0.92 & 0.90 \\ 
-    & 3 & 0.97 & 0.99 & 1.00 \\ 
-    \hline
-    \multirow{2}{*}{Nuclear} & marm & 1.00 & 1.00 & 1.00 \\ 
-    & pall & 0.99 & 1.00 & 1.00 \\ 
-    \hline
-  \end{tabular}
-  \label{tab:miss_tests}
-\end{table}
-
-\begin{table}[ht]
-  \centering
-  \caption{Results from classification model estimates of the secondary, multi-species dataset.
-Models are random forest (RF), linear discriminate analysis (LDA), and multinomial logistic regression (MLR).
-Comparison of in-sample and out-of-sample AUC of the best performing model, along with the number of predictors.
-AUC values range between 0.5 (no better than random) and 1 (perfect classification).}
-  \begin{tabular}{ l r r r }
-    \hline
-    & \# predictors & In-sample AUC & Out-of-sample AUC \\ 
-    \hline
-    \hline
-    RF &   11 & 1.000 & 0.999 \\ 
-    MLR &    6 & 1.000 & 0.998 \\ 
-    LDA &   10 & 1.000 & 1.000 \\ 
-    \hline
-  \end{tabular}
-  \label{tab:second_res}
-\end{table}
-
 \end{document}

### 2015-09-10 doc/tabs_4.tex
submitted to syst biol

--- 2015-09-10 doc/tabs.tex
@@
 
+\setcounter{table}{3}
+
 \begin{document}
-
-\begin{table}[ht]
-  \centering
-  \caption{AUC values for the best model of each classification scheme for both the observed (training) data and the generalized (testing) data.
-Results from all three different supervised learning approaches are shown here.
-AUC values range between 0.5 and 1.
-}
-  \begin{tabular}{ l | c c | c c | c c }
-    \hline
-    & \multicolumn{2}{| l |}{random forest} & 
-    \multicolumn{2}{| l |}{multinomial logistic regression} & 
-    \multicolumn{2}{| l}{linear discriminate analysis} \\
-    Scheme & Observed & Generalized & Observed & Generalized & Observed & Geneeralized \\ 
-    \hline
-    \hline
-    Morph 1 & 0.63 & 0.73 & 0.75 & 0.79 & 0.75 & 0.80 \\ 
-    Morph 2 & 0.61 & 0.58 & 0.76 & 0.77 & 0.76 & 0.77 \\ 
-    Mito 1& 0.63 & 0.62 & 0.75 & 0.63 & 0.75 & 0.63 \\ 
-    Mito 2 & 0.77 & 0.67 & 0.80 & 0.64 & 0.80 & 0.63 \\ 
-    Mito 3 & 0.56 & 0.64 & 0.71 & 0.74 & 0.71 & 0.73 \\ 
-    Nuclear & 0.56 & 0.67 & 0.74 & 0.62 & 0.74 & 0.77 \\ 
-    \hline
-  \end{tabular}
-  \label{tab:comp}
-\end{table}
-
-\begin{table}
-  \centering
-  \caption{Results of bootstrap comparisons between the scheme with the highest mean AUC value and all other schemes, defined P(best - other \(>\) 0).
-An asterix indicates the best scheme.
-This was done for each of the three modeling techniques included in this study: random forest (RF), multinomial logistic regression (MLR), and linear discriminate analysis (LDA).
-Probabilities are the percent of comparisons that are greater than the observed difference in means.}
-  \begin{tabular}{ l r r r }
-    \hline
-    Scheme & RF & MLR & LDA \\
-    \hline
-    \hline
-    Morph 1 & \(\ast\) & \(\ast\) & 1 \\
-    Morph 2 & 0.79 & 0.55 & 1 \\
-    Mito 1 & 0.89 & 0.94 & 1 \\ 
-    Mito 2 & 0.82 & 0.57 & \(\ast\) \\ 
-    Mito 3 & 0.79 & 0.69 & 0.73 \\ 
-    Nuclear & 0.79 & 0.96 & 0.96 \\ 
-    \hline
-  \end{tabular}
-  \label{tab:gen_tests}
-\end{table}
-
-\begin{table}
-  \centering
-  \caption{Results of comparisons between correctly and incorrectly classified observations from the testing data set, defined P(distance - simulated \(>\) 0).
-For each scheme, the classifications with at least 10 observations were tested.
-This was done for each of the three modeling techniques included in this study: random forest (RF), multinomial logistic regression (MLR), and linear discriminate analysis (LDA).}
-  \begin{tabular}{ l | l r r r }
-    \hline
-    Scheme & Class & RF & MLR & LDA \\
-    \hline
-    \hline
-    \multirow{2}{*}{Morph 1} & CCR & 0.77 & 1.00 & 1.00 \\ 
-    & marm & 0.87 & 0.93 & 1.00 \\ 
-    \hline
-    \multirow{2}{*}{Morph 2} & CCR & 0.88 & 1.00 & 1.00 \\ 
-    & marm & 1.00 & 1.00 & 0.98 \\ 
-    \hline
-    \multirow{2}{*}{Mito 1} & CCR & 0.94 & 0.99 & 0.98 \\ 
-    & marm & 0.99 & 0.98 & 0.47 \\ 
-    \hline
-    \multirow{2}{*}{Mito 2} & marm & 0.85 & 0.97 & 0.96 \\ 
-    & pall & 0.94 & 1.00 & 1.00 \\ 
-    \hline
-    \multirow{2}{*}{Mito 3} & 1 & 0.40 & 0.92 & 0.90 \\ 
-    & 3 & 0.97 & 0.99 & 1.00 \\ 
-    \hline
-    \multirow{2}{*}{Nuclear} & marm & 1.00 & 1.00 & 1.00 \\ 
-    & pall & 0.99 & 1.00 & 1.00 \\ 
-    \hline
-  \end{tabular}
-  \label{tab:miss_tests}
-\end{table}
 

### 2015-09-10 doc/draft_turtle.tex
qc issues

@@
 }
-%  \begin{tabular}{ l | c c | c c | c c }
-%    \hline
-%    & \multicolumn{2}{| l |}{random forest} & 
-%    \multicolumn{2}{| l |}{multinomial logistic regression} & 
-%    \multicolumn{2}{| l}{linear discriminate analysis} \\
-%    Scheme & Observed & Generalized & Observed & Generalized & Observed & Geneeralized \\ 
-%    \hline
-%    \hline
-%    Morph 1 & 0.63 & 0.73 & 0.75 & 0.79 & 0.75 & 0.80 \\ 
-%    Morph 2 & 0.61 & 0.58 & 0.76 & 0.77 & 0.76 & 0.77 \\ 
-%    Mito 1& 0.63 & 0.62 & 0.75 & 0.63 & 0.75 & 0.63 \\ 
-%    Mito 2 & 0.77 & 0.67 & 0.80 & 0.64 & 0.80 & 0.63 \\ 
-%    Mito 3 & 0.56 & 0.64 & 0.71 & 0.74 & 0.71 & 0.73 \\ 
-%    Nuclear & 0.56 & 0.67 & 0.74 & 0.62 & 0.74 & 0.77 \\ 
-%    \hline
-%  \end{tabular}
+  \begin{tabular}{ l | c c | c c | c c }
+    \hline
+    & \multicolumn{2}{| l |}{random forest} & 
+    \multicolumn{2}{| l |}{multinomial logistic regression} & 
+    \multicolumn{2}{| l}{linear discriminate analysis} \\
+    Scheme & Observed & Generalized & Observed & Generalized & Observed & Geneeralized \\ 
+    \hline
+    \hline
+    Morph 1 & 0.63 & 0.73 & 0.75 & 0.79 & 0.75 & 0.80 \\ 
+    Morph 2 & 0.61 & 0.58 & 0.76 & 0.77 & 0.76 & 0.77 \\ 
+    Mito 1& 0.63 & 0.62 & 0.75 & 0.63 & 0.75 & 0.63 \\ 
+    Mito 2 & 0.77 & 0.67 & 0.80 & 0.64 & 0.80 & 0.63 \\ 
+    Mito 3 & 0.56 & 0.64 & 0.71 & 0.74 & 0.71 & 0.73 \\ 
+    Nuclear & 0.56 & 0.67 & 0.74 & 0.62 & 0.74 & 0.77 \\ 
+    \hline
+  \end{tabular}
   \label{tab:comp}
@@
 Probabilities are the percent of comparisons that are greater than the observed difference in means.}
-%  \begin{tabular}{ l r r r }
-%    \hline
-%    Scheme & RF & MLR & LDA \\
-%    \hline
-%    \hline
-%    Morph 1 & \(\ast\) & \(\ast\) & 1 \\
-%    Morph 2 & 0.79 & 0.55 & 1 \\
-%    Mito 1 & 0.89 & 0.94 & 1 \\ 
-%    Mito 2 & 0.82 & 0.57 & \(\ast\) \\ 
-%    Mito 3 & 0.79 & 0.69 & 0.73 \\ 
-%    Nuclear & 0.79 & 0.96 & 0.96 \\ 
-%    \hline
-%  \end{tabular}
+  \begin{tabular}{ l r r r }
+    \hline
+    Scheme & RF & MLR & LDA \\
+    \hline
+    \hline
+    Morph 1 & \(\ast\) & \(\ast\) & 1 \\
+    Morph 2 & 0.79 & 0.55 & 1 \\
+    Mito 1 & 0.89 & 0.94 & 1 \\ 
+    Mito 2 & 0.82 & 0.57 & \(\ast\) \\ 
+    Mito 3 & 0.79 & 0.69 & 0.73 \\ 
+    Nuclear & 0.79 & 0.96 & 0.96 \\ 
+    \hline
+  \end{tabular}
   \label{tab:gen_tests}
@@
 This was done for each of the three modeling techniques included in this study: random forest (RF), multinomial logistic regression (MLR), and linear discriminate analysis (LDA).}
-%  \begin{tabular}{ l | l r r r }
-%    \hline
-%    Scheme & Class & RF & MLR & LDA \\
-%    \hline
-%    \hline
-%    \multirow{2}{*}{Morph 1} & CCR & 0.77 & 1.00 & 1.00 \\ 
-%    & marm & 0.87 & 0.93 & 1.00 \\ 
-%    \hline
-%    \multirow{2}{*}{Morph 2} & CCR & 0.88 & 1.00 & 1.00 \\ 
-%    & marm & 1.00 & 1.00 & 0.98 \\ 
-%    \hline
-%    \multirow{2}{*}{Mito 1} & CCR & 0.94 & 0.99 & 0.98 \\ 
-%    & marm & 0.99 & 0.98 & 0.47 \\ 
-%    \hline
-%    \multirow{2}{*}{Mito 2} & marm & 0.85 & 0.97 & 0.96 \\ 
-%    & pall & 0.94 & 1.00 & 1.00 \\ 
-%    \hline
-%    \multirow{2}{*}{Mito 3} & 1 & 0.40 & 0.92 & 0.90 \\ 
-%    & 3 & 0.97 & 0.99 & 1.00 \\ 
-%    \hline
-%    \multirow{2}{*}{Nuclear} & marm & 1.00 & 1.00 & 1.00 \\ 
-%    & pall & 0.99 & 1.00 & 1.00 \\ 
-%    \hline
-%  \end{tabular}
+  \begin{tabular}{ l | l r r r }
+    \hline
+    Scheme & Class & RF & MLR & LDA \\
+    \hline
+    \hline
+    \multirow{2}{*}{Morph 1} & CCR & 0.77 & 1.00 & 1.00 \\ 
+    & marm & 0.87 & 0.93 & 1.00 \\ 
+    \hline
+    \multirow{2}{*}{Morph 2} & CCR & 0.88 & 1.00 & 1.00 \\ 
+    & marm & 1.00 & 1.00 & 0.98 \\ 
+    \hline
+    \multirow{2}{*}{Mito 1} & CCR & 0.94 & 0.99 & 0.98 \\ 
+    & marm & 0.99 & 0.98 & 0.47 \\ 
+    \hline
+    \multirow{2}{*}{Mito 2} & marm & 0.85 & 0.97 & 0.96 \\ 
+    & pall & 0.94 & 1.00 & 1.00 \\ 
+    \hline
+    \multirow{2}{*}{Mito 3} & 1 & 0.40 & 0.92 & 0.90 \\ 
+    & 3 & 0.97 & 0.99 & 1.00 \\ 
+    \hline
+    \multirow{2}{*}{Nuclear} & marm & 1.00 & 1.00 & 1.00 \\ 
+    & pall & 0.99 & 1.00 & 1.00 \\ 
+    \hline
+  \end{tabular}
   \label{tab:miss_tests}
@@
 AUC values range between 0.5 (no better than random) and 1 (perfect classification).}
-%  \begin{tabular}{ l r r r }
-%    \hline
-%    & \# predictors & In-sample AUC & Out-of-sample AUC \\ 
-%    \hline
-%    \hline
-%    RF &   11 & 1.000 & 0.999 \\ 
-%    MLR &    6 & 1.000 & 0.998 \\ 
-%    LDA &   10 & 1.000 & 1.000 \\ 
-%    \hline
-%  \end{tabular}
+  \begin{tabular}{ l r r r }
+    \hline
+    & \# predictors & In-sample AUC & Out-of-sample AUC \\ 
+    \hline
+    \hline
+    RF &   11 & 1.000 & 0.999 \\ 
+    MLR &    6 & 1.000 & 0.998 \\ 
+    LDA &   10 & 1.000 & 1.000 \\ 
+    \hline
+  \end{tabular}
   \label{tab:second_res}
@@
   \centering
-%  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/plastra}
+  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/plastra}
   \caption{Depiction of general plastral shape of \textit{E. marmorata} and position of the 19 landmark used in this study. Anterior is towards the top of the figure.}
@@
   \centering
-%  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/gap_res}
+  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/gap_res}
   \caption{Results from PAM clustering of the Riemannian shape distance for 8 different number of clusters. Vertical lines are 1 standard deviation of the mean determined from 500 resamples.}
@@
   \centering
-%  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/gap_map}
+  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/gap_map}
   \caption{Comparison of geographic distribution of clustered observations from the 2 clustering PAM solution.
@@
   \centering
-%  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/sel_val}
+  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/sel_val}
   \caption{Graphical representation of the AUC values from model selection for multinomial logistic regression and linear discriminate analysis, respecitively.
@@
   \centering
-%  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/var_imp}
+  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/var_imp}
   \caption{Variable importance from the random forest models for each of the six classification schemes.
@@
   \centering
-%  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/gen_res}
+  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/gen_res}
   \caption{Bootstrap distributions for genearlized AUC values for each of the classification schemes.
@@
   \centering
-%  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/seven_boot}
+  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/seven_boot}
   \caption{Distribution of 1000 bootstrap replicates of the out-of-sample AUC values from the secondary, multi-species dataset.

### 2016-01-20 doc/draft_turtle.tex
new data group. moved clusters out. redoing work

@@
 
+\begin{figure}[ht]
+  \centering
+  \includegraphics[height = 0.5\textheight, width = \textwidth, keepaspectratio = true]{figure/plastra}
+  \caption{Depiction of general plastral shape of \textit{E. marmorata} and position of the 19 landmark used in this study. Anterior is towards the top of the figure.}
+  \label{fig:plastra}
+\end{figure}
+
 Following previous work on plastron variation \citep{Angielczyk2007,Angielczyk2011,Angielczyk2013a}, we used TpsDig 2.04 \citep{Rohlf2005} to digitize 19 landmarks (Fig.
@@
 \subsection{Unsupervised learning}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[height = 0.5\textheight, width = \textwidth, keepaspectratio = true]{figure/gap_res}
+  \caption{Results from PAM clustering of the Riemannian shape distance for 8 different number of clusters. Vertical lines are 1 standard deviation of the mean determined from 500 resamples.}
+  \label{fig:gap}
+\end{figure}
+
 Comparison of gap statistic values from PAM clustering show that the optimal, minimal number of clusters is most likely one (Fig.
@@
 Because of this, we assert that there is no means of naturally partitioning plastron shape into distinct subgroups with out reference to external information.
-% PAM stuff
-% figure
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[height = 0.5\textheight, width = \textwidth, keepaspectratio = true]{figure/gap_map}
+  \caption{Comparison of geographic distribution of clustered observations from the 2 clustering PAM solution.
+Colour and shape correspond to each of the groups.
+There is clearly no geographic signal in the data.}
+  \label{fig:gap_map}
+\end{figure}
 
@@
 
+\begin{figure}[ht]
+  \centering
+  \includegraphics[height = 0.5\textheight, width = \textwidth, keepaspectratio = true]{figure/var_imp}
+  \caption{Variable importance from the random forest models for each of the six classification schemes.
+Importance is measured as the mean decrease in Gini Index, which is a measure of the strength by which that variable determines CART structure.
+Indices that are farther to the right indicate greater variable importance.}
+  \label{fig:var_imp}
+\end{figure}
+
 As part of fitting a random forest model, a ranking of variable importance also is determined.
@@
 
+\begin{figure}[ht]
+  \centering
+  \includegraphics[height = 0.5\textheight, width = \textwidth, keepaspectratio = true]{figure/gen_res}
+  \caption{Bootstrap distributions for genearlized AUC values for each of the classification schemes.
+Each row corresponds to a different modeling approach: multinomial logistic regression (MLR), random forest (RF), linear discriminate analysis (LDA), and LDA using best variables from random forest.
+Each distribution corresponds to 1000 bootstrap replicates.}
+  \label{fig:gen_hist}
+\end{figure}
+
 Observed AUC values for all of the optimal models are not exceptionally high (\ref{fig:sel}).
@@
 
-% generalization
+\begin{figure}[ht]
+  \centering
+  \includegraphics[height = 0.5\textheight, width = \textwidth, keepaspectratio = true]{figure/sel_val}
+  \caption{Graphical representation of the AUC values from model selection for multinomial logistic regression and linear discriminate analysis, respecitively.
+AUC model selection is based on greatest AUC value.
+The horizontal axis corresponds to the cummulative number of axes included in the model of interest.
+A highlighted point corresponds to the AUC best model for that classification scheme.}
+  \label{fig:sel}
+\end{figure}
+
+\begin{table}[ht]
+  \centering
+  \caption{AUC values for the best model of each classification scheme for both the observed (training) data and the generalized (testing) data.
+Results from all three different supervised learning approaches are shown here.
+AUC values range between 0.5 and 1.
+}
+  \begin{tabular}{ l | c c | c c | c c }
+    \hline
+    & \multicolumn{2}{| l |}{random forest} & 
+    \multicolumn{2}{| l |}{multinomial logistic regression} & 
+    \multicolumn{2}{| l}{linear discriminate analysis} \\
+    Scheme & Observed & Generalized & Observed & Generalized & Observed & Geneeralized \\ 
+    \hline
+    \hline
+    Morph 1 & 0.63 & 0.73 & 0.75 & 0.79 & 0.75 & 0.80 \\ 
+    Morph 2 & 0.61 & 0.58 & 0.76 & 0.77 & 0.76 & 0.77 \\ 
+    Mito 1& 0.63 & 0.62 & 0.75 & 0.63 & 0.75 & 0.63 \\ 
+    Mito 2 & 0.77 & 0.67 & 0.80 & 0.64 & 0.80 & 0.63 \\ 
+    Mito 3 & 0.56 & 0.64 & 0.71 & 0.74 & 0.71 & 0.73 \\ 
+    Nuclear & 0.56 & 0.67 & 0.74 & 0.62 & 0.74 & 0.77 \\ 
+    \hline
+  \end{tabular}
+  \label{tab:comp}
+\end{table}
+
 Mean AUC values for the model generalizations, in most cases, are approximately equal to the observed AUC values from the training data set (Table \ref{tab:comp}).
@@
 LDA, Table \ref{tab:gen_tests}).
-% misfits
+
+\begin{table}[ht]
+  \centering
+  \caption{Results of bootstrap comparisons between the scheme with the highest mean AUC value and all other schemes, defined P(best - other \(>\) 0).
+An asterix indicates the best scheme.
+This was done for each of the three modeling techniques included in this study: random forest (RF), multinomial logistic regression (MLR), and linear discriminate analysis (LDA).
+Probabilities are the percent of comparisons that are greater than the observed difference in means.}
+  \begin{tabular}{ l r r r }
+    \hline
+    Scheme & RF & MLR & LDA \\
+    \hline
+    \hline
+    Morph 1 & \(\ast\) & \(\ast\) & 1 \\
+    Morph 2 & 0.79 & 0.55 & 1 \\
+    Mito 1 & 0.89 & 0.94 & 1 \\ 
+    Mito 2 & 0.82 & 0.57 & \(\ast\) \\ 
+    Mito 3 & 0.79 & 0.69 & 0.73 \\ 
+    Nuclear & 0.79 & 0.96 & 0.96 \\ 
+    \hline
+  \end{tabular}
+  \label{tab:gen_tests}
+\end{table}
+
 Differences in mean shape between correctly and incorrectly classified observations from test set frequently were statistically significant, though there are exceptions.
@@
 These results demonstrate that when there are strong distinctions between the states of the classification schemes, the methods used here can recover them.
+
+\begin{table}[ht]
+  \centering
+  \caption{Results from classification model estimates of the secondary, multi-species dataset.
+Models are random forest (RF), linear discriminate analysis (LDA), and multinomial logistic regression (MLR).
+Comparison of in-sample and out-of-sample AUC of the best performing model, along with the number of predictors.
+AUC values range between 0.5 (no better than random) and 1 (perfect classification).}
+  \begin{tabular}{ l r r r }
+    \hline
+    & \# predictors & In-sample AUC & Out-of-sample AUC \\ 
+    \hline
+    \hline
+    RF &   11 & 1.000 & 0.999 \\ 
+    MLR &    6 & 1.000 & 0.998 \\ 
+    LDA &   10 & 1.000 & 1.000 \\ 
+    \hline
+  \end{tabular}
+  \label{tab:second_res}
+\end{table}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[height = 0.5\textheight, width = \textwidth, keepaspectratio = true]{figure/seven_boot}
+  \caption{Distribution of 1000 bootstrap replicates of the out-of-sample AUC values from the secondary, multi-species dataset.
+The AUC values are from the random forest (RF), linear discriminate analysis (LDA), and multinomial logistic regression (MLR).
+AUC values range between 0.5 (no better than random) and 1 (perfect classification).}
+  \label{fig:seven_boot}
+\end{figure}
 
@@
 
-\clearpage
-
-\begin{table}[ht]
-  \centering
-  \caption{AUC values for the best model of each classification scheme for both the observed (training) data and the generalized (testing) data.
-Results from all three different supervised learning approaches are shown here.
-AUC values range between 0.5 and 1.
-}
-  \begin{tabular}{ l | c c | c c | c c }
-    \hline
-    & \multicolumn{2}{| l |}{random forest} & 
-    \multicolumn{2}{| l |}{multinomial logistic regression} & 
-    \multicolumn{2}{| l}{linear discriminate analysis} \\
-    Scheme & Observed & Generalized & Observed & Generalized & Observed & Geneeralized \\ 
-    \hline
-    \hline
-    Morph 1 & 0.63 & 0.73 & 0.75 & 0.79 & 0.75 & 0.80 \\ 
-    Morph 2 & 0.61 & 0.58 & 0.76 & 0.77 & 0.76 & 0.77 \\ 
-    Mito 1& 0.63 & 0.62 & 0.75 & 0.63 & 0.75 & 0.63 \\ 
-    Mito 2 & 0.77 & 0.67 & 0.80 & 0.64 & 0.80 & 0.63 \\ 
-    Mito 3 & 0.56 & 0.64 & 0.71 & 0.74 & 0.71 & 0.73 \\ 
-    Nuclear & 0.56 & 0.67 & 0.74 & 0.62 & 0.74 & 0.77 \\ 
-    \hline
-  \end{tabular}
-  \label{tab:comp}
-\end{table}
-
-\begin{table}[ht]
-  \centering
-  \caption{Results of bootstrap comparisons between the scheme with the highest mean AUC value and all other schemes, defined P(best - other \(>\) 0).
-An asterix indicates the best scheme.
-This was done for each of the three modeling techniques included in this study: random forest (RF), multinomial logistic regression (MLR), and linear discriminate analysis (LDA).
-Probabilities are the percent of comparisons that are greater than the observed difference in means.}
-  \begin{tabular}{ l r r r }
-    \hline
-    Scheme & RF & MLR & LDA \\
-    \hline
-    \hline
-    Morph 1 & \(\ast\) & \(\ast\) & 1 \\
-    Morph 2 & 0.79 & 0.55 & 1 \\
-    Mito 1 & 0.89 & 0.94 & 1 \\ 
-    Mito 2 & 0.82 & 0.57 & \(\ast\) \\ 
-    Mito 3 & 0.79 & 0.69 & 0.73 \\ 
-    Nuclear & 0.79 & 0.96 & 0.96 \\ 
-    \hline
-  \end{tabular}
-  \label{tab:gen_tests}
-\end{table}
-
-\begin{table}[ht]
-  \centering
-  \caption{Results of comparisons between correctly and incorrectly classified observations from the testing data set, defined P(distance - simulated \(>\) 0).
+%\begin{table}[ht]
+%  \centering
+%  \caption{Results of comparisons between correctly and incorrectly classified observations from the testing data set, defined P(distance - simulated \(>\) 0).
 For each scheme, the classifications with at least 10 observations were tested.
 This was done for each of the three modeling techniques included in this study: random forest (RF), multinomial logistic regression (MLR), and linear discriminate analysis (LDA).}
-  \begin{tabular}{ l | l r r r }
-    \hline
-    Scheme & Class & RF & MLR & LDA \\
-    \hline
-    \hline
-    \multirow{2}{*}{Morph 1} & CCR & 0.77 & 1.00 & 1.00 \\ 
-    & marm & 0.87 & 0.93 & 1.00 \\ 
-    \hline
-    \multirow{2}{*}{Morph 2} & CCR & 0.88 & 1.00 & 1.00 \\ 
-    & marm & 1.00 & 1.00 & 0.98 \\ 
-    \hline
-    \multirow{2}{*}{Mito 1} & CCR & 0.94 & 0.99 & 0.98 \\ 
-    & marm & 0.99 & 0.98 & 0.47 \\ 
-    \hline
-    \multirow{2}{*}{Mito 2} & marm & 0.85 & 0.97 & 0.96 \\ 
-    & pall & 0.94 & 1.00 & 1.00 \\ 
-    \hline
-    \multirow{2}{*}{Mito 3} & 1 & 0.40 & 0.92 & 0.90 \\ 
-    & 3 & 0.97 & 0.99 & 1.00 \\ 
-    \hline
-    \multirow{2}{*}{Nuclear} & marm & 1.00 & 1.00 & 1.00 \\ 
-    & pall & 0.99 & 1.00 & 1.00 \\ 
-    \hline
-  \end{tabular}
-  \label{tab:miss_tests}
-\end{table}
-
-\begin{table}[ht]
-  \centering
-  \caption{Results from classification model estimates of the secondary, multi-species dataset.
-Models are random forest (RF), linear discriminate analysis (LDA), and multinomial logistic regression (MLR).
-Comparison of in-sample and out-of-sample AUC of the best performing model, along with the number of predictors.
-AUC values range between 0.5 (no better than random) and 1 (perfect classification).}
-  \begin{tabular}{ l r r r }
-    \hline
-    & \# predictors & In-sample AUC & Out-of-sample AUC \\ 
-    \hline
-    \hline
-    RF &   11 & 1.000 & 0.999 \\ 
-    MLR &    6 & 1.000 & 0.998 \\ 
-    LDA &   10 & 1.000 & 1.000 \\ 
-    \hline
-  \end{tabular}
-  \label{tab:second_res}
-\end{table}
-
-\clearpage
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/plastra}
-  \caption{Depiction of general plastral shape of \textit{E. marmorata} and position of the 19 landmark used in this study. Anterior is towards the top of the figure.}
-  \label{fig:plastra}
-\end{figure}
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/gap_res}
-  \caption{Results from PAM clustering of the Riemannian shape distance for 8 different number of clusters. Vertical lines are 1 standard deviation of the mean determined from 500 resamples.}
-  \label{fig:gap}
-\end{figure}
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/gap_map}
-  \caption{Comparison of geographic distribution of clustered observations from the 2 clustering PAM solution.
-Colour and shape correspond to each of the groups.
-There is clearly no geographic signal in the data.}
-  \label{fig:gap_map}
-\end{figure}
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/sel_val}
-  \caption{Graphical representation of the AUC values from model selection for multinomial logistic regression and linear discriminate analysis, respecitively.
-AUC model selection is based on greatest AUC value.
-The horizontal axis corresponds to the cummulative number of axes included in the model of interest.
-A highlighted point corresponds to the AUC best model for that classification scheme.}
-  \label{fig:sel}
-\end{figure}
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/var_imp}
-  \caption{Variable importance from the random forest models for each of the six classification schemes.
-Importance is measured as the mean decrease in Gini Index, which is a measure of the strength by which that variable determines CART structure.
-Indices that are farther to the right indicate greater variable importance.}
-  \label{fig:var_imp}
-\end{figure}
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/gen_res}
-  \caption{Bootstrap distributions for genearlized AUC values for each of the classification schemes.
-Each row corresponds to a different modeling approach: multinomial logistic regression (MLR), random forest (RF), linear discriminate analysis (LDA), and LDA using best variables from random forest.
-Each distribution corresponds to 1000 bootstrap replicates.}
-  \label{fig:gen_hist}
-\end{figure}
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/seven_boot}
-  \caption{Distribution of 1000 bootstrap replicates of the out-of-sample AUC values from the secondary, multi-species dataset.
-The AUC values are from the random forest (RF), linear discriminate analysis (LDA), and multinomial logistic regression (MLR).
-AUC values range between 0.5 (no better than random) and 1 (perfect classification).}
-  \label{fig:seven_boot}
-\end{figure}
+%  \begin{tabular}{ l | l r r r }
+%    \hline
+%    Scheme & Class & RF & MLR & LDA \\
+%    \hline
+%    \hline
+%    \multirow{2}{*}{Morph 1} & CCR & 0.77 & 1.00 & 1.00 \\ 
+%    & marm & 0.87 & 0.93 & 1.00 \\ 
+%    \hline
+%    \multirow{2}{*}{Morph 2} & CCR & 0.88 & 1.00 & 1.00 \\ 
+%    & marm & 1.00 & 1.00 & 0.98 \\ 
+%    \hline
+%    \multirow{2}{*}{Mito 1} & CCR & 0.94 & 0.99 & 0.98 \\ 
+%    & marm & 0.99 & 0.98 & 0.47 \\ 
+%    \hline
+%    \multirow{2}{*}{Mito 2} & marm & 0.85 & 0.97 & 0.96 \\ 
+%    & pall & 0.94 & 1.00 & 1.00 \\ 
+%    \hline
+%    \multirow{2}{*}{Mito 3} & 1 & 0.40 & 0.92 & 0.90 \\ 
+%    & 3 & 0.97 & 0.99 & 1.00 \\ 
+%    \hline
+%    \multirow{2}{*}{Nuclear} & marm & 1.00 & 1.00 & 1.00 \\ 
+%    & pall & 0.99 & 1.00 & 1.00 \\ 
+%    \hline
+%  \end{tabular}
+%  \label{tab:miss_tests}
+%\end{table}
 

### 2016-01-20 R/clear_cut.r

@@ -38,7 +38,9 @@ theme_update(axis.text = element_text(size = 20),
              strip.text = element_text(size = 20))
 
 # actually start doing analysis...
-newturt <- list.files('../data/new_turtle', full.names = TRUE)
+newturt <- list.files('../data/new_turtle', 
+                      pattern = 'adult', 
+                      full.names = TRUE)
 turt <- llply(newturt, function(x) read.csv(x, header = FALSE))
 centroids <- llply(turt, function(x) x[, ncol(x)])
 numbers <- llply(turt, function(x) x[, 1:2])

### 2016-01-20 R/make_plots.r

@@ -48,88 +48,6 @@ goog.map <- get_map(location = c(longmi,
 ggmap(goog.map)
 
 
-## PCA results
-#morph <- ggpairs(adult, columns = c(1:3),
-#                 upper = 'blank',
-#                 params = c(LabelSize = 2, gridLabelSize = 2, size = 1))
-#fits <- procGPA(land.adult)
-#links <- c(1:7, 13:8)
-#snd.links <- c(2, 8, 3, 9, 4, 10, 5, 11, 6, 12)
-#land <- mshape.plot(fits, links = links, snd.links = snd.links)
-#land <- land + coord_equal()
-#morph <- putPlot(morph, land, 1, 3)
-#pdf(file = '../doc/figure/pca_res.pdf')
-#print(morph)
-#dev.off()
-#
-#
-## variation along first two PCs
-#mt <- mshape(land.adult)
-#pc1.max <- which.max(adult$PC1)
-#pc1.min <- which.min(adult$PC1)
-#pc2.max <- which.max(adult$PC2)
-#pc2.min <- which.min(adult$PC2)
-#
-#ex.lab <- function(var, value) {
-#  value <- as.character(value)
-#  if(var == 'lab') {
-#    value[value == '1'] <- 'min'
-#    value[value == '2'] <- 'mean'
-#    value[value == '3'] <- 'max'
-#  }
-#  return(value)
-#}
-#
-#inits <- list(land.adult[, , pc1.min], land.adult[, , pc1.max],
-#              mt,
-#              land.adult[, , pc2.min], land.adult[, , pc2.max], 
-#              mt)
-#secs <- inits <- lapply(inits, as.data.frame)
-#inits <- lapply(inits, function(x) x[links, ])
-#inits <- lapply(inits, function(x) cbind(x, rbind(x[-1, ], x[1, ])))
-#inits <- lapply(inits, function(x) {
-#                names(x) <- c('V1', 'V2', 'V3', 'V4')
-#                x})
-#inits <- Reduce(rbind, inits)
-#pcl <- unlist(lapply(c('PC1', 'PC2'), function(x) rep(x, 3 * nrow(mt))))
-#pc.typ <- c(rep('min', nrow(mt)), rep('max', nrow(mt)))
-#pc.mm <- rep('mean', nrow(mt))
-#pc.typ <- c(pc.typ, pc.mm, pc.typ, pc.mm)
-#pcvar <- cbind(inits, pcl, pc.typ)
-#
[…]

### 2016-01-26 R/supervised_mung.r
trach analysis works. redo all at 10 samp per pred

@@ -9,7 +9,7 @@ source('../R/mung.r')
 set.seed(1)
 
 # formulas
-max.ad <- nrow(adult) / 50
+max.ad <- nrow(adult) / 10
 tvar.a <- paste('PC', 1:max.ad, sep = '')
 groups <- list('sh1', 'sh2', 'sh3', 'sh4', 'sh5', 'spinks')
 tform.a <- lapply(groups, function(x, y) make.form(y, x), y = tvar.a)

### 2016-01-26 R/make_plots.r
changed variable number. had to updage labeller function because ggplot update

@@ -14,7 +14,7 @@ library(plyr)
 
 source('../R/plotting_functions.r')
 
-source_url('https://raw.github.com/JoFrhwld/FAAV/master/r/stat-ellipse.R')
+#source_url('https://raw.github.com/JoFrhwld/FAAV/master/r/stat-ellipse.R') # 
 
 load('../data/gen.RData')
 
@@ -135,14 +135,12 @@ ggsave(file = '../doc/figure/sel_val.png', plot = gsel,
 
 
 # generalize plot
-gen.name <- function(var, value) {
+gen.name <- function(value) {
   value <- as.character(value)
-  if(var == 'L1') {
-    value[value == 'mn'] <- 'MLR'
-    value[value == 'rf'] <- 'RF'
-    value[value == 'lda'] <- 'LDA'
-    value[value == 'lrf'] <- 'selected LDA'
-  }
+  value[value == 'mn'] <- 'MLR'
+  value[value == 'rf'] <- 'RF'
+  value[value == 'lda'] <- 'LDA'
+  value[value == 'lrf'] <- 'selected LDA'
   return(value)
 }
 
@@ -156,7 +154,8 @@ lr <- Reduce(cbind, Map(function(x) x$t, lrf))
 colnames(lr) <- groups
 zz <- list(rf = oo, mn = uu, lda = vv, lrf = lr)
 dd <- melt(zz)
-gdist <- ggplot(dd, aes(x = value, fill = Var2))
+names(dd)[3] <- 'val'
+gdist <- ggplot(dd, aes(x = val, fill = Var2))
 gdist <- gdist + geom_histogram(aes(y = ..density..),
                                 alpha = 0.3,
                                 size = 0.1,
@@ -167,7 +166,7 @@ gdist <- gdist + scale_fill_manual(name = '', values = cbp,
                                    labels = c('Morph 1', 'Morph 2',
                                               'Mito 1', 'Nuclear',
                                               'Mito 2', 'Mito 3'))
-gdist <- gdist + facet_grid(L1 ~ ., labeller = gen.name)
+gdist <- gdist + facet_grid(L1 ~ ., labeller = labeller(L1 = gen.name))
 gdist <- gdist + labs(x = 'AUC')
 gdist <- gdist + coord_cartesian(xlim = c(0, 1))
 gdist <- gdist + theme(legend.title = element_blank(),

### 2016-01-28 R/clear_cut.r
included size as predictor. all graphs updated

@@ -42,8 +42,8 @@ newturt <- list.files('../data/new_turtle',
                       pattern = 'adult', 
                       full.names = TRUE)
 turt <- llply(newturt, function(x) read.csv(x, header = FALSE))
-centroids <- llply(turt, function(x) x[, ncol(x)])
 numbers <- llply(turt, function(x) x[, 1:2])
+centroids <- llply(turt, function(x) x[, ncol(x)])
 turt <- llply(turt, function(x) x[, -c(1:2, ncol(x))])
 # number, museum #, lands...., centroid
 turt <- Reduce(rbind, turt)
@@ -51,10 +51,11 @@ turt.align <- df2array(turt, n.land = 26, n.dim = 2)
 turt.proc <- procGPA(turt.align)
 turt.scores <- turt.proc$scores
 
+
 turt.name <- laply(str_split(newturt, '\\/'), function(x) x[length(x)])
 turt.name <- str_trim(str_extract(turt.name, '\\s(.*?)\\s'))
 turt.scores <- data.frame(name = rep(turt.name, times = laply(numbers, nrow)), 
-                          turt.scores)
+                          unlist(centroids), turt.scores)
 
 
 # supervised learning
@@ -65,18 +66,18 @@ train <- turt.scores[part, ]
 test <- turt.scores[-part, ]
 
 # go over all theoretical fits based on max pred
-fort.fit <- rfe(x = train[, seq(from = 2, to = max.ad + 1)], 
+fort.fit <- rfe(x = train[, seq(from = 2, to = max.ad + 2)], 
                  y = train[, 1], 
                  rfeControl = rf.ctrl, ntree = 100, metric = 'ROC', 
                  sizes = 1:max.ad)
 lda.fit <- list()
 for(ii in seq(from = 2, to = max.ad)) {
-  lda.fit[[ii - 1]] <- lda(train[, seq(from = 2, to = ii + 1)], train[, 1])
+  lda.fit[[ii - 1]] <- lda(train[, seq(from = 2, to = ii + 2)], train[, 1])
 }
 
 mnom.fit <- list()
 for(ii in seq(from = 2, to = max.ad)) {
-  dat <- as.matrix(train[, seq(from = 2, to = ii + 1)])
+  dat <- as.matrix(train[, seq(from = 2, to = ii + 2)])
   colnames(dat) <- NULL
[…]

### 2016-02-12 doc/draft_turtle.tex
beginning to redo turtle manuscript

@@
 
+% additionally
+% comparison with two groups
+
 \section{Materials and Methods}
 \subsection{Specimens, sampling, morphometrics}
-We collected landmark-based morphometric data from 354 adult \textit{E.
-marmorata} museum specimens.
-These specimens are a subset of those included in \citet{Angielczyk2007}, \citet{Angielczyk2011}, and \citet{Angielczyk2013a}, essentially representing the adult individuals included in those papers.
+Three different landmark-based morphometric datasets describing plastron variation were assembled for this analysis: \textit{E.
+marmorata} specimens from across its geographic range, \textit{Trachemys} specimens from both subspecies, and specimens from seven distinct species from the Emydidae family.
 We chose to focus on adults because significant changes in plastron shape occur over the course of ontogeny in \textit{E.
@@
 
+The first dataset included 354 adult \textit{E.
+marmorata} museum specimens; a subset of those included in \citet{Angielczyk2007}, \citet{Angielczyk2011}, and \citet{Angielczyk2013a}.
 We assigned a classification to each specimen for the different binning schemes based on geographic occurrence data recorded in museum collection archives.
@@
 
-\begin{figure}[ht]
+% trachemys
+%   how many
+%   where are they from?
+%   these are new
+
+For the third dataset, we analyzed 578 total specimens from the following species: \textit{Emys blandigii}, \textit{Terrapene coahuila}, \textit{Clemmys guttata}, \textit{Glyptemys insculpta}, \textit{Glyptemys muhlenbergii}, \textit{Emys orbicularis}, and \textit{Terrapene ornata}.
+Like the first data set, these specimens are a subset of those used in \citet{Angielczyk2011} and \citet{Angielczyk2013a}.
+
+\begin{figure}[h]
   \centering
@@
 Seventeen of the landmarks are at the endpoints or intersection of the keratinous plastral scutes that cover the platron.
-Twelve of the landmarks were symmetrical across the axis of symmetry and, in order to prevent issues related to degrees of freedom and other similar concerns \citep{Klingenberg2002}, we reflected these landmarks across the axis of symmetry (i.e.
+Twelve of the landmarks were symmetrical across the axis of symmetry and, in order to limit issues surrounding asymmetry and incomplete specimens \citep{Klingenberg2002}, we reflected these landmarks across the axis of symmetry (i.e.
 midline) prior to analysis and used the average position of each symmetrical pair.
@@
 We compared clustering solutions using the gap statistic, which is a measure of goodness of clustering \citep{Tibshirani2001a}.
-
-%The gap statistic is defined
-%\[Gap_{n}(k) = E^{*}_{n}[\log(W_{k})] - \log(W_{k})\] 
-%where \(W_{k}\) is
-%\[W_{k} = \sum^{k}_{r = 1}{\frac{1}{2n_{r}} (\sum_{i,i' \in C_{r}} d_{ii'})}\].
-%\(d_{ii'}\) is the dispersion of the clustering solution or the sum of the pairwise dissimilarities between observations in each cluster and their respective mediods (\(C\)) for all clusters \(r\).
-This value is averaged and compared to the expected dispersion (\(E^{*}_{n}\)) of a sample \(n\) from a reference distribution.
-In this case, the reference distribution was estimated from 500 resamples.
+This method has been applied to morphometric data before and represents a comparatively simple approach for estimating latent classification CITATION.
 
@@
 The other two methods, however, are not.
-In all cases, the optimal number of PCs used as predictors was chosen via maximum within-sample AUC value, explained below.
+In all cases, the optimal number of PCs used as predictors was chosen via maximum within-sample receiver operating characteristic values, explained below.
+
+In classification studies, such as this one, a common metric of performance is area under the receiver operating characteristic curve (AUC).
+AUC is an estimate of the relationship between the false positive and true positive rates, as opposed to just the true positive rate (accuracy).
+This relationship is especially useful in cases such as this study where misclassification needs to be minimized just as much as an accurate classification must be obtained.
+AUC ranges between 0.5 and 1, with 0.5 indicating classification no better than random and 1 indicating perfect classification \citep{Hastie2009}.
+
+The standard AUC calculation is defined for binary classifications, however in our application there are multiple categories.
+The alternative calculation that we used follows an all-against-one strategy where the individual AUC values for each class versus all others are averaged to produce a multiclass AUC \citep{Hand2001}.
+To estimate confidence intervals on the out-of-sample AUC values, we performed a nonparametric bootstrap in which the true and estimated classifications were resampled with replacement 1000 times.
 
@@
 Random forest models are an extension of classification and regression trees (CART) \citep{Breiman1984,Breiman2001}.
-The goal of CARTs is to use a series of different features (i.e.
-predictors) to estimate the class of an observation.
-In top-down induction of decision trees for each member of a given set of predictor variables, attribute value tests are used to estimate the differences between classes.
-This process, called recursive partitioning, is then repeated on each subset.
-The recursion continues until the resulting observations all share the same class or no more meaningful partitions are possible.
-The resulting model is a tree structure by which observations are classified at each intersection via the estimated cutoff points from the attribute tests made during model fitting.
-% figure to help explain the ``tests''
-
 In a random forest model, many CARTs are built from a random subsample of both the features and the observations (specimens).
@@
 This is a generally useful property for studies in which the goal is to describe and model the differences between classes and the relative importance of different predictors.
-
+Random forest models were fit using the ``randomForest'' R package CITATION.
 In this analysis, we used 1000 subtrees to estimate the random forest model parameters.
-We estimated the best set of predictors necessary for each classification scheme using a recursive feature selection algorithm, and we chose the optimal number of PCs to include based on the AUC of the model.
-Following the backwards selection algorithm implemented in \texttt{caret} \citep{KuhnMAN2013}, the maximum number of features were included in the initial model, their importance ranked, and the AUC of the model calculated.
+
+We estimated the best set of predictors in the random forest model using a recursive feature selection algorithm, with AUC as our optimality criterion.
+Following the iterative backwards selection algorithm implemented in \texttt{caret} \citep{KuhnMAN2013}, the maximum number of features were included in the initial model, their importance ranked, and the AUC of the model calculated.
 The lowest ranked feature was then removed, and the AUC of the model recalculated.
-This was repeated until only one feature, remained.
 Because PCs were kept in order of importance and not in relation to the amount of variance each PC described, the PCs are not included in the order of ascending eigenvalue.
-
-In classification studies, such as this one, a common metric of performance is area under the receiver operating characteristic curve (AUC).
-AUC is an estimate of the relationship between the false positive and true positive rates, as opposed to just the true positive rate (accuracy).
-This relationship is especially useful in cases such as this study where misclassification needs to be minimized just as much as an accurate classification must be obtained.
-AUC ranges between 0.5 and 1, with 0.5 indicating classification no better than random and 1 indicating perfect classification \citep{Hastie2009}.
-
-The standard AUC calculation is defined for binary classifications, however in our application there are multiple categories.
-The alternative calculation that we used follows an all-against-one strategy where the individual AUC values for each class versus all others are averaged to produce a multiclass AUC \citep{Hand2001}.
-To estimate confidence intervals on the out-of-sample AUC values, we performed a nonparametric bootstrap in which the true and estimated classifications were resampled with replacement.
-This was done 1000 times.
 
@@
 
-To determine if there were systematic differences in plastron shape between the correctly and incorrectly classified samples, we used a permutation test to estimate if the dissimilarity between the correctly and incorrectly classified individuals was significantly greater than random.
-The group labels were permuted 1000 times and the distance between the new centroids was calculated.
-The number of permutations less than the empirical difference divided by 1000 gives a \textit{p}-value for the test.
-Significant results indicate that correctly and incorrectly classified specimens are systematically different.
-This was done only for classes where there were 10 or more observations.
-
-\subsection{Comparison with clear-cut example}
-In addition to the above analysis of classification schemes of \textit{E.
-marmorata/pallida}, we applied the above approach to a selection of seven morphologically distinct emydine species.
-This additional analysis was done in order to confirm the efficacy of our supervised learning approach.
-The seven species analyzed were \textit{Emys blandigii}, \textit{Terrapene coahuila}, \textit{Clemmys guttata}, \textit{Glyptemys insculpta}, \textit{Glyptemys muhlenbergii}, \textit{Emys orbicularis}, and \textit{Terrapene ornata} with a total of 578 specimens analyzed in total.
-Again, these data were a subset of those used in \citet{Angielczyk2011} and \citet{Angielczyk2013a}.
-
-As with the \textit{E.
-marmorata/pallida} analysis, we analyzed plastron shape variation.
-The support for a seven category classification scheme was evaluated using random forest, linear discriminate analysis, and multinomial logistic regression where the number of PCs used as predictors ranged from two to 11.
-Data were split into training and testing sets and model performance for both sets was evaluated using the AUC metric.
-
 \section{Results}
@@
 
-\begin{figure}[ht]
+Gap statistic values from the PAM clustering are mostly equivalent for all potential clusters, indicating that none of the proposed clusters are statistically different from each other (Fig.
+\ref{fig:gap}).
+No clustering result is statistically greater than that with only a single cluster, because of this we favor the interpretation that there is no latent structure in the data detected in this analysis.
+While the solution with two groupings does have the greatest mean gap statistic, there is no geographical signal in the results of this clustering solution (Fig.
+\ref{fig:gap_map}).
+Additionally, there is no correspondence between these two clusters and the sex of the specimens INSERT STATS.
+Because of this, we assert that there is no means of naturally partitioning \textit{E.
+marmorata} plastron shape into distinct subgroups with out reference to external information.
+
+\begin{figure}[h]
   \centering
@@
 
-Comparison of gap statistic values from PAM clustering show that the optimal, minimal number of clusters is most likely one (Fig.
-\ref{fig:gap}).
-There is some ambiguity in choice because, although it is not statistically different from a solution with only one group, the solution with two groupings does have the greatest mean gap statistic.
-However, there is no geographical signal in the results of this clustering solution (Fig.
-\ref{fig:gap_map}).
-Because of this, we assert that there is no means of naturally partitioning plastron shape into distinct subgroups with out reference to external information.
-
-\begin{figure}[ht]
+\begin{figure}[h]
   \centering
@@
 
-\begin{figure}[ht]
+\begin{figure}[h]
   \centering
@@
   \includegraphics[height = 0.5\textheight, width = \textwidth, keepaspectratio = true]{figure/sel_val}
-  \caption{Graphical representation of the AUC values from model selection for multinomial logistic regression and linear discriminate analysis, respecitively.
+  \caption{Graphical representation of the AUC values from model selection for multinomial logistic regression and linear discriminate analysis, respectively.
 AUC model selection is based on greatest AUC value.
@@
     \multicolumn{2}{| l}{linear discriminate analysis} \\
-    Scheme & Observed & Generalized & Observed & Generalized & Observed & Geneeralized \\ 
+    Scheme & Observed & Generalized & Observed & Generalized & Observed & Generalized \\ 
     \hline
@@
 \ref{fig:gen_hist}).
-Although the scheme with two species has the greatest AUC point estimate for each modeling approach, this scheme is not significantly greater than any other except in some limited cases (e.g.
-LDA, Table \ref{tab:gen_tests}).
-
-\begin{table}[ht]
-  \centering
-  \caption{Results of bootstrap comparisons between the scheme with the highest mean AUC value and all other schemes, defined P(best - other \(>\) 0).
-An asterix indicates the best scheme.
-This was done for each of the three modeling techniques included in this study: random forest (RF), multinomial logistic regression (MLR), and linear discriminate analysis (LDA).
-Probabilities are the percent of comparisons that are greater than the observed difference in means.}
-  \begin{tabular}{ l r r r }
-    \hline
-    Scheme & RF & MLR & LDA \\
-    \hline
-    \hline
-    Morph 1 & \(\ast\) & \(\ast\) & 1 \\
-    Morph 2 & 0.79 & 0.55 & 1 \\
-    Mito 1 & 0.89 & 0.94 & 1 \\ 
-    Mito 2 & 0.82 & 0.57 & \(\ast\) \\ 
-    Mito 3 & 0.79 & 0.69 & 0.73 \\ 
-    Nuclear & 0.79 & 0.96 & 0.96 \\ 
-    \hline
-  \end{tabular}
-  \label{tab:gen_tests}
-\end{table}
-
-Differences in mean shape between correctly and incorrectly classified observations from test set frequently were statistically significant, though there are exceptions.
-Again, this test was to determine if the mean shapes were statistically different or not.
-The frequency of these results, however, is important because it means that the different models are poor predictors of class membership.
-This may be because differences in plastron shape do not align with the any of the hypothesized classification schemes.
-
-\subsection{Comparison with clear-cut example}
-In contrast with the above results, the analysis of the seven morphologically-distinct species demonstrates near perfect classification of the both the in-sample and out-of-sample datasets (Table \ref{tab:second_res}).
+
+\subsection{Results from other datasets}
+% trachemys
+% The analysis of the two subspecies of \uppercase{Trachemys}
+%   clearly identifiable
+%   in-sample is near perfect
+%   generalization is near perfect
+
+% seven species
+In contrast with the analysis of \textit{E.
+marmorata}, the analysis of the seven morphologically-distinct species demonstrates near perfect classification of the both the in-sample and out-of-sample datasets (Table \ref{tab:second_res}).
 Additionally, the distribution of ROC scores from 1000 bootstrap replicates are tightly clustered near ROC = 1 (Fig.
@@
 
-The unsupervised learning results indicate only a single group of observations being optimal, which is consistent with the results from the generalizations of the supervised learning models.
+The unsupervised learning results indicate no grouping is optimal or particularly better than a single grouping, which is consistent with the results from the generalizations of the supervised learning models.
 The classification schemes used in the supervised learning models correspond, loosely, to unsupervised learning solutions with multiple groups.

### 2016-02-16 doc/draft_turtle.tex
redid cluster. lots of small changes to document. needs some re-writing

@@
 marmorata/``pallida''} based on plastron shape.
-In contrast, the analysis of the ``clear-cut'' example demonstrates a near perfect classification, which demonstrates that the methods can recover correct results when an appropriate signal exists.
+In contrast, the analysis of the ``clear-cut'' example demonstrates a near perfect classification, which demonstrates that the methods can recover correct results when a signal exists.
 Explanations for the lack of grouping in \textit{E.
@@
 These considerations have sparked interest in whether geometric morphometric analyses can capture fine-scale variation that can be used for identifying cryptic species.
-This would make the task of idetifying and maintaining endagered or conserved groups much easier and could contribute to improved classifications of extinct taxa and populations.
+This would make the task of identifying and maintaining endangered or conserved groups much easier and could contribute to improved classifications of extinct taxa and populations.
 Most such studies focus on using morphometrics to discover differences between taxa that were identified by other means \citep{Polly2003,Zelditch2004,Gaubert2005b,Gunduz2007,Polly2007a,Demandt2009}.
@@
 The basic statistical mechanics are supplemented by randomization, sorting, and partitioning algorithms, along with the maximization or minimization of summary statistics, in order to best estimate a general model for all data, both sampled and unsampled \citep{Hastie2009}.
-Machine learning approaches have found use in medical research, epidemiology, economics and automated image identification such as handwritten zip codes \citep{Hastie2009}.
+Machine learning approaches have found use in medical research, epidemiology, economics and automated identification of images such as handwritten zip codes \citep{Hastie2009}.
 The two major classes of machine learning methods are unsupervised and supervised learning.
-Unsupervised learning methods are used with unlabeled data where the underlying structure is estimated, and they are analogous to clustering and density estimation methods \citep{Kaufman1990}.
+Unsupervised learning methods are used with unlabeled data where the underlying structure is estimated; they are analogous to clustering and density estimation methods \citep{Kaufman1990}.
 Supervised learning methods are used with labeled data where the final output of data is known and the rules for going from input to output are inferred.
@@
 Geometric morphometric approaches to identifying differences in morphological variation between different classes, including cryptic species, mostly have used methods like linear discriminate analysis and canonical variates analysis \citep{Polly2003,Zelditch2004,Gaubert2005b,Gunduz2007,Polly2007a,Francoy2009,Sztencel-Jabonka2009,MitrovskiBogdanovic2013}.
-Because of their similarity to multivariate approaches like principle components analysis (PCA), these methods are comparatively straightforward ways of understanding the differences in morphology between classes.
+Because of their similarity to multivariate approaches like principal components analysis (PCA), these methods are comparatively straightforward ways of understanding the differences in morphology between classes.
 They also benefit from producing results that can be easily visualized, which aids in the interpretation and presentation of data and results.
@@
 
-Of particular importance in the context of cryptic diversity in \textit{E.
+Of particular relevance in the context of cryptic diversity in \textit{E.
 marmorata} is the morphometric analysis of carapace shape carried out by \citet{Holland1992}, who compared populations of \textit{E.
 marmorata} from three areas of its range.
-This study concluded that geographic distance was a poor indicator of mophological differentiation, and instead hypothesized that geographic features such as breaks between different drainage basis are probably more important barriers to dispersal and interbreeding.
-Additionally, \citep{Holland1992} suggested that morphological differences were more pronounced as the magnitude of barriers and distance increased, but this variation required many variables to adequately capture, implying only very subtle morphological differentiation between putatively distinct populations.
+This study concluded that geographic distance was a poor indicator of morphological differentiation, and instead hypothesized that geographic features such as breaks between different drainage basis are probably more important barriers to dispersal and interbreeding.
+Additionally, \citet{Holland1992} suggested that morphological differences were more pronounced as the magnitude of barriers and distance increased, but this variation required many variables to adequately capture, implying only very subtle morphological differentiation between putatively distinct populations.
 Finally, Holland concluded that \textit{E.
-marmorata} is best classified as three distinct species: a northern species, souther species, and a Columbia Basin species.
+marmorata} is best classified as three distinct species: a northern species, southern species, and a Columbia Basin species.
 This classification is similar to that of \citet{Seeliger1945}, except elevated to the species level and without recognition of a distinct Baja species.
@@
 
+In order to test the efficacy of our approach, we analyzed two additional datasets: the two subspecies of \textit{Trachemys scripta scripta} and \textit{Trachemys scripta elgans}, and seven distinct species from the Emydidae family.
+The former dataset represents a situation similar to the \textit{E.
+marmorata} scenario except there are two valid subspecies CITATIONS.
+The latter dataset represents a set of morphologically distinct species.
+Both of these datasets help constrain the effectiveness of the methods used here and can help in interpreting the \textit{E.
+marmorata} scenario.
+
 % additionally
@@
 Three different landmark-based morphometric datasets describing plastron variation were assembled for this analysis: \textit{E.
-marmorata} specimens from across its geographic range, \textit{Trachemys} specimens from both subspecies, and specimens from seven distinct species from the Emydidae family.
+marmorata} specimens from across its geographic range, \textit{T.
+scripta} specimens from both subspecies, and specimens from seven distinct species from the Emydidae family.
 We chose to focus on adults because significant changes in plastron shape occur over the course of ontogeny in \textit{E.
@@
 
-% trachemys
-%   how many
-%   where are they from?
-%   these are new
+The second dataset is a compilation of XXX specimens of \textit{T.
+scripta} with approximatly equal representation of both the \textit{T.
+scripta scripta} and \textit{T.
+scripta elegans} subspecies.
+This landmark data was collected specifically for this study.
 
@@
 
-We conducted this analysis using the \texttt{cluster} package for R \citep{Maechler2013}.
+Only the \textit{E. marmorata} dataset was analyzed using unsupervised approaches. We conducted this analysis using the \texttt{cluster} package for R \citep{Maechler2013}.
 
@@
 The other two methods, however, are not.
-In all cases, the optimal number of PCs used as predictors was chosen via maximum within-sample receiver operating characteristic values, explained below.
+The set of possible predictors included the centroid size of each specimen along with the value of that specimen along all 22 PC axes.
+In all cases, the optimal subset of PCs included in the final model was chosen via maximum within-sample receiver operating characteristic values, explained below.
 
@@
 LDA is very similar to PCA except that instead of finding the linear combination of features that maximize the amount of explained variance in the data, LDA maximizes the differences between classes.
-The results of this analysis produces a transformation matrix by which the original features can be transformed to reflect the best discrimination between the classes.
+The results from this approach produce a transformation matrix by which the original features can be transformed to reflect the best discrimination between the classes.
 In this study, we applied LDA to the eigenscores from a subset of the total number of PCs, ranging from two to six in increasing order of complexity.
@@
 We estimated the best set of predictors in the random forest model using a recursive feature selection algorithm, with AUC as our optimality criterion.
-Following the iterative backwards selection algorithm implemented in \texttt{caret} \citep{KuhnMAN2013}, the maximum number of features were included in the initial model, their importance ranked, and the AUC of the model calculated.
-The lowest ranked feature was then removed, and the AUC of the model recalculated.
+Following the iterative backwards selection algorithm implemented in \texttt{caret} \citep{KuhnMAN2013}, the maximum number of features possible were included in the initial model, their importance ranked, and the AUC of the model calculated.
+The lowest-ranked feature was then removed, and the AUC of the model recalculated.
 Because PCs were kept in order of importance and not in relation to the amount of variance each PC described, the PCs are not included in the order of ascending eigenvalue.
@@
 In this analysis, we used 75\% of samples as the training set while the remaining 25\% were used as the testing set.
-
-It is common for some out of sample observations to be misclassified.
-This misclassification may be due to the model not accurately representing shape variance, systematic differences between the training and test sets, or systematic differences between the accurately and inaccurately classified samples.
 Testing and training sets are determined completely at random within each class and with respect to shape.
-Results were not effected by the individual specimen class assignments to the testing or training sets.
+Results were not affected by the individual specimen class assignments to the testing or training sets.
+
+Supervised learning methods were applied to all three datasets.
 
@@
 \ref{fig:gap_map}).
-Additionally, there is no correspondence between these two clusters and the sex of the specimens INSERT STATS.
+Additionally, there is no correspondence between these two clusters and the sex of the specimens (\(\chi^{2}\) test 0.7685, df = 1, \textit{p}-value 0.38).
 Because of this, we assert that there is no means of naturally partitioning \textit{E.
-marmorata} plastron shape into distinct subgroups with out reference to external information.
+marmorata} plastron shape into distinct subgroups without reference to external information.
 
@@
 AUC--based model selection revealed some important patterns of variation and congruence between the classification schemes and the actual data.
-Generally, the best performing models tended to include as many PCs as possible \ref{fig:sel}).
+Generally, the best performing models tended to include as many PCs as possible (Fig.
+\ref{fig:sel}).
 Note that the best random forest models were determined via recursive feature selection, so PCs were not included in order of percent variance explained.
@@
 As part of fitting a random forest model, a ranking of variable importance also is determined.
-Interestingly, the order of variable importance is not the same as the order of the PCs (\ref{fig:var_imp}).
+Interestingly, the order of variable importance is not the same as the order of the PCs (Fig.
+\ref{fig:var_imp}).
 This means that the variance describing the differences between the classes does not align with the major axes of variance (i.e.
@@
 
-Observed AUC values for all of the optimal models are not exceptionally high (\ref{fig:sel}).
+Observed AUC values for all of the optimal models are not exceptionally high (Fig.
+\ref{fig:sel}).
 In most cases the different proposed classification schemes are generally poor descriptors of the observed variation.
 It appears that the data set is overwhelmed by noise, making any accurate classifications difficult at best.
-This observation is cemented with the generalizations of the models to the testing data set (\ref{fig:gen_hist}).
+This observation is cemented with the generalizations of the models to the testing data set (Fig.
+\ref{fig:gen_hist}).
 
@@
 Mean AUC values for the model generalizations, in most cases, are approximately equal to the observed AUC values from the training data set (Table \ref{tab:comp}).
-The  cases in which the AUC from the  generalizations is less than the observed, indicate poor model fit and a poor classification scheme.
-AUC values from model generalization, or estimating testing data set membership, does not indicate a clear ``best'' classification scheme (Fig.
+The  cases in which the AUC from the  generalizations is less than the observed indicate poor model fit and a poor classification scheme.
+Comparison of AUC values from the model generalizations do not indicate a clear ``best'' classification scheme (Fig.
 \ref{fig:gen_hist}).
@@
 \subsection{Results from other datasets}
+
+In contrast to the muddled results from the analysis of \textit{E.
+marmorata}, the analysis of \textit{T.
+scripta scripta} to \textit{T.
+scripta elegans} indicates that these two subspecies are morphologically distinct enough to be classified with near perfect accuracy (Table \ref{tab:second_res}, with both in-sample and out-of-sample classification having AUC values of approximately 1 for all methods.
+The analysis of the seven morphologically-distinct species also demonstrates near perfect classification of both the in-sample and out-of-sample datasets (Table \ref{tab:second_res}).
+
+For both datasets, the distribution of ROC scores from 1000 bootstrap replicates are tightly clustered near AUC = 1 (Fig.
+\ref{fig:other_boot}), in contrast to the results from the \textit{E.
+marmorata} case (Fig.
+\ref{fig:gen_hist}).
+These results demonstrate that when there are strong distinctions between the states of the classification schemes, the methods used here can recover them.
+% insert table here with the AUC values for in and out of sample results
+%   just like below for the 7 clear cut
 % trachemys
@@
 
-% seven species
-In contrast with the analysis of \textit{E.
-marmorata}, the analysis of the seven morphologically-distinct species demonstrates near perfect classification of the both the in-sample and out-of-sample datasets (Table \ref{tab:second_res}).
-Additionally, the distribution of ROC scores from 1000 bootstrap replicates are tightly clustered near ROC = 1 (Fig.
-\ref{fig:seven_boot}), which is contrast to the results from the \textit{E.
-marmorata} case (Fig.
-\ref{fig:gen_hist}).
-These results demonstrate that when there are strong distinctions between the states of the classification schemes, the methods used here can recover them.
-
 \begin{table}[ht]
-  \centering
+  \begin{subtable}[t]{\linewidth}
+    \begin{tabular}{ l r r r }
+      \hline
+      & \# predictors & In-sample AUC & Out-of-sample AUC \\ 
+      \hline
+      \hline
+      RF &   10 & 0.965 & 1.000 \\ 
+      LDA &    7 & 1.000 & 0.993 \\ 
+      MnL &    5 & 1.000 & 0.979 \\ 
+      \hline
+    \end{tabular}
+  \end{subtable}
+
+  \begin{subtable}[t]{\linewidth}
+    \begin{tabular}{ l r r r }
+      \hline
+      & \# predictors & In-sample AUC & Out-of-sample AUC \\ 
+      \hline
+      \hline
+      RF &   11 & 1.000 & 0.999 \\ 
+      LDA &   10 & 1.000 & 1.000 \\ 
+      MnL &    6 & 1.000 & 0.998 \\ 
+      \hline
+    \end{tabular}
+  \end{subtable}
   \caption{Results from classification model estimates of the secondary, multi-species dataset.
@@
 AUC values range between 0.5 (no better than random) and 1 (perfect classification).}
-  \begin{tabular}{ l r r r }
-    \hline
-    & \# predictors & In-sample AUC & Out-of-sample AUC \\ 
-    \hline
-    \hline
-    RF &   11 & 1.000 & 0.999 \\ 
-    MLR &    6 & 1.000 & 0.998 \\ 
-    LDA &   10 & 1.000 & 1.000 \\ 
-    \hline
-  \end{tabular}
   \label{tab:second_res}
@@
   \centering
-  \includegraphics[height = 0.5\textheight, width = \textwidth, keepaspectratio = true]{figure/seven_boot}
+  \begin{subfigure}[h]{0.45\linewidth}
+    \includegraphics[height = 0.3\textheight, width = \textwidth, keepaspectratio = true]{figure/trach_boot}
+    \caption{\textit{Tracheyms}}
+    \label{fig:trach_boot}
+  \end{subfigure}
+  \begin{subfigure}[h]{0.45\linewidth}
+    \includegraphics[height = 0.3\textheight, width = \textwidth, keepaspectratio = true]{figure/seven_boot}
+    \caption{Seven}
+    \label{fig:seven_boot}
+  \end{subfigure}
   \caption{Distribution of 1000 bootstrap replicates of the out-of-sample AUC values from the secondary, multi-species dataset.
@@
 AUC values range between 0.5 (no better than random) and 1 (perfect classification).}
-  \label{fig:seven_boot}
+  \label{fig:other_boot}
 \end{figure}
@@
 marmorata} subgroups, and the machine learning methods had no trouble accurately classifying the specimens correctly.
-These results demonstrate that plastron shape is normally a good marker for species delineation iin the closest relatives of \textit{E.
+These results demonstrate that plastron shape is normally a good marker for species delineation in the closest relatives of \textit{E.
 marmorata}, and that our lack of results for \textit{E.
@@
 pallida} does not, on its own, preclude the recognition of this taxon.
-However, this apparent lack of congruence does prompt a reexamination of the methods and concepts that led to that taxonomic revision, especially considering that plastron shape is normally a strong indicator of species differences amoung emydines.
+However, this apparent lack of congruence does prompt a reexamination of the methods and concepts that led to that taxonomic revision, especially considering that plastron shape is normally a strong indicator of species differences among emydines.
 In other words, before we can assess the significance of the morphological non-diagnosablity, it is essential to evaluate the methods and concepts that led to the initial taxonomic revision.
@@
 Because the geography, natural history, demonstrated genetic admixture of \textit{E.
-marmorata}, and comparisons with other morphologically-diagnosable species conflict with the recognition of \textit{E.
-pallida}, we hypothesize that out inability to classify the morphological data by proposed species is because \textit{E.
+marmorata}, and comparisons with other morphologically diagnosable species conflict with the recognition of \textit{E.
+pallida}, we hypothesize that our inability to classify the morphological data by proposed species is because \textit{E.
 pallida} is not a distinct species.
-We agree with \citet{Carstens2013} that ``the inferences drawn from species delimitation studies should be conservative, for in most contexts it is better to fail to delimit species than it is to falsely delimit entities that do not represent actual evolutionary lineages.''
 
@@
 marmorata} truly reflects the presence of a single species, or if it is an artifact of plastron shape being a poor morphological marker for phylogenetic and phylogeographic divergences, because we could not carry out our morphometric analyses on the specimens from which the genetic data were obtained.
-The comparisons with the other emydine species, suggests it is because \textit{E.
+The comparisons with the other emydine species suggests it is because \textit{E.
 marmorata} is a single species.
@@
 From a morphological standpoint, support for the validity of ``\textit{E.
-Pallida}'' may come from other aspects of morphology, such as carapace shape or other features.
+pallida}'' may come from other aspects of morphology, such as carapace shape or other features.
 Likewise, further investigation of the phylogeographic utility of plastron shape in other turtle species will help to clarify whether the lack of differentiation seen in \textit{E.

### 2016-02-16 R/supervised_mung.r

@@ -66,7 +66,7 @@ centroid <- rawturt[, n.land + 1]  # size of each observation
 geo <- cbind(lat = meta.adult$lat, long = meta.adult$long)
 
 fit <- procGPA(land.adult)
-#adult.dist <- riem.matrix(land.adult)
+adult.dist <- riem.matrix(land.adult)
 
 adult <- cbind(data.frame(size = centroid, fit$stdscores),
                meta.adult)

### 2016-02-18 doc/draft_turtle.tex
some changes manuscript to reflect edits

@@
 We used multiple machine learning methods, both unsupervised and supervised, to compare different classification hypotheses and asked whether it is possible to determine which amongst a set of classification hypotheses is best.
-In addition, we applied the above approach to a “clear-cut” example of seven unambiguously distinct species closely related to \textit{E.
+  In addition, we applied the above approach to two additional datasets: the subspecies of \textit{Trachemys scripta} and a set of seven unambiguously distinct species closely related to \textit{E.
 marmorata}.
@@
 marmorata/``pallida''} based on plastron shape.
-In contrast, the analysis of the ``clear-cut'' example demonstrates a near perfect classification, which demonstrates that the methods can recover correct results when a signal exists.
+In contrast, the analysis of the other datasets demonstrate a near perfect classification, which demonstrates that the methods can recover correct results when they exist.
 Explanations for the lack of grouping in \textit{E.
-marmorata} include that possibility that genetic differentiation is not associated with plastron shape variation below the species level and/or that local selective pressures (e.g., from hydrological regime) overwhelm morphological differentiation.
+marmorata} include that possibility that genetic differentiation is not associated with plastron shape variation in this species complex and/or that local selective pressures (e.g., from hydrological regime) overwhelm morphological differentiation.
 A reconsideration of the methods used to delimit \textit{E.
@@
 marmorata} based on variation in plastron (ventral shell) shape in order to determine whether this character is consistent with any of the past divisions based on other morphological features or molecular data.
-We are particularly interested in whether any aspect of plastron shape can be used to reliably diagnose \citet{Spinks2014} proposed species, and if so, the nature of that shape variation.
+We are particularly interested in whether any aspect of plastron shape can be used to reliably diagnose the proposed species of \citet{Spinks2014}, and if so, the nature of that shape variation.
+
+% Why did we analyze plastron shape?
+%   ammenable to 2d GMM
+%   easy to compare across species
 
@@
 This changes the classification of certain individuals near the boundaries between groups, providing a test of the robustness of the classification schemes.
+Sex information was only know for a subset of the total dataset and was not included as a predictor of classification.
+Sex information was used to determine if observations cluster by sex or not.
 
 The second dataset is a compilation of XXX specimens of \textit{T.
-scripta} with approximatly equal representation of both the \textit{T.
+scripta} with approximately equal representation of both the \textit{T.
 scripta scripta} and \textit{T.
@@
 This landmark data was collected specifically for this study.
-
 For the third dataset, we analyzed 578 total specimens from the following species: \textit{Emys blandigii}, \textit{Terrapene coahuila}, \textit{Clemmys guttata}, \textit{Glyptemys insculpta}, \textit{Glyptemys muhlenbergii}, \textit{Emys orbicularis}, and \textit{Terrapene ornata}.
@@
 We conducted all subsequent analyses on the resulting ``half'' plastra.
-We superimposed the plastral landmark configurations using generalized Procrustes analysis \citep{Dryden1998a}, after which, we calculated the principal components (PC) of shape using the \texttt{shapes} package for R \citep{2013,Dryden2013}.
+We superimposed the plastral landmark configurations using generalized Procrustes analysis \citep{Dryden1998a}, after which, we calculated the principal components (PC) of shape using the \texttt{shapes} package for R \citep{2015,Dryden2013}.
 
@@
 We compared clustering solutions using the gap statistic, which is a measure of goodness of clustering \citep{Tibshirani2001a}.
-This method has been applied to morphometric data before and represents a comparatively simple approach for estimating latent classification CITATION.
+This method has been applied to morphometric data before and represents a comparatively simple approach for estimating latent classification \citep{Claude2008}.
 
@@
 We used three different supervised learning, or classification, approaches: linear discriminate analysis, multinomial logistic regression, and random forests.
-Linear discriminate analysis, also known as canonical variate analysis, is commonly used in studies of geometric morphometric data \citep{Zelditch2004,Mitteroecker2011}.
+Linear discriminate analysis, also known as canonical variate analysis, is commonly used in studies of geometric morphometric data \citep{Zelditch2004,Claude2008,Mitteroecker2011}.
 The other two methods, however, are not.
@@
 Similar to the odds ratios calculated from the coefficients of a logistic regression, the relative risk of a classification can be determined from the coefficients of the model.
+Multinomial logistic regression models were implemented using the ``nnet'' R package \citep{Venables2002a}.
 
@@
 This is a generally useful property for studies in which the goal is to describe and model the differences between classes and the relative importance of different predictors.
-Random forest models were fit using the ``randomForest'' R package CITATION.
+Random forest models were fit using the ``randomForest'' R package \citep{Liaw2002}.
 In this analysis, we used 1000 subtrees to estimate the random forest model parameters.
@@
 \ref{fig:gap_map}).
-Additionally, there is no correspondence between these two clusters and the sex of the specimens (\(\chi^{2}\) test 0.7685, df = 1, \textit{p}-value 0.38).
+Additionally, there is no correspondence between these two clusters and the sex of the specimens (\(\chi^{2}\) 0.7685, df = 1, \textit{p}-value 0.38).
 Because of this, we assert that there is no means of naturally partitioning \textit{E.
@@
 \citet{Spinks2014} elevated \textit{E.
-pallida} based on a Bayesian species delimitation analysis of SNP data using BPP \citep{Yang2010b}.
+pallida} based on a species delimitation analysis of SNP data using BPP \citep{Yang2010b}.
 However, \citet{Spinks2014} did not heed the caveats about species delimitation methods raised by \citet{Carstens2013}.
@@
 
-        \bibitem[{{R Core Team}(2013)}]{2013}
-        {R Core Team}. 2013. R: A Language and Environment for Statistical Computing. R
+        \bibitem[{{R Core Team}(2015)}]{2015}
+        {R Core Team}. 2015. R: A Language and Environment for Statistical Computing. R
         Foundation for Statistical Computing Vienna, Austria.
-        \bibAnnoteFile{2013}
+        \bibAnnoteFile{2015}
 
@@
 
+        \bibitem[{Liaw and Wiener(2002)}]{Liaw2002}
+        Law, A. and M. Wiener. 2002. {Classification and Regression by randomForest.} 
+        R News 2(3):18--22.
+        \bibAnnoteFile{Liaw2002}
+
       \bibitem[{Yang and Rannala(2010)}]{Yang2010b}
@@
         \bibAnnoteFile{Zelditch2004}
+
+        \bibitem[{Claude(2008)}]{Claude2008}
+        Claude, J. 2004. {Morphometrics with R}. Springer-Verlang, New York.
+        \bibAnnoteFile{Claude2008}
 

### 2016-02-29 doc/resubmit_letter.tex
response to reviewers letter started

\documentclass{letter}
\usepackage{microtype}
\usepackage{letterbib}
\usepackage{natbib}
\frenchspacing

\signature{Peter D Smits}
\address{Committee on Evolutionary Biology \\ University of Chicago \\
  1025 E. 57th Street \\ Culver Hall 402 \\ Chicago, IL 60637 \\
psmits@uchicago.edu}

\begin{document}
\begin{letter}{Editor \\ \textit{Systematic Biology}}
  \opening{Dear Editor,}

  Please find enclosed the manuscript entitled ``How cryptic is cryptic diversity? Machine learning approaches to classifying morphological variation in the Pacific Pond Turtle (\textit{Emys marmorata})'' which we are submitting for consideration at the journal \textit{Systematic Biology}. This manuscript is authored by myself (Peter D Smits), Kenneth D Angielczyk, James F Parham, and Bryan Stuart. This paper was previously submitted to this journal with manuscript ID USYB-2015-207.

  We believe our analysis and results are of interest to the community as we present a non-standard approach for evaluating the intensity of cryptic diversity and the efficacy of different classification schemes.

  Morphological and molecular attempts to classify the turtle species \textit{Emys marmorata} have produced three different potential sub-species or species divisions. In this study we compared these hypotheses of how to subdivide \textit{Emys marmorata} using multiple machine learning approaches. We analyzed over 500 specimens of this species from over its entire geographic range. Additionally, instead of classification accuracy, we use a metric of the ratio of true-positive and false-positive rates for both in-sample and out-of-sample model performance. Finally, in order to test whether our approaches were valid, we compared these results to those from two other independent datasets: the subspecies of \textit{Trachemys stripta} and seven morphologically distinct turtle species within the same subfamily.

  Our results demonstrate, that while our approaches can very precisely identify morphologically distinct groupings, the morphological diversity of the \textit{Emy marmorata} species ``complex'' is inconsistent with any of the proposed classification schemes. We propose that these results may be due to the possibility that genetic differentiation is not associated with plastron shape variation \textit{below} the species level and/or that local selective pressures (e.g., from hydrological regime) overwhelm morphological differentiation. Additionally, a reconsideration of the methods used to delimit \textit{E. ``pallida,''} the lack of barriers to gene flow, the strong evidence for widespread admixture between lineages, and the fact that plastron shape can be used to differentiate other emydine species suggest that its lack of diagnosability most likely reflects the non-distinctiveness of this proposed taxon.

  The reviews and comments from the Editor, Associate Editor, an anonymous reviewer, and Julien Claude were very helpful and we believe the manuscript has been much improved for them. Our responses to those reviewers follow.

  Thank you for considering our work. Please send all correspondence regarding this manuscript to me via my email address (psmits@uchicago.edu).

  \encl{Article, figures, tables.}

  \section{Editor}
  \textbf{My main concern first struck me at Line 147, which I think is the first mention of plastron shape. At that point, I realized it wasn’t clear to me what aspects of shell shape were used in previous studies. Was plastron shape considered by Holland or Seeliger? Basically, I’d just like to know more about how the data used in this study were used in previous studies (if at all). Why use plastron shape rather than other features? Why not the triangular inguinal scales instead (or as well)? Without a clear connection to previous studies, the relevance of the plastron data for this question is not obvious. I presume that it is and I’m just ignorant, but if we accept this for publication, I won’t be the only ignorant reader, so this should be made very clear.}

  %
  %
  %

  \textbf{1) Line 41: ``idetifying'', ``endagered'' - typos \\
    2) Line 73: “principle” should be ``principal'' here. \\
    3) Line 123: Should be ``Holland (1992)'' \\
    4) Line 127: ``souther'' - typo \\
    5) Line 166: The binning schemes have to be clearly described. They are called ``Mito 1'', ``Mito 2'', etc., in Figure 4. What are these? Reviewer 1 points this out as well. \\
    6) Line 216: ``The results of this analysis produces'' - subject-verb agreement issue here \\
    7) Line 245: ``the maximum number of features were included'' - subject-verb agreement \\
    8) Line 246: Hyphenate ``lowest-ranked feature''\\
    9) Line 247: Delete comma after ``feature''. \\
    10) Line 251: This is not an appropriate place to define the acronym ``AUC'', because you have already used it a few times. You need to introduce it earlier (reviewer 1 agrees).\\
    11) Line 270: ``out of sample'' should be ``out-of-sample''\\
    12) Line 274: ``effected'' should be ``affected'' here \\
  13) Line 304: ``with out'' should be ``without'' }

  Corrections made where appropriate.

  \textbf{14) Lines 299-305: As you note, figure 2 shows that the solution with two groupings has the greatest mean gap statistic, so I think it’s misleading to cite Fig. 2 when asserting that ``Comparison of gap statistic values from PAM clustering show that the optimal, minimal number of clusters is most likely one''. I understand the argument about the lack of a geographic signal, but unless I am severely misunderstanding something (always possible), Fig. 2, by itself, tells me two groupings is best. This needs to be clarified. }

  %
  %
  %

  \textbf{15) Line 308: Should be ``best-performing models''\\
    16) Line 309: ``as possible 6)'' -- ??? \\
    17) Line 328: Delete comma after ``observed'' here. \\
    18) Line 329-330: ``AUC values…or estimating testing data set membership, does not indicate'' - subject-verb agreement issue here \\
    19) Line 341: No hyphen needed on ``morphologically distinct species''\\
  20) Line 342: ``of the both the'' - typo }

  Corrections made where appropriate.

  \textbf{21) Line 343: What are ``ROC scores''? This acronym has not been introduced, as far as I can tell. }

  The methods section and text in general have been adjusted to introduce ``ROC'', ``AUC'', and similar terms earlier.

  \textbf{22) Line 344: ``which is contrast to'' - Poor wording here, or something. }

  Correction made.

  \textbf{23) Line 350: Again, I disagree with this; it is not what Figure 2 shows. Now, the geographical argument may be a good one, but it is accessory and is being invoked a posteriori to reject two groupings despite what Figure 2 shows. }

  %
  %
  %

  \textbf{24) Line 360: Again, why use plastron shape? If it was used by others to delineate two (or more) species within this group, o.k., but right now that is not clear. }

  %
  %
  %

  \textbf{25) Line 373: ``Nevertheless, it is important to note that plastron shape is an extremely effective method for differentiating members of the other seven species we investigated'' - Doesn’t this sort of run counter to the statement in line 364 (``Plastron shape does not seem to preserve a strong phylogenetic signal at the interspecific level in emydine turtles'')? Or is it that plastron is fine for classification even though it lacks strong phylogenetic signal? }

  %
  %
  %

  \textbf{26) Table 2 caption: ``asterix'' should be ``asterisk''. \\
    27) Line 379: ``iin'' – typo \\
    28) Line 401: ``morphologyand'' - typo \\
    29) Line 415: No need for a hyphen in ``morphologically diagnosable''\\
  30) Line 416: ``that out'' should be ``that our'' }

  Corrections made where appropriate.

  \textbf{31) Line 419: As I often tell my student, you should use quotes sparingly. This is the third quote from Carstens et al. (2013), and that just strikes me as somewhat lazy writing. I think simply paraphrasing Carstens et al. (2013) would be adequate in all three cases. Carstens is a fine writer, but I don’t feel that these quotes are particularly amazing crystallizations of a crucial concept that need to be repeated verbatim. }

  Quotation removed and text re-enforced.

  \textbf{32) Line 432: ``The comparisons with other emydine species, suggests'' - Watch subject-verb agreement, and delete comma after “species”. \\
  33) Line 437: No need to capitalize ``pallida'' here. }

  Corrections made where appropriate.

  \section{Anonymous reviewer \# 1}

  \textbf{I have two primary criticisms of the paper. First, the presentation of some material is not always conducive to a reader’s full understanding of this study. Aside from numerous typos or grammatical issues (see below), in some cases the material presented is incomplete, or not given enough context for the reader to understand. The major issues for me were as follows: the Results section contains multiple numbers in parentheses - are these supposed to refer to figure numbers? It would seem like they are, and that the authors just neglected to put ``Fig.'' in front, but they don’t show up in the proper order, so I can’t be sure. ``AUC'' first shows up in line 212 but is not defined until 251; it should be defined at its first appearance. Table 1 provides abbreviations for a series of ``Schemes'', but I can’t find in the text where such schemes are explicitly described. The authors should provide this information (``Morph 1'' corresponds to X classification in Y reference, for example). }

  %
  %
  %

  \textbf{Figure 3 purports to show the geographic distribution of groups arrived at for a certain solution; however, I’m pretty sure that I don’t see 354 points, which would correspond to all of the specimens in the data set. And this made me question my earlier assumption that the input for the grouping methods consisted of data for individual turtles. If the input matrices consisted of something else, like means for a given locality, the authors should state this. And it does make sense that the clustering should be performed on mean values; otherwise, it seems like some of the first clusters to come out, at least of the unsupervised approaches, would just separate males and females. So the authors should describe how they handled the issue of sexual dimorphism in plastron shape (well-known in many turtle species) in E. marmorata. }

 %
 %
 %

  \textbf{My second criticism has to do with the choice of data set for the author’s ``clear-cut example''. In one view, this study doesn’t need a ``clear-cut'' example for comparison at all. This is the case if the question of the study is: ``Is there evidence for morphological differentiation in the plastra of E. marmorata, consistent with some previously-proposed classifications?'' But on another view, an appropriate comparator is necessary. This is the case if the question of the study is: ``Is there as much evidence in the plastra of E. marmorata for grouping members of this species into multiple subspecies, as there is in other turtle species which have already been grouped into multiple subspecies?''. And of course, a comparator is also necessary to answer the question: ``Can the methods used in this paper ever produce a taxonomically-appropriate clustering?''\\
    The authors clearly intend the latter for their study. They almost could not have chosen a set of taxa better-suited to successfully finding differences (maybe not at all within a given subfamily of turtles). The species they have chosen differ in pretty much all of the major factors that are known to affect the plastra of turtles: size, plastral kinesis, habitat, and degree of phylogenetic relatedness. Perhaps this is why they chose the taxa that they did, but if so, then I’m not sure that the results are particularly germane to the rest of the analyses. Of course the methods will produce an appropriate grouping for that set of taxa - I cannot imagine any published methods not doing so. So these results don’t really provide a useful comparator with the E. marmorata groupings; they just show that the methods used aren’t utterly abysmal. \\
    Instead, I think that the authors need to be able to answer the second question that I proposed above. They need to be able to show that these methods can successfully group turtle plastra which show the same magnitude of variation that is expected in subspecies or recently-diverged species or cryptic species in related groups. Grouping species within Emydinae won’t provide this. I’m not sure what the best comparisons would be. Perhaps Chrysemys subspecies, or C. picta versus C. dorsalis. Perhaps Terrapene species or subspecies. But the authors should pick and justify an appropriate comparison. And if the methods cannot successfully group these taxa either, then the relevance of their failure to group E. marmorata will be put in its proper context. \\
  The paper is otherwise strong. Again, I think the authors did an excellent job of synthesizing results from multiple methods, and the study is important (and, needless to say, appropriate for Systematic Biology). But I think the authors need to clean up their presentation and choose a more appropriate comparison or validation of the relevance of their results. }

  %
  %
  %

  \textbf{Minor editorial comments follow. Line numbers are provided, along with suggested corrections: \\
    15 - what is meant by an ``appropriate'' signal? \\
    41 - ``identifying'' \\
    59 - ``identification of images'' (otherwise the sentence asserts that handwritten zip codes are a type of automated imagine identification) \\
    62 - ``estimated; they'' \\
    127 - ``southern'' \\
    150 - ``the proposed species of Spinks et al.'' \\
    160 - can you provide numbers for sexes as well? \\
    247 - ``feature remained'' \\
    274 - ``affected'' \\
    291 - can you provide numbers of specimens and sexes here as well? \\
    304 - ``without'' \\
    335 - ``sets'' \\
    379 - ``in'' \\
    384 - to be fair, streamlining is probably not relevant for T. ornata, and possibly much less so for G. insculpta and maybe even G. muhlenbergii \\
    391 - ``among'' \\
    401 - ``morphology and'' \\
    404 - well, geography, at least, was certainly brought to bear on the classification, yes? Even if life history, morphology, and behavior were not. \\
    416 - ``our'' \\
    432 - ``species suggest'' \\
    437 - ``pallida'' \\
    582 - ``2015'' \\
  583 - provide issue and pages, or ``in press'' }

  Corrections made where appropriate.

  \section{Reviewer \# 2 Julien Claude}

  \textbf{0. non-significant remarks \\
    -l. 41 idetifying -\(>\) identifying \\
    -l 189. update your citation in R. \\
    -l. 212. The explanation of AUC is coming late, you should avoid using abbreviation if the reader do not understand its meaning, at least you should give the meaning of each letter as soon a you are using the abbreviation. \\
    -l 379: iin -\(>\) in \\
  -explain which package/functions you have been using for LDA, random forest, and polynomial regression.}

  Changes made where appropriate. In particular, the explanation of AUC is moved earlier in the methods along with a few other aspects of terminology.

  \textbf{Major remarks: }

  \textbf{1. Interrogation concerning the choice of the methods \\\\
    1.1. The authors are using three supervised techniques but only one unsupervised technique to test or evaluate the clustering pattern in their turtle. Three of these methods (LDA, and PAM) have been used with morphometrics (PAM in Claude 2008: Morphometrics with R), there is a good example of regression trees applied to Kangaroo measurements in Faraway 2006 (Extending the linear model in R). I am not sure for multinomial logistic regression, but see my remark below : it may not be the best\\\\
  1.2. I am surprised that only PAM was used as unique non supervised method (with the use of the gap stat to select the number of group). This is certainly one of the simplest method with the kmeans, but there are at least two other methodologies that the authors could have selected because they have considerable advantages on PAM: the one of geneland (see Guillot et al., 2012, Systematic Biology) that has the advantage to consider geography or not and may work with the data of the authors (since they are spatialized; and Mclust (Fraley and Rafteri, 2002) which allows different model of covariance and variances within groups. l. 46, the work of Guillot et al. 2012 could be cited for species delimitation. l. 198, the use of PAM is not new in morphometrics, I have been using it in 2008 (maybe you should cite this example). }

  PCA is technically an unsupervised method, it just isn't a clustering method.
  %
  %
  %

  \textbf{1.3. l. 77 you say ``Most previous morphometric studies did not assess which amongst a set of alternative classification hypotheses was optimal.''; It is maybe true, but some have done the job, and I gave some possibilities in R and packages in Claude 2008. Here again you are omitting Guillot et al., 2012. \\\\
    1.4- why using only 1 unsupervised techniques, and three supervised techniques (on which two present certain weaknesses according to your data) ?\\\\
    As far as I understand, the paper is not intented to compare the behaviour of supervised techniques, but it is using several ones\ldots By looking at the results of bootstrap for AUC, it seems that LDA is the method that is clearly the less sensitive to sampling artefact (it is really clear for the trial with the 7 species), and in average it seems LDA reach a better score than others. I would consider only this in the ms, keeping the two other for appendices. It is rather easy to understand why: in the LDA the metric used is the mahalanobis distance : this allows somehow to disregard variation that can be due to intragroup variation. This does not seem to be the case for the two other methods. As stated by the authors, PCs are considered as independant features by forest models, and this may not be wonderful\ldots indeed PC scores depend on the ordination of individuals in the shape space and therefore of the sampling (they can be influenced by allometry, grouping structures, etc) and the superimposition\ldots they are not really independant variables and not measuring intergroup variation (even if intergroup variation is well structured). \\\\
  -\(>\) use mclust and or geneland instead or with PAM, put in appendix the methods/results that may have behaviroral problems}

  Principal components are considered independent features in all models where they are used as predictors of any variable; this happens in random forest, multinomial logistic regression, the linear formulation of LDA, all of them. If random forest has problems because of this fact, all methods have problems because of this fact.

  Also PCs are, by definition, independent descriptors of the observed variation. Principal component regression is an extremely common technique for dimenstional reduction and is essentially what we're doing here. Our large multivariate dataset is not ammenable to regression, so we use PCA to get independent/uncorrelated predictors of class. This approach is very kosher see \textit{Elements of Statistical Learning} by Tibshiri et al. for evidence.

  We are not perporting that species evolve along PC axes or that groups are defined by axes; we just state that groups may represent sections of morphospace defined by regions of each PC. The PCs are just means of reducing dimensions and putting everything in a nice coordinate system; these coordinates then help us identify regions of morphospace that may correspond to a group of interest.

  Additionally, we do not agree with your suggestion to move the other methods to the appendix. That undermines the entire point of this paper.
  %
  %
  %

  \textbf{2. Interrogation regarding the nature of the data. \\\\
  2.1.- l 188. explain whether you work on raw procrustes coordinates or on tangent coordinates. Explain why you are using PCs instead of tangent coordinates. }

  %
  %
  %

  \textbf{2.2.- can the inconclusive results come from the fact that measurement variation is great by comparison to interindividual variation ? Nowhere you estimate measurement error by comparion to interindividual variation in your data set (see Yezerinac et al., 1992 Syst. Bioloy; Claude et al., 2003 BJLS). How many percents of shape variation is it ? from my experience, measuring landmark on plastron in 2 D can produce an important error rate; is it your case ? are we able to recognize individuals ? I guess that the methods you are using are certainly sensitive to measurement error; but this is not estimated in your ms.\\\\
   l. 356. This can be also because intragroup variation is high by comparison to intergroup variation for the set of landmarks you selected (allometry, developmental noise, etc\ldots). Maybe shape sex dimorphism is stronger than intergroup difference. Why not focusing only on males or females (in the intro you say that sex shape dimorphism has not been evaluated)? }

  %
  %
  %

  \textbf{2.3. - Why do you symmetrise your individuals ? Actually asymmetry could not be geographically structured, I do not see clearly what kind of problems you could have with degrees of freedom. In symmetrising as you do, I hardly understand how you can compute the Riemanian distance ? does that mean that your individuals were superimposed on the basis of half the paired coordinates ? In doing so; are you considering that the landmarks on the symmetry axis are not constrained by the symmetry ? How much fluctuating asymetry do you have by comparison to measurement error and interindividual variation ? \\\\
  -\(>\) estimate the importance of asymmetry, digitization error, and sex dimorphism }

  %
  %
  %

  \textbf{3. Number of PCs retained in LDA and other supervised techniques. 
    I really do not understand why the authors restricted the number of PCs to 6. In exploring the results obtained, it seems that the best AUC is usually reached for the maximal number of PC retained; I guess it would be better if you increase the number of PCs, why not using all the non Null PCs (you can follow Chiari \& Claude approach 2012 -Compte rendus Biologie), there are 354 observations, 13 ``operating'' landmarks (-\(>\) at the end you have less that 26 non null PCs, which is considerably less than the number of observation - the maximal number of expected groups). In general, you can reasonnably use all of the variables rather than to reduce the dimensionality of your dataset. remark than in PAM, you are using variation in the whole shape space since you input the shape distance between two observations. \\\\
  -\(>\) use the selection of PCs that give the best cross validated classification rather than a small set }

  All these methods involve the estimation of minimum 1 parameter per PC used as a predictor of class. We restricted the number of PCs because we wanted a minimum sample size per PC so that our models were not overfit and bias towards the in-sample data; this would make out-of-sample predictions much weaker. However, following your suggestion, we have increased the number of features used per model. We've also included centroid size as a predictor of class.

  %
  %
  %

  \textbf{4. Group numbers and PAM \\\\
  Pam and marmorata study. according to fig 2., l. 299, the best clustering is for 2, not 1. You can explain later why 1 should be prefered on 2. 2 may come from the presence of a strong sex shape dimorphism but this is not evaluated. I do not understand how the vertical bars were obtained for the gap stat. The resampling should be explained here. But if this resampling correspond to the sampling error for the gap stat: note that according to your results: you can hesitate between 1, 2, 3, 4, 5 and 8 according to your plot, these partitions do not significanly differ amont each others. This part is rather inconclusive; rather than it clearly conclude for 1 group. }
  %
  %
  %

  \textbf{5. differences between well and wrongly assigned groups. \\\\
    l 276 to 282; I do not understand why the authors use a complex strategy to estimate differences between these groups, and the way it is run is not completely clear. Why not using Manova and look at the Pillai values and stat ? \\\\
  -\(>\) Better explain that part or make it simpler. }

  This section has been removed because it did not contribute to the over all text.

  \textbf{6. Problems in considering that PCs are ``independant'' shape variables. \\\\
    Note that grouping structure should not necessarily follow the direction of eigenvectors. The direction of eigenvectors depends on many things and is not necessarily determined by the groups. The direction may depends on allometries, or different number of observations within every groups and eventually different sex ratio within groups. Allometry is not necessarily orthogonal to group differentiation (because groups may differ in their allometry patterns - see for instance Claude, 2013-). The presence of allometry is not evaluated in your data set; the importance of shape sex dimorphism is not evaluated. \\\\
    -\(>\) avoid interpreting that a differentiation on PC 1 is of greatest importance than elsewhere. PC1 may represent just 15\% of total shape variation (there is still 85\% of remaing variation, and group structure is rarely colinear to PCs) \\\\
  -\(>\) investigate the role of allometry }

  %
  %
  %

  \textbf{7. why not including size in the analyses ? \\\\
    You are not including size (even if you are using only adults) in your supervised learning method, but you recognize that postnatal ontogeny can be important. However, incorporating size in your analyses may increase the differentiation between groups (in LDA, considering the fact that size can covary with shape; may influence ordination of individuals greatly). Allometry may be present in your data set and may obscure group differentiation (see Chaval et al., Mammalia, 2015). \\\\
  -\(>\) perform the same analysis with size }

  Size has now been included as a predictor of class in all analyses.

  \textbf{8. intro and background too long\ldots, machine learning section in methods too long \\\\
  This part can be reduced, the readership of systematic biology is aware about clustering and partitioning techniques. the readership is also familiar with geometric morphometrics and problems regarding species delimitation. The readership knows also how a LDA works. }

  Adjustments made as requested.

  \textbf{9. omitted references\ldots \\\\
  I was surprised that none of the publication that I have authored were mentionned in the text, while they could have been considered. As I have no competive interests with the authors, I thought it rather strange. Among these figure the selection of number of PCs for LDA, the use of unsupervised techniques, and the use of the combination between geography and morphology; which may be important to consider for improving this version. }

  No harm or foul was meant. Citations have bene updated to reflect your request.

\end{letter}
\bibliographystyle{plain}
\bibliography{newbib}
\end{document}

### 2016-03-01 doc/draft_turtle.tex
response to reviewers letter started
updated most figures. added some edits relating to those figures themselves.

@@
 \usepackage{subcaption, wrapfig}
+\usepackage{rotating}
 \usepackage{titlesec}
@@
 
-% Why did we analyze plastron shape?
-%   ammenable to 2d GMM
-%   easy to compare across species
+\uppercase{Why did we analyze plastron shape?}
 
@@
 marmorata} scenario.
-
-% additionally
-% comparison with two groups
 
@@
 Therefore our specimen classifications were based solely on the geographic information, not explicit assignment using molecular data.
-Because the exact barriers between different biogeographic regions are unknown and unclear, we represented each hypothesis with two schemes for a total of six different schemes.
-The schemes differed based on where geographic boundaries were assigned.
+Because the exact barriers between different biogeographic regions are unknown and unclear, we represented some hypothesis with two schemes for a total of six different schemes.
+These schemes differed based on where geographic boundaries were assigned.
 This changes the classification of certain individuals near the boundaries between groups, providing a test of the robustness of the classification schemes.
@@
 Sex information was used to determine if observations cluster by sex or not.
+The scheme names are as follows: Mito 1 and 2 correspond to \citet{Spinks2005}, Mito 3 corresponds to \citet{Spinks2010}, Morph 1 and Morph 2 correspond to \citet{Holland1992}, and Nuclear corresponds to \citet{Spinks2014}.
 
@@
 In all cases, the optimal subset of PCs included in the final model was chosen via maximum within-sample receiver operating characteristic values, explained below.
+Additionally, all of the following models assume that each of the predictors (PC axes) are statistically independent; this is a reasonable assumption because by definition principal components are statistically independent.
+Using PCs as predictors in regression and classification methods is called PCA regression is a common technique for for dimensionality reduction and for accounting for correlation between variables \citep{Hastie2009}.
+The PCs are a coordinate system where combinations of axes and ranges of those axes can describe groupings; we are not assuming that evolution happens ``along'' PC axes \citep{Uyeda2015} but use them as a coordinate system for describing observed shapes.
 
@@
   \includegraphics[height = 0.5\textheight, width = \textwidth, keepaspectratio = true]{figure/gap_res}
-  \caption{Results from PAM clustering of the Riemannian shape distance for 8 different number of clusters. Vertical lines are 1 standard deviation of the mean determined from 500 resamples.}
+  \caption{Results from PAM clustering of the Riemannian shape distance for 8 different number of clusters.
+Vertical lines are 1 standard deviation of the mean determined from 500 resamples.
+354 points are plotted, some of which overlap.}
   \label{fig:gap}
@@
 Colour and shape correspond to each of the groups.
-There is clearly no geographic signal in the data.}
+There is clearly no geographic signal in the data.
+Points are slightly transparent as 354 points are plotted but many of them overlap.}
   \label{fig:gap_map}
@@
 
-\begin{figure}[h]
-  \centering
-  \includegraphics[height = 0.5\textheight, width = \textwidth, keepaspectratio = true]{figure/var_imp}
-  \caption{Variable importance from the random forest models for each of the six classification schemes.
-Importance is measured as the mean decrease in Gini Index, which is a measure of the strength by which that variable determines CART structure.
-Indices that are farther to the right indicate greater variable importance.}
-  \label{fig:var_imp}
-\end{figure}
-
 As part of fitting a random forest model, a ranking of variable importance also is determined.
@@
 
-\begin{figure}[ht]
-  \centering
-  \includegraphics[height = 0.5\textheight, width = \textwidth, keepaspectratio = true]{figure/gen_res}
-  \caption{Bootstrap distributions for genearlized AUC values for each of the classification schemes.
-Each row corresponds to a different modeling approach: multinomial logistic regression (MLR), random forest (RF), linear discriminate analysis (LDA), and LDA using best variables from random forest.
-Each distribution corresponds to 1000 bootstrap replicates.}
-  \label{fig:gen_hist}
-\end{figure}
-
 Observed AUC values for all of the optimal models are not exceptionally high (Fig.
@@
 
+Mean AUC values for the model generalizations, in most cases, are approximately equal to the observed AUC values from the training data set (Table \ref{tab:comp}).
+The  cases in which the AUC from the  generalizations is less than the observed indicate poor model fit and a poor classification scheme.
+Comparison of AUC values from the model generalizations do not indicate a clear ``best'' classification scheme (Fig.
+\ref{fig:gen_hist}).
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[height = 0.8\textheight, width = 0.8\textwidth, keepaspectratio = true]{figure/var_imp}
+  \caption{Variable importance from the random forest models for each of the six classification schemes.
+Importance is measured as the mean decrease in Gini Index, which is a measure of the strength by which that variable determines CART structure.
+Indices that are farther to the right indicate greater variable importance.}
+  \label{fig:var_imp}
+\end{figure}
+
 \begin{figure}[ht]
   \centering
-  \includegraphics[height = 0.5\textheight, width = \textwidth, keepaspectratio = true]{figure/sel_val}
+  \includegraphics[height = 0.8\textheight, width = \textwidth, keepaspectratio = true]{figure/gen_res}
+  \caption{Bootstrap distributions for genearlized AUC values for each of the classification schemes.
+Columns correspond to different modeling approachs: multinomial logistic regression (MLR), random forest (RF), linear discriminate analysis (LDA), and LDA using best variables from random forest.
+Rows correspond to different classification schemes.
+Each distribution corresponds to 1000 bootstrap replicates.}
+  \label{fig:gen_hist}
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[height = 0.8\textheight, width = 0.8\textwidth, keepaspectratio = true]{figure/sel_val}
   \caption{Graphical representation of the AUC values from model selection for multinomial logistic regression and linear discriminate analysis, respectively.
@@
 
-\begin{table}[ht]
+\clearpage
+
+\begin{sidewaystable}[ht]
   \centering
@@
 }
-  \begin{tabular}{ l | c c | c c | c c }
+  \begin{tabular}{l | r r r | r r r | r r r | r r r }
     \hline
-    & \multicolumn{2}{| l |}{random forest} & 
-    \multicolumn{2}{| l |}{multinomial logistic regression} & 
-    \multicolumn{2}{| l}{linear discriminate analysis} \\
-    Scheme & Observed & Generalized & Observed & Generalized & Observed & Generalized \\ 
+    Scheme & \multicolumn{3}{c |}{RF} & \multicolumn{3}{c |}{MLR} & \multicolumn{3}{c |}{LDA} & \multicolumn{3}{c}{LDA + RF} \\
+    & Train & Test & SD(Test) & Train & Test & SD(Test) & Train & Test & SD(Test) & Train & Test & SD(Test) \\ 
     \hline
-    \hline
-    Morph 1 & 0.63 & 0.73 & 0.75 & 0.79 & 0.75 & 0.80 \\ 
-    Morph 2 & 0.61 & 0.58 & 0.76 & 0.77 & 0.76 & 0.77 \\ 
-    Mito 1& 0.63 & 0.62 & 0.75 & 0.63 & 0.75 & 0.63 \\ 
-    Mito 2 & 0.77 & 0.67 & 0.80 & 0.64 & 0.80 & 0.63 \\ 
-    Mito 3 & 0.56 & 0.64 & 0.71 & 0.74 & 0.71 & 0.73 \\ 
-    Nuclear & 0.56 & 0.67 & 0.74 & 0.62 & 0.74 & 0.77 \\ 
+    Mito 1 & 0.67 & 0.74 & 0.05 & 0.90 & 0.67 & 0.10 & 0.87 & 0.73 & 0.05 & 0.67 & 0.73 & 0.05 \\ 
+    Mito 2 & 0.77 & 0.70 & 0.05 & 0.91 & 0.58 & 0.05 & 0.89 & 0.73 & 0.05 & 0.77 & 0.74 & 0.05 \\ 
+    Mito 3 & 0.74 & 0.66 & 0.05 & 0.91 & 0.54 & 0.05 & 0.88 & 0.65 & 0.05 & 0.74 & 0.66 & 0.05 \\ 
+    Morph 1 & 0.81 & 0.71 & 0.06 & 0.86 & 0.73 & 0.05 & 0.85 & 0.74 & 0.05 & 0.81 & 0.79 & 0.05 \\ 
+    Morph 2 & 0.68 & 0.72 & 0.05 & 0.89 & 0.67 & 0.04 & 0.86 & 0.73 & 0.05 & 0.68 & 0.72 & 0.05 \\ 
+    Nuclear & 0.64 & 0.63 & 0.06 & 0.84 & 0.59 & 0.06 & 0.82 & 0.58 & 0.06 & 0.64 & 0.58 & 0.08 \\ 
    \hline
@@
   \label{tab:comp}
-\end{table}
-
-Mean AUC values for the model generalizations, in most cases, are approximately equal to the observed AUC values from the training data set (Table \ref{tab:comp}).
-The  cases in which the AUC from the  generalizations is less than the observed indicate poor model fit and a poor classification scheme.
-Comparison of AUC values from the model generalizations do not indicate a clear ``best'' classification scheme (Fig.
-\ref{fig:gen_hist}).
-
+\end{sidewaystable}
+
+\clearpage
 \subsection{Results from other datasets}
@@
   \centering
-  \begin{subfigure}[h]{0.45\linewidth}
-    \includegraphics[height = 0.3\textheight, width = \textwidth, keepaspectratio = true]{figure/trach_boot}
+  \begin{subfigure}[h]{0.8\linewidth}
+    \includegraphics[height = 0.8\textheight, width = \textwidth, keepaspectratio = true]{figure/trach_boot}
     \caption{\textit{Tracheyms}}
@@
   \end{subfigure}
-  \begin{subfigure}[h]{0.45\linewidth}
-    \includegraphics[height = 0.3\textheight, width = \textwidth, keepaspectratio = true]{figure/seven_boot}
+  \\
+  \begin{subfigure}[h]{0.8\linewidth}
+    \includegraphics[height = 0.8\textheight, width = \textwidth, keepaspectratio = true]{figure/seven_boot}
     \caption{Seven}
@@
     gastropod radiation from Lake Malawi}. Proceedings of the National Academy of
-        Sciences .
+    Sciences 110:13892--13897.
     \bibAnnoteFile{VanBocxlaer2013}
+
+    \bibitem[{Uyeda et~al.(2008)Uyeda, Caetano, and Pennell}]{Uyeda2015}
+    Uyeda, Josef C., D.~S.~Caetano and M.~W.~Pennell, {Comparative Analysis of 
+    Principal Components Can be Misleading}. Systematic Biology 64:677--689.
+    \bibAnnoteFile{Uyeda2015}
 
@@
 
-%\begin{table}[ht]
-%  \centering
-%  \caption{Results of comparisons between correctly and incorrectly classified observations from the testing data set, defined P(distance - simulated \(>\) 0).
-For each scheme, the classifications with at least 10 observations were tested.
-This was done for each of the three modeling techniques included in this study: random forest (RF), multinomial logistic regression (MLR), and linear discriminate analysis (LDA).}
-%  \begin{tabular}{ l | l r r r }
-%    \hline
-%    Scheme & Class & RF & MLR & LDA \\
-%    \hline
-%    \hline
-%    \multirow{2}{*}{Morph 1} & CCR & 0.77 & 1.00 & 1.00 \\ 
-%    & marm & 0.87 & 0.93 & 1.00 \\ 
-%    \hline
-%    \multirow{2}{*}{Morph 2} & CCR & 0.88 & 1.00 & 1.00 \\ 
-%    & marm & 1.00 & 1.00 & 0.98 \\ 
-%    \hline
-%    \multirow{2}{*}{Mito 1} & CCR & 0.94 & 0.99 & 0.98 \\ 
-%    & marm & 0.99 & 0.98 & 0.47 \\ 
-%    \hline
-%    \multirow{2}{*}{Mito 2} & marm & 0.85 & 0.97 & 0.96 \\ 
-%    & pall & 0.94 & 1.00 & 1.00 \\ 
-%    \hline
-%    \multirow{2}{*}{Mito 3} & 1 & 0.40 & 0.92 & 0.90 \\ 
-%    & 3 & 0.97 & 0.99 & 1.00 \\ 
-%    \hline
-%    \multirow{2}{*}{Nuclear} & marm & 1.00 & 1.00 & 1.00 \\ 
-%    & pall & 0.99 & 1.00 & 1.00 \\ 
-%    \hline
-%  \end{tabular}
-%  \label{tab:miss_tests}
-%\end{table}
-
 \end{document}

### 2016-03-01 R/make_plots.r

@@ -67,7 +67,7 @@ rf.imp$gr <- mapvalues(rf.imp$gr, unique(rf.imp$gr),
                          'Nuclear', 'Mito 2', 'Mito 3'))
 
 grf <- ggplot(rf.imp, aes(x = gini, y = pc))
-grf <- grf + geom_point(size = 5)
+grf <- grf + geom_point(size = 3)
 #grf <- grf + geom_bar(stat = 'identity', width = 0.5)
 grf <- grf + facet_grid(. ~ gr)
 grf <- grf + labs(x = 'Mean decrease\nin Gini Index', 
@@ -75,7 +75,7 @@ grf <- grf + labs(x = 'Mean decrease\nin Gini Index',
 grf <- grf + theme(axis.title = element_text(size = 20),
                    axis.title.y = element_text())
 ggsave(file = '../doc/figure/var_imp.png', plot = grf,
-       width = 15, height = 5)
+       width = 8, height = 10)
 
 # selection accumulation curves
 m.auc <- llply(tm.a.analysis$auc, function(x) {
@@ -130,7 +130,7 @@ gsel <- gsel + labs(x = 'Cummulative number of\nprincipal components',
                     y = 'Value')
 gsel <- gsel + coord_cartesian(ylim = c(0.5, 1))
 ggsave(file = '../doc/figure/sel_val.png', plot = gsel,
-       width = 15, height = 10)
+       width = 10, height = 8)
 
 
 
@@ -155,20 +155,20 @@ colnames(lr) <- groups
 zz <- list(rf = oo, mn = uu, lda = vv, lrf = lr)
 dd <- melt(zz)
 names(dd)[3] <- 'val'
-gdist <- ggplot(dd, aes(x = val, fill = Var2))
+gdist <- ggplot(dd, aes(x = val))
 gdist <- gdist + geom_histogram(aes(y = ..density..),
-                                alpha = 0.3,
+                                #alpha = 0.3,
                                 size = 0.1,
                                 position = 'identity',
-                                colour = 'darkgrey',
+                                #colour = 'darkgrey',
                                 binwidth = 1/100)
 gdist <- gdist + scale_fill_manual(name = '', values = cbp,
                                    labels = c('Morph 1', 'Morph 2',
                                               'Mito 1', 'Nuclear',
                                               'Mito 2', 'Mito 3'))
[…]

### 2016-03-03 doc/draft_turtle.tex
ken comments. editor and first reviewer almost finished. second reviewer hurts me inside.

@@
 \author[3]{James F Parham}%\thanks{jparham@fullerton.edu}}
+\author[4]{Bryan Stuart}%\thanks{bryan.stuart@naturalsciences.org}}
 \affil[1]{Committee on Evolutionary Biology, University of Chicago}
@@
 \affil[3]{Department of Geological Sciences, California State University -- Fullerton}
+\affil[4]{Section of Research and Collections, North Carolina Museum of Sciences}
 
@@
 We used multiple machine learning methods, both unsupervised and supervised, to compare different classification hypotheses and asked whether it is possible to determine which amongst a set of classification hypotheses is best.
-  In addition, we applied the above approach to two additional datasets: the subspecies of \textit{Trachemys scripta} and a set of seven unambiguously distinct species closely related to \textit{E.
+  In addition, we applied the above approach to two additional datasets: two well-supported subspecies of \textit{Trachemys scripta} and a set of seven unambiguously distinct species closely related to \textit{E.
 marmorata}.
@@
 marmorata/``pallida''} based on plastron shape.
-In contrast, the analysis of the other datasets demonstrate a near perfect classification, which demonstrates that the methods can recover correct results when they exist.
+In contrast, the analysis of the other datasets demonstrate a near perfect classification, which demonstrates that application of the methods to plastron shape data can recover correct results in similar situations.
 Explanations for the lack of grouping in \textit{E.
@@
 A reconsideration of the methods used to delimit \textit{E.
-``pallida,''} the lack of barriers to gene flow, the strong evidence for widespread admixture between lineages, and the fact that plastron shape can be used to differentiate other emydine species suggest that its lack of diagnosability most likely reflects the non-distinctiveness of this proposed taxon.
+``pallida,''} the lack of barriers to gene flow, the strong evidence for widespread admixture between lineages, and the fact that plastron shape can be used to differentiate other emydid species and subspecies suggest that its lack of diagnosability most likely reflects the non-distinctiveness of this proposed taxon.
 \end{abstract}
@@
 However, the quality of size as a classifier of sex can vary greatly between populations \citep{Holland1992} because of the magnitude of size differences among populations \citep{Lubcke2007,Germano2009}.
-However, the effect of sexual dimorphism on shape, \textit{sensu} \citet{Kendall1977a}, has not been assessed \citep{Holland1992,Lubcke2007,Germano2008}.
+The effect of sexual dimorphism on shape, \textit{sensu} \citet{Kendall1977a}, has not been assessed \citep{Holland1992,Lubcke2007,Germano2008}.
 
@@
 marmorata} is the morphometric analysis of carapace shape carried out by \citet{Holland1992}, who compared populations of \textit{E.
-marmorata} from three areas of its range.
+marmorata} from three areas of the species range.
 This study concluded that geographic distance was a poor indicator of morphological differentiation, and instead hypothesized that geographic features such as breaks between different drainage basis are probably more important barriers to dispersal and interbreeding.
@@
 Finally, Holland concluded that \textit{E.
-marmorata} is best classified as three distinct species: a northern species, southern species, and a Columbia Basin species.
+marmorata} is best classified as three distinct species: a northern species, a southern species, and a Columbia Basin species.
 This classification is similar to that of \citet{Seeliger1945}, except elevated to the species level and without recognition of a distinct Baja species.
@@
 Analyses with nuclear DNA \citep{Spinks2010} and single-nucleotide polymorphism (SNP) data suggest a primarily north--south division in \textit{E.
-marmorata}, although the dataset differed from that of \citet{Spinks2005} in the location of this break point.
-These studies discussed the potential taxonomic implications of their results, with \citet{Spinks2014} going so far as to strongly advocate for the recognition of at least two species (\emph{E.
+marmorata}, although these datasets differed from that of \citet{Spinks2005} in the location of the break point.
+All three studies discussed the potential taxonomic implications of their results, with \citet{Spinks2014} going so far as to strongly advocate for the recognition of at least two species (\emph{E.
 marmorata} and \emph{E.
@@
 However, they did not discuss in detail the morphological characters that would help to diagnose these species beyond those specified by \citet{Seeliger1945}.
-Given that these characters are somewhat variable within the proposed species, and that \citet{Holland1992} described shell shape variation that might be consistent with this taxonomy, a geometric morphometric analysis of shell shape might provide a reliable way to diagnose groups (whether species or subspecies) within \textit{E.
+Given that these characters are variable within the proposed species, and that \citet{Holland1992} described shell shape variation that might be consistent with this taxonomy, a geometric morphometric analysis of shell shape might provide a reliable way to diagnose groups (whether species or subspecies) within \textit{E.
 marmorata}.
@@
 
-In order to test the efficacy of our approach, we analyzed two additional datasets: the two subspecies of \textit{Trachemys scripta scripta} and \textit{Trachemys scripta elgans}, and seven distinct species from the Emydidae family.
+In order to test the efficacy of our approach, we analyzed two additional datasets: two subspecies of \textit{Trachemys} (\textit{T.
+scripta scripta} and \textit{T.
+scripta elgans}), and seven distinct species from the subfamily Emyidinae.
 The former dataset represents a situation similar to the \textit{E.
 marmorata} scenario except there are two valid subspecies CITATIONS.
-The latter dataset represents a set of morphologically distinct species.
+The latter dataset represents a set of well-defined morphologically distinct species.
 Both of these datasets help constrain the effectiveness of the methods used here and can help in interpreting the \textit{E.
@@
 marmorata} specimens from across its geographic range, \textit{T.
-scripta} specimens from both subspecies, and specimens from seven distinct species from the Emydidae family.
+scripta} specimens from both subspecies, and specimens from seven distinct emydine species.
 We chose to focus on adults because significant changes in plastron shape occur over the course of ontogeny in \textit{E.
@@
 When precise latitude and longitude information were not available we estimated them from locality information.
-Because \citet{Spinks2005}, \citet{Spinks2010}, and \citet{Spinks2014} did not use vouchered specimens we were not able to directly sample individuals in their studies.
+Because \citet{Spinks2005}, \citet{Spinks2010}, and \citet{Spinks2014} did not use vouchered specimens we were not able to directly sample the individuals in their studies.
 Therefore our specimen classifications were based solely on the geographic information, not explicit assignment using molecular data.
@@
 
-The second dataset is a compilation of XXX specimens of \textit{T.
-scripta} with approximately equal representation of both the \textit{T.
-scripta scripta} and \textit{T.
-scripta elegans} subspecies.
-This landmark data was collected specifically for this study.
+The second dataset is a compilation of 101 specimens of \textit{T.
+scripta}: 51 specimens of \textit{T.
+scripta scripta} and 50 specimens of \textit{T.
+scripta scripta}.
+These landmark data are new to this study.
 For the third dataset, we analyzed 578 total specimens from the following species: \textit{Emys blandigii}, \textit{Terrapene coahuila}, \textit{Clemmys guttata}, \textit{Glyptemys insculpta}, \textit{Glyptemys muhlenbergii}, \textit{Emys orbicularis}, and \textit{Terrapene ornata}.
@@
 
-Following previous work on plastron variation \citep{Angielczyk2007,Angielczyk2011,Angielczyk2013a}, we used TpsDig 2.04 \citep{Rohlf2005} to digitize 19 landmarks (Fig.
+Following previous work on plastron shape \citep{Angielczyk2007,Angielczyk2011,Angielczyk2013a}, we used TpsDig 2.04 \citep{Rohlf2005} to digitize 19 landmarks (Fig.
 \ref{fig:plastra}).
-Seventeen of the landmarks are at the endpoints or intersection of the keratinous plastral scutes that cover the platron.
-Twelve of the landmarks were symmetrical across the axis of symmetry and, in order to limit issues surrounding asymmetry and incomplete specimens \citep{Klingenberg2002}, we reflected these landmarks across the axis of symmetry (i.e.
+Seventeen of the landmarks are at the endpoints or intersection of the keratinous plastral scutes that cover the plastron.
+Twelve of the landmarks were symmetrical across the axis of symmetry.
+Because damage prevented the digitization of all the symmetric landmarks in some specimens, we reflected landmarks across the axis of symmetry (i.e.
 midline) prior to analysis and used the average position of each symmetrical pair.
@@
 We compared clustering solutions using the gap statistic, which is a measure of goodness of clustering \citep{Tibshirani2001a}.
+Gap statistic estimates were recalculated from bootstrap resampled 500 times in order to get confidence intervals for those estimates as described in \citet{Tibshirani2001a}.
 This method has been applied to morphometric data before and represents a comparatively simple approach for estimating latent classification \citep{Claude2008}.
 
-Only the \textit{E. marmorata} dataset was analyzed using unsupervised approaches. We conducted this analysis using the \texttt{cluster} package for R \citep{Maechler2013}.
+We only use the unsupervised approach with the \textit{E.
+marmorata} dataset was analyzed using unsupervised approaches.
+We conducted this analysis using the \texttt{cluster} package for R \citep{Maechler2013}.
 
@@
 The other two methods, however, are not.
-The set of possible predictors included the centroid size of each specimen along with the value of that specimen along all 22 PC axes.
+For the \textit{E.
+marmorata} dataset, the set of possible predictors included the centroid size of each specimen along with the value of that specimen along the first 22 PC axes.
+For the other two datasets, less than the maximum number of PCs were included in analysis because for each predictor included, at least one parameter is estimated.
+Given the lower sample sizes of these two datasets, we set the maximum number of PCs equal to the sample size divided by 10, rounded down.
 In all cases, the optimal subset of PCs included in the final model was chosen via maximum within-sample receiver operating characteristic values, explained below.
-Additionally, all of the following models assume that each of the predictors (PC axes) are statistically independent; this is a reasonable assumption because by definition principal components are statistically independent.
+
+All of the supervised approaches used here assume that each of the predictors (PC axes) are statistically independent; this is a reasonable assumption because by definition principal components are statistically independent.
 Using PCs as predictors in regression and classification methods is called PCA regression is a common technique for for dimensionality reduction and for accounting for correlation between variables \citep{Hastie2009}.
@@
 This relationship is especially useful in cases such as this study where misclassification needs to be minimized just as much as an accurate classification must be obtained.
-AUC ranges between 0.5 and 1, with 0.5 indicating classification no better than random and 1 indicating perfect classification \citep{Hastie2009}.
+AUC ranges between 0.5 and 1, with 0.5 indicating a classification no better than random and 1 indicating a perfect classification \citep{Hastie2009}.
 
@@
 In total, this produced nine different LDA scaling matrices.
+We implemented LDA models using the ``MASS'' R package \citep{Venables2002a}.
 
@@
 Similar to the odds ratios calculated from the coefficients of a logistic regression, the relative risk of a classification can be determined from the coefficients of the model.
-Multinomial logistic regression models were implemented using the ``nnet'' R package \citep{Venables2002a}.
+We implemented the multinomial logistic regression models using the ``nnet'' R package \citep{Venables2002a}.
 
@@
 This is a generally useful property for studies in which the goal is to describe and model the differences between classes and the relative importance of different predictors.
-Random forest models were fit using the ``randomForest'' R package \citep{Liaw2002}.
+We fit the random forest models using the ``randomForest'' R package \citep{Liaw2002}.
 In this analysis, we used 1000 subtrees to estimate the random forest model parameters.
@@
 
-Gap statistic values from the PAM clustering are mostly equivalent for all potential clusters, indicating that none of the proposed clusters are statistically different from each other (Fig.
+Gap statistic values from the PAM clustering are mostly equivalent for all potential clusters, indicating that none of the proposed clusters are statistically different from each other as evidenced by the confidence intervals around these estimates (Fig.
 \ref{fig:gap}).
-No clustering result is statistically greater than that with only a single cluster, because of this we favor the interpretation that there is no latent structure in the data detected in this analysis.
+No clustering result is statistically greater than that with only a single cluster, because of this we favor the interpretation that there is no latent structure in the data detected by this analysis.
 While the solution with two groupings does have the greatest mean gap statistic, there is no geographical signal in the results of this clustering solution (Fig.
@@
 Additionally, there is no correspondence between these two clusters and the sex of the specimens (\(\chi^{2}\) 0.7685, df = 1, \textit{p}-value 0.38).
-Because of this, we assert that there is no means of naturally partitioning \textit{E.
-marmorata} plastron shape into distinct subgroups without reference to external information.
+Because of this, we assert that 1) there is no means of naturally partitioning \textit{E.
+marmorata} plastron shape into distinct subgroups without reference to external information and 2) sexual dimorphism most likely is not confounding these analyses.
 
@@
   \caption{Results from PAM clustering of the Riemannian shape distance for 8 different number of clusters.
-Vertical lines are 1 standard deviation of the mean determined from 500 resamples.
-354 points are plotted, some of which overlap.}
+Vertical lines are 2 standard deviation of the mean determined from 500 resamples as defined in \citep{Tibshirani2001a}.}
   \label{fig:gap}
@@
 There is clearly no geographic signal in the data.
-Points are slightly transparent as 354 points are plotted but many of them overlap.}
+Points are slightly transparent as 354 points are plotted but many of them overlap.
+354 points are plotted, some of which overlap.}
   \label{fig:gap_map}
@@
 scripta scripta} to \textit{T.
-scripta elegans} indicates that these two subspecies are morphologically distinct enough to be classified with near perfect accuracy (Table \ref{tab:second_res}, with both in-sample and out-of-sample classification having AUC values of approximately 1 for all methods.
+scripta elegans} indicates that these two subspecies are sufficiently morphologically distinct to be classified with near perfect accuracy (Table \ref{tab:second_res}, with both in-sample and out-of-sample classification having AUC values of approximately 1 for all methods.
 The analysis of the seven morphologically-distinct species also demonstrates near perfect classification of both the in-sample and out-of-sample datasets (Table \ref{tab:second_res}).
@@
 \ref{fig:gen_hist}).
-These results demonstrate that when there are strong distinctions between the states of the classification schemes, the methods used here can recover them.
+These results demonstrate that when there are distinctions between the states of the classification schemes, the methods used here can recover them.
 % insert table here with the AUC values for in and out of sample results
@@
 
-The results of this study indicate that there is no clear grouping of \textit{E. marmorata} based on plastron shape.
-
-The unsupervised learning results indicate no grouping is optimal or particularly better than a single grouping, which is consistent with the results from the generalizations of the supervised learning models.
+The results of this study show that there is no clear grouping of \textit{E. marmorata} based on plastron shape.
+
+The unsupervised learning results indicate that no grouping is optimal or particularly better than a single grouping, which is consistent with the results from the generalizations of the supervised learning models.
 The classification schemes used in the supervised learning models correspond, loosely, to unsupervised learning solutions with multiple groups.
@@
 from hydrological regime) overwhelm morphological differentiation.
-This makes sense given that the plastron is involved in both protection and hydrodynamics, and not necessary mate choice \citep{Rivera2008,Rivera2011,Stayton2011,Rivera2014} and that shell shape in \textit{E.
+This makes sense given that shell shape is influenced by selection for both protection and streamlining, but not necessary mate choice \citep{Rivera2008,Rivera2011,Stayton2011,Rivera2014,Pollyb} and that shell shape in \textit{E.
 marmorata} is known to vary among populations inhabiting water bodies with different flow regimes \citep{Holland1992,Lubcke2007,Germano2009}.
@@
 
-Both the low AUC values (\(< 0.9\)) and the significant difference between the correctly and incorrectly classified observations support the conclusion that none of the hypothesized classification schemes are good descriptors of the observed plastral variation within \textit{E.
+Both the low out-of-sample AUC values (\(< 0.9\)) and the significant difference between the correctly and incorrectly classified observations support the conclusion that none of the hypothesized classification schemes are good descriptors of the observed plastral variation within \textit{E.
 marmorata}.
 
-Nevertheless, it is important to note that plastron shape is an extremely effective method for differentiating members of the other seven species we investigated.
+Nevertheless, it is important to note that plastron shape is an extremely effective method for differentiating classes in the additional datasets we investigated.
 The magnitude of shape differences between the species (measured as Procrustes distance between species' mean shapes) is approximately an order of magnitude greater than the differences between the \textit{E.
 marmorata} subgroups, and the machine learning methods had no trouble accurately classifying the specimens correctly.
-These results demonstrate that plastron shape is normally a good marker for species delineation in the closest relatives of \textit{E.
+However, the magnitude of the shape differences between the \textit{T.
+scripta} subspecies is comparable to theose separating the different \textit{E.
+marmorata} subgroups, yet even in this case the machine learning methods returned an almost perfect classification.
+These results demonstrate that plastron shape is normally a good marker for differentiating real subgroups in close relatives of \textit{E.
 marmorata}, and that our lack of results for \textit{E.
@@
 Invoking issues such as the role of the plastron in protection or the need for streamlining are insufficient because the other species are expected to be subject to similar constraints \citep{Stayton2011,Pollyb}.
+While it may seem counterintuitive that plastron shape is both useful for species delimitation but has weak or absent phylogenetic signal, it is important to note that these are different goals.
+While phylogenetically similar species may not be morphologically similar, the variation within a species is much less than the variation between species which leads to plastron shape being a useful tool for species delimitation but not phylogenetic reconstruction.
 
@@
 pallida} does not, on its own, preclude the recognition of this taxon.
-However, this apparent lack of congruence does prompt a reexamination of the methods and concepts that led to that taxonomic revision, especially considering that plastron shape is normally a strong indicator of species differences among emydines.
+However, this apparent lack of congruence does prompt a reexamination of the methods and concepts that led to that taxonomic revision, especially considering that plastron shape is demonstrably capable of differentiating species and subspecies among emydids.
 In other words, before we can assess the significance of the morphological non-diagnosablity, it is essential to evaluate the methods and concepts that led to the initial taxonomic revision.
@@
 However, \citet{Spinks2014} did not heed the caveats about species delimitation methods raised by \citet{Carstens2013}.
-In addition to specifically addressing the shortcomings of validation methods such as BPP that rely on guide trees and ``should be interpreted with caution,'' \citet{Carstens2013} also strongly emphasize that ``Inferences regarding species boundaries based on genetic data alone are likely inadequate, and species delimitation should be conducted with consideration of the life history, geographical distribution, morphology and behaviour (where applicable) of the focal system\dots'' These caveats evoke the development of the Unified Species Concept \citep{Dayrat2005a,DeQueiroz2007b}, Integrative Taxonomy \citep{Padial2010}, and other pluralist approaches to species delimitation.
+In addition to specifically addressing the shortcomings of validation methods such as BPP that rely on guide trees and ``should be interpreted with caution,'' \citet{Carstens2013} also strongly emphasized that ``Inferences regarding species boundaries based on genetic data alone are likely inadequate, and species delimitation should be conducted with consideration of the life history, geographical distribution, morphology and behaviour (where applicable) of the focal system\dots'' These caveats evoke the development of the Unified Species Concept \citep{Dayrat2005a,DeQueiroz2007b}, Integrative Taxonomy \citep{Padial2010}, and other pluralist approaches to species delimitation.
 None of these considerations were brought to bear on the \textit{E.
@@
 Because the geography, natural history, demonstrated genetic admixture of \textit{E.
-marmorata}, and comparisons with other morphologically diagnosable species conflict with the recognition of \textit{E.
+marmorata}, and comparisons with other morphologically diagnosable species and subspecies conflict with the recognition of \textit{E.
 pallida}, we hypothesize that our inability to classify the morphological data by proposed species is because \textit{E.
@@
 Finally, it is important to note that the data and analyses we present do not let us definitively say whether the apparent lack of morphological divergence within \textit{E.
-marmorata} truly reflects the presence of a single species, or if it is an artifact of plastron shape being a poor morphological marker for phylogenetic and phylogeographic divergences, because we could not carry out our morphometric analyses on the specimens from which the genetic data were obtained.
-The comparisons with the other emydine species suggests it is because \textit{E.
+marmorata} truly reflects the presence of a single species, or if it is an artifact of plastron shape being a poor morphological marker for phylogenetic and phylogeographic divergences in the case of \textit{E.
+marmorata}, because we could not carry out our morphometric analyses on the specimens from which the genetic data were obtained.
+The comparisons with the other emydid taxa suggest that it is because \textit{E.
 marmorata} is a single species.
@@
 Likewise, further investigation of the phylogeographic utility of plastron shape in other turtle species will help to clarify whether the lack of differentiation seen in \textit{E.
-marmoarata}, and the strong differentiation among the other emydines, is typical or an unusual case.
+marmoarata}, and the strong differentiation among the other emydids, is typical or an unusual case.
 

### 2016-03-03 doc/resubmit_letter.tex
ken comments. editor and first reviewer almost finished. second reviewer hurts me inside.

@@
 
-  %
-  %
-  %
+  The results of PAM clustering have been clarified.
+The biggest detail is remembering the confidence intervals around the mean.
+In the previous draft, only a single standard deviation was plotted which may lead to confusion.
+We've changed this so that two standard deviations are plotted as confidence intervals; this makes the result that no single classification scheme is optimal much more obvious.
 
@@
 
-  %
-  %
-  %
+  These typographic errors and content omissions were fixed where applicable. Additionally, the sources of the schemes have been identified early in the Methods section.
 
@@
 
- %
- %
- %
+  354 points are plotted it is just that many of them are plotted on top of each other. We've made these points slightly transparent to help understand how many points there are and where they occur.
+
+  Additionally, we've tested if the two-cluster PAM solution corresponded to the sexes of the specimens.
+While we did not have complete sex information for every specimen, we found that this clustering solution has no correspondence with sex.
 
@@
 marmorata for grouping members of this species into multiple subspecies, as there is in other turtle species which have already been grouped into multiple subspecies?''.
-And of course, a comparator is also necessary to answer the question: ``Can the methods used in this paper ever produce a taxonomically-appropriate clustering?''\\
+And of course, a comparator is also necessary to answer the question: ``Can the methods used in this paper ever produce a taxonomically-appropriate clustering?''\\\\
     The authors clearly intend the latter for their study.
@@
 So these results don’t really provide a useful comparator with the E.
-marmorata groupings; they just show that the methods used aren’t utterly abysmal. \\
+marmorata groupings; they just show that the methods used aren’t utterly abysmal. \\\\
     Instead, I think that the authors need to be able to answer the second question that I proposed above.
@@
 And if the methods cannot successfully group these taxa either, then the relevance of their failure to group E.
-marmorata will be put in its proper context. \\
+marmorata will be put in its proper context. \\\\
   The paper is otherwise strong.
@@
 
-  %
-  %
-  %
+  You are correct that our goal was the to address the second question.
+Following your suggestion, we've included an additional dataset looking at plastron shape variation between the two subspecies \textit{Trachemys scripta}.
+Between these two additional datasets we believe we have demonstrated that our approach can work when classes are actually distinct.
 

### 2016-03-03 R/trac_analysis.r
updated figures

@@ -196,9 +196,8 @@ lda.boot <- boot(data = tester, statistic = boot.roc, R = 1000)
 # make some output graphs
 test.gen <- melt(data.frame(RF = rf.boot$t, MLR = mnl.boot$t, LDA = lda.boot$t))
 gen.gg <- ggplot(test.gen, aes(x = value))
-gen.gg <- gen.gg + geom_histogram(aes(y = ..density..), 
-                                  position = 'identity')
-gen.gg <- gen.gg + labs(x = 'AUC')
+gen.gg <- gen.gg + geom_histogram(position = 'identity', binwidth = 1/50)
+gen.gg <- gen.gg + labs(x = 'AUC', y = 'Count')
 gen.gg <- gen.gg + facet_grid(variable ~ .)
 ggsave(file = '../doc/figure/trach_boot.png', plot = gen.gg, 
        width = 15, height = 10)

### 2016-04-04 doc/draft_turtle.tex
including all the digitization error stuff. half way there.

@@
 
+We estimated the possible effect of digitization error CITATIONS on our results by comparing within (replicated) specimen Procrustes distances to the distances between classification scheme centroids.
+10 randomly selected specimen both for this study and an additional four times.
+These 50 landmark configurations were then Procrustes superimposed.
+A range of four Procrustes distances were then calculated as the distance between the original configutation and each replicate configuration.
+These distances were then compared to the Procrustes distances between the centroids of the different classification schemes of \textit{E.
+marmorata} being analyzed.
+
 \subsection{Machine learning analyses}
@@
 \section{Results}
+
+\subsection{Effect of digitization}
+
+% within specimen (between replicates) procrustes distance
+% between specimens and centroid procrustes distance for each classification scheme
+% what graphs if any?
+%   just put them in an online appendix?
 

### 2016-04-04 R/digit_error.r

@@ -1,0 +1,56 @@
+library(shapes)
+library(geomorph)
+source('../R/df2array.r')
+source('../R/supervised_mung.r')
+
+procdist <- function(x, y) {
+  dis <- c()
+  for(ii in seq(dim(y)[3])) {
+    dif <- x - y[, , ii]  # diff of x coord, y coord
+    dif <- dif^2
+    dis[ii] <- sum(rowSums(dif))
+  }
+  dis
+}
+
+
+by.scheme <- list()
+scheme <- c('sh1', 'sh2', 'sh3', 'sh4', 'sh5', 'spinks')
+for(jj in seq(6)) {
+  by.class <- list()
+  sch <- as.character(adult[, scheme[ii]])
+  sch.t <- unique(sch)
+  for(ii in seq(length(sch.t))) {
+    sch.w <- sch == sch.t[ii]
+    by.class[[ii]] <- fit$rotated[, , sch.w]
+  }
+  by.scheme[[jj]] <- by.class
+}
+
+class.dist <- list()
+for(ii in seq(length(by.scheme))) {
+  class.mean <- llply(by.scheme[[ii]], mshape)
+  class.dist[[ii]] <- Map(function(x, y) procdist(x, y), 
+                          class.mean, by.scheme[[ii]])
+}
+# have the proc dist from mean shape for every class in every scheme
+
+
+
+rep.turt <- read.table('../data/replicate turtles fixed.txt', 
+                       header = FALSE, stringsAsFactors = FALSE)
+rep.turt <- df2array(rep.turt, n.land = 26, n.dim = 2)
+gpa.turt <- procGPA(rep.turt)
+
+#riemdist(rep.turt[, , 1], rep.turt[, , 2])
+#ssriemdist(rep.turt[, , 1], rep.turt[, , 2])  # scaled
+
+
+unit <- seq(1, 50, by = 5)
+within.dist <- list()
+for(ii in seq(length(unit))) {
+  within.dist[[ii]] <- procdist(gpa.turt$rotated[, , unit[ii]], 
+                         gpa.turt$rotated[, , (unit[ii] + 1):(unit[ii] + 4)])
+}
+
+

### 2016-04-04 R/shape_distance.r

@@ -9,8 +9,8 @@ riem.matrix <- function(config) {
   n <- dim(config)[3]
 
   pr <- combn(n, 2)
-  dis <- apply(pr, 2, function(x) riemdist(config[, , x[1]], 
-                                           config[, , x[2]]))
+  dis <- apply(pr, 2, function(x) ssriemdist(config[, , x[1]], 
+                                             config[, , x[2]]))
   mat <- matrix(ncol = n, nrow = n)
   for(ii in seq(ncol(pr))) {
     mat[pr[1, ii], pr[2, ii]] <- dis[ii]

### 2016-04-05 doc/draft_turtle.tex
fixed methods for digitzation analysis

@@
 These 50 landmark configurations were then Procrustes superimposed.
-A range of four Procrustes distances were then calculated as the distance between the original configutation and each replicate configuration.
-These distances were then compared to the Procrustes distances between the centroids of the different classification schemes of \textit{E.
+A range of four Procrustes distances were then calculated as the average of the pairwise distances between each of the replicate configurations.
+These distances were then compared to the average pairwise Procrustes distance between the class centroids for each classification schemes of \textit{E.
 marmorata} being analyzed.

### 2016-05-04 doc/turtles_statistical_learning.tex
overhaul of methods section; essentially redoing the papers innards

--- 2016-04-05 doc/draft_turtle.tex
@@
 \title{How cryptic is cryptic diversity? Machine learning approaches to classifying morphological variation in the Pacific Pond Turtle (\textit{Emys marmorata})}
+%\title{Supervised learning approaches to quantifying cryptic diversity: picking amongst taxonomic hypotheses of Pacific Pond Turtle (\textit{Emys marmorata})}
 \author[1]{Peter D Smits}%\thanks{psmits@uchicago.edu}}
@@
 \begin{abstract}
-  We investigate the morphometric identification of cryptic species using machine learning approaches by examining their implications for a recently proposed cryptic turtle species (\textit{Emys pallida}).
-We collected landmark-based morphometric data from 354 adult \textit{E.
-marmorata/``pallida''} museum specimens.
-We assigned a classification to each specimen for six different binning schemes based on geographic occurrence data recorded in museum collection archives.
-We used multiple machine learning methods, both unsupervised and supervised, to compare different classification hypotheses and asked whether it is possible to determine which amongst a set of classification hypotheses is best.
-  In addition, we applied the above approach to two additional datasets: two well-supported subspecies of \textit{Trachemys scripta} and a set of seven unambiguously distinct species closely related to \textit{E.
-marmorata}.
-The results of this study indicate that there is no clear grouping of \textit{E.
-marmorata/``pallida''} based on plastron shape.
-In contrast, the analysis of the other datasets demonstrate a near perfect classification, which demonstrates that application of the methods to plastron shape data can recover correct results in similar situations.
-Explanations for the lack of grouping in \textit{E.
-marmorata} include that possibility that genetic differentiation is not associated with plastron shape variation in this species complex and/or that local selective pressures (e.g., from hydrological regime) overwhelm morphological differentiation.
-A reconsideration of the methods used to delimit \textit{E.
-``pallida,''} the lack of barriers to gene flow, the strong evidence for widespread admixture between lineages, and the fact that plastron shape can be used to differentiate other emydid species and subspecies suggest that its lack of diagnosability most likely reflects the non-distinctiveness of this proposed taxon.
 \end{abstract}
@@
 
-% cryptic diversity
-%   most species are still deliminated solely based on morphology
-Molecular systematics has repeatedly demonstrated the existence of cryptic species that can only be diagnosed using genetic data \citep{Stuart2006,Bickford2007,SchlickSteiner2007,Pfenninger2007,Clare2011,Funk2012}.
-In attempts to streamline the documentation of biodiversity, several methods of species delimitation that rely almost entirely on genetic data have recently been proposed \citep{Pons2006,Carstens2010,Hausdorf2010,O'Meara2010,Yang2010b,Huelsenbeck2011b}.
-Although strong caveats on the utility of these methods have been raised \citep{Bauer2000,Carstens2013}, they are nevertheless being used to name species \citep{Leache2010,Spinks2014}.
-
-The majority of extant taxa, and almost all extinct taxa, are delimited by morphology alone.
-This disjunction complicates interpretations of variation and diversity in deep time, as apparent morphological stasis may not reflect the true underlying diversity \citep{Eldredge1972,Gould1977a,Hunt2008,VanBocxlaer2013}.
-Similarly, for many museum specimens of extant taxa (e.g.
-those preserved in formalin), it is difficult to acquire the genetic data needed for non-morphological species delimitation methods.
-
-These considerations have sparked interest in whether geometric morphometric analyses can capture fine-scale variation that can be used for identifying cryptic species.
-This would make the task of identifying and maintaining endangered or conserved groups much easier and could contribute to improved classifications of extinct taxa and populations.
-Most such studies focus on using morphometrics to discover differences between taxa that were identified by other means \citep{Polly2003,Zelditch2004,Gaubert2005b,Gunduz2007,Polly2007a,Demandt2009}.
-Additionally, there has been work on automated taxon identification and classification of taxa into groups \citep{Baylac2003,Dobigny2003,MacLeod2007,VandenBrink2011}.
-
-Here, we investigate the morphometric identification of cryptic species using machine learning approaches.
-In particular, we ask whether it is possible to determine which amongst a set of classification hypotheses is best and examine the implications of the results for a recently proposed set of cryptic turtle species.
-
-% past work on automatic taxon identification and older approaches to classifying taxa
-% why use machine learning methods
-\subsection{Background and study system}
-Machine learning is an extension of known statistical methodology \citep{Hastie2009} that emphasizes high predictive accuracy and generality at the expense of the interpretability of individual parameters.
-The basic statistical mechanics are supplemented by randomization, sorting, and partitioning algorithms, along with the maximization or minimization of summary statistics, in order to best estimate a general model for all data, both sampled and unsampled \citep{Hastie2009}.
-Machine learning approaches have found use in medical research, epidemiology, economics and automated identification of images such as handwritten zip codes \citep{Hastie2009}.
-The two major classes of machine learning methods are unsupervised and supervised learning.
-Unsupervised learning methods are used with unlabeled data where the underlying structure is estimated; they are analogous to clustering and density estimation methods \citep{Kaufman1990}.
-Supervised learning methods are used with labeled data where the final output of data is known and the rules for going from input to output are inferred.
-These are analogous to classification and regression models \citep{Breiman1984}.
-Our application of the approaches used in this study illustrates only a sampling of the various methods available for clustering observations and fitting classification models.
-
-Geometric morphometric approaches to identifying differences in morphological variation between different classes, including cryptic species, mostly have used methods like linear discriminate analysis and canonical variates analysis \citep{Polly2003,Zelditch2004,Gaubert2005b,Gunduz2007,Polly2007a,Francoy2009,Sztencel-Jabonka2009,MitrovskiBogdanovic2013}.
-Because of their similarity to multivariate approaches like principal components analysis (PCA), these methods are comparatively straightforward ways of understanding the differences in morphology between classes.
-They also benefit from producing results that can be easily visualized, which aids in the interpretation and presentation of data and results.
-Most previous morphometric studies did not assess which amongst a set of alternative classification hypotheses was optimal.
-For example, studies such as those of \citet{Caumul2005a} and \citet{Polly2007a} focused on comparing different aspects of morphology and their fidelity to a classification scheme instead of comparing the fidelity of one aspect of morphology to multiple classification schemes.
-In this context, the study of \citet{Cardini2009a} is noteworthy because they compared morphological variation in marmots at the population, regional, and species level and determined the fidelity of shape to divisions at each of these levels.
-
-Here, we used multiple machine learning methods, both unsupervised and supervised, to compare different classification hypotheses.
-These methods provide different advantages for understanding how to classify taxa, as well as the accuracy of the resulting classifications.
-Although machine learning methods such as neural networks have been applied to studying shape variation \citep{Baylac2003,Dobigny2003,MacLeod2007,VandenBrink2011}, including in the context of automated taxon identification and classification of groups, the number of cases remains limited.
-In the current study, we not only consider pure classification accuracy but also use a statistic of classification strength that reflects the rate at which taxa are both accurately and inaccurately classified.
-
-We analyzed the problem of whether there are distinct subspecies or cryptic species within the western pond turtle, \textit{Emys marmorata} \citep{Baird1852} (formerly \emph{Clemmys marmorata}; see \citealp{Feldman2002}).
-\textit{Emys marmorata} is distributed from northern Washington State, USA to Baja California, Mexico.
-Traditionally, \textit{E.
-marmorata} was classified into two named subspecies: the northern \textit{E.
-marmorata marmorata} and the southern \textit{Emys marmorata pallida} \citep{Seeliger1945}, with a central Californian intergrade zone in between.
-\textit{Emys marmorata marmorata} is differentiated from \textit{E.
-marmorata pallida} by the presence of a pair of triangular inguinal scales and darker neck markings.
-The triangular inguinal plates can sometimes be present in \textit{E.
-marmorata pallida} although they are considerably smaller.
-\citet{Seeliger1945} did not formally include the Baja California populations of \textit{E.
-marmorata} in either taxon, implying the existence of a third distinct but unnamed subspecies.
-
-Previous work on morphological variation in \textit{E.
-marmorata} has focused primarily on differentiation between populations over a portion of the species' total range \citep{Lubcke2007,Germano2008,Germano2009,Bury2010}; comparatively few studies have included specimens from across the entire range \citep{Holland1992}.
-Most of these studies considered how local biotic and abiotic factors may contribute to differences in carapace length and found that size can vary greatly between different populations \citep{Lubcke2007,Germano2008,Germano2009}.
-There also has been interest in size-based sexual dimorphism in \textit{E.
-marmorata} \citep{Holland1992,Lubcke2007,Germano2009}, with males being on average larger than females based on total carapace length and other linear measurements.
-However, the quality of size as a classifier of sex can vary greatly between populations \citep{Holland1992} because of the magnitude of size differences among populations \citep{Lubcke2007,Germano2009}.
-The effect of sexual dimorphism on shape, \textit{sensu} \citet{Kendall1977a}, has not been assessed \citep{Holland1992,Lubcke2007,Germano2008}.
-
-Of particular relevance in the context of cryptic diversity in \textit{E.
-marmorata} is the morphometric analysis of carapace shape carried out by \citet{Holland1992}, who compared populations of \textit{E.
-marmorata} from three areas of the species range.
-This study concluded that geographic distance was a poor indicator of morphological differentiation, and instead hypothesized that geographic features such as breaks between different drainage basis are probably more important barriers to dispersal and interbreeding.
-Additionally, \citet{Holland1992} suggested that morphological differences were more pronounced as the magnitude of barriers and distance increased, but this variation required many variables to adequately capture, implying only very subtle morphological differentiation between putatively distinct populations.
-Finally, Holland concluded that \textit{E.
-marmorata} is best classified as three distinct species: a northern species, a southern species, and a Columbia Basin species.
-This classification is similar to that of \citet{Seeliger1945}, except elevated to the species level and without recognition of a distinct Baja species.
-
-More recently, the phylogeography of \textit{E.
-marmorata} and the possibility of cryptic diversity was investigated using molecular data \citep{Spinks2005,Spinks2010,Spinks2014}.
-Based on mitochondrial DNA, \citet{Spinks2005} recognized four subclades within \textit{E.
-marmorata}, a northern clade, a San Joaquin Valley clade, a Santa Barbara clade, and a southern clade.
-Analyses with nuclear DNA \citep{Spinks2010} and single-nucleotide polymorphism (SNP) data suggest a primarily north--south division in \textit{E.
-marmorata}, although these datasets differed from that of \citet{Spinks2005} in the location of the break point.
-All three studies discussed the potential taxonomic implications of their results, with \citet{Spinks2014} going so far as to strongly advocate for the recognition of at least two species (\emph{E.
-marmorata} and \emph{E.
-pallida}), and a possible third based on populations in Baja California.
-However, they did not discuss in detail the morphological characters that would help to diagnose these species beyond those specified by \citet{Seeliger1945}.
-Given that these characters are variable within the proposed species, and that \citet{Holland1992} described shell shape variation that might be consistent with this taxonomy, a geometric morphometric analysis of shell shape might provide a reliable way to diagnose groups (whether species or subspecies) within \textit{E.
-marmorata}.
-
-In this study, we attempt to estimate the best classification scheme of \textit{E.
-marmorata} based on variation in plastron (ventral shell) shape in order to determine whether this character is consistent with any of the past divisions based on other morphological features or molecular data.
-We are particularly interested in whether any aspect of plastron shape can be used to reliably diagnose the proposed species of \citet{Spinks2014}, and if so, the nature of that shape variation.
-
-\uppercase{Why did we analyze plastron shape?}
-
-Because of unclear geographic boundaries between subgroups of \textit{E.
-marmorata}, we compare multiple hypotheses of morphologically-- and molecularly--based classification.
-We hypothesize that if morphological variation corresponds to class assignment, it should be possible to determine the best classification hypothesis of \textit{E.
-marmorata} from amongst multiple candidate hypotheses.
-However, if morphological variation does not correspond to any of the standing hypothesis, then supervised learning model generalization performance will be poor.
-
-In order to test the efficacy of our approach, we analyzed two additional datasets: two subspecies of \textit{Trachemys} (\textit{T.
-scripta scripta} and \textit{T.
-scripta elgans}), and seven distinct species from the subfamily Emyidinae.
-The former dataset represents a situation similar to the \textit{E.
-marmorata} scenario except there are two valid subspecies CITATIONS.
-The latter dataset represents a set of well-defined morphologically distinct species.
-Both of these datasets help constrain the effectiveness of the methods used here and can help in interpreting the \textit{E.
-marmorata} scenario.
-
 \section{Materials and Methods}
 \subsection{Specimens, sampling, morphometrics}
-Three different landmark-based morphometric datasets describing plastron variation were assembled for this analysis: \textit{E.
-marmorata} specimens from across its geographic range, \textit{T.
-scripta} specimens from both subspecies, and specimens from seven distinct emydine species.
+Three different landmark-based morphometric datasets describing plastron variation were assembled for this analysis: specimens from seven distinct emydine species, \textit{T.
+scripta} specimens from both subspecies, and \textit{E.
+marmorata} specimens from across its geographic range.
 We chose to focus on adults because significant changes in plastron shape occur over the course of ontogeny in \textit{E.
 marmorata} and other emydines \citep{Angielczyk2013a}.
+% three turtle datasets
+%   seven species
+%   Trachemys species
+%   Emys complex
+% 26 landmarks
 
-The first dataset included 354 adult \textit{E.
+The first dataset is a compilation of 101 specimens of \textit{T.
+scripta}: 51 specimens of \textit{T.
+scripta scripta} and 50 specimens of \textit{T.
+scripta scripta}.
+These landmark data are new to this study.
+
+The second dataset, we analyzed 578 total specimens from the following species: \textit{Emys blandigii}, \textit{Terrapene coahuila}, \textit{Clemmys guttata}, \textit{Glyptemys insculpta}, \textit{Glyptemys muhlenbergii}, \textit{Emys orbicularis}, and \textit{Terrapene ornata}.
+Like the first data set, these specimens are a subset of those used in \citet{Angielczyk2011} and \citet{Angielczyk2013a}.
+
+The final dataset dataset included 354 adult \textit{E.
 marmorata} museum specimens; a subset of those included in \citet{Angielczyk2007}, \citet{Angielczyk2011}, and \citet{Angielczyk2013a}.
@@
 The scheme names are as follows: Mito 1 and 2 correspond to \citet{Spinks2005}, Mito 3 corresponds to \citet{Spinks2010}, Morph 1 and Morph 2 correspond to \citet{Holland1992}, and Nuclear corresponds to \citet{Spinks2014}.
-
-The second dataset is a compilation of 101 specimens of \textit{T.
-scripta}: 51 specimens of \textit{T.
-scripta scripta} and 50 specimens of \textit{T.
-scripta scripta}.
-These landmark data are new to this study.
-For the third dataset, we analyzed 578 total specimens from the following species: \textit{Emys blandigii}, \textit{Terrapene coahuila}, \textit{Clemmys guttata}, \textit{Glyptemys insculpta}, \textit{Glyptemys muhlenbergii}, \textit{Emys orbicularis}, and \textit{Terrapene ornata}.
-Like the first data set, these specimens are a subset of those used in \citet{Angielczyk2011} and \citet{Angielczyk2013a}.
 
@@
 
+\subsection{Biasing effects}
+% digitization
 We estimated the possible effect of digitization error CITATIONS on our results by comparing within (replicated) specimen Procrustes distances to the distances between classification scheme centroids.
@@
 
-\subsection{Machine learning analyses}
-\subsubsection{Unsupervised learning}
-In order to preserve the relationship between all landmark configurations in shape space, we measured the dissimilarity between observations using Kendall's Riemannian shape distance or \(\rho\) \citep{Kendall1984a,Dryden1998a}.
-We chose this metric because shape space, or the set of all possible shape configurations following Procrustes superimposition, is a Riemannian manifold and thus non-Euclidean \citep{Dryden1998a}.
-\(\rho\) varies between 0 and \(\pi / 2\) when there is no reflection invariance, which should not be a concern in the case of the half plastral landmark configurations used in the study.
+% sex
+% permutation test
+%   is the distance between sex means bigger than expected by random?
 
-We divisively clustered the shape dissimilarity matrix using partitioning around mediods clustering (PAM), a method similar to \textit{k}-means clustering except that instead of minimizing the sum of squared Euclidean distances between observations and centroids, the sum of squared dissimilarities between observations and mediods is minimized \citep{Kaufman1990}.
-Because the optimal number of clusters of shape configurations in the study was unknown, being possibly three, four, or some other value, we estimated clustering solutions in which the number of clusters varied between one and eight.
-We compared clustering solutions using the gap statistic, which is a measure of goodness of clustering \citep{Tibshirani2001a}.
-Gap statistic estimates were recalculated from bootstrap resampled 500 times in order to get confidence intervals for those estimates as described in \citet{Tibshirani2001a}.
-This method has been applied to morphometric data before and represents a comparatively simple approach for estimating latent classification \citep{Claude2008}.
+\subsection{Supervised learning approaches}
+% predictors/features/covariates
+%   1:25 PCs
+%   size
+%   size X PC1
+% analogy to PCA regression
+The maximum set of possible predictors or features used for any model are the first 25 principal components (PCs), scaled centroid size, and the interaction between scaled centroid size and PC 1.
+Size and the interaction between size and PC 1 were included as predictors in order to account for a possible interaction between size and shape over the duration of an individual as well as potential size differences between classes, even if this is unlikely \citep{Seeliger1945,Holland1992}.
+We say ``maximum set'' because the best or selected models based on 5-fold cross-validation does need not to, nor will they likely, include all predictors possible (see below).
 
-We only use the unsupervised approach with the \textit{E.
-marmorata} dataset was analyzed using unsupervised approaches.
-We conducted this analysis using the \texttt{cluster} package for R \citep{Maechler2013}.
+This approach is in many ways analogous to PCA regresion.
+PCA regression takes advantage of two aspects of PCA for improving regression fit \citep{Hastie2009}.
+Because the PCs of shape are by definition orthagonal, allowing them to easily as independent predictors or features of class membership without fear of colinearity.
 
-\subsubsection{Supervised learning}
-We used three different supervised learning, or classification, approaches: linear discriminate analysis, multinomial logistic regression, and random forests.
-Linear discriminate analysis, also known as canonical variate analysis, is commonly used in studies of geometric morphometric data \citep{Zelditch2004,Claude2008,Mitteroecker2011}.
-The other two methods, however, are not.
-For the \textit{E.
-marmorata} dataset, the set of possible predictors included the centroid size of each specimen along with the value of that specimen along the first 22 PC axes.
-For the other two datasets, less than the maximum number of PCs were included in analysis because for each predictor included, at least one parameter is estimated.
-Given the lower sample sizes of these two datasets, we set the maximum number of PCs equal to the sample size divided by 10, rounded down.
-In all cases, the optimal subset of PCs included in the final model was chosen via maximum within-sample receiver operating characteristic values, explained below.
+% auc of roc as performance measure
+%   relationship between false positive rate and true positive rate
+%   varies between 0.5 and 1 (random -- perfect)
+%   better performance than accuracy when groups unbalanced
+In classification studies, such as this one, a common metric of performance is the receiver operating characteristic (ROC) which is the relationship between the false and true positive rates \citep{Hastie2009}.
+The area under the ROC curce (AUC) is then the derived estimate of the model performance; AUC ranges from 0.5 to 1 which corresponding to performance similar to random guesses and perfect classificationrates, respectively \citep{Hastie2009}.
+Both ROC and AUC are preferrable to simple classificaiton accuracy when class membership is unbalanced, as it is in these analyses \citep{Hastie2009}.
+The standard ROC and AUC calculations are defined only for binary classifications.
+To generalize this approach for mulitple response classes, we used an all-against-one strategy where the model AUC is the average of the AUC values from the multiple binary comparisons of one class compared to all others \citep{Hand2001}.
 
-All of the supervised approaches used here assume that each of the predictors (PC axes) are statistically independent; this is a reasonable assumption because by definition principal components are statistically independent.
-Using PCs as predictors in regression and classification methods is called PCA regression is a common technique for for dimensionality reduction and for accounting for correlation between variables \citep{Hastie2009}.
-The PCs are a coordinate system where combinations of axes and ranges of those axes can describe groupings; we are not assuming that evolution happens ``along'' PC axes \citep{Uyeda2015} but use them as a coordinate system for describing observed shapes.
+% training and testing paradigm
+%   training dataset
+%     for multiple models with between 3 and 28 predictors
+%     5-fold cross validation of each model
+%     best model had greatest mean AUC value
+%     selected model is most parsimonious model within 1 SD of ROC of best model
+%   testing dataset
+%     estimate class of out-of-sample testing dataset
+%     compare average AUC across models for each scheme
+% caret package for R
+We adopted a training and testing paradigm for selecting parsimonious models and estimating their overall error rates \citep{Hastie2009,Kuhn2013}.
+Within-sample model performance is inherently biased upwards, so model evaluation requires overcoming this bias.
+With very large sample sizes, as in this study, part of the sample can be used as the ``training set'' and the remainder acts as the ``testing set.'' In this approach, following all cleaning and vetting, the data is split into a training dataset and a testing dataset.
+The former is used for fitting the model where as the later is used for measuring model performance, a process called model generalization.
+In this analysis, we used 80\% of samples as the training set while the remaining 20\% were used as the testing set.
 
-In classification studies, such as this one, a common metric of performance is area under the receiver operating characteristic curve (AUC).
-AUC is an estimate of the relationship between the false positive and true positive rates, as opposed to just the true positive rate (accuracy).
-This relationship is especially useful in cases such as this study where misclassification needs to be minimized just as much as an accurate classification must be obtained.
-AUC ranges between 0.5 and 1, with 0.5 indicating a classification no better than random and 1 indicating a perfect classification \citep{Hastie2009}.
+For a given supervised learning method, we compared the fit of 27 models as the average AUC from 10 rounds of 5-fold cross-validation.
+Cross-validation is an approach for estimating the average out-of-sample predictive error of a model by simulating out-of-sample data from the training data itself \citep{Hastie2009}.
+In a single round of \(k\)-fold cross-validation, the training data is divided into \(k\) blocks where the model is fit to \(k - 1\) blocks and the values of the \(k\)th block are predicted; this is then repeated for all combinations of blocks.
+Within each round, the predictive performance metrics is averaged across all folds.
+Finally, the predictive performance metric is the averaged across all rounds of \(k\)-fold cross-validation.
+This process was implemented using the R package \texttt{caret} CITATION.
 
-The standard AUC calculation is defined for binary classifications, however in our application there are multiple categories.
-The alternative calculation that we used follows an all-against-one strategy where the individual AUC values for each class versus all others are averaged to produce a multiclass AUC \citep{Hand2001}.
-To estimate confidence intervals on the out-of-sample AUC values, we performed a nonparametric bootstrap in which the true and estimated classifications were resampled with replacement 1000 times.
+For a given supervised learning method, the ``best'' trained model is that the highest mean AUC as estimated from 5-fold cross-validation.
+The selected or final model, however, is the next most parsimonious model that is within one standard error of the best model; this is a variant on the ``one-standard error'' rule from \citet{Hastie2009}.
 
-Linear discriminate analysis (LDA) attempts to find a linear combination of predictors that best model two or more classes.
-LDA is very similar to PCA except that instead of finding the linear combination of features that maximize the amount of explained variance in the data, LDA maximizes the differences between classes.
-The results from this approach produce a transformation matrix by which the original features can be transformed to reflect the best discrimination between the classes.
-In this study, we applied LDA to the eigenscores from a subset of the total number of PCs, ranging from two to six in increasing order of complexity.
-In total, this produced nine different LDA scaling matrices.
-We implemented LDA models using the ``MASS'' R package \citep{Venables2002a}.
-
-Multinomial logistic regression is an extension of logistic regression, where instead of a binary response there are three or more response classes \citep{Venables2002a}.
-Similar to the odds ratios calculated from the coefficients of a logistic regression, the relative risk of a classification can be determined from the coefficients of the model.
-We implemented the multinomial logistic regression models using the ``nnet'' R package \citep{Venables2002a}.
-
-Random forest models are an extension of classification and regression trees (CART) \citep{Breiman1984,Breiman2001}.
-In a random forest model, many CARTs are built from a random subsample of both the features and the observations (specimens).
-This process is then repeated many times and the parameters of the final model are chosen as the mode of estimates from the distribution of CARTs \citep{Breiman2001}.
-In addition to classifying the observations, this procedure allows for the features to be ranked in order of importance.
-This is a generally useful property for studies in which the goal is to describe and model the differences between classes and the relative importance of different predictors.
-We fit the random forest models using the ``randomForest'' R package \citep{Liaw2002}.
-In this analysis, we used 1000 subtrees to estimate the random forest model parameters.
-
-We estimated the best set of predictors in the random forest model using a recursive feature selection algorithm, with AUC as our optimality criterion.
-Following the iterative backwards selection algorithm implemented in \texttt{caret} \citep{KuhnMAN2013}, the maximum number of features possible were included in the initial model, their importance ranked, and the AUC of the model calculated.
-The lowest-ranked feature was then removed, and the AUC of the model recalculated.
-Because PCs were kept in order of importance and not in relation to the amount of variance each PC described, the PCs are not included in the order of ascending eigenvalue.
-
-The ultimate measure of model fit is accurately predicting the values of unobserved samples \citep{Hastie2009,Kuhn2013}.
-Within-sample performance is inherently biased upwards, so model evaluation requires overcoming this bias.
-With very large sample sizes, as in this study, part of the sample can be used as the ``training set'' and the remainder acts as the ``testing set.'' The former is used for fitting the model where as the later is used for measuring model performance, a process called model generalization.
-In this analysis, we used 75\% of samples as the training set while the remaining 25\% were used as the testing set.
-Testing and training sets are determined completely at random within each class and with respect to shape.
-Results were not affected by the individual specimen class assignments to the testing or training sets.
-
-Supervised learning methods were applied to all three datasets.
+% use multiple methods
+%   multinomial logistic regression (nnet)
+%   linear discriminate analysis (MASS)
+%   penalized discriminate analysis (mda)
+%   neural network (nnet)
+%   random forest (randomForest)
+% each method has different assumptions and treat data differently
+%   all assume that predictors have additive effect (e.g. independent)
+The supervised learning methods used here were multinomial logistic regression (MLR), linear discrminiate analysis (LDA), penalized discriminate analysis (PDA), single-hidden-layer neutral network (NN), and random forests (RF).
+Each of these methods makes different assumptions, treat data differently, and can produce very different classification results \citep{Hastie2009}.
+The common assumption of all of these methods is that the predictors or features are independent and/or have additive effects on prediction \citep{Hastie2009}.
 
@@
 
-\subsection{Effect of digitization}
-
-% within specimen (between replicates) procrustes distance
-% between specimens and centroid procrustes distance for each classification scheme
-% what graphs if any?
-%   just put them in an online appendix?
-
-\subsection{Unsupervised learning}
-
-Gap statistic values from the PAM clustering are mostly equivalent for all potential clusters, indicating that none of the proposed clusters are statistically different from each other as evidenced by the confidence intervals around these estimates (Fig.
-\ref{fig:gap}).
-No clustering result is statistically greater than that with only a single cluster, because of this we favor the interpretation that there is no latent structure in the data detected by this analysis.
-While the solution with two groupings does have the greatest mean gap statistic, there is no geographical signal in the results of this clustering solution (Fig.
-\ref{fig:gap_map}).
-Additionally, there is no correspondence between these two clusters and the sex of the specimens (\(\chi^{2}\) 0.7685, df = 1, \textit{p}-value 0.38).
-Because of this, we assert that 1) there is no means of naturally partitioning \textit{E.
-marmorata} plastron shape into distinct subgroups without reference to external information and 2) sexual dimorphism most likely is not confounding these analyses.
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[height = 0.5\textheight, width = \textwidth, keepaspectratio = true]{figure/gap_res}
-  \caption{Results from PAM clustering of the Riemannian shape distance for 8 different number of clusters.
-Vertical lines are 2 standard deviation of the mean determined from 500 resamples as defined in \citep{Tibshirani2001a}.}
-  \label{fig:gap}
-\end{figure}
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[height = 0.5\textheight, width = \textwidth, keepaspectratio = true]{figure/gap_map}
-  \caption{Comparison of geographic distribution of clustered observations from the 2 clustering PAM solution.
-Colour and shape correspond to each of the groups.
-There is clearly no geographic signal in the data.
-Points are slightly transparent as 354 points are plotted but many of them overlap.
-354 points are plotted, some of which overlap.}
-  \label{fig:gap_map}
-\end{figure}
+\subsection{Biasing effects}
 
 \subsection{Supervised learning}
-% model selection/fit
-AUC--based model selection revealed some important patterns of variation and congruence between the classification schemes and the actual data.
-Generally, the best performing models tended to include as many PCs as possible (Fig.
-\ref{fig:sel}).
-Note that the best random forest models were determined via recursive feature selection, so PCs were not included in order of percent variance explained.
-That almost all LDA and multinomial logistic regression models were as complex as possible indicates that the differences between the different groups within each classification scheme are very small.
 
-As part of fitting a random forest model, a ranking of variable importance also is determined.
-Interestingly, the order of variable importance is not the same as the order of the PCs (Fig.
-\ref{fig:var_imp}).
-This means that the variance describing the differences between the classes does not align with the major axes of variance (i.e.
-the PCs).
-Such a result is expected if variation between classes was extremely fine grained and not a part of the principal form or function of the plastron.
-Moreover, this result is consistent with the results from the AUC--based model selection for the multinomial logistic regression and LDA models.
-
-Observed AUC values for all of the optimal models are not exceptionally high (Fig.
-\ref{fig:sel}).
-In most cases the different proposed classification schemes are generally poor descriptors of the observed variation.
-It appears that the data set is overwhelmed by noise, making any accurate classifications difficult at best.
-This observation is cemented with the generalizations of the models to the testing data set (Fig.
-\ref{fig:gen_hist}).
-
-Mean AUC values for the model generalizations, in most cases, are approximately equal to the observed AUC values from the training data set (Table \ref{tab:comp}).
-The  cases in which the AUC from the  generalizations is less than the observed indicate poor model fit and a poor classification scheme.
-Comparison of AUC values from the model generalizations do not indicate a clear ``best'' classification scheme (Fig.
-\ref{fig:gen_hist}).
-
-\begin{figure}[h]
+\begin{figure}[ht]
   \centering
-  \includegraphics[height = 0.8\textheight, width = 0.8\textwidth, keepaspectratio = true]{figure/var_imp}
-  \caption{Variable importance from the random forest models for each of the six classification schemes.
-Importance is measured as the mean decrease in Gini Index, which is a measure of the strength by which that variable determines CART structure.
-Indices that are farther to the right indicate greater variable importance.}
-  \label{fig:var_imp}
+  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/other_model_sel}
+  \caption{CAPTION}
+  \label{fig:other_sel}
 \end{figure}
@@
   \centering
-  \includegraphics[height = 0.8\textheight, width = \textwidth, keepaspectratio = true]{figure/gen_res}
-  \caption{Bootstrap distributions for genearlized AUC values for each of the classification schemes.
-Columns correspond to different modeling approachs: multinomial logistic regression (MLR), random forest (RF), linear discriminate analysis (LDA), and LDA using best variables from random forest.
-Rows correspond to different classification schemes.
-Each distribution corresponds to 1000 bootstrap replicates.}
-  \label{fig:gen_hist}
+  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/other_oos_sel}
+  \caption{CAPTION}
+  \label{fig:other_oos}
 \end{figure}
@@
   \centering
-  \includegraphics[height = 0.8\textheight, width = 0.8\textwidth, keepaspectratio = true]{figure/sel_val}
-  \caption{Graphical representation of the AUC values from model selection for multinomial logistic regression and linear discriminate analysis, respectively.
-AUC model selection is based on greatest AUC value.
-The horizontal axis corresponds to the cummulative number of axes included in the model of interest.
-A highlighted point corresponds to the AUC best model for that classification scheme.}
-  \label{fig:sel}
+  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/emys_model_sel}
+  \caption{CAPTION}
+  \label{fig:emys_sel}
 \end{figure}
-
-\clearpage
-
-\begin{sidewaystable}[ht]
-  \centering
-  \caption{AUC values for the best model of each classification scheme for both the observed (training) data and the generalized (testing) data.
-Results from all three different supervised learning approaches are shown here.
-AUC values range between 0.5 and 1.
-}
-  \begin{tabular}{l | r r r | r r r | r r r | r r r }
-    \hline
-    Scheme & \multicolumn{3}{c |}{RF} & \multicolumn{3}{c |}{MLR} & \multicolumn{3}{c |}{LDA} & \multicolumn{3}{c}{LDA + RF} \\
-    & Train & Test & SD(Test) & Train & Test & SD(Test) & Train & Test & SD(Test) & Train & Test & SD(Test) \\ 
-    \hline
-    Mito 1 & 0.67 & 0.74 & 0.05 & 0.90 & 0.67 & 0.10 & 0.87 & 0.73 & 0.05 & 0.67 & 0.73 & 0.05 \\ 
-    Mito 2 & 0.77 & 0.70 & 0.05 & 0.91 & 0.58 & 0.05 & 0.89 & 0.73 & 0.05 & 0.77 & 0.74 & 0.05 \\ 
-    Mito 3 & 0.74 & 0.66 & 0.05 & 0.91 & 0.54 & 0.05 & 0.88 & 0.65 & 0.05 & 0.74 & 0.66 & 0.05 \\ 
-    Morph 1 & 0.81 & 0.71 & 0.06 & 0.86 & 0.73 & 0.05 & 0.85 & 0.74 & 0.05 & 0.81 & 0.79 & 0.05 \\ 
-    Morph 2 & 0.68 & 0.72 & 0.05 & 0.89 & 0.67 & 0.04 & 0.86 & 0.73 & 0.05 & 0.68 & 0.72 & 0.05 \\ 
-    Nuclear & 0.64 & 0.63 & 0.06 & 0.84 & 0.59 & 0.06 & 0.82 & 0.58 & 0.06 & 0.64 & 0.58 & 0.08 \\ 
-   \hline
-\end{tabular}
-  \label{tab:comp}
-\end{sidewaystable}
-
-\clearpage
-\subsection{Results from other datasets}
-
-In contrast to the muddled results from the analysis of \textit{E.
-marmorata}, the analysis of \textit{T.
-scripta scripta} to \textit{T.
-scripta elegans} indicates that these two subspecies are sufficiently morphologically distinct to be classified with near perfect accuracy (Table \ref{tab:second_res}, with both in-sample and out-of-sample classification having AUC values of approximately 1 for all methods.
-The analysis of the seven morphologically-distinct species also demonstrates near perfect classification of both the in-sample and out-of-sample datasets (Table \ref{tab:second_res}).
-
-For both datasets, the distribution of ROC scores from 1000 bootstrap replicates are tightly clustered near AUC = 1 (Fig.
-\ref{fig:other_boot}), in contrast to the results from the \textit{E.
-marmorata} case (Fig.
-\ref{fig:gen_hist}).
-These results demonstrate that when there are distinctions between the states of the classification schemes, the methods used here can recover them.
-% insert table here with the AUC values for in and out of sample results
-%   just like below for the 7 clear cut
-% trachemys
-% The analysis of the two subspecies of \uppercase{Trachemys}
-%   clearly identifiable
-%   in-sample is near perfect
-%   generalization is near perfect
-
-\begin{table}[ht]
-  \begin{subtable}[t]{\linewidth}
-    \begin{tabular}{ l r r r }
-      \hline
-      & \# predictors & In-sample AUC & Out-of-sample AUC \\ 
-      \hline
-      \hline
-      RF &   10 & 0.965 & 1.000 \\ 
-      LDA &    7 & 1.000 & 0.993 \\ 
-      MnL &    5 & 1.000 & 0.979 \\ 
-      \hline
-    \end{tabular}
-  \end{subtable}
-
-  \begin{subtable}[t]{\linewidth}
-    \begin{tabular}{ l r r r }
-      \hline
-      & \# predictors & In-sample AUC & Out-of-sample AUC \\ 
-      \hline
-      \hline
-      RF &   11 & 1.000 & 0.999 \\ 
-      LDA &   10 & 1.000 & 1.000 \\ 
-      MnL &    6 & 1.000 & 0.998 \\ 
-      \hline
-    \end{tabular}
-  \end{subtable}
-  \caption{Results from classification model estimates of the secondary, multi-species dataset.
-Models are random forest (RF), linear discriminate analysis (LDA), and multinomial logistic regression (MLR).
-Comparison of in-sample and out-of-sample AUC of the best performing model, along with the number of predictors.
-AUC values range between 0.5 (no better than random) and 1 (perfect classification).}
-  \label{tab:second_res}
-\end{table}
 
@@
   \centering
-  \begin{subfigure}[h]{0.8\linewidth}
-    \includegraphics[height = 0.8\textheight, width = \textwidth, keepaspectratio = true]{figure/trach_boot}
-    \caption{\textit{Tracheyms}}
-    \label{fig:trach_boot}
-  \end{subfigure}
-  \\
-  \begin{subfigure}[h]{0.8\linewidth}
-    \includegraphics[height = 0.8\textheight, width = \textwidth, keepaspectratio = true]{figure/seven_boot}
-    \caption{Seven}
-    \label{fig:seven_boot}
-  \end{subfigure}
-  \caption{Distribution of 1000 bootstrap replicates of the out-of-sample AUC values from the secondary, multi-species dataset.
-The AUC values are from the random forest (RF), linear discriminate analysis (LDA), and multinomial logistic regression (MLR).
-AUC values range between 0.5 (no better than random) and 1 (perfect classification).}
-  \label{fig:other_boot}
+  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/emys_oos_sel}
+  \caption{CAPTION}
+  \label{fig:emys_oos}
 \end{figure}
@@
 \section{Discussion}
-
-The results of this study show that there is no clear grouping of \textit{E. marmorata} based on plastron shape.
-
-The unsupervised learning results indicate that no grouping is optimal or particularly better than a single grouping, which is consistent with the results from the generalizations of the supervised learning models.
-The classification schemes used in the supervised learning models correspond, loosely, to unsupervised learning solutions with multiple groups.
-Because unsupervised learning solutions with multiple groups are poor descriptors of the observed variation, it is important to see this generally supported by the supervised learning results.
-
-The results from fitting the various supervised learning models to each of the classification schemes generally shows that no one scheme is ``best.'' Possible explanations include that the genetic differentiation is not associated with plastron shape variation and/or that local selective pressures (e.g.
-from hydrological regime) overwhelm morphological differentiation.
-This makes sense given that shell shape is influenced by selection for both protection and streamlining, but not necessary mate choice \citep{Rivera2008,Rivera2011,Stayton2011,Rivera2014,Pollyb} and that shell shape in \textit{E.
-marmorata} is known to vary among populations inhabiting water bodies with different flow regimes \citep{Holland1992,Lubcke2007,Germano2009}.
-Plastron shape does not seem to preserve a strong phylogenetic signal at the interspecific level in emydine turtles \citep{Angielczyk2011}, and our current results suggest that this may be the case for phylogeographic signal within emydine species as well.
-A final possibility (explored below) is that the proposed classification schemes themselves do not represent significant evolutionary lineages.
-
-Both the low out-of-sample AUC values (\(< 0.9\)) and the significant difference between the correctly and incorrectly classified observations support the conclusion that none of the hypothesized classification schemes are good descriptors of the observed plastral variation within \textit{E.
-marmorata}.
-
-Nevertheless, it is important to note that plastron shape is an extremely effective method for differentiating classes in the additional datasets we investigated.
-The magnitude of shape differences between the species (measured as Procrustes distance between species' mean shapes) is approximately an order of magnitude greater than the differences between the \textit{E.
-marmorata} subgroups, and the machine learning methods had no trouble accurately classifying the specimens correctly.
-However, the magnitude of the shape differences between the \textit{T.
-scripta} subspecies is comparable to theose separating the different \textit{E.
-marmorata} subgroups, yet even in this case the machine learning methods returned an almost perfect classification.
-These results demonstrate that plastron shape is normally a good marker for differentiating real subgroups in close relatives of \textit{E.
-marmorata}, and that our lack of results for \textit{E.
-marmorata} is not simply a shortcoming of the methods we applied.
-Indeed, they beg the question of what factors have suppressed morphological differentiation of plastron shape in \textit{E.
-marmorata} and \textit{E.
-pallida} if they are distinct species.
-Invoking issues such as the role of the plastron in protection or the need for streamlining are insufficient because the other species are expected to be subject to similar constraints \citep{Stayton2011,Pollyb}.
-While it may seem counterintuitive that plastron shape is both useful for species delimitation but has weak or absent phylogenetic signal, it is important to note that these are different goals.
-While phylogenetically similar species may not be morphologically similar, the variation within a species is much less than the variation between species which leads to plastron shape being a useful tool for species delimitation but not phylogenetic reconstruction.
-
-\subsection{Is there more than one species of Western Pond Turtle?}
-
-The lack of morphological support for the distinctiveness of \textit{E.
-pallida} does not, on its own, preclude the recognition of this taxon.
-However, this apparent lack of congruence does prompt a reexamination of the methods and concepts that led to that taxonomic revision, especially considering that plastron shape is demonstrably capable of differentiating species and subspecies among emydids.
-In other words, before we can assess the significance of the morphological non-diagnosablity, it is essential to evaluate the methods and concepts that led to the initial taxonomic revision.
-
-\citet{Spinks2014} elevated \textit{E.
-pallida} based on a species delimitation analysis of SNP data using BPP \citep{Yang2010b}.
-However, \citet{Spinks2014} did not heed the caveats about species delimitation methods raised by \citet{Carstens2013}.
-In addition to specifically addressing the shortcomings of validation methods such as BPP that rely on guide trees and ``should be interpreted with caution,'' \citet{Carstens2013} also strongly emphasized that ``Inferences regarding species boundaries based on genetic data alone are likely inadequate, and species delimitation should be conducted with consideration of the life history, geographical distribution, morphology and behaviour (where applicable) of the focal system\dots'' These caveats evoke the development of the Unified Species Concept \citep{Dayrat2005a,DeQueiroz2007b}, Integrative Taxonomy \citep{Padial2010}, and other pluralist approaches to species delimitation.
-None of these considerations were brought to bear on the \textit{E.
-marmorata} system until now, and in doing so we find the proposal that \textit{E.
-pallida} is a distinct species to be lacking in a normally robust morphological marker.
-
-The natural history and geographical distribution of \textit{E.
-marmorata} and \textit{E.
-pallida} also make the recognition of these taxa implausible.
-The data from \citet{Spinks2014} show extensive introgression and admixture in Central California, which is expected because there are no significant barriers to gene flow in this region.
-Combined with the well-demonstrated ability for testudinoid turtles, including emydids and even \textit{Emys}, to hybridize (e.g.
-\citealt{Buskirk2005,Spinks2009,Parham2013}) it is hard to imagine how \textit{E.
-marmorata} and \textit{E.
-pallida} could maintain their integrity in the face of such admixture.
-Because the geography, natural history, demonstrated genetic admixture of \textit{E.
-marmorata}, and comparisons with other morphologically diagnosable species and subspecies conflict with the recognition of \textit{E.
-pallida}, we hypothesize that our inability to classify the morphological data by proposed species is because \textit{E.
-pallida} is not a distinct species.
-
-We fully agree with \citet{Spinks2014} that \textit{E.
-marmorata} (\textit{sensu lato}) is a species deserving of strong conservation efforts, and we do not wish to trivialize this need.
-Moreover, the genetic diversity uncovered by the analysis of \citet{Spinks2014} should be explicitly accounted for in any conservation plan.
-Given the apparent lack of morphological distinction, however, we consider that this diversity should be considered Evolutionary Significant Units or Distinct Population Segments instead of distinct species.
-
-Finally, it is important to note that the data and analyses we present do not let us definitively say whether the apparent lack of morphological divergence within \textit{E.
-marmorata} truly reflects the presence of a single species, or if it is an artifact of plastron shape being a poor morphological marker for phylogenetic and phylogeographic divergences in the case of \textit{E.
-marmorata}, because we could not carry out our morphometric analyses on the specimens from which the genetic data were obtained.
-The comparisons with the other emydid taxa suggest that it is because \textit{E.
-marmorata} is a single species.
-However, tests of both our preferred conclusion (\textit{E.
-marmorata} as a single species) and that \citet{Spinks2014} should include morphological and molecular analyses of the same set of voucher specimens, as well as additional tests of species delimitation using alternative methods and corroborating evidence as suggested by \citet{Carstens2013}.
-From a morphological standpoint, support for the validity of ``\textit{E.
-pallida}'' may come from other aspects of morphology, such as carapace shape or other features.
-Likewise, further investigation of the phylogeographic utility of plastron shape in other turtle species will help to clarify whether the lack of differentiation seen in \textit{E.
-marmoarata}, and the strong differentiation among the other emydids, is typical or an unusual case.
 
@@
 
-\clearpage
-\begin{thebibliography}{77}
-\providecommand{\natexlab}[1]{#1}
-\providecommand{\selectlanguage}[1]{\relax}
-\providecommand{\bibAnnoteFile}[1]{%
-  \IfFileExists{#1}{\begin{quotation}\noindent\textsc{Key:} #1\\
-    \textsc{Annotation:}\ \input{#1}\end{quotation}}{}}
-  \providecommand{\bibAnnote}[2]{%
-    \begin{quotation}\noindent\textsc{Key:} #1\\
-      \textsc{Annotation:}\ #2\end{quotation}}
-
-    \bibitem[{Angielczyk and Feldman(2013)}]{Angielczyk2013a}
-    Angielczyk, K.~D. and C.~R. Feldman. 2013. {Are diminutive turtles
-    miniaturized? The ontogeny of plastron shape in emydine turtles}. Biological
-    Journal of the Linnean Society 108:727--755.
-    \bibAnnoteFile{Angielczyk2013a}
-
-    \bibitem[{Angielczyk et~al.(2011)Angielczyk, Feldman, and
-    Miller}]{Angielczyk2011}
-    Angielczyk, K.~D., C.~R. Feldman, and G.~R. Miller. 2011. {Adaptive evolution
-    of plastron shape in emydine turtles.} Evolution 65:377--394.
-    \bibAnnoteFile{Angielczyk2011}
-
-    \bibitem[{Angielczyk and Sheets(2007)}]{Angielczyk2007}
-    Angielczyk, K.~D. and H.~D. Sheets. 2007. {Investigation of simulated tectonic
-    deformation in fossils using geometric morphometrics}. Paleobiology
-    33:125--148.
-    \bibAnnoteFile{Angielczyk2007}
-
-    \bibitem[{Baird and Girard(1852)}]{Baird1852}
-    Baird, S.~F. and C.~Girard. 1852. {Descriptions of new species of reptiles
-      collected by the U.S. Exploring Expedition under the command of Capt. Charles
-    Wilkes}. Proceedings of the National Academy of Sciences Philadelphia
-    6:174--177.
-    \bibAnnoteFile{Baird1852}
-
-    \bibitem[{Bauer et~al.(2000)Bauer, Parham, Brown, Stuart, Grismer, Papenfuss,
-      Bohme, Savage, Carranza, Grismer, Wagner, Schmitz, Ananjeva, and
-    Inger}]{Bauer2000}
-    Bauer, A.~M., J.~F. Parham, R.~M. Brown, B.~L. Stuart, L.~Grismer, T.~J.
-    Papenfuss, W.~Bohme, J.~M. Savage, S.~Carranza, J.~L. Grismer, P.~Wagner,
-    A.~Schmitz, N.~B. Ananjeva, and R.~F. Inger. 2000. {Availability of new
-      Baysian-delimited gecko names and the importance of character-based species
-    descriptions}. Proceedings of the Royal Society B: Biological Sciences
-    278:490--492.
-    \bibAnnoteFile{Bauer2000}
-
-    \bibitem[{Baylac et~al.(2003)Baylac, Villemant, and Simbolotti}]{Baylac2003}
-    Baylac, M., C.~Villemant, and G.~Simbolotti. 2003. {Combining geometric
-      morphometrics with pattern recognition for the investigation of species
-    complexes}. Biological Journal of the Linnean Society 80:89--98.
-    \bibAnnoteFile{Baylac2003}
-
-    \bibitem[{Bickford et~al.(2007)Bickford, Lohman, Sodhi, Ng, Meier, Winker,
-    Ingram, and Das}]{Bickford2007}
-    Bickford, D., D.~J. Lohman, N.~S. Sodhi, P.~K.~L. Ng, R.~Meier, K.~Winker,
-    K.~K. Ingram, and I.~Das. 2007. {Cryptic species as a window on diversity and
-    conservation.} Trends in ecology \& evolution 22:148--55.
-    \bibAnnoteFile{Bickford2007}
-
-    \bibitem[{Breiman(2001)}]{Breiman2001}
-    Breiman, L. 2001. {Random Forests}. Machine Learning 45:5--32.
-    \bibAnnoteFile{Breiman2001}
-
-    \bibitem[{Breiman et~al.(1984)Breiman, Friedman, Stone, and
-    Olshen}]{Breiman1984}
-    Breiman, L., J.~Friedman, C.~J. Stone, and R.~A. Olshen. 1984. {Classification
-    and regression trees}. Wadsworth International Group, Belmont.
-    \bibAnnoteFile{Breiman1984}
-
-    \bibitem[{Bury et~al.(2010)Bury, Germano, and Bury}]{Bury2010}
-    Bury, R.~B., D.~J. Germano, and G.~W. Bury. 2010. {Population Structure and
-      Growth of the Turtle Actinemys marmorata from the Klamath–Siskiyou
-    Ecoregion: Age, Not Size, Matters}. Copeia 2010:443--451.
-    \bibAnnoteFile{Bury2010}
-
-    \bibitem[{Buskirk et~al.(2005)Buskirk, Parham, and Feldman}]{Buskirk2005}
-    Buskirk, S.~W., J.~F. Parham, and C.~R. Feldman. 2005. {On the hybridisation
-    between two distantly related Asian turtles (Testudines: Scalia x Mauremys)}.
-    Salamandra 41:21--26.
-    \bibAnnoteFile{Buskirk2005}
-
-    \bibitem[{Cardini et~al.(2009)Cardini, Nagorsen, O'Higgins, Polly, {Thorington
-    Jr}, and Tongiorgi}]{Cardini2009a}
-    Cardini, A., D.~Nagorsen, P.~O'Higgins, P.~D. Polly, R.~W. {Thorington Jr}, and
-    P.~Tongiorgi. 2009. {Detecting biological distinctiveness using geometric
-    morphometrics: an example case from the Vancouver Island marmot}. Ethology
-    Ecology \& Evolution 21:209--223.
-    \bibAnnoteFile{Cardini2009a}
-
-    \bibitem[{Carstens and Dewey(2010)}]{Carstens2010}
-    Carstens, B.~C. and T.~A. Dewey. 2010. {Species Delimitation Using a Combined
-      Coalescent and Information-Theoretic Approach: An Example from North American
-    Myotis Bats}. Systematic Biology 59:400--414.
-    \bibAnnoteFile{Carstens2010}
-
-    \bibitem[{Carstens et~al.(2013)Carstens, Pelletier, Reid, and
-    Satler}]{Carstens2013}
-    Carstens, B.~C., T.~a. Pelletier, N.~M. Reid, and J.~D. Satler. 2013. {How to
-    fail at species delimitation.} Molecular ecology 22:4369--83.
-    \bibAnnoteFile{Carstens2013}
-
-    \bibitem[{Caumul and Polly(2005)}]{Caumul2005a}
-    Caumul, R. and P.~D. Polly. 2005. {Phylogenetic and environmental components of
-      morphological variation: skull, mandible, and molar shape in marmots
-    (Marmota, Rodentia).} Evolution; international journal of organic evolution
-    59:2460--72.
-    \bibAnnoteFile{Caumul2005a}
-
-    \bibitem[{Clare(2011)}]{Clare2011}
-    Clare, E.~L. 2011. {Cryptic species? Patterns of maternal and paternal gene
-    flow in eight neotropical bats.} PloS one 6:e21460.
-    \bibAnnoteFile{Clare2011}
-
-    \bibitem[{Dayrat(2005)}]{Dayrat2005a}
-    Dayrat, B. 2005. {Towards integrative taxonomy}. Biological Journal of the
-    Linnean Society 85:407--415.
-    \bibAnnoteFile{Dayrat2005a}
-
-    \bibitem[{{De Queiroz}(2007)}]{DeQueiroz2007b}
-    {De Queiroz}, K. 2007. {Species concepts and species delimitation.} Systematic
-    Biology 56:879--86.
-    \bibAnnoteFile{DeQueiroz2007b}
-
-    \bibitem[{Demandt and Bergek(2009)}]{Demandt2009}
-    Demandt, M.~H. and S.~Bergek. 2009. {Identification of cyprinid hybrids by
-    using geometric morphometrics and microsatellites}. Journal of Applied
-    Ichthyology 25:695--701.
-    \bibAnnoteFile{Demandt2009}
-
-    \bibitem[{Dobigny et~al.(2003)Dobigny, Granjon, Aniskin, Ba, and
-    Voloboulev}]{Dobigny2003}
-    Dobigny, G., L.~Granjon, V.~Aniskin, K.~Ba, and V.~Voloboulev. 2003. {A new
-    sigling species of Taterillus (Muridae, Gerbillinae) from West Agrica}.
-    Mammalian Biology 68:299--316.
-    \bibAnnoteFile{Dobigny2003}
-
-    \bibitem[{Dryden(2013)}]{Dryden2013}
-    Dryden, I.~L. 2013. shapes: Statistical shape analysis. R package version
-    1.1-8.
-    \bibAnnoteFile{Dryden2013}
-
-    \bibitem[{Dryden and Mardia(1998)}]{Dryden1998a}
-    Dryden, I.~L. and K.~Y. Mardia. 1998. {Statistical shape analysis}. Wiley, New
-    York.
-    \bibAnnoteFile{Dryden1998a}
-
-    \bibitem[{Eldredge and Gould(1972)}]{Eldredge1972}
-    Eldredge, N. and S.~J. Gould. 1972. {Punctuated equilibria: an alternative to
-    phyletic gradualism}. Pages~82--115 \emph{in} Models in Paleobiology
-    (T.~J.~M. Schopf, ed.). Freeman Cooper, San Francisco.
-    \bibAnnoteFile{Eldredge1972}
-
-    \bibitem[{Feldman and Parham(2002)}]{Feldman2002}
-    Feldman, C.~R. and J.~F. Parham. 2002. {Molecular phylogenetics of emydine
-    turtles: taxonomic revision and the evolution of shell kinesis.} Molecular
-    Phylogenetics and Evolution 22:388--98.
-    \bibAnnoteFile{Feldman2002}
-
-    \bibitem[{Francoy et~al.(2009)Francoy, Silva, Nunes-Silva, Menezes, and
-    Imperatriz-Fonseca}]{Francoy2009}
-    Francoy, T.~M., R.~A.~O. Silva, P.~Nunes-Silva, C.~Menezes, and V.~L.
-    Imperatriz-Fonseca. 2009. {Gender identification of five genera of stingless
-    bees (Apidae, Meliponini) based on wing morphology}. Genetics and molecular
-    research 8:207--214.
-    \bibAnnoteFile{Francoy2009}
-
-    \bibitem[{Funk et~al.(2012)Funk, Caminer, and Ron}]{Funk2012}
-    Funk, W.~C., M.~Caminer, and S.~R. Ron. 2012. {High levels of cryptic species
-    diversity uncovered in Amazonian frogs.} Proceedings of the Royal Society B:
-    Biological Sciences 279:1806--14.
-    \bibAnnoteFile{Funk2012}
-
-    \bibitem[{Gaubert et~al.(2005)Gaubert, Taylor, Fernandes, Bruford, and
-    Veron}]{Gaubert2005b}
-    Gaubert, P., P.~J. Taylor, C.~a. Fernandes, M.~W. Bruford, and G.~Veron. 2005.
-    {Patterns of cryptic hybridization revealed using an integrative approach: a
-      case study on genets (Carnivora, Viverridae, Genetta spp.) from the southern
-    African subregion}. Biological Journal of the Linnean Society 86:11--33.
-    \bibAnnoteFile{Gaubert2005b}
-
-    \bibitem[{Germano and Bury(2009)}]{Germano2009}
-    Germano, D.~J. and R.~B. Bury. 2009. {Variation in body size, growth, and
-      population structure of Actinemys marmorata from lentic and lotic habitats in
-    Southern Oregon}. Journal of Herpetology 43:510--520.
-    \bibAnnoteFile{Germano2009}
-
-    \bibitem[{Germano and Rathbun(2008)}]{Germano2008}
-    Germano, D.~J. and G.~B. Rathbun. 2008. {Growth, population structure, and
-      reproduction of western pond turtles (Actinemys marmorata) on the Central
-    Coast of California}. Chelonian Conservation and Biology 7:188--194.
-    \bibAnnoteFile{Germano2008}
-
-    \bibitem[{Gould and Eldredge(1977)}]{Gould1977a}
-    Gould, S.~J. and N.~Eldredge. 1977. {Punctuated equilibria: the tempo and mode
-    of evolution reconsidered}. Paleobiology 3:115--151.
-    \bibAnnoteFile{Gould1977a}
-
-    \bibitem[{G\"{u}nd\"{u}z et~al.(2007)G\"{u}nd\"{u}z, Jaarola, Tez, Yeniyurt,
-    Polly, and Searle}]{Gunduz2007}
-    G\"{u}nd\"{u}z, I., M.~Jaarola, C.~Tez, C.~Yeniyurt, P.~D. Polly, and J.~B.
-    Searle. 2007. {Multigenic and morphometric differentiation of ground
-      squirrels (Spermophilus, Scuiridae, Rodentia) in Turkey, with a description
-    of a new species.} Molecular phylogenetics and evolution 43:916--35.
-    \bibAnnoteFile{Gunduz2007}
-
-    \bibitem[{Hand and Till(2001)}]{Hand2001}
-    Hand, D.~J. and R.~J. Till. 2001. {A Simple Generalisation of the Area Under
-    the ROC Curve for Multiple Class Classification Problems}. Machine Learning
-    45:171--186.
-    \bibAnnoteFile{Hand2001}
-
-    \bibitem[{Hastie et~al.(2009)Hastie, Tibshirani, and Friedman}]{Hastie2009}
-    Hastie, T., R.~Tibshirani, and J.~Friedman. 2009. {The elements of statistical
-    learning: data mining, inference, and prediction}. 2nd ed. Springer, New
-    York.
-    \bibAnnoteFile{Hastie2009}
-
-    \bibitem[{Hausdorf and Hennig(2010)}]{Hausdorf2010}
-    Hausdorf, B. and C.~Hennig. 2010. {Species delimitation using dominant and
-    codominant multilocus markers.} Systematic biology 59:491--503.
-    \bibAnnoteFile{Hausdorf2010}
-
-    \bibitem[{Holland(1992)}]{Holland1992}
-    Holland, D.~C. 1992. {Level and pattern in morphological variation: a
-    phylogeographic study of the western pond turtle (Clemmys marmorata)}. Ph.D.
-    thesis University of Southwestern Louisiana.
-    \bibAnnoteFile{Holland1992}
-
-    \bibitem[{Huelsenbeck et~al.(2011)Huelsenbeck, Andolfatto, and
-    Huelsenbeck}]{Huelsenbeck2011b}
-    Huelsenbeck, J.~P., P.~Andolfatto, and E.~T. Huelsenbeck. 2011. {Structurama:
-    bayesian inference of population structure.} Evolutionary bioinformatics
-    online 7:55--9.
-    \bibAnnoteFile{Huelsenbeck2011b}
-
-    \bibitem[{Hunt(2008)}]{Hunt2008}
-    Hunt, G. 2008. {Gradual or pulsed evolution: when should punctuational
-    explanations be preferred?} Paleobiology 34:360--377.
-    \bibAnnoteFile{Hunt2008}
-
-    \bibitem[{Kaufman and Rousseeuw(1990)}]{Kaufman1990}
-    Kaufman, L. and P.~J. Rousseeuw. 1990. {Finding groups in data : an
-    introduction to cluster analysis}. Wiley, New York.
-    \bibAnnoteFile{Kaufman1990}
-
-    \bibitem[{Kendall(1977)}]{Kendall1977a}
-    Kendall, D.~G. 1977. {The diffusion of shape}. Advances in Applied Probability
-    9:428--430.
-    \bibAnnoteFile{Kendall1977a}
-
-    \bibitem[{Kendall(1984)}]{Kendall1984a}
-    Kendall, D.~G. 1984. {Shape Manifolds, Procrustean Metrics, and Complex
-    Projective Spaces}. Bulletin of the London Mathematical Society 16:81--121.
-    \bibAnnoteFile{Kendall1984a}
-
-    \bibitem[{Klingenberg et~al.(2002)Klingenberg, Barluenga, and
-    Meyer}]{Klingenberg2002}
-    Klingenberg, C.~P., M.~Barluenga, and A.~Meyer. 2002. {Shape analysis of
-    symetric structures: quantifying variation among individuals and asymmetry}.
-    Evolution 56:1909--1920.
-    \bibAnnoteFile{Klingenberg2002}
-
-    \bibitem[{Kuhn(2013)}]{KuhnMAN2013}
-    Kuhn, M. 2013. caret: Classification and Regression Training. R package version
-    5.15-61.
-    \bibAnnoteFile{KuhnMAN2013}
-
-    \bibitem[{Kuhn and Johnson(2013)}]{Kuhn2013}
-    Kuhn, M. and K.~Johnson. 2013. {Applied predictive modeling}. Springer, New
-    York, NY.
-    \bibAnnoteFile{Kuhn2013}
-
-    \bibitem[{Leach\'{e} and Fujita(2010)}]{Leache2010}
-    Leach\'{e}, A.~D. and M.~K. Fujita. 2010. {Bayesian species delimitation in
-    West African forest geckos (Hemidactylus fasciatus).} Proceedings. Biological
-    sciences / The Royal Society 277:3071--7.
-    \bibAnnoteFile{Leache2010}
-
-    \bibitem[{Lubcke and Wilson(2007)}]{Lubcke2007}
-    Lubcke, G.~M. and D.~S. Wilson. 2007. {Variation in shell morphology of the
-      Western Pond Turtle (Actinemys marmorata Baird and Giarard) from three
-    aquativ habitats in Northern California}. Journal of Herpetology 41:107--114.
-    \bibAnnoteFile{Lubcke2007}
-
-    \bibitem[{MacLeod(2007)}]{MacLeod2007}
-    MacLeod, N. 2007. {Automated taxon identification in systematics: theory,
-    approaches and applications}. CRC Press, Boca Raton.
-    \bibAnnoteFile{MacLeod2007}
-
-    \bibitem[{Maechler et~al.(2013)Maechler, Rousseeuw, Struyf, Hubert, and
-    Hornik}]{Maechler2013}
-    Maechler, M., P.~Rousseeuw, A.~Struyf, M.~Hubert, and K.~Hornik. 2013. cluster:
-    Cluster Analysis Basics and Extensions. R package version 1.14.4.
-    \bibAnnoteFile{Maechler2013}
-
-    \bibitem[{Mitrovski-Bogdanovic et~al.(2013)Mitrovski-Bogdanovic, Petrovic,
-      Mitrovic, Ivanovic, \v{Z}ikic, Star\'{y}, Vorburger, and
-    Tomanovic}]{MitrovskiBogdanovic2013}
-    Mitrovski-Bogdanovic, A., A.~Petrovic, M.~Mitrovic, A.~Ivanovic, V.~\v{Z}ikic,
-    P.~Star\'{y}, C.~Vorburger, and v.~Tomanovic. 2013. {Identification of two
-      cryptic species within the Praon abjectum group (Hymenoptera: Braconidae:
-    Aphidiinae) using molecular markers and geometric morphometrics}. Annals of
-    the entomological society of America 106:170--180.
-    \bibAnnoteFile{MitrovskiBogdanovic2013}
-
-    \bibitem[{Mitteroecker and Bookstein(2011)}]{Mitteroecker2011}
-    Mitteroecker, P. and F.~Bookstein. 2011. {Linear Discrimination, Ordination,
-    and the Visualization of Selection Gradients in Modern Morphometrics}.
-    Evolutionary Biology 38:100--114.
-    \bibAnnoteFile{Mitteroecker2011}
-
-    \bibitem[{O'Meara(2010)}]{O'Meara2010}
-    O'Meara, B.~C. 2010. {New heuristic methods for joint species delimitation and
-    species tree inference.} Systematic biology 59:59--73.
-    \bibAnnoteFile{O'Meara2010}
-
-    \bibitem[{Padial et~al.(2010)Padial, Miralles, {De la Riva}, and
-    Vences}]{Padial2010}
-    Padial, J.~M., A.~Miralles, I.~{De la Riva}, and M.~Vences. 2010. {The
-    integrative future of taxonomy}. Frontiers in Zoology 7:1--14.
-    \bibAnnoteFile{Padial2010}
-
-    \bibitem[{Parham et~al.(2013)Parham, Papenfuss, Dijk, Wilson, Marte, Schettino,
-    and {Brian Simison}}]{Parham2013}
-    Parham, J.~F., T.~J. Papenfuss, P.~P.~V. Dijk, B.~S. Wilson, C.~Marte, L.~R.
-    Schettino, and W.~{Brian Simison}. 2013. {Genetic introgression and
-      hybridization in Antillean freshwater turtles (Trachemys) revealed by
-    coalescent analyses of mitochondrial and cloned nuclear markers.} Molecular
-    phylogenetics and evolution 67:176--87.
-    \bibAnnoteFile{Parham2013}
-
-    \bibitem[{Pfenninger and Schwenk(2007)}]{Pfenninger2007}
-    Pfenninger, M. and K.~Schwenk. 2007. {Cryptic animal species are homogeneously
-    distributed among taxa and biogeographical regions.} BMC evolutionary biology
-    7:121.
-    \bibAnnoteFile{Pfenninger2007}
-
-    \bibitem[{Polly(2003)}]{Polly2003}
-    Polly, P.~D. 2003. {Paleophylogeography of Sorex araneus: molar shape as a
-    morphological marker for fossil shrews}. Mammalia 68:233--243.
-    \bibAnnoteFile{Polly2003}
-
-    \bibitem[{Polly(2007)}]{Polly2007a}
-    Polly, P.~D. 2007. {Phylogeographic differentiation in Sorex araneus:
-    morphology in relation to geography and karyotype}. Russian Journal of
-    Theriology 6:73--84.
-    \bibAnnoteFile{Polly2007a}
-
-    \bibitem[{Polly et~al.()Polly, Stayton, Dumont, Pierce, Rayfield, and
-    Angielczyk}]{Pollyb}
-    Polly, P.~D., C.~T. Stayton, E.~R. Dumont, S.~E. Pierce, E.~J. Rayfield, and
-    K.~D. Angielczyk. ???? {Combining geometric morphometrics and finite element
-    analysis with evolutionary modeling: towards a synthesis}. Journal of
-    Vertebrate Paleontology .
-    \bibAnnoteFile{Pollyb}
-
-    \bibitem[{Pons et~al.(2006)Pons, Barraclough, Gomez-Zurita, Cardoso, Duran,
-    Hazell, Kamoun, Sumlin, and Vogler}]{Pons2006}
-    Pons, J., T.~Barraclough, J.~Gomez-Zurita, A.~Cardoso, D.~Duran, S.~Hazell,
-    S.~Kamoun, W.~Sumlin, and A.~Vogler. 2006. {Sequence-Based Species
-    Delimitation for the DNA Taxonomy of Undescribed Insects}. Systematic Biology
-    55:595--609.
-    \bibAnnoteFile{Pons2006}
-
-    \bibitem[{{R Core Team}(2015)}]{2015}
-    {R Core Team}. 2015. R: A Language and Environment for Statistical Computing. R
-    Foundation for Statistical Computing Vienna, Austria.
-    \bibAnnoteFile{2015}
-
-    \bibitem[{Rivera(2008)}]{Rivera2008}
-    Rivera, G. 2008. {Ecomorphological variation in shell shape of the freshwater
-    turtle Pseudemys concinna inhabiting different aquatic flow regimes.}
-    Integrative and comparative biology 48:769--87.
-    \bibAnnoteFile{Rivera2008}
-
-    \bibitem[{Rivera et~al.(2014)Rivera, Davis, Godwin, and Adams}]{Rivera2014}
-    Rivera, G., J.~N. Davis, J.~C. Godwin, and D.~C. Adams. 2014. {Repeatability of
-    Habitat-Associated Divergence in Shell Shape of Turtles}. Evolutionary
-    Biology Pages~29--37.
-    \bibAnnoteFile{Rivera2014}
-
-    \bibitem[{Rivera and Stayton(2011)}]{Rivera2011}
-    Rivera, G. and C.~T. Stayton. 2011. {Finite element modeling of shell shape in
-      the freshwater turtle Pseudemys concinna reveals a trade-off between
-    mechanical strength and hydrodynamic efficiency.} Journal of morphology
-    272:1192--203.
-    \bibAnnoteFile{Rivera2011}
-
-    \bibitem[{Rohlf(2005)}]{Rohlf2005}
-    Rohlf, F.~J. 2005. {TpsDig 2.04}.
-    \bibAnnoteFile{Rohlf2005}
-
-    \bibitem[{Schilck-Steiner et~al.(2007)Schilck-Steiner, Seifert, Stauffer,
-    Christian, Crozier, and Steiner}]{SchlickSteiner2007}
-    Schilck-Steiner, B.~C., B.~Seifert, C.~Stauffer, E.~Christian, R.~H. Crozier,
-    and F.~M. Steiner. 2007. {Without morphology, cryptic species stay in
-    taxonomic crypsis following discovery}. Trends in ecology \& evolution
-    22:391--392.
-    \bibAnnoteFile{SchlickSteiner2007}
-
-    \bibitem[{Seeliger(1945)}]{Seeliger1945}
-    Seeliger, L.~M. 1945. {Variation in the Pacific Mud Turtle}. Copeia
-    1945:150--159.
-    \bibAnnoteFile{Seeliger1945}
-
-    \bibitem[{Spinks and Shaffer(2005)}]{Spinks2005}
-    Spinks, P.~Q. and H.~B. Shaffer. 2005. {Range-wide molecular analysis of the
-      western pond turtle (Emys marmorata): cryptic variation, isolation by
-    distance, and their conservation implications.} Molecular ecology
-    14:2047--64.
-    \bibAnnoteFile{Spinks2005}
-
-    \bibitem[{Spinks and Shaffer(2009)}]{Spinks2009}
-    Spinks, P.~Q. and H.~B. Shaffer. 2009. {Conflicting mitochondrial and nuclear
-      phylogenies for the widely disjunct Emys (Testudines: Emydidae) species
-    complex, and what they tell us about biogeography and hybridization.}
-    Systematic biology 58:1--20.
-    \bibAnnoteFile{Spinks2009}
-
-    \bibitem[{Spinks et~al.(2014)Spinks, Thomson, and {Bradley
-    Shaffer}}]{Spinks2014}
-    Spinks, P.~Q., R.~C. Thomson, and H.~{Bradley Shaffer}. 2014. {The advantages
-      of going large: genome wide SNPs clarify the complex population history and
-    systematics of the threatened western pond turtle}. Molecular Ecology
-    Pages~n/a--n/a.
-    \bibAnnoteFile{Spinks2014}
-
-    \bibitem[{Spinks et~al.(2010)Spinks, Thomson, and Shaffer}]{Spinks2010}
-    Spinks, P.~Q., R.~C. Thomson, and H.~B. Shaffer. 2010. {Nuclear gene
-      phylogeography reveals the historical legacy of an ancient inland sea on
-    lineages of the western pond turtle, Emys marmorata in California.} Molecular
-    ecology 19:542--56.
-    \bibAnnoteFile{Spinks2010}
-
-    \bibitem[{Stayton(2011)}]{Stayton2011}
-    Stayton, C.~T. 2011. {Biomechanics on the half shell: functional performance
-      influences patterns of morphological variation in the emydid turtle
-    carapace.} Zoology (Jena, Germany) 114:213--23.
-    \bibAnnoteFile{Stayton2011}
-
-    \bibitem[{Stuart et~al.(2006)Stuart, Inger, and Voris}]{Stuart2006}
-    Stuart, B.~L., R.~F. Inger, and H.~K. Voris. 2006. {High level of cryptic
-      species diversity revealed by sympatric lineages of Southeast Asian forest
-    frogs.} Biology letters 2:470--4.
-    \bibAnnoteFile{Stuart2006}
-
-    \bibitem[{Sztencel-Jabłonka et~al.(2009)Sztencel-Jabłonka, Jones, and
-    BogdanowicZ}]{Sztencel-Jabonka2009}
-    Sztencel-Jabłonka, A., G.~Jones, and W.~BogdanowicZ. 2009. {Skull Morphology
-      of Two Cryptic Bat Species: Pipistrellus pipistrellus and P. pygmaeus — A
-    3D Geometric Morphometrics Approach with Landmark Reconstruction}. Acta
-    Chiropterologica 11:113--126.
-    \bibAnnoteFile{Sztencel-Jabonka2009}
-
-    \bibitem[{Tibshirani et~al.(2001)Tibshirani, Walther, and
-    Hastie}]{Tibshirani2001a}
-    Tibshirani, R., G.~Walther, and T.~Hastie. 2001. {Estimating the number of
-    clusters in a data set via the gap statistic}. Journal of the Royal
-    Statistical Society: Series B (Statistical Methodology) 63:411--423.
-    \bibAnnoteFile{Tibshirani2001a}
-
-    \bibitem[{{Van Bocxlaer} and Hunt(2013)}]{VanBocxlaer2013}
-    {Van Bocxlaer}, B. and G.~Hunt. 2013. {Morphological stasis in an ongoing
-    gastropod radiation from Lake Malawi}. Proceedings of the National Academy of
-    Sciences 110:13892--13897.
-    \bibAnnoteFile{VanBocxlaer2013}
-
-    \bibitem[{Uyeda et~al.(2008)Uyeda, Caetano, and Pennell}]{Uyeda2015}
-    Uyeda, Josef C., D.~S.~Caetano and M.~W.~Pennell, {Comparative Analysis of 
-    Principal Components Can be Misleading}. Systematic Biology 64:677--689.
-    \bibAnnoteFile{Uyeda2015}
-
-    \bibitem[{van~den Brink and Bokma(2011)}]{VandenBrink2011}
-    van~den Brink, V. and F.~Bokma. 2011. {Morphometric shape analysis using
-      learning vector quantization neural networks — an example distinguishing
-    two microtine vole species}. Annales Zoologici Fennici 48:359--364.
-    \bibAnnoteFile{VandenBrink2011}
-
-    \bibitem[{Venables and Ripley(2002)}]{Venables2002a}
-    Venables, W. and B.~D. Ripley. 2002. {Modern applied statistics with S}. 4th
-    ed. Springer, New York.
-    \bibAnnoteFile{Venables2002a}
-
-    \bibitem[{Liaw and Wiener(2002)}]{Liaw2002}
-    Law, A. and M. Wiener. 2002. {Classification and Regression by randomForest.} 
-    R News 2(3):18--22.
-    \bibAnnoteFile{Liaw2002}
-
-    \bibitem[{Yang and Rannala(2010)}]{Yang2010b}
-    Yang, Z. and B.~Rannala. 2010. {Bayesian species delimitation using multilocus
-    sequence data.} Proceedings of the National Academy of Sciences 107:9264--9.
-    \bibAnnoteFile{Yang2010b}
-
-    \bibitem[{Zelditch et~al.(2004)Zelditch, Swiderski, and Sheets}]{Zelditch2004}
-    Zelditch, M.~L., D.~L. Swiderski, and H.~D. Sheets. 2004. {Geometric
-    morphometrics for biologists: a primer}. Elsevier Academic Press, Amsterdam.
-    \bibAnnoteFile{Zelditch2004}
-
-    \bibitem[{Claude(2008)}]{Claude2008}
-    Claude, J. 2004. {Morphometrics with R}. Springer-Verlang, New York.
-    \bibAnnoteFile{Claude2008}
-
-\end{thebibliography}
+\bibliographystyle{abbrvnat}
+\bibliography{turtle,packages}
 

### 2016-05-06 doc/turtles_statistical_learning.tex
fixed some of the labels for the figures; need to make sure schemes are correct because confusion

@@
 
-\title{How cryptic is cryptic diversity? Machine learning approaches to classifying morphological variation in the Pacific Pond Turtle (\textit{Emys marmorata})}
-%\title{Supervised learning approaches to quantifying cryptic diversity: picking amongst taxonomic hypotheses of Pacific Pond Turtle (\textit{Emys marmorata})}
+%\title{How cryptic is cryptic diversity? Machine learning approaches to classifying morphological variation in the Pacific Pond Turtle (\textit{Emys marmorata})}
+\title{Supervised learning approaches to classifying morphological viariation: picking amongst taxonomic hypotheses for the Pacific Pond Turtle (\textit{Emys marmorata})}
 \author[1]{Peter D Smits}%\thanks{psmits@uchicago.edu}}
@@
 Both ROC and AUC are preferrable to simple classificaiton accuracy when class membership is unbalanced, as it is in these analyses \citep{Hastie2009}.
-The standard ROC and AUC calculations are defined only for binary classifications.
-To generalize this approach for mulitple response classes, we used an all-against-one strategy where the model AUC is the average of the AUC values from the multiple binary comparisons of one class compared to all others \citep{Hand2001}.
+The standard ROC and AUC calculations are defined only for binary classifications, which is not the case for our seven species and \textit{Emys} complex datasets.
+To generalize this approach for situations with mulitple response classes, we used an all-against-one strategy where the model AUC is the average of the AUC values from the multiple binary comparisons of one class compared to all others \citep{Hand2001}.
 
@@
 %   all assume that predictors have additive effect (e.g. independent)
-The supervised learning methods used here were multinomial logistic regression (MLR), linear discrminiate analysis (LDA), penalized discriminate analysis (PDA), single-hidden-layer neutral network (NN), and random forests (RF).
-Each of these methods makes different assumptions, treat data differently, and can produce very different classification results \citep{Hastie2009}.
-The common assumption of all of these methods is that the predictors or features are independent and/or have additive effects on prediction \citep{Hastie2009}.
+Instead of relying on a single supervised learning method, we chose to use an ensemble of mulitple approaches so that the congruence bewteen the them could be used as a means of ``support'' for one conclusion or another.
+The supervised learning methods used here are described in Table \ref{tab:methods}.
+Each of these methods makes different assumptions, treat data differently, and can produce different qualities of classification results depending on the nature of the data \citep{Hastie2009}.
+
+\begin{table}
+  \centering
+  \begin{tabular}{l l l l}
+    \hline
+    Method name & abbreviation & R package & citation \\
+    \hline
+    multinomial logistic regression & MLR & nnet & citation \\
+    linear discriminate analysis & LDA & MASS & citation \\
+    penalized discrminiate analysis & PDA & mda & citation \\
+    single-hidden-layer neural network & NN & nnet & citation \\
+    random forests & RF & randomForest & citation \\
+    \hline
+  \end{tabular}
+  \caption{table of the methods}
+  \label{tab:methods}
+\end{table}
 

### 2016-05-06 R/clear_cut.r

@@ -151,6 +151,15 @@ selc.melt <- cbind(selc.melt,
                             roc.out, select.roc)))
 selc.melt$npred <- as.numeric(as.character(selc.melt$npred))
 
+# map values of schemes to useful names
+# map values of models to correct abbreviation
+roc.melt$model <- mapvalues(roc.melt$model, from = unique(roc.melt$model),
+                            to = c('MLR', 'NN', 'LDA', 'PDA', 'RF'))
+high.melt$model <- mapvalues(high.melt$model, from = unique(high.melt$model),
+                            to = c('MLR', 'NN', 'LDA', 'PDA', 'RF'))
+selc.melt$model <- mapvalues(selc.melt$model, from = unique(selc.melt$model),
+                            to = c('MLR', 'NN', 'LDA', 'PDA', 'RF'))
+
 
 roc.plot <- ggplot(roc.melt, aes(x = npred, y = ROC))
 roc.plot <- roc.plot + geom_linerange(aes(ymin = ROCmin, ymax = ROCmax))
@@ -185,6 +194,11 @@ oos.melt <- Reduce(rbind, Map(function(x, y)
 oos.melt <- data.frame(oos.melt)
 oos.melt$value <- as.numeric(as.character(oos.melt$value))
 oos.mean <- ddply(oos.melt, .(scheme), summarize, mean = mean(value))
+
+# map values of schemes to useful names
+# map values of models to correct abbreviation
+oos.melt$model <- mapvalues(oos.melt$model, from = unique(oos.melt$model),
+                            to = c('MLR', 'NN', 'LDA', 'PDA', 'RF'))
 
 oos.plot <- ggplot(oos.melt, aes(x = model, y = value))
 oos.plot <- oos.plot + geom_point() + facet_grid(. ~ scheme)

### 2016-06-01 doc/turtles_statistical_learning.tex
put the intro back in the document. graphs of all the pc plots for the paper. sexual dimorphism test graph. bias methods section/

@@
 \begin{abstract}
+  We investigate the morphometric identification of cryptic species using machine learning approaches by examining their implications for a recently proposed cryptic turtle species (\textit{Emys pallida}).
+We collected landmark-based morphometric data from 354 adult \textit{E.
+marmorata/``pallida''} museum specimens.
+We assigned a classification to each specimen for six different binning schemes based on geographic occurrence data recorded in museum collection archives.
+We used multiple machine learning methods, both unsupervised and supervised, to compare different classification hypotheses and asked whether it is possible to determine which amongst a set of classification hypotheses is best.
+  In addition, we applied the above approach to two additional datasets: two well-supported subspecies of \textit{Trachemys scripta} and a set of seven unambiguously distinct species closely related to \textit{E.
+marmorata}.
+The results of this study indicate that there is no clear grouping of \textit{E.
+marmorata/``pallida''} based on plastron shape.
+In contrast, the analysis of the other datasets demonstrate a near perfect classification, which demonstrates that application of the methods to plastron shape data can recover correct results in similar situations.
+Explanations for the lack of grouping in \textit{E.
+marmorata} include that possibility that genetic differentiation is not associated with plastron shape variation in this species complex and/or that local selective pressures (e.g., from hydrological regime) overwhelm morphological differentiation.
+A reconsideration of the methods used to delimit \textit{E.
+``pallida,''} the lack of barriers to gene flow, the strong evidence for widespread admixture between lineages, and the fact that plastron shape can be used to differentiate other emydid species and subspecies suggest that its lack of diagnosability most likely reflects the non-distinctiveness of this proposed taxon.
 \end{abstract}
@@
 \section{Introduction}
+% cryptic diversity
+%   most species are still deliminated solely based on morphology
+Molecular systematics has repeatedly demonstrated the existence of cryptic species that can only be diagnosed using genetic data \citep{Stuart2006,Bickford2007,SchlickSteiner2007,Pfenninger2007,Clare2011,Funk2012}.
+In attempts to streamline the documentation of biodiversity, several methods of species delimitation that rely almost entirely on genetic data have recently been proposed \citep{Pons2006,Carstens2010,Hausdorf2010,O'Meara2010,Yang2010b,Huelsenbeck2011b}.
+Although strong caveats on the utility of these methods have been raised \citep{Bauer2000,Carstens2013}, they are nevertheless being used to name species \citep{Leache2010,Spinks2014}.
+
+The majority of extant taxa, and almost all extinct taxa, are delimited by morphology alone.
+This disjunction complicates interpretations of variation and diversity in deep time, as apparent morphological stasis may not reflect the true underlying diversity \citep{Eldredge1972,Gould1977a,Hunt2008,VanBocxlaer2013}.
+Similarly, for many museum specimens of extant taxa (e.g.
+those preserved in formalin), it is difficult to acquire the genetic data needed for non-morphological species delimitation methods.
+
+These considerations have sparked interest in whether geometric morphometric analyses can capture fine-scale variation that can be used for identifying cryptic species.
+This would make the task of identifying and maintaining endangered or conserved groups much easier and could contribute to improved classifications of extinct taxa and populations.
+Most such studies focus on using morphometrics to discover differences between taxa that were identified by other means \citep{Polly2003,Zelditch2004,Gaubert2005b,Gunduz2007,Polly2007a,Demandt2009}.
+Additionally, there has been work on automated taxon identification and classification of taxa into groups \citep{Baylac2003,Dobigny2003,MacLeod2007,VandenBrink2011}.
+
+Here, we investigate the morphometric identification of cryptic species using machine learning approaches.
+In particular, we ask whether it is possible to determine which amongst a set of classification hypotheses is best and examine the implications of the results for a recently proposed set of cryptic turtle species.
+
+% past work on automatic taxon identification and older approaches to classifying taxa
+% why use machine learning methods
+\subsection{Background and study system}
+Machine learning is an extension of known statistical methodology \citep{Hastie2009} that emphasizes high predictive accuracy and generality at the expense of the interpretability of individual parameters.
+The basic statistical mechanics are supplemented by randomization, sorting, and partitioning algorithms, along with the maximization or minimization of summary statistics, in order to best estimate a general model for all data, both sampled and unsampled \citep{Hastie2009}.
+Machine learning approaches have found use in medical research, epidemiology, economics and automated identification of images such as handwritten zip codes \citep{Hastie2009}.
+The two major classes of machine learning methods are unsupervised and supervised learning.
+Unsupervised learning methods are used with unlabeled data where the underlying structure is estimated; they are analogous to clustering and density estimation methods \citep{Kaufman1990}.
+Supervised learning methods are used with labeled data where the final output of data is known and the rules for going from input to output are inferred.
+These are analogous to classification and regression models \citep{Breiman1984}.
+Our application of the approaches used in this study illustrates only a sampling of the various methods available for clustering observations and fitting classification models.
+
+Geometric morphometric approaches to identifying differences in morphological variation between different classes, including cryptic species, mostly have used methods like linear discriminate analysis and canonical variates analysis \citep{Polly2003,Zelditch2004,Gaubert2005b,Gunduz2007,Polly2007a,Francoy2009,Sztencel-Jabonka2009,MitrovskiBogdanovic2013}.
+Because of their similarity to multivariate approaches like principal components analysis (PCA), these methods are comparatively straightforward ways of understanding the differences in morphology between classes.
+They also benefit from producing results that can be easily visualized, which aids in the interpretation and presentation of data and results.
+Most previous morphometric studies did not assess which amongst a set of alternative classification hypotheses was optimal.
+For example, studies such as those of \citet{Caumul2005a} and \citet{Polly2007a} focused on comparing different aspects of morphology and their fidelity to a classification scheme instead of comparing the fidelity of one aspect of morphology to multiple classification schemes.
+In this context, the study of \citet{Cardini2009a} is noteworthy because they compared morphological variation in marmots at the population, regional, and species level and determined the fidelity of shape to divisions at each of these levels.
+
+Here, we used multiple machine learning methods, both unsupervised and supervised, to compare different classification hypotheses.
+These methods provide different advantages for understanding how to classify taxa, as well as the accuracy of the resulting classifications.
+Although machine learning methods such as neural networks have been applied to studying shape variation \citep{Baylac2003,Dobigny2003,MacLeod2007,VandenBrink2011}, including in the context of automated taxon identification and classification of groups, the number of cases remains limited.
+In the current study, we not only consider pure classification accuracy but also use a statistic of classification strength that reflects the rate at which taxa are both accurately and inaccurately classified.
+
+We analyzed the problem of whether there are distinct subspecies or cryptic species within the western pond turtle, \textit{Emys marmorata} \citep{Baird1852} (formerly \emph{Clemmys marmorata}; see \citealp{Feldman2002}).
+\textit{Emys marmorata} is distributed from northern Washington State, USA to Baja California, Mexico.
+Traditionally, \textit{E.
+marmorata} was classified into two named subspecies: the northern \textit{E.
+marmorata marmorata} and the southern \textit{Emys marmorata pallida} \citep{Seeliger1945}, with a central Californian intergrade zone in between.
+\textit{Emys marmorata marmorata} is differentiated from \textit{E.
+marmorata pallida} by the presence of a pair of triangular inguinal scales and darker neck markings.
+The triangular inguinal plates can sometimes be present in \textit{E.
+marmorata pallida} although they are considerably smaller.
+\citet{Seeliger1945} did not formally include the Baja California populations of \textit{E.
+marmorata} in either taxon, implying the existence of a third distinct but unnamed subspecies.
+
+Previous work on morphological variation in \textit{E.
+marmorata} has focused primarily on differentiation between populations over a portion of the species' total range \citep{Lubcke2007,Germano2008,Germano2009,Bury2010}; comparatively few studies have included specimens from across the entire range \citep{Holland1992}.
+Most of these studies considered how local biotic and abiotic factors may contribute to differences in carapace length and found that size can vary greatly between different populations \citep{Lubcke2007,Germano2008,Germano2009}.
+There also has been interest in size-based sexual dimorphism in \textit{E.
+marmorata} \citep{Holland1992,Lubcke2007,Germano2009}, with males being on average larger than females based on total carapace length and other linear measurements.
+However, the quality of size as a classifier of sex can vary greatly between populations \citep{Holland1992} because of the magnitude of size differences among populations \citep{Lubcke2007,Germano2009}.
+The effect of sexual dimorphism on shape, \textit{sensu} \citet{Kendall1977a}, has not been assessed \citep{Holland1992,Lubcke2007,Germano2008}.
+
+Of particular relevance in the context of cryptic diversity in \textit{E.
+marmorata} is the morphometric analysis of carapace shape carried out by \citet{Holland1992}, who compared populations of \textit{E.
+marmorata} from three areas of the species range.
+This study concluded that geographic distance was a poor indicator of morphological differentiation, and instead hypothesized that geographic features such as breaks between different drainage basis are probably more important barriers to dispersal and interbreeding.
+Additionally, \citet{Holland1992} suggested that morphological differences were more pronounced as the magnitude of barriers and distance increased, but this variation required many variables to adequately capture, implying only very subtle morphological differentiation between putatively distinct populations.
+Finally, Holland concluded that \textit{E.
+marmorata} is best classified as three distinct species: a northern species, a southern species, and a Columbia Basin species.
+This classification is similar to that of \citet{Seeliger1945}, except elevated to the species level and without recognition of a distinct Baja species.
+
+More recently, the phylogeography of \textit{E.
+marmorata} and the possibility of cryptic diversity was investigated using molecular data \citep{Spinks2005,Spinks2010,Spinks2014}.
+Based on mitochondrial DNA, \citet{Spinks2005} recognized four subclades within \textit{E.
+marmorata}, a northern clade, a San Joaquin Valley clade, a Santa Barbara clade, and a southern clade.
+Analyses with nuclear DNA \citep{Spinks2010} and single-nucleotide polymorphism (SNP) data suggest a primarily north--south division in \textit{E.
+marmorata}, although these datasets differed from that of \citet{Spinks2005} in the location of the break point.
+All three studies discussed the potential taxonomic implications of their results, with \citet{Spinks2014} going so far as to strongly advocate for the recognition of at least two species (\emph{E.
+marmorata} and \emph{E.
+pallida}), and a possible third based on populations in Baja California.
+However, they did not discuss in detail the morphological characters that would help to diagnose these species beyond those specified by \citet{Seeliger1945}.
+Given that these characters are variable within the proposed species, and that \citet{Holland1992} described shell shape variation that might be consistent with this taxonomy, a geometric morphometric analysis of shell shape might provide a reliable way to diagnose groups (whether species or subspecies) within \textit{E.
+marmorata}.
+
+In this study, we attempt to estimate the best classification scheme of \textit{E.
+marmorata} based on variation in plastron (ventral shell) shape in order to determine whether this character is consistent with any of the past divisions based on other morphological features or molecular data.
+We are particularly interested in whether any aspect of plastron shape can be used to reliably diagnose the proposed species of \citet{Spinks2014}, and if so, the nature of that shape variation.
+
+\uppercase{Why did we analyze plastron shape? Peter doesn't have a good answer.}
+
+Because of unclear geographic boundaries between subgroups of \textit{E.
+marmorata}, we compare multiple hypotheses of morphologically-- and molecularly--based classification.
+We hypothesize that if morphological variation corresponds to class assignment, it should be possible to determine the best classification hypothesis of \textit{E.
+marmorata} from amongst multiple candidate hypotheses.
+However, if morphological variation does not correspond to any of the standing hypothesis, then supervised learning model generalization performance will be poor.
 
@@
 A range of four Procrustes distances were then calculated as the average of the pairwise distances between each of the replicate configurations.
-These distances were then compared to the average pairwise Procrustes distance between the class centroids for each classification schemes of \textit{E.
-marmorata} being analyzed.
 
-% sex
-% permutation test
-%   is the distance between sex means bigger than expected by random?
+The relative magnitude of digitization error was calculated as the ratio between the average of the within-species replicate distances and the average distance between any two configurations in the replicate dataset.
+The goal of this ratio is to determine if the within group distances are on average smaller than the between individual distances; a value of 0 indicates perfect grouping, a value of 1 indicates no difference between grouping and no grouping, and a value of 1+ indicates that the grouping is counter-intuitive to the data.
+
+Turtles are well known to demonstrate strong sexual dimorphism in carapace shape CITATION.
+To test for issues surrounding sexual dimorphism in our \textit{E.
+marmorata} data set, we used a simple permutation test to determine if the distance between the mean female and male shapes is greater than expected when the sex labels are randomly shuffled.
+Because not all of our specimens have sex identifications associated with them, this analysis was done using a subset of the data.
 
@@
 
-\subsection{Biasing effects}
+\subsection{Geometric morphometrics}
+
+\begin{figure}[ht]
+  \centering
+  \begin{subfigure}[b]{0.4\textwidth}
+    \caption{}
+    \includegraphics[width = \textwidth]{figure/cc7_pc_graph}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.4\textwidth}
+    \caption{}
+    \includegraphics[width = \textwidth]{figure/tra_pc_graph}
+  \end{subfigure}
+  \caption{CAPTION}
+  \label{fig:other_pca}
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/emys_pc_graph}
+  \caption{CAPTION}
+  \label{fig:emys_pca}
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/sex_test_hist}
+  \caption{CAPTION}
+  \label{fig:sex_test}
+\end{figure}
 

### 2016-06-01 R/digit_error.r

@@ -167,7 +167,7 @@ species.ratio2 <- mean(within.species) / mean(turt.dist[lower.tri(turt.dist)])
 
 # ratio of
 #   average within group distance / average between individuals distance
-#   1+ = grouping not meaningful
+#   1+ = grouping is counter-intuitive
 #   1 = no grouping
 #   0 = amazing grouping
 # scheme.ratio

### 2016-06-01 R/sex_diff.r

@@ -2,6 +2,9 @@ library(shapes)
 library(reshape2)
 library(stringr)
 library(geomorph)
+library(ggplot2)
+library(scales)
+library(grid)
 source('../R/df2array.r')
 source('../R/supervised_mung.r')
 
@@ -58,4 +61,14 @@ for(ii in seq(1000)) {
   null[ii] <- sum(rowSums((sm[[1]] - sm[[2]])^2))
 }
 
+dfg <- data.frame(null = null)
 
+
+theme_set(theme_minimal())
+dfgp <- ggplot(dfg, aes(x = null))
+dfgp <- dfgp + geom_histogram(colour = 'darkgrey', fill = 'lightgrey')
+dfgp <- dfgp + geom_vline(xintercept = sex.mean, size = 2)
+dfgp <- dfgp + scale_x_continuous(breaks = pretty_breaks(4))
+dfgp <- dfgp + labs(x = 'Procrustes distance', y = 'Frequency')
+ggsave(plot = dfgp, filename = '../doc/figure/sex_test_hist.png',
+       width = 4, height = 3)

### 2016-06-08 doc/draft_turtle.tex
started captions. added package citations

@@
 
-\subsection{Unsupervised learning}
-
-Gap statistic values from the PAM clustering are mostly equivalent for all potential clusters, indicating that none of the proposed clusters are statistically different from each other as evidenced by the confidence intervals around these estimates (Fig.
-\ref{fig:gap}).
-No clustering result is statistically greater than that with only a single cluster, because of this we favor the interpretation that there is no latent structure in the data detected by this analysis.
-While the solution with two groupings does have the greatest mean gap statistic, there is no geographical signal in the results of this clustering solution (Fig.
-\ref{fig:gap_map}).
-Additionally, there is no correspondence between these two clusters and the sex of the specimens (\(\chi^{2}\) 0.7685, df = 1, \textit{p}-value 0.38).
-Because of this, we assert that 1) there is no means of naturally partitioning \textit{E.
-marmorata} plastron shape into distinct subgroups without reference to external information and 2) sexual dimorphism most likely is not confounding these analyses.
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[height = 0.5\textheight, width = \textwidth, keepaspectratio = true]{figure/gap_res}
-  \caption{Results from PAM clustering of the Riemannian shape distance for 8 different number of clusters.
-Vertical lines are 2 standard deviation of the mean determined from 500 resamples as defined in \citep{Tibshirani2001a}.}
-  \label{fig:gap}
-\end{figure}
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[height = 0.5\textheight, width = \textwidth, keepaspectratio = true]{figure/gap_map}
-  \caption{Comparison of geographic distribution of clustered observations from the 2 clustering PAM solution.
-Colour and shape correspond to each of the groups.
-There is clearly no geographic signal in the data.
-Points are slightly transparent as 354 points are plotted but many of them overlap.
-354 points are plotted, some of which overlap.}
-  \label{fig:gap_map}
-\end{figure}
-
-\subsection{Supervised learning}
-% model selection/fit
 AUC--based model selection revealed some important patterns of variation and congruence between the classification schemes and the actual data.
@@
 
-For both datasets, the distribution of ROC scores from 1000 bootstrap replicates are tightly clustered near AUC = 1 (Fig.
-\ref{fig:other_boot}), in contrast to the results from the \textit{E.
-marmorata} case (Fig.
-\ref{fig:gen_hist}).
-These results demonstrate that when there are distinctions between the states of the classification schemes, the methods used here can recover them.
 % insert table here with the AUC values for in and out of sample results

### 2016-06-08 doc/turtles_statistical_learning.tex
started captions. added package citations

@@
 We conducted all subsequent analyses on the resulting ``half'' plastra.
-We superimposed the plastral landmark configurations using generalized Procrustes analysis \citep{Dryden1998a}, after which, we calculated the principal components (PC) of shape using the \texttt{shapes} package for R \citep{2015,Dryden2013}.
+We superimposed the plastral landmark configurations using generalized Procrustes analysis \citep{Dryden1998a}, after which, we calculated the principal components (PC) of shape using the \texttt{shapes} package for R \citep{R2016,Dryden2013}.
 
@@
 Finally, the predictive performance metric is the averaged across all rounds of \(k\)-fold cross-validation.
-This process was implemented using the R package \texttt{caret} CITATION.
+This process was implemented using the R package \texttt{caret} \citep{KuhnMAN2013}.
 
@@
     \hline
-    multinomial logistic regression & MLR & nnet & citation \\
-    linear discriminate analysis & LDA & MASS & citation \\
-    penalized discrminiate analysis & PDA & mda & citation \\
-    single-hidden-layer neural network & NN & nnet & citation \\
-    random forests & RF & randomForest & citation \\
+    multinomial logistic regression & MLR & nnet & \citet{Venables2002} \\
+    linear discriminate analysis & LDA & MASS & \citet{Venables2002} \\
+    penalized discrminiate analysis & PDA & mda & \citet{mdapack} \\
+    single-hidden-layer neural network & NN & nnet & \citet{Venables2002} \\
+    random forests & RF & randomForest & \citet{Liaw2002} \\
     \hline
@@
 
+Analysis of the seven morphologically-distinct species and the \textit{T.
+scripta scripta}--\textit{T.
+scripta elegans} datasets indicate that these classifications are sufficiently morphologically distinct.
+Both in-sample and out-of-sample classification have AUC values of approximately 1 for all methods (Fig.
+\ref{fig:other_sel}, \ref{fig:other_oos}).
+For both datasets, the ROC scores from testing datasets are tightly clustered near AUC = 1 (Fig.
+\ref{fig:other_oos}).
+These results demonstrate that when there are distinctions between the states of the classification schemes, the methods used here can recover them.
+
 \begin{figure}[ht]
@@
 
+AUC--based model selection revealed some important patterns of variation and congruence between the classification schemes and the actual data.
+Generally, the best performing models tended to include about half the total number of possible PCs (Fig.
+\ref{fig:emys_sel}).
+As was the case with the \textit{Trachemys} dataset (Fig.
+\ref{fig:other_sel}), the neural net models are the least well-performing models compared to all others used (Fig.
+\ref{fig:emys_sel}).
+
 \begin{figure}[ht]
@@
   \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/emys_model_sel}
-  \caption{CAPTION}
+  \caption{AUC values for models of varying complexity fit to the \textit{Emys marmorata} training datasets for each classification scheme.
+Standard errors on AUC estimates are calculated from 10 rounds of 5-fold cross-validation.
+Indicated are the best performing and the selected models, in red and blue respectively.
+In some cases, there is no difference in complexity between the best and selected models.}
   \label{fig:emys_sel}
 \end{figure}
+
+Observed AUC values for all of the optimal models are not exceptionally high (Fig.
+\ref{fig:emys_sel}).
+In most cases the different proposed classification schemes are generally poor descriptors of the observed variation.
+It appears that the data set is overwhelmed by noise, making any accurate classifications difficult at best.
+This observation is cemented with the generalizations of the models to the testing data set (Fig.
+\ref{fig:emys_oos}).
 
@@
   \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/emys_oos_sel}
-  \caption{CAPTION}
+  \caption{Comparison of out-of-sample AUC estimates from the predictions of selected models (Fig.
+\ref{fig:emys_sel}), grouped by classification scheme.
+The horizontal line in each panel corresponds to the average AUC value across all models of that classification scheme.}
   \label{fig:emys_oos}
@@
 
+Mean AUC values for the model generalizations, in most cases, are approximately equal to the observed AUC values from the training data set (Fig.
+\ref{fig:emys_sel}, \ref{fig:emys_oos}).
+The  cases in which the AUC from the  generalizations is less than the observed indicate poor model fit and a poor classification scheme.
+Comparison of AUC values from the model generalizations do not indicate a clear ``best'' classification scheme (Fig.
+\ref{fig:emys_sel}, \ref{fig:emys_oos}).
+
 \section{Discussion}
+
+The results of this study show that there is no clear grouping of \textit{E. marmorata} based on plastron shape.
+
+The unsupervised learning results indicate that no grouping is optimal or particularly better than a single grouping, which is consistent with the results from the generalizations of the supervised learning models.
+The classification schemes used in the supervised learning models correspond, loosely, to unsupervised learning solutions with multiple groups.
+Because unsupervised learning solutions with multiple groups are poor descriptors of the observed variation, it is important to see this generally supported by the supervised learning results.
+
+The results from fitting the various supervised learning models to each of the classification schemes generally shows that no one scheme is ``best.'' Possible explanations include that the genetic differentiation is not associated with plastron shape variation and/or that local selective pressures (e.g.
+from hydrological regime) overwhelm morphological differentiation.
+This makes sense given that shell shape is influenced by selection for both protection and streamlining, but not necessary mate choice \citep{Rivera2008,Rivera2011,Stayton2011,Rivera2014,Pollyb} and that shell shape in \textit{E.
+marmorata} is known to vary among populations inhabiting water bodies with different flow regimes \citep{Holland1992,Lubcke2007,Germano2009}.
+Plastron shape does not seem to preserve a strong phylogenetic signal at the interspecific level in emydine turtles \citep{Angielczyk2011}, and our current results suggest that this may be the case for phylogeographic signal within emydine species as well.
+A final possibility (explored below) is that the proposed classification schemes themselves do not represent significant evolutionary lineages.
+
+Both the low out-of-sample AUC values (\(< 0.9\)) and the significant difference between the correctly and incorrectly classified observations support the conclusion that none of the hypothesized classification schemes are good descriptors of the observed plastral variation within \textit{E.
+marmorata}.
+
+Nevertheless, it is important to note that plastron shape is an extremely effective method for differentiating classes in the additional datasets we investigated.
+The magnitude of shape differences between the species (measured as Procrustes distance between species' mean shapes) is approximately an order of magnitude greater than the differences between the \textit{E.
+marmorata} subgroups, and the machine learning methods had no trouble accurately classifying the specimens correctly.
+However, the magnitude of the shape differences between the \textit{T.
+scripta} subspecies is comparable to theose separating the different \textit{E.
+marmorata} subgroups, yet even in this case the machine learning methods returned an almost perfect classification.
+These results demonstrate that plastron shape is normally a good marker for differentiating real subgroups in close relatives of \textit{E.
+marmorata}, and that our lack of results for \textit{E.
+marmorata} is not simply a shortcoming of the methods we applied.
+Indeed, they beg the question of what factors have suppressed morphological differentiation of plastron shape in \textit{E.
+marmorata} and \textit{E.
+pallida} if they are distinct species.
+Invoking issues such as the role of the plastron in protection or the need for streamlining are insufficient because the other species are expected to be subject to similar constraints \citep{Stayton2011,Pollyb}.
+While it may seem counterintuitive that plastron shape is both useful for species delimitation but has weak or absent phylogenetic signal, it is important to note that these are different goals.
+While phylogenetically similar species may not be morphologically similar, the variation within a species is much less than the variation between species which leads to plastron shape being a useful tool for species delimitation but not phylogenetic reconstruction.
+
+\subsection{Is there more than one species of Western Pond Turtle?}
+
+The lack of morphological support for the distinctiveness of \textit{E.
+pallida} does not, on its own, preclude the recognition of this taxon.
+However, this apparent lack of congruence does prompt a reexamination of the methods and concepts that led to that taxonomic revision, especially considering that plastron shape is demonstrably capable of differentiating species and subspecies among emydids.
+In other words, before we can assess the significance of the morphological non-diagnosablity, it is essential to evaluate the methods and concepts that led to the initial taxonomic revision.
+
+\citet{Spinks2014} elevated \textit{E.
+pallida} based on a species delimitation analysis of SNP data using BPP \citep{Yang2010b}.
+However, \citet{Spinks2014} did not heed the caveats about species delimitation methods raised by \citet{Carstens2013}.
+In addition to specifically addressing the shortcomings of validation methods such as BPP that rely on guide trees and ``should be interpreted with caution,'' \citet{Carstens2013} also strongly emphasized that ``Inferences regarding species boundaries based on genetic data alone are likely inadequate, and species delimitation should be conducted with consideration of the life history, geographical distribution, morphology and behaviour (where applicable) of the focal system\dots'' These caveats evoke the development of the Unified Species Concept \citep{Dayrat2005a,DeQueiroz2007b}, Integrative Taxonomy \citep{Padial2010}, and other pluralist approaches to species delimitation.
+None of these considerations were brought to bear on the \textit{E.
+marmorata} system until now, and in doing so we find the proposal that \textit{E.
+pallida} is a distinct species to be lacking in a normally robust morphological marker.
+
+The natural history and geographical distribution of \textit{E.
+marmorata} and \textit{E.
+pallida} also make the recognition of these taxa implausible.
+The data from \citet{Spinks2014} show extensive introgression and admixture in Central California, which is expected because there are no significant barriers to gene flow in this region.
+Combined with the well-demonstrated ability for testudinoid turtles, including emydids and even \textit{Emys}, to hybridize (e.g.
+\citealt{Buskirk2005,Spinks2009,Parham2013}) it is hard to imagine how \textit{E.
+marmorata} and \textit{E.
+pallida} could maintain their integrity in the face of such admixture.
+Because the geography, natural history, demonstrated genetic admixture of \textit{E.
+marmorata}, and comparisons with other morphologically diagnosable species and subspecies conflict with the recognition of \textit{E.
+pallida}, we hypothesize that our inability to classify the morphological data by proposed species is because \textit{E.
+pallida} is not a distinct species.
+
+We fully agree with \citet{Spinks2014} that \textit{E.
+marmorata} (\textit{sensu lato}) is a species deserving of strong conservation efforts, and we do not wish to trivialize this need.
+Moreover, the genetic diversity uncovered by the analysis of \citet{Spinks2014} should be explicitly accounted for in any conservation plan.
+Given the apparent lack of morphological distinction, however, we consider that this diversity should be considered Evolutionary Significant Units or Distinct Population Segments instead of distinct species.
+
+Finally, it is important to note that the data and analyses we present do not let us definitively say whether the apparent lack of morphological divergence within \textit{E.
+marmorata} truly reflects the presence of a single species, or if it is an artifact of plastron shape being a poor morphological marker for phylogenetic and phylogeographic divergences in the case of \textit{E.
+marmorata}, because we could not carry out our morphometric analyses on the specimens from which the genetic data were obtained.
+The comparisons with the other emydid taxa suggest that it is because \textit{E.
+marmorata} is a single species.
+However, tests of both our preferred conclusion (\textit{E.
+marmorata} as a single species) and that \citet{Spinks2014} should include morphological and molecular analyses of the same set of voucher specimens, as well as additional tests of species delimitation using alternative methods and corroborating evidence as suggested by \citet{Carstens2013}.
+From a morphological standpoint, support for the validity of ``\textit{E.
+pallida}'' may come from other aspects of morphology, such as carapace shape or other features.
+Likewise, further investigation of the phylogeographic utility of plastron shape in other turtle species will help to clarify whether the lack of differentiation seen in \textit{E.
+marmoarata}, and the strong differentiation among the other emydids, is typical or an unusual case.
 

### 2016-06-08 R/clear_cut.r

@@ -1,5 +1,4 @@
 library(shapes)
-library(MuMIn)
 library(geomorph)
 library(plyr)
 library(stringr)
@@ -25,7 +24,17 @@ source('../R/support_functions.r')
 source('../R/multiclass_roc.r')
 source('../R/train_and_test.r')
 
+theme_set(theme_bw())
+cbp <- c('#000000', '#E69F00', '#56B4E9', '#009E73', 
+         '#F0E442', '#0072B2', '#D55E00', '#CC79A7')
+cbp.long <- c('#000000', '#004949', '#009292', '#FF7DB6', '#FFB6DB', 
+              '#490092', '#006DDB', '#B66DFF', '#6DB6FF', '#B6DBFF', 
+              '#920000', '#924900', '#DBD100', '#24FF24', '#FFFF6D')
+
+grab <- laply(seq(5), function(x) seq(from = x, to = length(cbp.long), by = 5))
+cbp.ord <- cbp.long[t(grab)]
 turt.out <- list()
+
 
 # actually start doing analysis...
 newturt <- list.files('../data/new_turtle', 

### 2016-06-08 doc/turtles_statistical_learning.tex
fixed a citation issue

@@
 from hydrological regime) overwhelm morphological differentiation.
-This makes sense given that shell shape is influenced by selection for both protection and streamlining, but not necessary mate choice \citep{Rivera2008,Rivera2011,Stayton2011,Rivera2014,Pollyb} and that shell shape in \textit{E.
+This makes sense given that shell shape is influenced by selection for both protection and streamlining, but not necessary mate choice \citep{Rivera2008,Rivera2011,Stayton2011,Rivera2014,Polly2016} and that shell shape in \textit{E.
 marmorata} is known to vary among populations inhabiting water bodies with different flow regimes \citep{Holland1992,Lubcke2007,Germano2009}.
@@
 pallida} if they are distinct species.
-Invoking issues such as the role of the plastron in protection or the need for streamlining are insufficient because the other species are expected to be subject to similar constraints \citep{Stayton2011,Pollyb}.
+Invoking issues such as the role of the plastron in protection or the need for streamlining are insufficient because the other species are expected to be subject to similar constraints \citep{Stayton2011,Polly2016}.
 While it may seem counterintuitive that plastron shape is both useful for species delimitation but has weak or absent phylogenetic signal, it is important to note that these are different goals.

### 2016-07-07 doc/turtles_statistical_learning.tex
measurement error section

@@
 We assigned a classification to each specimen for six different binning schemes based on geographic occurrence data recorded in museum collection archives.
-We used multiple machine learning methods, both unsupervised and supervised, to compare different classification hypotheses and asked whether it is possible to determine which amongst a set of classification hypotheses is best.
+We used multiple supervised machine learning methods to compare different classification hypotheses and asked whether it is possible to determine which amongst a set of classification hypotheses is best.
   In addition, we applied the above approach to two additional datasets: two well-supported subspecies of \textit{Trachemys scripta} and a set of seven unambiguously distinct species closely related to \textit{E.
@@
 
-Here, we used multiple machine learning methods, both unsupervised and supervised, to compare different classification hypotheses.
+Here, we used multiple supervised machine learning methods to compare different classification hypotheses.
 These methods provide different advantages for understanding how to classify taxa, as well as the accuracy of the resulting classifications.
@@
   \centering
+  \caption{Table of the supervised learning methods used in this analysis.}
   \begin{tabular}{l l l l}
@@
   \end{tabular}
-  \caption{table of the methods}
   \label{tab:methods}
@@
 \subsection{Geometric morphometrics}
+
+The results of the PCA of plastron shape in both the seven species and \textit{Trachemys} datasets demonstrate strong association between shape and the recognized classification schemes (Fig.
+\ref{fig:other_pca}).
+Additionally, there appears to be both little difference in centroid size across individuals as well as little correspondence between centroid size and the PC axes.
 
@@
   \end{subfigure}
-  \caption{CAPTION}
+  \caption{Scatterplots of the first two axes from the PCA results for both the seven species (a) and \textit{Trachemys} (b) datasets.
+Point color corresponds to the species within each dataset.
+Point size is proportional to individual centroid size.
+In parentheses next to the axis labels are the percent of total variation accounted for by that axis.}
   \label{fig:other_pca}
 \end{figure}
+
+The results of the PCA of plastron shape in the \textit{Emys marmorata} dataset show no clear connection between plastron shape and any of the six proposed classification schemes (Fig.
+\ref{fig:emys_pca}).
+The first PC axis of shape variation appears to be very structured by differences in individual centroid size (Fig.
+\ref{fig:emys_pca}); this was the motivation for including centroid size and its interaction with PC1 as predictors in all of the supervised learning models.
 
@@
   \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/emys_pc_graph}
-  \caption{CAPTION}
+  \caption{Plot of the first two axes of the PCA of the \textit{Emys marmorata} dataset.
+Each panel corresponds to each of the six different proposed classification schemes analyzed as part of this study.
+Point color corresponds to the classes within each scheme, with the titles corresponding to subspecies or regional codes.
+Point size is proportional to centroid size.}
   \label{fig:emys_pca}
@@
 
+Analysis of the differences between sexes of \textit{E.
+marmorata} indicates that sex does not appear to strongly structure differences in shape (Fig.
+\ref{fig:sex_test}).
+The difference between the centroids of both sexes is very small; the sexes overlap much more than expected from a null distribution formed by permuting the sex-labels.
+
 \begin{figure}[ht]
   \centering
-  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/sex_test_hist}
-  \caption{CAPTION}
+  \includegraphics[height = 0.3\textheight, width = \textwidth, keepaspectratio = true]{figure/sex_test_hist}
+  \caption{Comparison of observed Procrustes distance between the centroids of each sex (vertical line) to a null distribution generated from 1000 permutations of the sex-labels.
+This result indicates that the difference between the centroids is as small/smaller than expected by random.}
   \label{fig:sex_test}
 \end{figure}
+
+Comparison of the within to between Procrustes distances of the digitization replicates gives an approximate estimate of the error between distinct groupings (Table \ref{tab:rep_res}).
+The ratio of the average within-individual distance to the average distance between individuals for the replicated datasets is 1.11; this indicates that the grouping is slightly counter-intuitive to the data.
+This value provides a baseline by which to understand how distinct the groupings are, where otherratios are compared to the correction ratio \(1.11/1\).
+
+The results from the seven species and \textit{Trachemys} datasets indicate that both of these classification schemes are more recognizable than not given our estimate of digitization error (Table \ref{tab:rep_res}).
+In contrast, the different \textit{E.
+marmorata} classification schemes appear to be barely be distinct, with their within:between ratios being approximate 1 which indicates that the differences between groupings are approximatly as different as any two random individuals are (Table \ref{tab:rep_res}).
+
+\begin{table}
+  \centering
+  \caption{Results from the within-individual to between-individual Procrustes distances for the replicated shapes, the seven species dataset, the \textit{Trachemys} dataset, and each of the \textit{Emys marmorata} classification schemes.}
+  \begin{tabular}{l l c c}
+    \textbf{Dataset} & \textbf{Scheme} & \textbf{Ratio} & \textbf{Corrected ratio} \\
+    \hline
+    Replicates & & 1.11 & \\
+    \hline
+    Seven species & & 0.33 & \\
+    \textit{Trachemys} & & 0.76 & \\
+    \hline
+    \textit{E. marmorata} & Mito 1 & 0.92 & 1.02 \\
+     & Mito 2 & 0.92 & 1.02 \\
+     & Mito 3 & 0.88 & 0.98 \\
+     & Morph 1 & 0.98 & 1.09 \\
+     & Morph 2 & 0.83 & 0.92 \\
+     & Nuclear & 0.90 & 1 \\
+    \hline
+  \end{tabular}
+  \label{tab:rep_res}
+\end{table}
 
@@
   \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/other_model_sel}
-  \caption{CAPTION}
+  \caption{Comparisons of model fit to the training dataset for each of the supervised learning methods; the results from the seven species dataset are presented in the left column, while those from the \textit{Trachemys} dataset are presented in the right column.
+Models were fit to datasets of varying complexity, with the number of parameters listed along the x-axis.
+Model fit is measured as the area under the receiver operating characteristic (AUC), which ranges from 0.5 to 1.
+Error bars correspond to one standard error estimated from 10 rounds of 5-fold cross-validation.
+The red dot corresponds to the model fit with the highest mean AUC while the blue dot corresponds to the model selected for further analysis.
+In some cases, there is no difference in complexity between the best and selected models.}
   \label{fig:other_sel}
@@
   \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/other_oos_sel}
-  \caption{CAPTION}
+  \caption{The results of out-of-sample predictive performance of the selected models for both the seven species (left) and \textit{Trachemys} datasets.
+Predictive performance is measured as the area under the receiver operating characteristic (AUC), which ranges from 0.5 to 1.
+Points correspond to the individual out-of-sample predictive performance of the specific model, indicated along the x-axis.
+The horizontal bars correspond the average out-of-sample predictive performance of all the models.}
   \label{fig:other_oos}
@@
 The results of this study show that there is no clear grouping of \textit{E. marmorata} based on plastron shape.
-
-The unsupervised learning results indicate that no grouping is optimal or particularly better than a single grouping, which is consistent with the results from the generalizations of the supervised learning models.
-The classification schemes used in the supervised learning models correspond, loosely, to unsupervised learning solutions with multiple groups.
-Because unsupervised learning solutions with multiple groups are poor descriptors of the observed variation, it is important to see this generally supported by the supervised learning results.
 

### 2016-07-07 doc/turtles_statistical_learning.tex
sent draft to coauthors

@@
 
-This approach is in many ways analogous to PCA regresion.
+This approach is in many ways analogous to PCA regression.
 PCA regression takes advantage of two aspects of PCA for improving regression fit \citep{Hastie2009}.
-Because the PCs of shape are by definition orthagonal, allowing them to easily as independent predictors or features of class membership without fear of colinearity.
+Because the PCs of shape are by definition orthogonal, allowing them to easily as independent predictors or features of class membership without fear of collinearity.
 
@@
 In classification studies, such as this one, a common metric of performance is the receiver operating characteristic (ROC) which is the relationship between the false and true positive rates \citep{Hastie2009}.
-The area under the ROC curce (AUC) is then the derived estimate of the model performance; AUC ranges from 0.5 to 1 which corresponding to performance similar to random guesses and perfect classificationrates, respectively \citep{Hastie2009}.
-Both ROC and AUC are preferrable to simple classificaiton accuracy when class membership is unbalanced, as it is in these analyses \citep{Hastie2009}.
+The area under the ROC curve (AUC) is then the derived estimate of the model performance; AUC ranges from 0.5 to 1 which corresponding to performance similar to random guesses and perfect classification rates, respectively \citep{Hastie2009}.
+Both ROC and AUC are preferable to simple classification accuracy when class membership is unbalanced, as it is in these analyses \citep{Hastie2009}.
 The standard ROC and AUC calculations are defined only for binary classifications, which is not the case for our seven species and \textit{Emys} complex datasets.
-To generalize this approach for situations with mulitple response classes, we used an all-against-one strategy where the model AUC is the average of the AUC values from the multiple binary comparisons of one class compared to all others \citep{Hand2001}.
+To generalize this approach for situations with multiple response classes, we used an all-against-one strategy where the model AUC is the average of the AUC values from the multiple binary comparisons of one class compared to all others \citep{Hand2001}.
 
@@
 %   all assume that predictors have additive effect (e.g. independent)
-Instead of relying on a single supervised learning method, we chose to use an ensemble of mulitple approaches so that the congruence bewteen the them could be used as a means of ``support'' for one conclusion or another.
+Instead of relying on a single supervised learning method, we chose to use an ensemble of multiple approaches so that the congruence between the them could be used as a means of ``support'' for one conclusion or another.
 The supervised learning methods used here are described in Table \ref{tab:methods}.
@@
 The ratio of the average within-individual distance to the average distance between individuals for the replicated datasets is 1.11; this indicates that the grouping is slightly counter-intuitive to the data.
-This value provides a baseline by which to understand how distinct the groupings are, where otherratios are compared to the correction ratio \(1.11/1\).
+This value provides a baseline by which to understand how distinct the groupings are, where other ratios are compared to the correction ratio \(1.11/1\).
 
@@
 In contrast, the different \textit{E.
-marmorata} classification schemes appear to be barely be distinct, with their within:between ratios being approximate 1 which indicates that the differences between groupings are approximatly as different as any two random individuals are (Table \ref{tab:rep_res}).
+marmorata} classification schemes appear to be barely be distinct, with their within:between ratios being approximate 1 which indicates that the differences between groupings are approximately as different as any two random individuals are (Table \ref{tab:rep_res}).
 
@@
     \hline
-    Seven species & & 0.33 & \\
-    \textit{Trachemys} & & 0.76 & \\
+    Seven species & & 0.33 & 0.37 \\
+    \textit{Trachemys} & & 0.76 & 0.84 \\
     \hline
@@
 However, the magnitude of the shape differences between the \textit{T.
-scripta} subspecies is comparable to theose separating the different \textit{E.
+scripta} subspecies is considered comparable to those separating the different \textit{E.
 marmorata} subgroups, yet even in this case the machine learning methods returned an almost perfect classification.

### 2017-07-10 doc/turtles_statistical_learning.tex
all models run with new schemes. story doesnt change. need more plots

@@
 %\title{How cryptic is cryptic diversity? Machine learning approaches to classifying morphological variation in the Pacific Pond Turtle (\textit{Emys marmorata})}
-\title{Supervised learning approaches to classifying morphological viariation: picking amongst taxonomic hypotheses for the Pacific Pond Turtle (\textit{Emys marmorata})}
+\title{Supervised learning approaches to classifying morphological variation: assessing taxonomic hypotheses for the Pacific Pond Turtle (\textit{Emys marmorata})}
 \author[1]{Peter D Smits}%\thanks{psmits@uchicago.edu}}
@@
 \begin{abstract}
-  We investigate the morphometric identification of cryptic species using machine learning approaches by examining their implications for a recently proposed cryptic turtle species (\textit{Emys pallida}).
+  We investigate morphometric identification of cryptic species using machine learning approaches by examining their implications for a recently proposed cryptic turtle species (\textit{Emys pallida}).
 We collected landmark-based morphometric data from 354 adult \textit{E.
 marmorata/``pallida''} museum specimens.
-We assigned a classification to each specimen for six different binning schemes based on geographic occurrence data recorded in museum collection archives.
-We used multiple supervised machine learning methods to compare different classification hypotheses and asked whether it is possible to determine which amongst a set of classification hypotheses is best.
-  In addition, we applied the above approach to two additional datasets: two well-supported subspecies of \textit{Trachemys scripta} and a set of seven unambiguously distinct species closely related to \textit{E.
+We assigned a classification to each specimen for six different binning schemes using geographic occurrence data recorded in museum collection archives.
+We used multiple supervised machine learning methods to compare different classification hypotheses and asked whether it is possible to determine which among this set of classification hypotheses is best.
+  In addition, we applied the above approach to two other datasets: two well-supported subspecies of \textit{Trachemys scripta} and a set of seven unambiguously distinct species closely related to \textit{E.
 marmorata}.
@@
 marmorata/``pallida''} based on plastron shape.
-In contrast, the analysis of the other datasets demonstrate a near perfect classification, which demonstrates that application of the methods to plastron shape data can recover correct results in similar situations.
+In contrast, analysis of the other datasets resulted in near perfect classifications, which demonstrates that application of the methods to plastron shape data can recover correct results in similar situations.
 Explanations for the lack of grouping in \textit{E.
@@
 A reconsideration of the methods used to delimit \textit{E.
-``pallida,''} the lack of barriers to gene flow, the strong evidence for widespread admixture between lineages, and the fact that plastron shape can be used to differentiate other emydid species and subspecies suggest that its lack of diagnosability most likely reflects the non-distinctiveness of this proposed taxon.
+``pallida,''} the lack of barriers to gene flow, the strong evidence for widespread admixture between lineages, and the fact that plastron shape can be used to differentiate other emydid species and subspecies suggest that the non-diagnosability of \textit{E.
+``pallida''} stems from it not being a distinct species.
 \end{abstract}
@@
 
-The majority of extant taxa, and almost all extinct taxa, are delimited by morphology alone.
+In contrast to those genetically-diagnosed species, the majority of extant taxa, and almost all extinct taxa, are delimited by morphology alone.
 This disjunction complicates interpretations of variation and diversity in deep time, as apparent morphological stasis may not reflect the true underlying diversity \citep{Eldredge1972,Gould1977a,Hunt2008,VanBocxlaer2013}.
-Similarly, for many museum specimens of extant taxa (e.g.
+It also has serious implications for our records of modern biodiversity: for many museum specimens of extant taxa (e.g.
 those preserved in formalin), it is difficult to acquire the genetic data needed for non-morphological species delimitation methods.
@@
 Here, we investigate the morphometric identification of cryptic species using machine learning approaches.
-In particular, we ask whether it is possible to determine which amongst a set of classification hypotheses is best and examine the implications of the results for a recently proposed set of cryptic turtle species.
+In particular, we ask whether it is possible to determine which among a set of classification hypotheses is best and examine the implications of the results for a recently proposed set of cryptic turtle species.
 
@@
 Machine learning approaches have found use in medical research, epidemiology, economics and automated identification of images such as handwritten zip codes \citep{Hastie2009}.
-The two major classes of machine learning methods are unsupervised and supervised learning.
+% additional/more specific CITATIONS?
+
+There are two major classes of machine learning method: unsupervised and supervised learning.
 Unsupervised learning methods are used with unlabeled data where the underlying structure is estimated; they are analogous to clustering and density estimation methods \citep{Kaufman1990}.
@@
 These are analogous to classification and regression models \citep{Breiman1984}.
-Our application of the approaches used in this study illustrates only a sampling of the various methods available for clustering observations and fitting classification models.
+Our application of the approaches used in this study illustrates only a sampling of the various methods available for fitting classification models.
+The classification methods used in this study were chosen because they are suited for more than two response classes.
 
@@
 Here, we used multiple supervised machine learning methods to compare different classification hypotheses.
-These methods provide different advantages for understanding how to classify taxa, as well as the accuracy of the resulting classifications.
+Each of these methods provide different advantages for understanding how to classify taxa, as well as the accuracy of the resulting classifications.
 Although machine learning methods such as neural networks have been applied to studying shape variation \citep{Baylac2003,Dobigny2003,MacLeod2007,VandenBrink2011}, including in the context of automated taxon identification and classification of groups, the number of cases remains limited.
@@
 marmorata} has focused primarily on differentiation between populations over a portion of the species' total range \citep{Lubcke2007,Germano2008,Germano2009,Bury2010}; comparatively few studies have included specimens from across the entire range \citep{Holland1992}.
-Most of these studies considered how local biotic and abiotic factors may contribute to differences in carapace length and found that size can vary greatly between different populations \citep{Lubcke2007,Germano2008,Germano2009}.
+Most of these studies considered how local biotic and abiotic factors may contribute to differences in carapace length, and they found that size can vary greatly between different populations \citep{Lubcke2007,Germano2008,Germano2009}.
 There also has been interest in size-based sexual dimorphism in \textit{E.
@@
 marmorata} from three areas of the species range.
-This study concluded that geographic distance was a poor indicator of morphological differentiation, and instead hypothesized that geographic features such as breaks between different drainage basis are probably more important barriers to dispersal and interbreeding.
-Additionally, \citet{Holland1992} suggested that morphological differences were more pronounced as the magnitude of barriers and distance increased, but this variation required many variables to adequately capture, implying only very subtle morphological differentiation between putatively distinct populations.
+Holland concluded that geographic distance was a poor indicator of morphological differentiation, and instead hypothesized that geographic features such as breaks between different drainage basis are probably more important barriers to dispersal and interbreeding.
+Additionally, he suggested that morphological differences were more pronounced as the magnitude of barriers and distance increased, but this variation required many variables to adequately capture, implying only very subtle morphological differentiation between putatively distinct populations.
 Finally, Holland concluded that \textit{E.
@@
 \subsection{Specimens, sampling, morphometrics}
-Three different landmark-based morphometric datasets describing plastron variation were assembled for this analysis: specimens from seven distinct emydine species, \textit{T.
-scripta} specimens from both subspecies, and \textit{E.
-marmorata} specimens from across its geographic range.
-We chose to focus on adults because significant changes in plastron shape occur over the course of ontogeny in \textit{E.
+Three different landmark-based morphometric datasets describing plastron variation were assembled for this analysis: 1) specimens from seven distinct emydine species; 2) \textit{T.
+scripta} specimens from the two main subspecies (\textit{T.
+scripta elgans} and \textit{T.
+scripta scripta}); and 3) \textit{E.
+marmorata} specimens from across the species' geographic range.
+The first two datasets are intended to serve as a basic test of whether machine learning techniques can differentiate species-level groupings of emydine turtles using plastron shape.
+Dataset 1 should be relatively easy, given the high level of plastron shape disparty that exists among these species \citep{Angieczyk2011}, whereas dataset 2 should be a case that is more comparable to our \textit{E.
+marmorata} dataset.
+We chose to focus on adults in all three cases because significant changes in plastron shape occur over the course of ontogeny in \textit{E.
 marmorata} and other emydines \citep{Angielczyk2013a}.
@@
 
-The first dataset is a compilation of 101 specimens of \textit{T.
+The first dataset we analyzed includes 578 total specimens from the following species: \textit{Clemmys guttata}, \textit{Emys blandigii}, \textit{Emys orbicularis}, \textit{Glyptemys insculpta}, \textit{Glyptemys muhlenbergii}, \textit{Terrapene coahuila}, and \textit{Terrapene ornata}.
+These specimens are a subset of those used in \citet{Angielczyk2011} and \citet{Angielczyk2013a}.
+
+The second dataset is a compilation of 101 specimens of two subspecies of \textit{T.
 scripta}: 51 specimens of \textit{T.
 scripta scripta} and 50 specimens of \textit{T.
-scripta scripta}.
+scripta elegans}.
 These landmark data are new to this study.
 
-The second dataset, we analyzed 578 total specimens from the following species: \textit{Emys blandigii}, \textit{Terrapene coahuila}, \textit{Clemmys guttata}, \textit{Glyptemys insculpta}, \textit{Glyptemys muhlenbergii}, \textit{Emys orbicularis}, and \textit{Terrapene ornata}.
-Like the first data set, these specimens are a subset of those used in \citet{Angielczyk2011} and \citet{Angielczyk2013a}.
-
 The final dataset dataset included 354 adult \textit{E.
-marmorata} museum specimens; a subset of those included in \citet{Angielczyk2007}, \citet{Angielczyk2011}, and \citet{Angielczyk2013a}.
+marmorata} museum specimens; representing a subset of those included in \citet{Angielczyk2007}, \citet{Angielczyk2011}, and \citet{Angielczyk2013a}.
 We assigned a classification to each specimen for the different binning schemes based on geographic occurrence data recorded in museum collection archives.
@@
 This changes the classification of certain individuals near the boundaries between groups, providing a test of the robustness of the classification schemes.
-Sex information was only know for a subset of the total dataset and was not included as a predictor of classification.
-Sex information was used to determine if observations cluster by sex or not.
+Sex information was known only for a subset of the total dataset and was not included as a predictor of classification.
+However, we did use sex information to determine if observations cluster by sex or not.
 The scheme names are as follows: Mito 1 and 2 correspond to \citet{Spinks2005}, Mito 3 corresponds to \citet{Spinks2010}, Morph 1 and Morph 2 correspond to \citet{Holland1992}, and Nuclear corresponds to \citet{Spinks2014}.
+% CHANGE
 
@@
 
-Following previous work on plastron shape \citep{Angielczyk2007,Angielczyk2011,Angielczyk2013a}, we used TpsDig 2.04 \citep{Rohlf2005} to digitize 19 landmarks (Fig.
+Following previous work on plastron shape \citep{Angielczyk2007,Angielczyk2011,Angielczyk2013a}, we used TpsDig 2.04 \citep{Rohlf2005} to digitize 19 two-dimensional landmarks (Fig.
 \ref{fig:plastra}).
@@
 % digitization
-We estimated the possible effect of digitization error CITATIONS on our results by comparing within (replicated) specimen Procrustes distances to the distances between classification scheme centroids.
-10 randomly selected specimen both for this study and an additional four times.
+We estimated the possible effect of digitization error \citep{Arnqvist1998,Cramon2007,Munoz-MunozF.2010} on our results by comparing within (replicated) specimen Procrustes distances to the distances between classification scheme centroids.
+Ten randomly selected \textit{E.
+marmorata} specimens were each digitized four times, with the original set of digitized coordinates serving as a fifth replicate.
 These 50 landmark configurations were then Procrustes superimposed.
-A range of four Procrustes distances were then calculated as the average of the pairwise distances between each of the replicate configurations.
+A range of four Procrustes distances were then calculated as the average of the pairwise distances between each of the replicate configurations of a given specimen.
 
-The relative magnitude of digitization error was calculated as the ratio between the average of the within-species replicate distances and the average distance between any two configurations in the replicate dataset.
+For each specimen, the difference in shape caused by digitization was calculated as the mean of all pairwise Procrustes distances between the five replicates of that specimen.
+The average distance between any two digitizations was calculated as the mean of all pairwise Procrustes distances between all replicates for all specimens.
+The ratio between these two values was used to assess the magnitude of variation caused by digitization.
 The goal of this ratio is to determine if the within group distances are on average smaller than the between individual distances; a value of 0 indicates perfect grouping, a value of 1 indicates no difference between grouping and no grouping, and a value of 1+ indicates that the grouping is counter-intuitive to the data.
 
-Turtles are well known to demonstrate strong sexual dimorphism in carapace shape CITATION.
-To test for issues surrounding sexual dimorphism in our \textit{E.
+Turtles are well known to demonstrate strong sexual dimorphism in plastron shape and morphology (e.g.
+the presence of a plastral concavity) CITATION.
+To test for biases resulting from sexual dimorphism in our \textit{E.
 marmorata} data set, we used a simple permutation test to determine if the distance between the mean female and male shapes is greater than expected when the sex labels are randomly shuffled.
@@
 \subsection{Supervised learning approaches}
+Instead of relying on a single supervised learning method, we chose to use an ensemble of multiple approaches so that the congruence between the them could be used as a means of ``support'' for one conclusion or another.
+The supervised learning methods used here are described in Table \ref{tab:methods}.
+Each of these methods makes different assumptions, treat data differently, and can produce different qualities of classification results depending on the nature of the data \citep{Hastie2009}.
+% use multiple methods
+%   multinomial logistic regression (nnet)
+%   linear discriminate analysis (MASS)
+%   penalized discriminate analysis (mda)
+%   neural network (nnet)
+%   random forest (randomForest)
+% each method has different assumptions and treat data differently
+%   all assume that predictors have additive effect (e.g. independent)
+
+\begin{table}
+  \centering
+  \caption{Table of the supervised learning methods used in this analysis.}
+  \begin{tabular}{l l l l}
+    \hline
+    Method name & abbreviation & R package & citation \\
+    \hline
+    multinomial logistic regression & MLR & nnet & \citet{Venables2002} \\
+    linear discriminate analysis & LDA & MASS & \citet{Venables2002} \\
+    penalized discrminiate analysis & PDA & mda & \citet{mdapack} \\
+    single-hidden-layer neural network & NN & nnet & \citet{Venables2002} \\
+    random forests & RF & randomForest & \citet{Liaw2002} \\
+    \hline
+  \end{tabular}
+  \label{tab:methods}
+\end{table}
+
 % predictors/features/covariates
@@
 % analogy to PCA regression
-The maximum set of possible predictors or features used for any model are the first 25 principal components (PCs), scaled centroid size, and the interaction between scaled centroid size and PC 1.
-Size and the interaction between size and PC 1 were included as predictors in order to account for a possible interaction between size and shape over the duration of an individual as well as potential size differences between classes, even if this is unlikely \citep{Seeliger1945,Holland1992}.
-We say ``maximum set'' because the best or selected models based on 5-fold cross-validation does need not to, nor will they likely, include all predictors possible (see below).
+The maximum set of possible predictors or features used for any model of our dataset is comprised of the first 25 principal components (PCs), scaled centroid size, and the interaction between scaled centroid size and PC 1.
+Additional interaction terms were not considered because of model complexity/sample size concerns.
+Size and the interaction between size and PC 1 were included as predictors t account for known ontogenetic variation in plastron shape \citep{Angieczyk2013a} as well as potential size differences between classes, even if this is unlikely \citep{Seeliger1945,Holland1992}.
+These data constitute a ``maximum set'' because the best or selected models based on five-fold cross-validation need not, and likely will not, include all predictors possible (see below).
 
-This approach is in many ways analogous to PCA regression.
-PCA regression takes advantage of two aspects of PCA for improving regression fit \citep{Hastie2009}.
-Because the PCs of shape are by definition orthogonal, allowing them to easily as independent predictors or features of class membership without fear of collinearity.
-
-% auc of roc as performance measure
-%   relationship between false positive rate and true positive rate
-%   varies between 0.5 and 1 (random -- perfect)
-%   better performance than accuracy when groups unbalanced
-In classification studies, such as this one, a common metric of performance is the receiver operating characteristic (ROC) which is the relationship between the false and true positive rates \citep{Hastie2009}.
-The area under the ROC curve (AUC) is then the derived estimate of the model performance; AUC ranges from 0.5 to 1 which corresponding to performance similar to random guesses and perfect classification rates, respectively \citep{Hastie2009}.
-Both ROC and AUC are preferable to simple classification accuracy when class membership is unbalanced, as it is in these analyses \citep{Hastie2009}.
-The standard ROC and AUC calculations are defined only for binary classifications, which is not the case for our seven species and \textit{Emys} complex datasets.
-To generalize this approach for situations with multiple response classes, we used an all-against-one strategy where the model AUC is the average of the AUC values from the multiple binary comparisons of one class compared to all others \citep{Hand2001}.
+Because our supervised learning models use PCs as predictors, this approach is in many ways analogous to PCA regression.
+PCA regression takes advantage of dimensionality reduction and orthogonal PCs aspects of PCA to improve regression fit \citep{Hastie2009}.
+Because the PCs of shape are by definition orthogonal, they can easily serve as independent predictors or features of class membership without fear of collinearity.
 
@@
 
-For a given supervised learning method, we compared the fit of 27 models as the average AUC from 10 rounds of 5-fold cross-validation.
-Cross-validation is an approach for estimating the average out-of-sample predictive error of a model by simulating out-of-sample data from the training data itself \citep{Hastie2009}.
-In a single round of \(k\)-fold cross-validation, the training data is divided into \(k\) blocks where the model is fit to \(k - 1\) blocks and the values of the \(k\)th block are predicted; this is then repeated for all combinations of blocks.
-Within each round, the predictive performance metrics is averaged across all folds.
+% auc of roc as performance measure
+%   relationship between false positive rate and true positive rate
+%   varies between 0.5 and 1 (random -- perfect)
+%   better performance than accuracy when groups unbalanced
+In classification studies, such as this one, a common metric of performance is the receiver operating characteristic (ROC) which is the relationship between the false and true positive rates \citep{Hastie2009}.
+The area under the ROC curve (AUC) is the derived estimate of the model performance; AUC ranges from 0.5 to 1 which correspond to performance similar to random guesses and perfect classification rates, respectively \citep{Hastie2009}.
+Both ROC and AUC are preferable to simple classification accuracy when class membership is unbalanced, as it is in these analyses \citep{Hastie2009}.
+The standard ROC and AUC calculations are defined only for binary classifications, which is not the case for our seven species and \textit{Emys} complex datasets.
+To generalize this approach for situations with multiple response classes, we used an all-against-one strategy where the model AUC is the average of the AUC values from the multiple binary comparisons of one class compared to all others \citep{Hand2001}.
+
+For a given supervised learning method, we compared the fit of 27 models as the average AUC from 10 rounds of five-fold cross-validation.
+Cross-validation is an approach for estimating the average out-of-sample predictive error of a model by simulating out-of-sample data from the training dataset itself \citep{Hastie2009}.
+In a single round of \(k\)-fold cross-validation, the training data are divided into \(k\) blocks where the model is fit to \(k - 1\) blocks and the values of the \(k\)th block are predicted.
+This is repeated for all combinations of blocks.
+Within each round, the predictive performance metrics are averaged across all folds.
 Finally, the predictive performance metric is the averaged across all rounds of \(k\)-fold cross-validation.
@@
 
-For a given supervised learning method, the ``best'' trained model is that the highest mean AUC as estimated from 5-fold cross-validation.
+For a given supervised learning method, the ``best'' trained model is that with the highest mean AUC as estimated from five-fold cross-validation.
 The selected or final model, however, is the next most parsimonious model that is within one standard error of the best model; this is a variant on the ``one-standard error'' rule from \citet{Hastie2009}.
-
-% use multiple methods
-%   multinomial logistic regression (nnet)
-%   linear discriminate analysis (MASS)
-%   penalized discriminate analysis (mda)
-%   neural network (nnet)
-%   random forest (randomForest)
-% each method has different assumptions and treat data differently
-%   all assume that predictors have additive effect (e.g. independent)
-Instead of relying on a single supervised learning method, we chose to use an ensemble of multiple approaches so that the congruence between the them could be used as a means of ``support'' for one conclusion or another.
-The supervised learning methods used here are described in Table \ref{tab:methods}.
-Each of these methods makes different assumptions, treat data differently, and can produce different qualities of classification results depending on the nature of the data \citep{Hastie2009}.
-
-\begin{table}
-  \centering
-  \caption{Table of the supervised learning methods used in this analysis.}
-  \begin{tabular}{l l l l}
-    \hline
-    Method name & abbreviation & R package & citation \\
-    \hline
-    multinomial logistic regression & MLR & nnet & \citet{Venables2002} \\
-    linear discriminate analysis & LDA & MASS & \citet{Venables2002} \\
-    penalized discrminiate analysis & PDA & mda & \citet{mdapack} \\
-    single-hidden-layer neural network & NN & nnet & \citet{Venables2002} \\
-    random forests & RF & randomForest & \citet{Liaw2002} \\
-    \hline
-  \end{tabular}
-  \label{tab:methods}
-\end{table}
+This rule is applied to ameliorate the chance of selecting an overly complex model.
 
@@
 The ratio of the average within-individual distance to the average distance between individuals for the replicated datasets is 1.11; this indicates that the grouping is slightly counter-intuitive to the data.
-This value provides a baseline by which to understand how distinct the groupings are, where other ratios are compared to the correction ratio \(1.11/1\).
+This value also provides a baseline by which to understand how distinct the groupings are, where other ratios are compared to the correction ratio \(1.11/1\).
 
@@
 In contrast, the different \textit{E.
-marmorata} classification schemes appear to be barely be distinct, with their within:between ratios being approximate 1 which indicates that the differences between groupings are approximately as different as any two random individuals are (Table \ref{tab:rep_res}).
+marmorata} classification schemes appear to be barely be distinct, with their within:between ratios approximating 1.
+This indicates that the magnitude of the differences between groupings is approximately the same as the difference as any two random individuals (Table \ref{tab:rep_res}).
 
@@
 
-Analysis of the seven morphologically-distinct species and the \textit{T.
+Analysis of the seven morphologically and genetically distinct species and the \textit{T.
 scripta scripta}--\textit{T.
-scripta elegans} datasets indicate that these classifications are sufficiently morphologically distinct.
-Both in-sample and out-of-sample classification have AUC values of approximately 1 for all methods (Fig.
+scripta elegans} datasets indicate that these classifications are sufficiently morphologically distinct to be differentiated on the basis of plastron shape.
+Both in-sample and out-of-sample classification have AUC values of approximately 1 for all methods, implying near-perfect classification rates (Fig.
 \ref{fig:other_sel}, \ref{fig:other_oos}).
@@
 \ref{fig:other_oos}).
-These results demonstrate that when there are distinctions between the states of the classification schemes, the methods used here can recover them.
+These results demonstrate that when there are distinctions between the states of the classification schemes (i.e.
+differences in plastron shape that correlate with the different taxonomic groups), the methods used here can recover them.
 
@@
 \ref{fig:emys_sel}).
-As was the case with the \textit{Trachemys} dataset (Fig.
-\ref{fig:other_sel}), the neural net models are the least well-performing models compared to all others used (Fig.
-\ref{fig:emys_sel}).
 
@@
 
-Observed AUC values for all of the optimal models are not exceptionally high (Fig.
+Observed AUC values for all of the optimal models are not exceptionally high for the \textit{E.
+marmorata} dataset(Fig.
 \ref{fig:emys_sel}).
 In most cases the different proposed classification schemes are generally poor descriptors of the observed variation.
-It appears that the data set is overwhelmed by noise, making any accurate classifications difficult at best.
+It appears that the data set is overwhelmed by noise (likely biological and analytical), making any accurate classifications difficult at best.
 This observation is cemented with the generalizations of the models to the testing data set (Fig.
@@
 
-The results of this study show that there is no clear grouping of \textit{E. marmorata} based on plastron shape.
+The results of this study show that there is no clear grouping scheme for \textit{E.
+marmorata} specimens based on plastron shape.
+Both the low out-of-sample AUC values and the significant difference between the correctly and incorrectly classified observations support the conclusion that none of the hypothesized classification schemes are good descriptions of the observed plastral variation within \textit{E.
+marmorata}.
 
-The results from fitting the various supervised learning models to each of the classification schemes generally shows that no one scheme is ``best.'' Possible explanations include that the genetic differentiation is not associated with plastron shape variation and/or that local selective pressures (e.g.
+The results from fitting the various supervised learning models to each of the classification schemes generally shows that no one scheme is ``best.'' An analytical explanation of this result is that the level of digitization error in the \textit{E.
+marmorata} dataset is so great as to swamp out any biological signal.
+We think this is unlikely because all of the specimens considered in our three analyses were digitized by one of us (K.D.A.), and digitization error was not a problem in the seven species or \textit{Trachemys} examples.
+There are also no features of the plastron of \textit{E.
+marmorata} that would make it significantly more difficult to accurately digitize than the plastra of the other speices.
+
+Biological explanations include the possibilty that genetic differentiation is not associated with plastron shape variation and/or that local selective pressures (e.g.
 from hydrological regime) overwhelm morphological differentiation.
-This makes sense given that shell shape is influenced by selection for both protection and streamlining, but not necessary mate choice \citep{Rivera2008,Rivera2011,Stayton2011,Rivera2014,Polly2016} and that shell shape in \textit{E.
+Both of these options seem plausible given that shell shape is influenced by selection for both protection and streamlining, but not necessary mate choice \citep{Rivera2008,Rivera2011,Stayton2011,Rivera2014,Polly2016} and that shell shape in \textit{E.
 marmorata} is known to vary among populations inhabiting water bodies with different flow regimes \citep{Holland1992,Lubcke2007,Germano2009}.
@@
 
-Both the low out-of-sample AUC values (\(< 0.9\)) and the significant difference between the correctly and incorrectly classified observations support the conclusion that none of the hypothesized classification schemes are good descriptors of the observed plastral variation within \textit{E.
-marmorata}.
-
-Nevertheless, it is important to note that plastron shape is an extremely effective method for differentiating classes in the additional datasets we investigated.
-The magnitude of shape differences between the species (measured as Procrustes distance between species' mean shapes) is approximately an order of magnitude greater than the differences between the \textit{E.
-marmorata} subgroups, and the machine learning methods had no trouble accurately classifying the specimens correctly.
+Despite the negative result for \textit{E.
+marmorata}, it is important to note that plastron shape is an extremely effective method for differentiating classes in the additional datasets we investigated.
+The magnitude of shape differences between the species (measured as Procrustes distance between the seven species' mean shapes) is approximately an order of magnitude greater than the differences between the \textit{E.
+marmorata} subgroups, and not surprisingly the machine learning methods had no trouble classifying the specimens correctly.
 However, the magnitude of the shape differences between the \textit{T.
-scripta} subspecies is considered comparable to those separating the different \textit{E.
+scripta} subspecies is comparable to those separating the different \textit{E.
 marmorata} subgroups, yet even in this case the machine learning methods returned an almost perfect classification.
@@
 marmorata} is not simply a shortcoming of the methods we applied.
-Indeed, they beg the question of what factors have suppressed morphological differentiation of plastron shape in \textit{E.
+Indeed, it begs the question of what factors have suppressed morphological differentiation of plastron shape in \textit{E.
 marmorata} and \textit{E.
@@
 Invoking issues such as the role of the plastron in protection or the need for streamlining are insufficient because the other species are expected to be subject to similar constraints \citep{Stayton2011,Polly2016}.
-While it may seem counterintuitive that plastron shape is both useful for species delimitation but has weak or absent phylogenetic signal, it is important to note that these are different goals.
-While phylogenetically similar species may not be morphologically similar, the variation within a species is much less than the variation between species which leads to plastron shape being a useful tool for species delimitation but not phylogenetic reconstruction.
+Although it may seem counterintuitive that plastron shape is both useful for species delimitation but has weak or absent phylogenetic signal, it is important to remember that these are different goals.
+While phylogenetically similar species may not be morphologically similar (e.g.
+compare the box turtles of the genus \textit{Terrapene} to the closely related spotted turtle \textit{Clemmys guttata}), the variation within a species typically is much less than the variation between species.
+Therefore, the consistent plastron shapes that characterize different emydid species leads to plastron shape being a useful tool for species delimination, even when other selective factors have overprinted similarities stemming from patterns of descent from common ancestors.
 
@@
 pallida} does not, on its own, preclude the recognition of this taxon.
-However, this apparent lack of congruence does prompt a reexamination of the methods and concepts that led to that taxonomic revision, especially considering that plastron shape is demonstrably capable of differentiating species and subspecies among emydids.
+However, this apparent lack of congruence does prompt a reexamination of the methods and concepts that led to that taxonomic revision, especially considering that plastron shape is demonstrably capable of differentiating species and subspecies among other emydids.
 In other words, before we can assess the significance of the morphological non-diagnosablity, it is essential to evaluate the methods and concepts that led to the initial taxonomic revision.
@@
 pallida} based on a species delimitation analysis of SNP data using BPP \citep{Yang2010b}.
-However, \citet{Spinks2014} did not heed the caveats about species delimitation methods raised by \citet{Carstens2013}.
+However, \citet{Spinks2014} did not heed the caveats about such species delimitation methods raised by \citet{Carstens2013}.
 In addition to specifically addressing the shortcomings of validation methods such as BPP that rely on guide trees and ``should be interpreted with caution,'' \citet{Carstens2013} also strongly emphasized that ``Inferences regarding species boundaries based on genetic data alone are likely inadequate, and species delimitation should be conducted with consideration of the life history, geographical distribution, morphology and behaviour (where applicable) of the focal system\dots'' These caveats evoke the development of the Unified Species Concept \citep{Dayrat2005a,DeQueiroz2007b}, Integrative Taxonomy \citep{Padial2010}, and other pluralist approaches to species delimitation.
@@
 marmorata} (\textit{sensu lato}) is a species deserving of strong conservation efforts, and we do not wish to trivialize this need.
-Moreover, the genetic diversity uncovered by the analysis of \citet{Spinks2014} should be explicitly accounted for in any conservation plan.
+Moreover, the genetic diversity uncovered by the analysis of \citet{Spinks2014} should be accounted for explicitly in any conservation plan.
 Given the apparent lack of morphological distinction, however, we consider that this diversity should be considered Evolutionary Significant Units or Distinct Population Segments instead of distinct species.
@@
 marmorata} truly reflects the presence of a single species, or if it is an artifact of plastron shape being a poor morphological marker for phylogenetic and phylogeographic divergences in the case of \textit{E.
-marmorata}, because we could not carry out our morphometric analyses on the specimens from which the genetic data were obtained.
-The comparisons with the other emydid taxa suggest that it is because \textit{E.
+marmorata}.
+This is because we could not carry out our morphometric analyses on the specimens from which the genetic data were obtained.
+The comparisons with the other emydid taxa suggest that our negative result is is because \textit{E.
 marmorata} is a single species.
 However, tests of both our preferred conclusion (\textit{E.
-marmorata} as a single species) and that \citet{Spinks2014} should include morphological and molecular analyses of the same set of voucher specimens, as well as additional tests of species delimitation using alternative methods and corroborating evidence as suggested by \citet{Carstens2013}.
+marmorata} as a single species) and that of \citet{Spinks2014} should include morphological and molecular analyses of the same set of voucher specimens, as well as additional tests of species delimitation using alternative methods and corroborating evidence as suggested by \citet{Carstens2013}.
 From a morphological standpoint, support for the validity of ``\textit{E.

### 2017-07-10 R/supervised_mung.r

@@ -24,36 +24,41 @@ land <- df2array(land, n.land, n.dim)
 
 # meta data
 #meta <- read.csv('../data/marmorata_meta2.csv')
-meta <- read.csv('../data/turtle_update.csv')
+meta <- read.csv('../data/marm_upbins.csv')
 meta[meta == "?"] <- NA
 meta[meta == "<NA>"] <- NA
+#names(meta) <- c('ind', 'spec', 'lat', 'long', 'year',
+#                 'sh1', 'sh2', 'sh3', 'sh4', 'sh5', 'spinks.pre', 'spinks',
+#                 'p.sex', 'b.sex', 'sex.pre', 'notes')
 names(meta) <- c('ind', 'spec', 'lat', 'long', 'year',
-                 'sh1', 'sh2', 'sh3', 'sh4', 'sh5', 'spinks.pre', 'spinks',
+                 'sp10.1', 'sp10.2', 'sp10.3', 'sp14.1', 'sp14.2', 'morph', 'ndu',
                  'p.sex', 'b.sex', 'sex.pre', 'notes')
-meta$sh4 <- as.factor(meta$sh4)
-meta$spinks <- as.factor(meta$spinks)
 
+# get rid of everything that doesn't have lat-long
+meta[, c('lat', 'long')] <- apply(meta[, c('lat', 'long')], 
+                                  2, function(x) as.numeric(as.character(x)))
+meta$long <- meta$long * -1
+land <- land[, , !is.na(meta$lat)]
+rawturt<- rawturt[!is.na(meta$lat), ]
+meta <- meta[!is.na(meta$lat), ]
+
+# get rid of everything that doesn't have assigns
 # clean off the missing important values
-imp <- c('lat', 'long', 'sh1', 'sh2', 'sh3', 'sh4', 'sh5', 'spinks', 'p.sex')
+imp <- c('lat', 'long', 'sp10.1', 'sp10.2', 'sp10.3', 'sp14.1', 'sp14.2', 'morph', 'p.sex')
 rms <- Reduce('|', alply(meta[, imp], 2, is.na))
 land <- land[, , !rms]
 meta <- meta[!rms, ]
 rawturt <- rawturt[!rms, ]
 
-# clean the geographic information
-meta[, c('lat', 'long')] <- apply(meta[, c('lat', 'long')], 
[…]

### 2017-09-05 doc/turtles_statistical_learning.tex
updated 7cc data. stuart and ken comments on manuscript.

@@
 %\title{How cryptic is cryptic diversity? Machine learning approaches to classifying morphological variation in the Pacific Pond Turtle (\textit{Emys marmorata})}
-\title{Supervised learning approaches to classifying morphological variation: assessing taxonomic hypotheses for the Pacific Pond Turtle (\textit{Emys marmorata})}
-\author[1]{Peter D Smits}%\thanks{psmits@uchicago.edu}}
-\author[1,2]{Kenneth D Angielczyk}%\thanks{kangielczyk@fieldmuseum.org}}
-\author[3]{James F Parham}%\thanks{jparham@fullerton.edu}}
-\author[4]{Bryan Stuart}%\thanks{bryan.stuart@naturalsciences.org}}
-\affil[1]{Committee on Evolutionary Biology, University of Chicago}
-\affil[2]{Integrative Research Center, Field Museum of Natural History}
-\affil[3]{Department of Geological Sciences, California State University -- Fullerton}
-\affil[4]{Section of Research and Collections, North Carolina Museum of Sciences}
+\title{Ensemble approaches for estimating congruence between species delimitation and morphological variation: comparing taxonomic hypotheses for the Pacific Pond Turtle (\textit{Emys marmorata})}
+\author[1]{Peter D Smits}%\thanks{peterdavidsmits@gmail.com}}
+\author[2,3]{Kenneth D Angielczyk}%\thanks{kangielczyk@fieldmuseum.org}}
+\author[4]{James F Parham}%\thanks{jparham@fullerton.edu}}
+\author[5]{Bryan L Stuart}%\thanks{bryan.stuart@naturalsciences.org}}
+\affil[1]{Department of Integrative Biology, University of California -- Berkeley}
+\affil[2]{Committee on Evolutionary Biology, University of Chicago}
+\affil[3]{Integrative Research Center, Field Museum of Natural History}
+\affil[4]{Department of Geological Sciences, California State University -- Fullerton}
+\affil[5]{Section of Research and Collections, North Carolina Museum of Natural Sciences}
 
@@
 \maketitle
-\noindent{\textbf{Corresponding author:} Peter D Smits, Committee on Evolutionary Biology, University of Chicago, 1025 E.
-57th Street, Culver Hall 402, Chicago, IL, 60637, USA; E-mail: \href{mailto:psmits@uchicago.edu}{psmits@uchicago.edu}}
+\noindent{\textbf{Corresponding author:} Peter D Smits, Department of Integrative Biology, University of California -- Berkeley, 3040 Valley Life Sciences Building \#5151, Berkeley, CA, 94720, USA; E-mail: \href{mailto:peterdavidsmits@gmail.com}{peterdavidsmits@gmail.com}}
 
@@
 \begin{abstract}
-  We investigate morphometric identification of cryptic species using machine learning approaches by examining their implications for a recently proposed cryptic turtle species (\textit{Emys pallida}).
-We collected landmark-based morphometric data from 354 adult \textit{E.
-marmorata/``pallida''} museum specimens.
-We assigned a classification to each specimen for six different binning schemes using geographic occurrence data recorded in museum collection archives.
-We used multiple supervised machine learning methods to compare different classification hypotheses and asked whether it is possible to determine which among this set of classification hypotheses is best.
-  In addition, we applied the above approach to two other datasets: two well-supported subspecies of \textit{Trachemys scripta} and a set of seven unambiguously distinct species closely related to \textit{E.
-marmorata}.
-The results of this study indicate that there is no clear grouping of \textit{E.
-marmorata/``pallida''} based on plastron shape.
-In contrast, analysis of the other datasets resulted in near perfect classifications, which demonstrates that application of the methods to plastron shape data can recover correct results in similar situations.
+%  We investigated the morphometric identification of cryptic species using machine learning approaches by examining their implications for a recently proposed cryptic turtle species (\textit{Emys pallida}).
+We collected landmark-based morphometric data from 366 adult \textit{E.
+marmorata/``E.
+pallida"} museum specimens.
+We assigned a classification to each specimen for six different binning schemes based on geographic occurrence data recorded in museum collection archives.
+We used an ensemble of supervised machine learning approaches to determine which classification hypothesis was best supported by the data.
+In addition, we applied the same approach to two clear-cut examples, one consisting of seven unambiguously distinct species closely related to \textit{E.
+marmorata}, and the other consisting of two subspecies of \textit{Trachemys scripta}.
+Our results indicate that there is no clear grouping of \textit{E.
+marmorata/``E.
+pallida"} based on plastron shape.
+In contrast, the analyses of the clear-cut examples produced near perfect classifications, demonstrating that the methods can recover correct results when an appropriate signal exists.
 Explanations for the lack of grouping in \textit{E.
-marmorata} include that possibility that genetic differentiation is not associated with plastron shape variation in this species complex and/or that local selective pressures (e.g., from hydrological regime) overwhelm morphological differentiation.
-A reconsideration of the methods used to delimit \textit{E.
-``pallida,''} the lack of barriers to gene flow, the strong evidence for widespread admixture between lineages, and the fact that plastron shape can be used to differentiate other emydid species and subspecies suggest that the non-diagnosability of \textit{E.
-``pallida''} stems from it not being a distinct species.
+marmorata} include the possibility that genetic differentiation is not associated with plastron shape variation below the species level and/or that local selective pressures (e.g., from hydrological regime) overwhelm morphological differentiation.
+A reconsideration of the methods used to delimit \textit{``E.
+pallida"}, the lack of barriers to gene flow, the strong evidence for widespread admixture between lineages, and the fact that plastron shape can be used to delineate other emydine species and sub-species suggest that its lack of diagnosability most likely reflects the non-distinctiveness of this proposed taxon.
 \end{abstract}
@@
 In contrast to those genetically-diagnosed species, the majority of extant taxa, and almost all extinct taxa, are delimited by morphology alone.
-This disjunction complicates interpretations of variation and diversity in deep time, as apparent morphological stasis may not reflect the true underlying diversity \citep{Eldredge1972,Gould1977a,Hunt2008,VanBocxlaer2013}.
+This disjunction complicates interpretations of variation and diversity in deep time, as apparent morphological stasis may not reflect the true underlying diversity \citep{Eldredge1972,Gould1977a,VanBocxlaer2013}.
 It also has serious implications for our records of modern biodiversity: for many museum specimens of extant taxa (e.g.
@@
 This would make the task of identifying and maintaining endangered or conserved groups much easier and could contribute to improved classifications of extinct taxa and populations.
-Most such studies focus on using morphometrics to discover differences between taxa that were identified by other means \citep{Polly2003,Zelditch2004,Gaubert2005b,Gunduz2007,Polly2007a,Demandt2009}.
-Additionally, there has been work on automated taxon identification and classification of taxa into groups \citep{Baylac2003,Dobigny2003,MacLeod2007,VandenBrink2011}.
+Most such studies focus on using morphometrics to discover differences between taxa that were identified by other means \citep{Polly2003,Zelditch2004,Gaubert2005b,Gunduz2007,Polly2007a,Demandt2009,Markolf2013,Fruciano2016}.
+Additionally, there has been work on automated taxon identification and classification of taxa into groups \citep{Baylac2003,Dobigny2003,MacLeod2007,VandenBrink2011,Vitek2017}.
 
 Here, we investigate the morphometric identification of cryptic species using machine learning approaches.
-In particular, we ask whether it is possible to determine which among a set of classification hypotheses is best and examine the implications of the results for a recently proposed set of cryptic turtle species.
+We use an ensemble learning approach where multiple methods are used in order to look for consensus amoung their results.
+We test our approach on three datasets: plastron shape of seven species of closely related turtles, plastron shape of two species of closely related turtles, and plastron shape of the \textit{Emys marmorata} species complex.
+In particular, we ask whether it is possible to determine which among a set of classification hypotheses best aligns with the observed morphology, and examine the implications of our results for the \textit{E.
+marmorata} complex.
 
@@
 
-Geometric morphometric approaches to identifying differences in morphological variation between different classes, including cryptic species, mostly have used methods like linear discriminate analysis and canonical variates analysis \citep{Polly2003,Zelditch2004,Gaubert2005b,Gunduz2007,Polly2007a,Francoy2009,Sztencel-Jabonka2009,MitrovskiBogdanovic2013}.
+Geometric morphometric approaches to identifying differences in morphological variation between different classes, including cryptic species, mostly have used methods like linear discriminate analysis and canonical variates analysis \citep{Polly2003,Zelditch2004,Gaubert2005b,Gunduz2007,Polly2007a,Francoy2009,Sztencel-Jabonka2009,MitrovskiBogdanovic2013,Dillard2017}.
 Because of their similarity to multivariate approaches like principal components analysis (PCA), these methods are comparatively straightforward ways of understanding the differences in morphology between classes.
@@
 
-Here, we used multiple supervised machine learning methods to compare different classification hypotheses.
+Here, we used an ensemble of supervised machine learning methods to compare the congruence of the morphological data to different classification hypotheses.
 Each of these methods provide different advantages for understanding how to classify taxa, as well as the accuracy of the resulting classifications.
-Although machine learning methods such as neural networks have been applied to studying shape variation \citep{Baylac2003,Dobigny2003,MacLeod2007,VandenBrink2011}, including in the context of automated taxon identification and classification of groups, the number of cases remains limited.
+Machine learning methods have been combined with geometric morphometric data to study shape variation in a variety of contexts, including automated taxon identification and classificaiton of groups \citep{Baylac2003,Dobigny2003,MacLeod2007,VandenBrink2011}.
+% MORE
+%Although machine learning methods such as neural networks have been applied to studying shape variation, including in the context of automated taxon identification and classification of groups, the number of cases remains limited.
 In the current study, we not only consider pure classification accuracy but also use a statistic of classification strength that reflects the rate at which taxa are both accurately and inaccurately classified.
@@
 Analyses with nuclear DNA \citep{Spinks2010} and single-nucleotide polymorphism (SNP) data suggest a primarily north--south division in \textit{E.
-marmorata}, although these datasets differed from that of \citet{Spinks2005} in the location of the break point.
+marmorata}, although these datasets differed from that of mitocondrial-based results of \citet{Spinks2005} in the location of the break point \citep{Spinks2014}.
 All three studies discussed the potential taxonomic implications of their results, with \citet{Spinks2014} going so far as to strongly advocate for the recognition of at least two species (\emph{E.
@@
 In this study, we attempt to estimate the best classification scheme of \textit{E.
-marmorata} based on variation in plastron (ventral shell) shape in order to determine whether this character is consistent with any of the past divisions based on other morphological features or molecular data.
-We are particularly interested in whether any aspect of plastron shape can be used to reliably diagnose the proposed species of \citet{Spinks2014}, and if so, the nature of that shape variation.
+marmorata} based on variation in plastron (ventral shell) shape in order to determine whether this character is consistent with any of the proposed taxonomies of the \textit{E.
+marmorata} complex.
 
-\uppercase{Why did we analyze plastron shape? Peter doesn't have a good answer.}
+We choose to analyze plastron shape for multiple reasons.
+First, it is very easy to collect geometric morphometric data on plastron shape from two-dimensional pictures as the structure is virtually flat.
+This approach allows both museum specimens and individuals in the field to be analyzed together.
+Second, previous work has suggested that there are strong differences in plastron shape among traditionally recognized emydine species \citep{Angielczyk2007,Angielczyk2011,Angielczyk2013a}.
+Finally, due to this previous study a large dataset was readily avaliable.
 
+In the case of the \textit{E.
+marmorata} species complex, we hypothesize that if one or more of the proposed classification schemes are consistent with the morphological data then our ensemble approach fit to those hypotheses will have higher out-of-sample predictive performance than the more inconsistent hypotheses.
+However, if all of the classification schemes lead to equal out-of-sample predictive performance then we would conclude that the molecular based hypotheses are inconsistent with whatever information is present in the morphological data.
 Because of unclear geographic boundaries between subgroups of \textit{E.
-marmorata}, we compare multiple hypotheses of morphologically-- and molecularly--based classification.
-We hypothesize that if morphological variation corresponds to class assignment, it should be possible to determine the best classification hypothesis of \textit{E.
-marmorata} from amongst multiple candidate hypotheses.
-However, if morphological variation does not correspond to any of the standing hypothesis, then supervised learning model generalization performance will be poor.
+marmorata}, we compare multiple permutations of the \citep{Spinks2010} and \citet{Spinks2014} hypotheses.
 
@@
 \subsection{Specimens, sampling, morphometrics}
-Three different landmark-based morphometric datasets describing plastron variation were assembled for this analysis: 1) specimens from seven distinct emydine species; 2) \textit{T.
+Three different geometric morphometric datasets describing turtle plastron variation were assembled for this analysis: 1) specimens from seven distinct emydine species; 2) \textit{T.
 scripta} specimens from the two main subspecies (\textit{T.
-scripta elgans} and \textit{T.
+scripta elegans} and \textit{T.
 scripta scripta}); and 3) \textit{E.
 marmorata} specimens from across the species' geographic range.
-The first two datasets are intended to serve as a basic test of whether machine learning techniques can differentiate species-level groupings of emydine turtles using plastron shape.
-Dataset 1 should be relatively easy, given the high level of plastron shape disparty that exists among these species \citep{Angieczyk2011}, whereas dataset 2 should be a case that is more comparable to our \textit{E.
-marmorata} dataset.
-We chose to focus on adults in all three cases because significant changes in plastron shape occur over the course of ontogeny in \textit{E.
+The first two datasets are intended to serve as a test of whether machine learning techniques can differentiate species-level groupings of emydine turtles using plastron shape.
+We expect that the first case represents a low complexity dataset because of the high level of plastron shape disparity that exists among these species \citep{Angielczyk2011}, whereas the second dataset should be relatively higher in complexity and more analogous to the \textit{E.
+marmorata} example.
+We predict that the \textit{E.
+marmorata} dataset should be of the highest complexity and our greatest challenge given \citep{Holland1992} finding that only very subtle differences existed between geographically-distinct populations.
+%We expect the first dataset represents a low-complexity dataset given the high level of plastron shape disparty that exists among these species \citep{Angielczyk2011}, whereas the second dataset should be relatively higher in complexity than the first.
+Ultimately, we expect that the third dataset to be of the highest complexity and our biggest analytical challenge.
+For all datasets we chose to focus on adults because significant changes in plastron shape occur over the course of ontogeny in \textit{E.
 marmorata} and other emydines \citep{Angielczyk2013a}.
@@
 
-The final dataset dataset included 354 adult \textit{E.
-marmorata} museum specimens; representing a subset of those included in \citet{Angielczyk2007}, \citet{Angielczyk2011}, and \citet{Angielczyk2013a}.
-We assigned a classification to each specimen for the different binning schemes based on geographic occurrence data recorded in museum collection archives.
-When precise latitude and longitude information were not available we estimated them from locality information.
+The final dataset is of 366 adult \textit{E.
+marmorata} museum specimens.
+These speciemens represent a subset of those included in \citet{Angielczyk2007}, \citet{Angielczyk2011}, and \citet{Angielczyk2013a}.
 Because \citet{Spinks2005}, \citet{Spinks2010}, and \citet{Spinks2014} did not use vouchered specimens we were not able to directly sample the individuals in their studies.
-Therefore our specimen classifications were based solely on the geographic information, not explicit assignment using molecular data.
-Because the exact barriers between different biogeographic regions are unknown and unclear, we represented some hypothesis with two schemes for a total of six different schemes.
-These schemes differed based on where geographic boundaries were assigned.
-This changes the classification of certain individuals near the boundaries between groups, providing a test of the robustness of the classification schemes.
-Sex information was known only for a subset of the total dataset and was not included as a predictor of classification.
-However, we did use sex information to determine if observations cluster by sex or not.
-The scheme names are as follows: Mito 1 and 2 correspond to \citet{Spinks2005}, Mito 3 corresponds to \citet{Spinks2010}, Morph 1 and Morph 2 correspond to \citet{Holland1992}, and Nuclear corresponds to \citet{Spinks2014}.
+Instead, our specimen classifications were based solely on the geographic information and not explicit assignment using molecular data.
+For each taxonomic hypothesis, specimens were assigned to one of the possible classes based on geographic occurrence data recorded in museum collection archives.
+In cases where precise latitude and longitude information were not available we estimated for them from other locality information.
+Because the exact barriers between different biogeographic regions are unknown and unclear, we represented each hypothesis with multiple possible realizations representing the classification uncertainty for specimens present at the geographic boundaries.
+The taxonomic hypotheses and sub-hypotheses for the \textit{E.
+marorata} used here are presented in Table \ref{tab:hypotheses}.
+
+Sex was known only for a subset of the total dataset and was not included as a predictor of classification.
+Instead, we estimated the degree by which specimens cluster morphologically by sex in order to determine how much of a potential biasing factor sexual dimorphism could be for our analysis of the \textit{E.
+marmorata} species complex (see below).
+%The scheme names are as follows: Mito 1 and 2 correspond to \citet{Spinks2005}, Mito 3 corresponds to \citet{Spinks2010}, Morph 1 and Morph 2 correspond to \citet{Holland1992}, and Nuclear corresponds to \citet{Spinks2014}.
 % CHANGE
@@
   \includegraphics[height = 0.5\textheight, width = \textwidth, keepaspectratio = true]{figure/plastra}
-  \caption{Depiction of general plastral shape of \textit{E. marmorata} and position of the 19 landmark used in this study. Anterior is towards the top of the figure.}
+  \caption{Depiction of general plastral shape of \textit{E. marmorata} and position of the 19 landmarks used in this study. Anterior is towards the top of the figure.}
   \label{fig:plastra}
 \end{figure}
+
+\begin{table}
+  \centering
+  \caption{Table of species delimitation hypotheses for \textit{E. marmorata}}
+  \begin{tabular}{l l l }
+    \hline
+    Abbreviation & Number of classes & citation \\
+    \hline
+    SP10.1 & 3 & \citet{Spinks2010} \\
+    SP10.2 & 3 & \citet{Spinks2010} \\
+    SP10.3 & 4 & \citet{Spinks2010} \\
+    SP14.1 & 2 & \citet{Spinks2014} \\
+    SP14.2 & 4 & \citet{Spinks2014} \\
+    \hline
+  \end{tabular}
+  \label{tab:hypotheses}
+\end{table}
 
@@
 We conducted all subsequent analyses on the resulting ``half'' plastra.
-We superimposed the plastral landmark configurations using generalized Procrustes analysis \citep{Dryden1998a}, after which, we calculated the principal components (PC) of shape using the \texttt{shapes} package for R \citep{R2016,Dryden2013}.
+We superimposed the plastral landmark configurations using generalized Procrustes analysis \citep{Dryden1998a}, after which we calculated the principal components (PC) of shape using the \texttt{shapes} package for R \citep{R2016,Dryden2013}.
 
@@
 These 50 landmark configurations were then Procrustes superimposed.
-A range of four Procrustes distances were then calculated as the average of the pairwise distances between each of the replicate configurations of a given specimen.
+A range of four Procrustes distances was then calculated as the average of the pairwise distances between each of the replicate configurations of a given specimen.
 
@@
 
-Turtles are well known to demonstrate strong sexual dimorphism in plastron shape and morphology (e.g.
-the presence of a plastral concavity) CITATION.
+\textit{Emys marmorata} is known to display sexual dimorhpism in plastron shape, particulary the presence of a plastra concavity in males \citep{Seeliger1945}.
 To test for biases resulting from sexual dimorphism in our \textit{E.
-marmorata} data set, we used a simple permutation test to determine if the distance between the mean female and male shapes is greater than expected when the sex labels are randomly shuffled.
-Because not all of our specimens have sex identifications associated with them, this analysis was done using a subset of the data.
+marmorata} dataset, we used a simple permutation test to determine if the distance between the mean female and male shapes is greater than expected when the sex labels are randomly shuffled.
+Because not all of our specimens have sex identifications associated with them, this analysis was done using a subset of the data (257 of 360).
 
 \subsection{Supervised learning approaches}
-Instead of relying on a single supervised learning method, we chose to use an ensemble of multiple approaches so that the congruence between the them could be used as a means of ``support'' for one conclusion or another.
-The supervised learning methods used here are described in Table \ref{tab:methods}.
-Each of these methods makes different assumptions, treat data differently, and can produce different qualities of classification results depending on the nature of the data \citep{Hastie2009}.
+Instead of relying on a single supervised learning method, we chose to use an ensemble approach where multiple model types are used in concert so that any congruence between the them increases our support for that conclusion over another \citep{Hastie2009}.
+The supervised learning methods used here are named in Table \ref{tab:methods}.
+Each of these methods makes different assumptions, treats data differently, and can produce different classification results depending on the nature of the data \citep{Hastie2009}.
+For example, multinomial logistic regression is a type of generalized linear model while random forest is an ensemble approach where multiple decision trees are fit to subsets of the full data and then averaged.
 % use multiple methods
@@
 Additional interaction terms were not considered because of model complexity/sample size concerns.
-Size and the interaction between size and PC 1 were included as predictors t account for known ontogenetic variation in plastron shape \citep{Angieczyk2013a} as well as potential size differences between classes, even if this is unlikely \citep{Seeliger1945,Holland1992}.
+Size and the interaction between size and PC 1 were included as predictors to account for known ontogenetic variation in plastron shape \citep{Angielczyk2013a} as well as potential size differences between classes, even if this is unlikely \citep{Seeliger1945,Holland1992}.
 These data constitute a ``maximum set'' because the best or selected models based on five-fold cross-validation need not, and likely will not, include all predictors possible (see below).
-
 Because our supervised learning models use PCs as predictors, this approach is in many ways analogous to PCA regression.
-PCA regression takes advantage of dimensionality reduction and orthogonal PCs aspects of PCA to improve regression fit \citep{Hastie2009}.
+PCA regression takes advantage of reduction and orthogonality PCs to improve regression fit \citep{Hastie2009}.
 Because the PCs of shape are by definition orthogonal, they can easily serve as independent predictors or features of class membership without fear of collinearity.
@@
 Within-sample model performance is inherently biased upwards, so model evaluation requires overcoming this bias.
-With very large sample sizes, as in this study, part of the sample can be used as the ``training set'' and the remainder acts as the ``testing set.'' In this approach, following all cleaning and vetting, the data is split into a training dataset and a testing dataset.
-The former is used for fitting the model where as the later is used for measuring model performance, a process called model generalization.
+With very large sample sizes, as in this study, part of the sample can be used as the ``training set'' and the remainder acts as the ``testing set.'' In this approach, following all cleaning and vetting, the data are split into a training dataset and a testing dataset.
+The former is used for fitting the model whereas the later is used for measuring model performance, a process called model generalization.
 In this analysis, we used 80\% of samples as the training set while the remaining 20\% were used as the testing set.
@@
 This process was implemented using the R package \texttt{caret} \citep{KuhnMAN2013}.
-
 For a given supervised learning method, the ``best'' trained model is that with the highest mean AUC as estimated from five-fold cross-validation.
@@
   \centering
-  \begin{subfigure}[b]{0.4\textwidth}
+  \begin{subfigure}[b]{0.7\textwidth}
     \caption{}
@@
   \end{subfigure}
-  \begin{subfigure}[b]{0.4\textwidth}
+
+  \begin{subfigure}[b]{0.7\textwidth}
     \caption{}
@@
   \end{subfigure}
-  \caption{Scatterplots of the first two axes from the PCA results for both the seven species (a) and \textit{Trachemys} (b) datasets.
-Point color corresponds to the species within each dataset.
-Point size is proportional to individual centroid size.
-In parentheses next to the axis labels are the percent of total variation accounted for by that axis.}
+  \caption{Two scatterplots of morphological differences from two of the three datasets analyzed in this study.
+(a) Scatterplot of the first two PCA axies from the landmarks from the seven different species dataset, and (b) the first two axes of variation from two subspecies of \textit{Trachemys} dataset.
+Point colors correspond to the categories within each dataset while point size is proportional to individual centroid size.
+In parentheses next to the axis labels are the percent of total variation accounted for by that axis.
+For both datasets there are clear distinctions between the different categories.}
   \label{fig:other_pca}
@@
 
-The results of the PCA of plastron shape in the \textit{Emys marmorata} dataset show no clear connection between plastron shape and any of the six proposed classification schemes (Fig.
+The results of the PCA of plastron shape in the \textit{Emys marmorata} dataset show no clear connection between plastron shape and any of the proposed classification schemes (Fig.
 \ref{fig:emys_pca}).
-The first PC axis of shape variation appears to be very structured by differences in individual centroid size (Fig.
+The first PC axis of shape variation appears to be primarily structured by differences in individual centroid size (Fig.
 \ref{fig:emys_pca}); this was the motivation for including centroid size and its interaction with PC1 as predictors in all of the supervised learning models.
@@
   \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/emys_pc_graph}
-  \caption{Plot of the first two axes of the PCA of the \textit{Emys marmorata} dataset.
-Each panel corresponds to each of the six different proposed classification schemes analyzed as part of this study.
-Point color corresponds to the classes within each scheme, with the titles corresponding to subspecies or regional codes.
-Point size is proportional to centroid size.}
+  \caption{Scatterplot of the first two axes of morphological variation in the \textit{Emys marmorata} dataset.
+Each panel corresponds to each of different proposed classification schemes analyzed as part of this study.
+Point color corresponds to the categories within each scheme, and the category names corresond to geographic regions.
+Point size is proportional to centroid size and the numbers in parantheses next to the axis labels are the percent of total variation accounted for by that axis.}
   \label{fig:emys_pca}
@@
 \ref{fig:sex_test}).
-The difference between the centroids of both sexes is very small; the sexes overlap much more than expected from a null distribution formed by permuting the sex-labels.
+The difference in mean shape between the sexes is very small; the sexes overlap slightly more than is expected from a null distribution based on permuting the sex-labels.
 
@@
 Comparison of the within to between Procrustes distances of the digitization replicates gives an approximate estimate of the error between distinct groupings (Table \ref{tab:rep_res}).
-The ratio of the average within-individual distance to the average distance between individuals for the replicated datasets is 1.11; this indicates that the grouping is slightly counter-intuitive to the data.
+The ratio of the average within-individual distance to the average distance between individuals for the replicated datasets is 1.11; this indicates that the grouping is slightly counter-intuitive to the data and is consistent with all shapes being very similar regardless of individual identity.
 This value also provides a baseline by which to understand how distinct the groupings are, where other ratios are compared to the correction ratio \(1.11/1\).
@@
   \centering
-  \caption{Results from the within-individual to between-individual Procrustes distances for the replicated shapes, the seven species dataset, the \textit{Trachemys} dataset, and each of the \textit{Emys marmorata} classification schemes.}
+  \caption{Results from the within-individual to between-individual Procrustes distances for the replicated plastron shape data.
+Results are presented for all three datasets analyzed here: the \textit{Trachemys} dataset, the seven species dataset, and each of the \textit{Emys marmorata} classification schemes.}
   \begin{tabular}{l l c c}
@@
     \hline
-    \textit{E. marmorata} & Mito 1 & 0.92 & 1.02 \\
-     & Mito 2 & 0.92 & 1.02 \\
-     & Mito 3 & 0.88 & 0.98 \\
-     & Morph 1 & 0.98 & 1.09 \\
-     & Morph 2 & 0.83 & 0.92 \\
-     & Nuclear & 0.90 & 1 \\
+    \textit{E. marmorata} & SP10.1 & 0.93 & 1.03 \\
+     & SP10.2 & 0.93 & 1.03 \\
+     & SP10.3 & 0.88 & 0.98 \\
+     & SP14.1 & 0.98 & 1.09 \\
+     & SP14.2 & 0.87 & 0.97 \\
     \hline
@@
   \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/other_model_sel}
-  \caption{Comparisons of model fit to the training dataset for each of the supervised learning methods; the results from the seven species dataset are presented in the left column, while those from the \textit{Trachemys} dataset are presented in the right column.
+  \caption{Comparisons of model fit to the training dataset for each of the supervised learning methods applied to the first two datasets; the results from the seven species dataset are presented in the left column, while those from the \textit{Trachemys} dataset are presented in the right column.
 Models were fit to datasets of varying complexity, with the number of parameters listed along the x-axis.
@@
   \caption{AUC values for models of varying complexity fit to the \textit{Emys marmorata} training datasets for each classification scheme.
+The x-axis corresponds to the total number of predictors included in each model, while the y-axis corresponds to the AUC value which is a measure of goodness of fit for classificaiton datasets.
+A model with a high AUC value corresponds to better classification performance than a model with a lower AUC value.
 Standard errors on AUC estimates are calculated from 10 rounds of 5-fold cross-validation.
@@
 
-Observed AUC values for all of the optimal models are not exceptionally high for the \textit{E.
-marmorata} dataset(Fig.
+Observed AUC values for all of the optimal models are lower for the \textit{E.
+marmorata} dataset (Fig.
 \ref{fig:emys_sel}).
 In most cases the different proposed classification schemes are generally poor descriptors of the observed variation.
-It appears that the data set is overwhelmed by noise (likely biological and analytical), making any accurate classifications difficult at best.
-This observation is cemented with the generalizations of the models to the testing data set (Fig.
+It appears that the dataset is overwhelmed by noise (likely biological and analytical), making any accurate classifications difficult at best.
+This observation is cemented with the generalizations of the models to the testing dataset (Fig.
 \ref{fig:emys_oos}).
@@
 
-Mean AUC values for the model generalizations, in most cases, are approximately equal to the observed AUC values from the training data set (Fig.
+Mean AUC values for the model generalizations, in most cases, are approximately equal to the observed AUC values from the training dataset (Fig.
 \ref{fig:emys_sel}, \ref{fig:emys_oos}).
@@
 
-The results of this study show that there is no clear grouping scheme for \textit{E.
-marmorata} specimens based on plastron shape.
+As expected, our ensemble approach yields high out-of-sample classification performance for the first two datasets.
+These results indicate that in cases of clear class seperation (Fig.
+\ref{fig:other_pca}) our approach is able to detect this and make good out-of-sample prediction.
+
+In the case of the \textit{E.
+marmorata} dataset, our results show that none of the proposed taxonomic hypotheses for the \textit{E.
+marmorata} species complex are more consistent with morphological differentiation than any other proposal (Fig.
+\ref{fig:emys_oos}).
 Both the low out-of-sample AUC values and the significant difference between the correctly and incorrectly classified observations support the conclusion that none of the hypothesized classification schemes are good descriptions of the observed plastral variation within \textit{E.
 marmorata}.
-
-The results from fitting the various supervised learning models to each of the classification schemes generally shows that no one scheme is ``best.'' An analytical explanation of this result is that the level of digitization error in the \textit{E.
+An analytical explanation of this result is that the level of digitization error in the \textit{E.
 marmorata} dataset is so great as to swamp out any biological signal.
@@
 marmorata} is known to vary among populations inhabiting water bodies with different flow regimes \citep{Holland1992,Lubcke2007,Germano2009}.
-Plastron shape does not seem to preserve a strong phylogenetic signal at the interspecific level in emydine turtles \citep{Angielczyk2011}, and our current results suggest that this may be the case for phylogeographic signal within emydine species as well.
+Plastron shape does not seem to preserve a strong phylogenetic signal at the interspecific level in emydine turtles, at least compared to the effect of the presence or absence of a plastral hinge \citep{Angielczyk2011}, and our current results suggest that this may be the case for phylogeographic signal within emydine species as well.
 A final possibility (explored below) is that the proposed classification schemes themselves do not represent significant evolutionary lineages.
@@
 compare the box turtles of the genus \textit{Terrapene} to the closely related spotted turtle \textit{Clemmys guttata}), the variation within a species typically is much less than the variation between species.
-Therefore, the consistent plastron shapes that characterize different emydid species leads to plastron shape being a useful tool for species delimination, even when other selective factors have overprinted similarities stemming from patterns of descent from common ancestors.
+Therefore, the consistent plastron shapes that characterize different emydid species leads to plastron shape being a useful tool for species delimitation, even when other selective factors have overprinted similarities stemming from patterns of descent from common ancestors.
 

### 2017-09-05 R/clear_cut.r

@@ -38,12 +38,31 @@ turt.out <- list()
 
 # actually start doing analysis...
 newturt <- list.files('../data/new_turtle', 
-                      pattern = 'adult', 
+                      pattern = 'txt', 
                       full.names = TRUE)
-turt <- llply(newturt, function(x) read.csv(x, header = FALSE))
-numbers <- llply(turt, function(x) x[, 1:2])
+turt <- llply(newturt, function(x) read.delim(x, header = FALSE, sep = ' '))
+# for some reason there are 2 dead columns...
+turt <- llply(turt, function(x) { 
+                x = x[1:27]
+                x})
+
+# need to get rid of the JRB specimens
+inturt <- list.files('../data/new_turtle', 
+                      pattern = 'list.csv', 
+                      full.names = TRUE)
+# blan, coa, gut, ins, muh, orb, orn, pic
+inturt <- inturt[c(1, 3, 4, 5, 6, 7, 8, 2)]
+numbers <- llply(inturt, function(x) read.csv(x, header = TRUE))
+
+# remove JRB before things get awkward
+spec.source <- llply(numbers, function(x) as.character(x[, 2]))
+to.rm <- llply(spec.source, function(x) str_detect(x, 'JRB'))
+
+turt <- Map(function(x, y) {x = x[!y, ]; x}, turt, to.rm)
+numbers <- Map(function(x, y) {x = x[!y, ]; x}, numbers, to.rm)
+
+# ok, onto the analysis
 centroids <- llply(turt, function(x) x[, ncol(x)])
-turt <- llply(turt, function(x) x[, -c(1:2, ncol(x))])
 # number, museum #, lands...., centroid
 turt <- Reduce(rbind, turt)
 turt.align <- df2array(turt, n.land = 26, n.dim = 2)
@@ -54,10 +73,12 @@ centroids <- scale(unlist(centroids))
 turt.name <- laply(str_split(newturt, '\\/'), function(x) x[length(x)])
 turt.name <- str_trim(str_extract(turt.name, '\\s(.*?)\\s'))
 turt.out[[1]] <- data.frame(sp = rep(turt.name, 
-                                        times = laply(numbers, nrow)), 
-                               size = centroids,
-                               inter = (centroids * turt.scores[, 1]),
-                               turt.scores, stringsAsFactors = FALSE)
+                                     times = laply(numbers, nrow)),
+                            size = centroids,
+                            inter = (centroids * turt.scores[, 1]),
[…]

### 2017-09-05 R/train_and_test.r

@@ -68,8 +68,8 @@ abrv.model <- function(method, adult, scheme, npred = 28) {
     traind <- list()
     for(ii in seq(from = 3, to = ncol(dataset))) {
       traind[[ii - 2]] <- 
-        train(form = responses ~ ., 
-              data = as.matrix(dataset[, seq(ii)]),
+        train(y = responses, 
+              x = dataset[, seq(ii)],
               method = method,
               trControl = ctrl,
               metric = 'ROC', 

### 2017-09-06 doc/turtles_statistical_learning.tex
removed white space around some plots. added a few citations

@@
 Each of these methods provide different advantages for understanding how to classify taxa, as well as the accuracy of the resulting classifications.
-Machine learning methods have been combined with geometric morphometric data to study shape variation in a variety of contexts, including automated taxon identification and classificaiton of groups \citep{Baylac2003,Dobigny2003,MacLeod2007,VandenBrink2011}.
+Machine learning methods have been combined with geometric morphometric data to study shape variation in a variety of contexts, including automated taxon identification and classificaiton of groups \citep{Baylac2003,Dobigny2003,MacLeod2007,VanBocxlaer2010,VandenBrink2011,Navega2015}.
 % MORE

### 2017-09-29 doc/turtles_statistical_learning.tex
updated analyses and manuscript to reflect different sampling approach

--- 2017-09-05 doc/turtles_statistical_learning.tex
@@
 \title{Ensemble approaches for estimating congruence between species delimitation and morphological variation: comparing taxonomic hypotheses for the Pacific Pond Turtle (\textit{Emys marmorata})}
-\author[1]{Peter D Smits}%\thanks{peterdavidsmits@gmail.com}}
+\author[1,2]{Peter D Smits}%\thanks{peterdavidsmits@gmail.com}}
 \author[2,3]{Kenneth D Angielczyk}%\thanks{kangielczyk@fieldmuseum.org}}
@@
 \affil[3]{Integrative Research Center, Field Museum of Natural History}
-\affil[4]{Department of Geological Sciences, California State University -- Fullerton}
+\affil[4]{John D. Cooper Archaeological and Paleontological Center, Department of Geological Sciences, California State University, Fullerton}
 \affil[5]{Section of Research and Collections, North Carolina Museum of Natural Sciences}
@@
 \begin{abstract}
-%  We investigated the morphometric identification of cryptic species using machine learning approaches by examining their implications for a recently proposed cryptic turtle species (\textit{Emys pallida}).
-We collected landmark-based morphometric data from 366 adult \textit{E.
+
+  We investigated the morphometric identification of cryptic species using machine learning approaches by examining their implications for a recently proposed cryptic turtle species (\textit{Emys pallida}).
+We collected landmark-based morphometric data from 532 adult \textit{E.
 marmorata/``E.
-pallida"} museum specimens.
+pallida''} museum specimens.
 We assigned a classification to each specimen for six different binning schemes based on geographic occurrence data recorded in museum collection archives.
@@
 marmorata}, and the other consisting of two subspecies of \textit{Trachemys scripta}.
-Our results indicate that there is no clear grouping of \textit{E.
+Our results indicate that there is no clear ``best'' grouping of \textit{E.
 marmorata/``E.
-pallida"} based on plastron shape.
+pallida''} based on plastron shape.
 In contrast, the analyses of the clear-cut examples produced near perfect classifications, demonstrating that the methods can recover correct results when an appropriate signal exists.
@@
 marmorata} include the possibility that genetic differentiation is not associated with plastron shape variation below the species level and/or that local selective pressures (e.g., from hydrological regime) overwhelm morphological differentiation.
-A reconsideration of the methods used to delimit \textit{``E.
-pallida"}, the lack of barriers to gene flow, the strong evidence for widespread admixture between lineages, and the fact that plastron shape can be used to delineate other emydine species and sub-species suggest that its lack of diagnosability most likely reflects the non-distinctiveness of this proposed taxon.
+A reconsideration of the methods used to delimit ``E.
+pallida,'' the lack of barriers to gene flow, the strong evidence for widespread admixture between lineages, and the fact that plastron shape can be used to delineate other emydine species and sub-species suggest that its lack of diagnosability most likely reflects the non-distinctiveness of this proposed taxon.
 \end{abstract}
@@
 Here, we investigate the morphometric identification of cryptic species using machine learning approaches.
-We use an ensemble learning approach where multiple methods are used in order to look for consensus amoung their results.
+We use an ensemble learning approach where multiple methods are used in order to look for consensus among their results.
 We test our approach on three datasets: plastron shape of seven species of closely related turtles, plastron shape of two species of closely related turtles, and plastron shape of the \textit{Emys marmorata} species complex.
@@
 \subsection{Background and study system}
-Machine learning is an extension of known statistical methodology \citep{Hastie2009} that emphasizes high predictive accuracy and generality at the expense of the interpretability of individual parameters.
-The basic statistical mechanics are supplemented by randomization, sorting, and partitioning algorithms, along with the maximization or minimization of summary statistics, in order to best estimate a general model for all data, both sampled and unsampled \citep{Hastie2009}.
-Machine learning approaches have found use in medical research, epidemiology, economics and automated identification of images such as handwritten zip codes \citep{Hastie2009}.
+Machine learning is an extension of known statistical methodology \citep{Hastie2009} that emphasizes high predictive accuracy and generality often at the expense of the interpretability of individual parameters.
+Basic statistical approaches are supplemented by randomization, sorting, and partitioning algorithms, along with the maximization or minimization of summary statistics, in order to best estimate a general model for all data, both sampled and unsampled \citep{Hastie2009}.
+Machine learning approaches have found use in medical research, epidemiology, economics, and automated identification of images such as handwritten zip codes \citep{Hastie2009}.
 % additional/more specific CITATIONS?
@@
 Supervised learning methods are used with labeled data where the final output of data is known and the rules for going from input to output are inferred.
-These are analogous to classification and regression models \citep{Breiman1984}.
-Our application of the approaches used in this study illustrates only a sampling of the various methods available for fitting classification models.
-The classification methods used in this study were chosen because they are suited for more than two response classes.
+These are analogous to classification and regression models \citep{Breiman1984,Hastie2009}.
+Our application of the supervised learning approaches used in this study illustrates only a sampling of the various methods available for fitting classification models.
+The specific methods used in this study were chosen because they are suited for cases with more two or more response classes.
 
-Geometric morphometric approaches to identifying differences in morphological variation between different classes, including cryptic species, mostly have used methods like linear discriminate analysis and canonical variates analysis \citep{Polly2003,Zelditch2004,Gaubert2005b,Gunduz2007,Polly2007a,Francoy2009,Sztencel-Jabonka2009,MitrovskiBogdanovic2013,Dillard2017}.
+Geometric morphometric approaches to identifying differences in morphological variation between classes, including cryptic species, have mostly relied on methods like linear discriminate analysis and canonical variates analysis \citep{Polly2003,Zelditch2004,Gaubert2005b,Gunduz2007,Polly2007a,Francoy2009,Sztencel-Jabonka2009,MitrovskiBogdanovic2013,Dillard2017}.
 Because of their similarity to multivariate approaches like principal components analysis (PCA), these methods are comparatively straightforward ways of understanding the differences in morphology between classes.
@@
 Each of these methods provide different advantages for understanding how to classify taxa, as well as the accuracy of the resulting classifications.
-Machine learning methods have been combined with geometric morphometric data to study shape variation in a variety of contexts, including automated taxon identification and classificaiton of groups \citep{Baylac2003,Dobigny2003,MacLeod2007,VandenBrink2011}.
+Machine learning methods have been combined with geometric morphometric data to study shape variation in a variety of contexts, including automated taxon identification and classification of groups \citep{Baylac2003,Dobigny2003,MacLeod2007,VanBocxlaer2010,VandenBrink2011,Navega2015}.
 % MORE
 %Although machine learning methods such as neural networks have been applied to studying shape variation, including in the context of automated taxon identification and classification of groups, the number of cases remains limited.
-In the current study, we not only consider pure classification accuracy but also use a statistic of classification strength that reflects the rate at which taxa are both accurately and inaccurately classified.
+In the current study, we not only consider pure classification accuracy but also use a statistic of classification strength that reflects the rate at which taxa are both accurately and inaccurately classified: the area under the Receiver Operating Characteristic curve \citep{Hastie2009}.
 
@@
 Second, previous work has suggested that there are strong differences in plastron shape among traditionally recognized emydine species \citep{Angielczyk2007,Angielczyk2011,Angielczyk2013a}.
-Finally, due to this previous study a large dataset was readily avaliable.
+Finally, due to this previous study a large dataset was readily available.
 
@@
 marmorata} species complex, we hypothesize that if one or more of the proposed classification schemes are consistent with the morphological data then our ensemble approach fit to those hypotheses will have higher out-of-sample predictive performance than the more inconsistent hypotheses.
-However, if all of the classification schemes lead to equal out-of-sample predictive performance then we would conclude that the molecular based hypotheses are inconsistent with whatever information is present in the morphological data.
+However, if all of the classification schemes lead to equal out-of-sample predictive performance then we would conclude that the proposed hypotheses are inconsistent with whatever information is present in the morphological data.
 Because of unclear geographic boundaries between subgroups of \textit{E.
@@
 We predict that the \textit{E.
-marmorata} dataset should be of the highest complexity and our greatest challenge given \citep{Holland1992} finding that only very subtle differences existed between geographically-distinct populations.
+marmorata} dataset should be of the highest complexity and our greatest challenge given the finding that only very subtle differences existed between geographically-distinct populations \citep{Holland1992}.
 %We expect the first dataset represents a low-complexity dataset given the high level of plastron shape disparty that exists among these species \citep{Angielczyk2011}, whereas the second dataset should be relatively higher in complexity than the first.
@@
 
-The final dataset is of 366 adult \textit{E.
-marmorata} museum specimens.
-These speciemens represent a subset of those included in \citet{Angielczyk2007}, \citet{Angielczyk2011}, and \citet{Angielczyk2013a}.
+The final dataset is of 532 adult E.
+marmorata museum specimens, though not all specimens were able to be assigned a class for all schemes (Fig.
+\ref{fig:map}).
+These specimens represent a subset of those included in \citet{Angielczyk2007}, \citet{Angielczyk2011}, and \citet{Angielczyk2013a}.
 Because  \citet{Spinks2005}, \citet{Spinks2010}, and \citet{Spinks2014} did not use vouchered specimens we were not able to directly sample the individuals in their studies.
 Instead, our specimen classifications were based solely on the geographic information and not explicit assignment using molecular data.
-For each taxonomic hypothesis, specimens were assigned to one of the possible classes based on geographic occurrence data recorded in museum collection archives.
+For each taxonomic hypothesis, specimens were assigned to one of the possible classes based on geographic occurrence data recorded in museum collections.
 In cases where precise latitude and longitude information were not available we estimated for them from other locality information.
 Because the exact barriers between different biogeographic regions are unknown and unclear, we represented each hypothesis with multiple possible realizations representing the classification uncertainty for specimens present at the geographic boundaries.
-The taxonomic hypotheses and sub-hypotheses for the \textit{E.
-marorata} used here are presented in Table \ref{tab:hypotheses}.
+The taxonomic hypotheses and sub-hypotheses for \textit{E.
+marmorata} used here are presented in Table \ref{tab:hypotheses} and Figure \ref{fig:map}.
+
+For \citet{Spinks2010} we used three binning schemes.
+All three schemes include a class for \textit{E.
+marmorata} specimens from northern populations (marm) as well as a class for those assigned to \textit{E.
+pallida} (pall) and an intergrade zone in the Central Coast Ranges (CCR).
+The schemes differ in the assignment of samples from the San Joaquin Valley (Fig.
+\ref{fig:map}).
+Scheme SP10.1 and SP10.2 differ in the assignment of specimens from the western San Joaquin Valley to either CCR or marm reflecting uncertainty regarding their genetic affinity as explained above.
+In scheme SP10.3 these specimens are assigned to a San Joaquin class reflecting the mitochondrial distinctiveness shown by \citet{Spinks2005}.
+For \citet{Spinks2014} we used two binning schemes with SP14.1 being based on their phylogenetic network analysis and SP14.2 being based on their Bayesian species delimitation analysis.
+The latter scheme requires the addition of two new classes, ``Baja'' and ``Foothill,'' to accommodate the genetic groupings recovered by the SNP Structure analysis that was used to create the guide tree for the BPP species delimitation analysis in \citet{Spinks2014}.
+Finally, we proposed a conservative morphological hypothesis (``Morph'') in order to compare the molecular hypotheses with something approximating the original taxonomic hypothesis for the group; this scheme is made up solely of the marm and pall classes from the SP10.3 scheme.
 
@@
 % CHANGE
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[height = 0.8\textheight, width = \textwidth, keepaspectratio = true]{figure/Ken_Ang_SpLoc_DRAFT}
+  \caption{Geographic distribution of specimens sampled for comparing the hypothesized subdivisions of \textit{Emys marmorata}.
+Each hypothesized scheme has two or more possible classes.
+Sample size differs between schemes because of our ability to confidently assign museum specimens to the various schemes.}
+  \label{fig:map}
+\end{figure}
 
@@
       SP14.2 & 4 & \citet{Spinks2014} \\
+      Morph & 2 & \citet{Spinks2010} \\
       \hline
@@
 We superimposed the plastral landmark configurations using generalized Procrustes analysis \citep{Dryden1998a}, after which we calculated the principal components (PC) of shape using the \texttt{shapes} package for R \citep{R2016,Dryden2013}.
+All specimens were used for superimposition, after which the subset labeled for each of the schemes were used in model training and testing (see below).
 
@@
 
-\textit{Emys marmorata} is known to display sexual dimorhpism in plastron shape, particulary the presence of a plastra concavity in males \citep{Seeliger1945}.
+\textit{Emys marmorata} is known to display sexual dimorphism in plastron shape, particularly the presence of a plastra concavity in males \citep{Seeliger1945}.
 To test for biases resulting from sexual dimorphism in our \textit{E.
 marmorata} dataset, we used a simple permutation test to determine if the distance between the mean female and male shapes is greater than expected when the sex labels are randomly shuffled.
-Because not all of our specimens have sex identifications associated with them, this analysis was done using a subset of the data (257 of 360).
+Because not all of our specimens have sex identifications associated with them, this analysis was done using a subset of the data (257 of 532).
 
 \subsection{Supervised learning approaches}
-Instead of relying on a single supervised learning method, we chose to use an ensemble approach where multiple model types are used in concert so that any congruence between the them increases our support for that conclusion over another \citep{Hastie2009}.
+Instead of relying on a single supervised learning method, we chose to use an ensemble approach where multiple model types are used in concert so that any congruence between them increases our support for that conclusion over another \citep{Hastie2009}.
 The supervised learning methods used here are named in Table \ref{tab:methods}.
 Each of these methods makes different assumptions, treats data differently, and can produce different classification results depending on the nature of the data \citep{Hastie2009}.
-For example, multinomial logistic regression is a type of generalized linear model while random forest is an ensemble approach where multiple decision trees are fit to subsets of the full data and then averaged.
+For example, multinomial logistic regression is a type of generalized linear model while random forest is itself an ensemble approach where multiple decision trees are fit to subsets of the full data and then averaged.
 % use multiple methods
@@
   \caption{Table of the supervised learning methods used in this analysis.}
+  \scalebox{0.9}{
   \begin{tabular}{l l l l}
@@
     \hline
-  \end{tabular}
+  \end{tabular}}
   \label{tab:methods}
@@
 The former is used for fitting the model whereas the later is used for measuring model performance, a process called model generalization.
-In this analysis, we used 80\% of samples as the training set while the remaining 20\% were used as the testing set.
+For each scheme, we limited the model training and testing to only those individuals with class labels for that scheme.
+In this analysis, we randomly divided 80\% of samples into the training set and the remaining 20\% into the testing set.
 
@@
 The selected or final model, however, is the next most parsimonious model that is within one standard error of the best model; this is a variant on the ``one-standard error'' rule from \citet{Hastie2009}.
-This rule is applied to ameliorate the chance of selecting an overly complex model.
+The purpose of this rule is to ameliorate the chances of selecting an overly complex model that will perform poorly when predicting the classes of out-of-sample data.
 
@@
 \ref{fig:other_pca}).
-Additionally, there appears to be both little difference in centroid size across individuals as well as little correspondence between centroid size and the PC axes.
 
@@
   \caption{Scatterplot of the first two axes of morphological variation in the \textit{Emys marmorata} dataset.
-Each panel corresponds to each of different proposed classification schemes analyzed as part of this study.
-Point color corresponds to the categories within each scheme, and the category names corresond to geographic regions.
-Point size is proportional to centroid size and the numbers in parantheses next to the axis labels are the percent of total variation accounted for by that axis.}
+Each panel corresponds to one of the six different classification schemes analyzed as part of this study (Tab.
+\ref{tab:hypotheses}).
+Point color corresponds to the categories within each scheme, and the class names correspond to geographic regions.
+Point size is proportional to centroid size of that specimen and the numbers in parentheses next to the axis labels are the percent of total variation accounted for along that axis.}
   \label{fig:emys_pca}
@@
 \ref{fig:sex_test}).
-The difference in mean shape between the sexes is very small; the sexes overlap slightly more than is expected from a null distribution based on permuting the sex-labels.
+The difference in mean shape between the sexes is very small; the sexes overlap about has much as expected given a null distribution based on permuting the sex-labels.
 
@@
     \hline
-    \textit{E. marmorata} & SP10.1 & 0.93 & 1.03 \\
-     & SP10.2 & 0.93 & 1.03 \\
-     & SP10.3 & 0.88 & 0.98 \\
-     & SP14.1 & 0.98 & 1.09 \\
-     & SP14.2 & 0.87 & 0.97 \\
+    \textit{E. marmorata} & SP10.1 & 0.99 & 1.10 \\
+     & SP10.2 & 1.00 & 1.11 \\
+     & SP10.3 & 0.94 & 1.04 \\
+     & SP14.1 & 1.01 & 1.12 \\
+     & SP14.2 & 0.93 & 1.04 \\
+     & Morph & 0.99 & 1.09 \\
     \hline
@@
   \caption{AUC values for models of varying complexity fit to the \textit{Emys marmorata} training datasets for each classification scheme.
-The x-axis corresponds to the total number of predictors included in each model, while the y-axis corresponds to the AUC value which is a measure of goodness of fit for classificaiton datasets.
+The x-axis corresponds to the total number of predictors included in each model, while the y-axis corresponds to the AUC value which is a measure of goodness of fit for classification datasets.
 A model with a high AUC value corresponds to better classification performance than a model with a lower AUC value.
@@
 Observed AUC values for all of the optimal models are lower for the \textit{E.
-marmorata} dataset (Fig.
-\ref{fig:emys_sel}).
+marmorata} dataset than for the other two datasets (Fig.
+\ref{fig:other_sel}, \ref{fig:emys_sel}).
 In most cases the different proposed classification schemes are generally poor descriptors of the observed variation.
@@
 \ref{fig:emys_sel}, \ref{fig:emys_oos}).
+Only in the case of the conservative morphological hypothesis (``Morph'') is the mean AUC value potentially distinct from that of other schemes; in this case mean AUC is lower than the average of the other five schemes which indicates that the moprhologically-based scheme performs more poorly than the molecularly-based ones.
+It is important to note, however, that the training and testing dataset for the ``Morph'' scheme is the smallest of the six schemes which may lead to poorer performance in in-sample and out-of-sample comparisons.
 
@@
 As expected, our ensemble approach yields high out-of-sample classification performance for the first two datasets.
-These results indicate that in cases of clear class seperation (Fig.
+These results indicate that in cases of clear class separation (Fig.
 \ref{fig:other_pca}) our approach is able to detect this and make good out-of-sample prediction.
@@
 Moreover, the genetic diversity uncovered by the analysis of \citet{Spinks2014} should be accounted for explicitly in any conservation plan.
-Given the apparent lack of morphological distinction, however, we consider that this diversity should be considered Evolutionary Significant Units or Distinct Population Segments instead of distinct species.
+Given the apparent lack of morphological distinction combined with the broad range of intergradation and other problems with the species hypothesis outlined above, we recommend that the populations elevated to \textit{E.
+pallida} by \citet{Spinks2014} are best considered Evolutionary Significant Units or Distinct or Population Segments instead of distinct species.
 

### 2017-09-29 R/digit_error.r

@@ -33,11 +33,14 @@ land.dist <- function(shapes) {
 
 
 # emys set
-scheme <- c('sp10.1', 'sp10.2', 'sp10.3', 'sp14.1', 'sp14.2')
+scheme <- c('sp10.1', 'sp10.2', 'sp10.3', 'sp14.1', 'sp14.2', 'morph')
 ff <- adult[, scheme]
 emys.split <- list()
 for(ii in seq(length(scheme))) {
-  emys.split[[ii]] <- split.land(fit$rotated, ff[, ii])
+  rms <- is.na(ff[, ii]) | ff[, ii] == ''
+  temp <- ff[!rms, ii]
+  tempfit <- fit$rotated[, , !rms]
+  emys.split[[ii]] <- split.land(tempfit, temp)
 }
 emys.dist <- land.dist(fit$rotated)
 within.scheme <- llply(emys.split, function(x) llply(x, land.dist))
@@ -62,7 +65,7 @@ for(kk in seq(length(scheme.mean))) {
 
 scheme.ratio <- laply(within.scheme, function(x) 
                       mean(x) / mean(emys.dist[lower.tri(emys.dist)]))
-
+#scheme.ratio * 1.11
 
 # replicated turtles
 

### 2017-09-29 R/supervised_mung.r

@@ -39,10 +39,10 @@ meta[, c('lat', 'long')] <- apply(meta[, c('lat', 'long')],
 meta[meta == '?' | meta == '??'] <- NA
 
 # get rid of everything that doesn't have lat-long
-#meta$long <- meta$long * -1
-#land <- land[, , !is.na(meta$lat)]
-#rawturt<- rawturt[!is.na(meta$lat), ]
-#meta <- meta[!is.na(meta$lat), ]
+#meta$long <- meta$long
+land <- land[, , !is.na(meta$lat)]
+rawturt<- rawturt[!is.na(meta$lat), ]
+meta <- meta[!is.na(meta$lat), ]
 
 # get rid of everything that doesn't have assigns
 # clean off the missing important values
@@ -59,7 +59,13 @@ jv <- grepl(pattern = '[jJh]', meta$p.sex)
 land.adult <- land[, , !jv]
 meta.adult <- meta[!jv, ]
 rawturt <- rawturt[!jv, ]
+
+rms <- is.na(meta.adult$sp10.1) | meta.adult$sp10.1 == ''
+land.adult <- land.adult[, , !rms]
+meta.adult <- meta.adult[!rms, ]
+rawturt <- rawturt[!rms, ]
 centroid <- rawturt[, n.land + 1]  # size of each observation
+
 
 geo <- cbind(lat = meta.adult$lat, long = meta.adult$long)
 
@@ -71,5 +77,12 @@ adult <- cbind(data.frame(size = centroid,
                           inter = centroid * fit$stdscores[, 1],
                           fit$stdscores),
                meta.adult)
+
+
+schemes <- c('sp10.1', 'sp10.2', 'sp10.3', 'sp14.1', 'sp14.2', 'morph')
+nspec <- aaply(meta.adult[, schemes], 2, function(x) {
+                 n <- is.na(x) | x == ''
+                 length(x[!n])})
+meta.adult <- meta.adult[!(is.na(meta.adult$sp10.1) | meta.adult$sp10.1 == ''), ]
 
 write.table(meta.adult, file = '../data/clean_meta_marm.csv')

### 2017-09-29 R/visual_class.r

@@ -137,14 +137,13 @@ scores <- Reduce(rbind,
 scores.df <- data.frame(scores)
 scores.df$class <- factor(ad, levels = level)
 scores.df$sch <- sch
-perc <- fit$percent[1:2]
 scores.df$centroid <- rep(rawturt[, n.land + 1], length(schemes))
 
 # need to be made human readable
 scores.df$class <- as.character(scores.df$class)
 scores.df <- scores.df[!is.na(scores.df$class), ]
 scores.df <- scores.df[scores.df$class != '', ]
-unique(scores.df$class)
+#unique(scores.df$class)
 scores.df$class <- mapvalues(scores.df$class, 
                              from = unique(scores.df$class), 
                              to = c('Central Coast', 
@@ -152,6 +151,8 @@ scores.df$class <- mapvalues(scores.df$class,
                                     'San Juan', 'Baja California', 
                                     'Sierra Foothills'))
 
+perc <- fit$percent[1:2]
+perc <- round(perc, digits = 1)
 emys.gg <- ggplot(scores.df, aes(x = PC1, y = PC2, 
                                  colour = class, size = centroid))
 emys.gg <- emys.gg + geom_point(alpha = 0.5)
@@ -161,7 +162,6 @@ emys.gg <- emys.gg + coord_fixed(ratio = 1,
                                  xlim = c(-3.5, 3.5), 
                                  ylim = c(-3.5, 3.5))
 emys.gg <- emys.gg + scale_colour_manual(values = cbp.ord)
-perc <- round(perc, digits = 1)
 emys.gg <- emys.gg + labs(x = paste0('PC 1 (', perc[1], '%)'), 
                           y = paste0('PC 2 (', perc[2], '%)'))
 ggsave(plot = emys.gg, filename = '../doc/figure/emys_pc_graph.png',

### 2017-10-11 R/supervised_mung.r
changed encoding of export

@@ -83,6 +83,9 @@ schemes <- c('sp10.1', 'sp10.2', 'sp10.3', 'sp14.1', 'sp14.2', 'morph')
 nspec <- aaply(meta.adult[, schemes], 2, function(x) {
                  n <- is.na(x) | x == ''
                  length(x[!n])})
-meta.adult <- meta.adult[!(is.na(meta.adult$sp10.1) | meta.adult$sp10.1 == ''), ]
 
-write.table(meta.adult, file = '../data/clean_meta_marm.csv')
+cliped <- meta.adult[, c('spec', 'lat', 'long', 
+                         'sp10.1', 'sp10.2', 'sp10.3', 
+                         'sp14.1', 'sp14.2', 
+                         'morph')]
+write.csv(cliped, file = '../data/clean_meta_marm.csv', fileEncoding = 'UTF-16LE')

### 2017-11-28 doc/turtles_statistical_learning.tex
changed encoding of export
data scare. remade all figures and did ken and jim edits. one last read, then off to coauthors before final submit

@@
 marmorata} include the possibility that genetic differentiation is not associated with plastron shape variation below the species level and/or that local selective pressures (e.g., from hydrological regime) overwhelm morphological differentiation.
-A reconsideration of the methods used to delimit ``E.
-pallida,'' the lack of barriers to gene flow, the strong evidence for widespread admixture between lineages, and the fact that plastron shape can be used to delineate other emydine species and sub-species suggest that its lack of diagnosability most likely reflects the non-distinctiveness of this proposed taxon.
+A reconsideration of the methods used to delimit \textit{``E.
+pallida,''} the lack of barriers to gene flow, the strong evidence for widespread admixture between lineages, and the fact that plastron shape can be used to delineate other emydine species and sub-species suggest that its lack of diagnosability most likely reflects the non-distinctiveness of this proposed taxon.
 \end{abstract}
@@
 We use an ensemble learning approach where multiple methods are used in order to look for consensus among their results.
-We test our approach on three datasets: plastron shape of seven species of closely related turtles, plastron shape of two species of closely related turtles, and plastron shape of the \textit{Emys marmorata} species complex.
+We test our approach on three datasets: plastron shape of seven species of closely related turtles, plastron shape of two subspecies of a single turtle species, and plastron shape of the \textit{Emys marmorata} species complex.
 In particular, we ask whether it is possible to determine which among a set of classification hypotheses best aligns with the observed morphology, and examine the implications of our results for the \textit{E.
@@
 
-Here, we used an ensemble of supervised machine learning methods to compare the congruence of the morphological data to different classification hypotheses.
-Each of these methods provide different advantages for understanding how to classify taxa, as well as the accuracy of the resulting classifications.
+Here, we used an ensemble of supervised machine learning methods to compare the congruence of morphological data to different classification hypotheses.
+Each of these methods provide different advantages for understanding how to classify specimens, as well as the accuracy of the resulting classifications.
 Machine learning methods have been combined with geometric morphometric data to study shape variation in a variety of contexts, including automated taxon identification and classification of groups \citep{Baylac2003,Dobigny2003,MacLeod2007,VanBocxlaer2010,VandenBrink2011,Navega2015}.
@@
 marmorata} is the morphometric analysis of carapace shape carried out by \citet{Holland1992}, who compared populations of \textit{E.
-marmorata} from three areas of the species range.
+marmorata} from three areas of the species' range.
 Holland concluded that geographic distance was a poor indicator of morphological differentiation, and instead hypothesized that geographic features such as breaks between different drainage basis are probably more important barriers to dispersal and interbreeding.
@@
 This approach allows both museum specimens and individuals in the field to be analyzed together.
-Second, previous work has suggested that there are strong differences in plastron shape among traditionally recognized emydine species \citep{Angielczyk2007,Angielczyk2011,Angielczyk2013a}.
-Finally, due to this previous study a large dataset was readily available.
+Second, previous work has suggested that there are strong differences in plastron shape among traditionally-recognized emydine species \citep{Angielczyk2007,Angielczyk2011,Angielczyk2013a}.
+Finally, due to these previous studies a large dataset was readily available.
 
@@
 
-The final dataset is of 532 adult E.
-marmorata museum specimens, though not all specimens were able to be assigned a class for all schemes (Fig.
+The final dataset is of 532 adult \textit{E.
+marmorata} museum specimens, though not all specimens were able to be assigned a class for all schemes (Fig.
 \ref{fig:map}).
@@
 For each taxonomic hypothesis, specimens were assigned to one of the possible classes based on geographic occurrence data recorded in museum collections.
-In cases where precise latitude and longitude information were not available we estimated for them from other locality information.
+In cases where precise latitude and longitude information were not available we estimated them from other locality information.
 Because the exact barriers between different biogeographic regions are unknown and unclear, we represented each hypothesis with multiple possible realizations representing the classification uncertainty for specimens present at the geographic boundaries.
@@
   \centering
-  \includegraphics[height = 0.8\textheight, width = \textwidth, keepaspectratio = true]{figure/Ken_Ang_SpLoc_DRAFT}
+  \includegraphics[height = 0.8\textheight, width = \textwidth, keepaspectratio = true]{figure/Ken_Ang_SpLoc_final}
   \caption{Geographic distribution of specimens sampled for comparing the hypothesized subdivisions of \textit{Emys marmorata}.
@@
 % digitization
-We estimated the possible effect of digitization error \citep{Arnqvist1998,Cramon2007,Munoz-MunozF.2010} on our results by comparing within (replicated) specimen Procrustes distances to the distances between classification scheme centroids.
-Ten randomly selected \textit{E.
+We estimated the possible effect of digitization error \citep{Arnqvist1998,Cramon2007,Munoz-MunozF.2010} on our results by comparing within-specimen (replicated) Procrustes distances to the distances between classification scheme centroids.
+Ten randomly-selected \textit{E.
 marmorata} specimens were each digitized four times, with the original set of digitized coordinates serving as a fifth replicate.
@@
 
-\textit{Emys marmorata} is known to display sexual dimorphism in plastron shape, particularly the presence of a plastra concavity in males \citep{Seeliger1945}.
+\textit{Emys marmorata} is known to display sexual dimorphism in plastral shape, particularly the presence of a plastra concavity in males \citep{Seeliger1945}.
 To test for biases resulting from sexual dimorphism in our \textit{E.
@@
 Each of these methods makes different assumptions, treats data differently, and can produce different classification results depending on the nature of the data \citep{Hastie2009}.
-For example, multinomial logistic regression is a type of generalized linear model while random forest is itself an ensemble approach where multiple decision trees are fit to subsets of the full data and then averaged.
+For example, multinomial logistic regression is a type of generalized linear model, whereas random forest is itself an ensemble approach where multiple decision trees are fit to subsets of the full dataset and then averaged.
 % use multiple methods
@@
 In contrast, the different \textit{E.
-marmorata} classification schemes appear to be barely be distinct, with their within:between ratios approximating 1.
-This indicates that the magnitude of the differences between groupings is approximately the same as the difference as any two random individuals (Table \ref{tab:rep_res}).
+marmorata} classification schemes appear to barely be distinct, with their within:between ratios approximating 1.
+This indicates that the magnitude of the differences between groupings is approximately the same as the difference between any two random individuals (Table \ref{tab:rep_res}).
 
@@
 
-Analysis of the seven morphologically and genetically distinct species and the \textit{T.
+Analysis of the seven morphologically- and genetically-distinct species and the \textit{T.
 scripta scripta}--\textit{T.
-scripta elegans} datasets indicate that these classifications are sufficiently morphologically distinct to be differentiated on the basis of plastron shape.
+scripta elegans} datasets indicate that these taxa are sufficiently morphologically distinct to be differentiated on the basis of plastron shape.
 Both in-sample and out-of-sample classification have AUC values of approximately 1 for all methods, implying near-perfect classification rates (Fig.
@@
 \ref{fig:other_oos}).
-These results demonstrate that when there are distinctions between the states of the classification schemes (i.e.
-differences in plastron shape that correlate with the different taxonomic groups), the methods used here can recover them.
+These results demonstrate that when there are distinctions between the states of the classification schemes (i.e., differences in plastron shape that correlate with the different taxonomic groups), the methods used here can recover them.
 
@@
 from hydrological regime) overwhelm morphological differentiation.
-Both of these options seem plausible given that shell shape is influenced by selection for both protection and streamlining, but not necessary mate choice \citep{Rivera2008,Rivera2011,Stayton2011,Rivera2014,Polly2016} and that shell shape in \textit{E.
+Both of these options seem plausible given that shell shape is influenced by selection for both protection and streamlining, but not necessary mate choice \citep{Rivera2008,Rivera2011,Stayton2011,Rivera2014,Polly2016}, and that shell shape in \textit{E.
 marmorata} is known to vary among populations inhabiting water bodies with different flow regimes \citep{Holland1992,Lubcke2007,Germano2009}.
@@
 marmorata} system until now, and in doing so we find the proposal that \textit{E.
-pallida} is a distinct species to be lacking in a normally robust morphological marker.
+pallida} is a distinct species to be lacking.
 
-The natural history and geographical distribution of \textit{E.
+In addition to lacking a robust morphological marker, the natural history and geographical distribution of \textit{E.
 marmorata} and \textit{E.
-pallida} also make the recognition of these taxa implausible.
-The data from \citet{Spinks2014} show extensive introgression and admixture in Central California, which is expected because there are no significant barriers to gene flow in this region.
+pallida} also make the recognition of these two taxa implausible.
+The mitochondrial data from \citet{Spinks2014} show extensive introgression and admixture in Central California, which is expected because there are no significant barriers to gene flow in this region.
+They also lack sampling from the populations between the two putative species in the San Francisco Bay Area, which we predict would likely show even more genetic mixing.
 Combined with the well-demonstrated ability for testudinoid turtles, including emydids and even \textit{Emys}, to hybridize (e.g.
@@
 pallida} could maintain their integrity in the face of such admixture.
-Because the geography, natural history, demonstrated genetic admixture of \textit{E.
+Any argument for the validity of \textit{E.
+pallida} as a distinct species needs to address these points.
+Because the geography, natural history, limited sampling from key areas, demonstrated genetic admixture of \textit{E.
 marmorata}, and comparisons with other morphologically diagnosable species and subspecies conflict with the recognition of \textit{E.
-pallida}, we hypothesize that our inability to classify the morphological data by proposed species is because \textit{E.
+pallida}, we hypothesize that \textit{E.
 pallida} is not a distinct species.
@@
 Given the apparent lack of morphological distinction combined with the broad range of intergradation and other problems with the species hypothesis outlined above, we recommend that the populations elevated to \textit{E.
-pallida} by \citet{Spinks2014} are best considered Evolutionary Significant Units or Distinct or Population Segments instead of distinct species.
+pallida} by \citet{Spinks2014} are best considered Evolutionary Significant Units or Distinct Population Segments instead of distinct species.
 
@@
 Sweet for field assistance and the California Department of Fish and Game for permits.
+We would also like to thank Marc Lambruschi (FMNH) for help with figure \ref{fig:map}.
 

### 2017-11-28 R/clear_cut.r

@@ -51,7 +51,7 @@ inturt <- list.files('../data/new_turtle',
                       pattern = 'list.csv', 
                       full.names = TRUE)
 # blan, coa, gut, ins, muh, orb, orn, pic
-inturt <- inturt[c(1, 3, 4, 5, 6, 7, 8, 2)]
+#inturt <- inturt[c(1, 3, 4, 5, 6, 7, 8, 2)]
 numbers <- llply(inturt, function(x) read.csv(x, header = TRUE))
 
 # remove JRB before things get awkward
@@ -101,15 +101,15 @@ turt.out[[2]] <- data.frame(sp = ids, size = centroids,
 
 meth <- c('multinom', 'nnet', 'lda', 'pda', 'rf')
 schemes <- c('cc7', 'trac')
-#results <- list()
-#for(ii in seq(length(meth))) {
-#  results[[ii]] <- abrv.model(method = meth[ii], 
-#                              adult = turt.out, 
-#                              scheme = schemes, 
-#                              npred = 25)
-#}
-#names(results) <- meth
-#save(results, file = '../data/others_cv_results.rdata')
+results <- list()
+for(ii in seq(length(meth))) {
+  results[[ii]] <- abrv.model(method = meth[ii], 
+                              adult = turt.out, 
+                              scheme = schemes, 
+                              npred = 25)
+}
+names(results) <- meth
+save(results, file = '../data/others_cv_results.rdata')
 load('../data/others_cv_results.rdata')
 
 

### 2017-11-28 R/train_and_test.r

@@ -49,3 +49,51 @@ use.model <- function(method, adult, scheme, npred = 28) {
                     training.dataset = trainsets, testing.dataset = testsets)
   final.res
 }
+
+
+abrv.model <- function(method, adult, scheme, npred = 28) {		
+
+  part <- llply(adult, function(x) 		
+                createDataPartition(x[, 1], p = 0.8, list = FALSE, times = 1))		
+
+  trainsets <- testsets <- list()		
+  train.res <- test.res <- list()		
+
+  train.res <- test.res <- list()		
+  for(kk in seq(length(scheme))) {		
+    trainsets[[kk]] <- adult[[kk]][part[[kk]], ]		
+    testsets[[kk]] <- adult[[kk]][-part[[kk]], ]		
+
+    dataset <- trainsets[[kk]][, 2:(npred + 1)]		
+    responses <- trainsets[[kk]][, 1]		
+
+    traind <- list()		
+    for(ii in seq(from = 3, to = ncol(dataset))) {		
+      traind[[ii - 2]] <- 		
+        train(y = responses, 		
+              x = dataset[, seq(ii)],		
+              method = method,		
+              trControl = ctrl,		
+              metric = 'ROC', 		
+              tuneLength = 1)		
+    }		
+    train.res[[kk]] <- traind		
+
+
+    # now test all the models against external data		
+    pclass <- llply(traind, function(x) 		
+                    predict.train(object = x, 		
+                                  newdata = testsets[[kk]][, 2:(npred + 1)], 		
+                                  type = 'raw'))		
+    prob <- llply(traind, function(x)		
+                  predict.train(object = x, 		
+                                newdata = testsets[[kk]][, 2:(npred + 1)], 		
+                                type = 'prob'))		
+    test.res[[kk]] <- Map(function(x, y) cbind(class = x, y), pclass, prob)		
+  }		
+
+  final.res <- list(training = train.res, testing = test.res,		
+                    training.dataset = trainsets, testing.dataset = testsets)		
+  final.res		
+}
+

### 2017-11-28 R/visual_class.r

@@ -42,7 +42,7 @@ inturt <- list.files('../data/new_turtle',
                       pattern = 'list.csv', 
                       full.names = TRUE)
 # blan, coa, gut, ins, muh, orb, orn, pic
-inturt <- inturt[c(1, 3, 4, 5, 6, 7, 8, 2)]
+#inturt <- inturt[c(1, 3, 4, 5, 6, 7, 8, 2)]
 numbers <- llply(inturt, function(x) read.csv(x, header = TRUE))
 
 # remove JRB before things get awkward

### 2017-11-29 doc/turtles_statistical_learning.tex
sent final to coauthors. submit soon?

@@
 \title{Ensemble approaches for estimating congruence between species delimitation and morphological variation: comparing taxonomic hypotheses for the Pacific Pond Turtle (\textit{Emys marmorata})}
-\author[1,2]{Peter D Smits}%\thanks{peterdavidsmits@gmail.com}}
+\author[1]{Peter D Smits}%\thanks{peterdavidsmits@gmail.com}}
 \author[2,3]{Kenneth D Angielczyk}%\thanks{kangielczyk@fieldmuseum.org}}
@@
 We used an ensemble of supervised machine learning approaches to determine which classification hypothesis was best supported by the data.
-In addition, we applied the same approach to two clear-cut examples, one consisting of seven unambiguously distinct species closely related to \textit{E.
+In addition, we applied the same approach to two clear-cut examples, one consisting of eight unambiguously distinct species closely related to \textit{E.
 marmorata}, and the other consisting of two subspecies of \textit{Trachemys scripta}.
@@
 We use an ensemble learning approach where multiple methods are used in order to look for consensus among their results.
-We test our approach on three datasets: plastron shape of seven species of closely related turtles, plastron shape of two subspecies of a single turtle species, and plastron shape of the \textit{Emys marmorata} species complex.
+We test our approach on three datasets: plastron shape of eight species of closely related turtles, plastron shape of two subspecies of a single turtle species, and plastron shape of the \textit{Emys marmorata} species complex.
 In particular, we ask whether it is possible to determine which among a set of classification hypotheses best aligns with the observed morphology, and examine the implications of our results for the \textit{E.
@@
 \subsection{Background and study system}
-Machine learning is an extension of known statistical methodology \citep{Hastie2009} that emphasizes high predictive accuracy and generality often at the expense of the interpretability of individual parameters.
+Machine learning is an extension of known statistical methodology \citep{Hastie2009} that emphasizes predictive accuracy and generality often at the expense of the interpretability of individual parameters.
 Basic statistical approaches are supplemented by randomization, sorting, and partitioning algorithms, along with the maximization or minimization of summary statistics, in order to best estimate a general model for all data, both sampled and unsampled \citep{Hastie2009}.
@@
 \subsection{Specimens, sampling, morphometrics}
-Three different geometric morphometric datasets describing turtle plastron variation were assembled for this analysis: 1) specimens from seven distinct emydine species; 2) \textit{T.
+Three different geometric morphometric datasets describing turtle plastron variation were assembled for this analysis: 1) specimens from eight distinct emydine species; 2) \textit{T.
 scripta} specimens from the two main subspecies (\textit{T.
@@
 % three turtle datasets
-%   seven species
+%   eight species
 %   Trachemys species
@@
 
-The first dataset we analyzed includes 578 total specimens from the following species: \textit{Clemmys guttata}, \textit{Emys blandigii}, \textit{Emys orbicularis}, \textit{Glyptemys insculpta}, \textit{Glyptemys muhlenbergii}, \textit{Terrapene coahuila}, and \textit{Terrapene ornata}.
+The first dataset we analyzed includes 578 total specimens from the following species: \textit{Chrysemys picta}, \textit{Clemmys guttata}, \textit{Emys blandigii}, \textit{Emys orbicularis}, \textit{Glyptemys insculpta}, \textit{Glyptemys muhlenbergii}, \textit{Terrapene coahuila}, and \textit{Terrapene ornata}.
 These specimens are a subset of those used in \citet{Angielczyk2011} and \citet{Angielczyk2013a}.
@@
   \centering
-  \includegraphics[height = 0.8\textheight, width = \textwidth, keepaspectratio = true]{figure/Ken_Ang_SpLoc_final}
+  \includegraphics[height = 0.8\textheight, width = \textwidth, keepaspectratio = true]{figure/Ken_Ang_SpLoc_final_comp}
   \caption{Geographic distribution of specimens sampled for comparing the hypothesized subdivisions of \textit{Emys marmorata}.
 Each hypothesized scheme has two or more possible classes.
-Sample size differs between schemes because of our ability to confidently assign museum specimens to the various schemes.}
+Sample size differs between schemes because of our ability to confidently assign museum specimens to the various schemes.
+The different classification abbreviations are as follows: \textit{E.
+marmorata} = ``marm'', \textit{E.
+pallida} = ``pall'', Central Coast Ranges = ``CCR'', San Joaquin Valley = ``SJ,'' Baya Peninsula = ``Baja,'', and Sierra Foothills = ``Foothill.''}
   \label{fig:map}
@@
 Both ROC and AUC are preferable to simple classification accuracy when class membership is unbalanced, as it is in these analyses \citep{Hastie2009}.
-The standard ROC and AUC calculations are defined only for binary classifications, which is not the case for our seven species and \textit{Emys} complex datasets.
+The standard ROC and AUC calculations are defined only for binary classifications, which is not the case for our eight species and \textit{Emys} complex datasets.
 To generalize this approach for situations with multiple response classes, we used an all-against-one strategy where the model AUC is the average of the AUC values from the multiple binary comparisons of one class compared to all others \citep{Hand2001}.
@@
 
-The results of the PCA of plastron shape in both the seven species and \textit{Trachemys} datasets demonstrate strong association between shape and the recognized classification schemes (Fig.
+The results of the PCA of plastron shape in both the eight species and \textit{Trachemys} datasets demonstrate strong association between shape and the recognized classification schemes (Fig.
 \ref{fig:other_pca}).
@@
   \caption{Two scatterplots of morphological differences from two of the three datasets analyzed in this study.
-(a) Scatterplot of the first two PCA axies from the landmarks from the seven different species dataset, and (b) the first two axes of variation from two subspecies of \textit{Trachemys} dataset.
+(a) Scatterplot of the first two PCA axies from the landmarks from the eight different species dataset, and (b) the first two axes of variation from two subspecies of \textit{Trachemys} dataset.
 Point colors correspond to the categories within each dataset while point size is proportional to individual centroid size.
@@
 
-The results from the seven species and \textit{Trachemys} datasets indicate that both of these classification schemes are more recognizable than not given our estimate of digitization error (Table \ref{tab:rep_res}).
+The results from the eight species and \textit{Trachemys} datasets indicate that both of these classification schemes are more recognizable than not given our estimate of digitization error (Table \ref{tab:rep_res}).
 In contrast, the different \textit{E.
@@
   \caption{Results from the within-individual to between-individual Procrustes distances for the replicated plastron shape data.
-Results are presented for all three datasets analyzed here: the \textit{Trachemys} dataset, the seven species dataset, and each of the \textit{Emys marmorata} classification schemes.}
+Results are presented for all three datasets analyzed here: the \textit{Trachemys} dataset, the eight species dataset, and each of the \textit{Emys marmorata} classification schemes.}
   \begin{tabular}{l l c c}
@@
 
-Analysis of the seven morphologically- and genetically-distinct species and the \textit{T.
+Analysis of the eight morphologically- and genetically-distinct species and the \textit{T.
 scripta scripta}--\textit{T.
@@
   \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/other_model_sel}
-  \caption{Comparisons of model fit to the training dataset for each of the supervised learning methods applied to the first two datasets; the results from the seven species dataset are presented in the left column, while those from the \textit{Trachemys} dataset are presented in the right column.
+  \caption{Comparisons of model fit to the training dataset for each of the supervised learning methods applied to the first two datasets; the results from the eight species dataset are presented in the left column, while those from the \textit{Trachemys} dataset are presented in the right column.
 Models were fit to datasets of varying complexity, with the number of parameters listed along the x-axis.
@@
   \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/other_oos_sel}
-  \caption{The results of out-of-sample predictive performance of the selected models for both the seven species (left) and \textit{Trachemys} datasets.
+  \caption{The results of out-of-sample predictive performance of the selected models for both the eight species (left) and \textit{Trachemys} datasets.
 Predictive performance is measured as the area under the receiver operating characteristic (AUC), which ranges from 0.5 to 1.
@@
 marmorata} dataset is so great as to swamp out any biological signal.
-We think this is unlikely because all of the specimens considered in our three analyses were digitized by one of us (K.D.A.), and digitization error was not a problem in the seven species or \textit{Trachemys} examples.
+We think this is unlikely because all of the specimens considered in our three analyses were digitized by one of us (K.D.A.), and digitization error was not a problem in the eight species or \textit{Trachemys} examples.
 There are also no features of the plastron of \textit{E.
@@
 marmorata}, it is important to note that plastron shape is an extremely effective method for differentiating classes in the additional datasets we investigated.
-The magnitude of shape differences between the species (measured as Procrustes distance between the seven species' mean shapes) is approximately an order of magnitude greater than the differences between the \textit{E.
+The magnitude of shape differences between the species (measured as Procrustes distance between the eight species' mean shapes) is approximately an order of magnitude greater than the differences between the \textit{E.
 marmorata} subgroups, and not surprisingly the machine learning methods had no trouble classifying the specimens correctly.

### 2017-12-05 doc/letter_editor.tex
submitted

@@
 \signature{Peter D Smits}
-\address{Committee on Evolutionary Biology \\ University of Chicago \\
-1025 E. 57th Street \\ Culver Hall 402 \\ Chicago, IL 60637 \\
-psmits@uchicago.edu}
+\address{Department of Integrative Biology\\ University of California Berkeley\\
+3040 Valley Life Sciences Bldg \#5050\\ Berkeley, CA 94720 \\
+psmits@berkeley.edu}
 
@@
 
-  Please find enclosed the manuscript entitled ``How cryptic is cryptic diversity?
-Machine learning approaches to classifying morphological variation in the Pacific Pond Turtle (\textit{Emys marmorata})'' which we are submitting for consideration at the journal \textit{Systematic Biology}.
-This manuscript is authored by myself (Peter D Smits), Kenneth D Angielczyk, and James F Parham.
+  Please find enclosed the manuscript entitled ``Ensemble approaches for estimating congruence between species delimitation and morphological variation: comparing taxonomic hypotheses for the Pacific Pond Turtle (\textit{Emys marmorata})'' which we are submitting for consideration at the journal \textit{Systematic Biology}.
+This manuscript is authored by myself (Peter D Smits), Kenneth D Angielczyk, James F Parham, and Bryan L Stuart.
 We believe our analysis and results are of interest to the community as we present a non-standard approach for evaluating the intensity of cryptic diversity and the efficacy of different classification schemes.
 
-  Morphological and molecular attempts to classify the turtle species \textit{Emys marmorata} have produced three different potential sub-species or species divisions.
-In this study we compared these hypotheses of how to subdivide \textit{Emys marmorata} using multiple machine learning approaches.
-We analyzed over 500 specimens of this species from over its entire geographic range.
-Additionally, instead of classification accuracy, we use a metric of the ratio of true-positive and false-positive rates for both in-sample and out-of-sample model performance.
-Finally, in order to test whether our approaches were valid, we compared these results to those from the analysis of seven morphologically distinct turtle species.
+  We investigated the morphometric identification of cryptic species using machine learning approaches by examining their implications for a recently proposed cryptic turtle species (\textit{Emys pallida}).
+Our dataset consists of geometric morphometric data collected from  532 adult \textit{E.
+marmorata/``E.
+pallida''} museum specimens.
+Each specimen was then assigned a classification for six different binning schemes based on geographic occurrence data recorded in museum collection archives.
+We analyzed our data using an ensemble of supervised machine learning approaches to determine which classification hypothesis was best supported by the data.
+In order to veryfiy our approach, we applied the same analysis to two alternate turtle morphology datasets, one consisting of eight unambiguously distinct species closely related to \textit{E.
+marmorata}, and the other consisting of two subspecies of \textit{Trachemys scripta}.
 
-  Our results demonstrate, that while our approaches can very precisely identify morphologically distinct groupings, the morphological diversity of the \textit{Emy marmorata} species ``complex'' is inconsistent with any of the proposed classification schemes.
-We propose that these results may be due to the possibility that genetic differentiation is not associated with plastron shape variation \textit{below} the species level and/or that local selective pressures (e.g., from hydrological regime) overwhelm morphological differentiation.
-Additionally, a reconsideration of the methods used to delimit \textit{E.
-``pallida,''} the lack of barriers to gene flow, the strong evidence for widespread admixture between lineages, and the fact that plastron shape can be used to differentiate other emydine species suggest that its lack of diagnosability most likely reflects the non-distinctiveness of this proposed taxon.
+  Our results indicate that there is no clear ``best'' grouping of \textit{E.
+marmorata/``E.
+pallida''} based on plastron shape.
+In contrast, the analyses of the clear-cut examples produced near perfect classifications, demonstrating that the methods can recover correct results when an appropriate signal exists.
+Explanations for the lack of grouping in \textit{E.
+marmorata} include the possibility that genetic differentiation is not associated with plastron shape variation below the species level and/or that local selective pressures (e.g., from hydrological regime) overwhelm morphological differentiation.
+A reconsideration of the methods used to delimit \textit{``E.
+pallida,''} the lack of barriers to gene flow, the strong evidence for widespread admixture between lineages, and the fact that plastron shape can be used to delineate other emydine species and sub-species suggest that its lack of diagnosability most likely reflects the non-distinctiveness of this proposed taxon.
 
-  Thank you for considering our work. Please send all correspondence regarding this manuscript to me via my email address (psmits@uchicago.edu).
+  An earlier version of this manuscript was submitted and subsequently rejected with an encouragement to resubmit (USYB-2015-207).
+Since this decision this study has undergone substatial revisions, the inclusion of new analysis, and is overall considerably different from the manuscript submitted in 2015.
+Because of these major chnages and the resubmission option expiring in October 2016, we are resubmitting this manuscript as a new article.
+
+  Thank you for considering our work. Please send all correspondence regarding this manuscript to me via my email address (psmits@berkeley.edu).
 
@@
 
-  \encl{Article; supplementary text, figures, tables.}
+  \encl{Article; figures, tables.}
 

### 2017-12-05 doc/turtles_statistical_learning.tex
submitted

@@
 \usepackage[font=small]{caption}
+\usepackage{float}
 \usepackage{subcaption, multirow, morefloats}
-\usepackage{subcaption, wrapfig}
+\usepackage{, wrapfig}
+\usepackage{afterpage}
 \usepackage{rotating}
@@
 \author[1]{Peter D Smits}%\thanks{peterdavidsmits@gmail.com}}
-\author[2,3]{Kenneth D Angielczyk}%\thanks{kangielczyk@fieldmuseum.org}}
-\author[4]{James F Parham}%\thanks{jparham@fullerton.edu}}
-\author[5]{Bryan L Stuart}%\thanks{bryan.stuart@naturalsciences.org}}
+\author[2]{Kenneth D Angielczyk}%\thanks{kangielczyk@fieldmuseum.org}}
+\author[3]{James F Parham}%\thanks{jparham@fullerton.edu}}
+\author[4]{Bryan L Stuart}%\thanks{bryan.stuart@naturalsciences.org}}
 \affil[1]{Department of Integrative Biology, University of California -- Berkeley}
-\affil[2]{Committee on Evolutionary Biology, University of Chicago}
-\affil[3]{Integrative Research Center, Field Museum of Natural History}
-\affil[4]{John D. Cooper Archaeological and Paleontological Center, Department of Geological Sciences, California State University, Fullerton}
-\affil[5]{Section of Research and Collections, North Carolina Museum of Natural Sciences}
+\affil[2]{Integrative Research Center, Field Museum of Natural History}
+\affil[3]{John D. Cooper Archaeological and Paleontological Center, Department of Geological Sciences, California State University, Fullerton}
+\affil[4]{Section of Research and Collections, North Carolina Museum of Natural Sciences}
 
@@
 \maketitle
-\noindent{\textbf{Corresponding author:} Peter D Smits, Department of Integrative Biology, University of California -- Berkeley, 3040 Valley Life Sciences Building \#5151, Berkeley, CA, 94720, USA; E-mail: \href{mailto:peterdavidsmits@gmail.com}{peterdavidsmits@gmail.com}}
+\noindent{\textbf{Corresponding author:} Peter D Smits, Department of Integrative Biology, University of California -- Berkeley, 3040 Valley Life Sciences Building \#5151, Berkeley, CA, 94720, USA; E-mail: \href{mailto:psmits@berkeley.edu}{psmits@berkeley.edu}
 
@@
 Most such studies focus on using morphometrics to discover differences between taxa that were identified by other means \citep{Polly2003,Zelditch2004,Gaubert2005b,Gunduz2007,Polly2007a,Demandt2009,Markolf2013,Fruciano2016}.
-Additionally, there has been work on automated taxon identification and classification of taxa into groups \citep{Baylac2003,Dobigny2003,MacLeod2007,VandenBrink2011,Vitek2017}.
+Additionally, there has been work on automated taxon identification and classification of taxa into groups \citep{Baylac2003,Dobigny2003,MacLeod2007,VandenBrink2011,Vitek2017}, as well as the development of models that combine genetic, phenotypic, and geographic data to infer evolutionary units of interest \citep{Guillot2012}.
 
@@
 
-Geometric morphometric approaches to identifying differences in morphological variation between classes, including cryptic species, have mostly relied on methods like linear discriminate analysis and canonical variates analysis \citep{Polly2003,Zelditch2004,Gaubert2005b,Gunduz2007,Polly2007a,Francoy2009,Sztencel-Jabonka2009,MitrovskiBogdanovic2013,Dillard2017}.
+Geometric morphometric approaches to identifying differences in morphological variation between classes, including cryptic species, have mostly relied on methods like linear discriminate analysis and canonical variates analysis \citep{Polly2003,Zelditch2004,Gaubert2005b,Gunduz2007,Polly2007a,Francoy2009,Sztencel-Jabonka2009,Edwards2011,MitrovskiBogdanovic2013,Dillard2017}.
 Because of their similarity to multivariate approaches like principal components analysis (PCA), these methods are comparatively straightforward ways of understanding the differences in morphology between classes.
@@
 We analyzed the problem of whether there are distinct subspecies or cryptic species within the western pond turtle, \textit{Emys marmorata} \citep{Baird1852} (formerly \emph{Clemmys marmorata}; see \citealp{Feldman2002}).
-\textit{Emys marmorata} is distributed from northern Washington State, USA to Baja California, Mexico.
+\textit{Emys marmorata} is distributed from northern Washington State, USA to Baja California, Mexico; populations in western Nevada may have been introduced by recent human activity or they could be a genuine part of the species' range \citep{Bury2017}.
 Traditionally, \textit{E.
@@
 \subsection{Specimens, sampling, morphometrics}
+
+\afterpage{\clearpage}
+\begin{figure}[ht]
+  \centering
+  \includegraphics[height = 0.8\textheight, width = \textwidth, keepaspectratio = true]{figure/Ken_Ang_SpLoc_final_comp}
+  \caption{Geographic distribution of specimens sampled for comparing the hypothesized subdivisions of \textit{Emys marmorata}.
+Each hypothesized scheme has two or more possible classes.
+Sample size differs between schemes because of our ability to confidently assign museum specimens to the various schemes.
+The number of localities shown on each map is less than the number of specimens sampled because some localities produced multiple specimens.
+The different classification abbreviations are as follows: \textit{E.
+marmorata} = ``marm'', \textit{E.
+pallida} = ``pall'', Central Coast Ranges = ``CCR'', San Joaquin Valley = ``SJ,'' Baya Peninsula = ``Baja,'', and Sierra Foothills = ``Foothill.''}
+  \label{fig:map}
+\end{figure}
+
+\afterpage{\clearpage}
+\begin{figure}[ht]
+  \centering
+  \includegraphics[height = 0.5\textheight, width = \textwidth, keepaspectratio = true]{figure/plastra}
+  \caption{Depiction of general plastral shape of \textit{E. marmorata} and position of the 19 landmarks used in this study. Anterior is towards the top of the figure.}
+  \label{fig:plastra}
+\end{figure}
+
 Three different geometric morphometric datasets describing turtle plastron variation were assembled for this analysis: 1) specimens from eight distinct emydine species; 2) \textit{T.
@@
 The first two datasets are intended to serve as a test of whether machine learning techniques can differentiate species-level groupings of emydine turtles using plastron shape.
-We expect that the first case represents a low complexity dataset because of the high level of plastron shape disparity that exists among these species \citep{Angielczyk2011}, whereas the second dataset should be relatively higher in complexity and more analogous to the \textit{E.
+We expect that the first case represents a low complexity dataset because of the high level of plastron shape disparity that exists among these species \citep{Claude2003a,Claude2006,Angielczyk2011}, whereas the second dataset should be relatively higher in complexity and more analogous to the \textit{E.
 marmorata} example.
@@
 
-Sex was known only for a subset of the total dataset and was not included as a predictor of classification.
-Instead, we estimated the degree by which specimens cluster morphologically by sex in order to determine how much of a potential biasing factor sexual dimorphism could be for our analysis of the \textit{E.
-marmorata} species complex (see below).
-%The scheme names are as follows: Mito 1 and 2 correspond to \citet{Spinks2005}, Mito 3 corresponds to \citet{Spinks2010}, Morph 1 and Morph 2 correspond to \citet{Holland1992}, and Nuclear corresponds to \citet{Spinks2014}.
-% CHANGE
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[height = 0.8\textheight, width = \textwidth, keepaspectratio = true]{figure/Ken_Ang_SpLoc_final_comp}
-  \caption{Geographic distribution of specimens sampled for comparing the hypothesized subdivisions of \textit{Emys marmorata}.
-Each hypothesized scheme has two or more possible classes.
-Sample size differs between schemes because of our ability to confidently assign museum specimens to the various schemes.
-The different classification abbreviations are as follows: \textit{E.
-marmorata} = ``marm'', \textit{E.
-pallida} = ``pall'', Central Coast Ranges = ``CCR'', San Joaquin Valley = ``SJ,'' Baya Peninsula = ``Baja,'', and Sierra Foothills = ``Foothill.''}
-  \label{fig:map}
-\end{figure}
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[height = 0.5\textheight, width = \textwidth, keepaspectratio = true]{figure/plastra}
-  \caption{Depiction of general plastral shape of \textit{E. marmorata} and position of the 19 landmarks used in this study. Anterior is towards the top of the figure.}
-  \label{fig:plastra}
-\end{figure}
-
 \begin{table}
@@
 \end{table}
+
+Sex was known only for a subset of the total dataset and was not included as a predictor of classification.
+Instead, we estimated the degree by which specimens cluster morphologically by sex in order to determine how much of a potential biasing factor sexual dimorphism could be for our analysis of the \textit{E.
+marmorata} species complex (see below).
+%The scheme names are as follows: Mito 1 and 2 correspond to \citet{Spinks2005}, Mito 3 corresponds to \citet{Spinks2010}, Morph 1 and Morph 2 correspond to \citet{Holland1992}, and Nuclear corresponds to \citet{Spinks2014}.
+% CHANGE
 
@@
 \subsection{Supervised learning approaches}
-Instead of relying on a single supervised learning method, we chose to use an ensemble approach where multiple model types are used in concert so that any congruence between them increases our support for that conclusion over another \citep{Hastie2009}.
-The supervised learning methods used here are named in Table \ref{tab:methods}.
-Each of these methods makes different assumptions, treats data differently, and can produce different classification results depending on the nature of the data \citep{Hastie2009}.
-For example, multinomial logistic regression is a type of generalized linear model, whereas random forest is itself an ensemble approach where multiple decision trees are fit to subsets of the full dataset and then averaged.
-% use multiple methods
-%   multinomial logistic regression (nnet)
-%   linear discriminate analysis (MASS)
-%   penalized discriminate analysis (mda)
-%   neural network (nnet)
-%   random forest (randomForest)
-% each method has different assumptions and treat data differently
-%   all assume that predictors have additive effect (e.g. independent)
 
@@
 \end{table}
+
+Instead of relying on a single supervised learning method, we chose to use an ensemble approach where multiple model types are used in concert so that any congruence between them increases our support for that conclusion over another \citep{Hastie2009}.
+The supervised learning methods used here are named in Table \ref{tab:methods}.
+Each of these methods makes different assumptions, treats data differently, and can produce different classification results depending on the nature of the data \citep{Hastie2009}.
+For example, multinomial logistic regression is a type of generalized linear model, whereas random forest is itself an ensemble approach where multiple decision trees are fit to subsets of the full dataset and then averaged.
+% use multiple methods
+%   multinomial logistic regression (nnet)
+%   linear discriminate analysis (MASS)
+%   penalized discriminate analysis (mda)
+%   neural network (nnet)
+%   random forest (randomForest)
+% each method has different assumptions and treat data differently
+%   all assume that predictors have additive effect (e.g. independent)
 
@@
 
-The results of the PCA of plastron shape in both the eight species and \textit{Trachemys} datasets demonstrate strong association between shape and the recognized classification schemes (Fig.
-\ref{fig:other_pca}).
-
-\begin{figure}[ht]
+\begin{figure}[h]
   \centering
@@
 
-The results of the PCA of plastron shape in the \textit{Emys marmorata} dataset show no clear connection between plastron shape and any of the proposed classification schemes (Fig.
-\ref{fig:emys_pca}).
-The first PC axis of shape variation appears to be primarily structured by differences in individual centroid size (Fig.
-\ref{fig:emys_pca}); this was the motivation for including centroid size and its interaction with PC1 as predictors in all of the supervised learning models.
-
+\afterpage{\clearpage}
 \begin{figure}[ht]
@@
 
-Analysis of the differences between sexes of \textit{E.
-marmorata} indicates that sex does not appear to strongly structure differences in shape (Fig.
-\ref{fig:sex_test}).
-The difference in mean shape between the sexes is very small; the sexes overlap about has much as expected given a null distribution based on permuting the sex-labels.
+The results of the PCA of plastron shape in both the eight species and \textit{Trachemys} datasets demonstrate strong association between shape and the recognized classification schemes (Fig.
+\ref{fig:other_pca}).
 
-\begin{figure}[ht]
+The results of the PCA of plastron shape in the \textit{Emys marmorata} dataset show no clear connection between plastron shape and any of the proposed classification schemes (Fig.
+\ref{fig:emys_pca}).
+The first PC axis of shape variation appears to be primarily structured by differences in individual centroid size (Fig.
+\ref{fig:emys_pca}); this was the motivation for including centroid size and its interaction with PC1 as predictors in all of the supervised learning models.
+
+\begin{figure}[h]
   \centering
@@
 
-Comparison of the within to between Procrustes distances of the digitization replicates gives an approximate estimate of the error between distinct groupings (Table \ref{tab:rep_res}).
-The ratio of the average within-individual distance to the average distance between individuals for the replicated datasets is 1.11; this indicates that the grouping is slightly counter-intuitive to the data and is consistent with all shapes being very similar regardless of individual identity.
-This value also provides a baseline by which to understand how distinct the groupings are, where other ratios are compared to the correction ratio \(1.11/1\).
-
-The results from the eight species and \textit{Trachemys} datasets indicate that both of these classification schemes are more recognizable than not given our estimate of digitization error (Table \ref{tab:rep_res}).
-In contrast, the different \textit{E.
-marmorata} classification schemes appear to barely be distinct, with their within:between ratios approximating 1.
-This indicates that the magnitude of the differences between groupings is approximately the same as the difference between any two random individuals (Table \ref{tab:rep_res}).
+Analysis of the differences between sexes of \textit{E.
+marmorata} indicates that sex does not appear to strongly structure differences in shape (Fig.
+\ref{fig:sex_test}).
+The difference in mean shape between the sexes is very small; the sexes overlap about has much as expected given a null distribution based on permuting the sex-labels.
 
@@
 
+Comparison of the within to between Procrustes distances of the digitization replicates gives an approximate estimate of the error between distinct groupings (Table \ref{tab:rep_res}).
+The ratio of the average within-individual distance to the average distance between individuals for the replicated datasets is 1.11; this indicates that the grouping is slightly counter-intuitive to the data and is consistent with all shapes being very similar regardless of individual identity.
+This value also provides a baseline by which to understand how distinct the groupings are, where other ratios are compared to the correction ratio \(1.11/1\).
+
+The results from the eight species and \textit{Trachemys} datasets indicate that both of these classification schemes are more recognizable than not given our estimate of digitization error (Table \ref{tab:rep_res}).
+In contrast, the different \textit{E.
+marmorata} classification schemes appear to barely be distinct, with their within:between ratios approximating 1.
+This indicates that the magnitude of the differences between groupings is approximately the same as the difference between any two random individuals (Table \ref{tab:rep_res}).
+
 \subsection{Supervised learning}
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[height = 0.5\textheight, width = \textwidth, keepaspectratio = true]{figure/other_model_sel}
+  \caption{Comparisons of model fit to the training dataset for each of the supervised learning methods applied to the first two datasets; the results from the eight species dataset are presented in the left column, while those from the \textit{Trachemys} dataset are presented in the right column.
+Models were fit to datasets of varying complexity, with the number of parameters listed along the x-axis.
+Model fit is measured as the area under the receiver operating characteristic (AUC), which ranges from 0.5 to 1.
+Error bars correspond to one standard error estimated from 10 rounds of 5-fold cross-validation.
+The red dot corresponds to the model fit with the highest mean AUC while the blue dot corresponds to the model selected for further analysis.
+In some cases, there is no difference in complexity between the best and selected models.}
+  \label{fig:other_sel}
+\end{figure}
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[height = 0.3\textheight, width = \textwidth, keepaspectratio = true]{figure/other_oos_sel}
+  \caption{The results of out-of-sample predictive performance of the selected models for both the eight species (left) and \textit{Trachemys} datasets.
+Predictive performance is measured as the area under the receiver operating characteristic (AUC), which ranges from 0.5 to 1.
+Points correspond to the individual out-of-sample predictive performance of the specific model, indicated along the x-axis.
+The horizontal bars correspond the average out-of-sample predictive performance of all the models.}
+  \label{fig:other_oos}
+\end{figure}
 
@@
 
-\begin{figure}[ht]
-  \centering
-  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/other_model_sel}
-  \caption{Comparisons of model fit to the training dataset for each of the supervised learning methods applied to the first two datasets; the results from the eight species dataset are presented in the left column, while those from the \textit{Trachemys} dataset are presented in the right column.
-Models were fit to datasets of varying complexity, with the number of parameters listed along the x-axis.
-Model fit is measured as the area under the receiver operating characteristic (AUC), which ranges from 0.5 to 1.
-Error bars correspond to one standard error estimated from 10 rounds of 5-fold cross-validation.
-The red dot corresponds to the model fit with the highest mean AUC while the blue dot corresponds to the model selected for further analysis.
-In some cases, there is no difference in complexity between the best and selected models.}
-  \label{fig:other_sel}
-\end{figure}
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/other_oos_sel}
-  \caption{The results of out-of-sample predictive performance of the selected models for both the eight species (left) and \textit{Trachemys} datasets.
-Predictive performance is measured as the area under the receiver operating characteristic (AUC), which ranges from 0.5 to 1.
-Points correspond to the individual out-of-sample predictive performance of the specific model, indicated along the x-axis.
-The horizontal bars correspond the average out-of-sample predictive performance of all the models.}
-  \label{fig:other_oos}
-\end{figure}
-
-AUC--based model selection revealed some important patterns of variation and congruence between the classification schemes and the actual data.
-Generally, the best performing models tended to include about half the total number of possible PCs (Fig.
-\ref{fig:emys_sel}).
-
+\afterpage{\clearpage}
 \begin{figure}[ht]
@@
 
-Observed AUC values for all of the optimal models are lower for the \textit{E.
-marmorata} dataset than for the other two datasets (Fig.
-\ref{fig:other_sel}, \ref{fig:emys_sel}).
-In most cases the different proposed classification schemes are generally poor descriptors of the observed variation.
-It appears that the dataset is overwhelmed by noise (likely biological and analytical), making any accurate classifications difficult at best.
-This observation is cemented with the generalizations of the models to the testing dataset (Fig.
-\ref{fig:emys_oos}).
-
+\afterpage{\clearpage}
 \begin{figure}[ht]
@@
 \end{figure}
+
+AUC--based model selection revealed some important patterns of variation and congruence between the classification schemes and the actual data.
+Generally, the best performing models tended to include about half the total number of possible PCs (Fig.
+\ref{fig:emys_sel}).
+
+Observed AUC values for all of the optimal models are lower for the \textit{E.
+marmorata} dataset than for the other two datasets (Fig.
+\ref{fig:other_sel}, \ref{fig:emys_sel}).
+In most cases the different proposed classification schemes are generally poor descriptors of the observed variation.
+It appears that the dataset is overwhelmed by noise (likely biological and analytical), making any accurate classifications difficult at best.
+This observation is cemented with the generalizations of the models to the testing dataset (Fig.
+\ref{fig:emys_oos}).
 
@@
 Sweet for field assistance and the California Department of Fish and Game for permits.
-We would also like to thank Marc Lambruschi (FMNH) for help with figure \ref{fig:map}.
+We would also like to thank M.
+Lambruschi (FMNH) for help with figure \ref{fig:map}.
 

### 2018-01-23 doc/letter_editor.tex
rejected from bjls. reformatted for peerj

@@
 \begin{document}
-\begin{letter}{Editor \\ \textit{Systematic Biology}}
+\begin{letter}{Editor \\ \textit{Biological Journal of the Linnean Society}}
   \opening{Dear Editor,}
 
-  Please find enclosed the manuscript entitled ``Ensemble approaches for estimating congruence between species delimitation and morphological variation: comparing taxonomic hypotheses for the Pacific Pond Turtle (\textit{Emys marmorata})'' which we are submitting for consideration at the journal \textit{Systematic Biology}.
+  Please find enclosed the manuscript entitled ``Ensemble approaches for estimating congruence between species delimitation and morphological variation: comparing taxonomic hypotheses for the Pacific Pond Turtle (\textit{Emys marmorata})'' which we are submitting for consideration for the \textit{Biological Journal of the Linnean Society}.
 This manuscript is authored by myself (Peter D Smits), Kenneth D Angielczyk, James F Parham, and Bryan L Stuart.
@@
 
-  An earlier version of this manuscript was submitted and subsequently rejected with an encouragement to resubmit (USYB-2015-207).
+  %An earlier version of this manuscript was submitted and subsequently rejected with an encouragement to resubmit (USYB-2015-207).
 Since this decision this study has undergone substatial revisions, the inclusion of new analysis, and is overall considerably different from the manuscript submitted in 2015.

### 2018-01-23 doc/tsl_amnat.tex
rejected from bjls. reformatted for peerj

--- 2017-12-05 doc/turtles_statistical_learning.tex
@@
-\documentclass[12pt,letterpaper]{article}
-
+\documentclass[11pt]{article}
+\usepackage[sc]{mathpazo} %Like Palatino with extensive math support
+\usepackage{fullpage}
+\usepackage[authoryear,sectionbib,sort]{natbib}
+\linespread{1.7}
+\usepackage[utf8]{inputenc}
+\usepackage{lineno}
 \usepackage{amsmath, amsthm}
@@
 \usepackage{microtype, parskip}
-\usepackage{natbib}
-\usepackage{lineno}
 \usepackage[font=small]{caption}
@@
 \usepackage{subcaption, multirow, morefloats}
-\usepackage{, wrapfig}
+\usepackage{wrapfig}
 \usepackage{afterpage}
 \usepackage{rotating}
-\usepackage{titlesec}
-\usepackage[nottoc,numbib]{tocbibind}
-\usepackage{authblk, attrib, fullpage}
 \usepackage{lineno}
 
-\frenchspacing
+%%%%%%%%%%%%%%%%%%%%%
+% LaTeX packages
+%%%%%%%%%%%%%%%%%%%%%
+% Please be sparing in your use of additional LaTeX packages, and
+% upload any required style files to Editorial Manager with the file
+% type "LaTeX ancillary files (.sty, .bst)."
 
-\captionsetup[subfigure]{position = top, labelfont = bf, textfont = normalfont, singlelinecheck = off, justification = raggedright}
+%%%%%%%%%%%%%%%%%%%%%
+% Line numbering
+%%%%%%%%%%%%%%%%%%%%%
+\usepackage{lineno}
+% Please use line numbering with your initial submission and
+% subsequent revisions. After acceptance, please comment out 
+% the commands \usepackage{lineno}, \linenumbers{} 
+% and \modulolinenumbers[3] below.
 
-\renewcommand{\section}[1]{%
-\bigskip
-\begin{center}
-\begin{Large}
-\normalfont\scshape #1
-\medskip
-\end{Large}
-\end{center}}
+\title{Ensemble approaches for estimating congruence between species delimitation and morphological variation: comparing taxonomic hypotheses for the Pacific Pond Turtle (\textit{Emys marmorata})}
 
-\renewcommand{\subsection}[1]{%
-\bigskip
-\begin{center}
-\begin{large}
-\normalfont\itshape #1
-\end{large}
-\end{center}}
+%%%%%%%%%%%%%%%%%%%%%
+% Authorship
+%%%%%%%%%%%%%%%%%%%%%
+% Please remove authorship information while your paper is under review,
+% unless you wish to waive your anonymity under double-blind review. 
+% Remember to uncomment the information after acceptance.
 
-\renewcommand{\subsubsection}[1]{%
-\vspace{2ex}
-\noindent
-\textit{#1.}---}
+\author{Peter D Smits$^{1,\ast,\dag}$ \\ 
+Kenneth D Angielczyk$^{2}$ \\ 
+Bryan L Stuart$^{3}$ \\
+James F Parham$^{4}$}
 
-\renewcommand{\tableofcontents}{}
-
-\bibpunct{(}{)}{;}{a}{}{,}  % this is a citation format command for natbib
-
-%\title{How cryptic is cryptic diversity? Machine learning approaches to classifying morphological variation in the Pacific Pond Turtle (\textit{Emys marmorata})}
-\title{Ensemble approaches for estimating congruence between species delimitation and morphological variation: comparing taxonomic hypotheses for the Pacific Pond Turtle (\textit{Emys marmorata})}
-\author[1]{Peter D Smits}%\thanks{peterdavidsmits@gmail.com}}
-\author[2]{Kenneth D Angielczyk}%\thanks{kangielczyk@fieldmuseum.org}}
-\author[3]{James F Parham}%\thanks{jparham@fullerton.edu}}
-\author[4]{Bryan L Stuart}%\thanks{bryan.stuart@naturalsciences.org}}
-\affil[1]{Department of Integrative Biology, University of California -- Berkeley}
-\affil[2]{Integrative Research Center, Field Museum of Natural History}
-\affil[3]{John D. Cooper Archaeological and Paleontological Center, Department of Geological Sciences, California State University, Fullerton}
-\affil[4]{Section of Research and Collections, North Carolina Museum of Natural Sciences}
+\date{}
 
 \begin{document}
+% The journal does not have numbered sections in the main portion of
+% articles. Please refrain from using section references such as
+% section~\ref{section:CountingOwlEggs}, and refer to sections by name
+% (e.g. section ``Counting Owl Eggs'').
+
+% Please note that we prefer (\citealt{Xiao2015}) to \citep{Xiao2015},
+% since \citep{} inserts a comma after "et al."
+
 \maketitle
-\noindent{\textbf{Corresponding author:} Peter D Smits, Department of Integrative Biology, University of California -- Berkeley, 3040 Valley Life Sciences Building \#5151, Berkeley, CA, 94720, USA; E-mail: \href{mailto:psmits@berkeley.edu}{psmits@berkeley.edu}
 
-\linenumbers
-\modulolinenumbers[2]
+\noindent{}1. Department of Integrative Biology, University of California -- Berkeley, Berkeley, CA, USA;
 
-\begin{abstract}
+\noindent{}2. Integrative Research Center, Field Museum of Natural History, Chicago, IL, USA;
+
+\noindent{}3. Section of Research and Collections, North Carolina Museum of Natural Sciences, Raleigh, NC, USA;
+
+\noindent{}4. John D. Cooper Archaeological and Paleontological Center, Department of Geological Sciences, California State University Fullerton, Fullerton, CA, USA.
+
+\noindent{}$\ast$ Corresponding author; e-mail: psmits@berkeley.edu
+
+\noindent{}$\dag$ ORCIDs: Smits, 0000-0003-4074-1447.
+
+\bigskip
+
+\textit{Manuscript elements}: Figure~1, figure~2, figure~3, figure~4, figure~5, figure~6, figure~7, figure~8, figure~9, table~1, table~2, table~3. Figures 1, 3, 4, 6, 8 are in color.
+
+\bigskip
+
+\textit{Keywords}: Geometric morphometrics, machine learning, conservation, Testudinae.
+
+\bigskip
+
+\textit{Manuscript type}: Article. 
+% Or e-article, note, e-note, natural history miscellany,
+% e-natural history miscellany, comment, reply, symposium, or
+% countdown to 150.
+
+\bigskip
+
+\noindent{\footnotesize Prepared using the suggested \LaTeX{} 
+template for \textit{Am.\ Nat.}}
+
+\linenumbers{}
+\modulolinenumbers[3]
+
+\newpage{}
+
+\section*{Abstract}
 
@@
 pallida,''} the lack of barriers to gene flow, the strong evidence for widespread admixture between lineages, and the fact that plastron shape can be used to delineate other emydine species and sub-species suggest that its lack of diagnosability most likely reflects the non-distinctiveness of this proposed taxon.
-\end{abstract}
 
-\section{Introduction}
-% cryptic diversity
-%   most species are still deliminated solely based on morphology
+\newpage{}
+
+\section*{Introduction}
+
 Molecular systematics has repeatedly demonstrated the existence of cryptic species that can only be diagnosed using genetic data \citep{Stuart2006,Bickford2007,SchlickSteiner2007,Pfenninger2007,Clare2011,Funk2012}.
@@
 % why use machine learning methods
-\subsection{Background and study system}
+\subsection*{Background and study system}
 Machine learning is an extension of known statistical methodology \citep{Hastie2009} that emphasizes predictive accuracy and generality often at the expense of the interpretability of individual parameters.
@@
 
-\section{Materials and Methods}
-\subsection{Specimens, sampling, morphometrics}
-
-\afterpage{\clearpage}
-\begin{figure}[ht]
-  \centering
-  \includegraphics[height = 0.8\textheight, width = \textwidth, keepaspectratio = true]{figure/Ken_Ang_SpLoc_final_comp}
-  \caption{Geographic distribution of specimens sampled for comparing the hypothesized subdivisions of \textit{Emys marmorata}.
-Each hypothesized scheme has two or more possible classes.
-Sample size differs between schemes because of our ability to confidently assign museum specimens to the various schemes.
-The number of localities shown on each map is less than the number of specimens sampled because some localities produced multiple specimens.
-The different classification abbreviations are as follows: \textit{E.
-marmorata} = ``marm'', \textit{E.
-pallida} = ``pall'', Central Coast Ranges = ``CCR'', San Joaquin Valley = ``SJ,'' Baya Peninsula = ``Baja,'', and Sierra Foothills = ``Foothill.''}
-  \label{fig:map}
-\end{figure}
-
-\afterpage{\clearpage}
-\begin{figure}[ht]
-  \centering
-  \includegraphics[height = 0.5\textheight, width = \textwidth, keepaspectratio = true]{figure/plastra}
-  \caption{Depiction of general plastral shape of \textit{E. marmorata} and position of the 19 landmarks used in this study. Anterior is towards the top of the figure.}
-  \label{fig:plastra}
-\end{figure}
+\section*{Methods}
+\subsection*{Specimens, sampling, morphometrics}
 
@@
 
-\begin{table}
-  \centering
-    \caption{Table of species delimitation hypotheses for \textit{E. marmorata}}
-    \begin{tabular}{l l l }
-      \hline
-      Abbreviation & Number of classes & citation \\
-      \hline
-      SP10.1 & 3 & \citet{Spinks2010} \\
-      SP10.2 & 3 & \citet{Spinks2010} \\
-      SP10.3 & 4 & \citet{Spinks2010} \\
-      SP14.1 & 2 & \citet{Spinks2014} \\
-      SP14.2 & 4 & \citet{Spinks2014} \\
-      Morph & 2 & \citet{Spinks2010} \\
-      \hline
-    \end{tabular}
-    \label{tab:hypotheses}
-\end{table}
-
 Sex was known only for a subset of the total dataset and was not included as a predictor of classification.
@@
 
-\subsection{Biasing effects}
+\subsection*{Biasing effects}
 % digitization
@@
 
-\subsection{Supervised learning approaches}
-
-\begin{table}
-  \centering
-  \caption{Table of the supervised learning methods used in this analysis.}
-  \scalebox{0.9}{
-  \begin{tabular}{l l l l}
-    \hline
-    Method name & abbreviation & R package & citation \\
-    \hline
-    multinomial logistic regression & MLR & nnet & \citet{Venables2002} \\
-    linear discriminate analysis & LDA & MASS & \citet{Venables2002} \\
-    penalized discrminiate analysis & PDA & mda & \citet{mdapack} \\
-    single-hidden-layer neural network & NN & nnet & \citet{Venables2002} \\
-    random forests & RF & randomForest & \citet{Liaw2002} \\
-    \hline
-  \end{tabular}}
-  \label{tab:methods}
-\end{table}
+\subsection*{Supervised learning approaches}
 
@@
 
-\section{Results}
+\section*{Results}
 
-\subsection{Geometric morphometrics}
-
-\begin{figure}[h]
-  \centering
-  \begin{subfigure}[b]{0.7\textwidth}
-    \caption{}
-    \includegraphics[width = \textwidth]{figure/cc7_pc_graph}
-  \end{subfigure}
-
-  \begin{subfigure}[b]{0.7\textwidth}
-    \caption{}
-    \includegraphics[width = \textwidth]{figure/tra_pc_graph}
-  \end{subfigure}
-  \caption{Two scatterplots of morphological differences from two of the three datasets analyzed in this study.
-(a) Scatterplot of the first two PCA axies from the landmarks from the eight different species dataset, and (b) the first two axes of variation from two subspecies of \textit{Trachemys} dataset.
-Point colors correspond to the categories within each dataset while point size is proportional to individual centroid size.
-In parentheses next to the axis labels are the percent of total variation accounted for by that axis.
-For both datasets there are clear distinctions between the different categories.}
-  \label{fig:other_pca}
-\end{figure}
-
-\afterpage{\clearpage}
-\begin{figure}[ht]
-  \centering
-  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/emys_pc_graph}
-  \caption{Scatterplot of the first two axes of morphological variation in the \textit{Emys marmorata} dataset.
-Each panel corresponds to one of the six different classification schemes analyzed as part of this study (Tab.
-\ref{tab:hypotheses}).
-Point color corresponds to the categories within each scheme, and the class names correspond to geographic regions.
-Point size is proportional to centroid size of that specimen and the numbers in parentheses next to the axis labels are the percent of total variation accounted for along that axis.}
-  \label{fig:emys_pca}
-\end{figure}
+\subsection*{Geometric morphometrics}
 
@@
 
-\begin{figure}[h]
-  \centering
-  \includegraphics[height = 0.3\textheight, width = \textwidth, keepaspectratio = true]{figure/sex_test_hist}
-  \caption{Comparison of observed Procrustes distance between the centroids of each sex (vertical line) to a null distribution generated from 1000 permutations of the sex-labels.
-This result indicates that the difference between the centroids is as small/smaller than expected by random.}
-  \label{fig:sex_test}
-\end{figure}
-
 Analysis of the differences between sexes of \textit{E.
@@
 The difference in mean shape between the sexes is very small; the sexes overlap about has much as expected given a null distribution based on permuting the sex-labels.
-
-\begin{table}
-  \centering
-  \caption{Results from the within-individual to between-individual Procrustes distances for the replicated plastron shape data.
-Results are presented for all three datasets analyzed here: the \textit{Trachemys} dataset, the eight species dataset, and each of the \textit{Emys marmorata} classification schemes.}
-  \begin{tabular}{l l c c}
-    \textbf{Dataset} & \textbf{Scheme} & \textbf{Ratio} & \textbf{Corrected ratio} \\
-    \hline
-    Replicates & & 1.11 & \\
-    \hline
-    Seven species & & 0.33 & 0.37 \\
-    \textit{Trachemys} & & 0.76 & 0.84 \\
-    \hline
-    \textit{E. marmorata} & SP10.1 & 0.99 & 1.10 \\
-     & SP10.2 & 1.00 & 1.11 \\
-     & SP10.3 & 0.94 & 1.04 \\
-     & SP14.1 & 1.01 & 1.12 \\
-     & SP14.2 & 0.93 & 1.04 \\
-     & Morph & 0.99 & 1.09 \\
-    \hline
-  \end{tabular}
-  \label{tab:rep_res}
-\end{table}
 
@@
 
-\subsection{Supervised learning}
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[height = 0.5\textheight, width = \textwidth, keepaspectratio = true]{figure/other_model_sel}
-  \caption{Comparisons of model fit to the training dataset for each of the supervised learning methods applied to the first two datasets; the results from the eight species dataset are presented in the left column, while those from the \textit{Trachemys} dataset are presented in the right column.
-Models were fit to datasets of varying complexity, with the number of parameters listed along the x-axis.
-Model fit is measured as the area under the receiver operating characteristic (AUC), which ranges from 0.5 to 1.
-Error bars correspond to one standard error estimated from 10 rounds of 5-fold cross-validation.
-The red dot corresponds to the model fit with the highest mean AUC while the blue dot corresponds to the model selected for further analysis.
-In some cases, there is no difference in complexity between the best and selected models.}
-  \label{fig:other_sel}
-\end{figure}
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[height = 0.3\textheight, width = \textwidth, keepaspectratio = true]{figure/other_oos_sel}
-  \caption{The results of out-of-sample predictive performance of the selected models for both the eight species (left) and \textit{Trachemys} datasets.
-Predictive performance is measured as the area under the receiver operating characteristic (AUC), which ranges from 0.5 to 1.
-Points correspond to the individual out-of-sample predictive performance of the specific model, indicated along the x-axis.
-The horizontal bars correspond the average out-of-sample predictive performance of all the models.}
-  \label{fig:other_oos}
-\end{figure}
+\subsection*{Supervised learning}
 
@@
 These results demonstrate that when there are distinctions between the states of the classification schemes (i.e., differences in plastron shape that correlate with the different taxonomic groups), the methods used here can recover them.
-
-\afterpage{\clearpage}
-\begin{figure}[ht]
-  \centering
-  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/emys_model_sel}
-  \caption{AUC values for models of varying complexity fit to the \textit{Emys marmorata} training datasets for each classification scheme.
-The x-axis corresponds to the total number of predictors included in each model, while the y-axis corresponds to the AUC value which is a measure of goodness of fit for classification datasets.
-A model with a high AUC value corresponds to better classification performance than a model with a lower AUC value.
-Standard errors on AUC estimates are calculated from 10 rounds of 5-fold cross-validation.
-Indicated are the best performing and the selected models, in red and blue respectively.
-In some cases, there is no difference in complexity between the best and selected models.}
-  \label{fig:emys_sel}
-\end{figure}
-
-\afterpage{\clearpage}
-\begin{figure}[ht]
-  \centering
-  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/emys_oos_sel}
-  \caption{Comparison of out-of-sample AUC estimates from the predictions of selected models (Fig.
-\ref{fig:emys_sel}), grouped by classification scheme.
-The horizontal line in each panel corresponds to the average AUC value across all models of that classification scheme.}
-  \label{fig:emys_oos}
-\end{figure}
 
@@
 
-\section{Discussion}
-
+\section*{Discussion}
 As expected, our ensemble approach yields high out-of-sample classification performance for the first two datasets.
@@
 
-\subsection{Is there more than one species of Western Pond Turtle?}
+\subsection*{Is there more than one species of Western Pond Turtle?}
 
@@
 
-\section*{Acknowledgements}
-Data collection for this project was supported in part by NSF DBI-0306158 (to KDA).
+%\section*{Conclusion}
+
+%%%%%%%%%%%%%%%%%%%%%
+% Acknowledgments
+%%%%%%%%%%%%%%%%%%%%%
+% You are encouraged to remove the Acknowledgments section while
+% your paper is under review (unless you wish to waive your anonymity
+% under double-blind review) if the Acknowledgments reveal your
+% identity. If you remove this section, you will need to add it back
+% in to your final files after acceptance.
+
+%\section*{Acknowledgments}
+%Data collection for this project was supported in part by NSF DBI-0306158 (to KDA).
 G.
@@
 
-\bibliographystyle{abbrvnat}
-\bibliography{turtle,packages}
+\newpage{}
+
+%%%%%%%%%%%%%%%%%%%%%
+% Bibliography
+%%%%%%%%%%%%%%%%%%%%%
+% You can either type your references following the examples below, or
+% compile your BiBTeX database and paste the contents of your .bbl file
+% here. The amnatnat.bst style file should work for this---but please
+% email the journal office at amnat at uchicago dot edu if you run into
+% any hitches with it!
+% The list below includes sample journal articles, book chapters, and
+% Dryad references.
+
+%\bibliographystyle{amnatnat}
+%\bibliography{turtle,packages}
+\begin{thebibliography}{86}
+\providecommand{\natexlab}[1]{#1}
+
+\bibitem[{Angielczyk and Feldman(2013)}]{Angielczyk2013a}
+Angielczyk, K.~D., and C.~R. Feldman. 2013.
+\newblock {Are diminutive turtles miniaturized? The ontogeny of plastron shape
+  in emydine turtles}.
+\newblock Biological Journal of the Linnean Society 108:727--755.
+
+\bibitem[{Angielczyk et~al.(2011)Angielczyk, Feldman, and
+  Miller}]{Angielczyk2011}
+Angielczyk, K.~D., C.~R. Feldman, and G.~R. Miller. 2011.
+\newblock {Adaptive evolution of plastron shape in emydine turtles.}
+\newblock Evolution 65:377--394.
+
+\bibitem[{Angielczyk and Sheets(2007)}]{Angielczyk2007}
+Angielczyk, K.~D., and H.~D. Sheets. 2007.
+\newblock {Investigation of simulated tectonic deformation in fossils using
+  geometric morphometrics}.
+\newblock Paleobiology 33:125--148.
+
+\bibitem[{Arnqvist and M{\aa}rtensson(1998)}]{Arnqvist1998}
+Arnqvist, G., and T.~M{\aa}rtensson. 1998.
+\newblock {Measurement error in geometric morphometrics: Empirical strategies
+  to assess and reduce its impact on measures of shape}.
+
+\bibitem[{Baird and Girard(1852)}]{Baird1852}
+Baird, S.~F., and C.~Girard. 1852.
+\newblock {Descriptions of new species of reptiles collected by the U.S.
+  Exploring Expedition under the command of Capt. Charles Wilkes}.
+\newblock Proceedings of the National Academy of Sciences Philadelphia
+  6:174--177.
+
+\bibitem[{Bauer et~al.(2000)Bauer, Parham, Brown, Stuart, Grismer, Papenfuss,
+  Bohme, Savage, Carranza, Grismer, Wagner, Schmitz, Ananjeva, and
+  Inger}]{Bauer2000}
+Bauer, A.~M., J.~F. Parham, R.~M. Brown, B.~L. Stuart, L.~Grismer, T.~J.
+  Papenfuss, W.~Bohme, J.~M. Savage, S.~Carranza, J.~L. Grismer, P.~Wagner,
+  A.~Schmitz, N.~B. Ananjeva, and R.~F. Inger. 2000.
+\newblock {Availability of new Baysian-delimited gecko names and the importance
+  of character-based species descriptions}.
+\newblock Proceedings of the Royal Society B: Biological Sciences 278:490--492.
+
+\bibitem[{Baylac et~al.(2003)Baylac, Villemant, and Simbolotti}]{Baylac2003}
+Baylac, M., C.~Villemant, and G.~Simbolotti. 2003.
+\newblock {Combining geometric morphometrics with pattern recognition for the
+  investigation of species complexes}.
+\newblock Biological Journal of the Linnean Society 80:89--98.
+
+\bibitem[{Bickford et~al.(2007)Bickford, Lohman, Sodhi, Ng, Meier, Winker,
+  Ingram, and Das}]{Bickford2007}
+Bickford, D., D.~J. Lohman, N.~S. Sodhi, P.~K.~L. Ng, R.~Meier, K.~Winker,
+  K.~K. Ingram, and I.~Das. 2007.
+\newblock {Cryptic species as a window on diversity and conservation.}
+\newblock Trends in ecology {\&} evolution 22:148--55.
+
+\bibitem[{Breiman et~al.(1984)Breiman, Friedman, Stone, and
+  Olshen}]{Breiman1984}
+Breiman, L., J.~Friedman, C.~J. Stone, and R.~A. Olshen. 1984.
+\newblock {Classification and regression trees}.
+\newblock Wadsworth International Group, Belmont.
+
+\bibitem[{Bury(2017)}]{Bury2017}
+Bury, R.~B. 2017.
+\newblock {Biogeography of Western Pond Turtles in the western Great Basin:
+  Dispersal Across a Northwest Passage ?}
+\newblock Western Wildlife2 4:72--80.
+
+\bibitem[{Bury et~al.(2010)Bury, Germano, and Bury}]{Bury2010}
+Bury, R.~B., D.~J. Germano, and G.~W. Bury. 2010.
+\newblock {Population Structure and Growth of the Turtle Actinemys marmorata
+  from the Klamath–Siskiyou Ecoregion: Age, Not Size, Matters}.
+\newblock Copeia 2010:443--451.
+
+\bibitem[{Buskirk et~al.(2005)Buskirk, Parham, and Feldman}]{Buskirk2005}
+Buskirk, S.~W., J.~F. Parham, and C.~R. Feldman. 2005.
+\newblock {On the hybridisation between two distantly related Asian turtles
+  (Testudines: Scalia x Mauremys)}.
+\newblock Salamandra 41:21--26.
+
+\bibitem[{Cardini et~al.(2009)Cardini, Nagorsen, O'Higgins, Polly, {Thorington
+  Jr}, and Tongiorgi}]{Cardini2009a}
+Cardini, A., D.~Nagorsen, P.~O'Higgins, P.~D. Polly, R.~W. {Thorington Jr}, and
+  P.~Tongiorgi. 2009.
+\newblock {Detecting biological distinctiveness using geometric morphometrics:
+  an example case from the Vancouver Island marmot}.
+\newblock Ethology Ecology {\&} Evolution 21:209--223.
+
+\bibitem[{Carstens and Dewey(2010)}]{Carstens2010}
+Carstens, B.~C., and T.~A. Dewey. 2010.
+\newblock {Species Delimitation Using a Combined Coalescent and
+  Information-Theoretic Approach: An Example from North American Myotis Bats}.
+\newblock Systematic Biology 59:400--414.
+
+\bibitem[{Carstens et~al.(2013)Carstens, Pelletier, Reid, and
+  Satler}]{Carstens2013}
+Carstens, B.~C., T.~a. Pelletier, N.~M. Reid, and J.~D. Satler. 2013.
+\newblock {How to fail at species delimitation.}
+\newblock Molecular ecology 22:4369--83.
+
+\bibitem[{Caumul and Polly(2005)}]{Caumul2005a}
+Caumul, R., and P.~D. Polly. 2005.
+\newblock {Phylogenetic and environmental components of morphological
+  variation: skull, mandible, and molar shape in marmots (Marmota, Rodentia).}
+\newblock Evolution; international journal of organic evolution 59:2460--72.
+
+\bibitem[{Clare(2011)}]{Clare2011}
+Clare, E.~L. 2011.
+\newblock {Cryptic species? Patterns of maternal and paternal gene flow in
+  eight neotropical bats.}
+\newblock PloS one 6:e21460.
+
+\bibitem[{Claude(2006)}]{Claude2006}
+Claude, J. 2006.
+\newblock {Convergence induced by plastral kinesis and geometric morphometric
+  assessment: a geometric morphometric assessment}.
+\newblock Fossil Turtle Research 1:34--45.
+
+\bibitem[{Claude et~al.(2003)Claude, Paradis, Tong, and Auffray}]{Claude2003a}
+Claude, J., E.~Paradis, H.~Tong, and J.~C. Auffray. 2003.
+\newblock {A geometric morphometric assessment of the effects
+  of$\backslash$nenvironment and cladogenesis on the evolution of
+  the$\backslash$nturtle shell}.
+\newblock Biological Journal of the Linnean Society 79:485--501.
+
+\bibitem[{Dayrat(2005)}]{Dayrat2005a}
+Dayrat, B. 2005.
+\newblock {Towards integrative taxonomy}.
+\newblock Biological Journal of the Linnean Society 85:407--415.
+
+\bibitem[{{De Queiroz}(2007)}]{DeQueiroz2007b}
+{De Queiroz}, K. 2007.
+\newblock {Species concepts and species delimitation.}
+\newblock Systematic Biology 56:879--86.
+
+\bibitem[{Demandt and Bergek(2009)}]{Demandt2009}
+Demandt, M.~H., and S.~Bergek. 2009.
+\newblock {Identification of cyprinid hybrids by using geometric morphometrics
+  and microsatellites}.
+\newblock Journal of Applied Ichthyology 25:695--701.
+
+\bibitem[{Dillard(2017)}]{Dillard2017}
+Dillard, K.~C. 2017.
+\newblock {A comparative analysis of geometric morphometrics across two
+  Pseudemys turtle species in east central Virginia}.
+\newblock Masters.
+\newblock Virginia Commonwealth University.
+
+\bibitem[{Dobigny et~al.(2003)Dobigny, Granjon, Aniskin, Ba, and
+  Voloboulev}]{Dobigny2003}
+Dobigny, G., L.~Granjon, V.~Aniskin, K.~Ba, and V.~Voloboulev. 2003.
+\newblock {A new sigling species of Taterillus (Muridae, Gerbillinae) from West
+  Agrica}.
+\newblock Mammalian Biology 68:299--316.
+
+\bibitem[{Dryden(2013)}]{Dryden2013}
+Dryden, I.~L. 2013.
+\newblock shapes: Statistical shape analysis.
+\newblock R package version 1.1-8.
+
+\bibitem[{Dryden and Mardia(1998)}]{Dryden1998a}
+Dryden, I.~L., and K.~Y. Mardia. 1998.
+\newblock {Statistical shape analysis}.
+\newblock Wiley, New York.
+
+\bibitem[{Edwards et~al.(2011)Edwards, Claude, {Van Vuuren}, and
+  Matthee}]{Edwards2011}
+Edwards, S., J.~Claude, B.~J. {Van Vuuren}, and C.~A. Matthee. 2011.
+\newblock {Evolutionary history of the Karoo bush rat, Myotomys unisulcatus
+  (Rodentia: Muridae): Disconcordance between morphology and genetics}.
+\newblock Biological Journal of the Linnean Society 102:510--526.
+
+\bibitem[{Eldredge and Gould(1972)}]{Eldredge1972}
+Eldredge, N., and S.~J. Gould. 1972.
+\newblock {Punctuated equilibria: an alternative to phyletic gradualism}.
+\newblock Pages 82--115 \emph{in} T.~J.~M. Schopf, ed. Models in Paleobiology.
+  Freeman Cooper, San Francisco.
+
+\bibitem[{Feldman and Parham(2002)}]{Feldman2002}
+Feldman, C.~R., and J.~F. Parham. 2002.
+\newblock {Molecular phylogenetics of emydine turtles: taxonomic revision and
+  the evolution of shell kinesis.}
+\newblock Molecular Phylogenetics and Evolution 22:388--98.
+
+\bibitem[{Francoy et~al.(2009)Francoy, Silva, Nunes-Silva, Menezes, and
+  Imperatriz-Fonseca}]{Francoy2009}
+Francoy, T.~M., R.~A.~O. Silva, P.~Nunes-Silva, C.~Menezes, and V.~L.
+  Imperatriz-Fonseca. 2009.
+\newblock {Gender identification of five genera of stingless bees (Apidae,
+  Meliponini) based on wing morphology}.
+\newblock Genetics and molecular research 8:207--214.
+
+\bibitem[{Fruciano et~al.(2016)Fruciano, Franchini, Raffini, Fan, and
+  Meyer}]{Fruciano2016}
+Fruciano, C., P.~Franchini, F.~Raffini, S.~Fan, and A.~Meyer. 2016.
+\newblock {Are sympatrically speciating Midas cichlid fish special? Patterns of
+  morphological and genetic variation in the closely related species
+  Archocentrus centrarchus}.
+\newblock Ecology and Evolution 6:4102--4114.
+
+\bibitem[{Funk et~al.(2012)Funk, Caminer, and Ron}]{Funk2012}
+Funk, W.~C., M.~Caminer, and S.~R. Ron. 2012.
+\newblock {High levels of cryptic species diversity uncovered in Amazonian
+  frogs.}
+\newblock Proceedings of the Royal Society B: Biological Sciences 279:1806--14.
+
+\bibitem[{Gaubert et~al.(2005)Gaubert, Taylor, Fernandes, Bruford, and
+  Veron}]{Gaubert2005b}
+Gaubert, P., P.~J. Taylor, C.~a. Fernandes, M.~W. Bruford, and G.~Veron. 2005.
+\newblock {Patterns of cryptic hybridization revealed using an integrative
+  approach: a case study on genets (Carnivora, Viverridae, Genetta spp.) from
+  the southern African subregion}.
+\newblock Biological Journal of the Linnean Society 86:11--33.
+
+\bibitem[{Germano and Bury(2009)}]{Germano2009}
+Germano, D.~J., and R.~B. Bury. 2009.
+\newblock {Variation in body size, growth, and population structure of
+  Actinemys marmorata from lentic and lotic habitats in Southern Oregon}.
+\newblock Journal of Herpetology 43:510--520.
+
+\bibitem[{Germano and Rathbun(2008)}]{Germano2008}
+Germano, D.~J., and G.~B. Rathbun. 2008.
+\newblock {Growth, population structure, and reproduction of western pond
+  turtles (Actinemys marmorata) on the Central Coast of California}.
+\newblock Chelonian Conservation and Biology 7:188--194.
+
+\bibitem[{Gould and Eldredge(1977)}]{Gould1977a}
+Gould, S.~J., and N.~Eldredge. 1977.
+\newblock {Punctuated equilibria: the tempo and mode of evolution
+  reconsidered}.
+\newblock Paleobiology 3:115--151.
+
+\bibitem[{Guillot et~al.(2012)Guillot, Renaud, Ledevin, Michaux, and
+  Claude}]{Guillot2012}
+Guillot, G., S.~Renaud, R.~Ledevin, J.~Michaux, and J.~Claude. 2012.
+\newblock {A unifying model for the analysis of phenotypic, genetic, and
+  geographic data}.
+\newblock Systematic Biology 61:897--911.
+
+\bibitem[{G{\"{u}}nd{\"{u}}z et~al.(2007)G{\"{u}}nd{\"{u}}z, Jaarola, Tez,
+  Yeniyurt, Polly, and Searle}]{Gunduz2007}
+G{\"{u}}nd{\"{u}}z, I., M.~Jaarola, C.~Tez, C.~Yeniyurt, P.~D. Polly, and J.~B.
+  Searle. 2007.
+\newblock {Multigenic and morphometric differentiation of ground squirrels
+  (Spermophilus, Scuiridae, Rodentia) in Turkey, with a description of a new
+  species.}
+\newblock Molecular phylogenetics and evolution 43:916--35.
+
+\bibitem[{Hand and Till(2001)}]{Hand2001}
+Hand, D.~J., and R.~J. Till. 2001.
+\newblock {A Simple Generalisation of the Area Under the ROC Curve for Multiple
+  Class Classification Problems}.
+\newblock Machine Learning 45:171--186.
+
+\bibitem[{Hastie et~al.(2009)Hastie, Tibshirani, and Friedman}]{Hastie2009}
+Hastie, T., R.~Tibshirani, and J.~Friedman. 2009.
+\newblock {The elements of statistical learning: data mining, inference, and
+  prediction}.
+\newblock 2nd ed. Springer, New York.
+
+\bibitem[{Hastie et~al.(2015)Hastie, Tibshirani, Leisch, Hornik, and
+  Ripley.}]{mdapack}
+Hastie, T., R.~Tibshirani, F.~Leisch, K.~Hornik, and B.~D. Ripley. 2015.
+\newblock mda: Mixture and Flexible Discriminant Analysis.
+\newblock R package version 0.4-8.
+
+\bibitem[{Hausdorf and Hennig(2010)}]{Hausdorf2010}
+Hausdorf, B., and C.~Hennig. 2010.
+\newblock {Species delimitation using dominant and codominant multilocus
+  markers.}
+\newblock Systematic biology 59:491--503.
+
+\bibitem[{Holland(1992)}]{Holland1992}
+Holland, D.~C. 1992.
+\newblock {Level and pattern in morphological variation: a phylogeographic
+  study of the western pond turtle (Clemmys marmorata)}.
+\newblock Ph.D. thesis.
+\newblock University of Southwestern Louisiana.
+
+\bibitem[{Huelsenbeck et~al.(2011)Huelsenbeck, Andolfatto, and
+  Huelsenbeck}]{Huelsenbeck2011b}
+Huelsenbeck, J.~P., P.~Andolfatto, and E.~T. Huelsenbeck. 2011.
+\newblock {Structurama: bayesian inference of population structure.}
+\newblock Evolutionary bioinformatics online 7:55--9.
+
+\bibitem[{Kaufman and Rousseeuw(1990)}]{Kaufman1990}
+Kaufman, L., and P.~J. Rousseeuw. 1990.
+\newblock {Finding groups in data : an introduction to cluster analysis}.
+\newblock Wiley, New York.
+
+\bibitem[{Kendall(1977)}]{Kendall1977a}
+Kendall, D.~G. 1977.
+\newblock {The diffusion of shape}.
+\newblock Advances in Applied Probability 9:428--430.
+
+\bibitem[{Kuhn(2013)}]{KuhnMAN2013}
+Kuhn, M. 2013.
+\newblock caret: Classification and Regression Training.
+\newblock R package version 5.15-61.
+
+\bibitem[{Kuhn and Johnson(2013)}]{Kuhn2013}
+Kuhn, M., and K.~Johnson. 2013.
+\newblock {Applied predictive modeling}.
+\newblock Springer, New York, NY.
+
+\bibitem[{Leach{\'{e}} and Fujita(2010)}]{Leache2010}
+Leach{\'{e}}, A.~D., and M.~K. Fujita. 2010.
+\newblock {Bayesian species delimitation in West African forest geckos
+  (Hemidactylus fasciatus).}
+\newblock Proceedings. Biological sciences / The Royal Society 277:3071--7.
+
+\bibitem[{Liaw and Wiener(2002)}]{Liaw2002}
+Liaw, A., and M.~Wiener. 2002.
+\newblock Classification and regression by randomforest.
+\newblock R News 2:18--22.
+
+\bibitem[{Lubcke and Wilson(2007)}]{Lubcke2007}
+Lubcke, G.~M., and D.~S. Wilson. 2007.
+\newblock {Variation in shell morphology of the Western Pond Turtle (Actinemys
+  marmorata Baird and Giarard) from three aquativ habitats in Northern
+  California}.
+\newblock Journal of Herpetology 41:107--114.
+
+\bibitem[{MacLeod(2007)}]{MacLeod2007}
+MacLeod, N. 2007.
+\newblock {Automated taxon identification in systematics: theory, approaches
+  and applications}.
+\newblock CRC Press, Boca Raton.
+
+\bibitem[{Markolf et~al.(2013)Markolf, Rakotonirina, Fichtel, von Grumbkow,
+  Brameier, and Kappeler}]{Markolf2013}
+Markolf, M., H.~Rakotonirina, C.~Fichtel, P.~von Grumbkow, M.~Brameier, and
+  P.~M. Kappeler. 2013.
+\newblock {True lemurs{\ldots}true species - species delimitation using
+  multiple data sources in the brown lemur complex}.
+\newblock BMC Evolutionary Biology 13:233.
+
+\bibitem[{Mitrovski-Bogdanovic et~al.(2013)Mitrovski-Bogdanovic, Petrovic,
+  Mitrovic, Ivanovic, {\v{Z}}ikic, Star{\'{y}}, Vorburger, and
+  Tomanovic}]{MitrovskiBogdanovic2013}
+Mitrovski-Bogdanovic, A., A.~Petrovic, M.~Mitrovic, A.~Ivanovic,
+  V.~{\v{Z}}ikic, P.~Star{\'{y}}, C.~Vorburger, and {\v{Z}}.~Tomanovic. 2013.
+\newblock {Identification of two cryptic species within the Praon abjectum
+  group (Hymenoptera: Braconidae: Aphidiinae) using molecular markers and
+  geometric morphometrics}.
+\newblock Annals of the entomological society of America 106:170--180.
+
+\bibitem[{{Munoz-Munoz F.} and {Perpinan D.}(2010)}]{Munoz-MunozF.2010}
+{Munoz-Munoz F.}, and {Perpinan D.} 2010.
+\newblock {Measurement error in morphometric studies: comparison between manual
+  and computerized methods}.
+\newblock Ann. Zool. 47:46--56.
+
+\bibitem[{Navega et~al.(2015)Navega, Vicente, Vieira, Ross, and
+  Cunha}]{Navega2015}
+Navega, D., R.~Vicente, D.~N. Vieira, A.~H. Ross, and E.~Cunha. 2015.
+\newblock {Sex estimation from the tarsal bones in a Portuguese sample: a
+  machine learning approach}.
+\newblock International Journal of Legal Medicine 129:651--659.
+
+\bibitem[{O'Meara(2010)}]{O'Meara2010}
+O'Meara, B.~C. 2010.
+\newblock {New heuristic methods for joint species delimitation and species
+  tree inference.}
+\newblock Systematic biology 59:59--73.
+
+\bibitem[{Padial et~al.(2010)Padial, Miralles, {De la Riva}, and
+  Vences}]{Padial2010}
+Padial, J.~M., A.~Miralles, I.~{De la Riva}, and M.~Vences. 2010.
+\newblock {The integrative future of taxonomy}.
+\newblock Frontiers in Zoology 7:1--14.
+
+\bibitem[{Parham et~al.(2013)Parham, Papenfuss, Dijk, Wilson, Marte, Schettino,
+  and {Brian Simison}}]{Parham2013}
+Parham, J.~F., T.~J. Papenfuss, P.~P.~V. Dijk, B.~S. Wilson, C.~Marte, L.~R.
+  Schettino, and W.~{Brian Simison}. 2013.
+\newblock {Genetic introgression and hybridization in Antillean freshwater
+  turtles (Trachemys) revealed by coalescent analyses of mitochondrial and
+  cloned nuclear markers.}
+\newblock Molecular phylogenetics and evolution 67:176--87.
+
+\bibitem[{Pfenninger and Schwenk(2007)}]{Pfenninger2007}
+Pfenninger, M., and K.~Schwenk. 2007.
+\newblock {Cryptic animal species are homogeneously distributed among taxa and
+  biogeographical regions.}
+\newblock BMC evolutionary biology 7:121.
+
+\bibitem[{Polly(2003)}]{Polly2003}
+Polly, P.~D. 2003.
+\newblock {Paleophylogeography of Sorex araneus: molar shape as a morphological
+  marker for fossil shrews}.
+\newblock Mammalia 68:233--243.
+
+\bibitem[{Polly(2007)}]{Polly2007a}
+---{}---{}---. 2007.
+\newblock {Phylogeographic differentiation in Sorex araneus: morphology in
+  relation to geography and karyotype}.
+\newblock Russian Journal of Theriology 6:73--84.
+
+\bibitem[{Polly et~al.(2016)Polly, Stayton, Dumont, Pierce, Rayfield, and
+  Angielczyk}]{Polly2016}
+Polly, P.~D., C.~T. Stayton, E.~R. Dumont, S.~E. Pierce, E.~J. Rayfield, and
+  K.~D. Angielczyk. 2016.
+\newblock {Combining geometric morphometrics and finite element analysis with
+  evolutionary modeling: towards a synthesis}.
+\newblock Journal of Vertebrate Paleontology 4634.
+
+\bibitem[{Pons et~al.(2006)Pons, Barraclough, Gomez-Zurita, Cardoso, Duran,
+  Hazell, Kamoun, Sumlin, and Vogler}]{Pons2006}
+Pons, J., T.~Barraclough, J.~Gomez-Zurita, A.~Cardoso, D.~Duran, S.~Hazell,
+  S.~Kamoun, W.~Sumlin, and A.~Vogler. 2006.
+\newblock {Sequence-Based Species Delimitation for the DNA Taxonomy of
+  Undescribed Insects}.
+\newblock Systematic Biology 55:595--609.
+
+\bibitem[{{R Core Team}(2016)}]{R2016}
+{R Core Team}. 2016.
+\newblock R: A Language and Environment for Statistical Computing.
+\newblock R Foundation for Statistical Computing, Vienna, Austria.
+
+\bibitem[{Rivera(2008)}]{Rivera2008}
+Rivera, G. 2008.
+\newblock {Ecomorphological variation in shell shape of the freshwater turtle
+  Pseudemys concinna inhabiting different aquatic flow regimes.}
+\newblock Integrative and comparative biology 48:769--87.
+
+\bibitem[{Rivera et~al.(2014)Rivera, Davis, Godwin, and Adams}]{Rivera2014}
+Rivera, G., J.~N. Davis, J.~C. Godwin, and D.~C. Adams. 2014.
+\newblock {Repeatability of Habitat-Associated Divergence in Shell Shape of
+  Turtles}.
+\newblock Evolutionary Biology pages 29--37.
+
+\bibitem[{Rivera and Stayton(2011)}]{Rivera2011}
+Rivera, G., and C.~T. Stayton. 2011.
+\newblock {Finite element modeling of shell shape in the freshwater turtle
+  Pseudemys concinna reveals a trade-off between mechanical strength and
+  hydrodynamic efficiency.}
+\newblock Journal of morphology 272:1192--203.
+
+\bibitem[{Rohlf(2005)}]{Rohlf2005}
+Rohlf, F.~J. 2005.
+\newblock {TpsDig 2.04}.
+
+\bibitem[{Schilck-Steiner et~al.(2007)Schilck-Steiner, Seifert, Stauffer,
+  Christian, Crozier, and Steiner}]{SchlickSteiner2007}
+Schilck-Steiner, B.~C., B.~Seifert, C.~Stauffer, E.~Christian, R.~H. Crozier,
+  and F.~M. Steiner. 2007.
+\newblock {Without morphology, cryptic species stay in taxonomic crypsis
+  following discovery}.
+\newblock Trends in ecology {\&} evolution 22:391--392.
+
+\bibitem[{Seeliger(1945)}]{Seeliger1945}
+Seeliger, L.~M. 1945.
+\newblock {Variation in the Pacific Mud Turtle}.
+\newblock Copeia 1945:150--159.
+
+\bibitem[{Spinks and Shaffer(2005)}]{Spinks2005}
+Spinks, P.~Q., and H.~B. Shaffer. 2005.
+\newblock {Range-wide molecular analysis of the western pond turtle (Emys
+  marmorata): cryptic variation, isolation by distance, and their conservation
+  implications.}
+\newblock Molecular ecology 14:2047--64.
+
+\bibitem[{Spinks and Shaffer(2009)}]{Spinks2009}
+---{}---{}---. 2009.
+\newblock {Conflicting mitochondrial and nuclear phylogenies for the widely
+  disjunct Emys (Testudines: Emydidae) species complex, and what they tell us
+  about biogeography and hybridization.}
+\newblock Systematic biology 58:1--20.
+
+\bibitem[{Spinks et~al.(2014)Spinks, Thomson, and {Bradley
+  Shaffer}}]{Spinks2014}
+Spinks, P.~Q., R.~C. Thomson, and H.~{Bradley Shaffer}. 2014.
+\newblock {The advantages of going large: genome wide SNPs clarify the complex
+  population history and systematics of the threatened western pond turtle}.
+\newblock Molecular Ecology pages n/a--n/a.
+
+\bibitem[{Spinks et~al.(2010)Spinks, Thomson, and Shaffer}]{Spinks2010}
+Spinks, P.~Q., R.~C. Thomson, and H.~B. Shaffer. 2010.
+\newblock {Nuclear gene phylogeography reveals the historical legacy of an
+  ancient inland sea on lineages of the western pond turtle, Emys marmorata in
+  California.}
+\newblock Molecular ecology 19:542--56.
+
+\bibitem[{Stayton(2011)}]{Stayton2011}
+Stayton, C.~T. 2011.
+\newblock {Biomechanics on the half shell: functional performance influences
+  patterns of morphological variation in the emydid turtle carapace.}
+\newblock Zoology (Jena, Germany) 114:213--23.
+
+\bibitem[{Stuart et~al.(2006)Stuart, Inger, and Voris}]{Stuart2006}
+Stuart, B.~L., R.~F. Inger, and H.~K. Voris. 2006.
+\newblock {High level of cryptic species diversity revealed by sympatric
+  lineages of Southeast Asian forest frogs.}
+\newblock Biology letters 2:470--4.
+
+\bibitem[{Sztencel-Jab{\l}onka et~al.(2009)Sztencel-Jab{\l}onka, Jones, and
+  BogdanowicZ}]{Sztencel-Jabonka2009}
+Sztencel-Jab{\l}onka, A., G.~Jones, and W.~BogdanowicZ. 2009.
+\newblock {Skull Morphology of Two Cryptic Bat Species: Pipistrellus
+  pipistrellus and P. pygmaeus — A 3D Geometric Morphometrics Approach with
+  Landmark Reconstruction}.
+\newblock Acta Chiropterologica 11:113--126.
+
+\bibitem[{{Van Bocxlaer} and Hunt(2013)}]{VanBocxlaer2013}
+{Van Bocxlaer}, B., and G.~Hunt. 2013.
+\newblock {Morphological stasis in an ongoing gastropod radiation from Lake
+  Malawi}.
+\newblock Proceedings of the National Academy of Sciences .
+
+\bibitem[{{Van Bocxlaer} and Schulthei{\ss}(2010)}]{VanBocxlaer2010}
+{Van Bocxlaer}, B., and R.~Schulthei{\ss}. 2010.
+\newblock {Comparison of morphometric techniques for shapes with few homologous
+  landmarks based on machine-learning approaches to biological discrimination}.
+\newblock Paleobiology 36:497--515.
+
+\bibitem[{van~den Brink and Bokma(2011)}]{VandenBrink2011}
+van~den Brink, V., and F.~Bokma. 2011.
+\newblock {Morphometric shape analysis using learning vector quantization
+  neural networks — an example distinguishing two microtine vole species}.
+\newblock Annales Zoologici Fennici 48:359--364.
+
+\bibitem[{Venables and Ripley(2002)}]{Venables2002}
+Venables, W.~N., and B.~D. Ripley. 2002.
+\newblock Modern Applied Statistics with S.
+\newblock 4th ed. Springer, New York.
+
+\bibitem[{Vitek et~al.(2017)Vitek, Manz, Gao, Bloch, Strait, and
+  Boyer}]{Vitek2017}
+Vitek, N.~S., C.~L. Manz, T.~Gao, J.~I. Bloch, S.~G. Strait, and D.~M. Boyer.
+  2017.
+\newblock {Semi-supervised determination of pseudocryptic morphotypes using
+  observer-free characterizations of anatomical alignment and shape}.
+\newblock Ecology and Evolution 7:5041--5055.
+
+\bibitem[{von Cramon-Taubadel et~al.(2007)von Cramon-Taubadel, Frazier, and
+  Lahr}]{Cramon2007}
+von Cramon-Taubadel, N., B.~C. Frazier, and M.~M. Lahr. 2007.
+\newblock {The problem of assessing landmark error in geometric morphometrics:
+  theory, methods, and modifications}.
+\newblock American journal of physical anthropology 132:535--544.
+
+\bibitem[{Yang and Rannala(2010)}]{Yang2010b}
+Yang, Z., and B.~Rannala. 2010.
+\newblock {Bayesian species delimitation using multilocus sequence data.}
+\newblock Proceedings of the National Academy of Sciences 107:9264--9.
+
+\bibitem[{Zelditch et~al.(2004)Zelditch, Swiderski, and Sheets}]{Zelditch2004}
+Zelditch, M.~L., D.~L. Swiderski, and H.~D. Sheets. 2004.
+\newblock {Geometric morphometrics for biologists: a primer}.
+\newblock Elsevier Academic Press, Amsterdam.
+
+\end{thebibliography}
+
+\newpage{}
+
+\section*{Tables}
+\renewcommand{\thetable}{\arabic{table}}
+\setcounter{table}{0}
+
+\begin{table}[h]
+  \caption{Table of species delimitation hypotheses for \textit{E. marmorata}}
+  \label{tab:hypotheses}
+  \centering
+  \begin{tabular}{l l l }
+    \hline
+      Abbreviation & Number of classes & citation \\
+      \hline
+      SP10.1 & 3 & \citet{Spinks2010} \\
+      SP10.2 & 3 & \citet{Spinks2010} \\
+      SP10.3 & 4 & \citet{Spinks2010} \\
+      SP14.1 & 2 & \citet{Spinks2014} \\
+      SP14.2 & 4 & \citet{Spinks2014} \\
+      Morph & 2 & \citet{Spinks2010} \\
+      \hline
+  \end{tabular}
+  %\bigskip{}
+  %\\
+  %{\footnotesize Note: Table titles should be short. Further details 
+  %should go in a `notes' area after the tabular environment, as shown 
+  %here. $^a$ Okapis are not native to Chicago, but they are to be met with 
+  %in both of the major Chicagoland zoos.}
+\end{table}
+
+\begin{table}[h]
+  \caption{Table of the supervised learning methods used in this analysis.}
+  \label{tab:methods}
+  \centering
+  \scalebox{0.9}{
+  \begin{tabular}{l l l l}
+    \hline
+    Method name & abbreviation & R package & citation \\
+    \hline
+    multinomial logistic regression & MLR & nnet & \citet{Venables2002} \\
+    linear discriminate analysis & LDA & MASS & \citet{Venables2002} \\
+    penalized discrminiate analysis & PDA & mda & \citet{mdapack} \\
+    single-hidden-layer neural network & NN & nnet & \citet{Venables2002} \\
+    random forests & RF & randomForest & \citet{Liaw2002} \\
+    \hline
+  \end{tabular}}
+\end{table}
+
+\begin{table}[h]
+  \caption{Results from the within-individual to between-individual Procrustes distances for the replicated plastron shape data.
+Results are presented for all three datasets analyzed here: the \textit{Trachemys} dataset, the eight species dataset, and each of the \textit{Emys marmorata} classification schemes.}
+  \label{tab:rep_res}
+  \centering
+  \begin{tabular}{l l c c}
+    \textbf{Dataset} & \textbf{Scheme} & \textbf{Ratio} & \textbf{Corrected ratio} \\
+    \hline
+    Replicates & & 1.11 & \\
+    \hline
+    Seven species & & 0.33 & 0.37 \\
+    \textit{Trachemys} & & 0.76 & 0.84 \\
+    \hline
+    \textit{E. marmorata} & SP10.1 & 0.99 & 1.10 \\
+     & SP10.2 & 1.00 & 1.11 \\
+     & SP10.3 & 0.94 & 1.04 \\
+     & SP14.1 & 1.01 & 1.12 \\
+     & SP14.2 & 0.93 & 1.04 \\
+     & Morph & 0.99 & 1.09 \\
+    \hline
+  \end{tabular}
+\end{table}
+
+\newpage{}
+
+%%%%%%%%%%%%%%%%%%%%%
+% Figure legends
+%%%%%%%%%%%%%%%%%%%%%
+% Please include all figure legends in a separate section at the end of 
+% the document. If you use \label{} and \ref{} to refer to your figures,
+% these can still work even if you comment out the %includegraphics{}
+% line. If you refer to figures as "fig. 1" (etc.) manually, the
+% legends can also appear simply as paragraphs.
+% For submission, please upload the relevant figure files separately to
+% Editorial Manager; Editorial Manager should insert them at the end of
+% the PDF automatically.
+% Figure legends should be concise, though they can be longer than the
+% titles of tables.
+
+\section*{Figure legends}
+
+\begin{figure}[h!]
+  \centering
+  \includegraphics[height = 0.6\textheight, width = \textwidth, keepaspectratio = true]{figure/Ken_Ang_SpLoc_final_comp}
+  \caption{Geographic distribution of specimens sampled for comparing the hypothesized subdivisions of \textit{Emys marmorata}.
+Each hypothesized scheme has two or more possible classes.
+Sample size differs between schemes because of our ability to confidently assign museum specimens to the various schemes.
+The number of localities shown on each map is less than the number of specimens sampled because some localities produced multiple specimens.
+The different classification abbreviations are as follows: \textit{E.
+marmorata} = ``marm'', \textit{E.
+pallida} = ``pall'', Central Coast Ranges = ``CCR'', San Joaquin Valley = ``SJ,'' Baya Peninsula = ``Baja,'', and Sierra Foothills = ``Foothill.''}
+  \label{fig:map}
+\end{figure}
+
+\begin{figure}[h!]
+  \centering
+  \includegraphics[height = 0.6\textheight, width = \textwidth, keepaspectratio = true]{figure/plastra}
+  \caption{Depiction of general plastral shape of \textit{E. marmorata} and position of the 19 landmarks used in this study. Anterior is towards the top of the figure.}
+  \label{fig:plastra}
+\end{figure}
+
+\begin{figure}[h!]
+  \centering
+  \begin{subfigure}[b]{0.45\textwidth}
+    \caption{}
+    \includegraphics[height = 0.6\textheight, width = \textwidth, keepaspectratio = true]{figure/cc7_pc_graph}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.45\textwidth}
+    \caption{}
+    \includegraphics[height = 0.6\textheight, width = \textwidth, keepaspectratio = true]{figure/tra_pc_graph}
+  \end{subfigure}
+  \caption{Two scatterplots of morphological differences from two of the three datasets analyzed in this study.
+(a) Scatterplot of the first two PCA axies from the landmarks from the eight different species dataset, and (b) the first two axes of variation from two subspecies of \textit{Trachemys} dataset.
+Point colors correspond to the categories within each dataset while point size is proportional to individual centroid size.
+In parentheses next to the axis labels are the percent of total variation accounted for by that axis.
+For both datasets there are clear distinctions between the different categories.}
+  \label{fig:other_pca}
+\end{figure}
+
+\begin{figure}[h!]
+  \centering
+  \includegraphics[height = 0.6\textheight, width = \textwidth, keepaspectratio = true]{figure/emys_pc_graph}
+  \caption{Scatterplot of the first two axes of morphological variation in the \textit{Emys marmorata} dataset.
+Each panel corresponds to one of the six different classification schemes analyzed as part of this study (Tab.
+\ref{tab:hypotheses}).
+Point color corresponds to the categories within each scheme, and the class names correspond to geographic regions.
+Point size is proportional to centroid size of that specimen and the numbers in parentheses next to the axis labels are the percent of total variation accounted for along that axis.}
+  \label{fig:emys_pca}
+\end{figure}
+
+\begin{figure}[h!]
+  \centering
+  \includegraphics[height = 0.6\textheight, width = \textwidth, keepaspectratio = true]{figure/sex_test_hist}
+  \caption{Comparison of observed Procrustes distance between the centroids of each sex (vertical line) to a null distribution generated from 1000 permutations of the sex-labels.
+This result indicates that the difference between the centroids is as small/smaller than expected by random.}
+  \label{fig:sex_test}
+\end{figure}
+
+\begin{figure}[h!]
+  \centering
+  \includegraphics[height = 0.6\textheight, width = \textwidth, keepaspectratio = true]{figure/other_model_sel}
+  \caption{Comparisons of model fit to the training dataset for each of the supervised learning methods applied to the first two datasets; the results from the eight species dataset are presented in the left column, while those from the \textit{Trachemys} dataset are presented in the right column.
+Models were fit to datasets of varying complexity, with the number of parameters listed along the x-axis.
+Model fit is measured as the area under the receiver operating characteristic (AUC), which ranges from 0.5 to 1.
+Error bars correspond to one standard error estimated from 10 rounds of 5-fold cross-validation.
+The red dot corresponds to the model fit with the highest mean AUC while the blue dot corresponds to the model selected for further analysis.
+In some cases, there is no difference in complexity between the best and selected models.}
+  \label{fig:other_sel}
+\end{figure}
+
+\begin{figure}[h!]
+  \centering
+  \includegraphics[height = 0.6\textheight, width = \textwidth, keepaspectratio = true]{figure/other_oos_sel}
+  \caption{The results of out-of-sample predictive performance of the selected models for both the eight species (left) and \textit{Trachemys} datasets.
+Predictive performance is measured as the area under the receiver operating characteristic (AUC), which ranges from 0.5 to 1.
+Points correspond to the individual out-of-sample predictive performance of the specific model, indicated along the x-axis.
+The horizontal bars correspond the average out-of-sample predictive performance of all the models.}
+  \label{fig:other_oos}
+\end{figure}
+
+\begin{figure}[h!]
+  \centering
+  \includegraphics[height = 0.6\textheight, width = \textwidth, keepaspectratio = true]{figure/emys_model_sel}
+  \caption{AUC values for models of varying complexity fit to the \textit{Emys marmorata} training datasets for each classification scheme.
+The x-axis corresponds to the total number of predictors included in each model, while the y-axis corresponds to the AUC value which is a measure of goodness of fit for classification datasets.
+A model with a high AUC value corresponds to better classification performance than a model with a lower AUC value.
+Standard errors on AUC estimates are calculated from 10 rounds of 5-fold cross-validation.
+Indicated are the best performing and the selected models, in red and blue respectively.
+In some cases, there is no difference in complexity between the best and selected models.}
+  \label{fig:emys_sel}
+\end{figure}
+
+\begin{figure}[h!]
+  \centering
+  \includegraphics[height = 0.6\textheight, width = \textwidth, keepaspectratio = true]{figure/emys_oos_sel}
+  \caption{Comparison of out-of-sample AUC estimates from the predictions of selected models (Fig.
+\ref{fig:emys_sel}), grouped by classification scheme.
+The horizontal line in each panel corresponds to the average AUC value across all models of that classification scheme.}
+  \label{fig:emys_oos}
+\end{figure}
+
+%\subsection*{Online figure legends}
+%
+%\renewcommand{\thefigure}{A\arabic{figure}}
+%\setcounter{figure}{0}
+%
+%\begin{figure}[h!]
+%  %\includegraphics{jumps20m}
+%  \caption{\textit{A}, the quick red fox proceeding to jump 20~m straight 
+%  into the air over not one, but several lazy dogs. \textit{B}, the quick 
+%  red fox landing gracefully despite the skepticism of naysayers.}
+%  \label{Fig:Jumps}
+%\end{figure}
+%
+%\begin{figure}[h!]
+%  %\includegraphics{jumps-nr-okapi}
+%  \caption{The quicker the red fox jumps, the likelier it is to land near 
+%  an okapi. For further details, see \citet{LemKapEx07}.}
+%  \label{Fig:JumpsOk}
+%\end{figure}
+%
+%\renewcommand{\thefigure}{B\arabic{figure}}
+%\setcounter{figure}{0}
 

### 2018-01-23 doc/turtles_peerj.tex
rejected from bjls. reformatted for peerj

--- 2017-12-05 doc/turtles_statistical_learning.tex
@@
-\documentclass[12pt,letterpaper]{article}
+%% Submissions for peer-review must enable line-numbering 
+%% using the lineno option in the \documentclass command.
+%%
+%% Preprints and camera-ready submissions do not need 
+%% line numbers, and should have this option removed.
+%%
+%% Please note that the line numbering option requires
+%% version 1.1 or newer of the wlpeerj.cls file, and
+%% the corresponding author info requires v1.2
 
+\documentclass[fleqn,10pt,lineno]{wlpeerj} % for journal submissions
+% \documentclass[fleqn,10pt]{wlpeerj} % for preprint submissions
 \usepackage{amsmath, amsthm}
@@
 
-\frenchspacing
+\title{Article Title}
+\title{Ensemble approaches for estimating congruence between species delimitation and morphological variation: comparing taxonomic hypotheses for the Pacific Pond Turtle (\textit{Emys marmorata})}
 
-\captionsetup[subfigure]{position = top, labelfont = bf, textfont = normalfont, singlelinecheck = off, justification = raggedright}
-
-\renewcommand{\section}[1]{%
-\bigskip
-\begin{center}
-\begin{Large}
-\normalfont\scshape #1
-\medskip
-\end{Large}
-\end{center}}
-
-\renewcommand{\subsection}[1]{%
-\bigskip
-\begin{center}
-\begin{large}
-\normalfont\itshape #1
-\end{large}
-\end{center}}
-
-\renewcommand{\subsubsection}[1]{%
-\vspace{2ex}
-\noindent
-\textit{#1.}---}
-
-\renewcommand{\tableofcontents}{}
-
-\bibpunct{(}{)}{;}{a}{}{,}  % this is a citation format command for natbib
-
-%\title{How cryptic is cryptic diversity? Machine learning approaches to classifying morphological variation in the Pacific Pond Turtle (\textit{Emys marmorata})}
-\title{Ensemble approaches for estimating congruence between species delimitation and morphological variation: comparing taxonomic hypotheses for the Pacific Pond Turtle (\textit{Emys marmorata})}
 \author[1]{Peter D Smits}%\thanks{peterdavidsmits@gmail.com}}
@@
 \affil[4]{Section of Research and Collections, North Carolina Museum of Natural Sciences}
+\corrauthor[1]{Peter D Smits}{psmits@berkeley.edu}
 
-\begin{document}
-\maketitle
-\noindent{\textbf{Corresponding author:} Peter D Smits, Department of Integrative Biology, University of California -- Berkeley, 3040 Valley Life Sciences Building \#5151, Berkeley, CA, 94720, USA; E-mail: \href{mailto:psmits@berkeley.edu}{psmits@berkeley.edu}
-
-\linenumbers
-\modulolinenumbers[2]
+% \keywords{Keyword1, Keyword2, Keyword3}
 
 \begin{abstract}
-
   We investigated the morphometric identification of cryptic species using machine learning approaches by examining their implications for a recently proposed cryptic turtle species (\textit{Emys pallida}).
@@
 
-\section{Introduction}
+\begin{document}
+
+\flushbottom
+\maketitle
+\thispagestyle{empty}
+
+\section*{Introduction}
 % cryptic diversity
@@
 
-\section{Materials and Methods}
-\subsection{Specimens, sampling, morphometrics}
+\section*{Methods}
+\subsection*{Specimens, sampling, morphometrics}
 
@@
 
-\subsection{Supervised learning approaches}
+\subsection*{Supervised learning approaches}
 
@@
 
-\section{Results}
+\section*{Results}
 
-\subsection{Geometric morphometrics}
+\subsection*{Geometric morphometrics}
 
@@
 
-\subsection{Supervised learning}
+\subsection*{Supervised learning}
 
@@
 
-\section{Discussion}
+\section*{Discussion}
 
@@
 
-\bibliographystyle{abbrvnat}
 \bibliography{turtle,packages}

### 2018-02-28 doc/letter_editor.tex
started edits for plos one

@@
 \begin{document}
-\begin{letter}{Editor \\ \textit{Biological Journal of the Linnean Society}}
+\begin{letter}{Editor \\ \textit{PeerJ}}
   \opening{Dear Editor,}
 
-  Please find enclosed the manuscript entitled ``Ensemble approaches for estimating congruence between species delimitation and morphological variation: comparing taxonomic hypotheses for the Pacific Pond Turtle (\textit{Emys marmorata})'' which we are submitting for consideration for the \textit{Biological Journal of the Linnean Society}.
+  Please find enclosed the manuscript entitled ``Ensemble approaches for estimating congruence between species delimitation and morphological variation: comparing taxonomic hypotheses for the Pacific Pond Turtle (\textit{Emys marmorata})'' which we are submitting for consideration for the \textit{PeerJ}.
 This manuscript is authored by myself (Peter D Smits), Kenneth D Angielczyk, James F Parham, and Bryan L Stuart.

### 2018-02-28 doc/turtles_peerj.tex
started edits for plos one

@@
 \author[2]{Kenneth D Angielczyk}%\thanks{kangielczyk@fieldmuseum.org}}
-\author[3]{James F Parham}%\thanks{jparham@fullerton.edu}}
-\author[4]{Bryan L Stuart}%\thanks{bryan.stuart@naturalsciences.org}}
+\author[3]{Bryan L Stuart}%\thanks{bryan.stuart@naturalsciences.org}}
+\author[4]{James F Parham}%\thanks{jparham@fullerton.edu}}
 \affil[1]{Department of Integrative Biology, University of California -- Berkeley}
 \affil[2]{Integrative Research Center, Field Museum of Natural History}
-\affil[3]{John D. Cooper Archaeological and Paleontological Center, Department of Geological Sciences, California State University, Fullerton}
-\affil[4]{Section of Research and Collections, North Carolina Museum of Natural Sciences}
+\affil[3]{Section of Research and Collections, North Carolina Museum of Natural Sciences}
+\affil[4]{John D. Cooper Archaeological and Paleontological Center, Department of Geological Sciences, California State University, Fullerton}
 \corrauthor[1]{Peter D Smits}{psmits@berkeley.edu}
@@
 
-\subsection*{Is there more than one species of Western Pond Turtle?}
+\section*{Conclusions}
 
@@
 
-\bibliography{turtle,packages}
+\bibliography{new_bib}
 

### 2018-02-28 doc/turtles_plosone.tex
started edits for plos one

--- 2018-02-28 doc/turtles_peerj.tex
@@
-%% Submissions for peer-review must enable line-numbering 
-%% using the lineno option in the \documentclass command.
-%%
-%% Preprints and camera-ready submissions do not need 
-%% line numbers, and should have this option removed.
-%%
-%% Please note that the line numbering option requires
-%% version 1.1 or newer of the wlpeerj.cls file, and
-%% the corresponding author info requires v1.2
+% Template for PLoS
+% Version 3.4 January 2017
+%
+% % % % % % % % % % % % % % % % % % % % % %
+%
+% -- IMPORTANT NOTE
+%
+% This template contains comments intended 
+% to minimize problems and delays during our production 
+% process. Please follow the template instructions
+% whenever possible.
+%
+% % % % % % % % % % % % % % % % % % % % % % % 
+%
+% Once your paper is accepted for publication, 
+% PLEASE REMOVE ALL TRACKED CHANGES in this file 
+% and leave only the final text of your manuscript. 
+% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
+% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
+%
+%
+% There are no restrictions on package use within the LaTeX files except that 
+% no packages listed in the template may be deleted.
+%
+% Please do not include colors or graphics in the text.
+%
+% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
+%
+% % % % % % % % % % % % % % % % % % % % % % %
+%
+% -- FIGURES AND TABLES
+%
+% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
+%
+% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
+% - Figures should be uploaded separately from your manuscript file. 
+% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
+% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
+% For figure citations, please use "Fig" instead of "Figure".
+% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
+%
+% Tables should be cell-based and may not contain:
+% - spacing/line breaks within cells to alter layout or alignment
+% - do not nest tabular environments (no tabular environments within tabular environments)
+% - no graphics or colored text (cell background color/shading OK)
+% See http://journals.plos.org/plosone/s/tables for table guidelines.
+%
+% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
+%
+% % % % % % % % % % % % % % % % % % % % % % % %
+%
+% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
+%
+% IMPORTANT
+% Below are a few tips to help format your equations and other special characters according to our specifications.
+For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
+%
+% For inline equations, please be sure to include all portions of an equation in the math environment.
+For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
+%
+% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
+%
+% Please add line breaks to long display equations when possible in order to fit size of the column. 
+%
+% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
+%
+% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2". 
+%
+% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
+%
+% % % % % % % % % % % % % % % % % % % % % % % % 
+%
+% Please contact latex@plos.org with any questions.
+%
+% % % % % % % % % % % % % % % % % % % % % % % %
 
-\documentclass[fleqn,10pt,lineno]{wlpeerj} % for journal submissions
-% \documentclass[fleqn,10pt]{wlpeerj} % for preprint submissions
-\usepackage{amsmath, amsthm}
-\usepackage{graphicx,hyperref}
-\usepackage{microtype, parskip}
-\usepackage{natbib}
-\usepackage{lineno}
-\usepackage[font=small]{caption}
-\usepackage{float}
-\usepackage{subcaption, multirow, morefloats}
-\usepackage{, wrapfig}
+\documentclass[10pt,letterpaper]{article}
+\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
+
+% amsmath and amssymb packages, useful for mathematical formulas and symbols
+\usepackage{amsmath,amssymb}
+
+% Use adjustwidth environment to exceed column width (see example table in text)
+\usepackage{changepage}
+
+% Use Unicode characters when possible
+\usepackage[utf8x]{inputenc}
+
+% textcomp package and marvosym package for additional characters
+\usepackage{textcomp,marvosym}
+
+% cite package, to clean up citations in the main text. Do not remove.
+\usepackage{cite}
+
+% Use nameref to cite supporting information files (see Supporting Information section for more info)
+\usepackage{nameref,hyperref}
+
+% line numbers
+\usepackage[right]{lineno}
+
+% ligatures disabled
+\usepackage{microtype}
+\DisableLigatures[f]{encoding = *, family = * }
+
+% color can be used to apply background shading to table cells only
+\usepackage[table]{xcolor}
+
+% array package and thick rules for tables
+\usepackage{array}
+
 \usepackage{afterpage}
-\usepackage{rotating}
-\usepackage{titlesec}
-\usepackage[nottoc,numbib]{tocbibind}
-\usepackage{authblk, attrib, fullpage}
-\usepackage{lineno}
+\usepackage{caption, subcaption, multirow, morefloats}
 
-\title{Article Title}
-\title{Ensemble approaches for estimating congruence between species delimitation and morphological variation: comparing taxonomic hypotheses for the Pacific Pond Turtle (\textit{Emys marmorata})}
+% create "+" rule type for thick vertical lines
+\newcolumntype{+}{!{\vrule width 2pt}}
 
-\author[1]{Peter D Smits}%\thanks{peterdavidsmits@gmail.com}}
-\author[2]{Kenneth D Angielczyk}%\thanks{kangielczyk@fieldmuseum.org}}
-\author[3]{Bryan L Stuart}%\thanks{bryan.stuart@naturalsciences.org}}
-\author[4]{James F Parham}%\thanks{jparham@fullerton.edu}}
-\affil[1]{Department of Integrative Biology, University of California -- Berkeley}
-\affil[2]{Integrative Research Center, Field Museum of Natural History}
-\affil[3]{Section of Research and Collections, North Carolina Museum of Natural Sciences}
-\affil[4]{John D. Cooper Archaeological and Paleontological Center, Department of Geological Sciences, California State University, Fullerton}
-\corrauthor[1]{Peter D Smits}{psmits@berkeley.edu}
+% create \thickcline for thick horizontal lines of variable length
+\newlength\savedwidth
+\newcommand\thickcline[1]{%
+  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
+  \cline{#1}%
+  \noalign{\vskip\arrayrulewidth}%
+  \noalign{\global\arrayrulewidth\savedwidth}%
+}
 
-% \keywords{Keyword1, Keyword2, Keyword3}
+% \thickhline command for thick horizontal lines that span the table
+\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
+\hline
+\noalign{\global\arrayrulewidth\savedwidth}}
 
-\begin{abstract}
+% Remove comment for double spacing
+%\usepackage{setspace} 
+%\doublespacing
+
+% Text layout
+\raggedright
+\setlength{\parindent}{0.5cm}
+\textwidth 5.25in 
+\textheight 8.75in
+
+% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
+% Captions will be left justified
+\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
+\renewcommand{\figurename}{Fig}
+
+% Use the PLoS provided BiBTeX style
+\bibliographystyle{plos2015}
+
+% Remove brackets from numbering in List of References
+\makeatletter
+\renewcommand{\@biblabel}[1]{\quad#1.}
+\makeatother
+
+% Leave date blank
+\date{}
+
+% Header and Footer with logo
+\usepackage{lastpage,fancyhdr,graphicx}
+\usepackage{epstopdf}
+\pagestyle{myheadings}
+\pagestyle{fancy}
+\fancyhf{}
+\setlength{\headheight}{27.023pt}
+\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
+\rfoot{\thepage/\pageref{LastPage}}
+\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
+\fancyheadoffset[L]{2.25in}
+\fancyfootoffset[L]{2.25in}
+\lfoot{\sf PLOS}
+
+%% Include all macros below
+
+\newcommand{\lorem}{{\bf LOREM}}
+\newcommand{\ipsum}{{\bf IPSUM}}
+
+%% END MACROS SECTION
+
+\begin{document}
+\vspace*{0.2in}
+
+% Title must be 250 characters or less.
+\begin{flushleft}
+{\Large
+\textbf\newline{Ensemble approaches for estimating congruence between species delimitation and morphological variation: comparing taxonomic hypotheses for the Pacific Pond Turtle (\textit{Emys marmorata})} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
+}
+\newline
+% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
+\\
+Peter D Smits\textsuperscript{1*},
+Kenneth D Angielczyk\textsuperscript{2},
+Bryan L Stuart\textsuperscript{3},
+James F Parham\textsuperscript{4},
+\\
+\bigskip
+\textbf{1} Department of Integrative Biology, University of California -- Berkeley, Berkeley, California, USA.
+\\
+\textbf{2} Integrative Research Center, Field Museum of Natural History, Chicago, Illinois, USA.
+\\
+\textbf{3} Section of Research and Collections, North Carolina Museum of Natural Sciences, Raleigh, North Carolina, USA.
+\\
+\textbf{4} John D. Cooper Archaeological and Paleontological Center, Department of Geological Sciences, California State University -- Fullerton, Fullerton, California, USA.
+\\
+\bigskip
+
+% Insert additional author notes using the symbols described below. Insert symbol callouts after author names as necessary.
+% 
+% Remove or comment out the author notes below if they aren't used.
+%
+% Primary Equal Contribution Note
+%\Yinyang These authors contributed equally to this work.
+
+% Additional Equal Contribution Note
+% Also use this double-dagger symbol for special authorship notes, such as senior authorship.
+%\ddag These authors also contributed equally to this work.
+
+% Current address notes
+%\textcurrency Current Address: Dept/Program/Center, Institution Name, City, State, Country % change symbol to "\textcurrency a" if more than one current address note
+% \textcurrency b Insert second current address 
+% \textcurrency c Insert third current address
+
+% Deceased author note
+%\dag Deceased
+
+% Group/Consortium Author Note
+%\textpilcrow Membership list can be found in the Acknowledgments section.
+
+% Use the asterisk to denote corresponding authorship and provide email address in note below.
+* psmits@berkeley.edu
+
+\end{flushleft}
+% Please keep the abstract below 300 words
+\section*{Abstract}
   We investigated the morphometric identification of cryptic species using machine learning approaches by examining their implications for a recently proposed cryptic turtle species (\textit{Emys pallida}).
@@
 pallida,''} the lack of barriers to gene flow, the strong evidence for widespread admixture between lineages, and the fact that plastron shape can be used to delineate other emydine species and sub-species suggest that its lack of diagnosability most likely reflects the non-distinctiveness of this proposed taxon.
-\end{abstract}
 
-\begin{document}
+% Please keep the Author Summary between 150 and 200 words
+% Use first person. PLOS ONE authors please skip this step. 
+% Author Summary not valid for PLOS ONE submissions.   
+%\section*{Author summary}
 
-\flushbottom
-\maketitle
-\thispagestyle{empty}
+\linenumbers
 
+% Use "Eq" instead of "Equation" for equation citations.
 \section*{Introduction}
+
 % cryptic diversity
 %   most species are still deliminated solely based on morphology
-Molecular systematics has repeatedly demonstrated the existence of cryptic species that can only be diagnosed using genetic data \citep{Stuart2006,Bickford2007,SchlickSteiner2007,Pfenninger2007,Clare2011,Funk2012}.
-In attempts to streamline the documentation of biodiversity, several methods of species delimitation that rely almost entirely on genetic data have recently been proposed \citep{Pons2006,Carstens2010,Hausdorf2010,O'Meara2010,Yang2010b,Huelsenbeck2011b}.
-Although strong caveats on the utility of these methods have been raised \citep{Bauer2000,Carstens2013}, they are nevertheless being used to name species \citep{Leache2010,Spinks2014}.
+Molecular systematics has repeatedly demonstrated the existence of cryptic species that can only be diagnosed using genetic data \cite{Stuart2006,Bickford2007,SchlickSteiner2007,Pfenninger2007,Clare2011,Funk2012}.
+In attempts to streamline the documentation of biodiversity, several methods of species delimitation that rely almost entirely on genetic data have recently been proposed \cite{Pons2006,Carstens2010,Hausdorf2010,O'Meara2010,Yang2010b,Huelsenbeck2011b}.
+Although strong caveats on the utility of these methods have been raised \cite{Bauer2000,Carstens2013}, they are nevertheless being used to name species \cite{Leache2010,Spinks2014}.
 
 In contrast to those genetically-diagnosed species, the majority of extant taxa, and almost all extinct taxa, are delimited by morphology alone.
-This disjunction complicates interpretations of variation and diversity in deep time, as apparent morphological stasis may not reflect the true underlying diversity \citep{Eldredge1972,Gould1977a,VanBocxlaer2013}.
+This disjunction complicates interpretations of variation and diversity in deep time, as apparent morphological stasis may not reflect the true underlying diversity \cite{Eldredge1972,Gould1977a,VanBocxlaer2013}.
 It also has serious implications for our records of modern biodiversity: for many museum specimens of extant taxa (e.g.
@@
 This would make the task of identifying and maintaining endangered or conserved groups much easier and could contribute to improved classifications of extinct taxa and populations.
-Most such studies focus on using morphometrics to discover differences between taxa that were identified by other means \citep{Polly2003,Zelditch2004,Gaubert2005b,Gunduz2007,Polly2007a,Demandt2009,Markolf2013,Fruciano2016}.
-Additionally, there has been work on automated taxon identification and classification of taxa into groups \citep{Baylac2003,Dobigny2003,MacLeod2007,VandenBrink2011,Vitek2017}, as well as the development of models that combine genetic, phenotypic, and geographic data to infer evolutionary units of interest \citep{Guillot2012}.
+Most such studies focus on using morphometrics to discover differences between taxa that were identified by other means \cite{Polly2003,Zelditch2004,Gaubert2005b,Gunduz2007,Polly2007a,Demandt2009,Markolf2013,Fruciano2016}.
+Additionally, there has been work on automated taxon identification and classification of taxa into groups \cite{Baylac2003,Dobigny2003,MacLeod2007,VandenBrink2011,Vitek2017}, as well as the development of models that combine genetic, phenotypic, and geographic data to infer evolutionary units of interest \cite{Guillot2012}.
 
@@
 \subsection*{Background and study system}
-Machine learning is an extension of known statistical methodology \citep{Hastie2009} that emphasizes predictive accuracy and generality often at the expense of the interpretability of individual parameters.
-Basic statistical approaches are supplemented by randomization, sorting, and partitioning algorithms, along with the maximization or minimization of summary statistics, in order to best estimate a general model for all data, both sampled and unsampled \citep{Hastie2009}.
-Machine learning approaches have found use in medical research, epidemiology, economics, and automated identification of images such as handwritten zip codes \citep{Hastie2009}.
+Machine learning is an extension of known statistical methodology \cite{Hastie2009} that emphasizes predictive accuracy and generality often at the expense of the interpretability of individual parameters.
+Basic statistical approaches are supplemented by randomization, sorting, and partitioning algorithms, along with the maximization or minimization of summary statistics, in order to best estimate a general model for all data, both sampled and unsampled \cite{Hastie2009}.
+Machine learning approaches have found use in medical research, epidemiology, economics, and automated identification of images such as handwritten zip codes \cite{Hastie2009}.
 % additional/more specific CITATIONS?
@@
 There are two major classes of machine learning method: unsupervised and supervised learning.
-Unsupervised learning methods are used with unlabeled data where the underlying structure is estimated; they are analogous to clustering and density estimation methods \citep{Kaufman1990}.
+Unsupervised learning methods are used with unlabeled data where the underlying structure is estimated; they are analogous to clustering and density estimation methods \cite{Kaufman1990}.
 Supervised learning methods are used with labeled data where the final output of data is known and the rules for going from input to output are inferred.
-These are analogous to classification and regression models \citep{Breiman1984,Hastie2009}.
+These are analogous to classification and regression models \cite{Breiman1984,Hastie2009}.
 Our application of the supervised learning approaches used in this study illustrates only a sampling of the various methods available for fitting classification models.
@@
 
-Geometric morphometric approaches to identifying differences in morphological variation between classes, including cryptic species, have mostly relied on methods like linear discriminate analysis and canonical variates analysis \citep{Polly2003,Zelditch2004,Gaubert2005b,Gunduz2007,Polly2007a,Francoy2009,Sztencel-Jabonka2009,Edwards2011,MitrovskiBogdanovic2013,Dillard2017}.
+Geometric morphometric approaches to identifying differences in morphological variation between classes, including cryptic species, have mostly relied on methods like linear discriminate analysis and canonical variates analysis \cite{Polly2003,Zelditch2004,Gaubert2005b,Gunduz2007,Polly2007a,Francoy2009,Sztencel-Jabonka2009,Edwards2011,MitrovskiBogdanovic2013,Dillard2017}.
 Because of their similarity to multivariate approaches like principal components analysis (PCA), these methods are comparatively straightforward ways of understanding the differences in morphology between classes.
@@
 Most previous morphometric studies did not assess which amongst a set of alternative classification hypotheses was optimal.
-For example, studies such as those of \citet{Caumul2005a} and \citet{Polly2007a} focused on comparing different aspects of morphology and their fidelity to a classification scheme instead of comparing the fidelity of one aspect of morphology to multiple classification schemes.
-In this context, the study of \citet{Cardini2009a} is noteworthy because they compared morphological variation in marmots at the population, regional, and species level and determined the fidelity of shape to divisions at each of these levels.
+For example, studies such as those of \cite{Caumul2005a} and \cite{Polly2007a} focused on comparing different aspects of morphology and their fidelity to a classification scheme instead of comparing the fidelity of one aspect of morphology to multiple classification schemes.
+In this context, the study of \cite{Cardini2009a} is noteworthy because they compared morphological variation in marmots at the population, regional, and species level and determined the fidelity of shape to divisions at each of these levels.
 
@@
 Each of these methods provide different advantages for understanding how to classify specimens, as well as the accuracy of the resulting classifications.
-Machine learning methods have been combined with geometric morphometric data to study shape variation in a variety of contexts, including automated taxon identification and classification of groups \citep{Baylac2003,Dobigny2003,MacLeod2007,VanBocxlaer2010,VandenBrink2011,Navega2015}.
+Machine learning methods have been combined with geometric morphometric data to study shape variation in a variety of contexts, including automated taxon identification and classification of groups \cite{Baylac2003,Dobigny2003,MacLeod2007,VanBocxlaer2010,VandenBrink2011,Navega2015}.
 % MORE
 %Although machine learning methods such as neural networks have been applied to studying shape variation, including in the context of automated taxon identification and classification of groups, the number of cases remains limited.
-In the current study, we not only consider pure classification accuracy but also use a statistic of classification strength that reflects the rate at which taxa are both accurately and inaccurately classified: the area under the Receiver Operating Characteristic curve \citep{Hastie2009}.
+In the current study, we not only consider pure classification accuracy but also use a statistic of classification strength that reflects the rate at which taxa are both accurately and inaccurately classified: the area under the Receiver Operating Characteristic curve \cite{Hastie2009}.
 
-We analyzed the problem of whether there are distinct subspecies or cryptic species within the western pond turtle, \textit{Emys marmorata} \citep{Baird1852} (formerly \emph{Clemmys marmorata}; see \citealp{Feldman2002}).
-\textit{Emys marmorata} is distributed from northern Washington State, USA to Baja California, Mexico; populations in western Nevada may have been introduced by recent human activity or they could be a genuine part of the species' range \citep{Bury2017}.
+We analyzed the problem of whether there are distinct subspecies or cryptic species within the western pond turtle, \textit{Emys marmorata} \cite{Baird1852} (formerly \emph{Clemmys marmorata}; see \cite{Feldman2002}).
+\textit{Emys marmorata} is distributed from northern Washington State, USA to Baja California, Mexico; populations in western Nevada may have been introduced by recent human activity or they could be a genuine part of the species' range \cite{Bury2017}.
 Traditionally, \textit{E.
 marmorata} was classified into two named subspecies: the northern \textit{E.
-marmorata marmorata} and the southern \textit{Emys marmorata pallida} \citep{Seeliger1945}, with a central Californian intergrade zone in between.
+marmorata marmorata} and the southern \textit{Emys marmorata pallida} \cite{Seeliger1945}, with a central Californian intergrade zone in between.
 \textit{Emys marmorata marmorata} is differentiated from \textit{E.
@@
 marmorata pallida} although they are considerably smaller.
-\citet{Seeliger1945} did not formally include the Baja California populations of \textit{E.
+\cite{Seeliger1945} did not formally include the Baja California populations of \textit{E.
 marmorata} in either taxon, implying the existence of a third distinct but unnamed subspecies.
@@
 Previous work on morphological variation in \textit{E.
-marmorata} has focused primarily on differentiation between populations over a portion of the species' total range \citep{Lubcke2007,Germano2008,Germano2009,Bury2010}; comparatively few studies have included specimens from across the entire range \citep{Holland1992}.
-Most of these studies considered how local biotic and abiotic factors may contribute to differences in carapace length, and they found that size can vary greatly between different populations \citep{Lubcke2007,Germano2008,Germano2009}.
+marmorata} has focused primarily on differentiation between populations over a portion of the species' total range \cite{Lubcke2007,Germano2008,Germano2009,Bury2010}; comparatively few studies have included specimens from across the entire range \cite{Holland1992}.
+Most of these studies considered how local biotic and abiotic factors may contribute to differences in carapace length, and they found that size can vary greatly between different populations \cite{Lubcke2007,Germano2008,Germano2009}.
 There also has been interest in size-based sexual dimorphism in \textit{E.
-marmorata} \citep{Holland1992,Lubcke2007,Germano2009}, with males being on average larger than females based on total carapace length and other linear measurements.
-However, the quality of size as a classifier of sex can vary greatly between populations \citep{Holland1992} because of the magnitude of size differences among populations \citep{Lubcke2007,Germano2009}.
-The effect of sexual dimorphism on shape, \textit{sensu} \citet{Kendall1977a}, has not been assessed \citep{Holland1992,Lubcke2007,Germano2008}.
+marmorata} \cite{Holland1992,Lubcke2007,Germano2009}, with males being on average larger than females based on total carapace length and other linear measurements.
+However, the quality of size as a classifier of sex can vary greatly between populations \cite{Holland1992} because of the magnitude of size differences among populations \cite{Lubcke2007,Germano2009}.
+The effect of sexual dimorphism on shape, \textit{sensu} \cite{Kendall1977a}, has not been assessed \cite{Holland1992,Lubcke2007,Germano2008}.
 
 Of particular relevance in the context of cryptic diversity in \textit{E.
-marmorata} is the morphometric analysis of carapace shape carried out by \citet{Holland1992}, who compared populations of \textit{E.
+marmorata} is the morphometric analysis of carapace shape carried out by \cite{Holland1992}, who compared populations of \textit{E.
 marmorata} from three areas of the species' range.
@@
 marmorata} is best classified as three distinct species: a northern species, a southern species, and a Columbia Basin species.
-This classification is similar to that of \citet{Seeliger1945}, except elevated to the species level and without recognition of a distinct Baja species.
+This classification is similar to that of \cite{Seeliger1945}, except elevated to the species level and without recognition of a distinct Baja species.
 
 More recently, the phylogeography of \textit{E.
-marmorata} and the possibility of cryptic diversity was investigated using molecular data \citep{Spinks2005,Spinks2010,Spinks2014}.
-Based on mitochondrial DNA, \citet{Spinks2005} recognized four subclades within \textit{E.
+marmorata} and the possibility of cryptic diversity was investigated using molecular data \cite{Spinks2005,Spinks2010,Spinks2014}.
+Based on mitochondrial DNA, \cite{Spinks2005} recognized four subclades within \textit{E.
 marmorata}, a northern clade, a San Joaquin Valley clade, a Santa Barbara clade, and a southern clade.
-Analyses with nuclear DNA \citep{Spinks2010} and single-nucleotide polymorphism (SNP) data suggest a primarily north--south division in \textit{E.
-marmorata}, although these datasets differed from that of mitocondrial-based results of \citet{Spinks2005} in the location of the break point \citep{Spinks2014}.
-All three studies discussed the potential taxonomic implications of their results, with \citet{Spinks2014} going so far as to strongly advocate for the recognition of at least two species (\emph{E.
+Analyses with nuclear DNA \cite{Spinks2010} and single-nucleotide polymorphism (SNP) data suggest a primarily north--south division in \textit{E.
+marmorata}, although these datasets differed from that of mitocondrial-based results of \cite{Spinks2005} in the location of the break point \cite{Spinks2014}.
+All three studies discussed the potential taxonomic implications of their results, with \cite{Spinks2014} going so far as to strongly advocate for the recognition of at least two species (\emph{E.
 marmorata} and \emph{E.
 pallida}), and a possible third based on populations in Baja California.
-However, they did not discuss in detail the morphological characters that would help to diagnose these species beyond those specified by \citet{Seeliger1945}.
-Given that these characters are variable within the proposed species, and that \citet{Holland1992} described shell shape variation that might be consistent with this taxonomy, a geometric morphometric analysis of shell shape might provide a reliable way to diagnose groups (whether species or subspecies) within \textit{E.
+However, they did not discuss in detail the morphological characters that would help to diagnose these species beyond those specified by \cite{Seeliger1945}.
+Given that these characters are variable within the proposed species, and that \cite{Holland1992} described shell shape variation that might be consistent with this taxonomy, a geometric morphometric analysis of shell shape might provide a reliable way to diagnose groups (whether species or subspecies) within \textit{E.
 marmorata}.
@@
 This approach allows both museum specimens and individuals in the field to be analyzed together.
-Second, previous work has suggested that there are strong differences in plastron shape among traditionally-recognized emydine species \citep{Angielczyk2007,Angielczyk2011,Angielczyk2013a}.
+Second, previous work has suggested that there are strong differences in plastron shape among traditionally-recognized emydine species \cite{Angielczyk2007,Angielczyk2011,Angielczyk2013a}.
 Finally, due to these previous studies a large dataset was readily available.
@@
 Because of unclear geographic boundaries between subgroups of \textit{E.
-marmorata}, we compare multiple permutations of the \citep{Spinks2010} and \citet{Spinks2014} hypotheses.
+marmorata}, we compare multiple permutations of the \cite{Spinks2010} and \cite{Spinks2014} hypotheses.
 
-\section*{Methods}
+\section*{Materials and methods}
 \subsection*{Specimens, sampling, morphometrics}
@@
 The first two datasets are intended to serve as a test of whether machine learning techniques can differentiate species-level groupings of emydine turtles using plastron shape.
-We expect that the first case represents a low complexity dataset because of the high level of plastron shape disparity that exists among these species \citep{Claude2003a,Claude2006,Angielczyk2011}, whereas the second dataset should be relatively higher in complexity and more analogous to the \textit{E.
+We expect that the first case represents a low complexity dataset because of the high level of plastron shape disparity that exists among these species \cite{Claude2003a,Claude2006,Angielczyk2011}, whereas the second dataset should be relatively higher in complexity and more analogous to the \textit{E.
 marmorata} example.
 We predict that the \textit{E.
-marmorata} dataset should be of the highest complexity and our greatest challenge given the finding that only very subtle differences existed between geographically-distinct populations \citep{Holland1992}.
-%We expect the first dataset represents a low-complexity dataset given the high level of plastron shape disparty that exists among these species \citep{Angielczyk2011}, whereas the second dataset should be relatively higher in complexity than the first.
+marmorata} dataset should be of the highest complexity and our greatest challenge given the finding that only very subtle differences existed between geographically-distinct populations \cite{Holland1992}.
+%We expect the first dataset represents a low-complexity dataset given the high level of plastron shape disparty that exists among these species \cite{Angielczyk2011}, whereas the second dataset should be relatively higher in complexity than the first.
 Ultimately, we expect that the third dataset to be of the highest complexity and our biggest analytical challenge.
 For all datasets we chose to focus on adults because significant changes in plastron shape occur over the course of ontogeny in \textit{E.
-marmorata} and other emydines \citep{Angielczyk2013a}.
+marmorata} and other emydines \cite{Angielczyk2013a}.
 % three turtle datasets
@@
 The first dataset we analyzed includes 578 total specimens from the following species: \textit{Chrysemys picta}, \textit{Clemmys guttata}, \textit{Emys blandigii}, \textit{Emys orbicularis}, \textit{Glyptemys insculpta}, \textit{Glyptemys muhlenbergii}, \textit{Terrapene coahuila}, and \textit{Terrapene ornata}.
-These specimens are a subset of those used in \citet{Angielczyk2011} and \citet{Angielczyk2013a}.
+These specimens are a subset of those used in \cite{Angielczyk2011} and \cite{Angielczyk2013a}.
 
@@
 \ref{fig:map}).
-These specimens represent a subset of those included in \citet{Angielczyk2007}, \citet{Angielczyk2011}, and \citet{Angielczyk2013a}.
-Because  \citet{Spinks2005}, \citet{Spinks2010}, and \citet{Spinks2014} did not use vouchered specimens we were not able to directly sample the individuals in their studies.
+These specimens represent a subset of those included in \cite{Angielczyk2007}, \cite{Angielczyk2011}, and \cite{Angielczyk2013a}.
+Because  \cite{Spinks2005}, \cite{Spinks2010}, and \cite{Spinks2014} did not use vouchered specimens we were not able to directly sample the individuals in their studies.
 Instead, our specimen classifications were based solely on the geographic information and not explicit assignment using molecular data.
@@
 
-For \citet{Spinks2010} we used three binning schemes.
+For \cite{Spinks2010} we used three binning schemes.
 All three schemes include a class for \textit{E.
@@
 Scheme SP10.1 and SP10.2 differ in the assignment of specimens from the western San Joaquin Valley to either CCR or marm reflecting uncertainty regarding their genetic affinity as explained above.
-In scheme SP10.3 these specimens are assigned to a San Joaquin class reflecting the mitochondrial distinctiveness shown by \citet{Spinks2005}.
-For \citet{Spinks2014} we used two binning schemes with SP14.1 being based on their phylogenetic network analysis and SP14.2 being based on their Bayesian species delimitation analysis.
-The latter scheme requires the addition of two new classes, ``Baja'' and ``Foothill,'' to accommodate the genetic groupings recovered by the SNP Structure analysis that was used to create the guide tree for the BPP species delimitation analysis in \citet{Spinks2014}.
+In scheme SP10.3 these specimens are assigned to a San Joaquin class reflecting the mitochondrial distinctiveness shown by \cite{Spinks2005}.
+For \cite{Spinks2014} we used two binning schemes with SP14.1 being based on their phylogenetic network analysis and SP14.2 being based on their Bayesian species delimitation analysis.
+The latter scheme requires the addition of two new classes, ``Baja'' and ``Foothill,'' to accommodate the genetic groupings recovered by the SNP Structure analysis that was used to create the guide tree for the BPP species delimitation analysis in \cite{Spinks2014}.
 Finally, we proposed a conservative morphological hypothesis (``Morph'') in order to compare the molecular hypotheses with something approximating the original taxonomic hypothesis for the group; this scheme is made up solely of the marm and pall classes from the SP10.3 scheme.
@@
       \hline
-      SP10.1 & 3 & \citet{Spinks2010} \\
-      SP10.2 & 3 & \citet{Spinks2010} \\
-      SP10.3 & 4 & \citet{Spinks2010} \\
-      SP14.1 & 2 & \citet{Spinks2014} \\
-      SP14.2 & 4 & \citet{Spinks2014} \\
-      Morph & 2 & \citet{Spinks2010} \\
+      SP10.1 & 3 & \cite{Spinks2010} \\
+      SP10.2 & 3 & \cite{Spinks2010} \\
+      SP10.3 & 4 & \cite{Spinks2010} \\
+      SP14.1 & 2 & \cite{Spinks2014} \\
+      SP14.2 & 4 & \cite{Spinks2014} \\
+      Morph & 2 & \cite{Spinks2010} \\
       \hline
@@
 marmorata} species complex (see below).
-%The scheme names are as follows: Mito 1 and 2 correspond to \citet{Spinks2005}, Mito 3 corresponds to \citet{Spinks2010}, Morph 1 and Morph 2 correspond to \citet{Holland1992}, and Nuclear corresponds to \citet{Spinks2014}.
+%The scheme names are as follows: Mito 1 and 2 correspond to \cite{Spinks2005}, Mito 3 corresponds to \cite{Spinks2010}, Morph 1 and Morph 2 correspond to \cite{Holland1992}, and Nuclear corresponds to \cite{Spinks2014}.
 % CHANGE
 
-Following previous work on plastron shape \citep{Angielczyk2007,Angielczyk2011,Angielczyk2013a}, we used TpsDig 2.04 \citep{Rohlf2005} to digitize 19 two-dimensional landmarks (Fig.
+Following previous work on plastron shape \cite{Angielczyk2007,Angielczyk2011,Angielczyk2013a}, we used TpsDig 2.04 \cite{Rohlf2005} to digitize 19 two-dimensional landmarks (Fig.
 \ref{fig:plastra}).
@@
 We conducted all subsequent analyses on the resulting ``half'' plastra.
-We superimposed the plastral landmark configurations using generalized Procrustes analysis \citep{Dryden1998a}, after which we calculated the principal components (PC) of shape using the \texttt{shapes} package for R \citep{R2016,Dryden2013}.
+We superimposed the plastral landmark configurations using generalized Procrustes analysis \cite{Dryden1998a}, after which we calculated the principal components (PC) of shape using the \texttt{shapes} package for R \cite{R2016,Dryden2013}.
 All specimens were used for superimposition, after which the subset labeled for each of the schemes were used in model training and testing (see below).
@@
 % digitization
-We estimated the possible effect of digitization error \citep{Arnqvist1998,Cramon2007,Munoz-MunozF.2010} on our results by comparing within-specimen (replicated) Procrustes distances to the distances between classification scheme centroids.
+We estimated the possible effect of digitization error \cite{Arnqvist1998,Cramon2007,Munoz-MunozF.2010} on our results by comparing within-specimen (replicated) Procrustes distances to the distances between classification scheme centroids.
 Ten randomly-selected \textit{E.
@@
 
-\textit{Emys marmorata} is known to display sexual dimorphism in plastral shape, particularly the presence of a plastra concavity in males \citep{Seeliger1945}.
+\textit{Emys marmorata} is known to display sexual dimorphism in plastral shape, particularly the presence of a plastra concavity in males \cite{Seeliger1945}.
 To test for biases resulting from sexual dimorphism in our \textit{E.
@@
     \hline
-    multinomial logistic regression & MLR & nnet & \citet{Venables2002} \\
-    linear discriminate analysis & LDA & MASS & \citet{Venables2002} \\
-    penalized discrminiate analysis & PDA & mda & \citet{mdapack} \\
-    single-hidden-layer neural network & NN & nnet & \citet{Venables2002} \\
-    random forests & RF & randomForest & \citet{Liaw2002} \\
+    multinomial logistic regression & MLR & nnet & \cite{Venables2002} \\
+    linear discriminate analysis & LDA & MASS & \cite{Venables2002} \\
+    penalized discrminiate analysis & PDA & mda & \cite{mdapack} \\
+    single-hidden-layer neural network & NN & nnet & \cite{Venables2002} \\
+    random forests & RF & randomForest & \cite{Liaw2002} \\
     \hline
@@
 
-Instead of relying on a single supervised learning method, we chose to use an ensemble approach where multiple model types are used in concert so that any congruence between them increases our support for that conclusion over another \citep{Hastie2009}.
+Instead of relying on a single supervised learning method, we chose to use an ensemble approach where multiple model types are used in concert so that any congruence between them increases our support for that conclusion over another \cite{Hastie2009}.
 The supervised learning methods used here are named in Table \ref{tab:methods}.
-Each of these methods makes different assumptions, treats data differently, and can produce different classification results depending on the nature of the data \citep{Hastie2009}.
+Each of these methods makes different assumptions, treats data differently, and can produce different classification results depending on the nature of the data \cite{Hastie2009}.
 For example, multinomial logistic regression is a type of generalized linear model, whereas random forest is itself an ensemble approach where multiple decision trees are fit to subsets of the full dataset and then averaged.
@@
 Additional interaction terms were not considered because of model complexity/sample size concerns.
-Size and the interaction between size and PC 1 were included as predictors to account for known ontogenetic variation in plastron shape \citep{Angielczyk2013a} as well as potential size differences between classes, even if this is unlikely \citep{Seeliger1945,Holland1992}.
+Size and the interaction between size and PC 1 were included as predictors to account for known ontogenetic variation in plastron shape \cite{Angielczyk2013a} as well as potential size differences between classes, even if this is unlikely \cite{Seeliger1945,Holland1992}.
 These data constitute a ``maximum set'' because the best or selected models based on five-fold cross-validation need not, and likely will not, include all predictors possible (see below).
 Because our supervised learning models use PCs as predictors, this approach is in many ways analogous to PCA regression.
-PCA regression takes advantage of reduction and orthogonality PCs to improve regression fit \citep{Hastie2009}.
+PCA regression takes advantage of reduction and orthogonality PCs to improve regression fit \cite{Hastie2009}.
 Because the PCs of shape are by definition orthogonal, they can easily serve as independent predictors or features of class membership without fear of collinearity.
@@
 % caret package for R
-We adopted a training and testing paradigm for selecting parsimonious models and estimating their overall error rates \citep{Hastie2009,Kuhn2013}.
+We adopted a training and testing paradigm for selecting parsimonious models and estimating their overall error rates \cite{Hastie2009,Kuhn2013}.
 Within-sample model performance is inherently biased upwards, so model evaluation requires overcoming this bias.
@@
 %   better performance than accuracy when groups unbalanced
-In classification studies, such as this one, a common metric of performance is the receiver operating characteristic (ROC) which is the relationship between the false and true positive rates \citep{Hastie2009}.
-The area under the ROC curve (AUC) is the derived estimate of the model performance; AUC ranges from 0.5 to 1 which correspond to performance similar to random guesses and perfect classification rates, respectively \citep{Hastie2009}.
-Both ROC and AUC are preferable to simple classification accuracy when class membership is unbalanced, as it is in these analyses \citep{Hastie2009}.
+In classification studies, such as this one, a common metric of performance is the receiver operating characteristic (ROC) which is the relationship between the false and true positive rates \cite{Hastie2009}.
+The area under the ROC curve (AUC) is the derived estimate of the model performance; AUC ranges from 0.5 to 1 which correspond to performance similar to random guesses and perfect classification rates, respectively \cite{Hastie2009}.
+Both ROC and AUC are preferable to simple classification accuracy when class membership is unbalanced, as it is in these analyses \cite{Hastie2009}.
 The standard ROC and AUC calculations are defined only for binary classifications, which is not the case for our eight species and \textit{Emys} complex datasets.
-To generalize this approach for situations with multiple response classes, we used an all-against-one strategy where the model AUC is the average of the AUC values from the multiple binary comparisons of one class compared to all others \citep{Hand2001}.
+To generalize this approach for situations with multiple response classes, we used an all-against-one strategy where the model AUC is the average of the AUC values from the multiple binary comparisons of one class compared to all others \cite{Hand2001}.
 
 For a given supervised learning method, we compared the fit of 27 models as the average AUC from 10 rounds of five-fold cross-validation.
-Cross-validation is an approach for estimating the average out-of-sample predictive error of a model by simulating out-of-sample data from the training dataset itself \citep{Hastie2009}.
+Cross-validation is an approach for estimating the average out-of-sample predictive error of a model by simulating out-of-sample data from the training dataset itself \cite{Hastie2009}.
 In a single round of \(k\)-fold cross-validation, the training data are divided into \(k\) blocks where the model is fit to \(k - 1\) blocks and the values of the \(k\)th block are predicted.
@@
 Finally, the predictive performance metric is the averaged across all rounds of \(k\)-fold cross-validation.
-This process was implemented using the R package \texttt{caret} \citep{KuhnMAN2013}.
+This process was implemented using the R package \texttt{caret} \cite{KuhnMAN2013}.
 For a given supervised learning method, the ``best'' trained model is that with the highest mean AUC as estimated from five-fold cross-validation.
-The selected or final model, however, is the next most parsimonious model that is within one standard error of the best model; this is a variant on the ``one-standard error'' rule from \citet{Hastie2009}.
+The selected or final model, however, is the next most parsimonious model that is within one standard error of the best model; this is a variant on the ``one-standard error'' rule from \cite{Hastie2009}.
 The purpose of this rule is to ameliorate the chances of selecting an overly complex model that will perform poorly when predicting the classes of out-of-sample data.
@@
 from hydrological regime) overwhelm morphological differentiation.
-Both of these options seem plausible given that shell shape is influenced by selection for both protection and streamlining, but not necessary mate choice \citep{Rivera2008,Rivera2011,Stayton2011,Rivera2014,Polly2016}, and that shell shape in \textit{E.
-marmorata} is known to vary among populations inhabiting water bodies with different flow regimes \citep{Holland1992,Lubcke2007,Germano2009}.
-Plastron shape does not seem to preserve a strong phylogenetic signal at the interspecific level in emydine turtles, at least compared to the effect of the presence or absence of a plastral hinge \citep{Angielczyk2011}, and our current results suggest that this may be the case for phylogeographic signal within emydine species as well.
+Both of these options seem plausible given that shell shape is influenced by selection for both protection and streamlining, but not necessary mate choice \cite{Rivera2008,Rivera2011,Stayton2011,Rivera2014,Polly2016}, and that shell shape in \textit{E.
+marmorata} is known to vary among populations inhabiting water bodies with different flow regimes \cite{Holland1992,Lubcke2007,Germano2009}.
+Plastron shape does not seem to preserve a strong phylogenetic signal at the interspecific level in emydine turtles, at least compared to the effect of the presence or absence of a plastral hinge \cite{Angielczyk2011}, and our current results suggest that this may be the case for phylogeographic signal within emydine species as well.
 A final possibility (explored below) is that the proposed classification schemes themselves do not represent significant evolutionary lineages.
@@
 pallida} if they are distinct species.
-Invoking issues such as the role of the plastron in protection or the need for streamlining are insufficient because the other species are expected to be subject to similar constraints \citep{Stayton2011,Polly2016}.
+Invoking issues such as the role of the plastron in protection or the need for streamlining are insufficient because the other species are expected to be subject to similar constraints \cite{Stayton2011,Polly2016}.
 Although it may seem counterintuitive that plastron shape is both useful for species delimitation but has weak or absent phylogenetic signal, it is important to remember that these are different goals.
@@
 
-\citet{Spinks2014} elevated \textit{E.
-pallida} based on a species delimitation analysis of SNP data using BPP \citep{Yang2010b}.
-However, \citet{Spinks2014} did not heed the caveats about such species delimitation methods raised by \citet{Carstens2013}.
-In addition to specifically addressing the shortcomings of validation methods such as BPP that rely on guide trees and ``should be interpreted with caution,'' \citet{Carstens2013} also strongly emphasized that ``Inferences regarding species boundaries based on genetic data alone are likely inadequate, and species delimitation should be conducted with consideration of the life history, geographical distribution, morphology and behaviour (where applicable) of the focal system\dots'' These caveats evoke the development of the Unified Species Concept \citep{Dayrat2005a,DeQueiroz2007b}, Integrative Taxonomy \citep{Padial2010}, and other pluralist approaches to species delimitation.
+\cite{Spinks2014} elevated \textit{E.
+pallida} based on a species delimitation analysis of SNP data using BPP \cite{Yang2010b}.
+However, \cite{Spinks2014} did not heed the caveats about such species delimitation methods raised by \cite{Carstens2013}.
+In addition to specifically addressing the shortcomings of validation methods such as BPP that rely on guide trees and ``should be interpreted with caution,'' \cite{Carstens2013} also strongly emphasized that ``Inferences regarding species boundaries based on genetic data alone are likely inadequate, and species delimitation should be conducted with consideration of the life history, geographical distribution, morphology and behaviour (where applicable) of the focal system\dots'' These caveats evoke the development of the Unified Species Concept \cite{Dayrat2005a,DeQueiroz2007b}, Integrative Taxonomy \cite{Padial2010}, and other pluralist approaches to species delimitation.
 None of these considerations were brought to bear on the \textit{E.
@@
 pallida} also make the recognition of these two taxa implausible.
-The mitochondrial data from \citet{Spinks2014} show extensive introgression and admixture in Central California, which is expected because there are no significant barriers to gene flow in this region.
+The mitochondrial data from \cite{Spinks2014} show extensive introgression and admixture in Central California, which is expected because there are no significant barriers to gene flow in this region.
 They also lack sampling from the populations between the two putative species in the San Francisco Bay Area, which we predict would likely show even more genetic mixing.
 Combined with the well-demonstrated ability for testudinoid turtles, including emydids and even \textit{Emys}, to hybridize (e.g.
-\citealt{Buskirk2005,Spinks2009,Parham2013}) it is hard to imagine how \textit{E.
+\cite{Buskirk2005,Spinks2009,Parham2013}) it is hard to imagine how \textit{E.
 marmorata} and \textit{E.
@@
 
-We fully agree with \citet{Spinks2014} that \textit{E.
+We fully agree with \cite{Spinks2014} that \textit{E.
 marmorata} (\textit{sensu lato}) is a species deserving of strong conservation efforts, and we do not wish to trivialize this need.
-Moreover, the genetic diversity uncovered by the analysis of \citet{Spinks2014} should be accounted for explicitly in any conservation plan.
+Moreover, the genetic diversity uncovered by the analysis of \cite{Spinks2014} should be accounted for explicitly in any conservation plan.
 Given the apparent lack of morphological distinction combined with the broad range of intergradation and other problems with the species hypothesis outlined above, we recommend that the populations elevated to \textit{E.
-pallida} by \citet{Spinks2014} are best considered Evolutionary Significant Units or Distinct Population Segments instead of distinct species.
+pallida} by \cite{Spinks2014} are best considered Evolutionary Significant Units or Distinct Population Segments instead of distinct species.
 
@@
 However, tests of both our preferred conclusion (\textit{E.
-marmorata} as a single species) and that of \citet{Spinks2014} should include morphological and molecular analyses of the same set of voucher specimens, as well as additional tests of species delimitation using alternative methods and corroborating evidence as suggested by \citet{Carstens2013}.
+marmorata} as a single species) and that of \cite{Spinks2014} should include morphological and molecular analyses of the same set of voucher specimens, as well as additional tests of species delimitation using alternative methods and corroborating evidence as suggested by \cite{Carstens2013}.
 From a morphological standpoint, support for the validity of ``\textit{E.
@@
 
-\section*{Acknowledgements}
+\section*{Supporting information}
+
+% Include only the SI item label in the paragraph heading. Use the \nameref{label} command to cite SI items in the text.
+\paragraph*{S1 Fig.}
+\label{S1_Fig}
+{\bf Bold the title sentence.} Add descriptive text after the title of the item (optional).
+
+\section*{Acknowledgments}
 Data collection for this project was supported in part by NSF DBI-0306158 (to KDA).
@@
 
+\nolinenumbers
+
+% Either type in your references using
+% \begin{thebibliography}{}
+% \bibitem{}
+% Text
+% \end{thebibliography}
+%
+% or
+%
+% Compile your BiBTeX database using our plos2015.bst
+% style file and paste the contents of your .bbl file
+% here. See http://journals.plos.org/plosone/s/latex for 
+% step-by-step instructions.
+% 
+%\begin{thebibliography}{10}
+%
+%\end{thebibliography}
 \bibliography{new_bib}

### 2018-02-28 doc/turtles_plosone.tex
plos one draft to coauthors

@@
 \subsection*{Background and study system}
-Machine learning is an extension of known statistical methodology \cite{Hastie2009} that emphasizes predictive accuracy and generality often at the expense of the interpretability of individual parameters.
+Machine learning is an extension of known statistical methodology that emphasizes predictive accuracy and generality often at the expense of the interpretability of individual parameters \cite{Hastie2009}.
 Basic statistical approaches are supplemented by randomization, sorting, and partitioning algorithms, along with the maximization or minimization of summary statistics, in order to best estimate a general model for all data, both sampled and unsampled \cite{Hastie2009}.
@@
 Most previous morphometric studies did not assess which amongst a set of alternative classification hypotheses was optimal.
-For example, studies such as those of \cite{Caumul2005a} and \cite{Polly2007a} focused on comparing different aspects of morphology and their fidelity to a classification scheme instead of comparing the fidelity of one aspect of morphology to multiple classification schemes.
-In this context, the study of \cite{Cardini2009a} is noteworthy because they compared morphological variation in marmots at the population, regional, and species level and determined the fidelity of shape to divisions at each of these levels.
+For example, studies such as those of Camul et al.
+\cite{Caumul2005a} and Polly \cite{Polly2007a} focused on comparing different aspects of morphology and their fidelity to a classification scheme instead of comparing the fidelity of one aspect of morphology to multiple classification schemes.
+In this context, the study of Cardini et al.
+\cite{Cardini2009a} is noteworthy because they compared morphological variation in marmots at the population, regional, and species level and determined the fidelity of shape to divisions at each of these levels.
 
@@
 marmorata pallida} although they are considerably smaller.
-\cite{Seeliger1945} did not formally include the Baja California populations of \textit{E.
+Seeliger \cite{Seeliger1945} did not formally include the Baja California populations of \textit{E.
 marmorata} in either taxon, implying the existence of a third distinct but unnamed subspecies.
@@
 However, the quality of size as a classifier of sex can vary greatly between populations \cite{Holland1992} because of the magnitude of size differences among populations \cite{Lubcke2007,Germano2009}.
-The effect of sexual dimorphism on shape, \textit{sensu} \cite{Kendall1977a}, has not been assessed \cite{Holland1992,Lubcke2007,Germano2008}.
+The effect of sexual dimorphism on shape, \textit{sensu} Kendall \cite{Kendall1977a}, has not been assessed \cite{Holland1992,Lubcke2007,Germano2008}.
 
 Of particular relevance in the context of cryptic diversity in \textit{E.
-marmorata} is the morphometric analysis of carapace shape carried out by \cite{Holland1992}, who compared populations of \textit{E.
+marmorata} is the morphometric analysis of carapace shape carried out by Holland \cite{Holland1992}, who compared populations of \textit{E.
 marmorata} from three areas of the species' range.
@@
 marmorata} and the possibility of cryptic diversity was investigated using molecular data \cite{Spinks2005,Spinks2010,Spinks2014}.
-Based on mitochondrial DNA, \cite{Spinks2005} recognized four subclades within \textit{E.
+Based on mitochondrial DNA, Spinks and Shaffer \cite{Spinks2005} recognized four subclades within \textit{E.
 marmorata}, a northern clade, a San Joaquin Valley clade, a Santa Barbara clade, and a southern clade.
 Analyses with nuclear DNA \cite{Spinks2010} and single-nucleotide polymorphism (SNP) data suggest a primarily north--south division in \textit{E.
-marmorata}, although these datasets differed from that of mitocondrial-based results of \cite{Spinks2005} in the location of the break point \cite{Spinks2014}.
-All three studies discussed the potential taxonomic implications of their results, with \cite{Spinks2014} going so far as to strongly advocate for the recognition of at least two species (\emph{E.
+marmorata}, although these datasets differed from that of mitocondrial-based results of Spinks and Shaffer \cite{Spinks2005} in the location of the break point \cite{Spinks2014}.
+All three studies discussed the potential taxonomic implications of their results, with Spinks et al.
+\cite{Spinks2014} going so far as to strongly advocate for the recognition of at least two species (\emph{E.
 marmorata} and \emph{E.
 pallida}), and a possible third based on populations in Baja California.
-However, they did not discuss in detail the morphological characters that would help to diagnose these species beyond those specified by \cite{Seeliger1945}.
-Given that these characters are variable within the proposed species, and that \cite{Holland1992} described shell shape variation that might be consistent with this taxonomy, a geometric morphometric analysis of shell shape might provide a reliable way to diagnose groups (whether species or subspecies) within \textit{E.
+However, they did not discuss in detail the morphological characters that would help to diagnose these species beyond those specified by Seeliger \cite{Seeliger1945}.
+Given that these characters are variable within the proposed species, and that Holland \cite{Holland1992} described shell shape variation that might be consistent with this taxonomy, a geometric morphometric analysis of shell shape might provide a reliable way to diagnose groups (whether species or subspecies) within \textit{E.
 marmorata}.
@@
 Because of unclear geographic boundaries between subgroups of \textit{E.
-marmorata}, we compare multiple permutations of the \cite{Spinks2010} and \cite{Spinks2014} hypotheses.
+marmorata}, we compare multiple permutations of the Spinks et al.
+\cite{Spinks2010} and Spinks et al.
+\cite{Spinks2014} hypotheses.
 
@@
 The first dataset we analyzed includes 578 total specimens from the following species: \textit{Chrysemys picta}, \textit{Clemmys guttata}, \textit{Emys blandigii}, \textit{Emys orbicularis}, \textit{Glyptemys insculpta}, \textit{Glyptemys muhlenbergii}, \textit{Terrapene coahuila}, and \textit{Terrapene ornata}.
-These specimens are a subset of those used in \cite{Angielczyk2011} and \cite{Angielczyk2013a}.
+These specimens are a subset of those used in previous studies \cite{Angielczyk2011,Angielczyk2013a}.
 
@@
 \ref{fig:map}).
-These specimens represent a subset of those included in \cite{Angielczyk2007}, \cite{Angielczyk2011}, and \cite{Angielczyk2013a}.
-Because  \cite{Spinks2005}, \cite{Spinks2010}, and \cite{Spinks2014} did not use vouchered specimens we were not able to directly sample the individuals in their studies.
+These specimens represent a subset of those included in previous studies \cite{Angielczyk2007,Angielczyk2011,Angielczyk2013a}.
+Because previous studies did not use vouchered specimens \cite{Spinks2005,Spinks2010,Spinks2014} we were not able to directly sample the individuals in their studies.
 Instead, our specimen classifications were based solely on the geographic information and not explicit assignment using molecular data.
@@
 The taxonomic hypotheses and sub-hypotheses for \textit{E.
-marmorata} used here are presented in Table \ref{tab:hypotheses} and Figure \ref{fig:map}.
+marmorata} used here are presented in Table \ref{tab:hypotheses} and Fig.
+\ref{fig:map}.
 
-For \cite{Spinks2010} we used three binning schemes.
+For our first three schemes \cite{Spinks2010} we used three binning schemes.
 All three schemes include a class for \textit{E.
@@
 Scheme SP10.1 and SP10.2 differ in the assignment of specimens from the western San Joaquin Valley to either CCR or marm reflecting uncertainty regarding their genetic affinity as explained above.
-In scheme SP10.3 these specimens are assigned to a San Joaquin class reflecting the mitochondrial distinctiveness shown by \cite{Spinks2005}.
-For \cite{Spinks2014} we used two binning schemes with SP14.1 being based on their phylogenetic network analysis and SP14.2 being based on their Bayesian species delimitation analysis.
-The latter scheme requires the addition of two new classes, ``Baja'' and ``Foothill,'' to accommodate the genetic groupings recovered by the SNP Structure analysis that was used to create the guide tree for the BPP species delimitation analysis in \cite{Spinks2014}.
+In scheme SP10.3 these specimens are assigned to a San Joaquin class reflecting the mitochondrial distinctiveness shown by Spinks and Shaffer \cite{Spinks2005}.
+For our second two schemes \cite{Spinks2014} we used two binning schemes with SP14.1 being based on their phylogenetic network analysis and SP14.2 being based on their Bayesian species delimitation analysis.
+The latter scheme requires the addition of two new classes, ``Baja'' and ``Foothill,'' to accommodate the genetic groupings recovered by the SNP Structure analysis that was used to create the guide tree for the BPP species delimitation analysis \cite{Spinks2014}.
 Finally, we proposed a conservative morphological hypothesis (``Morph'') in order to compare the molecular hypotheses with something approximating the original taxonomic hypothesis for the group; this scheme is made up solely of the marm and pall classes from the SP10.3 scheme.
@@
 Because not all of our specimens have sex identifications associated with them, this analysis was done using a subset of the data (257 of 532).
+This analysis was also done for each of the classification schemes in order to determine if sex is a biasing factor within these groups.
+As with the total dataset, only a limited number of specimens in each classification scheme have sex information, and the sex ratios within each of the classification schemes are as follows: SP10.1 202 F/199 M, SP10.2 202/199, SP10.3 202/199, SP14.1 177/178, SP14.2 136/125, Morph 112/120.
 
@@
 % analogy to PCA regression
-The maximum set of possible predictors or features used for any model of our dataset is comprised of the first 25 principal components (PCs), scaled centroid size, and the interaction between scaled centroid size and PC 1.
+The maximum set of possible predictors or features used for any model of our dataset is comprised of the first 25 principal components (PCs), scaled centroid size, the interaction between scaled centroid size and PC 1, and the interaction between scaled centroid size and PC 2.
 Additional interaction terms were not considered because of model complexity/sample size concerns.
@@
 For a given supervised learning method, the ``best'' trained model is that with the highest mean AUC as estimated from five-fold cross-validation.
-The selected or final model, however, is the next most parsimonious model that is within one standard error of the best model; this is a variant on the ``one-standard error'' rule from \cite{Hastie2009}.
+The selected or final model, however, is the next most parsimonious model that is within one standard error of the best model; this is a variant on the ``one-standard error'' rule \cite{Hastie2009}.
 The purpose of this rule is to ameliorate the chances of selecting an overly complex model that will perform poorly when predicting the classes of out-of-sample data.
@@
 
+\afterpage{\clearpage}
+\begin{figure}[h]
+  \centering
+  \includegraphics[height = 0.8\textheight, width = \textwidth, keepaspectratio = true]{figure/sex_test_hist_grouped}
+  \caption{Comparison of observed Procrustes distance between the centroids of each sex (vertical line) to a null distribution generated from 1000 permutations of the sex-labels for each of the six classification schemes.
+This result indicates that the difference between the centroids is as small/smaller than expected by random.
+For explination of classification scheme names see Table \ref{tab:hypotheses}.}
+  \label{fig:sex_test_group}
+\end{figure}
+
 Analysis of the differences between sexes of \textit{E.
@@
 The difference in mean shape between the sexes is very small; the sexes overlap about has much as expected given a null distribution based on permuting the sex-labels.
+The results form sex comparisons within each classification scheme have similar results with none of the observed differences being significantly different from the permuted distribution, indicating that sex is most likely not a biasing factor in our analyses (Fig.
+\ref{fig:sex_test_group}).
 
@@
     \hline
-    Replicates & & 1.11 & \\
+    Replicates & & 1.06 & \\
     \hline
-    Seven species & & 0.33 & 0.37 \\
-    \textit{Trachemys} & & 0.76 & 0.84 \\
+    Seven species & & 0.43 & 0.46 \\
+    \textit{Trachemys} & & 0.76 & 0.81 \\
     \hline
-    \textit{E. marmorata} & SP10.1 & 0.99 & 1.10 \\
-     & SP10.2 & 1.00 & 1.11 \\
-     & SP10.3 & 0.94 & 1.04 \\
-     & SP14.1 & 1.01 & 1.12 \\
-     & SP14.2 & 0.93 & 1.04 \\
-     & Morph & 0.99 & 1.09 \\
+    \textit{E. marmorata} & SP10.1 & 0.99 & 1.05 \\
+     & SP10.2 & 1.00 & 1.06 \\
+     & SP10.3 & 0.95 & 1.01 \\
+     & SP14.1 & 1.01 & 1.07 \\
+     & SP14.2 & 0.94 & 1.00 \\
+     & Morph & 0.99 & 1.05 \\
     \hline
@@
 
+Spinks et al.
 \cite{Spinks2014} elevated \textit{E.
 pallida} based on a species delimitation analysis of SNP data using BPP \cite{Yang2010b}.
-However, \cite{Spinks2014} did not heed the caveats about such species delimitation methods raised by \cite{Carstens2013}.
-In addition to specifically addressing the shortcomings of validation methods such as BPP that rely on guide trees and ``should be interpreted with caution,'' \cite{Carstens2013} also strongly emphasized that ``Inferences regarding species boundaries based on genetic data alone are likely inadequate, and species delimitation should be conducted with consideration of the life history, geographical distribution, morphology and behaviour (where applicable) of the focal system\dots'' These caveats evoke the development of the Unified Species Concept \cite{Dayrat2005a,DeQueiroz2007b}, Integrative Taxonomy \cite{Padial2010}, and other pluralist approaches to species delimitation.
+However, Spinks et al.
+\cite{Spinks2014} did not heed the caveats about such species delimitation methods raised by Carstens et al.
+\cite{Carstens2013}.
+In addition to specifically addressing the shortcomings of validation methods such as BPP that rely on guide trees and ``should be interpreted with caution,'' Carstens et al.
+\cite{Carstens2013} also strongly emphasized that ``Inferences regarding species boundaries based on genetic data alone are likely inadequate, and species delimitation should be conducted with consideration of the life history, geographical distribution, morphology and behaviour (where applicable) of the focal system\dots'' These caveats evoke the development of the Unified Species Concept \cite{Dayrat2005a,DeQueiroz2007b}, Integrative Taxonomy \cite{Padial2010}, and other pluralist approaches to species delimitation.
 None of these considerations were brought to bear on the \textit{E.
@@
 pallida} also make the recognition of these two taxa implausible.
-The mitochondrial data from \cite{Spinks2014} show extensive introgression and admixture in Central California, which is expected because there are no significant barriers to gene flow in this region.
+The mitochondrial data from Spinks et al.
+\cite{Spinks2014} show extensive introgression and admixture in Central California, which is expected because there are no significant barriers to gene flow in this region.
 They also lack sampling from the populations between the two putative species in the San Francisco Bay Area, which we predict would likely show even more genetic mixing.
@@
 
-We fully agree with \cite{Spinks2014} that \textit{E.
+We fully agree with Spinks et al.
+\cite{Spinks2014} that \textit{E.
 marmorata} (\textit{sensu lato}) is a species deserving of strong conservation efforts, and we do not wish to trivialize this need.
-Moreover, the genetic diversity uncovered by the analysis of \cite{Spinks2014} should be accounted for explicitly in any conservation plan.
+Moreover, the genetic diversity uncovered by the analysis of Spinks et al.
+\cite{Spinks2014} should be accounted for explicitly in any conservation plan.
 Given the apparent lack of morphological distinction combined with the broad range of intergradation and other problems with the species hypothesis outlined above, we recommend that the populations elevated to \textit{E.
@@
 However, tests of both our preferred conclusion (\textit{E.
-marmorata} as a single species) and that of \cite{Spinks2014} should include morphological and molecular analyses of the same set of voucher specimens, as well as additional tests of species delimitation using alternative methods and corroborating evidence as suggested by \cite{Carstens2013}.
+marmorata} as a single species) and that of Spinks et al.
+\cite{Spinks2014} should include morphological and molecular analyses of the same set of voucher specimens, as well as additional tests of species delimitation using alternative methods and corroborating evidence as suggested by Carstens et al.
+\cite{Carstens2013}.
 From a morphological standpoint, support for the validity of ``\textit{E.

### 2018-02-28 R/sex_diff.r

@@ -7,6 +7,7 @@ library(scales)
 library(grid)
 source('../R/df2array.r')
 source('../R/supervised_mung.r')
+source('../R/multiplot.r')
 
 split.land <- function(shapes, f) {
   ll <- length(unique(f))
@@ -31,6 +32,9 @@ land.dist <- function(shapes) {
   }
   distmat
 }
+
+
+
 
 
 tsex <- as.character(adult$p.sex)  # lots of regex to get this sexy
@@ -72,3 +76,54 @@ dfgp <- dfgp + scale_x_continuous(breaks = pretty_breaks(4))
 dfgp <- dfgp + labs(x = 'Procrustes distance', y = 'Frequency')
 ggsave(plot = dfgp, filename = '../doc/figure/sex_test_hist.png',
        width = 4, height = 3)
+
+
+# by groups
+gg <- c('sp10.1', 'sp10.2', 'sp10.3', 'sp14.1', 'sp14.2', 'morph')
+ggl <- list()
+for(jj in seq(length(gg))) {
+  tt <- adult[, c(gg[jj], 'p.sex')]
+  tt <- tt[!is.na(tt[, 1]), ]
+  tt[, 2] <- as.character(tt[, 2])
+  fe <- grep(pattern = '[fF]', tt[, 2], perl = TRUE)
+  ma <- grep(pattern = '[mM]', tt[, 2], perl = TRUE)
+  na <- grep(pattern = '[nN]', tt[, 2], perl = TRUE)
+  tt[fe, 2] <- 'F'
+  tt[ma, 2] <- 'M'
+
+  sexk <- which(tt[, 2] %in% c('F', 'M'))
+  tsex <- tt[sexk, 2]
+  sex.shape <- fit$rotated[, , (sexk)]
+
+
+  # reference
+  sex.split <- split.land(sex.shape, tsex)
+  sex.mean <- llply(sex.split, mshape)
+  sex.mean <- sum(rowSums((sex.mean[[1]] - sex.mean[[2]])^2))
+
+  null <- c()
+  for(ii in seq(1000)) {
+    tt <- sample(tsex)
+    ss <- split.land(sex.shape, tt)
+    sm <- llply(ss, mshape)
+    null[ii] <- sum(rowSums((sm[[1]] - sm[[2]])^2))
+  }
+
+  dfg <- data.frame(null = null)
+
+  print(sum(null > sex.mean) / length(null))
+
+  theme_set(theme_minimal())
+  dfgp <- ggplot(dfg, aes(x = null))
[…]

### 2018-03-07 doc/turtles_plosone.tex
ken and jim comments. sex ratio table

@@
   {\Large
-\textbf\newline{Ensemble approaches for estimating congruence between species delimitation and morphological variation: comparing taxonomic hypotheses for the Pacific Pond Turtle (\textit{Emys marmorata})} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
+  \textbf\newline{Ensemble approaches for estimating congruence between species delimitation and morphological variation: a case study using the Pacific Pond Turtle (\textit{Emys marmorata})} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
   }
@@
   We investigated the morphometric identification of cryptic species using machine learning approaches by examining their implications for a recently proposed cryptic turtle species (\textit{Emys pallida}).
-We collected landmark-based morphometric data from 532 adult \textit{E.
+We used landmark-based morphometric data to quantify plastron shape in 532 adult \textit{E.
 marmorata/``E.
@@
 We used an ensemble of supervised machine learning approaches to determine which classification hypothesis was best supported by the data.
-In addition, we applied the same approach to two clear-cut examples, one consisting of eight unambiguously distinct species closely related to \textit{E.
-marmorata}, and the other consisting of two subspecies of \textit{Trachemys scripta}.
-Our results indicate that there is no clear ``best'' grouping of \textit{E.
+In addition, we applied the same approach to two clear-cut examples, one consisting of seven unambiguously distinct species closely related to \textit{E.
+marmorata} and an outgroup (\textit{Chrysemys picta}), and the other consisting of two subspecies of \textit{Trachemys scripta}.
+The analyses of the clear-cut examples produced near perfect classifications, demonstrating that plastron shape typically is a useful marker for differentiating turtle species, and that the methods can recover correct results when an appropriate signa exists.
+There is no clear ``best'' grouping of \textit{E.
 marmorata/``E.
 pallida''} based on plastron shape.
-In contrast, the analyses of the clear-cut examples produced near perfect classifications, demonstrating that the methods can recover correct results when an appropriate signal exists.
 Explanations for the lack of grouping in \textit{E.
@@
 A reconsideration of the methods used to delimit \textit{``E.
-pallida,''} the lack of barriers to gene flow, the strong evidence for widespread admixture between lineages, and the fact that plastron shape can be used to delineate other emydine species and sub-species suggest that its lack of diagnosability most likely reflects the non-distinctiveness of this proposed taxon.
+pallida,''} the lack of barriers to gene flow, the strong evidence for widespread admixture between lineages, and the fact that plastron shape can be used to delineate other emydid species and sub-species suggest that its lack of diagnosability most likely reflects the non-distinctiveness of this proposed taxon.
 
@@
 These considerations have sparked interest in whether geometric morphometric analyses can capture fine-scale variation that can be used for identifying cryptic species.
-This would make the task of identifying and maintaining endangered or conserved groups much easier and could contribute to improved classifications of extinct taxa and populations.
-Most such studies focus on using morphometrics to discover differences between taxa that were identified by other means \cite{Polly2003,Zelditch2004,Gaubert2005b,Gunduz2007,Polly2007a,Demandt2009,Markolf2013,Fruciano2016}.
+Most studies focus on using morphometrics to discover differences between taxa that were identified by other means \cite{Polly2003,Zelditch2004,Gaubert2005b,Gunduz2007,Polly2007a,Demandt2009,Markolf2013,Fruciano2016}.
 Additionally, there has been work on automated taxon identification and classification of taxa into groups \cite{Baylac2003,Dobigny2003,MacLeod2007,VandenBrink2011,Vitek2017}, as well as the development of models that combine genetic, phenotypic, and geographic data to infer evolutionary units of interest \cite{Guillot2012}.
@@
 We use an ensemble learning approach where multiple methods are used in order to look for consensus among their results.
-We test our approach on three datasets: plastron shape of eight species of closely related turtles, plastron shape of two subspecies of a single turtle species, and plastron shape of the \textit{Emys marmorata} species complex.
+We test our approach on three datasets: plastron (ventral shell) shape of seven species of closely related turtles and one outgroup, plastron shape of two subspecies of a single turtle species, and plastron shape of the \textit{Emys marmorata} species complex.
 In particular, we ask whether it is possible to determine which among a set of classification hypotheses best aligns with the observed morphology, and examine the implications of our results for the \textit{E.
@@
 \subsection*{Background and study system}
-Machine learning is an extension of known statistical methodology that emphasizes predictive accuracy and generality often at the expense of the interpretability of individual parameters \cite{Hastie2009}.
+Machine learning is an extension of known statistical methodology that emphasizes predictive accuracy and generality, often at the expense of the interpretability of individual parameters \cite{Hastie2009}.
 Basic statistical approaches are supplemented by randomization, sorting, and partitioning algorithms, along with the maximization or minimization of summary statistics, in order to best estimate a general model for all data, both sampled and unsampled \cite{Hastie2009}.
@@
 
-There are two major classes of machine learning method: unsupervised and supervised learning.
+There are two major classes of machine learning methods: unsupervised and supervised learning.
 Unsupervised learning methods are used with unlabeled data where the underlying structure is estimated; they are analogous to clustering and density estimation methods \cite{Kaufman1990}.
@@
 Our application of the supervised learning approaches used in this study illustrates only a sampling of the various methods available for fitting classification models.
-The specific methods used in this study were chosen because they are suited for cases with more two or more response classes.
+The specific methods we used were chosen because they are suited for cases with more two or more response classes.
 
@@
 Analyses with nuclear DNA \cite{Spinks2010} and single-nucleotide polymorphism (SNP) data suggest a primarily north--south division in \textit{E.
-marmorata}, although these datasets differed from that of mitocondrial-based results of Spinks and Shaffer \cite{Spinks2005} in the location of the break point \cite{Spinks2014}.
+marmorata}, although these datasets differed from the mitocondrial-based results of Spinks and Shaffer \cite{Spinks2005} in the location of the break point \cite{Spinks2014}.
 All three studies discussed the potential taxonomic implications of their results, with Spinks et al.
@@
 We choose to analyze plastron shape for multiple reasons.
-First, it is very easy to collect geometric morphometric data on plastron shape from two-dimensional pictures as the structure is virtually flat.
-This approach allows both museum specimens and individuals in the field to be analyzed together.
+First, it is very easy to collect geometric morphometric data on plastron shape from two-dimensional photographs as the structure is virtually flat.
+This approach allows both museum specimens and individuals photographed in the field to be analyzed together.
 Second, previous work has suggested that there are strong differences in plastron shape among traditionally-recognized emydine species \cite{Angielczyk2007,Angielczyk2011,Angielczyk2013a}.
@@
 
-Three different geometric morphometric datasets describing turtle plastron variation were assembled for this analysis: 1) specimens from eight distinct emydine species; 2) \textit{T.
+Three different geometric morphometric datasets describing turtle plastron variation were assembled for this analysis: 1) specimens from seven distinct emydine species and a deriochelyine outgroup (\textit{C.
+picta}); 2) \textit{T.
 scripta} specimens from the two main subspecies (\textit{T.
@@
 marmorata} specimens from across the species' geographic range.
-The first two datasets are intended to serve as a test of whether machine learning techniques can differentiate species-level groupings of emydine turtles using plastron shape.
+The first two datasets are intended to serve as a test of whether machine learning techniques can differentiate species-level groupings of emydid turtles using plastron shape.
 We expect that the first case represents a low complexity dataset because of the high level of plastron shape disparity that exists among these species \cite{Claude2003a,Claude2006,Angielczyk2011}, whereas the second dataset should be relatively higher in complexity and more analogous to the \textit{E.
@@
 We predict that the \textit{E.
-marmorata} dataset should be of the highest complexity and our greatest challenge given the finding that only very subtle differences existed between geographically-distinct populations \cite{Holland1992}.
+marmorata} dataset should be of the highest complexity and our greatest challenge given the previous finding that only very subtle differences existed between geographically-distinct populations \cite{Holland1992}.
 %We expect the first dataset represents a low-complexity dataset given the high level of plastron shape disparty that exists among these species \cite{Angielczyk2011}, whereas the second dataset should be relatively higher in complexity than the first.
@@
 These specimens represent a subset of those included in previous studies \cite{Angielczyk2007,Angielczyk2011,Angielczyk2013a}.
-Because previous studies did not use vouchered specimens \cite{Spinks2005,Spinks2010,Spinks2014} we were not able to directly sample the individuals in their studies.
+Because previous genetic studies did not use vouchered specimens \cite{Spinks2005,Spinks2010,Spinks2014} we were not able to directly sample the individuals in their studies.
 Instead, our specimen classifications were based solely on the geographic information and not explicit assignment using molecular data.
@@
 
-For our first three schemes \cite{Spinks2010} we used three binning schemes.
+We used three main binning schemes.
 All three schemes include a class for \textit{E.
@@
 In scheme SP10.3 these specimens are assigned to a San Joaquin class reflecting the mitochondrial distinctiveness shown by Spinks and Shaffer \cite{Spinks2005}.
-For our second two schemes \cite{Spinks2014} we used two binning schemes with SP14.1 being based on their phylogenetic network analysis and SP14.2 being based on their Bayesian species delimitation analysis.
-The latter scheme requires the addition of two new classes, ``Baja'' and ``Foothill,'' to accommodate the genetic groupings recovered by the SNP Structure analysis that was used to create the guide tree for the BPP species delimitation analysis \cite{Spinks2014}.
-Finally, we proposed a conservative morphological hypothesis (``Morph'') in order to compare the molecular hypotheses with something approximating the original taxonomic hypothesis for the group; this scheme is made up solely of the marm and pall classes from the SP10.3 scheme.
+We used two versions of the binning scheme based on the results of Spinks et al.
+\cite{Spinks2014}: SP14.1 reflects their phylogenetic network analysis and SP14.2 corresponds to their Bayesian species delimitation analysis.
+SP14.2 requires the addition of two new classes, ``Baja'' and ``Foothill,'' to accommodate the genetic groupings recovered by the SNP Structure analysis that was used to create the guide tree for the BPP species delimitation analysis \cite{Spinks2014}.
+Finally, we proposed a conservative morphological scheme (``Morph'') in order to compare the molecular hypotheses with something approximating the original taxonomic hypothesis for the group.
+This scheme is made up solely of the marm and pall classes from the SP10.3 scheme.
 
@@
 Sex was known only for a subset of the total dataset and was not included as a predictor of classification.
-Instead, we estimated the degree by which specimens cluster morphologically by sex in order to determine how much of a potential biasing factor sexual dimorphism could be for our analysis of the \textit{E.
+Instead, we estimated the degree to which specimens cluster morphologically by sex in order to determine how much of a potential biasing factor sexual dimorphism could be for our analysis of the \textit{E.
 marmorata} species complex (see below).
@@
 \ref{fig:plastra}).
-Seventeen of the landmarks are at the endpoints or intersection of the keratinous plastral scutes that cover the plastron.
+Seventeen of the landmarks are at the endpoints or intersection of the keratinous scutes that cover the plastron.
 Twelve of the landmarks were symmetrical across the axis of symmetry.
 Because damage prevented the digitization of all the symmetric landmarks in some specimens, we reflected landmarks across the axis of symmetry (i.e.
-midline) prior to analysis and used the average position of each symmetrical pair.
+midline) prior to analysis and used the average position of each symmetric pair.
 In cases where damage or incompleteness prevented symmetric landmarks from being determined, we used only the single member of the pair.
@@
 marmorata} specimens were each digitized four times, with the original set of digitized coordinates serving as a fifth replicate.
-These 50 landmark configurations were then Procrustes superimposed.
+These 50 landmark configurations were Procrustes superimposed.
 A range of four Procrustes distances was then calculated as the average of the pairwise distances between each of the replicate configurations of a given specimen.
@@
 The ratio between these two values was used to assess the magnitude of variation caused by digitization.
-The goal of this ratio is to determine if the within group distances are on average smaller than the between individual distances; a value of 0 indicates perfect grouping, a value of 1 indicates no difference between grouping and no grouping, and a value of 1+ indicates that the grouping is counter-intuitive to the data.
-
-\textit{Emys marmorata} is known to display sexual dimorphism in plastral shape, particularly the presence of a plastra concavity in males \cite{Seeliger1945}.
+The goal of this ratio is to determine if the within group distances are smaller on average than the between individual distances; a value of 0 indicates perfect grouping, a value of 1 indicates no difference between grouping and no grouping, and a value of 1+ indicates that the grouping is counter-intuitive to the data.
+
+\textit{Emys marmorata} is known to display sexual dimorphism in plastral shape, particularly the presence of a plastral concavity in males \cite{Seeliger1945}.
 To test for biases resulting from sexual dimorphism in our \textit{E.
@@
 This analysis was also done for each of the classification schemes in order to determine if sex is a biasing factor within these groups.
-As with the total dataset, only a limited number of specimens in each classification scheme have sex information, and the sex ratios within each of the classification schemes are as follows: SP10.1 202 F/199 M, SP10.2 202/199, SP10.3 202/199, SP14.1 177/178, SP14.2 136/125, Morph 112/120.
+As with the total dataset, only a limited number of specimens in each classification scheme have sex information, and the observed sex ratios for each of the classes within the classification schemes are summarized in Table \ref{tab:sex_ratio}.
+Importantly, these within-class sex ratios are based only on the subset of specimens with sex information.
+
+\begin{table}
+  \centering
+  \caption{Table of within-class sex ratios from those speciems with sex information for ecah classification schemes used in this analysis.}
+    \begin{tabular}{l l}
+      \hline
+      Scheme & class ratios \\
+      \hline
+      SP10.1 & CCR: 70 F/60 M; Marm: 112/113; Pall: 20/26 \\
+      SP10.2 &  CCR: 63/50; Marm: 119/123; Pall: 20/26 \\
+      SP10.3 & CCR: 63/50; Marm: 92/95; Pall: 20/26; SJ: 27/28 \\
+      SP14.1 & Marm: 112/117; Pall: 65/61 \\
+      SP14.2 & Baja: 0/5; Foothill: 20/19; Marm: 70/63; Pall: 45/38\\
+      Morph & Marm: 92/94; Pall: 20/26 \\
+      \hline
+    \end{tabular}
+  \label{tab:sex_ratio}
+\end{table}
 
@@
 Additional interaction terms were not considered because of model complexity/sample size concerns.
-Size and the interaction between size and PC 1 were included as predictors to account for known ontogenetic variation in plastron shape \cite{Angielczyk2013a} as well as potential size differences between classes, even if this is unlikely \cite{Seeliger1945,Holland1992}.
+Size and the interaction between size and PCs 1 and 2 were included as predictors to account for known ontogenetic variation in plastron shape \cite{Angielczyk2013a} as well as potential size differences between classes, even if this is unlikely \cite{Seeliger1945,Holland1992}.
 These data constitute a ``maximum set'' because the best or selected models based on five-fold cross-validation need not, and likely will not, include all predictors possible (see below).
 Because our supervised learning models use PCs as predictors, this approach is in many ways analogous to PCA regression.
-PCA regression takes advantage of reduction and orthogonality PCs to improve regression fit \cite{Hastie2009}.
+PCA regression takes advantage of the reduction and orthogonality PCs to improve regression fit \cite{Hastie2009}.
 Because the PCs of shape are by definition orthogonal, they can easily serve as independent predictors or features of class membership without fear of collinearity.
@@
 In classification studies, such as this one, a common metric of performance is the receiver operating characteristic (ROC) which is the relationship between the false and true positive rates \cite{Hastie2009}.
-The area under the ROC curve (AUC) is the derived estimate of the model performance; AUC ranges from 0.5 to 1 which correspond to performance similar to random guesses and perfect classification rates, respectively \cite{Hastie2009}.
+The area under the ROC curve (AUC) is the derived estimate of model performance; AUC ranges from 0.5 to 1 which correspond to performance similar to random guesses and perfect classification rates, respectively \cite{Hastie2009}.
 Both ROC and AUC are preferable to simple classification accuracy when class membership is unbalanced, as it is in these analyses \cite{Hastie2009}.
@@
 The first PC axis of shape variation appears to be primarily structured by differences in individual centroid size (Fig.
-\ref{fig:emys_pca}); this was the motivation for including centroid size and its interaction with PC1 as predictors in all of the supervised learning models.
+\ref{fig:emys_pca}); a linear regression of PC 1 vs log centroid size shows that size explains approximately 81 percent of the variation along this axis.
+This was the motivation for including centroid size and its interaction with PC1 as predictors in all of the supervised learning models.
+Size typically explains less than three percent of the variance of the remaining PC axes (e.g.
+2.5\% on PC 2).
 
@@
 In contrast, the different \textit{E.
-marmorata} classification schemes appear to barely be distinct, with their within:between ratios approximating 1.
+marmorata} classification schemes appear to be barely distinct, with their within:between ratios approximating 1.
 This indicates that the magnitude of the differences between groupings is approximately the same as the difference between any two random individuals (Table \ref{tab:rep_res}).
@@
 scripta elegans} datasets indicate that these taxa are sufficiently morphologically distinct to be differentiated on the basis of plastron shape.
-Both in-sample and out-of-sample classification have AUC values of approximately 1 for all methods, implying near-perfect classification rates (Fig.
+Both in-sample and out-of-sample classifications have AUC values of approximately 1 for all methods, implying near-perfect classification rates (Fig.
 \ref{fig:other_sel}, \ref{fig:other_oos}).
@@
 These results indicate that in cases of clear class separation (Fig.
-\ref{fig:other_pca}) our approach is able to detect this and make good out-of-sample prediction.
+\ref{fig:other_pca}) our approach is able to detect this and make good out-of-sample predictions.
 
@@
 There are also no features of the plastron of \textit{E.
-marmorata} that would make it significantly more difficult to accurately digitize than the plastra of the other speices.
-
-Biological explanations include the possibilty that genetic differentiation is not associated with plastron shape variation and/or that local selective pressures (e.g.
+marmorata} that would make it significantly more difficult to accurately digitize than the plastra of the other species.
+
+Biological explanations include the possibility that genetic differentiation is not associated with plastron shape variation and/or that local selective pressures (e.g.
 from hydrological regime) overwhelm morphological differentiation.
@@
 marmorata}, it is important to note that plastron shape is an extremely effective method for differentiating classes in the additional datasets we investigated.
-The magnitude of shape differences between the species (measured as Procrustes distance between the eight species' mean shapes) is approximately an order of magnitude greater than the differences between the \textit{E.
+The magnitude of shape differences between the eight species (measured as Procrustes distance between the eight species' mean shapes) is approximately an order of magnitude greater than the differences between the \textit{E.
 marmorata} subgroups, and not surprisingly the machine learning methods had no trouble classifying the specimens correctly.
@@
 pallida} does not, on its own, preclude the recognition of this taxon.
-However, this apparent lack of congruence does prompt a reexamination of the methods and concepts that led to that taxonomic revision, especially considering that plastron shape is demonstrably capable of differentiating species and subspecies among other emydids.
+However, the apparent lack of congruence between phenotypic and genetic data does prompt a reexamination of the methods and concepts that led to that taxonomic revision, especially considering that plastron shape is demonstrably capable of differentiating species and subspecies among other emydids.
 In other words, before we can assess the significance of the morphological non-diagnosablity, it is essential to evaluate the methods and concepts that led to the initial taxonomic revision.
@@
 
-In addition to lacking a robust morphological marker, the natural history and geographical distribution of \textit{E.
+The only other range-wide analysis of morphological variation in the E.
+marmorata complex is that of Seeliger \cite{Seeliger1945}, which was cited by Spinks et al.
+\cite{Spinks2014} as being “consistent” with their results.
+A closer examination of Seeliger \cite{Seeliger1945} raises questions about that claim.
+For example, Spinks et al.
+\cite{Spinks2014} claim that Seeliger \cite{Seeliger1945} suggested that “an area of extensive intergradation was restricted to the San Joaquin Valley.” But Seeliger actually hypothesized that her \textit{E.
+m.
+pallida} and \textit{E.
+m.
+marmorata} also intergrade in the vicinity of the San Francisco Bay (Coast Ranges), and lists localities in  Alameda, Contra Costa, and Santa Clara counties where this occurs.
+Spinks et al.
+lack samples from the Bay Area, but the closest specimens to the east of the Bay Area show signs of intergradation.
+Thus, the putative genetic distinctiveness of \textit{E.
+marmorata} and \textit{E.
+pallida} may be partially due to a sampling artifact.
+
+A second issue is that the morphological characters that Seeliger \cite{Seeliger1945} used to distinguish the two putative taxa show considerable variation.
+The condition of the inguinal scales supposedly can be used to diagnose \textit{E.
+marmorata} (present and larger) and \textit{E.
+pallida} (smaller or absent).
+However, she noted that 11\% of of the \textit{E.
+marmorata marmorata} specimens in her dataset lack scales, 40\% of her \textit{E.
+marmorata pallida} specimens possess them, and specimens from inland central California have “all types of inguinals” (p.
+156).
+Although she qualitatively described differences in the size of the inguinals in the subspecies, she did not quantify these, and modern methods for quantifying shape differences in the scales she observed lay decades in the future.
+Seeliger \cite{Seeliger1945} also mentions distinct tendencies in coloration of the neck between the subspecies, but this character shows a similar pattern of variation to the inguinal scales.
+Nineteen percent and 33\% of \textit{E.
+marmorata marmorata} and \textit{E.
+marmorata pallida} specimens, respectively, possess the coloration pattern expected of the other subspecies, and specimens from the areas of intergradation “have both types of coloration and neither is predominant” (p.157).
+Further studies of variation in these characters is clearly warranted, especially in the broad intergrade zones.
+
+The natural history and geographical distribution of \textit{E.
 marmorata} and \textit{E.
@@
 The mitochondrial data from Spinks et al.
-\cite{Spinks2014} show extensive introgression and admixture in Central California, which is expected because there are no significant barriers to gene flow in this region.
-They also lack sampling from the populations between the two putative species in the San Francisco Bay Area, which we predict would likely show even more genetic mixing.
+\cite{Spinks2014} show extensive introgression and admixture in Central California, which is likely not just a mitochondrial sweep because there are no significant barriers to nuclear gene flow in this region.
+Spinks et al.
+also lack sampling from the populations between the two putative species in the San Francisco Bay Area, which we predict would likely show even more genetic mixing.
 Combined with the well-demonstrated ability for testudinoid turtles, including emydids and even \textit{Emys}, to hybridize (e.g.
@@
 pallida} as a distinct species needs to address these points.
+Other emydines show fairly high levels of genetic diversity that is often geographically structured (e.g.
+\textit{G.
+insculpta} \cite{Amato2008}, \textit{C.
+guttata} \cite{Davy2014}, and \textit{E.
+blandingii} \cite{Sethuraman2014}) but are still regarded as single species.
 Because the geography, natural history, limited sampling from key areas, demonstrated genetic admixture of \textit{E.
@@
 marmorata}.
-This is because we could not carry out our morphometric analyses on the specimens from which the genetic data were obtained.
-The comparisons with the other emydid taxa suggest that our negative result is is because \textit{E.
-marmorata} is a single species.
-However, tests of both our preferred conclusion (\textit{E.
+Further tests of both our preferred conclusion (\textit{E.
 marmorata} as a single species) and that of Spinks et al.
-\cite{Spinks2014} should include morphological and molecular analyses of the same set of voucher specimens, as well as additional tests of species delimitation using alternative methods and corroborating evidence as suggested by Carstens et al.
+\cite{Spinks2014} should include morphological and molecular analyses of specimens from within the Bay Area intergrade zone, ideally the same set of voucher specimens.
+We also recommend additional tests of species delimitation using alternative methods and corroborating evidence as suggested by Carstens et al.
 \cite{Carstens2013}.
@@
 % 
-%\begin{thebibliography}{10}
-%
-%\end{thebibliography}
-\bibliography{new_bib}
+%\bibliography{new_bib}
+\begin{thebibliography}{10}
+
+  \bibitem{Stuart2006}
+    Stuart BL, Inger RF, Voris HK.
+    \newblock {High level of cryptic species diversity revealed by sympatric
+    lineages of Southeast Asian forest frogs.}
+    \newblock Biology letters. 2006;2(3):470--4.
+    \newblock doi:{10.1098/rsbl.2006.0505}.
+
+  \bibitem{Bickford2007}
+    Bickford D, Lohman DJ, Sodhi NS, Ng PKL, Meier R, Winker K, et~al.
+    \newblock {Cryptic species as a window on diversity and conservation.}
+    \newblock Trends in ecology {\&} evolution. 2007;22(3):148--55.
+    \newblock doi:{10.1016/j.tree.2006.11.004}.
+
+  \bibitem{SchlickSteiner2007}
+    Schilck-Steiner BC, Seifert B, Stauffer C, Christian E, Crozier RH, Steiner FM.
+    \newblock {Without morphology, cryptic species stay in taxonomic crypsis
+    following discovery}.
+    \newblock Trends in ecology {\&} evolution. 2007;22(8):391--392.
+    \newblock doi:{10.1016/j.tree.2007.05.003}.
+
+  \bibitem{Pfenninger2007}
+    Pfenninger M, Schwenk K.
+    \newblock {Cryptic animal species are homogeneously distributed among taxa and
+    biogeographical regions.}
+    \newblock BMC evolutionary biology. 2007;7:121.
+    \newblock doi:{10.1186/1471-2148-7-121}.
+
+  \bibitem{Clare2011}
+    Clare EL.
+    \newblock {Cryptic species? Patterns of maternal and paternal gene flow in
+    eight neotropical bats.}
+    \newblock PloS one. 2011;6(7):e21460.
+    \newblock doi:{10.1371/journal.pone.0021460}.
+
+  \bibitem{Funk2012}
+    Funk WC, Caminer M, Ron SR.
+    \newblock {High levels of cryptic species diversity uncovered in Amazonian
+    frogs.}
+    \newblock Proceedings of the Royal Society B: Biological Sciences.
+    2012;279(1734):1806--14.
+    \newblock doi:{10.1098/rspb.2011.1653}.
+
+  \bibitem{Pons2006}
+    Pons J, Barraclough T, Gomez-Zurita J, Cardoso A, Duran D, Hazell S, et~al.
+    \newblock {Sequence-Based Species Delimitation for the DNA Taxonomy of
+    Undescribed Insects}.
+    \newblock Systematic Biology. 2006;55(4):595--609.
+    \newblock doi:{10.1080/10635150600852011}.
+
+  \bibitem{Carstens2010}
+    Carstens BC, Dewey TA.
+    \newblock {Species Delimitation Using a Combined Coalescent and
+    Information-Theoretic Approach: An Example from North American Myotis Bats}.
+    \newblock Systematic Biology. 2010;59(4):400--414.
+
+  \bibitem{Hausdorf2010}
+    Hausdorf B, Hennig C.
+    \newblock {Species delimitation using dominant and codominant multilocus
+    markers.}
+\newblock Systematic biology. 2010;59(5):491--503.
+\newblock doi:{10.1093/sysbio/syq039}.
+
+\bibitem{O'Meara2010}
+  O'Meara BC.
+\newblock {New heuristic methods for joint species delimitation and species
+  tree inference.}
+\newblock Systematic biology. 2010;59(1):59--73.
+\newblock doi:{10.1093/sysbio/syp077}.
+
+\bibitem{Yang2010b}
+  Yang Z, Rannala B.
+\newblock {Bayesian species delimitation using multilocus sequence data.}
+\newblock Proceedings of the National Academy of Sciences.
+  2010;107(20):9264--9.
+\newblock doi:{10.1073/pnas.0913022107}.
+
+\bibitem{Huelsenbeck2011b}
+  Huelsenbeck JP, Andolfatto P, Huelsenbeck ET.
+\newblock {Structurama: bayesian inference of population structure.}
+\newblock Evolutionary bioinformatics online. 2011;7:55--9.
+\newblock doi:{10.4137/EBO.S6761}.
+
+\bibitem{Bauer2000}
+  Bauer AM, Parham JF, Brown RM, Stuart BL, Grismer L, Papenfuss TJ, et~al.
+\newblock {Availability of new Baysian-delimited gecko names and the importance
+  of character-based species descriptions}.
+\newblock Proceedings of the Royal Society B: Biological Sciences.
+  2000;278:490--492.
+\newblock doi:{10.2307/1467045}.
+
+\bibitem{Carstens2013}
+  Carstens BC, Pelletier Ta, Reid NM, Satler JD.
+\newblock {How to fail at species delimitation.}
+\newblock Molecular ecology. 2013;22(17):4369--83.
+\newblock doi:{10.1111/mec.12413}.
+
+\bibitem{Leache2010}
+  Leach{\'{e}} AD, Fujita MK.
+\newblock {Bayesian species delimitation in West African forest geckos
+  (Hemidactylus fasciatus).}
+\newblock Proceedings Biological sciences / The Royal Society.
+  2010;277(1697):3071--7.
+\newblock doi:{10.1098/rspb.2010.0662}.
+
+\bibitem{Spinks2014}
+  Spinks PQ, Thomson RC, {Bradley Shaffer} H.
+\newblock {The advantages of going large: genome wide SNPs clarify the complex
+  population history and systematics of the threatened western pond turtle}.
+\newblock Molecular Ecology. 2014; p. n/a--n/a.
+\newblock doi:{10.1111/mec.12736}.
+
+\bibitem{Eldredge1972}
+  Eldredge N, Gould SJ.
+\newblock {Punctuated equilibria: an alternative to phyletic gradualism}.
+\newblock In: Schopf TJM, editor. Models in Paleobiology. San Francisco:
+  Freeman Cooper; 1972. p. 82--115.
+
+\bibitem{Gould1977a}
+  Gould SJ, Eldredge N.
+\newblock {Punctuated equilibria: the tempo and mode of evolution
+  reconsidered}.
+\newblock Paleobiology. 1977;3(2):115--151.
+
+\bibitem{VanBocxlaer2013}
+  {Van Bocxlaer} B, Hunt G.
+\newblock {Morphological stasis in an ongoing gastropod radiation from Lake
+  Malawi}.
+\newblock Proceedings of the National Academy of Sciences.
+  2013;doi:{10.1073/pnas.1308588110}.
+
+\bibitem{Polly2003}
+  Polly PD.
+\newblock {Paleophylogeography of Sorex araneus: molar shape as a morphological
+  marker for fossil shrews}.
+\newblock Mammalia. 2003;68(2):233--243.
+
+\bibitem{Zelditch2004}
+  Zelditch ML, Swiderski DL, Sheets HD.
+\newblock {Geometric morphometrics for biologists: a primer}.
+\newblock Amsterdam: Elsevier Academic Press; 2004.
+
+\bibitem{Gaubert2005b}
+  Gaubert P, Taylor PJ, Fernandes Ca, Bruford MW, Veron G.
+\newblock {Patterns of cryptic hybridization revealed using an integrative
+  approach: a case study on genets (Carnivora, Viverridae, Genetta spp.) from
+  the southern African subregion}.
+\newblock Biological Journal of the Linnean Society. 2005;86(1):11--33.
+\newblock doi:{10.1111/j.1095-8312.2005.00518.x}.
+
+\bibitem{Gunduz2007}
+  G{\"{u}}nd{\"{u}}z I, Jaarola M, Tez C, Yeniyurt C, Polly PD, Searle JB.
+\newblock {Multigenic and morphometric differentiation of ground squirrels
+  (Spermophilus, Scuiridae, Rodentia) in Turkey, with a description of a new
+  species.}
+\newblock Molecular phylogenetics and evolution. 2007;43(3):916--35.
+\newblock doi:{10.1016/j.ympev.2007.02.021}.
+
+\bibitem{Polly2007a}
+  Polly PD.
+\newblock {Phylogeographic differentiation in Sorex araneus: morphology in
+  relation to geography and karyotype}.
+\newblock Russian Journal of Theriology. 2007;6(1):73--84.
+
+\bibitem{Demandt2009}
+  Demandt MH, Bergek S.
+\newblock {Identification of cyprinid hybrids by using geometric morphometrics
+  and microsatellites}.
+\newblock Journal of Applied Ichthyology. 2009;25(6):695--701.
+\newblock doi:{10.1111/j.1439-0426.2009.01329.x}.
+
+\bibitem{Markolf2013}
+  Markolf M, Rakotonirina H, Fichtel C, von Grumbkow P, Brameier M, Kappeler PM.
+\newblock {True lemurs{\ldots}true species - species delimitation using
+  multiple data sources in the brown lemur complex}.
+\newblock BMC Evolutionary Biology. 2013;13(1):233.
+\newblock doi:{10.1186/1471-2148-13-233}.
+
+\bibitem{Fruciano2016}
+  Fruciano C, Franchini P, Raffini F, Fan S, Meyer A.
+\newblock {Are sympatrically speciating Midas cichlid fish special? Patterns of
+  morphological and genetic variation in the closely related species
+  Archocentrus centrarchus}.
+\newblock Ecology and Evolution. 2016;6(12):4102--4114.
+\newblock doi:{10.1002/ece3.2184}.
+
+\bibitem{Baylac2003}
+  Baylac M, Villemant C, Simbolotti G.
+\newblock {Combining geometric morphometrics with pattern recognition for the
+  investigation of species complexes}.
+\newblock Biological Journal of the Linnean Society. 2003;80:89--98.
+
+\bibitem{Dobigny2003}
+  Dobigny G, Granjon L, Aniskin V, Ba K, Voloboulev V.
+\newblock {A new sigling species of Taterillus (Muridae, Gerbillinae) from West
+  Agrica}.
+\newblock Mammalian Biology. 2003;68:299--316.
+
+\bibitem{MacLeod2007}
+  MacLeod N.
+\newblock {Automated taxon identification in systematics: theory, approaches
+  and applications}.
+\newblock Boca Raton: CRC Press; 2007.
+
+\bibitem{VandenBrink2011}
+  van~den Brink V, Bokma F.
+\newblock {Morphometric shape analysis using learning vector quantization
+  neural networks — an example distinguishing two microtine vole species}.
+\newblock Annales Zoologici Fennici. 2011;48:359--364.
+
+\bibitem{Vitek2017}
+  Vitek NS, Manz CL, Gao T, Bloch JI, Strait SG, Boyer DM.
+\newblock {Semi-supervised determination of pseudocryptic morphotypes using
+  observer-free characterizations of anatomical alignment and shape}.
+\newblock Ecology and Evolution. 2017;7(14):5041--5055.
+\newblock doi:{10.1002/ece3.3058}.
+
+\bibitem{Guillot2012}
+  Guillot G, Renaud S, Ledevin R, Michaux J, Claude J.
+\newblock {A unifying model for the analysis of phenotypic, genetic, and
+  geographic data}.
+\newblock Systematic Biology. 2012;61(6):897--911.
+\newblock doi:{10.1093/sysbio/sys038}.
+
+\bibitem{Hastie2009}
+  Hastie T, Tibshirani R, Friedman J.
+\newblock {The elements of statistical learning: data mining, inference, and
+  prediction}.
+\newblock 2nd ed. New York: Springer; 2009.
+
+\bibitem{Kaufman1990}
+  Kaufman L, Rousseeuw PJ.
+\newblock {Finding groups in data : an introduction to cluster analysis}.
+\newblock New York: Wiley; 1990.
+
+\bibitem{Breiman1984}
+  Breiman L, Friedman J, Stone CJ, Olshen RA.
+\newblock {Classification and regression trees}.
+\newblock Belmont: Wadsworth International Group; 1984.
+
+\bibitem{Francoy2009}
+  Francoy TM, Silva RAO, Nunes-Silva P, Menezes C, Imperatriz-Fonseca VL.
+\newblock {Gender identification of five genera of stingless bees (Apidae,
+  Meliponini) based on wing morphology}.
+\newblock Genetics and molecular research. 2009;8(1):207--214.
+
+\bibitem{Sztencel-Jabonka2009}
+  Sztencel-Jab{\l}onka A, Jones G, BogdanowicZ W.
+\newblock {Skull Morphology of Two Cryptic Bat Species: Pipistrellus
+  pipistrellus and P. pygmaeus — A 3D Geometric Morphometrics Approach with
+  Landmark Reconstruction}.
+\newblock Acta Chiropterologica. 2009;11(1):113--126.
+\newblock doi:{10.3161/150811009X465730}.
+
+\bibitem{Edwards2011}
+  Edwards S, Claude J, {Van Vuuren} BJ, Matthee CA.
+\newblock {Evolutionary history of the Karoo bush rat, Myotomys unisulcatus
+  (Rodentia: Muridae): Disconcordance between morphology and genetics}.
+\newblock Biological Journal of the Linnean Society. 2011;102(3):510--526.
+\newblock doi:{10.1111/j.1095-8312.2010.01583.x}.
+
+\bibitem{MitrovskiBogdanovic2013}
+  Mitrovski-Bogdanovic A, Petrovic A, Mitrovic M, Ivanovic A, {\v{Z}}ikic V,
+  Star{\'{y}} P, et~al.
+\newblock {Identification of two cryptic species within the Praon abjectum
+  group (Hymenoptera: Braconidae: Aphidiinae) using molecular markers and
+  geometric morphometrics}.
+\newblock Annals of the entomological society of America. 2013;106(2):170--180.
+
+\bibitem{Dillard2017}
+  Dillard KC.
+\newblock {A comparative analysis of geometric morphometrics across two
+  Pseudemys turtle species in east central Virginia} [Masters].
+\newblock Virginia Commonwealth University; 2017.
+
+\bibitem{Caumul2005a}
+  Caumul R, Polly PD.
+\newblock {Phylogenetic and environmental components of morphological
+  variation: skull, mandible, and molar shape in marmots (Marmota, Rodentia).}
+\newblock Evolution; international journal of organic evolution.
+  2005;59(11):2460--72.
+
+\bibitem{Cardini2009a}
+  Cardini A, Nagorsen D, O'Higgins P, Polly PD, {Thorington Jr} RW, Tongiorgi P.
+\newblock {Detecting biological distinctiveness using geometric morphometrics:
+  an example case from the Vancouver Island marmot}.
+\newblock Ethology Ecology {\&} Evolution. 2009;21:209--223.
+
+\bibitem{VanBocxlaer2010}
+  {Van Bocxlaer} B, Schulthei{\ss} R.
+\newblock {Comparison of morphometric techniques for shapes with few homologous
+  landmarks based on machine-learning approaches to biological discrimination}.
+\newblock Paleobiology. 2010;36(3):497--515.
+
+\bibitem{Navega2015}
+  Navega D, Vicente R, Vieira DN, Ross AH, Cunha E.
+\newblock {Sex estimation from the tarsal bones in a Portuguese sample: a
+  machine learning approach}.
+\newblock International Journal of Legal Medicine. 2015;129(3):651--659.
+\newblock doi:{10.1007/s00414-014-1070-5}.
+
+\bibitem{Baird1852}
+  Baird SF, Girard C.
+\newblock {Descriptions of new species of reptiles collected by the U.S.
+  Exploring Expedition under the command of Capt. Charles Wilkes}.
+\newblock Proceedings of the National Academy of Sciences Philadelphia.
+  1852;6:174--177.
+
+\bibitem{Feldman2002}
+  Feldman CR, Parham JF.
+\newblock {Molecular phylogenetics of emydine turtles: taxonomic revision and
+  the evolution of shell kinesis.}
+\newblock Molecular Phylogenetics and Evolution. 2002;22(3):388--98.
+\newblock doi:{10.1006/mpev.2001.1070}.
+
+\bibitem{Bury2017}
+  Bury RB.
+\newblock {Biogeography of Western Pond Turtles in the western Great Basin:
+  Dispersal Across a Northwest Passage ?}
+\newblock Western Wildlife2. 2017;4:72--80.
+
+\bibitem{Seeliger1945}
+  Seeliger LM.
+\newblock {Variation in the Pacific Mud Turtle}.
+\newblock Copeia. 1945;1945(3):150--159.
+
+\bibitem{Lubcke2007}
+  Lubcke GM, Wilson DS.
+\newblock {Variation in shell morphology of the Western Pond Turtle (Actinemys
+  marmorata Baird and Giarard) from three aquativ habitats in Northern
+  California}.
+\newblock Journal of Herpetology. 2007;41(1):107--114.
+
+\bibitem{Germano2008}
+  Germano DJ, Rathbun GB.
+\newblock {Growth, population structure, and reproduction of western pond
+  turtles (Actinemys marmorata) on the Central Coast of California}.
+\newblock Chelonian Conservation and Biology. 2008;7(2):188--194.
+
+\bibitem{Germano2009}
+  Germano DJ, Bury RB.
+\newblock {Variation in body size, growth, and population structure of
+  Actinemys marmorata from lentic and lotic habitats in Southern Oregon}.
+\newblock Journal of Herpetology. 2009;43(3):510--520.
+
+\bibitem{Bury2010}
+  Bury RB, Germano DJ, Bury GW.
+\newblock {Population Structure and Growth of the Turtle Actinemys marmorata
+  from the Klamath–Siskiyou Ecoregion: Age, Not Size, Matters}.
+\newblock Copeia. 2010;2010(3):443--451.
+\newblock doi:{10.1643/CH-08-096}.
+
+\bibitem{Holland1992}
+  Holland DC.
+\newblock {Level and pattern in morphological variation: a phylogeographic
+  study of the western pond turtle (Clemmys marmorata)}.
+\newblock University of Southwestern Louisiana; 1992.
+
+\bibitem{Kendall1977a}
+  Kendall DG.
+\newblock {The diffusion of shape}.
+\newblock Advances in Applied Probability. 1977;9(3):428--430.
+
+\bibitem{Spinks2005}
+  Spinks PQ, Shaffer HB.
+\newblock {Range-wide molecular analysis of the western pond turtle (Emys
+  marmorata): cryptic variation, isolation by distance, and their conservation
+  implications.}
+\newblock Molecular ecology. 2005;14(7):2047--64.
+\newblock doi:{10.1111/j.1365-294X.2005.02564.x}.
+
+\bibitem{Spinks2010}
+  Spinks PQ, Thomson RC, Shaffer HB.
+\newblock {Nuclear gene phylogeography reveals the historical legacy of an
+  ancient inland sea on lineages of the western pond turtle, Emys marmorata in
+  California.}
+\newblock Molecular ecology. 2010;19(3):542--56.
+\newblock doi:{10.1111/j.1365-294X.2009.04451.x}.
+
+\bibitem{Angielczyk2007}
+  Angielczyk KD, Sheets HD.
+\newblock {Investigation of simulated tectonic deformation in fossils using
+  geometric morphometrics}.
+\newblock Paleobiology. 2007;33(1):125--148.
+
+\bibitem{Angielczyk2011}
+  Angielczyk KD, Feldman CR, Miller GR.
+\newblock {Adaptive evolution of plastron shape in emydine turtles.}
+\newblock Evolution. 2011;65(2):377--394.
+\newblock doi:{10.1111/j.1558-5646.2010.01118.x}.
+
+\bibitem{Angielczyk2013a}
+  Angielczyk KD, Feldman CR.
+\newblock {Are diminutive turtles miniaturized? The ontogeny of plastron shape
+  in emydine turtles}.
+\newblock Biological Journal of the Linnean Society. 2013;108(4):727--755.
+\newblock doi:{10.1111/bij.12010}.
+
+\bibitem{Claude2003a}
+  Claude J, Paradis E, Tong H, Auffray JC.
+\newblock {A geometric morphometric assessment of the effects
+  of$\backslash$nenvironment and cladogenesis on the evolution of
+  the$\backslash$nturtle shell}.
+\newblock Biological Journal of the Linnean Society.
+  2003;79(December):485--501.
+\newblock doi:{10.1046/j.1095-8312.2003.00198.x}.
+
+\bibitem{Claude2006}
+  Claude J.
+\newblock {Convergence induced by plastral kinesis and geometric morphometric
+  assessment: a geometric morphometric assessment}.
+\newblock Fossil Turtle Research. 2006;1:34--45.
+
+\bibitem{Rohlf2005}
+  Rohlf FJ. {TpsDig 2.04}; 2005.
+
+\bibitem{Dryden1998a}
+  Dryden IL, Mardia KY.
+\newblock {Statistical shape analysis}.
+\newblock New York: Wiley; 1998.
+
+\bibitem{R2016}
+  {R Core Team}. R: A Language and Environment for Statistical Computing; 2016.
+\newblock Available from: \url{http://www.R-project.org/}.
+
+\bibitem{Dryden2013}
+  Dryden IL. shapes: Statistical shape analysis; 2013.
+\newblock Available from: \url{http://CRAN.R-project.org/package=shapes}.
+
+\bibitem{Arnqvist1998}
+  Arnqvist G, M{\aa}rtensson T. {Measurement error in geometric morphometrics:
+  Empirical strategies to assess and reduce its impact on measures of shape};
+  1998.
+
+\bibitem{Cramon2007}
+  von Cramon-Taubadel N, Frazier BC, Lahr MM.
+\newblock {The problem of assessing landmark error in geometric morphometrics:
+  theory, methods, and modifications}.
+\newblock American journal of physical anthropology. 2007;132(4):535--544.
+\newblock doi:{10.1002/ajpa}.
+
+\bibitem{Munoz-MunozF.2010}
+  {Munoz-Munoz F }, {Perpinan D }.
+\newblock {Measurement error in morphometric studies: comparison between manual
+  and computerized methods}.
+\newblock Ann Zool. 2010;47(1):46--56.
+
+\bibitem{Venables2002}
+  Venables WN, Ripley BD.
+\newblock Modern Applied Statistics with S.
+\newblock 4th ed. New York: Springer; 2002.
+\newblock Available from: \url{http://www.stats.ox.ac.uk/pub/MASS4}.
+
+\bibitem{mdapack}
+  Hastie T, Tibshirani R, Leisch F, Hornik K, Ripley BD. mda: Mixture and
+  Flexible Discriminant Analysis; 2015.
+\newblock Available from: \url{https://CRAN.R-project.org/package=mda}.
+
+\bibitem{Liaw2002}
+  Liaw A, Wiener M.
+\newblock Classification and Regression by randomForest.
+\newblock R News. 2002;2(3):18--22.
+
+\bibitem{Kuhn2013}
+  Kuhn M, Johnson K.
+\newblock {Applied predictive modeling}.
+\newblock New York, NY: Springer; 2013.
+
+\bibitem{Hand2001}
+  Hand DJ, Till RJ.
+\newblock {A Simple Generalisation of the Area Under the ROC Curve for Multiple
+  Class Classification Problems}.
+\newblock Machine Learning. 2001;45:171--186.
+
+\bibitem{KuhnMAN2013}
+  Kuhn M. caret: Classification and Regression Training; 2013.
+\newblock Available from: \url{http://CRAN.R-project.org/package=caret}.
+
+\bibitem{Rivera2008}
+  Rivera G.
+\newblock {Ecomorphological variation in shell shape of the freshwater turtle
+  Pseudemys concinna inhabiting different aquatic flow regimes.}
+\newblock Integrative and comparative biology. 2008;48(6):769--87.
+\newblock doi:{10.1093/icb/icn088}.
+
+\bibitem{Rivera2011}
+  Rivera G, Stayton CT.
+\newblock {Finite element modeling of shell shape in the freshwater turtle
+  Pseudemys concinna reveals a trade-off between mechanical strength and
+  hydrodynamic efficiency.}
+\newblock Journal of morphology. 2011;272(10):1192--203.
+\newblock doi:{10.1002/jmor.10974}.
+
+\bibitem{Stayton2011}
+  Stayton CT.
+\newblock {Biomechanics on the half shell: functional performance influences
+  patterns of morphological variation in the emydid turtle carapace.}
+\newblock Zoology (Jena, Germany). 2011;114(4):213--23.
+\newblock doi:{10.1016/j.zool.2011.03.002}.
+
+\bibitem{Rivera2014}
+  Rivera G, Davis JN, Godwin JC, Adams DC.
+\newblock {Repeatability of Habitat-Associated Divergence in Shell Shape of
+  Turtles}.
+\newblock Evolutionary Biology. 2014; p. 29--37.
+\newblock doi:{10.1007/s11692-013-9243-6}.
+
+\bibitem{Polly2016}
+  Polly PD, Stayton CT, Dumont ER, Pierce SE, Rayfield EJ, Angielczyk KD.
+\newblock {Combining geometric morphometrics and finite element analysis with
+  evolutionary modeling: towards a synthesis}.
+\newblock Journal of Vertebrate Paleontology. 2016;4634(March).
+\newblock doi:{10.1080/02724634.2016.1111225}.
+
+\bibitem{Dayrat2005a}
+  Dayrat B.
+\newblock {Towards integrative taxonomy}.
+\newblock Biological Journal of the Linnean Society. 2005;85:407--415.
+
+\bibitem{DeQueiroz2007b}
+  {De Queiroz} K.
+\newblock {Species concepts and species delimitation.}
+\newblock Systematic Biology. 2007;56(6):879--86.
+\newblock doi:{10.1080/10635150701701083}.
+
+\bibitem{Padial2010}
+  Padial JM, Miralles A, {De la Riva} I, Vences M.
+\newblock {The integrative future of taxonomy}.
+\newblock Frontiers in Zoology. 2010;7(16):1--14.
+
+\bibitem{Buskirk2005}
+  Buskirk SW, Parham JF, Feldman CR.
+\newblock {On the hybridisation between two distantly related Asian turtles
+  (Testudines: Scalia x Mauremys)}.
+\newblock Salamandra. 2005;41:21--26.
+
+\bibitem{Spinks2009}
+  Spinks PQ, Shaffer HB.
+\newblock {Conflicting mitochondrial and nuclear phylogenies for the widely
+  disjunct Emys (Testudines: Emydidae) species complex, and what they tell us
+  about biogeography and hybridization.}
+\newblock Systematic biology. 2009;58(1):1--20.
+\newblock doi:{10.1093/sysbio/syp005}.
+
+\bibitem{Parham2013}
+  Parham JF, Papenfuss TJ, Dijk PPV, Wilson BS, Marte C, Schettino LR, et~al.
+\newblock {Genetic introgression and hybridization in Antillean freshwater
+  turtles (Trachemys) revealed by coalescent analyses of mitochondrial and
+  cloned nuclear markers.}
+\newblock Molecular phylogenetics and evolution. 2013;67(1):176--87.
+\newblock doi:{10.1016/j.ympev.2013.01.004}.
+
+\bibitem{Amato2008}
+  Amato ML, Brooks RJ, Fu J.
+\newblock {A phylogeographic analysis of populations of the wood turtle
+  (Glyptemys insculpta) throughout its range}.
+\newblock Molecular Ecology. 2008;17(2):570--581.
+\newblock doi:{10.1111/j.1365-294X.2007.03580.x}.
+
+\bibitem{Davy2014}
+  Davy CM, Murphy RW.
+\newblock {Conservation genetics of the endangered Spotted Turtle (
+  {\textless}i{\textgreater}Clemmys guttata{\textless}/i{\textgreater} )
+  illustrate the risks of “bottleneck tests”}.
+\newblock Canadian Journal of Zoology. 2014;92(2):149--162.
+\newblock doi:{10.1139/cjz-2013-0188}.
+
+\bibitem{Sethuraman2014}
+  Sethuraman A, McGaugh SE, Becker ML, Chandler CH, Christiansen JL, Hayden S,
+  et~al.
+\newblock {Population genetics of Blanding's turtle (Emys blandingii) in the
+  midwestern United States}.
+\newblock Conservation Genetics. 2014;15(1):61--73.
+\newblock doi:{10.1007/s10592-013-0521-8}.
+
+\end{thebibliography}
 

### 2018-03-09 doc/plos_latex_template.tex
combined cc7g and tracg so it behaves

--- 2018-02-28 doc/turtles_plosone.tex
@@
 % DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
-% - Figures should be uploaded separately from your manuscript file. 
+% - Figures should be uprloaded separately from your manuscript file. 
 % - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
@@
 
-\usepackage{afterpage}
-\usepackage{caption, subcaption, multirow, morefloats}
-
 % create "+" rule type for thick vertical lines
@@
 {\Large
-\textbf\newline{Ensemble approaches for estimating congruence between species delimitation and morphological variation: comparing taxonomic hypotheses for the Pacific Pond Turtle (\textit{Emys marmorata})} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
+\textbf\newline{Title of submission to PLOS journals} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
 }
@@
 \\
-Peter D Smits\textsuperscript{1*},
-Kenneth D Angielczyk\textsuperscript{2},
-Bryan L Stuart\textsuperscript{3},
-James F Parham\textsuperscript{4},
+Name1 Surname\textsuperscript{1,2\Yinyang},
+Name2 Surname\textsuperscript{2\Yinyang},
+Name3 Surname\textsuperscript{2,3\textcurrency},
+Name4 Surname\textsuperscript{2},
+Name5 Surname\textsuperscript{2\ddag},
+Name6 Surname\textsuperscript{2\ddag},
+Name7 Surname\textsuperscript{1,2,3*},
+with the Lorem Ipsum Consortium\textsuperscript{\textpilcrow}
 \\
 \bigskip
-\textbf{1} Department of Integrative Biology, University of California -- Berkeley, Berkeley, California, USA.
+\textbf{1} Affiliation Dept/Program/Center, Institution Name, City, State, Country
 \\
-\textbf{2} Integrative Research Center, Field Museum of Natural History, Chicago, Illinois, USA.
+\textbf{2} Affiliation Dept/Program/Center, Institution Name, City, State, Country
 \\
-\textbf{3} Section of Research and Collections, North Carolina Museum of Natural Sciences, Raleigh, North Carolina, USA.
-\\
-\textbf{4} John D. Cooper Archaeological and Paleontological Center, Department of Geological Sciences, California State University -- Fullerton, Fullerton, California, USA.
+\textbf{3} Affiliation Dept/Program/Center, Institution Name, City, State, Country
 \\
@@
 % Primary Equal Contribution Note
-%\Yinyang These authors contributed equally to this work.
+\Yinyang These authors contributed equally to this work.
 
@@
 % Also use this double-dagger symbol for special authorship notes, such as senior authorship.
-%\ddag These authors also contributed equally to this work.
+\ddag These authors also contributed equally to this work.
 
 % Current address notes
-%\textcurrency Current Address: Dept/Program/Center, Institution Name, City, State, Country % change symbol to "\textcurrency a" if more than one current address note
+\textcurrency Current Address: Dept/Program/Center, Institution Name, City, State, Country % change symbol to "\textcurrency a" if more than one current address note
 % \textcurrency b Insert second current address 
@@
 % Deceased author note
-%\dag Deceased
+\dag Deceased
 
 % Group/Consortium Author Note
-%\textpilcrow Membership list can be found in the Acknowledgments section.
+\textpilcrow Membership list can be found in the Acknowledgments section.
 
 % Use the asterisk to denote corresponding authorship and provide email address in note below.
-* psmits@berkeley.edu
+* correspondingauthor@institute.edu
 
@@
 \section*{Abstract}
-  We investigated the morphometric identification of cryptic species using machine learning approaches by examining their implications for a recently proposed cryptic turtle species (\textit{Emys pallida}).
-We collected landmark-based morphometric data from 532 adult \textit{E.
-marmorata/``E.
-pallida''} museum specimens.
-We assigned a classification to each specimen for six different binning schemes based on geographic occurrence data recorded in museum collection archives.
-We used an ensemble of supervised machine learning approaches to determine which classification hypothesis was best supported by the data.
-In addition, we applied the same approach to two clear-cut examples, one consisting of eight unambiguously distinct species closely related to \textit{E.
-marmorata}, and the other consisting of two subspecies of \textit{Trachemys scripta}.
-Our results indicate that there is no clear ``best'' grouping of \textit{E.
-marmorata/``E.
-pallida''} based on plastron shape.
-In contrast, the analyses of the clear-cut examples produced near perfect classifications, demonstrating that the methods can recover correct results when an appropriate signal exists.
-Explanations for the lack of grouping in \textit{E.
-marmorata} include the possibility that genetic differentiation is not associated with plastron shape variation below the species level and/or that local selective pressures (e.g., from hydrological regime) overwhelm morphological differentiation.
-A reconsideration of the methods used to delimit \textit{``E.
-pallida,''} the lack of barriers to gene flow, the strong evidence for widespread admixture between lineages, and the fact that plastron shape can be used to delineate other emydine species and sub-species suggest that its lack of diagnosability most likely reflects the non-distinctiveness of this proposed taxon.
+Lorem ipsum dolor sit amet, consectetur adipiscing elit.
+Curabitur eget porta erat.
+Morbi consectetur est vel gravida pretium.
+Suspendisse ut dui eu ante cursus gravida non sed sem.
+Nullam sapien tellus, commodo id velit id, eleifend volutpat quam.
+Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus.
+Nunc pellentesque pretium diam, quis maximus dolor faucibus id.
+Nunc convallis sodales ante, ut ullamcorper est egestas vitae.
+Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.
 
@@
 % Author Summary not valid for PLOS ONE submissions.   
-%\section*{Author summary}
+\section*{Author summary}
+Lorem ipsum dolor sit amet, consectetur adipiscing elit.
+Curabitur eget porta erat.
+Morbi consectetur est vel gravida pretium.
+Suspendisse ut dui eu ante cursus gravida non sed sem.
+Nullam sapien tellus, commodo id velit id, eleifend volutpat quam.
+Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus.
+Nunc pellentesque pretium diam, quis maximus dolor faucibus id.
+Nunc convallis sodales ante, ut ullamcorper est egestas vitae.
+Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.
 
@@
 \section*{Introduction}
+Lorem ipsum dolor sit~\cite{bib1} amet, consectetur adipiscing elit.
+Curabitur eget porta erat.
+Morbi consectetur est vel gravida pretium.
+Suspendisse ut dui eu ante cursus gravida non sed sem.
+Nullam Eq~(\ref{eq:schemeP}) sapien tellus, commodo id velit id, eleifend volutpat quam.
+Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus.
+Nunc pellentesque pretium diam, quis maximus dolor faucibus id.~\cite{bib2} Nunc convallis sodales ante, ut ullamcorper est egestas vitae.
+Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.
 
-% cryptic diversity
-%   most species are still deliminated solely based on morphology
-Molecular systematics has repeatedly demonstrated the existence of cryptic species that can only be diagnosed using genetic data \cite{Stuart2006,Bickford2007,SchlickSteiner2007,Pfenninger2007,Clare2011,Funk2012}.
-In attempts to streamline the documentation of biodiversity, several methods of species delimitation that rely almost entirely on genetic data have recently been proposed \cite{Pons2006,Carstens2010,Hausdorf2010,O'Meara2010,Yang2010b,Huelsenbeck2011b}.
-Although strong caveats on the utility of these methods have been raised \cite{Bauer2000,Carstens2013}, they are nevertheless being used to name species \cite{Leache2010,Spinks2014}.
-
-In contrast to those genetically-diagnosed species, the majority of extant taxa, and almost all extinct taxa, are delimited by morphology alone.
-This disjunction complicates interpretations of variation and diversity in deep time, as apparent morphological stasis may not reflect the true underlying diversity \cite{Eldredge1972,Gould1977a,VanBocxlaer2013}.
-It also has serious implications for our records of modern biodiversity: for many museum specimens of extant taxa (e.g.
-those preserved in formalin), it is difficult to acquire the genetic data needed for non-morphological species delimitation methods.
-
-These considerations have sparked interest in whether geometric morphometric analyses can capture fine-scale variation that can be used for identifying cryptic species.
-This would make the task of identifying and maintaining endangered or conserved groups much easier and could contribute to improved classifications of extinct taxa and populations.
-Most such studies focus on using morphometrics to discover differences between taxa that were identified by other means \cite{Polly2003,Zelditch2004,Gaubert2005b,Gunduz2007,Polly2007a,Demandt2009,Markolf2013,Fruciano2016}.
-Additionally, there has been work on automated taxon identification and classification of taxa into groups \cite{Baylac2003,Dobigny2003,MacLeod2007,VandenBrink2011,Vitek2017}, as well as the development of models that combine genetic, phenotypic, and geographic data to infer evolutionary units of interest \cite{Guillot2012}.
-
-Here, we investigate the morphometric identification of cryptic species using machine learning approaches.
-We use an ensemble learning approach where multiple methods are used in order to look for consensus among their results.
-We test our approach on three datasets: plastron shape of eight species of closely related turtles, plastron shape of two subspecies of a single turtle species, and plastron shape of the \textit{Emys marmorata} species complex.
-In particular, we ask whether it is possible to determine which among a set of classification hypotheses best aligns with the observed morphology, and examine the implications of our results for the \textit{E.
-marmorata} complex.
-
-% past work on automatic taxon identification and older approaches to classifying taxa
-% why use machine learning methods
-\subsection*{Background and study system}
-Machine learning is an extension of known statistical methodology \cite{Hastie2009} that emphasizes predictive accuracy and generality often at the expense of the interpretability of individual parameters.
-Basic statistical approaches are supplemented by randomization, sorting, and partitioning algorithms, along with the maximization or minimization of summary statistics, in order to best estimate a general model for all data, both sampled and unsampled \cite{Hastie2009}.
-Machine learning approaches have found use in medical research, epidemiology, economics, and automated identification of images such as handwritten zip codes \cite{Hastie2009}.
-% additional/more specific CITATIONS?
-
-There are two major classes of machine learning method: unsupervised and supervised learning.
-Unsupervised learning methods are used with unlabeled data where the underlying structure is estimated; they are analogous to clustering and density estimation methods \cite{Kaufman1990}.
-Supervised learning methods are used with labeled data where the final output of data is known and the rules for going from input to output are inferred.
-These are analogous to classification and regression models \cite{Breiman1984,Hastie2009}.
-Our application of the supervised learning approaches used in this study illustrates only a sampling of the various methods available for fitting classification models.
-The specific methods used in this study were chosen because they are suited for cases with more two or more response classes.
-
-Geometric morphometric approaches to identifying differences in morphological variation between classes, including cryptic species, have mostly relied on methods like linear discriminate analysis and canonical variates analysis \cite{Polly2003,Zelditch2004,Gaubert2005b,Gunduz2007,Polly2007a,Francoy2009,Sztencel-Jabonka2009,Edwards2011,MitrovskiBogdanovic2013,Dillard2017}.
-Because of their similarity to multivariate approaches like principal components analysis (PCA), these methods are comparatively straightforward ways of understanding the differences in morphology between classes.
-They also benefit from producing results that can be easily visualized, which aids in the interpretation and presentation of data and results.
-Most previous morphometric studies did not assess which amongst a set of alternative classification hypotheses was optimal.
-For example, studies such as those of \cite{Caumul2005a} and \cite{Polly2007a} focused on comparing different aspects of morphology and their fidelity to a classification scheme instead of comparing the fidelity of one aspect of morphology to multiple classification schemes.
-In this context, the study of \cite{Cardini2009a} is noteworthy because they compared morphological variation in marmots at the population, regional, and species level and determined the fidelity of shape to divisions at each of these levels.
-
-Here, we used an ensemble of supervised machine learning methods to compare the congruence of morphological data to different classification hypotheses.
-Each of these methods provide different advantages for understanding how to classify specimens, as well as the accuracy of the resulting classifications.
-Machine learning methods have been combined with geometric morphometric data to study shape variation in a variety of contexts, including automated taxon identification and classification of groups \cite{Baylac2003,Dobigny2003,MacLeod2007,VanBocxlaer2010,VandenBrink2011,Navega2015}.
-% MORE
-%Although machine learning methods such as neural networks have been applied to studying shape variation, including in the context of automated taxon identification and classification of groups, the number of cases remains limited.
-In the current study, we not only consider pure classification accuracy but also use a statistic of classification strength that reflects the rate at which taxa are both accurately and inaccurately classified: the area under the Receiver Operating Characteristic curve \cite{Hastie2009}.
-
-We analyzed the problem of whether there are distinct subspecies or cryptic species within the western pond turtle, \textit{Emys marmorata} \cite{Baird1852} (formerly \emph{Clemmys marmorata}; see \cite{Feldman2002}).
-\textit{Emys marmorata} is distributed from northern Washington State, USA to Baja California, Mexico; populations in western Nevada may have been introduced by recent human activity or they could be a genuine part of the species' range \cite{Bury2017}.
-Traditionally, \textit{E.
-marmorata} was classified into two named subspecies: the northern \textit{E.
-marmorata marmorata} and the southern \textit{Emys marmorata pallida} \cite{Seeliger1945}, with a central Californian intergrade zone in between.
-\textit{Emys marmorata marmorata} is differentiated from \textit{E.
-marmorata pallida} by the presence of a pair of triangular inguinal scales and darker neck markings.
-The triangular inguinal plates can sometimes be present in \textit{E.
-marmorata pallida} although they are considerably smaller.
-\cite{Seeliger1945} did not formally include the Baja California populations of \textit{E.
-marmorata} in either taxon, implying the existence of a third distinct but unnamed subspecies.
-
-Previous work on morphological variation in \textit{E.
-marmorata} has focused primarily on differentiation between populations over a portion of the species' total range \cite{Lubcke2007,Germano2008,Germano2009,Bury2010}; comparatively few studies have included specimens from across the entire range \cite{Holland1992}.
-Most of these studies considered how local biotic and abiotic factors may contribute to differences in carapace length, and they found that size can vary greatly between different populations \cite{Lubcke2007,Germano2008,Germano2009}.
-There also has been interest in size-based sexual dimorphism in \textit{E.
-marmorata} \cite{Holland1992,Lubcke2007,Germano2009}, with males being on average larger than females based on total carapace length and other linear measurements.
-However, the quality of size as a classifier of sex can vary greatly between populations \cite{Holland1992} because of the magnitude of size differences among populations \cite{Lubcke2007,Germano2009}.
-The effect of sexual dimorphism on shape, \textit{sensu} \cite{Kendall1977a}, has not been assessed \cite{Holland1992,Lubcke2007,Germano2008}.
-
-Of particular relevance in the context of cryptic diversity in \textit{E.
-marmorata} is the morphometric analysis of carapace shape carried out by \cite{Holland1992}, who compared populations of \textit{E.
-marmorata} from three areas of the species' range.
-Holland concluded that geographic distance was a poor indicator of morphological differentiation, and instead hypothesized that geographic features such as breaks between different drainage basis are probably more important barriers to dispersal and interbreeding.
-Additionally, he suggested that morphological differences were more pronounced as the magnitude of barriers and distance increased, but this variation required many variables to adequately capture, implying only very subtle morphological differentiation between putatively distinct populations.
-Finally, Holland concluded that \textit{E.
-marmorata} is best classified as three distinct species: a northern species, a southern species, and a Columbia Basin species.
-This classification is similar to that of \cite{Seeliger1945}, except elevated to the species level and without recognition of a distinct Baja species.
-
-More recently, the phylogeography of \textit{E.
-marmorata} and the possibility of cryptic diversity was investigated using molecular data \cite{Spinks2005,Spinks2010,Spinks2014}.
-Based on mitochondrial DNA, \cite{Spinks2005} recognized four subclades within \textit{E.
-marmorata}, a northern clade, a San Joaquin Valley clade, a Santa Barbara clade, and a southern clade.
-Analyses with nuclear DNA \cite{Spinks2010} and single-nucleotide polymorphism (SNP) data suggest a primarily north--south division in \textit{E.
-marmorata}, although these datasets differed from that of mitocondrial-based results of \cite{Spinks2005} in the location of the break point \cite{Spinks2014}.
-All three studies discussed the potential taxonomic implications of their results, with \cite{Spinks2014} going so far as to strongly advocate for the recognition of at least two species (\emph{E.
-marmorata} and \emph{E.
-pallida}), and a possible third based on populations in Baja California.
-However, they did not discuss in detail the morphological characters that would help to diagnose these species beyond those specified by \cite{Seeliger1945}.
-Given that these characters are variable within the proposed species, and that \cite{Holland1992} described shell shape variation that might be consistent with this taxonomy, a geometric morphometric analysis of shell shape might provide a reliable way to diagnose groups (whether species or subspecies) within \textit{E.
-marmorata}.
-
-In this study, we attempt to estimate the best classification scheme of \textit{E.
-marmorata} based on variation in plastron (ventral shell) shape in order to determine whether this character is consistent with any of the proposed taxonomies of the \textit{E.
-marmorata} complex.
-
-We choose to analyze plastron shape for multiple reasons.
-First, it is very easy to collect geometric morphometric data on plastron shape from two-dimensional pictures as the structure is virtually flat.
-This approach allows both museum specimens and individuals in the field to be analyzed together.
-Second, previous work has suggested that there are strong differences in plastron shape among traditionally-recognized emydine species \cite{Angielczyk2007,Angielczyk2011,Angielczyk2013a}.
-Finally, due to these previous studies a large dataset was readily available.
-
-In the case of the \textit{E.
-marmorata} species complex, we hypothesize that if one or more of the proposed classification schemes are consistent with the morphological data then our ensemble approach fit to those hypotheses will have higher out-of-sample predictive performance than the more inconsistent hypotheses.
-However, if all of the classification schemes lead to equal out-of-sample predictive performance then we would conclude that the proposed hypotheses are inconsistent with whatever information is present in the morphological data.
-Because of unclear geographic boundaries between subgroups of \textit{E.
-marmorata}, we compare multiple permutations of the \cite{Spinks2010} and \cite{Spinks2014} hypotheses.
+\begin{eqnarray}
+\label{eq:schemeP}
+	\mathrm{P_Y} = \underbrace{H(Y_n) - H(Y_n|\mathbf{V}^{Y}_{n})}_{S_Y} + \underbrace{H(Y_n|\mathbf{V}^{Y}_{n})- H(Y_n|\mathbf{V}^{X,Y}_{n})}_{T_{X\rightarrow Y}},
+\end{eqnarray}
 
 \section*{Materials and methods}
-\subsection*{Specimens, sampling, morphometrics}
+\subsection*{Etiam eget sapien nibh}
 
-\afterpage{\clearpage}
-\begin{figure}[ht]
-  \centering
-  \includegraphics[height = 0.8\textheight, width = \textwidth, keepaspectratio = true]{figure/Ken_Ang_SpLoc_final_comp}
-  \caption{Geographic distribution of specimens sampled for comparing the hypothesized subdivisions of \textit{Emys marmorata}.
-Each hypothesized scheme has two or more possible classes.
-Sample size differs between schemes because of our ability to confidently assign museum specimens to the various schemes.
-The number of localities shown on each map is less than the number of specimens sampled because some localities produced multiple specimens.
-The different classification abbreviations are as follows: \textit{E.
-marmorata} = ``marm'', \textit{E.
-pallida} = ``pall'', Central Coast Ranges = ``CCR'', San Joaquin Valley = ``SJ,'' Baya Peninsula = ``Baja,'', and Sierra Foothills = ``Foothill.''}
-  \label{fig:map}
+% For figure citations, please use "Fig" instead of "Figure".
+Nulla mi mi, Fig~\ref{fig1} venenatis sed ipsum varius, volutpat euismod diam.
+Proin rutrum vel massa non gravida.
+Quisque tempor sem et dignissim rutrum.
+Lorem ipsum dolor sit amet, consectetur adipiscing elit.
+Morbi at justo vitae nulla elementum commodo eu id massa.
+In vitae diam ac augue semper tincidunt eu ut eros.
+Fusce fringilla erat porttitor lectus cursus, \nameref{S1_Video} vel sagittis arcu lobortis.
+Aliquam in enim semper, aliquam massa id, cursus neque.
+Praesent faucibus semper libero.
+
+% Place figure captions after the first paragraph in which they are cited.
+\begin{figure}[!h]
+\caption{{\bf Bold the figure title.}
+Figure caption text here, please use this space for the figure panel descriptions instead of using subfigure commands. A: Lorem ipsum dolor sit amet. B: Consectetur adipiscing elit.}
+\label{fig1}
 \end{figure}
 
-\afterpage{\clearpage}
-\begin{figure}[ht]
+% Results and Discussion can be combined.
+\section*{Results}
+Nulla mi mi, venenatis sed ipsum varius, Table~\ref{table1} volutpat euismod diam.
+Proin rutrum vel massa non gravida.
+Quisque tempor sem et dignissim rutrum.
+Lorem ipsum dolor sit amet, consectetur adipiscing elit.
+Morbi at justo vitae nulla elementum commodo eu id massa.
+In vitae diam ac augue semper tincidunt eu ut eros.
+Fusce fringilla erat porttitor lectus cursus, vel sagittis arcu lobortis.
+Aliquam in enim semper, aliquam massa id, cursus neque.
+Praesent faucibus semper libero.
+
+% Place tables after the first paragraph in which they are cited.
+\begin{table}[!ht]
+\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
 \centering
-  \includegraphics[height = 0.5\textheight, width = \textwidth, keepaspectratio = true]{figure/plastra}
-  \caption{Depiction of general plastral shape of \textit{E. marmorata} and position of the 19 landmarks used in this study. Anterior is towards the top of the figure.}
-  \label{fig:plastra}
-\end{figure}
-
-Three different geometric morphometric datasets describing turtle plastron variation were assembled for this analysis: 1) specimens from eight distinct emydine species; 2) \textit{T.
-scripta} specimens from the two main subspecies (\textit{T.
-scripta elegans} and \textit{T.
-scripta scripta}); and 3) \textit{E.
-marmorata} specimens from across the species' geographic range.
-The first two datasets are intended to serve as a test of whether machine learning techniques can differentiate species-level groupings of emydine turtles using plastron shape.
-We expect that the first case represents a low complexity dataset because of the high level of plastron shape disparity that exists among these species \cite{Claude2003a,Claude2006,Angielczyk2011}, whereas the second dataset should be relatively higher in complexity and more analogous to the \textit{E.
-marmorata} example.
-We predict that the \textit{E.
-marmorata} dataset should be of the highest complexity and our greatest challenge given the finding that only very subtle differences existed between geographically-distinct populations \cite{Holland1992}.
-%We expect the first dataset represents a low-complexity dataset given the high level of plastron shape disparty that exists among these species \cite{Angielczyk2011}, whereas the second dataset should be relatively higher in complexity than the first.
-Ultimately, we expect that the third dataset to be of the highest complexity and our biggest analytical challenge.
-For all datasets we chose to focus on adults because significant changes in plastron shape occur over the course of ontogeny in \textit{E.
-marmorata} and other emydines \cite{Angielczyk2013a}.
-% three turtle datasets
-%   eight species
-%   Trachemys species
-%   Emys complex
-% 26 landmarks
-
-The first dataset we analyzed includes 578 total specimens from the following species: \textit{Chrysemys picta}, \textit{Clemmys guttata}, \textit{Emys blandigii}, \textit{Emys orbicularis}, \textit{Glyptemys insculpta}, \textit{Glyptemys muhlenbergii}, \textit{Terrapene coahuila}, and \textit{Terrapene ornata}.
-These specimens are a subset of those used in \cite{Angielczyk2011} and \cite{Angielczyk2013a}.
-
-The second dataset is a compilation of 101 specimens of two subspecies of \textit{T.
-scripta}: 51 specimens of \textit{T.
-scripta scripta} and 50 specimens of \textit{T.
-scripta elegans}.
-These landmark data are new to this study.
-
-The final dataset is of 532 adult \textit{E.
-marmorata} museum specimens, though not all specimens were able to be assigned a class for all schemes (Fig.
-\ref{fig:map}).
-These specimens represent a subset of those included in \cite{Angielczyk2007}, \cite{Angielczyk2011}, and \cite{Angielczyk2013a}.
-Because  \cite{Spinks2005}, \cite{Spinks2010}, and \cite{Spinks2014} did not use vouchered specimens we were not able to directly sample the individuals in their studies.
-Instead, our specimen classifications were based solely on the geographic information and not explicit assignment using molecular data.
-For each taxonomic hypothesis, specimens were assigned to one of the possible classes based on geographic occurrence data recorded in museum collections.
-In cases where precise latitude and longitude information were not available we estimated them from other locality information.
-Because the exact barriers between different biogeographic regions are unknown and unclear, we represented each hypothesis with multiple possible realizations representing the classification uncertainty for specimens present at the geographic boundaries.
-The taxonomic hypotheses and sub-hypotheses for \textit{E.
-marmorata} used here are presented in Table \ref{tab:hypotheses} and Figure \ref{fig:map}.
-
-For \cite{Spinks2010} we used three binning schemes.
-All three schemes include a class for \textit{E.
-marmorata} specimens from northern populations (marm) as well as a class for those assigned to \textit{E.
-pallida} (pall) and an intergrade zone in the Central Coast Ranges (CCR).
-The schemes differ in the assignment of samples from the San Joaquin Valley (Fig.
-\ref{fig:map}).
-Scheme SP10.1 and SP10.2 differ in the assignment of specimens from the western San Joaquin Valley to either CCR or marm reflecting uncertainty regarding their genetic affinity as explained above.
-In scheme SP10.3 these specimens are assigned to a San Joaquin class reflecting the mitochondrial distinctiveness shown by \cite{Spinks2005}.
-For \cite{Spinks2014} we used two binning schemes with SP14.1 being based on their phylogenetic network analysis and SP14.2 being based on their Bayesian species delimitation analysis.
-The latter scheme requires the addition of two new classes, ``Baja'' and ``Foothill,'' to accommodate the genetic groupings recovered by the SNP Structure analysis that was used to create the guide tree for the BPP species delimitation analysis in \cite{Spinks2014}.
-Finally, we proposed a conservative morphological hypothesis (``Morph'') in order to compare the molecular hypotheses with something approximating the original taxonomic hypothesis for the group; this scheme is made up solely of the marm and pall classes from the SP10.3 scheme.
-
-\begin{table}
-  \centering
-    \caption{Table of species delimitation hypotheses for \textit{E. marmorata}}
-    \begin{tabular}{l l l }
+\caption{
+{\bf Table caption Nulla mi mi, venenatis sed ipsum varius, volutpat euismod diam.}}
+\begin{tabular}{|l+l|l|l|l|l|l|l|}
 \hline
-      Abbreviation & Number of classes & citation \\
-      \hline
-      SP10.1 & 3 & \cite{Spinks2010} \\
-      SP10.2 & 3 & \cite{Spinks2010} \\
-      SP10.3 & 4 & \cite{Spinks2010} \\
-      SP14.1 & 2 & \cite{Spinks2014} \\
-      SP14.2 & 4 & \cite{Spinks2014} \\
-      Morph & 2 & \cite{Spinks2010} \\
-      \hline
+\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \thickhline
+$cell1 row1$ & cell2 row 1 & cell3 row 1 & cell4 row 1 & cell5 row 1 & cell6 row 1 & cell7 row 1 & cell8 row 1\\ \hline
+$cell1 row2$ & cell2 row 2 & cell3 row 2 & cell4 row 2 & cell5 row 2 & cell6 row 2 & cell7 row 2 & cell8 row 2\\ \hline
+$cell1 row3$ & cell2 row 3 & cell3 row 3 & cell4 row 3 & cell5 row 3 & cell6 row 3 & cell7 row 3 & cell8 row 3\\ \hline
 \end{tabular}
-    \label{tab:hypotheses}
+\begin{flushleft} Table notes Phasellus venenatis, tortor nec vestibulum mattis, massa tortor interdum felis, nec pellentesque metus tortor nec nisl.
+Ut ornare mauris tellus, vel dapibus arcu suscipit sed.
+\end{flushleft}
+\label{table1}
+\end{adjustwidth}
 \end{table}
 
-Sex was known only for a subset of the total dataset and was not included as a predictor of classification.
-Instead, we estimated the degree by which specimens cluster morphologically by sex in order to determine how much of a potential biasing factor sexual dimorphism could be for our analysis of the \textit{E.
-marmorata} species complex (see below).
-%The scheme names are as follows: Mito 1 and 2 correspond to \cite{Spinks2005}, Mito 3 corresponds to \cite{Spinks2010}, Morph 1 and Morph 2 correspond to \cite{Holland1992}, and Nuclear corresponds to \cite{Spinks2014}.
-% CHANGE
+%PLOS does not support heading levels beyond the 3rd (no 4th level headings).
+\subsection*{\lorem\ and \ipsum\ nunc blandit a tortor}
+\subsubsection*{3rd level heading} 
+Maecenas convallis mauris sit amet sem ultrices gravida.
+Etiam eget sapien nibh.
+Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
+Curabitur fringilla pulvinar lectus consectetur pellentesque.
+Quisque augue sem, tincidunt sit amet feugiat eget, ullamcorper sed velit.
+Sed non aliquet felis.
+Lorem ipsum dolor sit amet, consectetur adipiscing elit.
+Mauris commodo justo ac dui pretium imperdiet.
+Sed suscipit iaculis mi at feugiat.
 
-Following previous work on plastron shape \cite{Angielczyk2007,Angielczyk2011,Angielczyk2013a}, we used TpsDig 2.04 \cite{Rohlf2005} to digitize 19 two-dimensional landmarks (Fig.
-\ref{fig:plastra}).
-Seventeen of the landmarks are at the endpoints or intersection of the keratinous plastral scutes that cover the plastron.
-Twelve of the landmarks were symmetrical across the axis of symmetry.
-Because damage prevented the digitization of all the symmetric landmarks in some specimens, we reflected landmarks across the axis of symmetry (i.e.
-midline) prior to analysis and used the average position of each symmetrical pair.
-In cases where damage or incompleteness prevented symmetric landmarks from being determined, we used only the single member of the pair.
-We conducted all subsequent analyses on the resulting ``half'' plastra.
-We superimposed the plastral landmark configurations using generalized Procrustes analysis \cite{Dryden1998a}, after which we calculated the principal components (PC) of shape using the \texttt{shapes} package for R \cite{R2016,Dryden2013}.
-All specimens were used for superimposition, after which the subset labeled for each of the schemes were used in model training and testing (see below).
+\begin{enumerate}
+	\item{react}
+	\item{diffuse free particles}
+	\item{increment time by dt and go to 1}
+\end{enumerate}
 
-\subsection*{Biasing effects}
-% digitization
-We estimated the possible effect of digitization error \cite{Arnqvist1998,Cramon2007,Munoz-MunozF.2010} on our results by comparing within-specimen (replicated) Procrustes distances to the distances between classification scheme centroids.
-Ten randomly-selected \textit{E.
-marmorata} specimens were each digitized four times, with the original set of digitized coordinates serving as a fifth replicate.
-These 50 landmark configurations were then Procrustes superimposed.
-A range of four Procrustes distances was then calculated as the average of the pairwise distances between each of the replicate configurations of a given specimen.
+\subsection*{Sed ac quam id nisi malesuada congue}
 
-For each specimen, the difference in shape caused by digitization was calculated as the mean of all pairwise Procrustes distances between the five replicates of that specimen.
-The average distance between any two digitizations was calculated as the mean of all pairwise Procrustes distances between all replicates for all specimens.
-The ratio between these two values was used to assess the magnitude of variation caused by digitization.
-The goal of this ratio is to determine if the within group distances are on average smaller than the between individual distances; a value of 0 indicates perfect grouping, a value of 1 indicates no difference between grouping and no grouping, and a value of 1+ indicates that the grouping is counter-intuitive to the data.
+Nulla mi mi, venenatis sed ipsum varius, volutpat euismod diam.
+Proin rutrum vel massa non gravida.
+Quisque tempor sem et dignissim rutrum.
+Lorem ipsum dolor sit amet, consectetur adipiscing elit.
+Morbi at justo vitae nulla elementum commodo eu id massa.
+In vitae diam ac augue semper tincidunt eu ut eros.
+Fusce fringilla erat porttitor lectus cursus, vel sagittis arcu lobortis.
+Aliquam in enim semper, aliquam massa id, cursus neque.
+Praesent faucibus semper libero.
 
-\textit{Emys marmorata} is known to display sexual dimorphism in plastral shape, particularly the presence of a plastra concavity in males \cite{Seeliger1945}.
-To test for biases resulting from sexual dimorphism in our \textit{E.
-marmorata} dataset, we used a simple permutation test to determine if the distance between the mean female and male shapes is greater than expected when the sex labels are randomly shuffled.
-Because not all of our specimens have sex identifications associated with them, this analysis was done using a subset of the data (257 of 532).
-
-\subsection*{Supervised learning approaches}
-
-\begin{table}
-  \centering
-  \caption{Table of the supervised learning methods used in this analysis.}
-  \scalebox{0.9}{
-  \begin{tabular}{l l l l}
-    \hline
-    Method name & abbreviation & R package & citation \\
-    \hline
-    multinomial logistic regression & MLR & nnet & \cite{Venables2002} \\
-    linear discriminate analysis & LDA & MASS & \cite{Venables2002} \\
-    penalized discrminiate analysis & PDA & mda & \cite{mdapack} \\
-    single-hidden-layer neural network & NN & nnet & \cite{Venables2002} \\
-    random forests & RF & randomForest & \cite{Liaw2002} \\
-    \hline
-  \end{tabular}}
-  \label{tab:methods}
-\end{table}
-
-Instead of relying on a single supervised learning method, we chose to use an ensemble approach where multiple model types are used in concert so that any congruence between them increases our support for that conclusion over another \cite{Hastie2009}.
-The supervised learning methods used here are named in Table \ref{tab:methods}.
-Each of these methods makes different assumptions, treats data differently, and can produce different classification results depending on the nature of the data \cite{Hastie2009}.
-For example, multinomial logistic regression is a type of generalized linear model, whereas random forest is itself an ensemble approach where multiple decision trees are fit to subsets of the full dataset and then averaged.
-% use multiple methods
-%   multinomial logistic regression (nnet)
-%   linear discriminate analysis (MASS)
-%   penalized discriminate analysis (mda)
-%   neural network (nnet)
-%   random forest (randomForest)
-% each method has different assumptions and treat data differently
-%   all assume that predictors have additive effect (e.g. independent)
-
-% predictors/features/covariates
-%   1:25 PCs
-%   size
-%   size X PC1
-% analogy to PCA regression
-The maximum set of possible predictors or features used for any model of our dataset is comprised of the first 25 principal components (PCs), scaled centroid size, and the interaction between scaled centroid size and PC 1.
-Additional interaction terms were not considered because of model complexity/sample size concerns.
-Size and the interaction between size and PC 1 were included as predictors to account for known ontogenetic variation in plastron shape \cite{Angielczyk2013a} as well as potential size differences between classes, even if this is unlikely \cite{Seeliger1945,Holland1992}.
-These data constitute a ``maximum set'' because the best or selected models based on five-fold cross-validation need not, and likely will not, include all predictors possible (see below).
-Because our supervised learning models use PCs as predictors, this approach is in many ways analogous to PCA regression.
-PCA regression takes advantage of reduction and orthogonality PCs to improve regression fit \cite{Hastie2009}.
-Because the PCs of shape are by definition orthogonal, they can easily serve as independent predictors or features of class membership without fear of collinearity.
-
-% training and testing paradigm
-%   training dataset
-%     for multiple models with between 3 and 28 predictors
-%     5-fold cross validation of each model
-%     best model had greatest mean AUC value
-%     selected model is most parsimonious model within 1 SD of ROC of best model
-%   testing dataset
-%     estimate class of out-of-sample testing dataset
-%     compare average AUC across models for each scheme
-% caret package for R
-We adopted a training and testing paradigm for selecting parsimonious models and estimating their overall error rates \cite{Hastie2009,Kuhn2013}.
-Within-sample model performance is inherently biased upwards, so model evaluation requires overcoming this bias.
-With very large sample sizes, as in this study, part of the sample can be used as the ``training set'' and the remainder acts as the ``testing set.'' In this approach, following all cleaning and vetting, the data are split into a training dataset and a testing dataset.
-The former is used for fitting the model whereas the later is used for measuring model performance, a process called model generalization.
-For each scheme, we limited the model training and testing to only those individuals with class labels for that scheme.
-In this analysis, we randomly divided 80\% of samples into the training set and the remaining 20\% into the testing set.
-
-% auc of roc as performance measure
-%   relationship between false positive rate and true positive rate
-%   varies between 0.5 and 1 (random -- perfect)
-%   better performance than accuracy when groups unbalanced
-In classification studies, such as this one, a common metric of performance is the receiver operating characteristic (ROC) which is the relationship between the false and true positive rates \cite{Hastie2009}.
-The area under the ROC curve (AUC) is the derived estimate of the model performance; AUC ranges from 0.5 to 1 which correspond to performance similar to random guesses and perfect classification rates, respectively \cite{Hastie2009}.
-Both ROC and AUC are preferable to simple classification accuracy when class membership is unbalanced, as it is in these analyses \cite{Hastie2009}.
-The standard ROC and AUC calculations are defined only for binary classifications, which is not the case for our eight species and \textit{Emys} complex datasets.
-To generalize this approach for situations with multiple response classes, we used an all-against-one strategy where the model AUC is the average of the AUC values from the multiple binary comparisons of one class compared to all others \cite{Hand2001}.
-
-For a given supervised learning method, we compared the fit of 27 models as the average AUC from 10 rounds of five-fold cross-validation.
-Cross-validation is an approach for estimating the average out-of-sample predictive error of a model by simulating out-of-sample data from the training dataset itself \cite{Hastie2009}.
-In a single round of \(k\)-fold cross-validation, the training data are divided into \(k\) blocks where the model is fit to \(k - 1\) blocks and the values of the \(k\)th block are predicted.
-This is repeated for all combinations of blocks.
-Within each round, the predictive performance metrics are averaged across all folds.
-Finally, the predictive performance metric is the averaged across all rounds of \(k\)-fold cross-validation.
-This process was implemented using the R package \texttt{caret} \cite{KuhnMAN2013}.
-For a given supervised learning method, the ``best'' trained model is that with the highest mean AUC as estimated from five-fold cross-validation.
-The selected or final model, however, is the next most parsimonious model that is within one standard error of the best model; this is a variant on the ``one-standard error'' rule from \cite{Hastie2009}.
-The purpose of this rule is to ameliorate the chances of selecting an overly complex model that will perform poorly when predicting the classes of out-of-sample data.
-
-\section*{Results}
-
-\subsection*{Geometric morphometrics}
-
-\begin{figure}[h]
-  \centering
-  \begin{subfigure}[b]{0.7\textwidth}
-    \caption{}
-    \includegraphics[width = \textwidth]{figure/cc7_pc_graph}
-  \end{subfigure}
-
-  \begin{subfigure}[b]{0.7\textwidth}
-    \caption{}
-    \includegraphics[width = \textwidth]{figure/tra_pc_graph}
-  \end{subfigure}
-  \caption{Two scatterplots of morphological differences from two of the three datasets analyzed in this study.
-(a) Scatterplot of the first two PCA axies from the landmarks from the eight different species dataset, and (b) the first two axes of variation from two subspecies of \textit{Trachemys} dataset.
-Point colors correspond to the categories within each dataset while point size is proportional to individual centroid size.
-In parentheses next to the axis labels are the percent of total variation accounted for by that axis.
-For both datasets there are clear distinctions between the different categories.}
-  \label{fig:other_pca}
-\end{figure}
-
-\afterpage{\clearpage}
-\begin{figure}[ht]
-  \centering
-  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/emys_pc_graph}
-  \caption{Scatterplot of the first two axes of morphological variation in the \textit{Emys marmorata} dataset.
-Each panel corresponds to one of the six different classification schemes analyzed as part of this study (Tab.
-\ref{tab:hypotheses}).
-Point color corresponds to the categories within each scheme, and the class names correspond to geographic regions.
-Point size is proportional to centroid size of that specimen and the numbers in parentheses next to the axis labels are the percent of total variation accounted for along that axis.}
-  \label{fig:emys_pca}
-\end{figure}
-
-The results of the PCA of plastron shape in both the eight species and \textit{Trachemys} datasets demonstrate strong association between shape and the recognized classification schemes (Fig.
-\ref{fig:other_pca}).
-
-The results of the PCA of plastron shape in the \textit{Emys marmorata} dataset show no clear connection between plastron shape and any of the proposed classification schemes (Fig.
-\ref{fig:emys_pca}).
-The first PC axis of shape variation appears to be primarily structured by differences in individual centroid size (Fig.
-\ref{fig:emys_pca}); this was the motivation for including centroid size and its interaction with PC1 as predictors in all of the supervised learning models.
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[height = 0.3\textheight, width = \textwidth, keepaspectratio = true]{figure/sex_test_hist}
-  \caption{Comparison of observed Procrustes distance between the centroids of each sex (vertical line) to a null distribution generated from 1000 permutations of the sex-labels.
-This result indicates that the difference between the centroids is as small/smaller than expected by random.}
-  \label{fig:sex_test}
-\end{figure}
-
-Analysis of the differences between sexes of \textit{E.
-marmorata} indicates that sex does not appear to strongly structure differences in shape (Fig.
-\ref{fig:sex_test}).
-The difference in mean shape between the sexes is very small; the sexes overlap about has much as expected given a null distribution based on permuting the sex-labels.
-
-\begin{table}
-  \centering
-  \caption{Results from the within-individual to between-individual Procrustes distances for the replicated plastron shape data.
-Results are presented for all three datasets analyzed here: the \textit{Trachemys} dataset, the eight species dataset, and each of the \textit{Emys marmorata} classification schemes.}
-  \begin{tabular}{l l c c}
-    \textbf{Dataset} & \textbf{Scheme} & \textbf{Ratio} & \textbf{Corrected ratio} \\
-    \hline
-    Replicates & & 1.11 & \\
-    \hline
-    Seven species & & 0.33 & 0.37 \\
-    \textit{Trachemys} & & 0.76 & 0.84 \\
-    \hline
-    \textit{E. marmorata} & SP10.1 & 0.99 & 1.10 \\
-     & SP10.2 & 1.00 & 1.11 \\
-     & SP10.3 & 0.94 & 1.04 \\
-     & SP14.1 & 1.01 & 1.12 \\
-     & SP14.2 & 0.93 & 1.04 \\
-     & Morph & 0.99 & 1.09 \\
-    \hline
-  \end{tabular}
-  \label{tab:rep_res}
-\end{table}
-
-Comparison of the within to between Procrustes distances of the digitization replicates gives an approximate estimate of the error between distinct groupings (Table \ref{tab:rep_res}).
-The ratio of the average within-individual distance to the average distance between individuals for the replicated datasets is 1.11; this indicates that the grouping is slightly counter-intuitive to the data and is consistent with all shapes being very similar regardless of individual identity.
-This value also provides a baseline by which to understand how distinct the groupings are, where other ratios are compared to the correction ratio \(1.11/1\).
-
-The results from the eight species and \textit{Trachemys} datasets indicate that both of these classification schemes are more recognizable than not given our estimate of digitization error (Table \ref{tab:rep_res}).
-In contrast, the different \textit{E.
-marmorata} classification schemes appear to barely be distinct, with their within:between ratios approximating 1.
-This indicates that the magnitude of the differences between groupings is approximately the same as the difference between any two random individuals (Table \ref{tab:rep_res}).
-
-\subsection*{Supervised learning}
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[height = 0.5\textheight, width = \textwidth, keepaspectratio = true]{figure/other_model_sel}
-  \caption{Comparisons of model fit to the training dataset for each of the supervised learning methods applied to the first two datasets; the results from the eight species dataset are presented in the left column, while those from the \textit{Trachemys} dataset are presented in the right column.
-Models were fit to datasets of varying complexity, with the number of parameters listed along the x-axis.
-Model fit is measured as the area under the receiver operating characteristic (AUC), which ranges from 0.5 to 1.
-Error bars correspond to one standard error estimated from 10 rounds of 5-fold cross-validation.
-The red dot corresponds to the model fit with the highest mean AUC while the blue dot corresponds to the model selected for further analysis.
-In some cases, there is no difference in complexity between the best and selected models.}
-  \label{fig:other_sel}
-\end{figure}
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[height = 0.3\textheight, width = \textwidth, keepaspectratio = true]{figure/other_oos_sel}
-  \caption{The results of out-of-sample predictive performance of the selected models for both the eight species (left) and \textit{Trachemys} datasets.
-Predictive performance is measured as the area under the receiver operating characteristic (AUC), which ranges from 0.5 to 1.
-Points correspond to the individual out-of-sample predictive performance of the specific model, indicated along the x-axis.
-The horizontal bars correspond the average out-of-sample predictive performance of all the models.}
-  \label{fig:other_oos}
-\end{figure}
-
-Analysis of the eight morphologically- and genetically-distinct species and the \textit{T.
-scripta scripta}--\textit{T.
-scripta elegans} datasets indicate that these taxa are sufficiently morphologically distinct to be differentiated on the basis of plastron shape.
-Both in-sample and out-of-sample classification have AUC values of approximately 1 for all methods, implying near-perfect classification rates (Fig.
-\ref{fig:other_sel}, \ref{fig:other_oos}).
-For both datasets, the ROC scores from testing datasets are tightly clustered near AUC = 1 (Fig.
-\ref{fig:other_oos}).
-These results demonstrate that when there are distinctions between the states of the classification schemes (i.e., differences in plastron shape that correlate with the different taxonomic groups), the methods used here can recover them.
-
-\afterpage{\clearpage}
-\begin{figure}[ht]
-  \centering
-  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/emys_model_sel}
-  \caption{AUC values for models of varying complexity fit to the \textit{Emys marmorata} training datasets for each classification scheme.
-The x-axis corresponds to the total number of predictors included in each model, while the y-axis corresponds to the AUC value which is a measure of goodness of fit for classification datasets.
-A model with a high AUC value corresponds to better classification performance than a model with a lower AUC value.
-Standard errors on AUC estimates are calculated from 10 rounds of 5-fold cross-validation.
-Indicated are the best performing and the selected models, in red and blue respectively.
-In some cases, there is no difference in complexity between the best and selected models.}
-  \label{fig:emys_sel}
-\end{figure}
-
-\afterpage{\clearpage}
-\begin{figure}[ht]
-  \centering
-  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/emys_oos_sel}
-  \caption{Comparison of out-of-sample AUC estimates from the predictions of selected models (Fig.
-\ref{fig:emys_sel}), grouped by classification scheme.
-The horizontal line in each panel corresponds to the average AUC value across all models of that classification scheme.}
-  \label{fig:emys_oos}
-\end{figure}
-
-AUC--based model selection revealed some important patterns of variation and congruence between the classification schemes and the actual data.
-Generally, the best performing models tended to include about half the total number of possible PCs (Fig.
-\ref{fig:emys_sel}).
-
-Observed AUC values for all of the optimal models are lower for the \textit{E.
-marmorata} dataset than for the other two datasets (Fig.
-\ref{fig:other_sel}, \ref{fig:emys_sel}).
-In most cases the different proposed classification schemes are generally poor descriptors of the observed variation.
-It appears that the dataset is overwhelmed by noise (likely biological and analytical), making any accurate classifications difficult at best.
-This observation is cemented with the generalizations of the models to the testing dataset (Fig.
-\ref{fig:emys_oos}).
-
-Mean AUC values for the model generalizations, in most cases, are approximately equal to the observed AUC values from the training dataset (Fig.
-\ref{fig:emys_sel}, \ref{fig:emys_oos}).
-The  cases in which the AUC from the  generalizations is less than the observed indicate poor model fit and a poor classification scheme.
-Comparison of AUC values from the model generalizations do not indicate a clear ``best'' classification scheme (Fig.
-\ref{fig:emys_sel}, \ref{fig:emys_oos}).
-Only in the case of the conservative morphological hypothesis (``Morph'') is the mean AUC value potentially distinct from that of other schemes; in this case mean AUC is lower than the average of the other five schemes which indicates that the moprhologically-based scheme performs more poorly than the molecularly-based ones.
-It is important to note, however, that the training and testing dataset for the ``Morph'' scheme is the smallest of the six schemes which may lead to poorer performance in in-sample and out-of-sample comparisons.
+\begin{itemize}
+	\item First bulleted item.
+	\item Second bulleted item.
+	\item Third bulleted item.
+\end{itemize}
 
 \section*{Discussion}
+Nulla mi mi, venenatis sed ipsum varius, Table~\ref{table1} volutpat euismod diam.
+Proin rutrum vel massa non gravida.
+Quisque tempor sem et dignissim rutrum.
+Lorem ipsum dolor sit amet, consectetur adipiscing elit.
+Morbi at justo vitae nulla elementum commodo eu id massa.
+In vitae diam ac augue semper tincidunt eu ut eros.
+Fusce fringilla erat porttitor lectus cursus, vel sagittis arcu lobortis.
+Aliquam in enim semper, aliquam massa id, cursus neque.
+Praesent faucibus semper libero~\cite{bib3}.
 
-As expected, our ensemble approach yields high out-of-sample classification performance for the first two datasets.
-These results indicate that in cases of clear class separation (Fig.
-\ref{fig:other_pca}) our approach is able to detect this and make good out-of-sample prediction.
+\section*{Conclusion}
 
-In the case of the \textit{E.
-marmorata} dataset, our results show that none of the proposed taxonomic hypotheses for the \textit{E.
-marmorata} species complex are more consistent with morphological differentiation than any other proposal (Fig.
-\ref{fig:emys_oos}).
-Both the low out-of-sample AUC values and the significant difference between the correctly and incorrectly classified observations support the conclusion that none of the hypothesized classification schemes are good descriptions of the observed plastral variation within \textit{E.
-marmorata}.
-An analytical explanation of this result is that the level of digitization error in the \textit{E.
-marmorata} dataset is so great as to swamp out any biological signal.
-We think this is unlikely because all of the specimens considered in our three analyses were digitized by one of us (K.D.A.), and digitization error was not a problem in the eight species or \textit{Trachemys} examples.
-There are also no features of the plastron of \textit{E.
-marmorata} that would make it significantly more difficult to accurately digitize than the plastra of the other speices.
+CO\textsubscript{2} Maecenas convallis mauris sit amet sem ultrices gravida.
+Etiam eget sapien nibh.
+Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
+Curabitur fringilla pulvinar lectus consectetur pellentesque.
+Quisque augue sem, tincidunt sit amet feugiat eget, ullamcorper sed velit.
 
-Biological explanations include the possibilty that genetic differentiation is not associated with plastron shape variation and/or that local selective pressures (e.g.
-from hydrological regime) overwhelm morphological differentiation.
-Both of these options seem plausible given that shell shape is influenced by selection for both protection and streamlining, but not necessary mate choice \cite{Rivera2008,Rivera2011,Stayton2011,Rivera2014,Polly2016}, and that shell shape in \textit{E.
-marmorata} is known to vary among populations inhabiting water bodies with different flow regimes \cite{Holland1992,Lubcke2007,Germano2009}.
-Plastron shape does not seem to preserve a strong phylogenetic signal at the interspecific level in emydine turtles, at least compared to the effect of the presence or absence of a plastral hinge \cite{Angielczyk2011}, and our current results suggest that this may be the case for phylogeographic signal within emydine species as well.
-A final possibility (explored below) is that the proposed classification schemes themselves do not represent significant evolutionary lineages.
-
-Despite the negative result for \textit{E.
-marmorata}, it is important to note that plastron shape is an extremely effective method for differentiating classes in the additional datasets we investigated.
-The magnitude of shape differences between the species (measured as Procrustes distance between the eight species' mean shapes) is approximately an order of magnitude greater than the differences between the \textit{E.
-marmorata} subgroups, and not surprisingly the machine learning methods had no trouble classifying the specimens correctly.
-However, the magnitude of the shape differences between the \textit{T.
-scripta} subspecies is comparable to those separating the different \textit{E.
-marmorata} subgroups, yet even in this case the machine learning methods returned an almost perfect classification.
-These results demonstrate that plastron shape is normally a good marker for differentiating real subgroups in close relatives of \textit{E.
-marmorata}, and that our lack of results for \textit{E.
-marmorata} is not simply a shortcoming of the methods we applied.
-Indeed, it begs the question of what factors have suppressed morphological differentiation of plastron shape in \textit{E.
-marmorata} and \textit{E.
-pallida} if they are distinct species.
-Invoking issues such as the role of the plastron in protection or the need for streamlining are insufficient because the other species are expected to be subject to similar constraints \cite{Stayton2011,Polly2016}.
-Although it may seem counterintuitive that plastron shape is both useful for species delimitation but has weak or absent phylogenetic signal, it is important to remember that these are different goals.
-While phylogenetically similar species may not be morphologically similar (e.g.
-compare the box turtles of the genus \textit{Terrapene} to the closely related spotted turtle \textit{Clemmys guttata}), the variation within a species typically is much less than the variation between species.
-Therefore, the consistent plastron shapes that characterize different emydid species leads to plastron shape being a useful tool for species delimitation, even when other selective factors have overprinted similarities stemming from patterns of descent from common ancestors.
-
-\section*{Conclusions}
-
-The lack of morphological support for the distinctiveness of \textit{E.
-pallida} does not, on its own, preclude the recognition of this taxon.
-However, this apparent lack of congruence does prompt a reexamination of the methods and concepts that led to that taxonomic revision, especially considering that plastron shape is demonstrably capable of differentiating species and subspecies among other emydids.
-In other words, before we can assess the significance of the morphological non-diagnosablity, it is essential to evaluate the methods and concepts that led to the initial taxonomic revision.
-
-\cite{Spinks2014} elevated \textit{E.
-pallida} based on a species delimitation analysis of SNP data using BPP \cite{Yang2010b}.
-However, \cite{Spinks2014} did not heed the caveats about such species delimitation methods raised by \cite{Carstens2013}.
-In addition to specifically addressing the shortcomings of validation methods such as BPP that rely on guide trees and ``should be interpreted with caution,'' \cite{Carstens2013} also strongly emphasized that ``Inferences regarding species boundaries based on genetic data alone are likely inadequate, and species delimitation should be conducted with consideration of the life history, geographical distribution, morphology and behaviour (where applicable) of the focal system\dots'' These caveats evoke the development of the Unified Species Concept \cite{Dayrat2005a,DeQueiroz2007b}, Integrative Taxonomy \cite{Padial2010}, and other pluralist approaches to species delimitation.
-None of these considerations were brought to bear on the \textit{E.
-marmorata} system until now, and in doing so we find the proposal that \textit{E.
-pallida} is a distinct species to be lacking.
-
-In addition to lacking a robust morphological marker, the natural history and geographical distribution of \textit{E.
-marmorata} and \textit{E.
-pallida} also make the recognition of these two taxa implausible.
-The mitochondrial data from \cite{Spinks2014} show extensive introgression and admixture in Central California, which is expected because there are no significant barriers to gene flow in this region.
-They also lack sampling from the populations between the two putative species in the San Francisco Bay Area, which we predict would likely show even more genetic mixing.
-Combined with the well-demonstrated ability for testudinoid turtles, including emydids and even \textit{Emys}, to hybridize (e.g.
-\cite{Buskirk2005,Spinks2009,Parham2013}) it is hard to imagine how \textit{E.
-marmorata} and \textit{E.
-pallida} could maintain their integrity in the face of such admixture.
-Any argument for the validity of \textit{E.
-pallida} as a distinct species needs to address these points.
-Because the geography, natural history, limited sampling from key areas, demonstrated genetic admixture of \textit{E.
-marmorata}, and comparisons with other morphologically diagnosable species and subspecies conflict with the recognition of \textit{E.
-pallida}, we hypothesize that \textit{E.
-pallida} is not a distinct species.
-
-We fully agree with \cite{Spinks2014} that \textit{E.
-marmorata} (\textit{sensu lato}) is a species deserving of strong conservation efforts, and we do not wish to trivialize this need.
-Moreover, the genetic diversity uncovered by the analysis of \cite{Spinks2014} should be accounted for explicitly in any conservation plan.
-Given the apparent lack of morphological distinction combined with the broad range of intergradation and other problems with the species hypothesis outlined above, we recommend that the populations elevated to \textit{E.
-pallida} by \cite{Spinks2014} are best considered Evolutionary Significant Units or Distinct Population Segments instead of distinct species.
-
-Finally, it is important to note that the data and analyses we present do not let us definitively say whether the apparent lack of morphological divergence within \textit{E.
-marmorata} truly reflects the presence of a single species, or if it is an artifact of plastron shape being a poor morphological marker for phylogenetic and phylogeographic divergences in the case of \textit{E.
-marmorata}.
-This is because we could not carry out our morphometric analyses on the specimens from which the genetic data were obtained.
-The comparisons with the other emydid taxa suggest that our negative result is is because \textit{E.
-marmorata} is a single species.
-However, tests of both our preferred conclusion (\textit{E.
-marmorata} as a single species) and that of \cite{Spinks2014} should include morphological and molecular analyses of the same set of voucher specimens, as well as additional tests of species delimitation using alternative methods and corroborating evidence as suggested by \cite{Carstens2013}.
-From a morphological standpoint, support for the validity of ``\textit{E.
-pallida}'' may come from other aspects of morphology, such as carapace shape or other features.
-Likewise, further investigation of the phylogeographic utility of plastron shape in other turtle species will help to clarify whether the lack of differentiation seen in \textit{E.
-marmoarata}, and the strong differentiation among the other emydids, is typical or an unusual case.
+Sed non aliquet felis.
+Lorem ipsum dolor sit amet, consectetur adipiscing elit.
+Mauris commodo justo ac dui pretium imperdiet.
+Sed suscipit iaculis mi at feugiat.
+Ut neque ipsum, luctus id lacus ut, laoreet scelerisque urna.
+Phasellus venenatis, tortor nec vestibulum mattis, massa tortor interdum felis, nec pellentesque metus tortor nec nisl.
+Ut ornare mauris tellus, vel dapibus arcu suscipit sed.
+Nam condimentum sem eget mollis euismod.
+Nullam dui urna, gravida venenatis dui et, tincidunt sodales ex.
+Nunc est dui, sodales sed mauris nec, auctor sagittis leo.
+Aliquam tincidunt, ex in facilisis elementum, libero lectus luctus est, non vulputate nisl augue at dolor.
+For more information, see \nameref{S1_Appendix}.
 
@@
 
+\paragraph*{S2 Fig.}
+\label{S2_Fig}
+{\bf Lorem ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
+Etiam eget sapien nibh.
+Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
+Curabitur fringilla pulvinar lectus consectetur pellentesque.
+
+\paragraph*{S1 File.}
+\label{S1_File}
+{\bf Lorem ipsum.}  Maecenas convallis mauris sit amet sem ultrices gravida.
+Etiam eget sapien nibh.
+Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
+Curabitur fringilla pulvinar lectus consectetur pellentesque.
+
+\paragraph*{S1 Video.}
+\label{S1_Video}
+{\bf Lorem ipsum.}  Maecenas convallis mauris sit amet sem ultrices gravida.
+Etiam eget sapien nibh.
+Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
+Curabitur fringilla pulvinar lectus consectetur pellentesque.
+
+\paragraph*{S1 Appendix.}
+\label{S1_Appendix}
+{\bf Lorem ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
+Etiam eget sapien nibh.
+Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
+Curabitur fringilla pulvinar lectus consectetur pellentesque.
+
+\paragraph*{S1 Table.}
+\label{S1_Table}
+{\bf Lorem ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
+Etiam eget sapien nibh.
+Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
+Curabitur fringilla pulvinar lectus consectetur pellentesque.
+
 \section*{Acknowledgments}
-Data collection for this project was supported in part by NSF DBI-0306158 (to KDA).
-G.
-Miller assisted with data collection and her participation in this research was supported by NSF REU DBI-0353797 (to R.
-Mooi of CAS).
-For access to emydine specimens, we thank: J.
-Vindum and R.
-Drewes (CAS); A.
-Resetar (FMNH); R.
-Feeney (LACM); C.
-Austin (LSUMNS); S.
-Sweet (MSE); J.McGuire and C.
-Conroy (MVZ); A.
-Wynn (NMNH); P.
-Collins (SBMNH); B.
-Hollingsworth (SDMNH); P.
-Holroyd (UCMP).
-We are grateful for S.
-Sweet for field assistance and the California Department of Fish and Game for permits.
-We would also like to thank M.
-Lambruschi (FMNH) for help with figure \ref{fig:map}.
+Cras egestas velit mauris, eu mollis turpis pellentesque sit amet.
+Interdum et malesuada fames ac ante ipsum primis in faucibus.
+Nam id pretium nisi.
+Sed ac quam id nisi malesuada congue.
+Sed interdum aliquet augue, at pellentesque quam rhoncus vitae.
 
@@
 % 
-%\begin{thebibliography}{10}
-%
-%\end{thebibliography}
-\bibliography{new_bib}
+\begin{thebibliography}{10}
+
+\bibitem{bib1}
+Conant GC, Wolfe KH.
+\newblock {{T}urning a hobby into a job: how duplicated genes find new
+  functions}.
+\newblock Nat Rev Genet. 2008 Dec;9(12):938--950.
+
+\bibitem{bib2}
+Ohno S.
+\newblock Evolution by gene duplication.
+\newblock London: George Alien \& Unwin Ltd. Berlin, Heidelberg and New York:
+  Springer-Verlag.; 1970.
+
+\bibitem{bib3}
+Magwire MM, Bayer F, Webster CL, Cao C, Jiggins FM.
+\newblock {{S}uccessive increases in the resistance of {D}rosophila to viral
+  infection through a transposon insertion followed by a {D}uplication}.
+\newblock PLoS Genet. 2011 Oct;7(10):e1002337.
+
+\end{thebibliography}
 

### 2018-03-09 doc/turtles_plosone.tex
combined cc7g and tracg so it behaves

@@
 
-\afterpage{\clearpage}
-\begin{figure}[ht]
-  \centering
-  \includegraphics[height = 0.8\textheight, width = \textwidth, keepaspectratio = true]{figure/Ken_Ang_SpLoc_final_comp}
-  \caption{Geographic distribution of specimens sampled for comparing the hypothesized subdivisions of \textit{Emys marmorata}.
-Each hypothesized scheme has two or more possible classes.
-Sample size differs between schemes because of our ability to confidently assign museum specimens to the various schemes.
-The number of localities shown on each map is less than the number of specimens sampled because some localities produced multiple specimens.
-The different classification abbreviations are as follows: \textit{E.
-marmorata} = ``marm'', \textit{E.
-pallida} = ``pall'', Central Coast Ranges = ``CCR'', San Joaquin Valley = ``SJ,'' Baya Peninsula = ``Baja,'', and Sierra Foothills = ``Foothill.''}
-  \label{fig:map}
-\end{figure}
-
-\afterpage{\clearpage}
-\begin{figure}[ht]
-  \centering
-  \includegraphics[height = 0.5\textheight, width = \textwidth, keepaspectratio = true]{figure/plastra}
-  \caption{Depiction of general plastral shape of \textit{E. marmorata} and position of the 19 landmarks used in this study. Anterior is towards the top of the figure.}
-  \label{fig:plastra}
-\end{figure}
-
 Three different geometric morphometric datasets describing turtle plastron variation were assembled for this analysis: 1) specimens from seven distinct emydine species and a deriochelyine outgroup (\textit{C.
@@
 marmorata} dataset should be of the highest complexity and our greatest challenge given the previous finding that only very subtle differences existed between geographically-distinct populations \cite{Holland1992}.
-%We expect the first dataset represents a low-complexity dataset given the high level of plastron shape disparty that exists among these species \cite{Angielczyk2011}, whereas the second dataset should be relatively higher in complexity than the first.
-Ultimately, we expect that the third dataset to be of the highest complexity and our biggest analytical challenge.
-For all datasets we chose to focus on adults because significant changes in plastron shape occur over the course of ontogeny in \textit{E.
-marmorata} and other emydines \cite{Angielczyk2013a}.
-% three turtle datasets
-%   eight species
-%   Trachemys species
-%   Emys complex
-% 26 landmarks
 
@@
 \ref{fig:map}.
-
-We used three main binning schemes.
-All three schemes include a class for \textit{E.
-marmorata} specimens from northern populations (marm) as well as a class for those assigned to \textit{E.
-pallida} (pall) and an intergrade zone in the Central Coast Ranges (CCR).
-The schemes differ in the assignment of samples from the San Joaquin Valley (Fig.
-\ref{fig:map}).
-Scheme SP10.1 and SP10.2 differ in the assignment of specimens from the western San Joaquin Valley to either CCR or marm reflecting uncertainty regarding their genetic affinity as explained above.
-In scheme SP10.3 these specimens are assigned to a San Joaquin class reflecting the mitochondrial distinctiveness shown by Spinks and Shaffer \cite{Spinks2005}.
-We used two versions of the binning scheme based on the results of Spinks et al.
-\cite{Spinks2014}: SP14.1 reflects their phylogenetic network analysis and SP14.2 corresponds to their Bayesian species delimitation analysis.
-SP14.2 requires the addition of two new classes, ``Baja'' and ``Foothill,'' to accommodate the genetic groupings recovered by the SNP Structure analysis that was used to create the guide tree for the BPP species delimitation analysis \cite{Spinks2014}.
-Finally, we proposed a conservative morphological scheme (``Morph'') in order to compare the molecular hypotheses with something approximating the original taxonomic hypothesis for the group.
-This scheme is made up solely of the marm and pall classes from the SP10.3 scheme.
 
@@
 
+\afterpage{\clearpage}
+\begin{figure}[ht]
+  \centering
+  %\includegraphics[height = 0.8\textheight, width = \textwidth, keepaspectratio = true]{figure/Ken_Ang_SpLoc_final_comp}
+  \caption{Geographic distribution of specimens sampled for comparing the hypothesized subdivisions of \textit{Emys marmorata}.
+Each hypothesized scheme has two or more possible classes.
+Sample size differs between schemes because of our ability to confidently assign museum specimens to the various schemes.
+The number of localities shown on each map is less than the number of specimens sampled because some localities produced multiple specimens.
+The different classification abbreviations are as follows: \textit{E.
+marmorata} = ``marm'', \textit{E.
+pallida} = ``pall'', Central Coast Ranges = ``CCR'', San Joaquin Valley = ``SJ,'' Baya Peninsula = ``Baja,'', and Sierra Foothills = ``Foothill.''}
+  \label{fig:map}
+\end{figure}
+
+We used three main binning schemes.
+All three schemes include a class for \textit{E.
+marmorata} specimens from northern populations (marm) as well as a class for those assigned to \textit{E.
+pallida} (pall) and an intergrade zone in the Central Coast Ranges (CCR).
+The schemes differ in the assignment of samples from the San Joaquin Valley (Fig.
+\ref{fig:map}).
+Scheme SP10.1 and SP10.2 differ in the assignment of specimens from the western San Joaquin Valley to either CCR or marm reflecting uncertainty regarding their genetic affinity as explained above.
+In scheme SP10.3 these specimens are assigned to a San Joaquin class reflecting the mitochondrial distinctiveness shown by Spinks and Shaffer \cite{Spinks2005}.
+We used two versions of the binning scheme based on the results of Spinks et al.
+\cite{Spinks2014}: SP14.1 reflects their phylogenetic network analysis and SP14.2 corresponds to their Bayesian species delimitation analysis.
+SP14.2 requires the addition of two new classes, ``Baja'' and ``Foothill,'' to accommodate the genetic groupings recovered by the SNP Structure analysis that was used to create the guide tree for the BPP species delimitation analysis \cite{Spinks2014}.
+Finally, we proposed a conservative morphological scheme (``Morph'') in order to compare the molecular hypotheses with something approximating the original taxonomic hypothesis for the group.
+This scheme is made up solely of the marm and pall classes from the SP10.3 scheme.
+
 Sex was known only for a subset of the total dataset and was not included as a predictor of classification.
@@
 marmorata} species complex (see below).
-%The scheme names are as follows: Mito 1 and 2 correspond to \cite{Spinks2005}, Mito 3 corresponds to \cite{Spinks2010}, Morph 1 and Morph 2 correspond to \cite{Holland1992}, and Nuclear corresponds to \cite{Spinks2014}.
-% CHANGE
 
@@
 All specimens were used for superimposition, after which the subset labeled for each of the schemes were used in model training and testing (see below).
+
+\begin{figure}[ht]
+  \centering
+  %\includegraphics[height = 0.5\textheight, width = \textwidth, keepaspectratio = true]{figure/plastra}
+  \caption{Depiction of general plastral shape of \textit{E. marmorata} and position of the 19 landmarks used in this study. Anterior is towards the top of the figure.}
+  \label{fig:plastra}
+\end{figure}
 
@@
 
+Instead of relying on a single supervised learning method, we chose to use an ensemble approach where multiple model types are used in concert so that any congruence between them increases our support for that conclusion over another \cite{Hastie2009}.
+The supervised learning methods used here are named in Table \ref{tab:methods}.
+Each of these methods makes different assumptions, treats data differently, and can produce different classification results depending on the nature of the data \cite{Hastie2009}.
+For example, multinomial logistic regression is a type of generalized linear model, whereas random forest is itself an ensemble approach where multiple decision trees are fit to subsets of the full dataset and then averaged.
+
 \begin{table}
@@
 
-Instead of relying on a single supervised learning method, we chose to use an ensemble approach where multiple model types are used in concert so that any congruence between them increases our support for that conclusion over another \cite{Hastie2009}.
-The supervised learning methods used here are named in Table \ref{tab:methods}.
-Each of these methods makes different assumptions, treats data differently, and can produce different classification results depending on the nature of the data \cite{Hastie2009}.
-For example, multinomial logistic regression is a type of generalized linear model, whereas random forest is itself an ensemble approach where multiple decision trees are fit to subsets of the full dataset and then averaged.
-% use multiple methods
-%   multinomial logistic regression (nnet)
-%   linear discriminate analysis (MASS)
-%   penalized discriminate analysis (mda)
-%   neural network (nnet)
-%   random forest (randomForest)
-% each method has different assumptions and treat data differently
-%   all assume that predictors have additive effect (e.g. independent)
-
-% predictors/features/covariates
-%   1:25 PCs
-%   size
-%   size X PC1
-% analogy to PCA regression
 The maximum set of possible predictors or features used for any model of our dataset is comprised of the first 25 principal components (PCs), scaled centroid size, the interaction between scaled centroid size and PC 1, and the interaction between scaled centroid size and PC 2.
@@
 
-% training and testing paradigm
-%   training dataset
-%     for multiple models with between 3 and 28 predictors
-%     5-fold cross validation of each model
-%     best model had greatest mean AUC value
-%     selected model is most parsimonious model within 1 SD of ROC of best model
-%   testing dataset
-%     estimate class of out-of-sample testing dataset
-%     compare average AUC across models for each scheme
-% caret package for R
 We adopted a training and testing paradigm for selecting parsimonious models and estimating their overall error rates \cite{Hastie2009,Kuhn2013}.
@@
 
-% auc of roc as performance measure
-%   relationship between false positive rate and true positive rate
-%   varies between 0.5 and 1 (random -- perfect)
-%   better performance than accuracy when groups unbalanced
 In classification studies, such as this one, a common metric of performance is the receiver operating characteristic (ROC) which is the relationship between the false and true positive rates \cite{Hastie2009}.
@@
 
+The results of the PCA of plastron shape in both the eight species and \textit{Trachemys} datasets demonstrate strong association between shape and the recognized classification schemes (Fig.
+\ref{fig:other_pca}).
+
 \begin{figure}[h]
   \centering
-  \begin{subfigure}[b]{0.7\textwidth}
-    \caption{}
-    \includegraphics[width = \textwidth]{figure/cc7_pc_graph}
-  \end{subfigure}
-
-  \begin{subfigure}[b]{0.7\textwidth}
-    \caption{}
-    \includegraphics[width = \textwidth]{figure/tra_pc_graph}
-  \end{subfigure}
+  %\includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/other_pc_graph}
   \caption{Two scatterplots of morphological differences from two of the three datasets analyzed in this study.
@@
 
-\afterpage{\clearpage}
+The results of the PCA of plastron shape in the \textit{Emys marmorata} dataset show no clear connection between plastron shape and any of the proposed classification schemes (Fig.
+\ref{fig:emys_pca}).
+The first PC axis of shape variation appears to be primarily structured by differences in individual centroid size (Fig.
+\ref{fig:emys_pca}); a linear regression of PC 1 vs log centroid size shows that size explains approximately 81 percent of the variation along this axis.
+This was the motivation for including centroid size and its interaction with PC1 as predictors in all of the supervised learning models.
+Size typically explains less than three percent of the variance of the remaining PC axes (e.g.
+2.5\% on PC 2).
+
 \begin{figure}[ht]
   \centering
-  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/emys_pc_graph}
+  %\includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/emys_pc_graph}
   \caption{Scatterplot of the first two axes of morphological variation in the \textit{Emys marmorata} dataset.
@@
 
-The results of the PCA of plastron shape in both the eight species and \textit{Trachemys} datasets demonstrate strong association between shape and the recognized classification schemes (Fig.
-\ref{fig:other_pca}).
-
-The results of the PCA of plastron shape in the \textit{Emys marmorata} dataset show no clear connection between plastron shape and any of the proposed classification schemes (Fig.
-\ref{fig:emys_pca}).
-The first PC axis of shape variation appears to be primarily structured by differences in individual centroid size (Fig.
-\ref{fig:emys_pca}); a linear regression of PC 1 vs log centroid size shows that size explains approximately 81 percent of the variation along this axis.
-This was the motivation for including centroid size and its interaction with PC1 as predictors in all of the supervised learning models.
-Size typically explains less than three percent of the variance of the remaining PC axes (e.g.
-2.5\% on PC 2).
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[height = 0.3\textheight, width = \textwidth, keepaspectratio = true]{figure/sex_test_hist}
-  \caption{Comparison of observed Procrustes distance between the centroids of each sex (vertical line) to a null distribution generated from 1000 permutations of the sex-labels.
-This result indicates that the difference between the centroids is as small/smaller than expected by random.}
-  \label{fig:sex_test}
-\end{figure}
-
-\afterpage{\clearpage}
-\begin{figure}[h]
-  \centering
-  \includegraphics[height = 0.8\textheight, width = \textwidth, keepaspectratio = true]{figure/sex_test_hist_grouped}
-  \caption{Comparison of observed Procrustes distance between the centroids of each sex (vertical line) to a null distribution generated from 1000 permutations of the sex-labels for each of the six classification schemes.
-This result indicates that the difference between the centroids is as small/smaller than expected by random.
-For explination of classification scheme names see Table \ref{tab:hypotheses}.}
-  \label{fig:sex_test_group}
-\end{figure}
-
 Analysis of the differences between sexes of \textit{E.
@@
 \ref{fig:sex_test_group}).
+
+\begin{figure}[h]
+  \centering
+  %\includegraphics[height = 0.3\textheight, width = \textwidth, keepaspectratio = true]{figure/sex_test_hist}
+  \caption{Comparison of observed Procrustes distance between the centroids of each sex (vertical line) to a null distribution generated from 1000 permutations of the sex-labels.
+This result indicates that the difference between the centroids is as small/smaller than expected by random.}
+  \label{fig:sex_test}
+\end{figure}
+
+\begin{figure}[h]
+  \centering
+  %\includegraphics[height = 0.8\textheight, width = \textwidth, keepaspectratio = true]{figure/sex_test_hist_grouped}
+  \caption{Comparison of observed Procrustes distance between the centroids of each sex (vertical line) to a null distribution generated from 1000 permutations of the sex-labels for each of the six classification schemes.
+This result indicates that the difference between the centroids is as small/smaller than expected by random.
+For explination of classification scheme names see Table \ref{tab:hypotheses}.}
+  \label{fig:sex_test_group}
+\end{figure}
+
+Comparison of the within to between Procrustes distances of the digitization replicates gives an approximate estimate of the error between distinct groupings (Table \ref{tab:rep_res}).
+The ratio of the average within-individual distance to the average distance between individuals for the replicated datasets is 1.11; this indicates that the grouping is slightly counter-intuitive to the data and is consistent with all shapes being very similar regardless of individual identity.
+This value also provides a baseline by which to understand how distinct the groupings are, where other ratios are compared to the correction ratio \(1.11/1\).
 
@@
 
-Comparison of the within to between Procrustes distances of the digitization replicates gives an approximate estimate of the error between distinct groupings (Table \ref{tab:rep_res}).
-The ratio of the average within-individual distance to the average distance between individuals for the replicated datasets is 1.11; this indicates that the grouping is slightly counter-intuitive to the data and is consistent with all shapes being very similar regardless of individual identity.
-This value also provides a baseline by which to understand how distinct the groupings are, where other ratios are compared to the correction ratio \(1.11/1\).
-
 The results from the eight species and \textit{Trachemys} datasets indicate that both of these classification schemes are more recognizable than not given our estimate of digitization error (Table \ref{tab:rep_res}).
@@
 
+Analysis of the eight morphologically- and genetically-distinct species and the \textit{T.
+scripta scripta}--\textit{T.
+scripta elegans} datasets indicate that these taxa are sufficiently morphologically distinct to be differentiated on the basis of plastron shape.
+Both in-sample and out-of-sample classifications have AUC values of approximately 1 for all methods, implying near-perfect classification rates (Fig.
+\ref{fig:other_sel}, \ref{fig:other_oos}).
+For both datasets, the ROC scores from testing datasets are tightly clustered near AUC = 1 (Fig.
+\ref{fig:other_oos}).
+These results demonstrate that when there are distinctions between the states of the classification schemes (i.e., differences in plastron shape that correlate with the different taxonomic groups), the methods used here can recover them.
+
 \begin{figure}[h]
   \centering
-  \includegraphics[height = 0.5\textheight, width = \textwidth, keepaspectratio = true]{figure/other_model_sel}
+  %\includegraphics[height = 0.5\textheight, width = \textwidth, keepaspectratio = true]{figure/other_model_sel}
   \caption{Comparisons of model fit to the training dataset for each of the supervised learning methods applied to the first two datasets; the results from the eight species dataset are presented in the left column, while those from the \textit{Trachemys} dataset are presented in the right column.
@@
   \centering
-  \includegraphics[height = 0.3\textheight, width = \textwidth, keepaspectratio = true]{figure/other_oos_sel}
+  %\includegraphics[height = 0.3\textheight, width = \textwidth, keepaspectratio = true]{figure/other_oos_sel}
   \caption{The results of out-of-sample predictive performance of the selected models for both the eight species (left) and \textit{Trachemys} datasets.
@@
 
-Analysis of the eight morphologically- and genetically-distinct species and the \textit{T.
-scripta scripta}--\textit{T.
-scripta elegans} datasets indicate that these taxa are sufficiently morphologically distinct to be differentiated on the basis of plastron shape.
-Both in-sample and out-of-sample classifications have AUC values of approximately 1 for all methods, implying near-perfect classification rates (Fig.
-\ref{fig:other_sel}, \ref{fig:other_oos}).
-For both datasets, the ROC scores from testing datasets are tightly clustered near AUC = 1 (Fig.
-\ref{fig:other_oos}).
-These results demonstrate that when there are distinctions between the states of the classification schemes (i.e., differences in plastron shape that correlate with the different taxonomic groups), the methods used here can recover them.
-
-\afterpage{\clearpage}
+AUC--based model selection revealed some important patterns of variation and congruence between the classification schemes and the actual data.
+Generally, the best performing models tended to include about half the total number of possible PCs (Fig.
+\ref{fig:emys_sel}).
+
 \begin{figure}[ht]
   \centering
-  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/emys_model_sel}
+  %\includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/emys_model_sel}
   \caption{AUC values for models of varying complexity fit to the \textit{Emys marmorata} training datasets for each classification scheme.
@@
 
-\afterpage{\clearpage}
-\begin{figure}[ht]
-  \centering
-  \includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/emys_oos_sel}
-  \caption{Comparison of out-of-sample AUC estimates from the predictions of selected models (Fig.
-\ref{fig:emys_sel}), grouped by classification scheme.
-The horizontal line in each panel corresponds to the average AUC value across all models of that classification scheme.}
-  \label{fig:emys_oos}
-\end{figure}
-
-AUC--based model selection revealed some important patterns of variation and congruence between the classification schemes and the actual data.
-Generally, the best performing models tended to include about half the total number of possible PCs (Fig.
-\ref{fig:emys_sel}).
-
 Observed AUC values for all of the optimal models are lower for the \textit{E.
@@
 \ref{fig:emys_oos}).
+
+\begin{figure}[ht]
+  \centering
+  %\includegraphics[height = \textheight, width = \textwidth, keepaspectratio = true]{figure/emys_oos_sel}
+  \caption{Comparison of out-of-sample AUC estimates from the predictions of selected models (Fig.
+\ref{fig:emys_sel}), grouped by classification scheme.
+The horizontal line in each panel corresponds to the average AUC value across all models of that classification scheme.}
+  \label{fig:emys_oos}
+\end{figure}
 
@@
 
-\section*{Supporting information}
-
-% Include only the SI item label in the paragraph heading. Use the \nameref{label} command to cite SI items in the text.
-\paragraph*{S1 Fig.}
-\label{S1_Fig}
-{\bf Bold the title sentence.} Add descriptive text after the title of the item (optional).
+%\section*{Supporting information}
+%
+%% Include only the SI item label in the paragraph heading. Use the \nameref{label} command to cite SI items in the text.
+%\paragraph*{S1 Fig.}
+%\label{S1_Fig}
+%{\bf Bold the title sentence.} Add descriptive text after the title of the item (optional).
 

### 2018-03-09 R/visual_class.r

@@ -95,19 +95,33 @@ splits <- split(scores.df, scores.df$who)
 
 # turt.proc$percent[1:2]  # percent of variation on PC
 
+splits[[1]]$g <- 'cc7'
+splits[[2]]$g <- 'tra'
+
+#splits <- dplyr::bind_rows(splits[[1]], splits[[2]], .id = 'g')
+
+
 clear.gg <- ggplot(splits[[1]], aes(x = PC1, y = PC2, 
                                     colour = species, size = centroid))
 clear.gg <- clear.gg + geom_point(alpha = 0.5)
 clear.gg <- clear.gg + scale_size_area(max_size = 3)
 clear.gg <- clear.gg + scale_colour_manual(values = cbp.ord)
+#clear.gg <- clear.gg + facet_wrap(~ g)
 
+cc7g <- clear.gg + labs(x = 'PC 1 (41.6%)', 
+                        y = 'PC 2 (19.1%)', 
+                        title = 'A.')
+cc7g <- cc7g + coord_fixed(ratio = 1)
+trag <- clear.gg %+% splits[[2]] + labs(x = 'PC 1 (37.9%)', 
+                                        y = 'PC 2 (17.3%)', 
+                                        title = 'B.')
+trag <- trag + coord_fixed(ratio = 1)
 
-cc7g <- clear.gg + labs(x = 'PC 1 (41.6%)', y = 'PC 2 (19.1%)')
-cc7g <- cc7g + coord_fixed(ratio = 1) + 
-  theme(plot.margin = unit(c(0, 0, 0, 0), 'mm'))
-trag <- clear.gg %+% splits[[2]] + labs(x = 'PC 1 (37.9%)', y = 'PC 2 (17.3%)')
-trag <- trag + coord_fixed(ratio = 1) +
-  theme(plot.margin = unit(c(0, 0, 0, 0), 'mm'))
+png(filename = '../doc/figure/other_pc_graph.png',
+    width = 850, height = 450)
+grid.arrange(cc7g, trag, ncol = 2)
+dev.off()
+
 
 ggsave(plot = cc7g, filename = '../doc/figure/cc7_pc_graph.png',
        width = 8, height = 6)

### 2019-02-13 doc/letter_editor.tex
needed to update ignore and restage

@@
 \begin{document}
-\begin{letter}{Editor \\ \textit{PeerJ}}
+\begin{letter}{Editor \\ \textit{PLoS One}}
   \opening{Dear Editor,}
 
-  Please find enclosed the manuscript entitled ``Ensemble approaches for estimating congruence between species delimitation and morphological variation: comparing taxonomic hypotheses for the Pacific Pond Turtle (\textit{Emys marmorata})'' which we are submitting for consideration for the \textit{PeerJ}.
+  Please find enclosed the manuscript entitled ``Ensemble approaches for estimating congruence between species delimitation and morphological variation: a case study using the Pacific Pond Turtle (\textit{Emys marmorata})'' which we are submitting for consideration for the \textit{PLoS One}.
 This manuscript is authored by myself (Peter D Smits), Kenneth D Angielczyk, James F Parham, and Bryan L Stuart.

### 2019-02-13 R/helper01_align_plot.r

@@ -1,6 +1,6 @@
-library(ggplot2)
-library(gtable)
-library(gridExtra)
+library(pacman)
+
+p_load(ggplot2, gtable, gridExtra)
 
 AlignPlots <- function(...) {
   LegendWidth <- function(x) x$grobs[[8]]$grobs[[1]]$widths[[4]]

### 2019-02-13 R/helper05_multiclass_roc.r

@@ -4,10 +4,11 @@
 #
 #
 # also based on caret::twoClassSummary
+library(pacman)
+
+p_load(pROC, Metrics, caret)
+
 multiClassSummary <- function(data, lev = NULL, model = NULL) {
-  require(pROC)
-  require(Metrics)
-  require(caret)
 
   #check data
   if(!all(levels(data[, 'pred']) == levels(data[, 'obs']))) {
